<!-- Google tag (gtag.js) -->
<script async src="https://www.googletagmanager.com/gtag/js?id=G-7PRVEBE1EF"></script>
<script>
  window.dataLayer = window.dataLayer || [];
  function gtag(){dataLayer.push(arguments);}
  gtag('js', new Date());

  gtag('config', 'G-7PRVEBE1EF');
</script>

# Soup-erb Beta Regression {#sec-beta}

```{r}
#| include: false

colourize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}
```

```{r setup, include=FALSE}
# Load required libraries
library(devtools)
library(readr)
library(ggplot2)
library(ggcorrplot)
library(cookbook)
library(tidyverse)
library(car)
library(lmtest)
library(fastDummies)

# Setting up Python dependencies
library(reticulate)
py_require("pandas") 
py_require("numpy")
py_require("matplotlib")
py_require("statsmodels")
py_require("pyreadr")
py_require("scikit-learn")
py_require("seaborn")
py_require("scipy")

pandas <- import("pandas")
numpy <- import("numpy")
matplotlib <- import("matplotlib")
statsmodels <- import("statsmodels")
pyreadr <- import("pyreadr")
sklearn <- import("sklearn")
seaborn <- import("seaborn")
scipy <- import("scipy")

# Load data in R
data <- beta_data

py$data <- r_to_py(beta_data)
```

::: {.LO}
::::{.LO-header}
Learning Objectives
::::
::::{.LO-container}
By the end of this chapter, you will be able to:

::::
:::

## Introduction

Imagine you are studying the conversion rate of a website. For each page, you record the proportion of visitors who complete a purchase. Some pages convert only a small fraction of visitors, while others perform much better. Your goal is to understand how features such as page load time, device type, or traffic source affect this conversion rate. The outcome you are modeling is numerical and continuous, but it also has clear limits. A conversion rate can never be less than zero and can never be greater than one. Standard linear regression is not well suited for this situation. It can produce predictions outside this range and assumes that the amount of variability is the same for all pages. Problems like this, where the response represents a proportion between zero and one, require a regression method that respects both the scale of the data and how the variability changes across observations. This is exactly where Beta regression comes to shine.





::: {.Warning}
::::{.Warning-header}
Fun fact!
::::
::::{.Warning-container}
**Soup-erb!** Soup that’s so heartwarming it feels like a cozy hug.
::::
:::

::: {#fig-beta-regression}
```{mermaid}
mindmap
  root((Regression 
  Analysis)
    Continuous <br/>Outcome Y
      {{Unbounded <br/>Outcome Y}}
        )Chapter 3: <br/>Ordinary <br/>Least Squares <br/>Regression(
          (Normal <br/>Outcome Y)
      {{Nonnegative <br/>Outcome Y}}
        )Chapter 4: <br/>Gamma Regression(
          (Gamma <br/>Outcome Y)
      {{Bounded <br/>Outcome Y <br/> between 0 and 1}}
        )Chapter 5: Beta <br/>Regression(
          (Beta <br/>Outcome Y)
    Discrete <br/>Outcome Y
```
:::
