<!-- Google tag (gtag.js) -->
<script async src="https://www.googletagmanager.com/gtag/js?id=G-7PRVEBE1EF"></script>
<script>
  window.dataLayer = window.dataLayer || [];
  function gtag(){dataLayer.push(arguments);}
  gtag('js', new Date());

  gtag('config', 'G-7PRVEBE1EF');
</script>

# Beta Regression {#sec-beta}

```{r}
#| include: false

colourize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}
```

```{r setup, include=FALSE}
# Load required libraries
library(devtools)
library(readr)
library(ggplot2)
library(ggcorrplot)
library(cookbook)
library(tidyverse)
library(car)
library(lmtest)
library(fastDummies)
library(betareg)

# Setting up Python dependencies
library(reticulate)

py_require("pandas") 
py_require("numpy")
py_require("matplotlib")
py_require("statsmodels")
py_require("pyreadr")
py_require("scikit-learn")
py_require("seaborn")
py_require("scipy")

pandas <- import("pandas")
numpy <- import("numpy")
matplotlib <- import("matplotlib")
statsmodels <- import("statsmodels")
pyreadr <- import("pyreadr")
sklearn <- import("sklearn")
seaborn <- import("seaborn")
scipy <- import("scipy")

# Load data in R
data <- Beta

py$data <- r_to_py(Beta)
```

::: {.LO}
::::{.LO-header}
Learning Objectives
::::
::::{.LO-container}
By the end of this chapter, you will be able to:

- Explain what **Beta regression** is and identify when it is the appropriate modelling choice for proportion data.
- Describe why **Ordinary Least Squares** is not suitable for proportion outcomes and how Beta regression addresses these limitations.
- Describe the **Beta distribution** and how its shape flexibility allows it to represent a wide variety of proportion data.
- Understand the role of the **logit link function** in Beta regression and how it connects the linear predictor to the mean proportion.
- Fit a Beta regression model using `R` or `Python` and interpret coefficients as **odds ratios** for the mean proportion.
- Evaluate model fit using **log-likelihood**, **AIC**, the **precision parameter**, and **residual diagnostics**.

::::
:::

## Introduction

Many real world problems involve modelling proportions. A common example is a website conversion rate, where each page is associated with a value between zero and one that represents the fraction of visitors who complete a purchase. Some pages convert only a small share of visitors, while others perform much better. Although these values are numerical and continuous, they are also restricted to a fixed range. A suitable model must respect these limits and reflect the fact that variability often changes depending on the average level of the proportion.

Traditional linear regression is not well suited for this type of data. It produces predictions outside the valid range and assumes that variability is constant across all observations. Problems like this call for a regression approach that is designed specifically for proportion outcomes. This is exactly where Beta regression comes to shine.

## What is Beta Regression

Beta regression is a regression model for response variables that are proportions, meaning they take values strictly between zero and one. It works by assuming the response follows a Beta distribution and modelling how the expected value of that distribution changes with the predictors.

A key advantage of this approach is that the Beta distribution can take many different shapes depending on the data. It can be symmetric, skewed to the left or right, or concentrated near the boundaries. This shape flexibility means the model can fit a wide variety of proportion data without needing to force it into a rigid template.

## The Beta Distribution

The Beta distribution is a continuous probability distribution defined on the interval between zero and one. A key feature of this distribution is its flexibility. Depending on its parameter values, it can take many shapes, including symmetric, left skewed, or right skewed forms, as well as distributions that place more mass near the boundaries.

```{python}
#| label: beta-distribution
#| echo: false
#| message: false
#| results: hide

import numpy as np
import matplotlib.pyplot as plt
from scipy.stats import beta

x = np.linspace(0.001, 0.999, 500)

parameters = [
    (0.5, 0.5),
    (2, 2),
    (2, 5),
    (5, 2),
    (1, 3),
    (3, 1),
]

fig, ax = plt.subplots(figsize=(9, 5))

for a, b in parameters:
    y = beta.pdf(x, a, b)
    ax.plot(x, y, label=f"alpha = {a}, beta = {b}")

ax.set_title("Examples of Beta Distributions on (0, 1)")
ax.set_xlabel("Proportion")
ax.set_ylabel("Density")
ax.set_xlim(0, 1)
ax.legend(frameon=False)

fig.tight_layout()
fig
```

The figure above shows several examples of Beta distributions with different parameter choices. Although the shapes vary widely, all distributions remain within the zero to one range, making the Beta distribution a natural choice for modelling proportion data.

The shape of a Beta distribution is controlled by two parameters, commonly called **alpha** and **beta**. When alpha is larger than beta, values tend to be closer to one. When beta is larger than alpha, values tend to be closer to zero. When the two parameters are similar, values cluster around the middle of the range. Together, alpha and beta also influence how spread out the data are, with smaller values producing more variability and larger values leading to tighter concentration around the average.

In practice, working directly with alpha and beta is not very convenient for regression modelling. These parameters describe the shape of the distribution, but they do not directly tell us how the average proportion changes when a predictor changes. To address this, Beta regression rewrites the distribution in terms of quantities that are easier to interpret, such as the **average proportion** and **how tightly the data cluster around that average**. This reexpression allows the model to relate the average proportion to predictor variables while still ensuring that all fitted values remain between zero and one. We explore this idea in more detail in the next section.

## Linking the Average Proportion to Predictors

In the [Gamma regression chapter](04-gamma.qmd), we introduced generalized linear models as a way to extend linear regression to response variables that do not follow a normal distribution. Beta regression follows the same framework and is another example of a generalized linear model, designed specifically for proportion data.

As with other generalized linear models, Beta regression does not model the response directly. Instead, it models the average of the response through a transformation that allows predictor variables to be related to the outcome using a familiar linear structure. This approach is especially important for proportion data, since the average must always remain between zero and one.

::: {.Tip}
::::{.Tip-header}
A Quick Refresh on GLMs
::::
::::{.Tip-container}
Generalized linear models are built from three components:

- **Random component** describes the distribution of the response variable  
- **Systematic component** combines predictor variables through a linear predictor  
- **Link function** connects the linear predictor to the average of the response  

These three pieces work together to extend linear regression to a wide range of response types.
::::
:::

For Beta regression, the three components of GLM take the following form:

- **Random component**: The response variable is assumed to follow a Beta distribution. This choice reflects the fact that the outcome is continuous and bounded between zero and one.

- **Systematic component**: Predictor variables are combined through a linear predictor. This part of the model describes how changes in the predictors are associated with changes in the average proportion.

- **Link function**: The link function connects the linear predictor to the average proportion. Its role is to ensure that the modeled average remains between zero and one. A commonly used choice is the logit function, which maps proportions to the real line and is also used in logistic regression.

## The Beta Regression Model

Putting the three components together, the Beta regression model can be written in mathematical form.

For each observation $i$, the response variable is assumed to follow a Beta distribution,

$$
Y_i \sim \text{Beta}(\mu_i \phi,\ (1 - \mu_i)\phi)
$$
Here, $\mu_i$ represents the average proportion for observation $i$, and $\phi$ is a parameter that controls how tightly the data cluster around that average. Larger values of $\phi$ correspond to less variability, while smaller values allow for more spread.

The average proportion $\mu_i$ is then linked to the predictors through a linear predictor using a link function. With the logit link, this takes the form

$$
\begin{aligned}
\text{logit}(\mu_i) &= \log\left(\frac{\mu_i}{1-\mu_i}\right) \\
                    &= \eta_i \\
                    &= \beta_0 + \beta_1 x_{i1} + \cdots + \beta_p x_{ip}
\end{aligned}
$$

Here, $\beta_0$ is the intercept and represents the baseline value of the transformed average proportion when all predictors are set to zero. The remaining coefficients $\beta_1, ..., \beta_p$ describe the association between each predictor and the average proportion on the logit scale, holding all other predictors constant. A positive coefficient indicates that larger values of the predictor are associated with a higher average proportion, while a negative coefficient indicates the opposite.

This formulation follows the same generalized linear model structure introduced in the [Gamma regression chapter](04-gamma.qmd). The random component specifies the distribution of the response, the systematic component combines predictors through a linear predictor, and the link function connects the linear predictor to the average of the response. Together, these components allow Beta regression to model proportion data in a flexible and interpretable way, while ensuring that fitted values always remain between zero and one.

## Case Study: Modeling Student Academic Performance

To demonstrate Beta regression in action, we will walk through a case study using a toy dataset that captures student academic performance. In this case study, our response variable is a final course grade, expressed as a percentage.

Because grades represent proportions that lie between zero and one hundred percent, they are naturally suited for Beta regression after being rescaled to the unit interval. Our goal is to understand how different aspects of a student’s academic life and daily habits are associated with their overall performance.

We will approach this case study using the data science workflow introduced in the [first chapter](01-intro.qmd). This ensures a structured approach to framing the problem, exploring the data, selecting an appropriate model, and interpreting the results.

### The Dataset

The toy dataset used in this case study represents a group of university students. Each row corresponds to a single student and includes information about study habits, lifestyle factors, and academic support. Below is a summary of the variables included in the dataset.

| **Variable Name**             | **Description**                                          |
| ----------------------------- | -------------------------------------------------------- |
| **Final Grade (%)**           | Final course grade expressed as a percentage.            |
| **Café Location**             | Primary café or study location used by the student.      |
| **Study Time**                | Self-reported amount of time spent studying.             |
| **Year in School**            | Academic year of the student (e.g., 1st Year, 4th Year). |
| **Hobbies**                   | Primary hobby outside of school.                         |
| **Primary Sport**             | Main sport played by the student, if any.                |
| **Living Arrangement**        | Living situation (e.g., alone, with friends).            |
| **Part-time Work Hours**      | Number of hours worked per week in a part-time job.      |
| **Study Group Participation** | Whether the student participates in a study group.       |
| **Hours of Sleep per Night**  | Average number of hours of sleep per night.              |
| **Has a Tutor**               | Whether the student receives tutoring support.           |


Here is a snapshot of the dataset:

<div style="overflow-x: auto;">

| **Final Grade (%)** | **Café Location**       | **Study Time** | **Year in School** | **Hobbies** | **Primary Sport** | **Living Arrangement** | **Part-time Work Hours** | **Study Group Participation** | **Hours of Sleep per Night** | **Has a Tutor** |
|---------------------|--------------------------|----------------|---------------------|-------------|-------------------|------------------------|---------------------------|-------------------------------|------------------------------|------------------|
| 96.99               | Koerner's Pub            | 5+ hours       | 1st Year            | Sports      | Soccer            | With Friends           | 15                        | Yes                           | 5.4                          | Yes              |
| 94.62               | Bean Around the World    | 5+ hours       | 4th Year            | Gaming      | Soccer            | Alone                  | 7                         | Yes                           | 5.4                          | No               |
| 90.30               | The Loop Café            | 5+ hours       | 2nd Year            | Sports      | None              | Alone                  | 6                         | No                            | 5.6                          | Yes              |
| 86.77               | Bean Around the World    | 5+ hours       | 1st Year            | Reading     | Basketball        | Alone                  | 10                        | No                            | 7.8                          | No               |

</div>


The dataset includes students across multiple academic years and a variety of living and working arrangements, giving a realistic cross-section of student life.

### The Problem We’re Trying to Solve

In this case study, our goal is to understand what factors are associated with higher or lower final grades. Since grades are continuous, bounded, and naturally interpreted as proportions, Beta regression is a suitable modeling choice.

The key question we aim to answer is:

> **Which academic and lifestyle factors are associated with differences in students’ final grades?**

Rather than focusing only on prediction, we are interested in understanding how study habits, support systems, and daily routines relate to academic outcomes. In the next section, we clarify our study design and begin applying the data science workflow to this problem.

## Study Design

Knowing what we want to learn, the next step is to be clear about the type of question we are asking.

This is an inferential task. We are interested in describing and interpreting relationships between variables, not building a model optimized for prediction. A predictive approach would focus on forecasting a specific student’s expected grade as accurately as possible. Here, we care more about which factors relate to performance and in what direction.

It is also worth noting that this is an observational dataset. Students were not randomly assigned to different study habits or support conditions, so the relationships we find are associational rather than causal. Observing that higher study time tends to accompany higher grades does not rule out the possibility that a third factor, such as student motivation, influences both.

## Data Collection and Wrangling

With our statistical question clearly defined, the next step is to ensure that the data is suitable for analysis. Although the dataset is already provided, it is still useful to reflect briefly on how this type of data might have been collected. Doing so helps clarify what the data represents and highlights potential limitations before moving on to preparation and modelling.

### Data Collection

For a study examining student academic performance, data like this could reasonably be collected through several common sources:

- **Academic Records**: Final grades are typically recorded by instructors or academic departments at the end of a course. These records provide a standardized and reliable measure of student performance.

- **Student Surveys**: Information on study habits, sleep, part-time work, tutoring, and study group participation is often collected through self-reported surveys. While this data provides valuable context, it may be subject to reporting bias or measurement error.

- **Institutional Data Systems**: Universities often combine academic records with student support and enrollment information, such as year in school or participation in academic programs.

Each of these sources comes with trade-offs. Academic records are objective but limited to performance outcomes, while survey-based variables may be noisy but offer insight into behaviors and lifestyle factors. Recognizing these limitations helps guide both exploratory analysis and interpretation later on.

### Data Wrangling

Once the data has been collected, the next step is to prepare it for modelling. Even small, well-structured datasets often require preprocessing before they can be analyzed effectively.

For Beta regression, data wrangling typically involves:

- Validating that the response variable represents a proportion.
- Rescaling the response variable so that it lies strictly between zero and one.
- Checking for boundary values such as exact zeros or ones.
- Converting categorical variables into numerical formats suitable for modelling.
- Inspecting missing values and potential inconsistencies in self-reported data.

Based on the data, there can, of course, be other things you may have to do but for now, let's walk through these steps using our student performance dataset, demonstrating each step in both R and Python.

#### 1. Preparing and Validating the Response Variable

In this case study, the response variable is the final course grade, recorded as a percentage. While this variable is continuous, it is currently measured on a scale from 0 to 100. Since Beta regression models proportions, the response must first be rescaled to lie between zero and one.

Before rescaling, we begin by examining the distribution of final grades to confirm that the values are within a reasonable range.

::: {.panel-tabset}
## R Code
```{r}
summary(data$`Final Grade (%)`)
```

## Python Code
```{python}
# Summary statistics for the response variable
print(data.columns.tolist())
print(data["Final Grade (%)"].describe())
```
:::

The summary confirms that all grades lie between 0 and 100, making them suitable for conversion to proportions. We therefore rescale the response by dividing each value by 100.

::: {.panel-tabset}
## R Code
```{r}
data$final_grade_prop <- data$`Final Grade (%)` / 100

# Drop original percentage column
data$`Final Grade (%)` <- NULL

summary(data$final_grade_prop)
```

## Python Code
```{python}
# Rescale grades to proportions
data["final_grade_prop"] = data["Final Grade (%)"] / 100

# Drop original percentage column
data = data.drop(columns=["Final Grade (%)"])

print(data["final_grade_prop"].describe())
```
:::

::: {.Heads-up}
::::{.Heads-up-header}
Heads-up!
::::
::::{.Heads-up-container}
At this stage, the structure of the response variable already suggests that a proportion-based model, such as Beta regression, will be a good fit. With that in mind, we perform the preprocessing needed for such a model here. In practice, this decision is often informed by exploratory analysis and may be revisited later, but introducing it now helps keep the workflow clear and focused.
::::
:::

After rescaling, the response variable now lies on the unit interval. The next step is to check whether any values are exactly zero or one, since standard Beta regression assumes outcomes are strictly between these boundaries.

#### 2. Checking for Boundary Values

With the rescaling done, we can now check whether any values land exactly on these boundaries.

::: {.panel-tabset}
## R Code
```{r}
sum(data$final_grade_prop <= 0)
sum(data$final_grade_prop >= 1)
```

## Python Code
```{python}
# Check for boundary values
print((data["final_grade_prop"] <= 0).sum())
print((data["final_grade_prop"] >= 1).sum())
```
:::

Oh no! The checks reveal that there are several observations with values equal to one. While this is not unusual for percentage-based data, it poses a problem for standard Beta regression, which assumes the response lies strictly between zero and one.

::: {.Heads-up}
::::{.Heads-up-header}
Heads-up!
::::
::::{.Heads-up-container}
When proportion data includes values exactly equal to zero or one, the assumptions of standard Beta regression are violated. This situation is common in practice, especially when proportions are derived from percentages or rounded measurements. There are several principled ways to handle boundary values, each with different trade-offs.

One approach is **boundary adjustment**, where values equal to zero or one are shifted slightly inward by a small amount. This method is simple, preserves the ordering of observations, and allows standard Beta regression to be applied without adding model complexity. It is often appropriate when boundary values arise from rounding or measurement limits rather than representing a fundamentally different process.

A second approach, proposed by Smithson and Verkuilen (2006), **applies a systematic transformation to the response variable**. Instead of only adjusting values at the boundaries, this method gently shifts all observations so that none of them are exactly zero or one.

Intuitively, the transformation treats the observed proportions as slightly uncertain and pulls extreme values a small distance toward the interior of the interval. This avoids hard cutoffs at zero and one while preserving the relative ordering of the data.

Mathematically, the transformation can be written as
$$
x' = \frac{x (N - 1) + s}{N}
$$

where $N$ is the sample size and $s$ is a small constant between 0 and 1, often chosen as $s=0.5$.

One way to think about this adjustment is that it spreads the data evenly across the open interval $(0,1)$, ensuring that no observation lies exactly on the boundaries. From a Bayesian perspective (for the Bayesianists out there), the constant $s$ can be interpreted as introducing a weak prior that prevents extreme values from being taken at face value. In practice, this approach is often viewed as more principled than manually nudging boundary values, though it also adds an extra layer of abstraction. It is most commonly used when the researcher wants to apply a consistent assumption across all observations, rather than singling out only the boundary cases for special treatment.

Another option is to use **zero–one–inflated Beta models**, which explicitly treat boundary values as arising from a separate process. These models are more flexible and appropriate when zeros or ones occur frequently and carry substantive meaning, but they also introduce additional parameters and complexity.

Finally, boundary observations can be **excluded**, though this is generally discouraged unless there is a strong substantive justification, as it removes potentially informative data.
::::
:::

In this case study, our goal is to focus on the core ideas of Beta regression and to keep the workflow accessible. Because the number of boundary values is small and likely driven by rounding of high grades, we proceed with a simple boundary adjustment and continue with standard Beta regression.

::: {.panel-tabset}
## R Code
```{r}
epsilon <- 0.001
data$final_grade_prop[data$final_grade_prop >= 1] <- 1 - epsilon
data$final_grade_prop[data$final_grade_prop <= 0] <- epsilon

summary(data$final_grade_prop)
```

## Python Code
```{python}
epsilon = 0.001
data.loc[data["final_grade_prop"] >= 1, "final_grade_prop"] = 1 - epsilon
data.loc[data["final_grade_prop"] <= 0, "final_grade_prop"] = epsilon

print(data["final_grade_prop"].describe())
```
:::

After this adjustment, all response values lie strictly between zero and one, making the data suitable for Beta regression.

#### 3. Encoding Categorical Variables

Several variables in the dataset are categorical, such as café location, year in school, hobbies, living arrangement, study group participation, and tutoring status. Regression models, including Beta regression, require numerical inputs, so these variables must be converted into a suitable numerical representation. We use one-hot encoding for this purpose.

As a quick refresh, one-hot encoding represents each category as a binary indicator variable. For a given categorical feature, each level becomes its own column that takes the value 1 when an observation belongs to that category and 0 otherwise. To avoid redundancy, one category is typically omitted and treated as the reference level.

::: {.panel-tabset}
## R Code
```{r}
# Convert categorical variables to factors
categorical_vars <- c(
  "Café Location",
  "Study Time",
  "Year in School",
  "Hobbies",
  "Primary Sport",
  "Living Arrangement",
  "Study Group Participation",
  "Has a Tutor"
)

data[categorical_vars] <- lapply(data[categorical_vars], as.factor)

# Create design matrix with dummy variables
design_matrix <- model.matrix(
  final_grade_prop ~ .,
  data = data
)

head(design_matrix)
```

## Python Code
```{python}
import pandas as pd

categorical_vars = [
  "Café Location",
  "Study Time",
  "Year in School",
  "Hobbies",
  "Primary Sport",
  "Living Arrangement",
  "Study Group Participation",
  "Has a Tutor"
]

s = data["Study Time"].astype("string")

# Replace en dash/em dash/minus with normal hyphen
s = s.str.replace(r"[–—−]", "-", regex=True).str.strip()

study_time_order = ["<1 hour", "1-3 hours", "3-5 hours", "5+ hours"]

data["Study Time"] = pd.Categorical(
    s,
    categories=study_time_order,
    ordered=True
)

# One-hot encode
data_encoded = pd.get_dummies(
    data,
    columns=categorical_vars,
    drop_first=True,
    dtype=float  # or dtype=int
)

data_encoded.head()
```
:::

After encoding, each categorical variable is represented numerically, making the dataset suitable for regression modeling. The resulting design matrix includes one column per predictor, with categorical effects interpreted relative to a reference category. We return to this point when interpreting the fitted model.

#### 4. Checking for Missing Values and Inconsistencies

A useful step before any analysis is to check for missing data. Missing values can affect model fitting and interpretation, especially when the dataset is small.

::: {.panel-tabset}
## R Code
```{r}
# Check for missing values in each column
colSums(is.na(data))
```

## Python Code
```{python}
# Check for missing values in each column
print(data.isna().sum())
```
:::

Fortunately, there are no missing values in this dataset. If missing values were present, several options would be available, including removing affected observations, imputing values, or revisiting how the data was collected. The appropriate choice depends on the extent and nature of the missingness, as well as the goals of the analysis.

::: {.Heads-up}
::::{.Heads-up-header}
Heads-up!
::::
::::{.Heads-up-container}
In addition to missing values, it is also good practice to scan for obvious inconsistencies, such as unexpected category labels or implausible numeric values. Earlier steps have already verified that the response variable lies within the required range. In a larger or messier dataset, it would also be sensible to check whether predictors such as hours of sleep or part-time work hours fall within reasonable bounds. We omit these additional checks here to keep the example focused, but the same principles apply in more complex analyses.
::::
:::

## Exploratory Data Analysis (EDA)

With the data cleaned and prepared, we now turn to our next step in our data science workflow: exploratory data analysis (EDA). The goal of this step is not to draw final conclusions, but to develop intuition about the response variable and its relationship with key predictors.

### Distribution of the Response Variable

We begin by examining the distribution of the final grade, expressed as a proportion. This helps us assess whether the response is bounded, skewed, or clustered near the boundaries.

::: {.panel-tabset}
## R Code
```{r}
hist(
  data$final_grade_prop,
  breaks = 10,
  main = "Distribution of Final Grades",
  xlab = "Final Grade (Proportion)",
  col = "lightgray",
  border = "white"
)

```

## Python Code
```{python}
import matplotlib.pyplot as plt

plt.hist(
  data["final_grade_prop"],
  bins=10
)

plt.xlabel("Final Grade (Proportion)")
plt.ylabel("Frequency")
plt.title("Distribution of Final Grades")
plt.show()
```
:::

The histogram, once again, shows that grades are continuous and bounded between zero and one, with the overall distribution being slightly left skewed. Great to see that students are mostly doing well here!

### Distribution of Predictor Variables

Let’s also examine the distributions of the predictor variables. This helps us understand the range, balance, and structure of the inputs before moving on to modelling. For continuous predictors, we stick to histograms to inspect their spread and shape. For categorical predictors, bar charts are sufficient to show how observations are distributed across categories.

#### Continuous Predictors

We begin with the continuous predictors: hours of sleep per night and part-time work hours.

::: {.panel-tabset}
## R Code
```{r}
par(mfrow = c(1, 2))

x <- data$`Hours of Sleep per Night`
br <- seq(min(x), max(x), length.out = 11)

# bins 1..10, with last bin including the rightmost edge like NumPy
bin_id <- findInterval(x, br, rightmost.closed = TRUE, all.inside = TRUE)

counts_like_numpy <- tabulate(bin_id, nbins = 10)

# Plot the histogram manually so the bar heights match counts_like_numpy exactly
plot(NA,
  xlim = range(br),
  ylim = c(0, max(counts_like_numpy, na.rm = TRUE)),
  main = "Hours of Sleep per Night",
  xlab = "Hours",
  ylab = "Frequency",
  xaxt = "n"
)
axis(1)
rect(
  xleft   = br[-length(br)],
  ybottom = 0,
  xright  = br[-1],
  ytop    = counts_like_numpy,
  col     = "lightgray",
  border  = "white"
)

hist(
  data$`Part-time Work Hours`,
  breaks = seq(
    min(data$`Part-time Work Hours`),
    max(data$`Part-time Work Hours`),
    length.out = 11
  ),
  main = "Part-time Work Hours",
  xlab = "Hours per Week",
  col = "lightgray",
  border = "white"
)

par(mfrow = c(1, 1))
```

## Python Code
```{python}
import matplotlib.pyplot as plt

fig, axes = plt.subplots(1, 2, figsize=(10, 4))

axes[0].hist(data["Hours of Sleep per Night"], bins=10)
axes[0].set_title("Hours of Sleep per Night")
axes[0].set_xlabel("Hours")
axes[0].set_ylabel("Frequency")

axes[1].hist(data["Part-time Work Hours"], bins=10)
axes[1].set_title("Part-time Work Hours")
axes[1].set_xlabel("Hours per Week")

plt.tight_layout()
plt.show()

```
:::

::: {.Heads-up}
::::{.Heads-up-header}
Heads-up!
::::
::::{.Heads-up-container}
The R code for Hours of Sleep per Night is intentionally more detailed than usual. This alternative approach ensures that the histogram uses the same bin boundaries and bin-assignment rules as Python, so the two plots are directly comparable. In practice, R’s default `hist()` function is usually sufficient.
::::
:::

Overall, both hours of sleep per night and part-time work hours appear to be approximately uniformly distributed, with no strong skewness or extreme concentration in specific ranges. This suggests that neither predictor is dominated by a narrow subset of values, which is helpful for subsequent regression modelling.

#### Categorical Predictors

For categorical predictors, histograms are no longer appropriate. Instead, we use bar charts to visualize how observations are distributed across categories. This allows us to quickly check whether certain groups are under-represented, over-represented, or potentially imbalanced.

::: {.panel-tabset}
## R Code
```{r}
par(mfrow = c(3, 3), mar = c(7, 4, 3, 1))

categorical_vars <- c(
  "Café Location",
  "Study Time",
  "Year in School",
  "Hobbies",
  "Primary Sport",
  "Living Arrangement",
  "Study Group Participation",
  "Has a Tutor"
)

for (v in categorical_vars) {
  barplot(
  table(data[[v]]),
  main = v,
  ylab = "Count",
  col = "lightgray",
  border = "white",
  las = 2
  )
}

par(mfrow = c(1, 1))
```

## Python Code
```{python}
import matplotlib.pyplot as plt

categorical_vars = [
  "Café Location",
  "Study Time",
  "Year in School",
  "Hobbies",
  "Primary Sport",
  "Living Arrangement",
  "Study Group Participation",
  "Has a Tutor"
]

fig, axes = plt.subplots(3, 3, figsize=(12, 9))
axes = axes.flatten()

for i, var in enumerate(categorical_vars):
  counts = data[var].value_counts()
  axes[i].bar(counts.index.astype(str), counts.values)
  axes[i].set_title(var)
  axes[i].set_ylabel("Count")
  axes[i].tick_params(axis="x", rotation=45)

# Turn off unused panels (9 total, 8 used)
for j in range(len(categorical_vars), len(axes)):
  axes[j].axis("off")

plt.tight_layout()
plt.show()
```
:::

Overall, the categorical predictors appear to be well balanced across their respective levels. No single category overwhelmingly dominates the dataset, and each level is represented by a reasonable number of observations. This balance is desirable in a regression setting, as it ensures that comparisons between categories are supported by sufficient data and reduces the risk of unstable or highly variable coefficient estimates.

### Relationships Between Predictors and the Response

So far, we have examined each variable individually. While univariate visualizations are useful for understanding distributions and balance, they provide limited insight into how predictors relate to the response variable. To build intuition about these relationships, we now turn to bivariate exploratory analysis, where predictors and the response are examined together. In what follows, we explore the relationship between the response and continuous predictors, followed by the relationship between the response and categorical predictors.

#### Continuous Predictors and the Response

When both the predictor and the response are continuous, scatterplots are a natural and effective visualization choice. Scatterplots allow us to inspect the form of the relationship, assess potential nonlinearity, and identify unusual observations or outliers.

::: {.panel-tabset}
## R Code
```{r}
par(mfrow = c(1, 2))

plot(
  data$`Hours of Sleep per Night`,
  data$final_grade_prop,
  xlab = "Hours of Sleep per Night",
  ylab = "Final Grade (Proportion)",
  main = "Sleep and Final Grade",
  pch = 19,
  col = "gray"
)

plot(
  data$`Part-time Work Hours`,
  data$final_grade_prop,
  xlab = "Part-time Work Hours",
  ylab = "Final Grade (Proportion)",
  main = "Work Hours and Final Grade",
  pch = 19,
  col = "gray"
)

par(mfrow = c(1, 1))
```

## Python Code
```{python}
import matplotlib.pyplot as plt

fig, axes = plt.subplots(1, 2, figsize=(10, 4))

axes[0].scatter(
  data["Hours of Sleep per Night"],
  data["final_grade_prop"],
  color="gray"
)
axes[0].set_xlabel("Hours of Sleep per Night")
axes[0].set_ylabel("Final Grade (Proportion)")
axes[0].set_title("Sleep and Final Grade")

axes[1].scatter(
  data["Part-time Work Hours"],
  data["final_grade_prop"],
  color="gray"
)
axes[1].set_xlabel("Part-time Work Hours")
axes[1].set_ylabel("Final Grade (Proportion)")
axes[1].set_title("Work Hours and Final Grade")

plt.tight_layout()
plt.show()
```
:::

At first glance, the scatterplots do not reveal a strong or obvious relationship between either continuous predictor and the final grade proportion. The points are widely dispersed, and no clear linear trend is immediately apparent. This suggests that, on their own, hours of sleep per night and part-time work hours may not have a simple or direct association with student performance.

However, this observation should be treated as hypothesis-generating rather than conclusive. Scatterplots reflect raw, unadjusted relationships and do not account for the influence of other variables in the dataset. It is therefore possible that these predictors matter only in combination with other factors, or that their effects become apparent after controlling for confounding variables in a regression model.

#### Categorical Predictors and the Response

When the predictor is categorical and the response is continuous, scatterplots are no longer appropriate. Instead, we compare the distribution of the response across groups defined by the categorical variable. Boxplots are a convenient choice because they summarize the center, spread, and potential outliers of the response within each category.

::: {.panel-tabset}
## R Code
```{r}
#| fig.height: 10
#| fig.width: 8
par(mfrow = c(3, 3), mar = c(7, 4, 3, 1))

categorical_vars <- c(
  "Café Location",
  "Study Time",
  "Year in School",
  "Hobbies",
  "Primary Sport",
  "Living Arrangement",
  "Study Group Participation",
  "Has a Tutor"
)

for (v in categorical_vars) {
  boxplot(
    data$final_grade_prop ~ data[[v]],
    main = v,
    ylab = "Final Grade (Proportion)",
    xlab = v,
    col = "lightgray",
    las = 2
  )
}

par(mfrow = c(1, 1))

```

## Python Code
```{python}
#| results: hide

import matplotlib.pyplot as plt

categorical_vars = [
  "Café Location",
  "Study Time",
  "Year in School",
  "Hobbies",
  "Primary Sport",
  "Living Arrangement",
  "Study Group Participation",
  "Has a Tutor"
]

fig, axes = plt.subplots(3, 3, figsize=(12, 9))
axes = axes.flatten()

for i, var in enumerate(categorical_vars):
  data.boxplot(
    column="final_grade_prop",
    by=var,
    ax=axes[i],
    grid=False
  )
  axes[i].set_title(var)
  axes[i].set_xlabel("")
  axes[i].set_ylabel("Final Grade (Proportion)")

# Turn off unused panel
for j in range(len(categorical_vars), len(axes)):
  _ = axes[j].axis("off")

plt.tight_layout()
plt.show()
```
:::

From the plots, we observe that study time is the only predictor with a clearly visible positive association with final grade. This finding is largely unsurprising and aligns well with common intuition: students who devote more time to studying tend to perform better academically.

At the same time, it is somewhat unexpected that the remaining categorical variables show little evidence of a strong marginal relationship with the response. Factors such as tutoring status, study group participation, living arrangement, or extracurricular interests might reasonably be expected to influence academic performance, yet no clear or consistent differences emerge when these variables are examined in isolation.

These observations are intriguing but should be interpreted with caution. The analysis at this stage is purely exploratory and reflects only superficial, marginal relationships. While these insights help guide our expectations, no substantive conclusions can be drawn until formal modelling is conducted, where the effects of multiple predictors can be assessed simultaneously.

## Data Modeling

After exploring the data and developing initial intuition through EDA, we now turn to the modelling stage. The purpose of this step is to formalize the relationships we observed and assess how different academic and lifestyle factors are associated with students’ final grades when considered jointly.

### Choosing a Suitable Regression Model

The choice of regression model should follow directly from what we have learned about the response variable and from the goals of the analysis.

In this case, the response variable, `final_grade_prop`, represents a student’s final grade expressed as a proportion. After preprocessing, all values lie strictly between zero and one. The EDA also showed that grades are not symmetrically distributed and tend to cluster toward higher values, with variability that appears to depend on the average grade level.

Equally important, our goal is not to predict future grades as accurately as possible, but to understand how study habits, lifestyle factors, and forms of academic support are associated with students’ performance.

Taken together, these considerations point toward Beta regression as an appropriate modeling choice. Beta regression is designed specifically for proportion data and naturally respects the bounds of the response. It allows the mean of the response to depend on predictors while accommodating heteroskedasticity, which is common in proportional outcomes.

By contrast, ordinary least squares assumes constant variance and does not constrain predictions to the unit interval. These assumptions are difficult to justify for grade data, making OLS a less suitable option for this analysis.

### Defining Modelling Parameters

With the modelling framework selected, the next step is to specify which variables to include. In this case study, we include all available predictors to capture a broad and realistic picture of student life.

The predictors reflect multiple dimensions of a student’s experience, including time allocation (such as study time, sleep, and part-time work hours), academic context (such as year in school), and sources of support (such as tutoring and study group participation), as well as several categorical variables describing personal routines and preferences.

Including the full set of predictors allows us to examine the association between each factor and the average final grade while holding the other variables constant. This multivariable perspective is especially important given that most predictors showed weak marginal relationships during EDA but may still play a role once considered together.

### Setting Up the Modelling Equation

Using Beta regression with a logit link, we model the average final grade proportion as a function of the predictors.

For each student $i$, the response is assumed to follow a Beta distribution,

$$
Y_i \sim \text{Beta}(\mu_i \phi,\ (1 - \mu_i)\phi)
$$


where $\mu_i$ denotes the average grade proportion and $\phi$ controls how concentrated grades are around that average.

The average proportion is linked to the predictors through the linear predictor:

<div style="overflow-x: auto; white-space: nowrap;">
$$
\begin{aligned}
\text{logit}(\mu_i) &= \beta_0 \\
&\quad + \beta_1 \times \text{part\_time\_work\_hours}_i \\
&\quad + \beta_2 \times \text{hours\_of\_sleep}_i \\
&\quad + \text{(terms for Study Time)} \\
&\quad + \text{(terms for Café Location)} \\
&\quad + \text{(terms for Year in School)} \\
&\quad + \text{(terms for Hobbies)} \\
&\quad + \text{(terms for Primary Sport)} \\
&\quad + \text{(terms for Living Arrangement)} \\
&\quad + \text{(terms for Study Group Participation)} \\
&\quad + \text{(terms for Has a Tutor)}
\end{aligned}
$$
</div>

The logit link maps the average grade proportion to the real line, allowing predictors to enter the model in a familiar linear form. Once the linear predictor is computed, we can return to the original grade scale using the inverse-logit transformation,

$$
\mu_i = \frac{\exp(\eta_i)}{1 + \exp(\eta_i)}
$$

where $\eta_i$ represents the linear combination of predictors.

Under this model, each coefficient describes how a predictor is associated with changes in the average final grade, on the logit scale, holding all other variables constant. Positive coefficients correspond to higher expected grades, while negative coefficients correspond to lower expected grades.

With the model structure defined, we are now ready to estimate the model parameters and examine how these academic and lifestyle factors relate to student performance in practice.

## Estimation

In this step, we fit the Beta regression model to the data and obtain numerical estimates for each coefficient. These estimates allow us to quantify how different academic and lifestyle factors are associated with students’ average final grades under the specified model.

::: {.Tip}
::::{.Tip-header}
How Beta Regression Is Fit
:::
::::{.Tip-container}
Like Gamma regression, Beta regression is typically estimated using maximum likelihood estimation (MLE). Rather than minimizing squared errors, MLE finds the parameter values that make the observed proportions most likely under the assumed Beta distribution. This approach naturally accounts for the bounded nature of the response and the relationship between the mean and variability.
:::
:::

### Fitting the Model

In R, Beta regression can be fit using the `betareg()` function, which is designed specifically for proportion outcomes. In Python, Beta regression is available through `statsmodels`, allowing us to specify the same link function and modelling structure.

::: {.panel-tabset}
## R Code
```{r r-fit}
library(betareg)

# Fit Beta regression model with logit link
beta_model <- betareg(
  final_grade_prop ~ .,
  data = data
)

# Display model summary
summary(beta_model)
```

## Python Code
```{python py-fit}
import statsmodels.api as sm
from statsmodels.othermod.betareg import BetaModel

y = data_encoded["final_grade_prop"].astype(float)

X = data_encoded.drop(columns=["final_grade_prop"])
X = sm.add_constant(X, has_constant="add") 

fit_py = BetaModel(y, X).fit()
print(fit_py.summary())
```
:::

### Interpreting the Coefficients

Based on the output, the (partially shown) fitted equation is:

<div style="overflow-x: auto; white-space: nowrap;">
$$
\begin{aligned}
\text{logit}(\mu_i) &= 0.6665 \\
&\quad - 0.0008 \times \text{part\_time\_work\_hours}_i \\
&\quad - 0.0145 \times \text{hours\_of\_sleep}_i \\
&\quad + \text{(terms for Study Time)} \\
&\quad + \text{(terms for Café Location)} \\
&\quad + \text{(terms for Year in School)} \\
&\quad + \text{(terms for Hobbies)} \\
&\quad + \text{(terms for Primary Sport)} \\
&\quad + \text{(terms for Living Arrangement)} \\
&\quad + \text{(terms for Study Group Participation)} \\
&\quad + \text{(terms for Has a Tutor)}
\end{aligned}
$$
</div>

While we can interpret the numbers given here, but before that, we should always check goodness of fit first. If the goodness of fit is bad, then the interpretation of these numbers could be meaningless.

## Goodness of Fit

Assessing goodness of fit for Beta regression is conceptually similar to other generalized linear models, but there are a few important differences compared to ordinary least squares regression.

### Log-likelihood, AIC, and BIC

Because Beta regression is fit using maximum likelihood, overall model fit is commonly assessed using:

- Log-likelihood
- Akaike Information Criterion (AIC)
- Bayesian Information Criterion (BIC)

Higher log-likelihood values indicate a better fit, while lower AIC and BIC values suggest a preferable balance between model fit and model complexity. These metrics are especially useful when comparing multiple Beta regression models fitted to the same response variable. In our fitted model, the log-likelihood is 1474.7, the AIC is −2903, and the BIC is −2791. Although we do not compare multiple models here, these metrics become especially informative when evaluating alternative model specifications.

### Pseudo-$R^2$

When using R’s `betareg()`, the model summary reports a pseudo-$R^2$ value. This quantity is defined as:

$$
R^{2}_{\text{pseudo}} = 1 - \frac{\ell(\text{null model})}{\ell(\text{fitted model})}
$$

where $\ell(\cdot)$ denotes the log-likelihood. This measure quantifies how much the fitted model improves upon an intercept-only model. From the R output, the pseudo-$R^2$ for our model is 0.6014, indicating an improvement over the null model.

::: {.Heads-up}
::::{.Heads-up-header}
Notes on pseudo-$R^2$
::::
::::{.Heads-up-container}
When interpreting pseudo-$R^2$, it is important to keep the following points in mind:

- This pseudo-$R^2$ is not directly comparable to the $R^2$ from linear regression.
- Values tend to be lower and should be interpreted relatively, not absolutely.
- The pseudo-$R^2$ is most useful for comparing competing models, rather than judging fit in isolation.

In contrast, Python’s `statsmodels.BetaModel` does not report a pseudo-$R^2$ by default, even though the same quantity could be computed manually from the log-likelihoods. This reflects a design choice rather than a difference in the underlying statistical model.
::::
:::

### Precision Parameter {#sec-beta-precision-parameter}

Unlike linear regression, Beta regression models not only the average response but also how much the observations vary around that average. This variability is summarized by the **precision parameter**, commonly denoted by $\phi$.

You can think of the precision parameter as describing how tightly the data cluster around the fitted mean:

- Larger values of $\phi$ indicate that observations are more tightly concentrated around the predicted mean.
- Smaller values of $\phi$ indicate greater variability in the response.

Recall from our earlier discussion that Beta regression can be expressed in terms of two quantities:

1. the average proportion ($\mu$), and
2. the precision parameter ($\phi$).

The regression coefficients estimate how the predictors affect the average proportion, while the precision parameter describes how much variability remains around those fitted values.

In R's `betareg()`, the precision parameter is estimated directly. In this example, the fitted model reports:

$$
\hat{\phi}=43.98
$$

Python's `statsmodels.BetaModel` reports the precision parameter on the log scale instead. The reported value of 3.78 corresponds to:

$$
\exp(3.78)=43.98
$$

which matches the estimate reported by `betareg()`.

A precision estimate of approximately 44 suggests that the observed grades are relatively tightly clustered around the values predicted by the model. In practice, the precision parameter is typically interpreted together with residual diagnostics rather than evaluated against a specific cutoff value.

### Residual Diagnostics {#sec-beta-residual-diagnostics}

As with other generalized linear models, residual diagnostics help us check whether a Beta regression model is capturing the main patterns in the data. In this section, we focus on Pearson residuals, which compare the observed values to the fitted mean values on a standardized scale.

::: {.panel-tabset}
## R Code
```{r}
library(betareg)

# Extract fitted means and residuals
mu_hat <- fitted(beta_model)
r_pearson <- residuals(beta_model, type = "pearson")

plot(mu_hat, r_pearson,
     xlab = "Fitted mean (mu_hat)",
     ylab = "Pearson residual",
     main = "Pearson residuals vs fitted")
abline(h = 0, lty = 2)

```

## Python Code
```{python}
#| results: hide

import matplotlib.pyplot as plt

mu_hat = fit_py.fittedvalues
pearson_resid = fit_py.resid_pearson

_ = plt.scatter(mu_hat, pearson_resid)
_ = plt.axhline(0, linestyle="--")
_ = plt.xlabel("Fitted mean")
_ = plt.ylabel("Pearson residual")
_ = plt.title("Pearson residuals vs fitted")

plt.show()
```
:::

The plots show Pearson residuals plotted against the fitted mean values. The residuals are centered around zero across the range of fitted values, indicating no systematic over or under-prediction. The vertical banding in the plot is not a warning sign. It appears because categorical predictors group observations at the same fitted mean, so instead of a smooth cloud of points, the plot shows columns. Within each column, the residuals are roughly symmetric around zero with no obvious outliers, which is what matters. While residual variability increases slightly for fitted means close to one, this behaviour is expected for Beta regression due to the bounded nature of the response.

### Others

Two other checks are also common in practice, though they are not shown here. A standard Q-Q plot is not well suited for Beta regression because Pearson residuals from this model are not expected to follow a normal distribution. Checking for influential observations using something like Cook's distance is also a good habit, as it can reveal whether any single observation is pulling the fitted coefficients more than it should.

## Results

Great! Nothing seems particularly concerning with goodness of fit. Having established that the model fits the data reasonably well, we can now finally turn to the substantive findings, interpreting results and answering our central question:

> Which aspects of students’ academic and daily routines are associated with differences in final course grades?

Recall that the model was fit using Beta regression with a logit link. As a result, coefficients describe changes in the log-odds of the mean grade proportion, rather than direct changes in grades themselves. Exponentiating a coefficient yields an odds ratio that describes how the odds of the mean grade proportion i.e., $\mu/(1-\mu)$ change multiplicatively as a predictor increases. Unlike in logistic regression, this is not the odds of a binary event such as passing or failing; it describes the expected proportion itself, holding all other variables constant.

Because the logit link is nonlinear, these effects should be interpreted in relative terms rather than as fixed percentage-point increases. The same change in log-odds can correspond to different changes in grade proportions depending on where a student lies on the grade scale.

### Predictors Associated with Final Grades

Among all variables included in the model, *Study Time* stands out as the dominant factor associated with academic performance.

Using students who study less than one hour as the reference group, we observe a clear and increasing association between reported study time and average grades:

- **1–3 hours of study**: The estimated coefficient is 0.458, corresponding to an odds ratio of

$$
e^{0.458} \approx 1.58
$$

This indicates that students who study 1–3 hours have noticeably higher odds of a higher mean grade proportion than those in the baseline group.

- **3–5 hours of study**: With a coefficient of 1.130, the odds ratio increases to

$$
e^{1.130} \approx 3.10
$$

This suggests that students in this category have roughly three times the odds of a higher mean grade compared to the reference group.

- **5+ hours of study**: The largest effect is observed here, with a coefficient of 1.706 and an odds ratio of

$$
e^{1.706} \approx 5.51
$$

Students who study five or more hours are therefore predicted to have substantially higher average grades than those who study less than one hour.

The pattern is monotonic and strong: as reported study time increases, the model consistently predicts higher average grade proportions. This relationship remains even after adjusting for sleep, work hours, academic year, and other lifestyle variables.


### Variables Without Strong Evidence of Association

In contrast, the remaining predictors in the model do not show statistically significant associations with final grades at conventional significance levels.

These include:

- Part-time work hours
- Hours of sleep per night
- Café location
- Year in school
- Hobbies
- Primary sport
- Living arrangement
- Study group participation
- Tutoring status

For example, the estimated effects of part-time work hours and sleep duration are small relative to their uncertainty, providing little evidence of a systematic relationship with grades once study time is accounted for. Similarly, variables describing where students study or whether they participate in academic support activities do not appear to differentiate average grades in this dataset.

::: {.Tip}
::::{.Tip-header}
Interpretation Reminder
::::
::::{.Tip-container}
It is worth emphasizing that these findings are conditional on the model specification and the available data. Lack of statistical significance does not imply that these factors are unimportant in all settings, only that the present analysis does not provide strong evidence of an association.
::::
:::

## Storytelling

To complete the data science workflow, we step back from coefficients and diagnostics and focus on how the results can be interpreted and applied in practice. This final stage, storytelling, is where statistical output is translated into insights that support real-world decision-making.

In this case study, the Beta regression model helps us understand how different aspects of students’ academic lives relate to their final course grades. Rather than treating grades as isolated numbers, the model allows us to view them as outcomes shaped by study habits, routines, and context. Below, we distill the key takeaways from the analysis and discuss what they imply for students, instructors, and academic support staff.

### Key Insights From the Model

Study time stands out as the primary driver of academic performance.
Across all model specifications, higher reported study time is consistently associated with higher average grades. The relationship is not only statistically significant, but also monotonic: students who study more tend to achieve better outcomes, even after accounting for sleep, work commitments, and other lifestyle factors. This reinforces a simple but powerful message: sustained study effort plays a central role in academic success.

From a practical perspective, this suggests that interventions aimed at improving time management, study planning, and sustained engagement with course material are likely to yield meaningful benefits. Academic advisors and instructors might focus less on where or how students study, and more on helping them build realistic, consistent study routines.

Many commonly discussed lifestyle factors show little evidence of association once study time is accounted for. Variables such as hours of sleep, part-time work hours, tutoring, study group participation, and living arrangement do not display strong independent associations with final grades in this model. While these factors are often assumed to influence academic outcomes, the analysis suggests that their effects may be indirect, context-dependent, or mediated through study time itself.

This does not mean that sleep or academic support are unimportant. Rather, it suggests that their influence may be more subtle than often assumed, or that they matter most when they enable students to study more effectively. In other words, these factors may support learning indirectly rather than directly translating into higher grades.

Environment and personal preferences appear less critical than effort in this dataset. Café location, hobbies, and sports participation show no clear relationship with grades once other variables are controlled for. This finding highlights an encouraging implication: students with very different routines and preferences can achieve similar outcomes, provided they invest sufficient time and effort into studying. Academic success does not appear to be tied to a specific lifestyle or environment, but rather to how students allocate their time.

::: {.Tip}
::::{.Tip-header}
Using the Results Responsibly
::::
::::{.Tip-container}
It is important to remember that this analysis is associational, not causal. The model does not prove that increasing study time will automatically raise grades for every student. Instead, it shows that, in this dataset, higher study time tends to co-occur with higher average grades after accounting for other measured factors.

Nevertheless, the results provide a useful evidence-based starting point for discussion. They help prioritize which factors are most strongly linked to performance and which may play a smaller or more indirect role. For educators and institutions, this kind of insight can guide where limited resources and attention are most likely to have impact.
::::
:::

## Conclusion

And that wraps up the Beta regression chapter. We have walked through a complete data science workflow, from problem formulation and data preparation to modelling, diagnostics, and interpretation and seen how Beta regression offers a principled and flexible framework for outcomes that are continuous, bounded between zero and one, and heteroskedastic.

More importantly, the chapter highlights that the value of Beta regression lies not just in fitting a model, but in using the model thoughtfully within a broader analytical process. Careful study design, appropriate preprocessing, exploratory analysis, model checking, and clear communication are what ultimately turn proportion data into actionable insight.

With these ideas in place, you are now ready to apply Beta regression confidently and communicate your findings in a way that supports informed decision-making.



::: {.Warning}
::::{.Warning-header}
Fun fact!
::::
::::{.Warning-container}
**Beet-a you didn’t know!** The beet plant belongs to the genus *Beta* (*Beta vulgaris*, to be precise). So the next time you run a Beta regression, remember that somewhere in a field, a humble root vegetable is quietly sharing your model’s name.
::::
:::

::: {#fig-beta-regression}
```{mermaid}
mindmap
  root((Regression 
  Analysis)
    Continuous <br/>Outcome Y
      {{Unbounded <br/>Outcome Y}}
        )Chapter 3: <br/>Ordinary <br/>Least Squares <br/>Regression(
          (Normal <br/>Outcome Y)
      {{Nonnegative <br/>Outcome Y}}
        )Chapter 4: <br/>Gamma Regression(
          (Gamma <br/>Outcome Y)
      {{Bounded <br/>Outcome Y <br/> between 0 and 1}}
        )Chapter 5: Beta <br/>Regression(
          (Beta <br/>Outcome Y)
    Discrete <br/>Outcome Y
```
:::

## Practice Exercises {#sec-beta-exercises}

The following practice exercises reinforce the main ideas from this chapter. They are designed to help you move between the **conceptual**, **computational**, and **interpretive** parts of Beta regression. Some questions focus on the model structure itself: the Beta random component, the logit link, the mean and precision parameterization, coefficient interpretation, and goodness-of-fit diagnostics. Others prepare you to think through applied modelling decisions for bounded proportion responses.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/children-teacher-rectangular-figures-7464611/).](img/exercises.png){width="550"}

As you work through the problems, keep in mind that Beta regression is useful for continuous responses that live on the unit interval, but the main lesson of the chapter is not only how to fit the model. It is also how to prepare proportion data responsibly, check whether the fitted model is adequate, and communicate the results without overstating what an observational analysis can show.

::: {.quiz}

### Conceptual Questions {#sec-beta-exercises-conceptual}

::: {.Question}
::::{.Question-header}
Question 5.1
::::
::::{.Question-container}
**True or False**

Beta regression is appropriate for any numerical response variable whose values are positive.
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 5.1
::::
::::{.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

**Correct answer:** False.

**Rationale:**

Beta regression is designed for **continuous proportion responses** that lie strictly between zero and one:

$$
0 < Y_i < 1.
$$

A positive response can still be unsuitable for Beta regression. For example, response time, income, sales revenue, and waiting time are positive but not bounded above by one. Those outcomes require other modelling strategies, such as Gamma regression for many positive continuous responses.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 5.2
::::
::::{.Question-container}
**Open-ended Question**

A website team records the conversion rate for each landing page, where each value is the proportion of visitors who made a purchase. Why might Beta regression be more appropriate than ordinary least squares for this response?
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 5.2
::::
::::{.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

Beta regression is more appropriate because the response is a **proportion**. Conversion rates must stay between zero and one, while ordinary least squares can produce fitted values outside that range.

Beta regression also allows the variability of the response to depend on the mean. This matters for proportion data because values near zero or one often have different variability than values near the middle of the unit interval.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 5.3
::::
::::{.Question-container}
**Multiple Choice**

Which response is most naturally suited to standard Beta regression?

**A.** Whether a student passed a course.

**B.** The number of assignments a student submitted late.

**C.** A student's final grade after rescaling it to a value strictly between zero and one.

**D.** The number of minutes a student spent studying.
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 5.3
::::
::::{.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

**Correct answer:** C.

**Rationale:**

Standard Beta regression is designed for continuous responses on the open unit interval. A rescaled grade proportion can be suitable if it satisfies

$$
0 < Y_i < 1.
$$

Option A is binary, Option B is a count, and Option D is a positive continuous duration.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 5.4
::::
::::{.Question-container}
**Open-ended Question**

What makes the Beta distribution flexible enough to model proportion data with **different shapes**? Give examples of the kinds of shapes it can represent.
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 5.4
::::
::::{.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

The Beta distribution can take many **different distributional shapes** depending on its two shape parameters, usually written $\alpha$ and $\beta$. It can be symmetric, skewed toward zero, skewed toward one, concentrated near the middle, or concentrated near the boundaries.

This flexibility is useful because real proportion data do not all look alike. For example, one dataset may contain conversion rates mostly near zero, while another may contain student grade proportions mostly near the high end of the scale.

::: {.panel-tabset}

## **`R`**

```{r}
x <- seq(0.001, 0.999, length.out = 500)

plot(
  x, dbeta(x, 0.5, 0.5),
  type = "l",
  ylim = c(0, 4),
  xlab = "Proportion",
  ylab = "Density",
  main = "Different Beta distribution shapes"
)

lines(x, dbeta(x, 2, 2), col = "blue", lwd = 2)
lines(x, dbeta(x, 2, 6), col = "red", lwd = 2)
lines(x, dbeta(x, 6, 2), col = "darkgreen", lwd = 2)

legend(
  "top",
  legend = c("Boundary-heavy", "Symmetric", "Skewed right", "Skewed left"),
  col = c("black", "blue", "red", "darkgreen"),
  lwd = 2,
  bty = "n"
)
```

## **`Python`**

```{python}
import numpy as np
import matplotlib.pyplot as plt
from scipy.stats import beta

x = np.linspace(0.001, 0.999, 500)

plt.figure()
plt.plot(x, beta.pdf(x, 0.5, 0.5), label="Boundary-heavy")
plt.plot(x, beta.pdf(x, 2, 2), label="Symmetric")
plt.plot(x, beta.pdf(x, 2, 6), label="Skewed right")
plt.plot(x, beta.pdf(x, 6, 2), label="Skewed left")
plt.xlabel("Proportion")
plt.ylabel("Density")
plt.title("Different Beta distribution shapes")
plt.legend()
plt.show()
```

:::

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 5.5
::::
::::{.Question-container}
**True or False**

Standard Beta regression can directly model responses that are exactly equal to zero or exactly equal to one.
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 5.5
::::
::::{.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

**Correct answer:** False.

**Rationale:**

The standard Beta distribution is defined on the **open** interval:

$$
0 < Y_i < 1.
$$

Exact zeros and ones are boundary values and need special attention. Depending on the problem, the analyst may need to transform the response, use a model that allows boundary values, or reconsider whether the response is being measured at the right level of detail.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 5.6
::::
::::{.Question-container}
**Open-ended Question**

In the student-performance case study, final grades were originally recorded as percentages. What two checks should be made before using them as the response in a standard Beta regression model?
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 5.6
::::
::::{.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

First, the grades should be rescaled from percentages to proportions:

$$
Y_i = \frac{\text{Final Grade Percentage}_i}{100}.
$$

Second, the analyst should check that the rescaled values lie strictly between zero and one:

$$
0 < Y_i < 1.
$$

This boundary check matters because standard Beta regression cannot directly handle exact zeros or exact ones.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 5.7
::::
::::{.Question-container}
**Multiple Choice**

Let $Y_i$ denote a continuous proportion response for observation $i$, and let $\mathbf{x}_i$ denote the regressors for that observation. In the Beta regression parameterization

$$
Y_i \mid \mathbf{x}_i \sim \text{Beta}(\mu_i\phi,\ (1-\mu_i)\phi),
$$

where $\mu_i=\mathbb{E}(Y_i\mid \mathbf{x}_i)$ is the conditional mean proportion and $\phi>0$ is the precision parameter, what does a larger value of $\phi$ indicate?

**A.** The response is no longer bounded between zero and one.

**B.** The observations are more tightly concentrated around the mean proportion.

**C.** The fitted mean proportion must equal zero.

**D.** The response has become a count.
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 5.7
::::
::::{.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

**Correct answer:** B.

**Rationale:**

The parameter $\phi$ is the **precision parameter**. Holding the mean $\mu_i$ fixed, the conditional variance is

$$
\operatorname{Var}(Y_i\mid \mathbf{x}_i)
=
\frac{\mu_i(1-\mu_i)}{1+\phi}.
$$

Thus, larger values of $\phi$ correspond to less variability around the conditional mean proportion. Smaller values of $\phi$ allow $Y_i$ to be more spread out around $\mu_i$.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 5.8
::::
::::{.Question-container}
**Open-ended Question**

Suppose a city records $Y_i$, the proportion of available bike-share docks that are occupied at station $i$ during the evening commute. Possible regressors include distance to the nearest train stop, number of nearby offices, rain status, and day-of-week indicators.

Before fitting a Beta regression model, what else needs to be specified or checked?
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 5.8
::::
::::{.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

Before fitting the model, the analyst should specify the response, regressors, random component, systematic component, and link function, and should check that the response values are strictly between zero and one.

The response is

$$
Y_i
=
\text{proportion of occupied bike-share docks at station } i.
$$

The regressors might include distance to the nearest train stop, number of nearby offices, rain status, and day-of-week indicators.

The **random component** is the Beta distribution for the response:

$$
Y_i \mid \mathbf{x}_i \sim \text{Beta}(\mu_i\phi,\ (1-\mu_i)\phi).
$$

The **systematic component** is the linear predictor:

$$
\eta_i
=
\beta_0 + \beta_1 x_{i1} + \cdots + \beta_k x_{ik}.
$$

The **link function** connects the mean proportion to the linear predictor. With the logit link,

$$
\text{logit}(\mu_i)
=
\log\left(\frac{\mu_i}{1-\mu_i}\right)
=
\eta_i.
$$

This setup also requires choosing how categorical regressors will be coded and confirming that no observed response values are exactly zero or exactly one.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 5.9
::::
::::{.Question-container}
**True or False**

Using a logit link in Beta regression helps ensure that the fitted mean proportion stays between zero and one.
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 5.9
::::
::::{.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

**Correct answer:** True.

**Rationale:**

The linear predictor $\eta_i$ can take any real value, but the inverse-logit transformation maps it back to the unit interval:

$$
\mu_i
=
\frac{\exp(\eta_i)}{1+\exp(\eta_i)}.
$$

This guarantees

$$
0 < \mu_i < 1.
$$

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 5.10
::::
::::{.Question-container}
**Multiple Choice**

In a Beta regression model for the proportion of recyclable material in a household's weekly trash stream, a coefficient of $\beta_j = 0$ means:

**A.** The household's recycling proportion is zero.

**B.** Holding the other regressors fixed, a one-unit increase in that regressor is associated with no change in the expected recycling proportion.

**C.** The response is binary.

**D.** The model has perfect goodness of fit.
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 5.10
::::
::::{.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

**Correct answer:** B.

**Rationale:**

With the logit link, exponentiating a coefficient gives a multiplicative effect on

$$
\frac{\mu_i}{1-\mu_i}.
$$

Since

$$
\exp(0)=1,
$$

a coefficient of zero corresponds to no multiplicative change in the odds of the mean proportion, holding the other regressors fixed.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 5.11
::::
::::{.Question-container}
**Open-ended Question**

Suppose a Beta regression model for the proportion of mobile-app users who complete onboarding estimates a coefficient of $0.40$ for an indicator that a push reminder was sent, using no reminder as the reference category. How would you interpret $\exp(0.40)$?
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 5.11
::::
::::{.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

The quantity $\exp(0.40)$ is approximately $1.49$. This means that, holding the other regressors fixed, app cohorts that received a push reminder are estimated to have about **1.49 times the odds of the mean onboarding-completion proportion** compared with cohorts that did not receive a reminder.

This is not the odds that one individual user completes onboarding. The response is a continuous completion proportion for a cohort, so the odds refer to

$$
\frac{\mu_i}{1-\mu_i},
$$

where $\mu_i$ is the fitted mean onboarding-completion proportion.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 5.12
::::
::::{.Question-container}
**Open-ended Question**

In a Beta regression model for the proportion of scheduled telehealth appointments completed by each clinic, what does it mean to interpret the weekend-availability coefficient while **holding other regressors fixed**?
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 5.12
::::
::::{.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

It means that the weekend-availability coefficient compares clinics that are the same with respect to the other regressors included in the model, such as patient volume, average appointment lead time, specialty mix, reminder system, and region, but differ in whether weekend appointments are available.

This is a conditional model interpretation. It does not mean the dataset contains perfectly matched clinics, and it does not prove that adding weekend availability causes a change in completion rates.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 5.13
::::
::::{.Question-container}
**Multiple Choice**

In @sec-beta-precision-parameter, the chapter introduces the Beta regression precision parameter, denoted by $\phi$. This parameter describes how tightly the response values cluster around the fitted mean $\mu_i$.

R's `betareg()` reports $\hat{\phi}$ directly. Python's `statsmodels.BetaModel` reports the same precision quantity on the log scale. If Python reports a log-precision value of $3.78$, what value should be compared with R's $\hat{\phi}$?

**A.** $3.78$

**B.** $\log(3.78)$

**C.** $\exp(3.78)$

**D.** $1/3.78$
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 5.13
::::
::::{.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

**Correct answer:** C.

**Rationale:**

If Python reports the precision parameter on the log scale, then the reported number is interpreted as

$$
\log(\hat{\phi}) = 3.78.
$$

To return to the precision scale, we exponentiate:

$$
\hat{\phi}=\exp(3.78).
$$

This is the value that should be compared with R's direct estimate of $\hat{\phi}$. The exponentiation matters because $\phi$ must be positive, while a log-scale parameter can range over the whole real line during model fitting.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 5.14
::::
::::{.Question-container}
**Open-ended Question**

Why is maximum likelihood estimation a natural fitting approach for Beta regression?
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 5.14
::::
::::{.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

Beta regression specifies a full probability distribution for each response value:

$$
Y_i \sim \text{Beta}(\mu_i\phi,\ (1-\mu_i)\phi).
$$

Maximum likelihood estimation chooses the parameter values that make the observed proportions most likely under that assumed distribution. This is more natural than minimizing squared errors because Beta regression needs to account for both the bounded response scale and the mean-variance relationship implied by the Beta distribution.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 5.15
::::
::::{.Question-container}
**Multiple Choice**

Which statement about AIC and BIC in Beta regression is most accurate?

**A.** They are most useful for comparing fitted models for the same response.

**B.** They prove that the fitted model is causal.

**C.** They must always be positive.

**D.** They replace the need to inspect residual diagnostics.
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 5.15
::::
::::{.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

**Correct answer:** A.

**Rationale:**

AIC and BIC summarize a trade-off between model fit and model complexity. They are especially useful when comparing candidate models fitted to the same response variable. They do not establish causality, and they do not replace model diagnostics.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 5.16
::::
::::{.Question-container}
**True or False**

If a Beta regression model has a good log-likelihood and a reasonable AIC, then residual diagnostics are unnecessary.
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 5.16
::::
::::{.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

**Correct answer:** False.

**Rationale:**

Overall fit statistics are useful, but they do not show every way a model can fail. Residual diagnostics can reveal patterns such as systematic underprediction, overprediction, unusual observations, or misspecification across parts of the fitted mean range.

For Beta regression, Pearson residuals versus fitted means provide a useful graphical check of whether residuals are centered around zero without obvious structure.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 5.17
::::
::::{.Question-container}
**Open-ended Question**

In the residual-diagnostics plot in @sec-beta-residual-diagnostics, observations appear in vertical bands. Why is this not automatically a warning sign?
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 5.17
::::
::::{.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

Vertical banding can occur when the model includes categorical regressors. Observations with the same categorical pattern can have similar or identical fitted means, so their residuals line up vertically.

The more important diagnostic question is whether the residuals are roughly centered around zero across the fitted mean range and whether there are clear patterns, strong asymmetry, or unusual observations that suggest lack of fit.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 5.18
::::
::::{.Question-container}
**Open-ended Question**

Why should the student-performance case study be interpreted as associational rather than causal?
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 5.18
::::
::::{.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

The dataset is observational. Students were not randomly assigned to study times, tutoring status, sleep patterns, part-time work hours, or living arrangements.

As a result, an estimated association does not prove that changing a regressor would cause a change in final grades. For example, higher study time may be related to motivation, course difficulty, prior preparation, or other variables not included in the model.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 5.19
::::
::::{.Question-container}
**Multiple Choice**

In a fitted Beta regression model for the proportion of hotel rooms occupied each night, suppose weekend-night status is statistically significant but lobby-renovation status is not. Which conclusion is most responsible?

**A.** Lobby renovations have no effect for any hotel in any setting.

**B.** The model provides evidence that weekend-night status is associated with mean room-occupancy proportion, while it does not provide strong evidence of an adjusted lobby-renovation association in this dataset.

**C.** The model proves that weekends cause higher hotel occupancy.

**D.** The fitted Beta regression model should be discarded automatically.
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 5.19
::::
::::{.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

**Correct answer:** B.

**Rationale:**

Statistical significance is evidence about the fitted model and dataset, not a universal truth. A nonsignificant lobby-renovation coefficient means the analysis does not provide strong evidence of an adjusted association for lobby renovation status in this dataset. It does not prove that renovations never matter.

Similarly, a significant weekend-night coefficient supports an adjusted association, but the observational design prevents a direct causal conclusion.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 5.20
::::
::::{.Question-container}
**Open-ended Question**

Imagine you are summarizing the student-performance Beta regression for an academic advising office. What should a good statistical story include, beyond simply listing coefficient estimates and $p$-values?
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 5.20
::::
::::{.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

A good statistical story should identify the main practical message, explain the direction and size of the most important associations, and connect the findings to the advising office's decisions.

For this case study, the story should emphasize that study time is the clearest adjusted association with final grade proportion in the fitted model. It should also explain that other lifestyle and support variables show weaker evidence after adjustment.

Finally, the story should state the limits of the analysis. The data are observational, so the results should not be framed as causal proof. The model can help guide discussion about study routines and support planning, but it should not be used as the only basis for consequential decisions about individual students.

</details>
::::
:::

### Coding Questions {#sec-beta-exercises-coding}

::: {.CodingCase}
:::: {.CodingCase-header}
Problem 1: Restaurant Table Occupancy
::::
:::: {.CodingCase-container}
A restaurant group wants to understand **table occupancy** across its shifts. Each row in the dataset represents one restaurant shift and records the proportion of seats filled during that shift, together with information about timing, weather, patio availability, local events, reservations, staffing, and discounts.

Occupancy rate is a natural setting for Beta regression because the response is **continuous**, **bounded**, and measured strictly between zero and one. A shift with occupancy rate $0.80$ means that, on average, 80% of available seats were filled during that shift. The response is not a count and it is not binary; it is a proportion.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-pixel-coffee-3702057/).](img/coffee.png){width="610"}

The coding questions below ask you to investigate a practical occupancy-planning problem:

1. Formulate the statistical questions.
2. Define the Beta regression model mathematically.
3. Understand and wrangle the variables.
4. Split the data before EDA.
5. Explore the training data for restaurant-specific patterns.
6. Fit and compare Beta regression models for expected occupancy.
7. Check whether the weather-by-patio relationship improves the model.
8. Evaluate held-out point and probabilistic prediction.
9. Communicate the findings in plain language.

@tbl-beta-exercise-occupancy-variables summarizes the variables in this dataset.

| Variable | Description |
|:---:|:---:|
| `occupancy_rate` | Proportion of seats filled during the shift. This is continuous and strictly between 0 and 1. |
| `day_of_week` | Day of the week. |
| `meal_period` | Meal period: `Brunch`, `Lunch`, or `Dinner`. |
| `hour` | Approximate starting hour of the shift using a 24-hour clock. |
| `week_of_year` | Week of the year, from 1 to 52. |
| `weather` | Weather during the shift: `Clear`, `Cloudy`, `Rain`, or `Heat`. |
| `temperature_f` | Outdoor temperature in degrees Fahrenheit. |
| `patio_open` | Whether patio seating was open: `No` or `Yes`. |
| `local_event` | Whether a nearby event occurred during the shift: `No` or `Yes`. |
| `reservation_count` | Number of reservations recorded before the shift. |
| `staff_count` | Number of staff scheduled for the shift. |
| `discount_percent` | Discount percentage offered during the shift. |

: Variables used in the restaurant-occupancy Beta regression exercise. {#tbl-beta-exercise-occupancy-variables .striped .hover}

The response `occupancy_rate` records the proportion of seats filled during each shift. The regressors describe timing, demand conditions, weather, service capacity, and promotions. Because this dataset is simulated for practice, the exercise is mainly about learning the Beta regression workflow rather than making claims about a specific restaurant company.
::::
:::

::: {.Tip}
:::: {.Tip-header}
Tip on libraries for this exercise!
::::
:::: {.Tip-container}
The complete dependencies are listed:

- `R`: [{readr}](https://readr.tidyverse.org), [{tidyverse}](https://www.tidyverse.org), [{ggplot2}](https://ggplot2.tidyverse.org), [{betareg}](https://cran.r-project.org/package=betareg), and [{reticulate}](https://rstudio.github.io/reticulate/).
- `Python`: [{pandas}](https://pandas.pydata.org/), [{numpy}](https://numpy.org/), [{matplotlib}](https://matplotlib.org/), and [{statsmodels}](https://www.statsmodels.org/stable/index.html).
::::
:::

#### Study Design

Before touching the data, translate the restaurant group's operational goal into statistical questions. In this example, the restaurant is not only asking whether occupancy is high or low. It wants to understand which conditions fill seats, whether promotions and patio availability matter, and whether the model can help plan a future shift.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-pixel-action-bound-3704046/).](img/magnifying-glass-2.png){width="420"}

::: {.Question}
:::: {.Question-header}
Question 5.21 — Framing an Occupancy Inquiry
::::
:::: {.Question-container}
Formulate an **inferential inquiry** that is specific to this restaurant occupancy dataset.

Your question should mention at least two of the following restaurant conditions: discounts, reservations, patio availability, weather, local events, and meal period.
::::
:::

::: {.Answer}
:::: {.Answer-header}
Answer 5.21
::::
:::: {.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

A suitable inferential inquiry is:

> **After accounting for timing and reservations, are larger discounts, patio-open shifts, and local events associated with higher expected table occupancy rate?**

This is an inferential question because it asks which shift characteristics are associated with the expected occupancy rate. It is not asking for the exact occupancy of one future shift, and it should not be interpreted causally without a stronger study design.

</details>
::::
:::

::: {.Question}
:::: {.Question-header}
Question 5.22 — A Planning Prediction
::::
:::: {.Question-container}
Formulate a **predictive inquiry** that a manager could use before relying on this model for future shift planning.
::::
:::

::: {.Answer}
:::: {.Answer-header}
Answer 5.22
::::
:::: {.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

A suitable predictive inquiry is:

> **Can a Beta regression model fitted on past shifts predict held-out occupancy rates better than an intercept-only baseline, both for expected occupancy and for the full predictive distribution?**

This question is predictive because it asks whether the fitted model generalizes to shifts that were not used for EDA, model selection, or fitting. It also distinguishes point prediction of expected occupancy from probabilistic prediction using the full fitted Beta distribution.

</details>
::::
:::

::: {.Question}
:::: {.Question-header}
Question 5.23 — Why a Beta Model Fits This Problem
::::
:::: {.Question-container}
Explain why a Beta regression model is appropriate for `occupancy_rate`. Then define the **observational units**, **response variable**, and **random component** mathematically.
::::
:::

::: {.Answer}
:::: {.Answer-header}
Answer 5.23
::::
:::: {.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

The observational units are **restaurant shifts**.

For the $i$th shift, the response variable is

$$
Y_i
=
\text{occupancy rate during the shift}.
$$

The random component is

$$
Y_i \mid \mathbf{x}_i
\sim
\operatorname{Beta}(\mu_i \phi,\ (1-\mu_i)\phi),
$$

where $\mu_i = \mathbb{E}(Y_i \mid \mathbf{x}_i)$ is the expected occupancy rate for the $i$th shift and $\phi > 0$ is the precision parameter. Under this parameterization,

$$
\operatorname{Var}(Y_i \mid \mathbf{x}_i)
=
\frac{\mu_i(1-\mu_i)}{1+\phi}.
$$

The systematic component will be specified after we examine the available regressors and baseline categories.

Beta regression is appropriate because `occupancy_rate` is a continuous proportion. A value such as $0.72$ means 72% of seats were filled on average during the shift. The response is not binary, not a count, and not unbounded.

</details>
::::
:::

#### Data Understanding

The dataset contains **one row per restaurant shift**. Refer to @tbl-beta-exercise-occupancy-variables. Before fitting a model, separate the variables into the parts of the restaurant operation they describe.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-emotion-bored-6230197/).](img/question-2.png){width="300"}

::: {.Question}
:::: {.Question-header}
Question 5.24 — Sorting the Restaurant Variables
::::
:::: {.Question-container}
Using the variables table, sort the variables into the following groups:

1. the response variable;
2. timing variables;
3. demand signals;
4. weather and seating variables;
5. staffing and promotion variables; and
6. categorical variables that will need indicator variables.

Do **not** use software yet.
::::
:::

::: {.Answer}
:::: {.Answer-header}
Answer 5.24
::::
:::: {.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

**Response:**

- `occupancy_rate`.

**Timing variables:**

- `hour`,
- `week_of_year`,
- `day_of_week`,
- `meal_period`.

**Demand signals:**

- `reservation_count`,
- `local_event`.

**Weather and seating variables:**

- `weather`,
- `temperature_f`,
- `patio_open`.

**Staffing and promotion variables:**

- `staff_count`,
- `discount_percent`.

**Categorical variables requiring indicator variables:**

- `day_of_week`,
- `meal_period`,
- `weather`,
- `patio_open`,
- `local_event`.

The categorical variables will require indicator variables once baseline categories are chosen.

</details>
::::
:::

::: {.Question}
:::: {.Question-header}
Question 5.25 — Choosing Meaningful Baselines
::::
:::: {.Question-container}
Choose baseline categories for the categorical regressors and explain what each baseline comparison will mean in this restaurant setting. Then assign notation to the response and the main continuous regressors.
::::
:::

::: {.Answer}
:::: {.Answer-header}
Answer 5.25
::::
:::: {.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

One reasonable baseline choice is:

- `Monday` for `day_of_week`;
- `Lunch` for `meal_period`;
- `Clear` for `weather`;
- `No` for `patio_open`;
- `No` for `local_event`.

With these baselines, coefficients compare other days to Monday shifts, brunch and dinner to lunch shifts, cloudy/rain/heat to clear weather, patio-open to patio-closed shifts, and local-event shifts to shifts without nearby events.

| Variable | Notation |
|:---:|:---:|
| `occupancy_rate` | $Y_i$ |
| `hour` | $x_{i,1}$ |
| `week_of_year` | $x_{i,2}$ |
| `temperature_f` | $x_{i,3}$ |
| `reservation_count` | $x_{i,4}$ |
| `staff_count` | $x_{i,5}$ |
| `discount_percent` | $x_{i,6}$ |
| `day_of_week` | Indicator regressors defined after selecting the baseline |
| `meal_period` | Indicator regressors defined after selecting the baseline |
| `weather` | Indicator regressors defined after selecting the baseline |
| `patio_open` | Indicator regressor defined after selecting the baseline |
| `local_event` | Indicator regressor defined after selecting the baseline |

For later modelling, it may be useful to replace `reservation_count` by $\log(1+\text{reservation_count})$ because reservations are a demand signal that may show diminishing returns once the restaurant is already close to full.

</details>
::::
:::

#### Data Wrangling

Before fitting a Beta regression model, verify that the variables have the appropriate storage types and coding. This exercise also creates two convenient modelling variables: `log_reservations`, which lets the reservation effect flatten out, and `discount_10`, which makes the discount coefficient easier to interpret per 10 percentage points.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-evaluation-measure-up-3976303/).](img/measurement.png){width="450"}

::: {.Question}
:::: {.Question-header}
Question 5.26 — Importing the Shift Records
::::
:::: {.Question-container}
Load `problem-1-data.csv` and store it as `occupancy_raw`. Then:

1. Display the first few observations.
2. Report the number of observations and variables.
3. Inspect the column names and storage types.
4. Confirm that the data describe restaurant shifts rather than individual customers or individual tables.
::::
:::

::: {.Answer}
:::: {.Answer-header}
Answer 5.26
::::
:::: {.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

::: {.panel-tabset}

## **`R`**

```{r}
library(readr)
library(tidyverse)
library(ggplot2)
library(betareg)

occupancy_data_path <- "./05-beta_files/simulations/problem-1-data.csv"
if (!file.exists(occupancy_data_path)) {
  occupancy_data_path <- "../docs/book/05-beta_files/simulations/problem-1-data.csv"
}

occupancy_raw <- read_csv(occupancy_data_path)

head(occupancy_raw)
dim(occupancy_raw)
glimpse(occupancy_raw)
```

## **`Python`**

```{python}
import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
import statsmodels.api as sm
from statsmodels.othermod.betareg import BetaModel
from pathlib import Path

occupancy_data_path = Path("./05-beta_files/simulations/problem-1-data.csv")
if not occupancy_data_path.exists():
    occupancy_data_path = Path("../docs/book/05-beta_files/simulations/problem-1-data.csv")

occupancy_raw = pd.read_csv(occupancy_data_path)

print(occupancy_raw.head())
print(occupancy_raw.shape)
print(occupancy_raw.dtypes)
```

:::

The imported dataset has one row per restaurant shift. The response is stored as a numeric proportion, continuous regressors are numeric, and categorical regressors are stored as character strings before their baseline order is made explicit.

</details>
::::
:::

::: {.Question}
:::: {.Question-header}
Question 5.27 — Preparing the Analysis Table
::::
:::: {.Question-container}
Using `occupancy_raw`:

1. Count missing values in each variable.
2. Drop rows with missing entries and store the result as `occupancy_data`.
3. Verify that `occupancy_rate` is strictly between zero and one.
4. Convert `day_of_week`, `meal_period`, `weather`, `patio_open`, and `local_event` to categorical variables using the baselines from **Question 5.25**.
5. Create `log_reservations = log(1 + reservation_count)`.
6. Create `discount_10 = discount_percent / 10`.
7. Report the number of shifts in each categorical level.
::::
:::

::: {.Answer}
:::: {.Answer-header}
Answer 5.27
::::
:::: {.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

::: {.panel-tabset}

## **`R`**

```{r}
occupancy_missing_summary <- occupancy_raw %>%
  summarize(across(everything(), ~ sum(is.na(.x)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_values"
  )

print(occupancy_missing_summary)

day_levels <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
meal_levels <- c("Lunch", "Brunch", "Dinner")
weather_levels <- c("Clear", "Cloudy", "Rain", "Heat")
yes_no_levels <- c("No", "Yes")

occupancy_data <- occupancy_raw %>%
  drop_na() %>%
  mutate(
    day_of_week = factor(day_of_week, levels = day_levels),
    meal_period = factor(meal_period, levels = meal_levels),
    weather = factor(weather, levels = weather_levels),
    patio_open = factor(patio_open, levels = yes_no_levels),
    local_event = factor(local_event, levels = yes_no_levels),
    log_reservations = log1p(reservation_count),
    discount_10 = discount_percent / 10
  )

occupancy_response_check <- tibble(
  quantity = c(
    "Number of retained shifts",
    "Minimum occupancy rate",
    "Maximum occupancy rate",
    "All rates are strictly between 0 and 1"
  ),
  value = c(
    nrow(occupancy_data),
    min(occupancy_data$occupancy_rate),
    max(occupancy_data$occupancy_rate),
    all(occupancy_data$occupancy_rate > 0 & occupancy_data$occupancy_rate < 1)
  )
)

print(occupancy_response_check)
table(occupancy_data$day_of_week)
table(occupancy_data$meal_period)
table(occupancy_data$weather)
table(occupancy_data$patio_open)
table(occupancy_data$local_event)
```

## **`Python`**

```{python}
occupancy_missing_summary = (
    occupancy_raw
    .isna()
    .sum()
    .rename("missing_values")
    .rename_axis("variable")
    .reset_index()
)

print(occupancy_missing_summary)

day_levels = ["Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"]
meal_levels = ["Lunch", "Brunch", "Dinner"]
weather_levels = ["Clear", "Cloudy", "Rain", "Heat"]
yes_no_levels = ["No", "Yes"]

occupancy_data = occupancy_raw.dropna().copy()
occupancy_data["day_of_week"] = pd.Categorical(
    occupancy_data["day_of_week"],
    categories=day_levels
)
occupancy_data["meal_period"] = pd.Categorical(
    occupancy_data["meal_period"],
    categories=meal_levels
)
occupancy_data["weather"] = pd.Categorical(
    occupancy_data["weather"],
    categories=weather_levels
)
occupancy_data["patio_open"] = pd.Categorical(
    occupancy_data["patio_open"],
    categories=yes_no_levels
)
occupancy_data["local_event"] = pd.Categorical(
    occupancy_data["local_event"],
    categories=yes_no_levels
)
occupancy_data["log_reservations"] = np.log1p(occupancy_data["reservation_count"])
occupancy_data["discount_10"] = occupancy_data["discount_percent"] / 10

occupancy_response_check = pd.DataFrame({
    "quantity": [
        "Number of retained shifts",
        "Minimum occupancy rate",
        "Maximum occupancy rate",
        "All rates are strictly between 0 and 1"
    ],
    "value": [
        len(occupancy_data),
        occupancy_data["occupancy_rate"].min(),
        occupancy_data["occupancy_rate"].max(),
        ((occupancy_data["occupancy_rate"] > 0) & (occupancy_data["occupancy_rate"] < 1)).all()
    ]
})

print(occupancy_response_check)
print(occupancy_data["day_of_week"].value_counts().reindex(day_levels))
print(occupancy_data["meal_period"].value_counts().reindex(meal_levels))
print(occupancy_data["weather"].value_counts().reindex(weather_levels))
print(occupancy_data["patio_open"].value_counts().reindex(yes_no_levels))
print(occupancy_data["local_event"].value_counts().reindex(yes_no_levels))
```

:::

There are a small number of missing values in `weather` and `reservation_count`, so dropping incomplete rows is reasonable here. The retained response values are strictly between zero and one, which satisfies the basic support requirement for standard Beta regression. The variables `log_reservations` and `discount_10` will make the later model more directly connected to the restaurant story: reservations are allowed to show diminishing returns, and discount effects can be read per 10 percentage points.

</details>
::::
:::

#### Exploratory Data Analysis

We now split the observations before making plots. The **training set** supports EDA, model fitting, and model diagnostics. The **testing set** remains untouched until the results stage, where it is used to check whether the occupancy model generalizes to shifts that did not guide the analysis.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-pixel-outlook-success-3704053/).](img/explore.png){width="400"}

::: {.Question}
:::: {.Question-header}
Question 5.28 — Holding Back Future Shifts
::::
:::: {.Question-container}
Using `occupancy_data`:

1. Create an 80/20 training/testing split in `R` using a seed of `123`.
2. Report how many shifts are assigned to training and testing.
3. Via [{reticulate}](https://rstudio.github.io/reticulate/), import the `R`-generated training and testing sets into `Python`.
4. Explain why the patio, discount, and reservation patterns should be explored only in the training data.
::::
:::

::: {.Answer}
:::: {.Answer-header}
Answer 5.28
::::
:::: {.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

::: {.panel-tabset}

## **`R`**

```{r}
set.seed(123)

occupancy_train_index <- sample(
  seq_len(nrow(occupancy_data)),
  size = floor(0.8 * nrow(occupancy_data))
)

occupancy_training_data <- occupancy_data[occupancy_train_index, ]
occupancy_testing_data <- occupancy_data[-occupancy_train_index, ]

occupancy_split_summary <- tibble(
  split = c("Training", "Testing"),
  rows = c(nrow(occupancy_training_data), nrow(occupancy_testing_data)),
  proportion = rows / nrow(occupancy_data)
)

print(occupancy_split_summary)
```

## **`Python`**

```{python}
occupancy_training_data = r.occupancy_training_data.copy().reset_index(drop=True)
occupancy_testing_data = r.occupancy_testing_data.copy().reset_index(drop=True)

categorical_vars = [
    "day_of_week",
    "meal_period",
    "weather",
    "patio_open",
    "local_event"
]
category_levels = {
    "day_of_week": day_levels,
    "meal_period": meal_levels,
    "weather": weather_levels,
    "patio_open": yes_no_levels,
    "local_event": yes_no_levels
}

for occupancy_subset in [occupancy_training_data, occupancy_testing_data]:
    for variable in categorical_vars:
        occupancy_subset[variable] = pd.Categorical(
            occupancy_subset[variable],
            categories=category_levels[variable]
        )

occupancy_split_summary_py = pd.DataFrame({
    "split": ["Training", "Testing"],
    "rows": [len(occupancy_training_data), len(occupancy_testing_data)],
    "proportion": [
        len(occupancy_training_data) / len(occupancy_data),
        len(occupancy_testing_data) / len(occupancy_data)
    ]
})

print(occupancy_split_summary_py)
```

:::

The training data should be used for all exploratory pattern-finding and model development. If the testing data are used to decide which restaurant patterns look important, then the later held-out evaluation is no longer a clean check of performance.

</details>
::::
:::

::: {.Question}
:::: {.Question-header}
Question 5.29 — Looking for Ceiling Pressure in Occupancy
::::
:::: {.Question-container}
Using only `occupancy_training_data`:

1. Summarize the distribution of `occupancy_rate`.
2. Plot a histogram of `occupancy_rate`.
3. Report the share of training shifts with occupancy above 0.90.
4. Explain why a response concentrated near high occupancy can still be compatible with Beta regression.
::::
:::

::: {.Answer}
:::: {.Answer-header}
Answer 5.29
::::
:::: {.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

::: {.panel-tabset}

## **`R`**

```{r}
occupancy_response_summary <- tibble(
  quantity = c(
    "Number of training shifts",
    "Minimum",
    "Median",
    "Mean",
    "Sample variance",
    "Maximum",
    "Share above 0.90"
  ),
  value = c(
    nrow(occupancy_training_data),
    min(occupancy_training_data$occupancy_rate),
    median(occupancy_training_data$occupancy_rate),
    mean(occupancy_training_data$occupancy_rate),
    var(occupancy_training_data$occupancy_rate),
    max(occupancy_training_data$occupancy_rate),
    mean(occupancy_training_data$occupancy_rate > 0.90)
  )
)

print(occupancy_response_summary)

ggplot(occupancy_training_data, aes(x = occupancy_rate)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(
    x = "Occupancy rate",
    y = "Number of shifts",
    title = "Distribution of occupancy rate in the training data"
  )
```

## **`Python`**

```{python}
occupancy_response = occupancy_training_data["occupancy_rate"]

occupancy_response_summary_py = pd.DataFrame({
    "quantity": [
        "Number of training shifts",
        "Minimum",
        "Median",
        "Mean",
        "Sample variance",
        "Maximum",
        "Share above 0.90"
    ],
    "value": [
        len(occupancy_response),
        occupancy_response.min(),
        occupancy_response.median(),
        occupancy_response.mean(),
        occupancy_response.var(ddof=1),
        occupancy_response.max(),
        (occupancy_response > 0.90).mean()
    ]
})

print(occupancy_response_summary_py.round(3))

plt.figure()
plt.hist(occupancy_training_data["occupancy_rate"], bins=30)
plt.xlabel("Occupancy rate")
plt.ylabel("Number of shifts")
plt.title("Distribution of occupancy rate in the training data")
plt.show()
```

:::

The response remains strictly between zero and one, with many shifts having fairly high occupancy. This is compatible with standard Beta regression because the response is a continuous proportion. The concentration near high occupancy is part of the modelling problem, not a reason to use a count or binary model.

</details>
::::
:::

::: {.Question}
:::: {.Question-header}
Question 5.30 — Discounts, Patio Weather, and Reservation Demand
::::
:::: {.Question-container}
Using only `occupancy_training_data`:

1. Plot mean `occupancy_rate` by hour, separated by `meal_period`.
2. Plot mean `occupancy_rate` by `discount_percent`.
3. Plot `occupancy_rate` by `weather` and `patio_open`.
4. Compute the patio lift within each weather category.
5. Plot mean `occupancy_rate` across binned values of `reservation_count`, separated by `local_event`.
6. For each plot, explain the visible pattern and why it matters for the restaurant modelling problem using a small table of findings.
::::
:::

::: {.Answer}
:::: {.Answer-header}
Answer 5.30
::::
:::: {.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

::: {.panel-tabset}

## **`R`**

```{r}
occupancy_hour_summary <- occupancy_training_data %>%
  mutate(hour_bin = floor(hour)) %>%
  group_by(hour_bin, meal_period) %>%
  summarize(mean_occupancy = mean(occupancy_rate), .groups = "drop")

ggplot(
  occupancy_hour_summary,
  aes(x = hour_bin, y = mean_occupancy, color = meal_period, group = meal_period)
) +
  geom_point() +
  geom_line() +
  theme_minimal() +
  theme(legend.position = "top") +
  labs(x = "Hour of shift", y = "Mean occupancy rate", color = "Meal period")

occupancy_discount_summary <- occupancy_training_data %>%
  group_by(discount_percent) %>%
  summarize(mean_occupancy = mean(occupancy_rate), .groups = "drop")

ggplot(occupancy_training_data, aes(x = discount_percent, y = occupancy_rate)) +
  geom_point(alpha = 0.25) +
  geom_line(
    data = occupancy_discount_summary,
    aes(y = mean_occupancy),
    color = "darkorange",
    linewidth = 1.1
  ) +
  geom_point(
    data = occupancy_discount_summary,
    aes(y = mean_occupancy),
    color = "darkorange",
    size = 2
  ) +
  theme_minimal() +
  labs(x = "Discount percent", y = "Occupancy rate")

ggplot(occupancy_training_data, aes(x = weather, y = occupancy_rate, fill = patio_open)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  theme(legend.position = "top") +
  labs(x = "Weather", y = "Occupancy rate", fill = "Patio open")

occupancy_patio_lift <- occupancy_training_data %>%
  group_by(weather, patio_open) %>%
  summarize(mean_occupancy = mean(occupancy_rate), .groups = "drop") %>%
  pivot_wider(names_from = patio_open, values_from = mean_occupancy) %>%
  mutate(patio_lift = Yes - No) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

print(occupancy_patio_lift)

occupancy_training_data %>%
  mutate(reservation_bin = cut(reservation_count, breaks = seq(0, 90, by = 10))) %>%
  group_by(reservation_bin, local_event) %>%
  summarize(mean_occupancy = mean(occupancy_rate), .groups = "drop") %>%
  ggplot(aes(x = reservation_bin, y = mean_occupancy, color = local_event, group = local_event)) +
  geom_point() +
  geom_line() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "top") +
  labs(x = "Reservation count", y = "Mean occupancy rate", color = "Local event")

occupancy_plot_findings <- tibble(
  Plot = c(
    "Hour and meal period",
    "Discount percent",
    "Weather and patio status",
    "Reservation count and local events"
  ),
  Finding = c(
    "Mean occupancy changes across the service day. Brunch and lunch shifts are concentrated earlier, while dinner shifts tend to operate at higher occupancy later in the day. This suggests that both hour and meal period should be included in the model.",
    "Occupancy generally increases as discounts rise, with the largest discounts showing the highest mean occupancy. The pattern is not perfectly linear at every discount level, but it supports treating promotions as a demand signal.",
    "Patio-open shifts usually have higher occupancy than patio-closed shifts, but the size of the gap changes by weather. Rain lowers occupancy overall, while clear or warmer conditions make patio seating more valuable.",
    "Occupancy rises quickly at low to moderate reservation counts and then flattens as shifts approach capacity. Local-event shifts tend to sit above non-event shifts, suggesting extra demand beyond reservations alone."
  )
)

print(occupancy_plot_findings)
```

## **`Python`**

```{python}
hour_means = (
    occupancy_training_data
    .assign(hour_bin=np.floor(occupancy_training_data["hour"]))
    .groupby(["hour_bin", "meal_period"], observed=True)["occupancy_rate"]
    .mean()
    .reset_index()
)

plt.figure()
for period in meal_levels:
    period_means = hour_means.loc[hour_means["meal_period"] == period]
    plt.plot(
        period_means["hour_bin"],
        period_means["occupancy_rate"],
        marker="o",
        label=period
    )
plt.xlabel("Hour of shift")
plt.ylabel("Mean occupancy rate")
plt.legend(title="Meal period", loc="upper center", bbox_to_anchor=(0.5, 1.18), ncol=3)
plt.tight_layout()
plt.show()

discount_means = (
    occupancy_training_data
    .groupby("discount_percent", observed=True)["occupancy_rate"]
    .mean()
)

plt.figure()
plt.scatter(
    occupancy_training_data["discount_percent"],
    occupancy_training_data["occupancy_rate"],
    alpha=0.25
)
plt.plot(discount_means.index, discount_means.values, marker="o")
plt.xlabel("Discount percent")
plt.ylabel("Occupancy rate")
plt.show()

weather_patio_groups = []
weather_patio_labels = []
for weather_level in weather_levels:
    for patio_level in yes_no_levels:
        values = occupancy_training_data.loc[
            (occupancy_training_data["weather"] == weather_level)
            & (occupancy_training_data["patio_open"] == patio_level),
            "occupancy_rate"
        ]
        weather_patio_groups.append(values)
        weather_patio_labels.append(f"{weather_level}\n{patio_level}")

plt.figure(figsize=(10, 5))
plt.boxplot(weather_patio_groups, tick_labels=weather_patio_labels)
plt.xlabel("Weather and patio open")
plt.ylabel("Occupancy rate")
plt.show()

weather_patio_means = pd.pivot_table(
    occupancy_training_data,
    values="occupancy_rate",
    index="weather",
    columns="patio_open",
    aggfunc="mean",
    observed=True
).reindex(index=weather_levels, columns=yes_no_levels)
weather_patio_means["patio_lift"] = (
    weather_patio_means["Yes"] - weather_patio_means["No"]
)
print(weather_patio_means.round(3))

reservation_bins = pd.cut(
    occupancy_training_data["reservation_count"],
    bins=np.arange(0, 91, 10)
)
reservation_means = pd.pivot_table(
    occupancy_training_data,
    values="occupancy_rate",
    index=reservation_bins,
    columns="local_event",
    aggfunc="mean",
    observed=True
).reindex(columns=yes_no_levels)

centers = [interval.mid for interval in reservation_means.index]

plt.figure()
for level in yes_no_levels:
    plt.plot(centers, reservation_means[level], marker="o", label=level)
plt.xlabel("Reservation count")
plt.ylabel("Mean occupancy rate")
plt.legend(title="Local event", loc="upper center", bbox_to_anchor=(0.5, 1.18), ncol=2)
plt.tight_layout()
plt.show()

occupancy_plot_findings_py = pd.DataFrame({
    "Plot": [
        "Hour and meal period",
        "Discount percent",
        "Weather and patio status",
        "Reservation count and local events"
    ],
    "Finding": [
        "Mean occupancy changes across the service day. Brunch and lunch shifts are concentrated earlier, while dinner shifts tend to operate at higher occupancy later in the day. This suggests that both hour and meal period should be included in the model.",
        "Occupancy generally increases as discounts rise, with the largest discounts showing the highest mean occupancy. The pattern is not perfectly linear at every discount level, but it supports treating promotions as a demand signal.",
        "Patio-open shifts usually have higher occupancy than patio-closed shifts, but the size of the gap changes by weather. Rain lowers occupancy overall, while clear or warmer conditions make patio seating more valuable.",
        "Occupancy rises quickly at low to moderate reservation counts and then flattens as shifts approach capacity. Local-event shifts tend to sit above non-event shifts, suggesting extra demand beyond reservations alone."
    ]
})

print(occupancy_plot_findings_py)
```

:::

The first plot points to service timing: occupancy is not constant across the day, and dinner shifts tend to be fuller than earlier service periods. The discount plot suggests that promotions are associated with higher occupancy, although this should be adjusted for other regressors before interpreting it as a modelled association. The weather and patio plot suggests that patio seating is helpful, but not equally helpful under every weather condition, which motivates checking an interaction later. The reservation plot shows demand increasing with reservations and then levelling off near capacity; local events appear to add demand beyond the reservation count alone. These are descriptive findings, not final model conclusions.

</details>
::::
:::

::: {.Question}
:::: {.Question-header}
Question 5.31 — Turning EDA Into a Restaurant Model Plan
::::
:::: {.Question-container}
Based on the training-set EDA, propose a model plan. Explain:

1. why Beta regression remains a reasonable starting model;
2. why `log_reservations` may be preferable to raw `reservation_count`;
3. why a `weather` by `patio_open` interaction is worth checking;
4. which other regressors should be included; and
5. why the testing data should remain unused until the final evaluation.
::::
:::

::: {.Answer}
:::: {.Answer-header}
Answer 5.31
::::
:::: {.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

| EDA component | Main training-set finding | Consequence for the modelling workflow |
|:---:|:---:|:---:|
| Response distribution | Occupancy rate is continuous, bounded, and strictly between zero and one. | Beta regression with a logit link remains a reasonable starting model. |
| Service timing | Occupancy differs by hour and meal period, with dinner shifts tending to run fuller. | Include `hour`, `meal_period`, `week_of_year`, and day-of-week controls. |
| Reservations | Occupancy rises with reservations but appears to flatten. | Use `log_reservations` to allow diminishing returns. |
| Discounts | Larger discounts correspond to higher mean occupancy. | Use `discount_10` so the coefficient is interpretable per 10 percentage points. |
| Weather and patio | Patio-open shifts have higher occupancy, but the size of the gap depends on weather. | Compare an additive model with a model that includes `weather * patio_open`. |
| Events | Local-event shifts tend to have higher occupancy. | Include `local_event` as a demand signal. |
| Data split | The testing data have not been used for EDA. | Keep them untouched for held-out evaluation. |

The EDA supports a Beta regression model with timing, demand, promotion, weather, seating, and staffing regressors. The most restaurant-specific modelling choice is to compare a model that treats weather and patio availability additively with one that lets the patio association differ by weather.

</details>
::::
:::

#### Beta Regression Modelling and Evaluation

The EDA suggests several restaurant conditions that may help explain expected occupancy rate. We now use the training data to estimate Beta regression models with a logit link. The main comparison is whether patio availability should have the same association across all weather categories or whether the patio story changes depending on weather.

::: {.Question}
:::: {.Question-header}
Question 5.32 — Selecting an Occupancy Model
::::
:::: {.Question-container}
Using only `occupancy_training_data`:

1. Write the random component, systematic component, and logit-link formulation for a reduced additive Beta regression model.
2. Write a full model that adds the `weather * patio_open` interaction.
3. Explain why the two models are nested, then state $H_0$ and $H_1$ for a likelihood-ratio test of the interaction.
4. Fit both models in `R` and `Python`, using the training data only.
5. Carry out the likelihood-ratio test at the 5% significance level. Does the larger weather-by-patio model give a statistically supported improvement over the additive model?
6. Report the fitted mean coefficients and exponentiated coefficients for the selected training model.

**Follow-up — check the better model.** If the larger model improves fit, select it and repeat the training diagnostics using this model: check finite estimates and standard errors, the fitted occupancy range, Pearson residuals, and grouped calibration, following **Question 5.33**. Diagnostics from the additive model do not establish that the larger model fits adequately. Otherwise, retain the additive model and proceed with that model in **Question 5.33**. Keep the testing data untouched during this decision and these checks.
::::
:::

::: {.Answer}
:::: {.Answer-header}
Answer 5.32
::::
:::: {.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

Let $Y_i$ be the occupancy rate for shift $i$, and let $\mu_i = \mathbb{E}(Y_i \mid \mathbf{x}_i)$ be the expected occupancy rate for that shift. Use the notation table from **Question 5.25**. Because the fitted model uses the transformed reservation and discount variables, define

$$
\tilde{x}_{i,4} = \log(1 + x_{i,4}),
\qquad
\tilde{x}_{i,6} = \frac{x_{i,6}}{10}.
$$

The random component is

$$
Y_i \mid \mathbf{x}_i
\sim
\operatorname{Beta}(\mu_i \phi,\ (1-\mu_i)\phi),
$$

with logit link

$$
\begin{aligned}
\operatorname{logit}(\mu_i) = \eta_i
&= \beta_0
+ \beta_1 x_{i,1}
+ \beta_2 x_{i,2}
+ \beta_3 x_{i,3}
+ \beta_4 \tilde{x}_{i,4}
+ \beta_5 x_{i,5}
+ \beta_6 \tilde{x}_{i,6}
+ \boldsymbol{\gamma}^{\top}\mathbf{z}_i.
\end{aligned}
$$

Here $\mathbf{z}_i$ contains the indicator regressors for the categorical variables listed in **Question 5.25**. The reference levels are Monday, lunch, clear weather, no patio, and no local event.

This is the reduced model, denoted by $\mathcal{M}_0$. The full model, denoted by $\mathcal{M}_1$, keeps the same terms but adds the weather-by-patio interaction:

$$
\operatorname{logit}(\mu_i)
=
\eta_i^{(0)}
+
\boldsymbol{\delta}^{\top}
\left(
\mathbf{w}_i \odot p_i
\right),
$$

where $\eta_i^{(0)}$ is the additive linear predictor above, $\mathbf{w}_i$ contains the non-baseline weather indicators, $p_i$ is the patio-open indicator, and $\mathbf{w}_i \odot p_i$ contains their interaction terms.

The reduced model is nested within the full model because setting every component of $\boldsymbol{\delta}$ equal to zero removes the interaction and returns the additive model. The likelihood-ratio test evaluates

$$
H_0:\boldsymbol{\delta}=\mathbf{0}
\qquad
\text{versus}
\qquad
H_1:\text{at least one component of }\boldsymbol{\delta}\text{ is nonzero}.
$$

The likelihood-ratio statistic is

$$
G^2
=
2
\left[
\ell(\widehat{\mathcal{M}}_1)
-
\ell(\widehat{\mathcal{M}}_0)
\right],
$$

and is compared with a $\chi^2$ reference distribution whose degrees of freedom are the number of additional interaction coefficients in the full model.

::: {.panel-tabset}

## **`R`**

```{r}
occupancy_additive_formula <- occupancy_rate ~ hour + week_of_year +
  temperature_f + log_reservations + staff_count + discount_10 +
  day_of_week + meal_period + weather + patio_open + local_event

occupancy_additive_beta_model <- betareg(
  occupancy_additive_formula,
  data = occupancy_training_data
)

occupancy_interaction_formula <- occupancy_rate ~ hour + week_of_year +
  temperature_f + log_reservations + staff_count + discount_10 +
  day_of_week + meal_period + weather * patio_open + local_event

occupancy_interaction_beta_model <- betareg(
  occupancy_interaction_formula,
  data = occupancy_training_data
)

occupancy_additive_loglik <- logLik(occupancy_additive_beta_model)
occupancy_interaction_loglik <- logLik(occupancy_interaction_beta_model)

occupancy_weather_patio_lrt_statistic <- 2 * (
  as.numeric(occupancy_interaction_loglik) -
    as.numeric(occupancy_additive_loglik)
)

occupancy_weather_patio_lrt_df <- attr(occupancy_interaction_loglik, "df") -
  attr(occupancy_additive_loglik, "df")

occupancy_weather_patio_lrt_p_value <- pchisq(
  occupancy_weather_patio_lrt_statistic,
  df = occupancy_weather_patio_lrt_df,
  lower.tail = FALSE
)

occupancy_weather_patio_lrt_summary <- tibble(
  Comparison = "Additive model vs weather-by-patio model",
  `LR statistic` = occupancy_weather_patio_lrt_statistic,
  df = occupancy_weather_patio_lrt_df,
  `p-value` = occupancy_weather_patio_lrt_p_value
) %>%
  mutate(
    `LR statistic` = round(`LR statistic`, 4),
    `p-value` = signif(`p-value`, 3)
  )

print(occupancy_weather_patio_lrt_summary)

if (occupancy_weather_patio_lrt_p_value < 0.05) {
  occupancy_selected_formula <- occupancy_interaction_formula
  occupancy_selected_beta_model <- occupancy_interaction_beta_model
  occupancy_selected_model_name <- "Weather-patio Beta"
} else {
  occupancy_selected_formula <- occupancy_additive_formula
  occupancy_selected_beta_model <- occupancy_additive_beta_model
  occupancy_selected_model_name <- "Additive Beta"
}

occupancy_selected_mean_coef <- coef(occupancy_selected_beta_model, model = "mean")

occupancy_selected_coef_table <- tibble(
  term = names(occupancy_selected_mean_coef),
  estimate = occupancy_selected_mean_coef,
  odds_multiplier = exp(estimate)
) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

print(occupancy_selected_model_name)
print(occupancy_selected_coef_table, n = Inf)
```

## **`Python`**

```{python}
model_vars = [
    "hour",
    "week_of_year",
    "temperature_f",
    "log_reservations",
    "staff_count",
    "discount_10",
    "day_of_week",
    "meal_period",
    "weather",
    "patio_open",
    "local_event"
]

categorical_vars = [
    "day_of_week",
    "meal_period",
    "weather",
    "patio_open",
    "local_event"
]

y = occupancy_training_data["occupancy_rate"].astype(float)

X_raw = occupancy_training_data[model_vars]
X_additive_encoded = pd.get_dummies(
    X_raw,
    columns=categorical_vars,
    drop_first=True
).astype(float)
X_additive = sm.add_constant(X_additive_encoded, has_constant="add")

occupancy_additive_beta_model = BetaModel(y, X_additive).fit()

def add_weather_patio_interactions(encoded_data):
    interaction_data = encoded_data.copy()
    patio_col = "patio_open_Yes"
    for weather_level in ["Cloudy", "Rain", "Heat"]:
        weather_col = f"weather_{weather_level}"
        if weather_col in interaction_data.columns and patio_col in interaction_data.columns:
            interaction_data[f"{weather_col}:patio_open_Yes"] = (
                interaction_data[weather_col] * interaction_data[patio_col]
            )
    return interaction_data

X_interaction_encoded = add_weather_patio_interactions(X_additive_encoded)
X_interaction = sm.add_constant(X_interaction_encoded, has_constant="add")

occupancy_interaction_beta_model = BetaModel(y, X_interaction).fit()

from scipy.stats import chi2

occupancy_weather_patio_lrt_statistic_py = 2 * (
    occupancy_interaction_beta_model.llf - occupancy_additive_beta_model.llf
)
occupancy_weather_patio_lrt_df_py = X_interaction.shape[1] - X_additive.shape[1]
occupancy_weather_patio_lrt_p_value_py = chi2.sf(
    occupancy_weather_patio_lrt_statistic_py,
    occupancy_weather_patio_lrt_df_py
)

occupancy_weather_patio_lrt_summary_py = pd.DataFrame({
    "Comparison": ["Additive model vs weather-by-patio model"],
    "LR statistic": [occupancy_weather_patio_lrt_statistic_py],
    "df": [occupancy_weather_patio_lrt_df_py],
    "p-value": [occupancy_weather_patio_lrt_p_value_py]
})

print(occupancy_weather_patio_lrt_summary_py.to_string(
    index=False,
    formatters={"LR statistic": "{:.4f}".format, "p-value": "{:.3g}".format}
))

if occupancy_weather_patio_lrt_p_value_py < 0.05:
    occupancy_selected_beta_model = occupancy_interaction_beta_model
    occupancy_selected_X = X_interaction
    occupancy_selected_X_encoded = X_interaction_encoded
    occupancy_selected_model_name_py = "Weather-patio Beta"
else:
    occupancy_selected_beta_model = occupancy_additive_beta_model
    occupancy_selected_X = X_additive
    occupancy_selected_X_encoded = X_additive_encoded
    occupancy_selected_model_name_py = "Additive Beta"

selected_mean_terms = occupancy_selected_beta_model.params.index[
    :occupancy_selected_X.shape[1]
]

occupancy_selected_coef_table_py = pd.DataFrame({
    "term": selected_mean_terms,
    "estimate": occupancy_selected_beta_model.params.loc[selected_mean_terms],
    "odds_multiplier": np.exp(
        occupancy_selected_beta_model.params.loc[selected_mean_terms]
    )
})

print(occupancy_selected_model_name_py)
print(occupancy_selected_coef_table_py.round(3))
```

:::

For this training split, the likelihood-ratio statistic is approximately $34.5$ on $3$ degrees of freedom, with $p \approx 1.57 \times 10^{-7}$. Thus, the larger weather-by-patio model gives a statistically supported improvement over the additive model and is selected at the 5% level. If the test had not rejected $H_0$, we would have retained the additive model for parsimony.

**Diagnostic follow-up.** Because the larger model is selected here, repeat the training checks using `occupancy_selected_beta_model`, which the code above assigns to the weather-by-patio fit. The code in **Answer 5.33** carries out these checks on the selected model: its estimates and standard errors are finite, fitted occupancy ranges from about $0.142$ to $0.990$, and the five calibration groups have observed and fitted mean occupancy within about $0.007$ of one another. Inspect the Pearson residual plot as well, including the largest absolute residual of about $3.66$, before accepting the model for the next stage; a significant likelihood-ratio test alone does not establish adequate fit. If the additive model were retained, the same code would check that model instead. Only after these training checks do we proceed to the frozen-specification testing analysis and held-out predictive evaluation.

For the selected model, the exponentiated coefficients describe multiplicative changes in the odds of the expected occupancy proportion, $\mu_i/(1-\mu_i)$. Coefficients involving `weather` and `patio_open` should be interpreted relative to the baseline categories from **Question 5.25**, and interaction terms should be interpreted as modifications to the corresponding additive weather and patio comparisons.

</details>
::::
:::

::: {.Question}
:::: {.Question-header}
Question 5.33 — Checking and Freezing the Selected Model
::::
:::: {.Question-container}
Using the selected training model from **Question 5.32**:

1. Verify that the selected model has finite coefficient estimates and model-based standard errors.
2. Check the range of fitted mean occupancy values on the training data.
3. Plot Pearson residuals against fitted mean occupancy values.
4. Create a grouped training calibration check by splitting fitted values into five groups and comparing observed mean occupancy with fitted mean occupancy.
5. Fit the frozen selected specification on `occupancy_testing_data` for coefficient-level inference. Report the mean coefficients, their exponentiated values, and 95% Wald confidence intervals for the exponentiated coefficients to support the inferential answer in **Question 5.35**.
6. Explain what must not be changed after the testing outcomes have been examined.
::::
:::

::: {.Answer}
:::: {.Answer-header}
Answer 5.33
::::
:::: {.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

::: {.panel-tabset}

## **`R`**

```{r}
occupancy_selected_se <- sqrt(diag(vcov(occupancy_selected_beta_model)))

occupancy_selected_diag <- occupancy_training_data %>%
  mutate(
    fitted_occupancy = fitted(occupancy_selected_beta_model),
    pearson_resid = residuals(occupancy_selected_beta_model, type = "pearson")
  )

occupancy_selected_check <- tibble(
  Quantity = c(
    "Selected training model",
    "Finite mean coefficients",
    "Finite model-based SEs",
    "Minimum fitted occupancy",
    "Maximum fitted occupancy",
    "Largest absolute Pearson residual"
  ),
  Value = c(
    occupancy_selected_model_name,
    all(is.finite(coef(occupancy_selected_beta_model, model = "mean"))),
    all(is.finite(occupancy_selected_se)),
    round(min(occupancy_selected_diag$fitted_occupancy), 3),
    round(max(occupancy_selected_diag$fitted_occupancy), 3),
    round(max(abs(occupancy_selected_diag$pearson_resid)), 3)
  )
)

print(occupancy_selected_check)

ggplot(occupancy_selected_diag, aes(x = fitted_occupancy, y = pearson_resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(
    x = "Fitted mean occupancy",
    y = "Pearson residual",
    title = occupancy_selected_model_name
  )

occupancy_selected_training_calibration <- occupancy_selected_diag %>%
  mutate(calibration_group = ntile(fitted_occupancy, 5)) %>%
  group_by(calibration_group) %>%
  summarize(
    Shifts = n(),
    `Mean fitted occupancy` = mean(fitted_occupancy),
    `Observed mean occupancy` = mean(occupancy_rate),
    .groups = "drop"
  )

print(
  occupancy_selected_training_calibration %>%
    mutate(across(where(is.numeric), ~ round(.x, 3)))
)

ggplot(
  occupancy_selected_training_calibration,
  aes(x = `Mean fitted occupancy`, y = `Observed mean occupancy`)
) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  geom_point(size = 2.5) +
  geom_line() +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  theme_minimal() +
  labs(x = "Mean fitted occupancy", y = "Observed mean occupancy")

occupancy_selected_mean_coef <- coef(
  occupancy_selected_beta_model,
  model = "mean"
)

occupancy_weather_patio_terms <- tibble(
  term = names(occupancy_selected_mean_coef),
  estimate = occupancy_selected_mean_coef,
  odds_multiplier = exp(estimate)
) %>%
  filter(str_detect(term, "weather|patio")) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

print(occupancy_weather_patio_terms)

occupancy_testing_inference_beta_model <- betareg(
  occupancy_selected_formula,
  data = occupancy_testing_data
)

occupancy_testing_inference_mean_coef <- coef(
  occupancy_testing_inference_beta_model,
  model = "mean"
)

occupancy_testing_inference_se <- sqrt(
  diag(vcov(occupancy_testing_inference_beta_model))
)[names(occupancy_testing_inference_mean_coef)]

occupancy_testing_inference_coef_table <- tibble(
  term = names(occupancy_testing_inference_mean_coef),
  estimate = occupancy_testing_inference_mean_coef,
  odds_multiplier = exp(estimate),
  odds_lower_95 = exp(estimate - qnorm(0.975) * occupancy_testing_inference_se),
  odds_upper_95 = exp(estimate + qnorm(0.975) * occupancy_testing_inference_se)
) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

print(occupancy_testing_inference_coef_table, n = Inf, width = Inf)
```

## **`Python`**

```{python}
occupancy_selected_diagnostics_py = occupancy_training_data.copy()
occupancy_selected_diagnostics_py["fitted_occupancy"] = (
    occupancy_selected_beta_model.fittedvalues
)
occupancy_selected_diagnostics_py["pearson_resid"] = (
    occupancy_selected_beta_model.resid_pearson
)

occupancy_selected_check_py = pd.DataFrame({
    "Quantity": [
        "Selected training model",
        "Finite coefficients",
        "Finite model-based SEs",
        "Minimum fitted occupancy",
        "Maximum fitted occupancy",
        "Largest absolute Pearson residual"
    ],
    "Value": [
        occupancy_selected_model_name_py,
        np.isfinite(occupancy_selected_beta_model.params).all(),
        np.isfinite(occupancy_selected_beta_model.bse).all(),
        round(occupancy_selected_diagnostics_py["fitted_occupancy"].min(), 3),
        round(occupancy_selected_diagnostics_py["fitted_occupancy"].max(), 3),
        round(occupancy_selected_diagnostics_py["pearson_resid"].abs().max(), 3)
    ]
})

print(occupancy_selected_check_py)

plt.figure()
plt.scatter(
    occupancy_selected_diagnostics_py["fitted_occupancy"],
    occupancy_selected_diagnostics_py["pearson_resid"],
    alpha=0.5
)
plt.axhline(0, color="red", linestyle="--")
plt.xlabel("Fitted mean occupancy")
plt.ylabel("Pearson residual")
plt.title(occupancy_selected_model_name_py)
plt.show()

occupancy_selected_diagnostics_py["calibration_group"] = pd.qcut(
    occupancy_selected_diagnostics_py["fitted_occupancy"].rank(method="first"),
    q=5,
    labels=False
) + 1

occupancy_selected_training_calibration_py = (
    occupancy_selected_diagnostics_py
    .groupby("calibration_group", observed=True)
    .agg(
        Shifts=("occupancy_rate", "size"),
        Mean_fitted_occupancy=("fitted_occupancy", "mean"),
        Observed_mean_occupancy=("occupancy_rate", "mean")
    )
    .reset_index()
)

print(occupancy_selected_training_calibration_py.round(3))

plt.figure()
plt.plot([0, 1], [0, 1], color="red", linestyle="--")
plt.plot(
    occupancy_selected_training_calibration_py["Mean_fitted_occupancy"],
    occupancy_selected_training_calibration_py["Observed_mean_occupancy"],
    marker="o"
)
plt.xlim(0, 1)
plt.ylim(0, 1)
plt.gca().set_aspect("equal", adjustable="box")
plt.xlabel("Mean fitted occupancy")
plt.ylabel("Observed mean occupancy")
plt.show()

selected_mean_terms = occupancy_selected_beta_model.params.index[
    :occupancy_selected_X.shape[1]
]
occupancy_weather_patio_terms_py = pd.DataFrame({
    "term": selected_mean_terms,
    "estimate": occupancy_selected_beta_model.params.loc[selected_mean_terms],
    "odds_multiplier": np.exp(
        occupancy_selected_beta_model.params.loc[selected_mean_terms]
    )
})
occupancy_weather_patio_terms_py = occupancy_weather_patio_terms_py[
    occupancy_weather_patio_terms_py["term"].str.contains("weather|patio")
]
print(occupancy_weather_patio_terms_py.round(3))

test_X_raw = occupancy_testing_data[model_vars]
test_X_additive_encoded = pd.get_dummies(
    test_X_raw,
    columns=categorical_vars,
    drop_first=True
).astype(float)
test_X_additive_encoded = test_X_additive_encoded.reindex(
    columns=X_additive_encoded.columns,
    fill_value=0
)

test_X_interaction_encoded = add_weather_patio_interactions(test_X_additive_encoded)
test_X_interaction_encoded = test_X_interaction_encoded.reindex(
    columns=X_interaction_encoded.columns,
    fill_value=0
)

if occupancy_selected_model_name_py == "Weather-patio Beta":
    test_X_selected_encoded = test_X_interaction_encoded
else:
    test_X_selected_encoded = test_X_additive_encoded

test_X_selected = sm.add_constant(test_X_selected_encoded, has_constant="add")
y_test = occupancy_testing_data["occupancy_rate"].astype(float)

occupancy_testing_inference_beta_model = BetaModel(y_test, test_X_selected).fit()

testing_mean_terms = occupancy_testing_inference_beta_model.params.index[
    :test_X_selected.shape[1]
]
testing_mean_ci = occupancy_testing_inference_beta_model.conf_int().loc[
    testing_mean_terms
]
occupancy_testing_inference_coef_table_py = pd.DataFrame({
    "term": testing_mean_terms,
    "estimate": occupancy_testing_inference_beta_model.params.loc[testing_mean_terms],
    "odds_multiplier": np.exp(
        occupancy_testing_inference_beta_model.params.loc[testing_mean_terms]
    ),
    "odds_lower_95": np.exp(testing_mean_ci.iloc[:, 0]),
    "odds_upper_95": np.exp(testing_mean_ci.iloc[:, 1])
})

print(occupancy_testing_inference_coef_table_py.round(3).to_string(index=False))
```

:::

The selected training model should have fitted mean occupancy values inside the unit interval, finite estimates, and no obvious residual pattern. The grouped calibration plot is an in-sample goodness-of-fit check: points close to the diagonal indicate that the fitted mean occupancy and observed mean occupancy agree reasonably well within fitted-value groups.

Only after the training-stage selection and checks do we fit the frozen selected specification on the testing data for coefficient-level inference. The specification is fixed at this point: after looking at the testing outcomes, we should not add or remove regressors, change transformations, drop the weather-by-patio interaction, or alter reference categories. For the predictive evaluation in **Question 5.34**, we use the training-fitted selected model, not the testing-set refit.

</details>
::::
:::

#### Results

The results stage now evaluates whether the final training-fitted model predicts held-out shifts well. This includes point prediction of the conditional mean and probabilistic prediction from the full fitted Beta distribution.

::: {.Question}
:::: {.Question-header}
Question 5.34 — Predictive Evaluation on Held-Out Shifts
::::
:::: {.Question-container}
Using the training-fitted selected model from **Question 5.32** as the final predictive model:

For this predictive evaluation, do not refit either model using the testing responses. The testing data should only be used to generate predictions and evaluate those predictions.

1. Fit an intercept-only Beta regression baseline using only `occupancy_training_data`.
2. For each testing observation, create a prediction table containing the observed response $y_i$, the predicted conditional mean $\hat{\mu}_i$, the predicted precision $\hat{\phi}_i$, and the fitted Beta shape parameters
   $$
   \hat{\alpha}_i = \hat{\mu}_i\hat{\phi}_i,
   \qquad
   \hat{\beta}_i = (1-\hat{\mu}_i)\hat{\phi}_i.
   $$
3. Compare point-prediction accuracy for the final model and baseline using
   $$
   \operatorname{MAE} = \frac{1}{n_{\text{test}}}\sum_i |y_i - \hat{\mu}_i|
   $$
   and
   $$
   \operatorname{RMSE} =
   \sqrt{
   \frac{1}{n_{\text{test}}}\sum_i (y_i - \hat{\mu}_i)^2
   }.
   $$
4. Compare probabilistic prediction using mean negative log predictive density,
   $$
   \operatorname{NLPD}
   =
   -\frac{1}{n_{\text{test}}}
   \sum_i \log\{f_i(y_i)\},
   $$
   where $f_i$ is the fitted Beta density for testing observation $i$.
5. Construct 90% predictive intervals and report both empirical coverage and average interval width.
6. Make a PIT histogram for the final model, where $u_i = F_i(y_i)$ and $F_i$ is the fitted Beta CDF.
7. Briefly conclude whether the final model improves point prediction and probabilistic prediction over the baseline.
::::
:::

::: {.Answer}
:::: {.Answer-header}
Answer 5.34
::::
:::: {.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

For held-out observation $i$, $\hat{\mu}_i$ is the point prediction for expected occupancy. The fitted Beta distribution uses both $\hat{\mu}_i$ and $\hat{\phi}_i$, so the probabilistic evaluation checks more than the conditional mean. All metrics below are computed on the held-out testing shifts, using models fitted only on the training data.

::: {.panel-tabset}

## **`R`**

```{r}
occupancy_baseline_beta_model <- betareg(
  occupancy_rate ~ 1,
  data = occupancy_training_data
)

evaluate_beta_prediction <- function(model, newdata, observed, model_name) {
  mu_hat <- as.numeric(predict(model, newdata = newdata, type = "response"))
  phi_hat <- as.numeric(predict(model, newdata = newdata, type = "precision"))

  alpha_hat <- mu_hat * phi_hat
  beta_hat <- (1 - mu_hat) * phi_hat

  lower_90 <- qbeta(0.05, shape1 = alpha_hat, shape2 = beta_hat)
  upper_90 <- qbeta(0.95, shape1 = alpha_hat, shape2 = beta_hat)
  pit <- pbeta(observed, shape1 = alpha_hat, shape2 = beta_hat)

  predictions <- tibble(
    observed = observed,
    mu_hat = mu_hat,
    phi_hat = phi_hat,
    alpha_hat = alpha_hat,
    beta_hat = beta_hat,
    lower_90 = lower_90,
    upper_90 = upper_90,
    PIT = pit
  )

  metrics <- tibble(
    Model = model_name,
    MAE = mean(abs(observed - mu_hat)),
    RMSE = sqrt(mean((observed - mu_hat)^2)),
    Mean_NLPD = -mean(dbeta(observed, shape1 = alpha_hat, shape2 = beta_hat, log = TRUE)),
    Coverage_90 = mean(observed >= lower_90 & observed <= upper_90),
    Average_Interval_Width = mean(upper_90 - lower_90)
  )

  list(metrics = metrics, predictions = predictions)
}

baseline_prediction_eval <- evaluate_beta_prediction(
  occupancy_baseline_beta_model,
  occupancy_testing_data,
  occupancy_testing_data$occupancy_rate,
  "Intercept-only Beta"
)

final_prediction_eval <- evaluate_beta_prediction(
  occupancy_selected_beta_model,
  occupancy_testing_data,
  occupancy_testing_data$occupancy_rate,
  occupancy_selected_model_name
)

occupancy_predictive_summary <- bind_rows(
  baseline_prediction_eval$metrics,
  final_prediction_eval$metrics
) %>%
  mutate(across(where(is.numeric), ~ round(.x, 4)))

print(occupancy_predictive_summary)

final_prediction_preview <- final_prediction_eval$predictions %>%
  select(observed, mu_hat, phi_hat, alpha_hat, beta_hat, lower_90, upper_90, PIT) %>%
  head() %>%
  mutate(across(where(is.numeric), ~ round(.x, 4)))

print(final_prediction_preview)

ggplot(final_prediction_eval$predictions, aes(x = PIT)) +
  geom_histogram(
    breaks = seq(0, 1, by = 0.1),
    fill = "gray70",
    color = "white",
    boundary = 0
  ) +
  theme_minimal() +
  labs(
    x = "PIT value",
    y = "Testing shifts",
    title = "Held-out PIT histogram for final Beta model"
  )
```

## **`Python`**

```{python}
from scipy.stats import beta

baseline_train_X = pd.DataFrame({
    "const": np.ones(len(occupancy_training_data))
})
baseline_test_X = pd.DataFrame({
    "const": np.ones(len(occupancy_testing_data))
})

occupancy_baseline_beta_model = BetaModel(y, baseline_train_X).fit()

test_X_raw = occupancy_testing_data[model_vars]
test_X_additive_encoded = pd.get_dummies(
    test_X_raw,
    columns=categorical_vars,
    drop_first=True
).astype(float)
test_X_additive_encoded = test_X_additive_encoded.reindex(
    columns=X_additive_encoded.columns,
    fill_value=0
)
test_X_interaction_encoded = add_weather_patio_interactions(
    test_X_additive_encoded
)
test_X_interaction_encoded = test_X_interaction_encoded.reindex(
    columns=X_interaction_encoded.columns,
    fill_value=0
)
test_X_interaction = sm.add_constant(test_X_interaction_encoded, has_constant="add")
test_X_additive = sm.add_constant(test_X_additive_encoded, has_constant="add")

if occupancy_selected_model_name_py == "Weather-patio Beta":
    test_X_selected = test_X_interaction
else:
    test_X_selected = test_X_additive

def evaluate_beta_prediction_py(model, exog, observed, model_name):
    observed = np.asarray(observed, dtype=float)
    mu_hat = np.asarray(model.predict(exog), dtype=float)
    phi_hat = np.repeat(np.exp(model.params.iloc[-1]), len(observed))

    if mu_hat.shape == () or len(mu_hat) != len(observed):
        mu_hat = np.repeat(np.ravel(mu_hat)[0], len(observed))

    alpha_hat = mu_hat * phi_hat
    beta_hat = (1 - mu_hat) * phi_hat

    lower_90 = beta.ppf(0.05, alpha_hat, beta_hat)
    upper_90 = beta.ppf(0.95, alpha_hat, beta_hat)
    pit = beta.cdf(observed, alpha_hat, beta_hat)

    predictions = pd.DataFrame({
        "observed": observed,
        "mu_hat": mu_hat,
        "phi_hat": phi_hat,
        "alpha_hat": alpha_hat,
        "beta_hat": beta_hat,
        "lower_90": lower_90,
        "upper_90": upper_90,
        "PIT": pit
    })

    metrics = pd.DataFrame({
        "Model": [model_name],
        "MAE": [np.mean(np.abs(observed - mu_hat))],
        "RMSE": [np.sqrt(np.mean((observed - mu_hat) ** 2))],
        "Mean_NLPD": [-np.mean(beta.logpdf(observed, alpha_hat, beta_hat))],
        "Coverage_90": [np.mean((observed >= lower_90) & (observed <= upper_90))],
        "Average_Interval_Width": [np.mean(upper_90 - lower_90)]
    })

    return {"metrics": metrics, "predictions": predictions}

y_test = occupancy_testing_data["occupancy_rate"].astype(float)

baseline_prediction_eval_py = evaluate_beta_prediction_py(
    occupancy_baseline_beta_model,
    baseline_test_X,
    y_test,
    "Intercept-only Beta"
)

final_prediction_eval_py = evaluate_beta_prediction_py(
    occupancy_selected_beta_model,
    test_X_selected,
    y_test,
    occupancy_selected_model_name_py
)

occupancy_predictive_summary_py = pd.concat(
    [
        baseline_prediction_eval_py["metrics"],
        final_prediction_eval_py["metrics"]
    ],
    ignore_index=True
)

print(occupancy_predictive_summary_py.round(4))

print(final_prediction_eval_py["predictions"].head().round(4))

plt.figure()
plt.hist(
    final_prediction_eval_py["predictions"]["PIT"],
    bins=np.linspace(0, 1, 11),
    edgecolor="white"
)
plt.xlabel("PIT value")
plt.ylabel("Testing shifts")
plt.title("Held-out PIT histogram for final Beta model")
plt.show()
```

:::

The final model should be judged against the intercept-only Beta baseline, not against the testing-set mean. Lower MAE and RMSE indicate better held-out point prediction of expected occupancy. Lower mean negative log predictive density indicates a better full predictive distribution because the model places more probability density near the occupancy rates that actually occurred. NLPD can be negative because Beta densities can be greater than 1, so the key comparison is still which model has the lower value on the same held-out response scale.

The 90% predictive interval coverage should be interpreted together with average interval width. Coverage close to 0.90 is desirable, but very wide intervals can cover many observations while still being unhelpful for planning. The PIT histogram should look roughly uniform if the final model's predictive distributions are reasonably calibrated on the held-out shifts.

</details>
::::
:::

#### Storytelling

The storytelling stage translates the technical results into a message that a restaurant operations team could use. The audience does not need every coefficient; it needs the main occupancy levers, the predictive-evaluation results, and the limits of the model.

::: {.Question}
:::: {.Question-header}
Question 5.35 — Telling the Occupancy Planning Story
::::
:::: {.Question-container}
Write a short storytelling summary for a restaurant operations team. Your summary should:

1. lead with the main practical message;
2. explicitly answer the inferential inquiry from **Question 5.21**: after accounting for timing and reservations, are larger discounts, patio-open shifts, and local events associated with higher expected occupancy? Use the fixed-specification testing-set estimates from **Question 5.33**, quantify at least two associations, and include a confidence interval for at least one; mention the reservation association and explain how the patio association depends on weather;
3. explain whether the final training-fitted model improves held-out point and probabilistic prediction over the baseline, citing results from **Question 5.34** such as MAE or RMSE, mean NLPD, and predictive-interval coverage and width;
4. describe one way the model could support staffing or promotion planning; and
5. state one important limitation.

Write in three to five sentences.
::::
:::

::: {.Answer}
:::: {.Answer-header}
Answer 5.35
::::
:::: {.Answer-container}
<details>
<summary><strong>Click here to reveal the answer!</strong></summary>

Use the testing-set refit in **Question 5.33** for the inferential statements and the frozen training fit evaluated in **Question 5.34** for prediction. The multipliers below refer to the odds of expected occupancy, $\mu/(1-\mu)$, holding other regressors fixed; they are not multipliers of occupancy itself. With an interaction, combine the patio coefficient with the relevant weather-by-patio coefficient: for example, the cloudy-weather patio multiplier is approximately $\exp(1.178 - 0.213) = 2.62$. The testing-set confidence intervals for the interaction multipliers include 1, so that refit does not give clear evidence of differences in the patio association across weather categories, despite the training-set selection result.

A short storytelling summary could be:

Reservations and shift conditions can help guide staffing: higher reservations are associated with higher expected occupancy, and, after accounting for timing, reservations, and other regressors, the testing-set fit also supports positive associations for discounts, local events, and patio availability. An extra 10 percentage points of discount is associated with about 1.77 times the odds of expected occupancy (95% confidence interval: 1.66 to 1.89), while local-event shifts have about 1.64 times the odds of shifts without events (95% confidence interval: 1.37 to 1.97). The estimated patio-open versus patio-closed multiplier varies with weather, from about 3.25 in clear weather to 2.62 in cloudy weather, although the testing data leave these weather differences uncertain. For held-out shifts, the training-fitted model reduces mean absolute error from about 17.6 to 6.1 occupancy percentage points and improves the full predictive distribution, with mean negative log predictive density falling from about $-0.35$ to $-1.28$. Its 90% predictive intervals average about 22.4 percentage points wide rather than the baseline's 66.1, but cover only 86.5% of testing shifts, so staffing plans should allow for forecast uncertainty, and these simulated observational associations do not establish that changing discounts or opening the patio causes higher occupancy.

</details>
::::
:::

:::
