<!-- Google tag (gtag.js) -->
<script async src="https://www.googletagmanager.com/gtag/js?id=G-7PRVEBE1EF"></script>
<script>
  window.dataLayer = window.dataLayer || [];
  function gtag(){dataLayer.push(arguments);}
  gtag('js', new Date());

  gtag('config', 'G-7PRVEBE1EF');
</script>

# Negative Binomial Regression {#sec-negative-binomial}

```{r}
#| include: false

colourize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}
```

::::: Warning
::: Warning-header
When to Use and Not Use Negative Binomial Regression
:::

::: Warning-container
Negative Binomial regression is useful when:

- The outcome variable is **discrete** and **count-type**, taking values $0, 1, 2, \dots$.
- The scientific or data-science inquiry focuses on how the expected count changes with a set of regressors.
- The outcome is **overdispersed**, meaning that the variance in the outcome variable is higher than what is plausible for a Classical Poisson model.

However, a negative binomial regression regression should not be used in the following scenarios:

- If the outcome is **binary**, use a [Negative Binomial regression](#sec-negative-binomial) model.
- If the outcome is count-type but the data does not exhibit overdispersion, start with a [Classical Poisson regression](#sec-classical-poisson) model introduced in @sec-classical-poisson.
- If the outcome is a count-type and the data exhibits **underdispersion**, meaning that the the variance in the outcome variable is lower than what is plausible for a Classical Poisson model, use a [Generalized Poisson regression](#sec-generalized-poisson) model.
:::
:::::

::: {#fig-negative-binomial-regression}
```{mermaid}
mindmap
  root((Regression 
  Analysis)
    Continuous <br/>Outcome Y
      {{Unbounded <br/>Outcome Y}}
        )Chapter 3: <br/>Ordinary <br/>Least Squares <br/>Regression(
          (Normal <br/>Outcome Y)
      {{Nonnegative <br/>Outcome Y}}
        )Chapter 4: <br/>Gamma Regression(
          (Gamma <br/>Outcome Y)
      {{Bounded <br/>Outcome Y <br/> between 0 and 1}}
        )Chapter 5: Beta <br/>Regression(
          (Beta <br/>Outcome Y)
      {{Nonnegative <br/>Survival <br/>Time Y}}
        )Chapter 6: <br/>Parametric <br/> Survival <br/>Regression(
          (Exponential <br/>Outcome Y)
          (Weibull <br/>Outcome Y)
          (Lognormal <br/>Outcome Y)
        )Chapter 7: <br/>Semiparametric <br/>Survival <br/>Regression(
          (Cox Proportional <br/>Hazards Model)
            (Hazard Function <br/>Outcome Y)
    Discrete <br/>Outcome Y
      {{Binary <br/>Outcome Y}}
        {{Ungrouped <br/>Data}}
          )Chapter 8: <br/>Binary Logistic <br/>Regression(
            (Bernoulli <br/>Outcome Y)
        {{Grouped <br/>Data}}
          )Chapter 9: <br/>Binomial Logistic <br/>Regression(
            (Binomial <br/>Outcome Y)
      {{Count <br/>Outcome Y}}
        {{Equidispersed <br/>Data}}
          )Chapter 10: <br/>Classical Poisson <br/>Regression(
            (Poisson <br/>Outcome Y)
        {{Overdispersed <br/>Data}}
          )Chapter 11: <br/>Negative Binomial <br/>Regression(
            (Negative Binomial <br/>Outcome Y)
```
:::

::: {.LO}
::::{.LO-header}
Learning Objectives
::::
::::{.LO-container}
By the end of this chapter, you will be able to:

- Determine when Classical Poisson regression is an **appropriate modelling choice**, and recognize situations where a Negative Binomial Regression model is a more appropriate choice.
- Frame **inferential** and **predictive inquiries** for a count-regression problem using the **data science workflow**.
- Specify a Poisson regression model as a generalized linear model with a **Negative Binomial random component**, a **linear predictor**, and a **log link function**.
- Write and interpret the **likelihood** and **log-likelihood functions** for a Negative Binomial regression model.
- Explain how **maximum likelihood estimation** is used to estimate Poisson regression coefficients.
- Interpret Poisson regression coefficients as **multiplicative changes in the expected count**, holding other regressors fixed.
- Assess model adequacy using **goodness-of-fit checks**, **residual-based summaries**, and the **Poisson mean-variance assumption**.
- Construct and interpret **confidence intervals** and **hypothesis tests** for Poisson regression coefficients.
- Evaluate **predictive performance** using test-set prediction accuracy metrics and baseline comparisons.
- Communicate Negative Binomial regression results responsibly for both inferential and predictive inquiries.
::::
:::

```{r setup-r, include=FALSE}
# Load required libraries
library(devtools)
library(readr)
library(tibble)
library(tidyr)
library(dplyr)
library(magrittr)
library(ggplot2)
library(gridExtra) # combine ggplot2 figures
library(ggExtra) # draw marginal histograms
library(broom)
library(knitr)
library(AER) # dispersiontest
library(MASS)

library(reticulate)
py_config()

py_require("pandas")
py_require("numpy")
py_require("seaborn")
py_require("pyreadr")
py_require("plotnine")
py_require("matplotlib")
py_require("statsmodels")
py_require("scikit-learn")
py_require("scipy")
py_require("pyreadr")

reticulate::py_install("plotnine")

pandas <- import("pandas")
numpy <- import("numpy")
seaborn <- import("seaborn")
pyreadr <- import("pyreadr")
matplotlib <- import("matplotlib")
statsmodels <- import("statsmodels")
sklearn <- import("sklearn")
scipy <- import("scipy")
plotnine <- import("plotnine")
pyreadr <- import("pyreadr")

options(htmlwidgets.TOJSON_ARGS = list(na = 'string')) 

print_as_dt <- function(x){
  DT::datatable(x,
                rownames = FALSE,
                options = list(
                  dom = 'ltipr', 
                  autoWidth = TRUE,
                  columnDefs = list(
                    list(className = 'dt-left', targets = "_all"),
                    list(width = '200px', targets = "_all")
                    )
                  )
                )
}

print_as_dt_all <- function(x){
  DT::datatable(
    x,
    rownames = FALSE,
    options = list(
      dom = 't',  # Only show the table (no filter/search/pagination info)
      paging = FALSE,
      scrollX = TRUE,
      autoWidth = TRUE,
      columnDefs = list(
        list(className = 'dt-left', targets = "_all")
        # Optionally remove width limitation or set only for specific columns
      )
    )
  )
}
```

```{python setup-py, include=FALSE}
import numpy as np
from plotnine import *
import statsmodels.api as sm
import statsmodels.formula.api as smf
# from statsmodels.formula.api import glm, ols
from statsmodels.genmod.families import Poisson
from matplotlib import pyplot as plt
from sklearn.metrics import mean_squared_error, mean_absolute_error
import scipy.stats as stats
```

## Introduction {#sec-negative-binomial-introduction}

## Case Study: Horseshoe Crab Satellites

The running case study in this chapter uses data on **horseshoe crab satellites**. At first glance, this dataset may feel quite different from the applied examples that data science students are used to seeing. It is not a web analytics dataset, a health-care dashboard, or a business experiment. However, it is a strong teaching example precisely because it gives us a clean and concrete count-regression problem: 

> **For each female horseshoe crab, we observe the *number of satellite males* around her.**

![Image by [*BarelyDevi*](https://pixabay.com/users/barelydevi-14723734/) via [*Pixabay*](https://pixabay.com/illustrations/crab-sea-seafood-ocean-beach-fish-5030504/).](img/crab.png){width="500"}

In the biological setting, a female horseshoe crab may arrive at a nesting site with an attached male. Other unattached males may gather around the nesting pair and compete for fertilization opportunities. These unattached males are called **satellite males**. Some female crabs have no satellite males nearby, while others have several. Therefore, the response variable is a non-negative integer count. It can be $0$, $1$, $2$, and so on, but it cannot be negative and it is not naturally continuous. This makes the dataset a useful starting point for Poisson regression.

The dataset used in this chapter comes from Brockmann's study of satellite male groups in *Limulus polyphemus* [@brockmann1996]. The version commonly used in statistics teaching contains one row per female crab. For each female crab, we observe the number of satellite males and several characteristics of the female, including body width, body weight, colour category, and spine condition. These variables let us study whether **observable female crab characteristics** are associated with the **expected number of satellite males**.

Note that this case study works well for our cookbook because it is small enough to explain carefully but rich enough to support a complete data science workflow. The response is a count, the regressors include both continuous and categorical variables, and the scientific question(s) can be approached from more than one modelling purpose:

- On the **inferential side**, the model can help us describe how the expected number of satellites is associated with female crab characteristics. 
- On the **predictive side**, the same regressors can be used to predict the expected number of satellites for a female crab and then evaluate how accurate those predictions are. W

We will formalize the two inquiries above in @sec-negative-binomial-study-design.

::: {.Tip}
:::: {.Tip-header}
Tip on why this biological dataset is worth using!
::::
:::: {.Tip-container}
The horseshoe-crab example is more than an old textbook dataset. It is connected to a real biological question about mating behaviour, competition, and observable traits. Brockmann's original study investigated why some nesting females attract larger groups of satellite males than others [@brockmann1996]. Related work has also studied male mating tactics in horseshoe crabs and the behavioural mechanisms behind attached and unattached males [@brockmann2002; @brockmann2009].

The dataset is especially useful for learning because the statistical structure is easy to state even if the biology is unfamiliar. The observational unit is a female crab. The response is the number of satellite males. Moreover, the regressors describe observable characteristics of the female crab. That said, this gives us a compact setting where the data science workflow can remain visible: define the inquiry, understand the data, specify the model, estimate the coefficients, check the model, and communicate what the results mean.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-pixel-nerd-3699334/).](img/feature.png){width="350"}

Horseshoe crabs also matter beyond this single modelling example. Their eggs are ecologically important food resources for migratory shorebirds in coastal systems, and horseshoe crab blood has historically been used in biomedical endotoxin testing [@botton2009; @maloney2018]. This broader context helps make the dataset less like an isolated classroom object and more like a compact entry point into ecological data analysis with count outcomes.

For our purposes, the key statistical lesson is that the response is a count. The number of satellite males can be $0$, $1$, $2$, and so on, but it cannot be negative and it is not naturally continuous. That is exactly the kind of modelling situation where Poisson regression becomes a meaningful starting point.
::::
:::

## Study Design: Framing the Inferential and Predictive Inquiries {#sec-negative-binomial-study-design}

Before fitting a Negative Binomial regression model, we need to clarify what the model is supposed to help us learn. This is the **study design** (as shown in @fig-ds-workflow-study-design) step of the data science workflow introduced in @sec-ds-workflow-study-design. At this stage, we are not yet choosing software functions or interpreting coefficients. Instead, we are defining the modelling purpose, the observational unit, the response variable, and the role of the regressors. 

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-pixel-action-bound-3704046/).](img/magnifying-glass-2.png){width="420"}

Hence, for the horseshoe crab case study, the study-design elements are summarized in @tbl-classical-poisson-study-design-variables. This table fixes the basic modelling ingredients before we decide how the Poisson regression model will be used. Note that @tbl-classical-poisson-study-design-variables is intentionally simple. At this point in the workflow, we are not yet fitting a model or interpreting coefficients. We are just making sure that the observational unit, response variable, and regressors are clear.

| **Role in the study design** | **Variable(s)** | **Description** |
|:---:|:---:|:---:|
| **Observational unit** | Female horseshoe crab | Each row corresponds to one female crab observed in the study. |
| **Response variable** | Number of satellite males | The count outcome: how many satellite males were observed near the female crab. |
| **Main regressors** | Body width, weight, colour category, spine condition | Observable characteristics of the female crab that may be associated with the expected number of satellite males. |

: Key elements of the horseshoe crab study design for the Poisson regression chapter. {#tbl-classical-poisson-study-design-variables .striped .hover}

Note that a single dataset can support different kinds of modelling inquiries. Therefore, we will use the same count response to illustrate two related but distinct modelling purposes: an **inferential inquiry** and a **predictive inquiry** as shown in @tbl-classical-poisson-study-design-inquiries. This distinction is critical since the modelling workflow is not judged in exactly the same way for both purposes.

| **Inquiry type** | **Main question** | **Main modelling emphasis** | **What we will look for later in the chapter** |
|:---:|:---:|:---:|:---:|
| **Inferential inquiry** | How is the expected number of satellite males associated with female crab characteristics? | Coefficients, uncertainty, hypothesis tests, confidence intervals, and interpretation. | Whether the estimated associations are meaningful, how uncertain they are, and whether they can be interpreted responsibly. |
| **Predictive inquiry** | How well can the available regressors predict the expected number of satellite males for a female crab? | Predicted expected counts, test-set accuracy, and comparison against a simple baseline. | Whether the fitted model improves predictive performance beyond a simple benchmark and whether the prediction errors are practically acceptable. |

: Two modelling inquiries supported by the horseshoe crab satellites case study. {#tbl-classical-poisson-study-design-inquiries .striped .hover}

The above inferential and predictive inquiries are connected but not interchangeable. A model can have interpretable coefficients but only modest predictive performance. Conversely, a model can give useful predictions without answering every scientific question about the mechanisms behind satellite male behaviour. Keeping the two inquiries separate helps us decide which results belong in the inferential interpretation and which results belong in the prediction-accuracy discussion.

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on inference versus prediction!
::::
:::: {.Heads-up-container}
For the inferential inquiry, the main object of interest is the relationship between the expected count and the regressors. We will pay attention to coefficient estimates, standard errors, confidence intervals, hypothesis tests, and the assumptions needed for the interpretation to be reasonable.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-emotion-bored-6230197/).](img/question-2.png){width="300"}

On the other hand, for the predictive inquiry, the main object of interest is how well the model predicts expected counts for observations not used to fit the model. We will therefore use a training/testing split, compute predicted expected counts on the test set, compare prediction errors against a baseline model, and summarize prediction accuracy with metrics such as **mean absolute error (MAE)** and **root mean squared error (RMSE)**.

Both inquiries require model checking. If the Poisson model fits the data poorly, then both the inferential interpretation and the predictive results may become less useful.
::::
:::

### Inferential Inquiry {#sec-classical-poisson-inferential-inquiry}

Specifically, the inferential inquiry asks:

> **How is the expected number of satellite males associated with female crab characteristics, holding other regressors fixed?**

This question is about **association and interpretation**. It does not ask whether a female crab characteristic causes satellite males to appear. The data are **observational**, and the case study does not establish a causal design. Instead, the inferential goal is to describe how the expected count of satellite males changes across observed female crab characteristics within a Poisson regression framework.

For example, one inferential question is whether wider female crabs tend to have a larger expected number of satellite males. Later, when we introduce the simple Poisson regression model in @sec-classical-poisson-simple-model, body width will be used as the first regressor because it gives a clear one-regressor entry point into the model. In the extended model in @sec-classical-poisson-additional-regressors, we will add more female crab characteristics so that we can interpret associations while holding other regressors fixed.

For this inquiry, the **results stage** will focus on coefficient estimates, uncertainty, and statistical interpretation. We will ask whether the estimated associations are consistent with the scientific question, whether the confidence intervals are informative, and whether hypothesis tests provide evidence for associations between the regressors and the expected count.

### Predictive Inquiry {#sec-classical-poisson-predictive-inquiry}

Now, the predictive inquiry asks:

> **How well can the available regressors be used to predict the expected number of satellite males for a female crab?**

This question is about **prediction accuracy**. The target of prediction in Poisson regression is the expected count, **not necessarily the exact observed count**. For a given female crab, the observed number of satellite males may differ from the model-predicted expected count because count outcomes are variable even when the model is well specified.

For this inquiry, the **results stage** will focus on out-of-sample prediction. We will fit the model using a training set and evaluate prediction accuracy on a testing set. The testing-set predictions will be compared against a simple baseline that predicts the training-set mean count for every test observation. This comparison is necessary because an accuracy metric is easier to interpret when we can ask whether the regression model does better than a simple benchmark.

The predictive inquiry will use metrics such as MAE and RMSE. MAE summarizes the typical absolute distance between the observed count and the predicted expected count, while RMSE penalizes larger prediction errors more strongly. These metrics will help us judge whether the Poisson regression model is useful for prediction, not only whether its coefficients are interpretable.

## Data Collection and Wrangling

The data used in this chapter are **secondary observational data**. They were originally collected in the study by @brockmann1996 on satellite male groups in horseshoe crabs and later made available in a form commonly used in a textbook on categorical data and count regression [@agresti2013]. Note that we are not designing a new experiment, assigning treatments, or controlling the biological setting. We are working with an existing observational dataset to determine whether female crab characteristics are associated with and can help predict the number of satellite males. Furthermore, the version of the dataset used comes from [Agresti's public GitHub repository](https://github.com/alanagresti/categorical-data/blob/master/Crabs.dat) for categorical-data examples. The GitHub page displays the file in the browser. In the code below, we use the corresponding raw file URL so that `R` and `Python` can import the data directly.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-evaluation-measure-up-3976303/).](img/measurement.png){width="450"}

According to the documentation for the teaching version of the dataset, each row corresponds to one female horseshoe crab [@agresti2013]. The variables record the female crab's colour, spine condition, carapace width, number of satellite males, and weight. The original data also include a binary indicator for whether at least one satellite male was present. That said, in this chapter, our response variable is the **count** of satellite males, so the binary indicator is not the primary outcome.

Now, before doing EDA, we will complete a small amount of data wrangling (as shown in @fig-ds-workflow-data-collection-wrangling). The goal is not to create a training/testing split yet. As in the data science workflow from @sec-intro (more specifically in @sec-ds-workflow-eda), the data split belongs to the EDA stage because we first need to inspect the data structure and then decide how to separate observations for modelling and evaluation. In this section, we only load the data, rename variables, and prepare the categorical regressors so that they are easier to interpret later. The main variables used in the chapter are summarized in @tbl-classical-poisson-wrangled-variables.

| **Variable after wrangling** | **Role** | **Description** |
|:---:|:---:|:---:|
| `satellites` | Response | Number of satellite males observed near the female crab. |
| `width_cm` | Regressor | Female crab carapace width, measured in centimetres. |
| `weight_kg` | Regressor | Female crab weight, measured in kilograms. |
| `color` | Regressor | Female crab colour category, treated as categorical. |
| `spine` | Regressor | Female crab spine condition, treated as categorical. |

: Main variables used after wrangling the horseshoe crab dataset. {#tbl-classical-poisson-wrangled-variables .striped .hover}

The `R` and `Python` code below follow the same wrangling logic. First, the data are read from the public teaching data file associated with the categorical-data examples in @agresti2013. Second, the columns are assigned readable names. Third, the colour and spine variables are converted to categorical variables with descriptive labels. Finally, the dataset is restricted to the variables needed for this chapter. More specifically, we have the following:

- In `R`, via the packages part of [{tidyverse}](https://www.tidyverse.org), we read the raw data directly from Agresti's GitHub repository. The file already includes column names, so we let `read_table()` read the header. Then, we rename the columns needed for this chapter: `sat` becomes `satellites`, `width` becomes `width_cm`, and `weight` becomes `weight_kg`. The variables `color` and `spine` are stored as numeric codes in the raw file, so we convert them into categorical variables with descriptive labels.
- In `Python`, via [{pandas}](https://pypi.org/project/pandas/), we repeats the same wrangling steps. The file already includes column names, so `read_csv()` reads the header directly. Then, `rename()` gives the response and continuous regressors more explicit names. The `map()` calls replace the numeric colour and spine codes with descriptive labels, and the final selection keeps the working variables used in this chapter.

::: {.panel-tabset}

## **`R` Code**

``` {.r}
# Loading library
library(tidyverse)

crabs_url <- paste0(
  "https://raw.githubusercontent.com/",
  "alanagresti/categorical-data/master/Crabs.dat"
)

crabs_raw <- read_table(crabs_url)

crabs <- crabs_raw |>
  transmute(
    satellites = sat,
    width_cm = width,
    weight_kg = weight,
    color = factor(
      color,
      levels = c(1, 2, 3, 4),
      labels = c(
        "Light",
        "Medium light",
        "Medium dark",
        "Dark"
      )
    ),
    spine = factor(
      spine,
      levels = c(1, 2, 3),
      labels = c(
        "Both good",
        "One worn or broken",
        "Both worn or broken"
      )
    )
  )
```

## **`Python` Code**

``` {.python}
# Importing library
import pandas as pd

crabs_url = (
    "https://raw.githubusercontent.com/"
    "alanagresti/categorical-data/master/Crabs.dat"
)

crabs_raw = pd.read_csv(
    crabs_url,
    sep=r"\s+"
)

color_labels = {
    1: "Light",
    2: "Medium light",
    3: "Medium dark",
    4: "Dark",
}

spine_labels = {
    1: "Both good",
    2: "One worn or broken",
    3: "Both worn or broken",

}

crabs = (
    crabs_raw
    .rename(
        columns={
            "sat": "satellites",
            "width": "width_cm",
            "weight": "weight_kg",
        }
    )
    .assign(
        color=lambda data_frame: data_frame["color"].map(color_labels),
        spine=lambda data_frame: data_frame["spine"].map(spine_labels),
    )
    [
        [
            "satellites",
            "width_cm",
            "weight_kg",
            "color",
            "spine",
        ]
    ]
)
```

:::

::: {.panel-tabset}

## **`R` Output**

```{r}
#| label: tbl-crabs-data-r
#| tbl-cap: "Full crabs data."
#| echo: false
#| message: false
#| warning: false

library(tidyverse)

crabs_url <- paste0(
  "https://raw.githubusercontent.com/",
  "alanagresti/categorical-data/master/Crabs.dat"
)

crabs_raw <- read_table(crabs_url)

crabs <- crabs_raw |>
  transmute(
    satellites = sat,
    width_cm = width,
    weight_kg = weight,
    color = factor(
      color,
      levels = c(1, 2, 3, 4),
      labels = c(
        "Light",
        "Medium light",
        "Medium dark",
        "Dark"
      )
    ),
    spine = factor(
      spine,
      levels = c(1, 2, 3),
      labels = c(
        "Both good",
        "One worn or broken",
        "Both worn or broken"
      )
    )
  )
  
# Showing the full dataset
crabs |> 
  print_as_dt()
```

## **`Python` Output**

```{python}
#| label: lst-classical-poisson-data-wrangling-py
#| lst-cap: "Loading and wrangling the horseshoe crab satellite data in Python."
#| echo: false
#| message: false
#| warning: false

import pandas as pd

crabs_url = (
    "https://raw.githubusercontent.com/"
    "alanagresti/categorical-data/master/Crabs.dat"
)

crabs_raw = pd.read_csv(
    crabs_url,
    sep=r"\s+"
)

color_labels = {
    1: "Light",
    2: "Medium light",
    3: "Medium dark",
    4: "Dark",
}

spine_labels = {
    1: "Both good",
    2: "One worn or broken",
    3: "Both worn or broken",

}

crabs = (
    crabs_raw
    .rename(
        columns={
            "sat": "satellites",
            "width": "width_cm",
            "weight": "weight_kg",
        }
    )
    .assign(
        color=lambda data_frame: data_frame["color"].map(color_labels),
        spine=lambda data_frame: data_frame["spine"].map(spine_labels),
    )
    [
        [
            "satellites",
            "width_cm",
            "weight_kg",
            "color",
            "spine",
        ]
    ]
)
```

```{r}
#| label: tbl-crabs-data-py
#| tbl-cap: "Full crabs data."
#| echo: false

# Showing the full dataset
crabs_py <- py$crabs
print_as_dt(crabs_py)
```

:::

After this wrangling step, the object `crabs` is the working dataset for subsequent sections. It contains one row per female crab, one count response, two continuous regressors, and two categorical regressors. The next step is EDA, where we will create the training/testing split needed , inspect the response distribution, and examine the regressors.

## Exploratory Data Analysis

EDA begins after the data have been collected and wrangled into the working object `crabs`. Following the workflow in @fig-ds-workflow-eda, this is also the stage where we create the training/testing split. The training set will be used for EDA, model fitting, and goodness-of-fit checking. These goodness-of-fit checks are applied to the model fitted on the training set because they act as the **workflow gate** before we decide whether the fitted Poisson regression model is adequate enough to use for the chapter's inferential and predictive purposes. Then, the testing set is held aside until the results stage, but it plays different roles for the two inquiries:

- For the **predictive inquiry**, the model fitted on the training set will be used to generate predicted expected counts for the testing observations, and the corresponding prediction accuracy outputs will be reported in the results stage. 
- For the **inferential inquiry**, after the training-set model has passed the goodness-of-fit stage, we will refit the selected model on the testing set to generate the inferential outputs reported in the results stage. This protects the final inferential claims from double dipping: the same observations used for EDA, model fitting, and diagnostics are not reused for the final coefficient-level inferential statements.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-pixel-outlook-success-3704053/).](img/explore.png){width="400"}

Before splitting the data, it is useful to classify the variables that will appear throughout the chapter. For the $i$th female crab, @tbl-classical-poisson-variables summarizes the response and the main regressors. Note that the response variable is the number of satellite males. This is the main reason Poisson regression is a natural starting point: $Y_i$ is a non-negative integer count. Then, the continuous regressors `width_cm` and `weight_kg` will help us examine whether larger female crabs tend to have higher expected numbers of satellites. On the other hand, the categorical regressors `color` and `spine` will help us examine whether visible phenotypic characteristics are associated with the response.

| **Variable** | **Role** | **Type** | **Notation** | **Description** |
|:---:|:---:|:---:|:---:|:---:|
| `satellites` | Response | Discrete count | $Y_i$ | Number of satellite males observed near the female crab. |
| `width_cm` | Regressor | Continuous | $x_{i,1}$ | Female crab carapace width, measured in centimetres. |
| `weight_kg` | Regressor | Continuous | $x_{i,2}$ | Female crab weight, measured in kilograms. |
| `color` | Regressor | Categorical | Encoded in @sec-classical-poisson-categorical-regressors using indicator regressors | Female crab colour category. |
| `spine` | Regressor | Categorical | Encoded in @sec-classical-poisson-categorical-regressors using indicator regressors | Female crab spine condition. |

: Variables used in the horseshoe crab Poisson regression workflow. {#tbl-classical-poisson-variables .striped .hover}

::: {.Tip}
:::: {.Tip-header}
Tip on the 50/50 split and more careful alternatives!
::::
:::: {.Tip-container}
This dataset contains only `r nrow(crabs)` observations. A standard predictive modelling workflow often uses a larger training set, such as an 80/20 training/testing split. However, in this chapter we need to support two goals at the same time: an **inferential** inquiry and a **predictive** inquiry.

Using a 50/50 split is a compromise. It gives the training set enough observations for EDA, model fitting, and goodness-of-fit checking, while also leaving a testing set large enough to support both the final **predictive accuracy assessment** and the final **inferential refit**. This choice is not perfect. A larger training set would usually help estimation and diagnostics, while a larger test set gives a more informative final assessment. With small datasets, we often have to choose between these inconveniences.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-pixel-to-learn-goal-3674110/).](img/mean.png){width="600"}

The 50/50 split used here is also a **simple random split**, which means it does not explicitly protect against imbalance across important variables. For example, if one female crab colour category or spine condition is rare, a simple random split could place too many of those crabs in one subset and too few in the other. That imbalance can affect both the inferential and predictive parts of the workflow.

A more careful split could stratify the data by an important categorical regressor, such as female crab colour, or by a coarser grouping of the response variable. Other strategies include repeated random splits, cross-validation, or bootstrap-based assessment. These approaches can reduce dependence on a single split and are widely used in predictive modelling and model validation; see @hastie2009, @kuhn2013, and @harrell2015 for broader discussions. They are worth trying as an extension, but the main chapter keeps a single 50/50 split so that the Poisson regression workflow remains transparent.
::::
:::

The `R` and `Python` code below first produce independent 50/50 random splits. These two splits use the same conceptual allocation and the same seed value, but they are not expected to select exactly the same observations because `R` and `Python` use different random-number machinery and different splitting implementations:

- @lst-classical-poisson-data-splitting-r uses the [{rsample}](https://rsample.tidymodels.org/) package to split the wrangled `crabs` dataset into training and testing sets. The argument `prop = 0.5` requests that approximately half of the observations be assigned to the training set. The remaining observations are assigned to the testing set. The sanity check prints the dimensions of both subsets and their observed proportions.
- @lst-classical-poisson-data-splitting-py performs the analogous 50/50 split in `Python` using `train_test_split()` from [{scikit-learn}](https://scikit-learn.org/stable/). The argument `test_size = 0.5` assigns approximately half of the observations to the testing set, with the rest assigned to the training set. The output is used only to demonstrate the analogous Python splitting workflow.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| lst-label: lst-classical-poisson-data-splitting-r
#| lst-cap: "Code to randomly split the horseshoe crab data into training and testing sets."
#| message: false
#| warning: false

# Loading libraries
library(rsample)
library(reticulate)

# Seed for reproducibility
set.seed(123)

# Randomly splitting into training and testing sets
crabs_data_splitting <- initial_split(
  crabs,
  prop = 0.5
)

# Assigning data points to training and testing sets
training_data <- training(crabs_data_splitting)
testing_data <- testing(crabs_data_splitting)

# Sanity checks
n_total <- nrow(crabs)
n_train <- nrow(training_data)
n_test <- nrow(testing_data)

cat(sprintf(
  "Training shape: %d %d\nTesting shape:  %d %d\n\nTraining proportion: %.3f\nTesting proportion:  %.3f\n",
  nrow(training_data), ncol(training_data),
  nrow(testing_data),  ncol(testing_data),
  n_train / n_total,
  n_test / n_total
))
```

## **`Python`**

```{python}
#| echo: true
#| lst-label: lst-classical-poisson-data-splitting-py
#| lst-cap: "Code to randomly split the horseshoe crab data into training and testing sets."
#| message: false
#| warning: false

# Importing function
from sklearn.model_selection import train_test_split

# Seed for reproducibility
random_state = 123

# Randomly splitting into training and testing sets
training_data_py_independent, testing_data_py_independent = train_test_split(
    crabs,
    test_size=0.5,
    random_state=random_state
)

# Sanity checks
n_total = len(crabs)
n_train = len(training_data_py_independent)
n_test = len(testing_data_py_independent)

print(
    f"Training shape: {training_data_py_independent.shape}\n"
    f"Testing shape:  {testing_data_py_independent.shape}\n\n"
    f"Training proportion: {n_train / n_total:.3f}\n"
    f"Testing proportion:  {n_test / n_total:.3f}"
)
```

:::

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on keeping `R` and `Python` aligned after the split!
::::
:::: {.Heads-up-container}
Using the same seed and split proportion in `R` and `Python` does not guarantee that the random splits will contain the same crabs. This is not an error. The two ecosystems use different splitting functions and pseudo-random number generators. Therefore, for the rest of this chapter, we use the `R`-generated training and testing sets as a common reference to ensure consistent coding outputs.
::::
:::

Via [{reticulate}](https://rstudio.github.io/reticulate/), @lst-classical-poisson-import-r-split-py imports the `R`-generated training and testing sets into the `Python` environment. This ensures the `R` and `Python` results are comparable since summaries, plots, fitted models, and prediction metrics will be based on the same observations. Henceforth, the `Python` code uses the same `training_data` and `testing_data` objects as the `R` workflow.

::: {.panel-tabset}

## **`Python`**

```{python}
#| echo: true
#| lst-label: lst-classical-poisson-import-r-split-py
#| lst-cap: "Code to import the R-generated training and testing sets into Python."
#| message: false
#| warning: false

# Importing R-generated training and testing sets via reticulate
training_data = r.training_data
testing_data = r.testing_data

# Ensuring categorical regressors are treated as categorical in Python
training_data["color"] = training_data["color"].astype("category")
training_data["spine"] = training_data["spine"].astype("category")
testing_data["color"] = testing_data["color"].astype("category")
testing_data["spine"] = testing_data["spine"].astype("category")
```

:::

The upcoming EDA will use only the training data. In this training split, we have `r nrow(training_data)` female crabs for exploration, model fitting, and goodness-of-fit checking. The testing data are kept aside until the results stage, where they will be used for two different purposes: final prediction accuracy for the predictive inquiry and refitting the selected model for the final inferential outputs.

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on how the split is used later!
::::
:::: {.Heads-up-container}
The same training/testing split supports both inquiry flavours, but the two flavours use the subsets differently after EDA:

- For the **predictive inquiry**, the Poisson regression model fitted on the training set will later be used to predict expected counts for the testing set. These test-set predictions will be used to compute prediction accuracy metrics in the results stage.
- For the **inferential inquiry**, the training set is used to explore the data, fit the candidate model, and run goodness-of-fit checks. If the model is adequate enough to proceed, the selected model specification will then be refit on the testing set to generate the final inferential outputs reported in the results stage. This helps avoid double dipping by separating the data used for exploration and diagnostics from the data used for final inferential reporting.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-pixel-to-learn-3674106/).](img/pieces.png){width="250"}
::::
:::

### Descriptive Summaries {#sec-classical-poisson-eda-general}

We begin with descriptive summaries because they tell us whether the data look compatible with the modelling task:

- For the **response**, we are especially interested in the number of zeros, the typical count, and the relationship between the sample mean and sample variance. 
- For the **regressors**, we want to understand the scale of the continuous variables and the distribution of the categorical variables.

Thus, we have the following:

- @tbl-classical-poisson-training-summary-r shows the training-set summaries for the response and the two continuous regressors. in `R`. For the count response, it reports the mean and variance because the Classical Poisson regression model will later require us to think carefully about the mean-variance relationship.
- @tbl-classical-poisson-training-summary-py shows the same summaries in `Python`. The code uses the `training_data` object imported from `R`, so the values should match the `R` summaries up to formatting.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-classical-poisson-training-summary-r
#| tbl-cap: "Descriptive summary statistics for the horseshoe crab training data."
#| message: false
#| warning: false

# Loading library to display tables
library(knitr)

training_summary <- training_data |>
  summarise(
    `Number of training observations` = n(),
    `Mean number of satellite males` = mean(satellites),
    `Variance of satellite males` = var(satellites),
    `Proportion with zero satellite males` = mean(satellites == 0),
    `Mean female crab width (cm)` = mean(width_cm),
    `SD female crab width (cm)` = sd(width_cm),
    `Mean female crab weight (kg)` = mean(weight_kg),
    `SD female crab weight (kg)` = sd(weight_kg)
  ) |>
  mutate(
    across(
      where(is.numeric),
      ~ round(.x, 2)
    )
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "Quantity",
    values_to = "Value"
  ) |>
  mutate(
    Value = if_else(
      Quantity == "Number of training observations",
      formatC(Value, format = "f", digits = 0),
      formatC(Value, format = "f", digits = 2)
    )
  ) 

training_summary |>
  kable(
    align = c("c", "c")
  )
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

training_summary = pd.DataFrame(
    {
        "Quantity": [
            "Number of training observations",
            "Mean number of satellite males",
            "Variance of satellite males",
            "Proportion with zero satellite males",
            "Mean female crab width (cm)",
            "SD female crab width (cm)",
            "Mean female crab weight (kg)",
            "SD female crab weight (kg)",
        ],
        "Value": [
            len(training_data),
            training_data["satellites"].mean(),
            training_data["satellites"].var(ddof=1),
            (training_data["satellites"] == 0).mean(),
            training_data["width_cm"].mean(),
            training_data["width_cm"].std(ddof=1),
            training_data["weight_kg"].mean(),
            training_data["weight_kg"].std(ddof=1),
        ],
    }
)

training_summary["Value"] = [
    f"{value:.0f}" if quantity == "Number of training observations"
    else f"{value:.2f}"
    for quantity, value in zip(
        training_summary["Quantity"],
        training_summary["Value"]
    )
]

classical_poisson_training_summary_py_html = (
    training_summary.to_html(
        index=False,
        border=0,
    )
)
```

```{r}
#| echo: false
#| label: tbl-classical-poisson-training-summary-py
#| tbl-cap: "Descriptive summary statistics for the horseshoe crab training data."
#| message: false
#| warning: false
#| results: asis

cat(
  "\n",
  py$classical_poisson_training_summary_py_html,
  "\n",
  sep = ""
)
```

:::

The table above gives a quick numerical profile of the training data before we fit any model. The training set contains `r nrow(training_data)` female crabs. The average number of satellite males is `r round(mean(training_data$satellites), 2)`, but the sample variance is `r round(var(training_data$satellites), 2)`, which is noticeably larger than the mean. We will return to this issue more formally in @sec-negative-binomial-overdispersion.

The proportion of crabs with zero satellite males is also important. In the training data, approximately `r round(mean(training_data$satellites == 0), 2)` of female crabs have no satellite males. This does not automatically mean that a Negative Binomial regression model is inappropriate, but it gives us a first reason to pay attention to the number of zeros when checking model adequacy later in the chapter.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-lecture-lecture-hall-3976296/).](img/sample.png){width="500"}

Now, the summaries for `width_cm` and `weight_kg` describe the scale of the continuous regressors. Female crab width has an average of `r round(mean(training_data$width_cm), 2)` cm, while weight has an average of `r round(mean(training_data$weight_kg), 2)` kg. These summaries help us interpret the range of body-size values before we examine whether body size appears associated with the number of satellite males.

Next, we summarize the categorical regressors `color` and `spine`. These summaries are useful because categorical imbalance can affect interpretation and prediction, especially with a small dataset.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-classical-poisson-color-summary-r
#| tbl-cap: "Descriptive summary of female crab colour categories in the training data."
#| message: false
#| warning: false

color_summary <- training_data |>
  count(color, name = "n") |>
  mutate(proportion = round(n / sum(n), 3))

color_summary |>
  rename(
    `Female crab colour` = color,
    `Number of female crabs` = n,
    `Proportion of training data` = proportion
  ) |>
  kable(
    align = c("c", "c", "c")
  )
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

color_summary = (
    training_data["color"]
    .value_counts()
    .rename_axis("color")
    .reset_index(name="n")
)

color_summary["proportion"] = (
    color_summary["n"] / color_summary["n"].sum()
).round(3)

color_summary_display = color_summary.rename(
    columns={
        "color": "Female crab colour",
        "n": "Number of female crabs",
        "proportion": "Proportion of training data",
    }
)

classical_poisson_color_summary_py_html = (
    color_summary_display.to_html(
        index=False,
        border=0,
    )
)
```

```{r}
#| echo: false
#| label: tbl-classical-poisson-color-summary-py
#| tbl-cap: "Descriptive summary of female crab colour categories in the training data."
#| message: false
#| warning: false
#| results: asis

cat(
  "\n",
  py$classical_poisson_color_summary_py_html,
  "\n",
  sep = ""
)
```

:::

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-classical-poisson-spine-summary-r
#| tbl-cap: "Descriptive summary of female crab spine condition categories in the training data."
#| message: false
#| warning: false

spine_summary <- training_data |>
  count(spine, name = "n") |>
  mutate(proportion = round(n / sum(n), 3))

spine_summary |>
  rename(
    `Female crab spine condition` = spine,
    `Number of female crabs` = n,
    `Proportion of training data` = proportion
  ) |>
  kable(
    align = c("c", "c", "c")
  )
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

spine_summary = (
    training_data["spine"]
    .value_counts()
    .rename_axis("spine")
    .reset_index(name="n")
)

spine_summary["proportion"] = (
    spine_summary["n"] / spine_summary["n"].sum()
).round(3)

spine_summary_display = spine_summary.rename(
    columns={
        "spine": "Female crab spine condition",
        "n": "Number of female crabs",
        "proportion": "Proportion of training data",
    }
)

classical_poisson_spine_summary_py_html = (
    spine_summary_display.to_html(
        index=False,
        border=0,
    )
)
```

```{r}
#| echo: false
#| label: tbl-classical-poisson-spine-summary-py
#| tbl-cap: "Descriptive summary of female crab spine condition categories in the training data."
#| message: false
#| warning: false
#| results: asis

cat(
  "\n",
  py$classical_poisson_spine_summary_py_html,
  "\n",
  sep = ""
)
```

:::

The above categorical summaries show that the training data **are not evenly distributed** across the female crab colour and spine categories. For colour, most crabs are in the `Medium light` category, followed by `Medium dark`, while the `Light` and `Dark` categories are less common. For spine condition, most crabs are in the `Both worn or broken` category, with fewer observations in the `Both good` and `One worn or broken` categories. Sparse categories can make coefficient estimates less stable and can make category-level comparisons harder to interpret. For now, we treat these summaries as exploratory checks; later, the model will help us assess whether colour and spine condition appear associated with the expected number of satellite males after accounting for the other regressors.

### Distribution of the Count Response {#sec-classical-poisson-eda-response}

We now examine the distribution of the response variable `satellites` in the **training data**. This is the first EDA output that directly speaks to the Poisson regression model. Before thinking about regressors, we need to understand the count we are trying to model: 

- *how often zero counts occur*, 
- *how concentrated the response is around small values*, 
- *whether there are unusually large counts*, and
- *whether the observed variability is already suggesting a possible goodness-of-fit issue*.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-teaching-to-learn-wiki-3976301/).](img/board.png){width="500"}

The response summaries, in @tbl-classical-poisson-response-summary-r, give a compact numerical view before we look at the response plot. The most important quantities for the Poisson regression story are the **mean**, the **variance**, the **proportion of zeros**, and the **maximum observed count**.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-classical-poisson-response-summary-r
#| tbl-cap: "Descriptive summary statistics for the distribution of male satellite counts in the training data."
#| message: false
#| warning: false

response_summary <- training_data |>
  summarise(
    `Number of training observations` = n(),
    `Number of zero satellite counts` = sum(satellites == 0),
    `Proportion of zero satellite counts` = mean(satellites == 0),
    `Minimum number of satellite males` = min(satellites),
    `Median number of satellite males` = median(satellites),
    `Mean number of satellite males` = mean(satellites),
    `Variance of satellite males` = var(satellites),
    `Variance-to-mean ratio` = var(satellites) / mean(satellites),
    `Maximum number of satellite males` = max(satellites)
  ) |>
  mutate(
    across(
      where(is.numeric),
      ~ round(.x, 2)
    )
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "Quantity",
    values_to = "Value"
  ) |>
  mutate(
    Value = if_else(
      Quantity %in% c(
        "Proportion of zero satellite counts",
        "Mean number of satellite males",
        "Variance of satellite males",
        "Variance-to-mean ratio"
      ),
      formatC(Value, format = "f", digits = 2),
      formatC(Value, format = "f", digits = 0)
    )
  )

response_summary |>
  kable(
    align = c("c", "c")
  )
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

response_summary = pd.DataFrame(
    {
        "Quantity": [
            "Number of training observations",
            "Number of zero satellite counts",
            "Proportion of zero satellite counts",
            "Minimum number of satellite males",
            "Median number of satellite males",
            "Mean number of satellite males",
            "Variance of satellite males",
            "Variance-to-mean ratio",
            "Maximum number of satellite males",
        ],
        "Value": [
            len(training_data),
            (training_data["satellites"] == 0).sum(),
            (training_data["satellites"] == 0).mean(),
            training_data["satellites"].min(),
            training_data["satellites"].median(),
            training_data["satellites"].mean(),
            training_data["satellites"].var(ddof=1),
            (
                training_data["satellites"].var(ddof=1)
                / training_data["satellites"].mean()
            ),
            training_data["satellites"].max(),
        ],
    }
)

two_decimal_rows = [
    "Proportion of zero satellite counts",
    "Mean number of satellite males",
    "Variance of satellite males",
    "Variance-to-mean ratio",
]

response_summary["Value"] = [
    f"{value:.2f}" if quantity in two_decimal_rows
    else f"{value:.0f}"
    for quantity, value in zip(
        response_summary["Quantity"],
        response_summary["Value"]
    )
]

classical_poisson_response_summary_py_html = (
    response_summary.to_html(
        index=False,
        border=0,
    )
)
```

```{r}
#| echo: false
#| label: tbl-classical-poisson-response-summary-py
#| tbl-cap: "Descriptive summary statistics for the distribution of male satellite counts in the training data."
#| message: false
#| warning: false
#| results: asis

cat(
  "\n",
  py$classical_poisson_response_summary_py_html,
  "\n",
  sep = ""
)
```

:::

The above response summaries already tell us several important things. First, **zero counts are common**: `r sum(training_data$satellites == 0)` out of `r nrow(training_data)` female crabs in the training set have no satellite males, corresponding to a proportion of `r round(mean(training_data$satellites == 0), 2)`. Second, **the response is right-skewed**: the median number of satellite males is `r median(training_data$satellites)`, while the maximum observed training-set count is `r max(training_data$satellites)`. Third, **the sample variance, `r round(var(training_data$satellites), 2)`, is much larger than the sample mean, `r round(mean(training_data$satellites), 2)`**. The variance-to-mean ratio is approximately `r round(var(training_data$satellites) / mean(training_data$satellites), 2)`, which is an early warning sign that the Classical Poisson mean-variance structure will need to be checked carefully in @sec-classical-poisson-equidispersion.

Note that the bar plot in @fig-classical-poisson-response-distribution-r (or @fig-classical-poisson-response-distribution-py) gives the same information visually. A bar plot is more appropriate than a smooth density plot because the response is a count. Each bar corresponds to a possible number of satellite males.

::: {.panel-tabset}

## **`R`**

```{r}
#| label: fig-classical-poisson-response-distribution-r
#| fig-cap: "Distribution of the number of satellite males in the training data."
#| warning: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

response_distribution_plot <- ggplot(training_data, aes(x = satellites)) +
  geom_bar(
    fill = "#0072B2",
    colour = "white",
    linewidth = 0.3
  ) +
  scale_x_continuous(
    breaks = seq(
      min(training_data$satellites),
      max(training_data$satellites),
      by = 1
    )
  ) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 15.5),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(
      size = 20,
      vjust = 0.5,
      margin = margin(r = 12)
    ),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x = "\n Number of satellite males",
    y = "Number of female crabs"
  ) +
  coord_cartesian(ylim = c(0, 35)) +
  scale_y_continuous(breaks = seq(0, 35, by = 5))

response_distribution_plot
```

## **`Python`**

```{python}
#| label: fig-classical-poisson-response-distribution-py
#| fig-cap: "Distribution of the number of satellite males in the training data."
#| warning: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

# Importing library
import matplotlib.pyplot as plt

satellite_counts = (
    training_data["satellites"]
    .value_counts()
    .sort_index()
)

response_distribution_figure, response_distribution_axis = plt.subplots(
    figsize=(14, 8)
)

response_distribution_axis.bar(
    satellite_counts.index,
    satellite_counts.values,
    edgecolor="white",
    linewidth=0.3,
    color="#0072B2"
)

response_distribution_axis.set_xlabel(
    "\n Number of satellite males",
    fontsize=20
)

response_distribution_axis.set_ylabel(
    "Number of female crabs",
    fontsize=20,
    labelpad=12
)

response_distribution_axis.tick_params(axis="both", labelsize=15.5)

response_distribution_axis.set_xticks(
    range(
        int(training_data["satellites"].min()),
        int(training_data["satellites"].max()) + 1
    )
)

response_distribution_axis.grid(
    True,
    which="major",
    axis="both",
    alpha=0.3
)

response_distribution_axis.grid(False, which="minor")

response_distribution_figure.tight_layout()
plt.show()
```

:::

The bar plot confirms that the training-set response is concentrated at zero and relatively small positive counts, with a thinner right tail extending to larger numbers of satellite males. This is critical for the **inferential inquiry** because the model must describe how the expected count changes with female crab characteristics while respecting the count scale. It is also important for the **predictive inquiry** because exact observed counts may be difficult to predict: many crabs have no satellites, while a smaller number have much larger counts. Therefore, later predictive results should focus on predicted **expected counts** and prediction errors, not on pretending that the model can deterministically identify the exact count for each crab.

We have to stress that, at this stage, the above response distribution does not prove that Poisson regression is inappropriate. It tells us what to look for. The large variance relative to the mean suggests **possible overdispersion**, and the visible number of zeros indicates we should later compare observed and fitted counts carefully. These issues will become part of the goodness-of-fit story before we decide whether the Classical Poisson model is adequate enough for the chapter's two inquiries.

### Satellite Counts and Continuous Regressors {#sec-classical-poisson-eda-continuous}

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-problem-technology-3974187/).](img/plug-2.png){width="400"}

The next part of the EDA examines the response against the two continuous regressors: **female crab width** and **female crab weight**. The inferential inquiry asks whether the expected number of satellite males is associated with female crab characteristics, so these plots are the first visual check of whether body size appears relevant. The predictive inquiry also benefits from this step, as a useful predictive model requires regressors that carry information about the response. Hence, we first summarize the response across **quartiles of female crab width and weight**. These tables are not formal model outputs. They are descriptive summaries that help us read the plots more carefully.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-classical-poisson-width-quartile-summary-r
#| tbl-cap: "Descriptive summary statistics of male satellite counts across width quartiles in the training data."
#| message: false
#| warning: false

width_quartile_summary <- training_data |>
  mutate(
    width_quartile = ntile(width_cm, 4),
    width_quartile = factor(
      width_quartile,
      levels = 1:4,
      labels = c("Q1: smallest widths", "Q2", "Q3", "Q4: largest widths")
    )
  ) |>
  group_by(width_quartile) |>
  summarise(
    n = n(),
    mean_width_cm = mean(width_cm),
    mean_satellites = mean(satellites),
    median_satellites = median(satellites),
    proportion_zero_satellites = mean(satellites == 0),
    .groups = "drop"
  ) |>
  mutate(
    across(
      where(is.numeric),
      ~ round(.x, 2)
    )
  )

width_quartile_summary |>
  rename(
    `Width quartile` = width_quartile,
    `Number of female crabs` = n,
    `Mean width (cm)` = mean_width_cm,
    `Mean number of satellite males` = mean_satellites,
    `Median number of satellite males` = median_satellites,
    `Proportion with zero satellite males` = proportion_zero_satellites
  ) |>
  kable(
    align = c("c", "c", "c", "c", "c", "c")
  )
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

width_quartile_data = training_data.copy()

width_quartile_data = (
    width_quartile_data
    .sort_values(
        "width_cm",
        kind="mergesort"
    )
    .reset_index(drop=True)
)

number_of_rows = len(width_quartile_data)
number_of_groups = 4

base_group_size = number_of_rows // number_of_groups
remainder = number_of_rows % number_of_groups

group_sizes = [
    base_group_size + 1 if group_index < remainder else base_group_size
    for group_index in range(number_of_groups)
]

quartile_labels = [
    "Q1: smallest widths",
    "Q2",
    "Q3",
    "Q4: largest widths",
]

width_quartile_data["width_quartile"] = np.repeat(
    quartile_labels,
    group_sizes
)

width_quartile_data["width_quartile"] = pd.Categorical(
    width_quartile_data["width_quartile"],
    categories=quartile_labels,
    ordered=True
)

width_quartile_summary = (
    width_quartile_data
    .groupby("width_quartile", observed=False)
    .agg(
        n=("satellites", "size"),
        mean_width_cm=("width_cm", "mean"),
        mean_satellites=("satellites", "mean"),
        median_satellites=("satellites", "median"),
        proportion_zero_satellites=("satellites", lambda x: (x == 0).mean()),
    )
    .reset_index()
    .round(2)
)

width_quartile_summary_display = width_quartile_summary.rename(
    columns={
        "width_quartile": "Width quartile",
        "n": "Number of female crabs",
        "mean_width_cm": "Mean width (cm)",
        "mean_satellites": "Mean number of satellite males",
        "median_satellites": "Median number of satellite males",
        "proportion_zero_satellites": (
            "Proportion with zero satellite males"
        ),
    }
)

classical_poisson_width_quartile_summary_py_html = (
    width_quartile_summary_display.to_html(
        index=False,
        border=0,
    )
)
```

```{r}
#| echo: false
#| label: tbl-classical-poisson-width-quartile-summary-py
#| tbl-cap: "Descriptive summary statistics of male satellite counts across width quartiles in the training data."
#| message: false
#| warning: false
#| results: asis

cat(
  "\n",
  py$classical_poisson_width_quartile_summary_py_html,
  "\n",
  sep = ""
)
```

:::


The above width-quartile summary shows a clear **positive descriptive pattern** between female crab width and the number of satellite males: 

- In the **smallest-width quartile**, female crabs have an average of `r width_quartile_summary$mean_satellites[width_quartile_summary$width_quartile == "Q1: smallest widths"]` satellite males, with a median of `r width_quartile_summary$median_satellites[width_quartile_summary$width_quartile == "Q1: smallest widths"]`. 
- In the **largest-width quartile**, the average increases to `r width_quartile_summary$mean_satellites[width_quartile_summary$width_quartile == "Q4: largest widths"]`, with a median of `r width_quartile_summary$median_satellites[width_quartile_summary$width_quartile == "Q4: largest widths"]`. 
- The **middle quartiles** also follow the same general ordering: the mean satellite count rises from `r width_quartile_summary$mean_satellites[width_quartile_summary$width_quartile == "Q2"]` in Q2 to `r width_quartile_summary$mean_satellites[width_quartile_summary$width_quartile == "Q3"]` in Q3. 

This monotone increase across width quartiles suggests that female crab width is an important body-size regressor to examine first. Now, let us proceed with the width plots.

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on smoothing and count-response scatterplots!
::::
:::: {.Heads-up-container}
In both @fig-classical-poisson-satellites-width-r (or @fig-classical-poisson-satellites-width-py), the orange curve is a **LOWESS/LOESS** smooth. **LOWESS** stands for **locally weighted scatterplot smoothing**, and **LOESS** stands for **locally estimated scatterplot smoothing**. Both refer to closely related smoothing methods that draw a flexible trend line through a scatterplot by fitting many small local regressions, each using observations close to a given value on the horizontal axis.

In this chapter, we use the LOWESS/LOESS curve only as a visual guide. It is not the Poisson regression model, and we will not interpret it as a fitted statistical model. Its purpose is simply to help us see whether satellite counts tend to be higher for wider female crabs.

Because we show this plot in both `R` and `Python`, we make the smoothing settings explicit. Although `R`'s `geom_smooth(method = "loess")` and Python's `lowess()` from [{statsmodels}](https://www.statsmodels.org/stable/index.html) are based on closely related ideas, their default settings are not identical. To make the two fitted curves visually comparable, we use the same smoothing fraction in both languages and align the local regression structure as closely as possible.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-emotion-pressure-6230208/).](img/surprise.png){width="475"}

Since `satellites` is a count response, the points naturally fall on horizontal bands at values such as $0$, $1$, $2$, and so on. These horizontal bands are useful: they remind us that the response is discrete, not continuous. When a continuous regressor such as `width_cm` is placed on the $x$-axis and a count response is placed on the $y$-axis, this banded structure is exactly what we should expect.
::::
:::

::: {.panel-tabset}

## **`R`**

```{r}
#| label: fig-classical-poisson-satellites-width-r
#| fig-cap: "Number of satellite males versus female crab width in the training data."
#| warning: false
#| message: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

satellites_width_plot <- ggplot(training_data, aes(x = width_cm, y = satellites)) +
  geom_point(
    alpha = 0.65,
    size = 2.3,
    colour = "#0072B2"
  ) +
  geom_smooth(
    method = "loess",
    formula = y ~ x,
    se = FALSE,
    linewidth = 1.2,
    colour = "#D55E00",
    method.args = list(
      span = 0.75,
      degree = 1,
      family = "gaussian"
    )
  )  +
  theme_bw() +
  theme(
    axis.text = element_text(size = 15.5),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(
      size = 20,
      vjust = 0.5,
      margin = margin(r = 12)
    ),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x = "\n Female crab width (cm)",
    y = "Number of satellite males"
  ) +
  scale_x_continuous(breaks = seq(22, 34, by = 2)) +
  scale_y_continuous(breaks = seq(0, 12, by = 2))

satellites_width_plot
```

## **`Python`**

```{python}
#| label: fig-classical-poisson-satellites-width-py
#| fig-cap: "Number of satellite males versus female crab width in the training data."
#| warning: false
#| message: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

# Importing function
from statsmodels.nonparametric.smoothers_lowess import lowess

width_x = training_data["width_cm"]
sat_y = training_data["satellites"]

smooth_width = lowess(
    endog=sat_y,
    exog=width_x,
    frac=0.75,
    it=0,
    return_sorted=True
)

satellites_width_figure, satellites_width_axis = plt.subplots(
    figsize=(14, 8)
)

satellites_width_axis.scatter(
    width_x,
    sat_y,
    alpha=0.65,
    s=35,
    color="#0072B2"
)

satellites_width_axis.plot(
    smooth_width[:, 0],
    smooth_width[:, 1],
    linewidth=1.2,
    color="#D55E00"
)

satellites_width_axis.set_xlabel(
    "\n Female crab width (cm)",
    fontsize=20
)

satellites_width_axis.set_ylabel(
    "Number of satellite males",
    fontsize=20,
    labelpad=12
)

satellites_width_axis.tick_params(axis="both", labelsize=15.5)
satellites_width_axis.set_xticks(range(22, 35, 2))
satellites_width_axis.set_yticks(range(0, 13, 2))
satellites_width_axis.grid(
    True,
    which="major",
    axis="both",
    alpha=0.3
)

satellites_width_axis.grid(False, which="minor")

satellites_width_figure.tight_layout()
plt.show()
```

:::

The width plot (from @fig-classical-poisson-satellites-width-r or @fig-classical-poisson-satellites-width-py) should be read together with the quartile (from @tbl-classical-poisson-width-quartile-summary-r). The blue points show the observed counts in the training set. Moreover, the orange LOWESS/LOESS curve is only an exploratory visual guide, but it suggests that **satellite counts tend to increase as female crab width increases**. This agrees with the quartile summary: the average satellite count rises from `r width_quartile_summary$mean_satellites[width_quartile_summary$width_quartile == "Q1: smallest widths"]` in the smallest-width quartile to `r width_quartile_summary$mean_satellites[width_quartile_summary$width_quartile == "Q4: largest widths"]` in the largest-width quartile. Then, the vertical spread around this increasing pattern is also important. Female crabs with similar widths can still have different numbers of satellite males. Some moderate-to-large crabs have zero or very few satellites, while some crabs have much larger counts. This tells us that width is a promising first regressor, but it will not fully explain the response on its own. We will use this EDA result to motivate the first Poisson regression model (see @sec-classical-poisson-simple-model), while leaving formal model specification and interpretation for the next sections.

We now repeat the same EDA logic for female crab weight. Since width and weight both describe body size, we should expect them to tell related but not necessarily identical stories.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-classical-poisson-weight-quartile-summary-r
#| tbl-cap: "Descriptive summary statistics of male satellite counts across weight quartiles in the training data."
#| message: false
#| warning: false

weight_quartile_summary <- training_data |>
  mutate(
    weight_quartile = ntile(weight_kg, 4),
    weight_quartile = factor(
      weight_quartile,
      levels = 1:4,
      labels = c("Q1: smallest weights", "Q2", "Q3", "Q4: largest weights")
    )
  ) |>
  group_by(weight_quartile) |>
  summarise(
    n = n(),
    mean_weight_kg = mean(weight_kg),
    mean_satellites = mean(satellites),
    median_satellites = median(satellites),
    proportion_zero_satellites = mean(satellites == 0),
    .groups = "drop"
  ) |>
  mutate(
    across(
      where(is.numeric),
      ~ round(.x, 2)
    )
  )

weight_quartile_summary |>
  rename(
    `Weight quartile` = weight_quartile,
    `Number of female crabs` = n,
    `Mean weight (kg)` = mean_weight_kg,
    `Mean number of satellite males` = mean_satellites,
    `Median number of satellite males` = median_satellites,
    `Proportion with zero satellite males` = proportion_zero_satellites
  ) |>
  kable(
    align = c("c", "c", "c", "c", "c", "c")
  )
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

weight_quartile_data = training_data.copy()

weight_quartile_data = (
    weight_quartile_data
    .sort_values(
        "weight_kg",
        kind="mergesort"
    )
    .reset_index(drop=True)
)

number_of_rows = len(weight_quartile_data)
number_of_groups = 4

base_group_size = number_of_rows // number_of_groups
remainder = number_of_rows % number_of_groups

group_sizes = [
    base_group_size + 1 if group_index < remainder else base_group_size
    for group_index in range(number_of_groups)
]

quartile_labels = [
    "Q1: smallest weights",
    "Q2",
    "Q3",
    "Q4: largest weights",
]

weight_quartile_data["weight_quartile"] = np.repeat(
    quartile_labels,
    group_sizes
)

weight_quartile_data["weight_quartile"] = pd.Categorical(
    weight_quartile_data["weight_quartile"],
    categories=quartile_labels,
    ordered=True
)

weight_quartile_summary = (
    weight_quartile_data
    .groupby("weight_quartile", observed=False)
    .agg(
        n=("satellites", "size"),
        mean_weight_kg=("weight_kg", "mean"),
        mean_satellites=("satellites", "mean"),
        median_satellites=("satellites", "median"),
        proportion_zero_satellites=("satellites", lambda x: (x == 0).mean()),
    )
    .reset_index()
    .round(2)
)

weight_quartile_summary_display = weight_quartile_summary.rename(
    columns={
        "weight_quartile": "Weight quartile",
        "n": "Number of female crabs",
        "mean_weight_kg": "Mean weight (kg)",
        "mean_satellites": "Mean number of satellite males",
        "median_satellites": "Median number of satellite males",
        "proportion_zero_satellites": (
            "Proportion with zero satellite males"
        ),
    }
)

classical_poisson_weight_quartile_summary_py_html = (
    weight_quartile_summary_display.to_html(
        index=False,
        border=0,
    )
)
```

```{r}
#| echo: false
#| label: tbl-classical-poisson-weight-quartile-summary-py
#| tbl-cap: "Descriptive summary statistics of male satellite counts across weight quartiles in the training data."
#| message: false
#| warning: false
#| results: asis

cat(
  "\n",
  py$classical_poisson_weight_quartile_summary_py_html,
  "\n",
  sep = ""
)
```

:::

The weight-quartile summary also points toward a **positive body-size pattern**:

- In the **smallest-weight quartile**, the average satellite count is `r weight_quartile_summary$mean_satellites[weight_quartile_summary$weight_quartile == "Q1: smallest weights"]`, with a median of `r weight_quartile_summary$median_satellites[weight_quartile_summary$weight_quartile == "Q1: smallest weights"]`. 
- In the **largest-weight quartile**, the average satellite count is `r weight_quartile_summary$mean_satellites[weight_quartile_summary$weight_quartile == "Q4: largest weights"]`, with a median of `r weight_quartile_summary$median_satellites[weight_quartile_summary$weight_quartile == "Q4: largest weights"]`. 

Thus, in the training data, heavier female crabs tend to have higher satellite counts. However, this is still an exploratory pattern, **not a biological conclusion**. @tbl-classical-poisson-weight-quartile-summary-r groups a continuous regressor into four broad bins, and it does not adjust for other female crab characteristics. In particular, the weight summary should be interpreted alongside the width summary, not in isolation. Width and weight are both body-size measurements, so part of the apparent weight pattern may overlap with the width pattern. This is one reason the chapter begins with a simple width-only model before considering an extended model with additional regressors. We will treat weight as a candidate body-size regressor to examine later, while leaving formal interpretation to the fitted regression model.

::: {.panel-tabset}

## **`R`**

```{r}
#| label: fig-classical-poisson-satellites-weight-r
#| fig-cap: "Number of satellite males versus female crab weight in the training data."
#| warning: false
#| message: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

satellites_weight_plot <- ggplot(
  training_data,
  aes(x = weight_kg, y = satellites)
) +
  geom_point(
    alpha = 0.65,
    size = 2.3,
    colour = "#0072B2"
  ) +
  geom_smooth(
    method = "loess",
    formula = y ~ x,
    se = FALSE,
    linewidth = 1.2,
    colour = "#D55E00",
    method.args = list(
      span = 0.75,
      degree = 1,
      family = "gaussian"
    )
  ) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 15.5),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(
      size = 20,
      vjust = 0.5,
      margin = margin(r = 12)
    ),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x = "\n Female crab weight (kg)",
    y = "Number of satellite males"
  ) +
  scale_x_continuous(breaks = seq(1, 5.5, by = 0.5)) +
  scale_y_continuous(breaks = seq(0, 12, by = 2))

satellites_weight_plot
```

## **`Python`**

```{python}
#| label: fig-classical-poisson-satellites-weight-py
#| fig-cap: "Number of satellite males versus female crab weight in the training data."
#| warning: false
#| message: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

weight_x = training_data["weight_kg"]
sat_y = training_data["satellites"]

smooth_weight = lowess(
    endog=sat_y,
    exog=weight_x,
    frac=0.75,
    it=0,
    return_sorted=True
)

satellites_weight_figure, satellites_weight_axis = plt.subplots(
    figsize=(14, 8)
)

satellites_weight_axis.scatter(
    weight_x,
    sat_y,
    alpha=0.65,
    s=35,
    color="#0072B2"
)

satellites_weight_axis.plot(
    smooth_weight[:, 0],
    smooth_weight[:, 1],
    linewidth=1.2,
    color="#D55E00"
)

satellites_weight_axis.set_xlabel(
    "\n Female crab weight (kg)",
    fontsize=20
)

satellites_weight_axis.set_ylabel(
    "Number of satellite males",
    fontsize=20,
    labelpad=12
)

satellites_weight_axis.tick_params(axis="both", labelsize=15.5)
satellites_weight_axis.set_xticks(np.arange(1, 5.6, 0.5))
satellites_weight_axis.set_yticks(range(0, 13, 2))

satellites_weight_axis.grid(
    True,
    which="major",
    axis="both",
    alpha=0.3
)

satellites_weight_axis.grid(False, which="minor")

satellites_weight_figure.tight_layout()
plt.show()
```

:::

The weight plot, in @fig-classical-poisson-satellites-weight-r or @fig-classical-poisson-satellites-weight-py, should be read with the same caution as the width plot. The horizontal bands again show that the response is a count, not a continuous outcome. The orange smooth gives an exploratory body-size trend, and the quartile table gives the corresponding numerical summary. The main EDA message is that **heavier female crabs tend to have higher satellite counts**, but the relationship is not deterministic. There are still zero counts and small positive counts across much of the weight range. Therefore, weight may add useful information in an extended model, but it should not be treated as a complete explanation of satellite counts.

### Satellite Counts and Categorical Regressors {#sec-classical-poisson-eda-categorical}

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-pixel-college-3702064/).](img/plug.png){width="500"}

Now, we compare satellite counts across the categorical regressors `color` and `spine`. Earlier, in @sec-classical-poisson-eda-general, the marginal summaries showed that these categories are not evenly represented in the training data. That imbalance is critical here because a category with few observations can have an unstable mean satellite count. Therefore, in this subsection, we combine two pieces of information: the number of observations in each category and the distribution of satellite counts within each category.

We begin with female crab colour.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-classical-poisson-color-satellite-summary-r
#| tbl-cap: "Descriptive summary statistics of male satellite counts by female crab colour in the training data."
#| message: false
#| warning: false

color_satellite_summary <- training_data |>
  group_by(color) |>
  summarise(
    n = n(),
    mean_satellites = mean(satellites),
    median_satellites = median(satellites),
    proportion_zero_satellites = mean(satellites == 0),
    .groups = "drop"
  ) |>
  mutate(
    across(
      where(is.numeric),
      ~ round(.x, 2)
    )
  )

color_satellite_summary |>
  rename(
    `Female crab colour` = color,
    `Number of female crabs` = n,
    `Mean number of satellite males` = mean_satellites,
    `Median number of satellite males` = median_satellites,
    `Proportion with zero satellite males` = proportion_zero_satellites
  ) |>
  kable(
    align = c("c", "c", "c", "c", "c")
  )
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

color_satellite_summary = (
    training_data
    .groupby("color", observed=False)
    .agg(
        n=("satellites", "size"),
        mean_satellites=("satellites", "mean"),
        median_satellites=("satellites", "median"),
        proportion_zero_satellites=("satellites", lambda x: (x == 0).mean()),
    )
    .reset_index()
    .round(2)
)

color_satellite_summary_display = color_satellite_summary.rename(
    columns={
        "color": "Female crab colour",
        "n": "Number of female crabs",
        "mean_satellites": "Mean number of satellite males",
        "median_satellites": "Median number of satellite males",
        "proportion_zero_satellites": (
            "Proportion with zero satellite males"
        ),
    }
)

classical_poisson_color_satellite_summary_py_html = (
    color_satellite_summary_display.to_html(
        index=False,
        border=0,
    )
)
```

```{r}
#| echo: false
#| label: tbl-classical-poisson-color-satellite-summary-py
#| tbl-cap: "Descriptive summary statistics of male satellite counts by female crab colour in the training data."
#| message: false
#| warning: false
#| results: asis

cat(
  "\n",
  py$classical_poisson_color_satellite_summary_py_html,
  "\n",
  sep = ""
)
```

:::

The colour summary, in @tbl-classical-poisson-color-satellite-summary-r, must be read in light of the category sizes reported earlier. In this training split, the most common colour category is `Medium light`, while `Light` and `Dark` are much less common. The satellite-count summaries suggest that colour may be associated with the response, but the evidence is uneven across categories. The category with the highest mean satellite count is ``r as.character(color_satellite_summary$color[which.max(color_satellite_summary$mean_satellites)])``, with a mean of `r max(color_satellite_summary$mean_satellites)`. The category with the lowest mean satellite count is ``r as.character(color_satellite_summary$color[which.min(color_satellite_summary$mean_satellites)])``, with a mean of `r min(color_satellite_summary$mean_satellites)`. These are descriptive differences only. **Because some colour categories are sparse, the extended model should treat colour comparisons cautiously.**

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on horizontal jitter in categorical plots!
::::
:::: {.Heads-up-container}
In the subsequent plots, the orange points are slightly **jittered horizontally** within each female crab category. This means that points are moved a little to the left or right so that overlapping observations are easier to see. Note that this horizontal jitter is only a plotting device. It does **not** change the observed number of satellite males. The vertical positions of the points remain the original count values, so the horizontal bands at $0$, $1$, $2$, and so on are preserved. These bands are still important because they remind us that `satellites` is a count response.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-emotion-fear-expression-6230192/).](img/curious.png){width="250"}

We use a seed in the plotting code so that the small left-right movements are reproducible. In other words, the same code produces the same visual arrangement each time the chapter is rendered.
::::
:::

::: {.panel-tabset}

## **`R`**

```{r}
#| label: fig-classical-poisson-satellites-color-r
#| fig-cap: "Number of satellite males by female crab colour in the training data."
#| warning: false
#| message: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

set.seed(123)

satellites_color_plot <- ggplot(
  training_data,
  aes(x = color, y = satellites)
) +
  geom_boxplot(
    fill = "#0072B2",
    alpha = 0.45,
    outlier.shape = NA
  ) +
  geom_point(
    position = position_jitter(
      width = 0.18,
      height = 0,
      seed = 123
    ),
    alpha = 0.65,
    size = 2.2,
    colour = "#D55E00"
  ) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 15.5),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(
      size = 20,
      vjust = 0.5,
      margin = margin(r = 12)
    ),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x = "\n Female crab colour",
    y = "Number of satellite males"
  ) +
  scale_y_continuous(breaks = seq(0, 12, by = 2))

satellites_color_plot
```

## **`Python`**

```{python}
#| label: fig-classical-poisson-satellites-color-py
#| fig-cap: "Number of satellite males by female crab colour in the training data."
#| warning: false
#| message: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

rng = np.random.default_rng(123)

color_order = [
    category
    for category in ["Light", "Medium light", "Medium dark", "Dark"]
    if category in list(training_data["color"].astype(str).unique())
]

color_data = [
    training_data.loc[
        training_data["color"].astype(str) == category,
        "satellites"
    ].to_numpy()
    for category in color_order
]

satellites_color_figure, satellites_color_axis = plt.subplots(
    figsize=(14, 8)
)

color_box = satellites_color_axis.boxplot(
    color_data,
    tick_labels=color_order,
    patch_artist=True,
    showfliers=False
)

for patch in color_box["boxes"]:
    patch.set_facecolor("#0072B2")
    patch.set_alpha(0.45)

for index, category in enumerate(color_order, start=1):
    y_values = training_data.loc[
        training_data["color"].astype(str) == category,
        "satellites"
    ].to_numpy()

    x_values = index + rng.uniform(
        -0.18,
        0.18,
        size=len(y_values)
    )

    satellites_color_axis.scatter(
        x_values,
        y_values,
        alpha=0.65,
        s=30,
        color="#D55E00"
    )

satellites_color_axis.set_xlabel(
    "\n Female crab colour",
    fontsize=20
)

satellites_color_axis.set_ylabel(
    "Number of satellite males",
    fontsize=20,
    labelpad=12
)

satellites_color_axis.tick_params(axis="both", labelsize=15.5)
satellites_color_axis.set_yticks(range(0, 13, 2))

satellites_color_axis.grid(
    True,
    which="major",
    axis="both",
    alpha=0.3
)

satellites_color_axis.grid(False, which="minor")

satellites_color_figure.tight_layout()
plt.show()
```

:::

The colour plot, in @fig-classical-poisson-satellites-color-r or @fig-classical-poisson-satellites-color-py, reinforces the table's cautionary message. The boxplots and horizontally jittered points show **substantial overlap** in satellite counts across colour categories. This means colour may still be useful in an extended model, but it is unlikely to separate the response cleanly on its own. The horizontal jitter is used only to reduce overlap within each colour category; it does not move the counts vertically, so the integer count bands remain visible.

We now repeat the same comparison for female crab spine condition.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-classical-poisson-spine-satellite-summary-r
#| tbl-cap: "Descriptive summary statistics of male satellite counts by female crab spine condition in the training data."
#| message: false
#| warning: false

spine_satellite_summary <- training_data |>
  group_by(spine) |>
  summarise(
    n = n(),
    mean_satellites = mean(satellites),
    median_satellites = median(satellites),
    proportion_zero_satellites = mean(satellites == 0),
    .groups = "drop"
  ) |>
  mutate(
    across(
      where(is.numeric),
      ~ round(.x, 2)
    )
  )

spine_satellite_summary |>
  rename(
    `Female crab spine condition` = spine,
    `Number of female crabs` = n,
    `Mean number of satellite males` = mean_satellites,
    `Median number of satellite males` = median_satellites,
    `Proportion with zero satellite males` = proportion_zero_satellites
  ) |>
  kable(
    align = c("c", "c", "c", "c", "c")
  )
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

spine_satellite_summary = (
    training_data
    .groupby("spine", observed=False)
    .agg(
        n=("satellites", "size"),
        mean_satellites=("satellites", "mean"),
        median_satellites=("satellites", "median"),
        proportion_zero_satellites=("satellites", lambda x: (x == 0).mean()),
    )
    .reset_index()
    .round(2)
)

spine_satellite_summary_display = spine_satellite_summary.rename(
    columns={
        "spine": "Female crab spine condition",
        "n": "Number of female crabs",
        "mean_satellites": "Mean number of satellite males",
        "median_satellites": "Median number of satellite males",
        "proportion_zero_satellites": (
            "Proportion with zero satellite males"
        ),
    }
)

classical_poisson_spine_satellite_summary_py_html = (
    spine_satellite_summary_display.to_html(
        index=False,
        border=0,
    )
)
```

```{r}
#| echo: false
#| label: tbl-classical-poisson-spine-satellite-summary-py
#| tbl-cap: "Descriptive summary statistics of male satellite counts by female crab spine condition in the training data."
#| message: false
#| warning: false
#| results: asis

cat(
  "\n",
  py$classical_poisson_spine_satellite_summary_py_html,
  "\n",
  sep = ""
)
```

:::

The spine summary, in @tbl-classical-poisson-spine-satellite-summary-r, is dominated by the `Both worn or broken` category, which was already the largest group in the marginal spine table. This imbalance is important. The category with the highest mean satellite count is ``r as.character(spine_satellite_summary$spine[which.max(spine_satellite_summary$mean_satellites)])``, with a mean of `r max(spine_satellite_summary$mean_satellites)`. The category with the lowest mean satellite count is ``r as.character(spine_satellite_summary$spine[which.min(spine_satellite_summary$mean_satellites)])``, with a mean of `r min(spine_satellite_summary$mean_satellites)`. However, because at least one spine category has relatively few observations, these descriptive differences should not be over-interpreted before modelling.

::: {.panel-tabset}

## **`R`**

```{r}
#| label: fig-classical-poisson-satellites-spine-r
#| fig-cap: "Number of satellite males by female crab spine condition in the training data."
#| warning: false
#| message: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

set.seed(123)

satellites_spine_plot <- ggplot(
  training_data,
  aes(x = spine, y = satellites)
) +
  geom_boxplot(
    fill = "#0072B2",
    alpha = 0.45,
    outlier.shape = NA
  ) +
  geom_point(
    position = position_jitter(
      width = 0.18,
      height = 0,
      seed = 123
    ),
    alpha = 0.65,
    size = 2.2,
    colour = "#D55E00"
  ) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 15.5),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(
      size = 20,
      vjust = 0.5,
      margin = margin(r = 12)
    ),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x = "\n Female crab spine condition",
    y = "Number of satellite males"
  ) +
  scale_y_continuous(breaks = seq(0, 12, by = 2))

satellites_spine_plot
```

## **`Python`**

```{python}
#| label: fig-classical-poisson-satellites-spine-py
#| fig-cap: "Number of satellite males by female crab spine condition in the training data."
#| warning: false
#| message: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

spine_order = [
    category
    for category in [
        "Both good",
        "One worn or broken",
        "Both worn or broken",
    ]
    if category in list(training_data["spine"].astype(str).unique())
]

spine_data = [
    training_data.loc[
        training_data["spine"].astype(str) == category,
        "satellites"
    ].to_numpy()
    for category in spine_order
]

satellites_spine_figure, satellites_spine_axis = plt.subplots(
    figsize=(14, 8)
)

spine_box = satellites_spine_axis.boxplot(
    spine_data,
    tick_labels=spine_order,
    patch_artist=True,
    showfliers=False
)

for patch in spine_box["boxes"]:
    patch.set_facecolor("#0072B2")
    patch.set_alpha(0.45)

for index, category in enumerate(spine_order, start=1):
    y_values = training_data.loc[
        training_data["spine"].astype(str) == category,
        "satellites"
    ].to_numpy()

    x_values = index + rng.uniform(
        -0.18,
        0.18,
        size=len(y_values)
    )

    satellites_spine_axis.scatter(
        x_values,
        y_values,
        alpha=0.65,
        s=30,
        color="#D55E00"
    )

satellites_spine_axis.set_xlabel(
    "\n Female crab spine condition",
    fontsize=20
)

satellites_spine_axis.set_ylabel(
    "Number of satellite males",
    fontsize=20,
    labelpad=12
)

satellites_spine_axis.tick_params(axis="both", labelsize=15.5)
satellites_spine_axis.set_yticks(range(0, 13, 2))

satellites_spine_axis.grid(
    True,
    which="major",
    axis="both",
    alpha=0.3
)

satellites_spine_axis.grid(False, which="minor")

satellites_spine_figure.tight_layout()
plt.show()
```

:::

The spine plot, in @fig-classical-poisson-satellites-spine-r or @fig-classical-poisson-satellites-spine-py, again shows **overlap** across categories. The largest category contains crabs with a wide range of satellite counts, including zero counts and larger positive counts. The smaller categories are harder to compare because they contain fewer observations. This makes spine a reasonable candidate regressor for the extended model, but not one that should be expected to dominate the body-size pattern seen for width and weight.

### Summary {#sec-classical-poisson-eda-summary}

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-blended-learning-6230153/).](img/blender.png){width="400"}

Before moving to the GLM formulation of Poisson regression, it is useful to summarize the main EDA findings. @tbl-classical-poisson-eda-summary records what we learned from the **training data** and how each finding will guide the next parts of the workflow.

| EDA component | Main finding from the training data | Why it matters for the Poisson regression workflow |
|:---:|:---:|:---:|
| Response distribution | The response `satellites` is a **non-negative integer count**. The training data contain `r sum(training_data$satellites == 0)` zero counts out of `r nrow(training_data)` observations. The mean number of satellite males is `r round(mean(training_data$satellites), 2)`, while the variance is `r round(var(training_data$satellites), 2)`, giving a variance-to-mean ratio of `r round(var(training_data$satellites) / mean(training_data$satellites), 2)`. | Poisson regression is a **natural starting point** because the response is a count. However, the variance is much larger than the mean, so the Classical Poisson mean-variance structure will need to be checked carefully later. |
| Count-response plot | The bar plot shows **many zero and small positive counts**, with a **thinner right tail** extending to larger satellite counts. | This reinforces that we are modelling a **discrete, right-skewed count response**. The large counts and the visible number of zeros motivate later observed-versus-fitted checks. |
| Width and satellite counts | The width-quartile summary shows a **clear positive descriptive pattern**: the mean satellite count rises from `r width_quartile_summary$mean_satellites[width_quartile_summary$width_quartile == "Q1: smallest widths"]` in the smallest-width quartile to `r width_quartile_summary$mean_satellites[width_quartile_summary$width_quartile == "Q4: largest widths"]` in the largest-width quartile. The width scatterplot shows the same general upward pattern, while preserving the horizontal bands created by the count response. | Width has the clearest descriptive body-size pattern in the EDA, so it is a sensible **first regressor for the simple Poisson regression model**. The vertical spread at similar widths also warns us that width alone will not fully explain the response. |
| Weight and satellite counts | The weight-quartile summary also suggests a **positive descriptive association**: the mean satellite count changes from `r weight_quartile_summary$mean_satellites[weight_quartile_summary$weight_quartile == "Q1: smallest weights"]` in the smallest-weight quartile to `r weight_quartile_summary$mean_satellites[weight_quartile_summary$weight_quartile == "Q4: largest weights"]` in the largest-weight quartile. The weight plot shows a related body-size pattern, again with substantial variability around the visual trend. | Weight is a **candidate body-size regressor for the extended model**. However, width and weight should not be interpreted in isolation because they both describe body size and may partly carry overlapping information. |
| Female crab colour | Colour groups have **uneven sample sizes**. The category with the highest mean satellite count is ``r as.character(color_satellite_summary$color[which.max(color_satellite_summary$mean_satellites)])``, with a mean of `r max(color_satellite_summary$mean_satellites)`. The category with the lowest mean satellite count is ``r as.character(color_satellite_summary$color[which.min(color_satellite_summary$mean_satellites)])``, with a mean of `r min(color_satellite_summary$mean_satellites)`. | Colour may be useful in the **extended model**, but category-level comparisons should be treated cautiously because some colour groups are **sparse** and the boxplots show **overlap** across categories. |
| Female crab spine condition | Spine groups are also **imbalanced**. The category with the highest mean satellite count is ``r as.character(spine_satellite_summary$spine[which.max(spine_satellite_summary$mean_satellites)])``, with a mean of `r max(spine_satellite_summary$mean_satellites)`. The category with the lowest mean satellite count is ``r as.character(spine_satellite_summary$spine[which.min(spine_satellite_summary$mean_satellites)])``, with a mean of `r min(spine_satellite_summary$mean_satellites)`. | Spine condition is worth checking in the **extended model**, but the EDA does not suggest that it should dominate the body-size regressors. The **imbalance** across spine categories also means coefficient-level interpretations should be cautious. |
| Predictive challenge | Across the continuous and categorical plots, female crabs with similar regressor values can still have quite different satellite counts. Zero counts also occur across multiple regions of the regressor space. | The predictive inquiry should be evaluated with **held-out testing data** and **prediction-error summaries**. Visual patterns from the training data can motivate model choices, but they are not enough to assess predictive performance. |

: Summary of the main EDA findings and their implications for the Poisson regression workflow. {#tbl-classical-poisson-eda-summary .striped .hover}

Overall, the EDA gives us a coherent modelling story. The response is a non-negative integer count with many zeros and variability that is much larger than the mean. Width shows the clearest descriptive association with satellite counts, while weight provides a related body-size signal that should be interpreted alongside width. Colour and spine condition may also contribute, but their category imbalance and overlapping count distributions call for caution. These findings do not establish final inferential conclusions; rather, they motivate the model-building sequence that follows. We now move to the GLM formulation so that the response distribution, regression structure, and coefficient interpretation are made explicit before estimation.

## Data Modelling


## Estimation

### Overdispersion

Let's revisit the `extended_poisson_model` fit in [Chapter 10](#sec-classical-poisson). This model showed substantial overdispersion relative to the Classical Poisson mean-variance assumption with a Pearson dispersion estimate of $3.082$ and a $p-$value < 0.001

-   We will test whether there is overdispersion in this Poisson regression model via function `dispersiontest()`, (from package `{AER}`). Note that `statsmodels` does not ship an equivalent function, so the `Python` code defines the same regression-based test as a small helper of the same name.
-   Let $Y_i$ be the $i$th Poisson response in the count regression model.

. . .

-   [Ideally in the presence of equidispersion]{style="color: purple;"}, $Y_i$ has the following parameters: \begin{gather*}
    \mathbb{E}(Y_i) = \lambda_i \\
    \text{Var}(Y_i) = \lambda_i.
    \end{gather*}

-   The test uses the following mathematical expression: $$
    \text{Var}(Y_i) = \overbrace{(1 + \gamma)}^\text{Dispersion Factor} \lambda_i,
    $$

. . .

-   Moreover, we have the following hypotheses: \begin{gather*}
    H_0: 1 + \gamma = 1 \\
    H_a: 1 + \gamma > 1.
    \end{gather*}
-   When there is evidence of overdispersion in our data, [we will reject $H_0$]{style="color: purple;"} under a pre-specified significance level $\alpha$.

::: {.panel-tabset}

## **`R` Code**

```{.r}
extended_poisson_model <- glm(
  satellites ~ width_cm + weight_kg + color + spine,
  family = poisson(link = "log"),
  data = training_data
)

dispersiontest(extended_poisson_model)
```

## **`Python` Code**

``` {.python}
import numpy as np
import scipy.stats as stats


def dispersiontest(results, alternative="greater"):
    """Cameron and Trivedi (1990) regression-based dispersion test.

    This is the `Python` counterpart of `AER::dispersiontest()` with
    `trafo = NULL`, i.e., the alternative is Var(Y_i) = (1 + gamma) * lambda_i.
    """
    y = np.asarray(results.model.endog, dtype=float)
    y_hat = np.asarray(results.fittedvalues, dtype=float)

    # Auxiliary quantity behind the test
    aux = ((y - y_hat) ** 2 - y) / y_hat

    test_statistic = np.sqrt(aux.size) * aux.mean() / aux.std(ddof=1)
    dispersion = aux.mean() + 1

    if alternative == "greater":
        p_value = stats.norm.sf(test_statistic)
    elif alternative == "less":
        p_value = stats.norm.cdf(test_statistic)
    elif alternative == "two-sided":
        p_value = 2 * stats.norm.sf(abs(test_statistic))
    else:
        raise ValueError(
            "alternative must be 'greater', 'less', or 'two-sided'."
        )

    return test_statistic, p_value, dispersion


# Fit Poisson model
extended_poisson_model = smf.glm(
    formula=(
        "satellites ~ width_cm + weight_kg "
        "+ C(color, Treatment(reference='Light')) "
        "+ C(spine, Treatment(reference='Both good'))"
    ),
    data=training_data,
    family=sm.families.Poisson()
).fit()

# Dispersion test
test_statistic, p_value, dispersion = dispersiontest(
    extended_poisson_model
)

print(f"Test statistic (z): {test_statistic:.4f}")
print(f"P-value: {p_value:.4g}")
print(f"Sample estimate (dispersion): {dispersion:.4f}")
```

:::

::: {.panel-tabset}

## **`R` Output**

```{r}
#| echo: false
#| message: false
#| warning: false
extended_poisson_model <- glm(
  satellites ~ width_cm + weight_kg + color + spine,
  family = poisson(link = "log"),
  data = training_data
)

dispersiontest(extended_poisson_model)
```

## **`Python` Output**

``` {python}
#| echo: false
#| message: false
#| warning: false
import numpy as np
import scipy.stats as stats


def dispersiontest(results, alternative="greater"):
    """Cameron and Trivedi (1990) regression-based dispersion test.

    This is the `Python` counterpart of `AER::dispersiontest()` with
    `trafo = NULL`, i.e., the alternative is Var(Y_i) = (1 + gamma) * lambda_i.
    """
    y = np.asarray(results.model.endog, dtype=float)
    y_hat = np.asarray(results.fittedvalues, dtype=float)

    # Auxiliary quantity behind the test
    aux = ((y - y_hat) ** 2 - y) / y_hat

    test_statistic = np.sqrt(aux.size) * aux.mean() / aux.std(ddof=1)
    dispersion = aux.mean() + 1

    if alternative == "greater":
        p_value = stats.norm.sf(test_statistic)
    elif alternative == "less":
        p_value = stats.norm.cdf(test_statistic)
    elif alternative == "two-sided":
        p_value = 2 * stats.norm.sf(abs(test_statistic))
    else:
        raise ValueError(
            "alternative must be 'greater', 'less', or 'two-sided'."
        )

    return test_statistic, p_value, dispersion


# Fit Poisson model
extended_poisson_model = smf.glm(
    formula=(
        "satellites ~ width_cm + weight_kg "
        "+ C(color, Treatment(reference='Light')) "
        "+ C(spine, Treatment(reference='Both good'))"
    ),
    data=training_data,
    family=sm.families.Poisson()
).fit()

# Dispersion test
test_statistic, p_value, dispersion = dispersiontest(
    extended_poisson_model
)

print(f"Test statistic (z): {test_statistic:.4f}")
print(f"P-value: {p_value:.4g}")
print(f"Sample estimate (dispersion): {dispersion:.4f}")
```
:::

With $\alpha = 0.05$, we reject $H_0$ since the $p\text{-value} < .001$. Hence, the `extended_poisson_model` has overdispersion.

### Fitting the Model

::: {.panel-tabset}

## **`R` Code**

```{.r}
negative_binomial_model <- glm.nb(satellites ~ width_cm,
  data = training_data
)
summary(negative_binomial_model)

extended_negative_binomial_model <- glm.nb(satellites ~ width_cm + weight_kg + color + spine,
  data = training_data
)
summary(extended_negative_binomial_model)
```

## **`Python` Code**

``` {.python}
negative_binomial_training_data = training_data.copy()

negative_binomial_model = smf.negativebinomial(
    formula="satellites ~ width_cm",
    data=negative_binomial_training_data
).fit()

negative_binomial_summary = pd.DataFrame(
    {
        "Term": negative_binomial_model.params.index,
        "Estimate": negative_binomial_model.params.values,
        "Standard error": negative_binomial_model.bse.values,
        "Exponentiated estimate": np.exp(
            negative_binomial_model.params.values
        ),
    }
)

extended_negative_binomial_model = smf.negativebinomial(
    formula="satellites ~ width_cm + weight_kg + color + spine",
    data=negative_binomial_training_data
).fit()

extended_negative_binomial_summary = pd.DataFrame(
    {
        "Term": extended_negative_binomial_model.params.index,
        "Estimate": extended_negative_binomial_model.params.values,
        "Standard error": extended_negative_binomial_model.bse.values,
        "Exponentiated estimate": np.exp(
            extended_negative_binomial_model.params.values
        ),
    }
)

negative_binomial_summary
extended_negative_binomial_summary
```

:::

::: {.panel-tabset}

## **`R` Output*

```{r}
#| echo: false
#| message: false
#| warning: false
negative_binomial_model <- glm.nb(satellites ~ width_cm,
  data = training_data
)
summary(negative_binomial_model)

extended_negative_binomial_model <- glm.nb(satellites ~ width_cm + weight_kg + color + spine,
  data = training_data
)
summary(extended_negative_binomial_model)
```

## **`Python` Output**

``` {python}
#| echo: false
#| message: false
#| warning: false
negative_binomial_training_data = training_data.copy()

negative_binomial_model = smf.negativebinomial(
    formula="satellites ~ width_cm",
    data=negative_binomial_training_data
).fit()

negative_binomial_summary = pd.DataFrame(
    {
        "Term": negative_binomial_model.params.index,
        "Estimate": negative_binomial_model.params.values,
        "Standard error": negative_binomial_model.bse.values,
        "Exponentiated estimate": np.exp(
            negative_binomial_model.params.values
        ),
    }
)

extended_negative_binomial_model = smf.negativebinomial(
    formula="satellites ~ width_cm + weight_kg + color + spine",
    data=negative_binomial_training_data
).fit()

extended_negative_binomial_summary = pd.DataFrame(
    {
        "Term": extended_negative_binomial_model.params.index,
        "Estimate": extended_negative_binomial_model.params.values,
        "Standard error": extended_negative_binomial_model.bse.values,
        "Exponentiated estimate": np.exp(
            extended_negative_binomial_model.params.values
        ),
    }
)

negative_binomial_summary
extended_negative_binomial_summary
```

:::


## Goodness of Fit


## Results


## Storytelling


## Practice Problems