<!-- Google tag (gtag.js) -->
<script async src="https://www.googletagmanager.com/gtag/js?id=G-7PRVEBE1EF"></script>
<script>
  window.dataLayer = window.dataLayer || [];
  function gtag(){dataLayer.push(arguments);}
  gtag('js', new Date());

  gtag('config', 'G-7PRVEBE1EF');
</script>

# Negative Binomial Regression {#sec-negative-binomial}

```{r}
#| include: false

colourize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}
```

::::: Warning
::: Warning-header
When to Use and Not Use Negative Binomial Regression
:::

::: Warning-container
Negative Binomial regression is useful when:

- The outcome variable is **discrete** and **count-type**, taking values $0, 1, 2, \dots$.
- The scientific or data-science inquiry focuses on how the expected count changes with a set of regressors.
- The outcome is **overdispersed**, meaning that the variance in the outcome variable is higher than what is plausible for a [Classical Poisson](https://alexrod61.github.io/regression-cookbook/book/10-classical-poisson.html) model.

However, a negative binomial regression regression should not be used in the following scenarios:

- If the outcome is **binary**, use a [Negative Binomial regression](#sec-negative-binomial) model.
- If the outcome is count-type and the data exhibits **equidispersion**, meaning that the conditional variance in the outcome variable is equal to the conditional mean, begin with a [Classical  regression](#sec-classical-) model.
- If the outcome is count-type and the data exhibits **underdispersion**, meaning that the the conditional variance in the outcome variable is lower than what is plausible for a Classical Poisson model, use a [Generalized Poisson regression](#sec-generalized-) model.
:::
:::::

::: {#fig-negative-binomial-regression}
```{mermaid}
mindmap
  root((Regression 
  Analysis)
    Continuous <br/>Outcome Y
      {{Unbounded <br/>Outcome Y}}
        )Chapter 3: <br/>Ordinary <br/>Least Squares <br/>Regression(
          (Normal <br/>Outcome Y)
      {{Nonnegative <br/>Outcome Y}}
        )Chapter 4: <br/>Gamma Regression(
          (Gamma <br/>Outcome Y)
      {{Bounded <br/>Outcome Y <br/> between 0 and 1}}
        )Chapter 5: Beta <br/>Regression(
          (Beta <br/>Outcome Y)
      {{Nonnegative <br/>Survival <br/>Time Y}}
        )Chapter 6: <br/>Parametric <br/> Survival <br/>Regression(
          (Exponential <br/>Outcome Y)
          (Weibull <br/>Outcome Y)
          (Lognormal <br/>Outcome Y)
        )Chapter 7: <br/>Semiparametric <br/>Survival <br/>Regression(
          (Cox Proportional <br/>Hazards Model)
            (Hazard Function <br/>Outcome Y)
    Discrete <br/>Outcome Y
      {{Binary <br/>Outcome Y}}
        {{Ungrouped <br/>Data}}
          )Chapter 8: <br/>Binary Logistic <br/>Regression(
            (Bernoulli <br/>Outcome Y)
        {{Grouped <br/>Data}}
          )Chapter 9: <br/>Binomial Logistic <br/>Regression(
            (Binomial <br/>Outcome Y)
      {{Count <br/>Outcome Y}}
        {{Equidispersed <br/>Data}}
          )Chapter 10: <br/>Classical Poisson <br/>Regression(
            ( <br/>Outcome Y)
        {{Overdispersed <br/>Data}}
          )Chapter 11: <br/>Negative Binomial <br/>Regression(
            (Negative Binomial <br/>Outcome Y)
```
:::

::: {.LO}
::::{.LO-header}
Learning Objectives
::::
::::{.LO-container}
By the end of this chapter, you will be able to:

- Determine when Negative Binomial Regression is an **appropriate modelling choice**, and recognize situations where a Classical  model is a more appropriate choice.
- Frame **inferential** and **predictive inquiries** for a count-regression problem using the **data science workflow**.
- Specify a Negative Binomial regression model as a generalized linear model with a **Negative Binomial random component**, a **linear predictor**, and a **log link function**.
- Write and interpret the **likelihood** and **log-likelihood functions** for a Negative Binomial regression model.
- Explain how **maximum likelihood estimation** is used to estimate  regression coefficients.
- Interpret  regression coefficients as **multiplicative changes in the expected count**, holding other regressors fixed.
- Assess model adequacy using **goodness-of-fit checks**, **residual-based summaries**, and the ** mean-variance assumption**.
- Construct and interpret **confidence intervals** and **hypothesis tests** for Negative Binomial regression coefficients.
- Evaluate **predictive performance** using test-set prediction accuracy metrics and baseline comparisons.
- Communicate Negative Binomial regression results responsibly for both inferential and predictive inquiries.
::::
:::

```{r setup-r, include=FALSE}
# Load required libraries
library(devtools)
library(readr)
library(tibble)
library(tidyr)
library(dplyr)
library(magrittr)
library(ggplot2)
library(gridExtra) # combine ggplot2 figures
library(ggExtra) # draw marginal histograms
library(broom)
library(knitr)
library(AER) # dispersiontest
library(MASS)

library(reticulate)
py_config()

py_require("pandas")
py_require("numpy")
py_require("seaborn")
py_require("pyreadr")
py_require("plotnine")
py_require("matplotlib")
py_require("statsmodels")
py_require("scikit-learn")
py_require("scipy")
py_require("pyreadr")

reticulate::py_install("plotnine")

pandas <- import("pandas")
numpy <- import("numpy")
seaborn <- import("seaborn")
pyreadr <- import("pyreadr")
matplotlib <- import("matplotlib")
statsmodels <- import("statsmodels")
sklearn <- import("sklearn")
scipy <- import("scipy")
plotnine <- import("plotnine")
pyreadr <- import("pyreadr")

options(htmlwidgets.TOJSON_ARGS = list(na = 'string')) 

print_as_dt <- function(x){
  DT::datatable(x,
                rownames = FALSE,
                options = list(
                  dom = 'ltipr', 
                  autoWidth = TRUE,
                  columnDefs = list(
                    list(className = 'dt-left', targets = "_all"),
                    list(width = '200px', targets = "_all")
                    )
                  )
                )
}

print_as_dt_all <- function(x){
  DT::datatable(
    x,
    rownames = FALSE,
    options = list(
      dom = 't',  # Only show the table (no filter/search/pagination info)
      paging = FALSE,
      scrollX = TRUE,
      autoWidth = TRUE,
      columnDefs = list(
        list(className = 'dt-left', targets = "_all")
        # Optionally remove width limitation or set only for specific columns
      )
    )
  )
}
```

```{python setup-py, include=FALSE}
import numpy as np
from plotnine import *
import statsmodels.api as sm
import statsmodels.formula.api as smf
# from statsmodels.formula.api import glm, ols
from statsmodels.genmod.families import Poisson
from matplotlib import pyplot as plt
from sklearn.metrics import mean_squared_error, mean_absolute_error
import scipy.stats as stats

def scrollable_table_html(data, **kwargs):
    kwargs.setdefault("index", False)
    kwargs.setdefault("border", 0)

    return (
        '<div class="table-responsive">'
        + data.to_html(
            **kwargs
        )
        + '</div>'
    )
```

## Introduction {#sec-negative-binomial-introduction}

**Negative Binomial regression** belongs to the generalized linear model (GLM) family, but it uses a Negative Binomial random component and a log link function instead of the Normal random component and identity link that appear in Ordinary Least-squares (OLS) regression. This allows the model to connect a set of regressors to the expected count while respecting the fact that fitted counts should not be negative.

## Case Study: Horseshoe Crab Satellites

The running case study in this chapter uses the dataset on **horseshoe crab satellites** from [Chapter 10](#sec-classical-). It is a good example for this chapter because it a **count regression** problem with an **overdispersed** response variable: 

> **For each female horseshoe crab, we observe the *number of satellite males* around her.**

![Image by [*BarelyDevi*](https://pixabay.com/users/barelydevi-14723734/) via [*Pixabay*](https://pixabay.com/illustrations/crab-sea-seafood-ocean-beach-fish-5030504/).](img/crab.png){width="500"}

In the biological setting, a female horseshoe crab may arrive at a nesting site with an attached male. Other unattached males may gather around the nesting pair and compete for fertilization opportunities. These unattached males are called **satellite males**. Some female crabs have no satellite males nearby, while others have several. Therefore, the response variable is a non-negative integer count. It can be $0$, $1$, $2$, and so on, but it cannot be negative and it is not naturally continuous. This makes the dataset a useful starting point for  regression.

In [Chapter 10](#sec-classical-), we saw that the variance in the number of satellite males is much larger than the mean number of satellite males. We will formally test **overdispersion** in @sec-overdispersion, and Negative Binomial regression can be a better modelling choice than  regression if there is statistical evidence of **overdispersion** in the data. The response is a count, the regressors include both continuous and categorical variables, and the scientific question(s) can be approached from more than one modelling purpose:

- On the **inferential side**, the model can help us describe how the expected number of satellites is associated with female crab characteristics. 
- On the **predictive side**, the same regressors can be used to predict the expected number of satellites for a female crab and then evaluate how accurate those predictions are. W

We will formalize the two inquiries above in @sec-negative-binomial-study-design.

::: {.Tip}
:::: {.Tip-header}
Tip on why this biological dataset is worth using!
::::
:::: {.Tip-container}
The horseshoe-crab example is more than an old textbook dataset. It is connected to a real biological question about mating behaviour, competition, and observable traits. Brockmann's original study investigated why some nesting females attract larger groups of satellite males than others [@brockmann1996]. Related work has also studied male mating tactics in horseshoe crabs and the behavioural mechanisms behind attached and unattached males [@brockmann2002; @brockmann2009].

The dataset is especially useful for learning because the statistical structure is easy to state even if the biology is unfamiliar. The observational unit is a female crab. The response is the number of satellite males. Moreover, the regressors describe observable characteristics of the female crab. That said, this gives us a compact setting where the data science workflow can remain visible: define the inquiry, understand the data, specify the model, estimate the coefficients, check the model, and communicate what the results mean.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-pixel-nerd-3699334/).](img/feature.png){width="350"}

Horseshoe crabs also matter beyond this single modelling example. Their eggs are ecologically important food resources for migratory shorebirds in coastal systems, and horseshoe crab blood has historically been used in biomedical endotoxin testing [@botton2009; @maloney2018]. This broader context helps make the dataset less like an isolated classroom object and more like a compact entry point into ecological data analysis with count outcomes.
::::
:::

## Study Design: Framing the Inferential and Predictive Inquiries {#sec-negative-binomial-study-design}

Before fitting a Negative Binomial regression model, we need to clarify what the model is supposed to help us learn. This is the **study design** (as shown in @fig-ds-workflow-study-design) step of the data science workflow introduced in @sec-ds-workflow-study-design. At this stage, we are not yet choosing software functions or interpreting coefficients. Instead, we are defining the modelling purpose, the observational unit, the response variable, and the role of the regressors. 

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-pixel-action-bound-3704046/).](img/magnifying-glass-2.png){width="420"}

Hence, for the horseshoe crab case study, the study-design elements are summarized in @tbl-negative-binomial-study-design-variables. This table fixes the basic modelling ingredients before we decide how the Negative Binomial regression model will be used. Note that @tbl-negative-binomial-study-design-variables is intentionally simple. At this point in the workflow, we are not yet fitting a model or interpreting coefficients. We are just making sure that the observational unit, response variable, and regressors are clear.

| **Role in the study design** | **Variable(s)** | **Description** |
|:---:|:---:|:---:|
| **Observational unit** | Female horseshoe crab | Each row corresponds to one female crab observed in the study. |
| **Response variable** | Number of satellite males | The count outcome: how many satellite males were observed near the female crab. |
| **Main regressors** | Body width, weight, colour category, spine condition | Observable characteristics of the female crab that may be associated with the expected number of satellite males. |

: Key elements of the horseshoe crab study design for the  regression chapter. {#tbl-negative-binomial-study-design-variables .striped .hover}

Note that a single dataset can support different kinds of modelling inquiries. Therefore, we will use the same count response to illustrate two related but distinct modelling purposes: an **inferential inquiry** and a **predictive inquiry** as shown in @tbl-negative-binomial-study-design-inquiries. This distinction is critical since the modelling workflow is not judged in exactly the same way for both purposes.

| **Inquiry type** | **Main question** | **Main modelling emphasis** | **What we will look for later in the chapter** |
|:---:|:---:|:---:|:---:|
| **Inferential inquiry** | How is the expected number of satellite males associated with female crab characteristics? | Coefficients, uncertainty, hypothesis tests, confidence intervals, and interpretation. | Whether the estimated associations are meaningful, how uncertain they are, and whether they can be interpreted responsibly. |
| **Predictive inquiry** | How well can the available regressors predict the expected number of satellite males for a female crab? | Predicted expected counts, test-set accuracy, and comparison against a simple baseline. | Whether the fitted model improves predictive performance beyond a simple benchmark and whether the prediction errors are practically acceptable. |

: Two modelling inquiries supported by the horseshoe crab satellites case study. {#tbl-negative-binomial-study-design-inquiries .striped .hover}

The above inferential and predictive inquiries are connected but not interchangeable. A model can have interpretable coefficients but only modest predictive performance. Conversely, a model can give useful predictions without answering every scientific question about the mechanisms behind satellite male behaviour. Keeping the two inquiries separate helps us decide which results belong in the inferential interpretation and which results belong in the prediction-accuracy discussion.

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on inference versus prediction!
::::
:::: {.Heads-up-container}
For the inferential inquiry, the main object of interest is the relationship between the expected count and the regressors. We will pay attention to coefficient estimates, standard errors, confidence intervals, hypothesis tests, and the assumptions needed for the interpretation to be reasonable.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-emotion-bored-6230197/).](img/question-2.png){width="300"}

On the other hand, for the predictive inquiry, the main object of interest is how well the model predicts expected counts for observations not used to fit the model. We will therefore use a training/testing split, compute predicted expected counts on the test set, compare prediction errors against a baseline model, and summarize prediction accuracy with metrics such as **mean absolute error (MAE)** and **root mean squared error (RMSE)**.

Both inquiries require model checking. If the Negative Binomial model fits the data poorly, then both the inferential interpretation and the predictive results may become less useful.
::::
:::

### Inferential Inquiry {#sec-negative-binomial-inferential-inquiry}

Specifically, the inferential inquiry asks:

> **How is the expected number of satellite males associated with female crab characteristics, holding other regressors fixed?**

This question is about **association and interpretation**. It does not ask whether a female crab characteristic causes satellite males to appear. The data are **observational**, and the case study does not establish a causal design. Instead, the inferential goal is to describe how the expected count of satellite males changes across observed female crab characteristics within a  regression framework.

For example, one inferential question is whether wider female crabs tend to have a larger expected number of satellite males. Later, when we introduce the simple Negative Binomial regression model in @sec-negative-binomial-simple-model, body width will be used as the first regressor because it gives a clear one-regressor entry point into the model. In the extended model in @sec-negative-binomial-additional-regressors, we will add more female crab characteristics so that we can interpret associations while holding other regressors fixed.

For this inquiry, the **results stage** will focus on coefficient estimates, uncertainty, and statistical interpretation. We will ask whether the estimated associations are consistent with the scientific question, whether the confidence intervals are informative, and whether hypothesis tests provide evidence for associations between the regressors and the expected count.

### Predictive Inquiry {#sec-negative-binomial-predictive-inquiry}

Now, the predictive inquiry asks:

> **How well can the available regressors be used to predict the expected number of satellite males for a female crab?**

This question is about **prediction accuracy**. The target of prediction in  regression is the expected count, **not necessarily the exact observed count**. For a given female crab, the observed number of satellite males may differ from the model-predicted expected count because count outcomes are variable even when the model is well specified.

For this inquiry, the **results stage** will focus on out-of-sample prediction. We will fit the model using a training set and evaluate prediction accuracy on a testing set. The testing-set predictions will be compared against a simple baseline that predicts the training-set mean count for every test observation. This comparison is necessary because an accuracy metric is easier to interpret when we can ask whether the regression model does better than a simple benchmark.

The predictive inquiry will use metrics such as MAE and RMSE. MAE summarizes the typical absolute distance between the observed count and the predicted expected count, while RMSE penalizes larger prediction errors more strongly. These metrics will help us judge whether the Negative Binomial regression model is useful for prediction, not only whether its coefficients are interpretable.

## Data Collection and Wrangling

The data used in this chapter are **secondary observational data**. They were originally collected in the study by @brockmann1996 on satellite male groups in horseshoe crabs and later made available in a form commonly used in a textbook on categorical data and count regression [@agresti2013]. Note that we are not designing a new experiment, assigning treatments, or controlling the biological setting. We are working with an existing observational dataset to determine whether female crab characteristics are associated with and can help predict the number of satellite males. Furthermore, the version of the dataset used comes from [Agresti's public GitHub repository](https://github.com/alanagresti/categorical-data/blob/master/Crabs.dat) for categorical-data examples. The GitHub page displays the file in the browser. In the code below, we use the corresponding raw file URL so that `R` and `Python` can import the data directly.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-evaluation-measure-up-3976303/).](img/measurement.png){width="450"}

According to the documentation for the teaching version of the dataset, each row corresponds to one female horseshoe crab [@agresti2013]. The variables record the female crab's colour, spine condition, carapace width, number of satellite males, and weight. The original data also include a binary indicator for whether at least one satellite male was present. That said, in this chapter, our response variable is the **count** of satellite males, so the binary indicator is not the primary outcome.

Now, before doing EDA, we will complete a small amount of data wrangling (as shown in @fig-ds-workflow-data-collection-wrangling). The goal is not to create a training/testing split yet. As in the data science workflow from @sec-intro (more specifically in @sec-ds-workflow-eda), the data split belongs to the EDA stage because we first need to inspect the data structure and then decide how to separate observations for modelling and evaluation. In this section, we only load the data, rename variables, and prepare the categorical regressors so that they are easier to interpret later. The main variables used in the chapter are summarized in @tbl-negative-binomial-wrangled-variables.

| **Variable after wrangling** | **Role** | **Description** |
|:---:|:---:|:---:|
| `satellites` | Response | Number of satellite males observed near the female crab. |
| `width_cm` | Regressor | Female crab carapace width, measured in centimetres. |
| `weight_kg` | Regressor | Female crab weight, measured in kilograms. |
| `color` | Regressor | Female crab colour category, treated as categorical. |
| `spine` | Regressor | Female crab spine condition, treated as categorical. |

: Main variables used after wrangling the horseshoe crab dataset. {#tbl-negative-binomial-wrangled-variables .striped .hover}

The `R` and `Python` code below follow the same wrangling logic. First, the data are read from the public teaching data file associated with the categorical-data examples in @agresti2013. Second, the columns are assigned readable names. Third, the colour and spine variables are converted to categorical variables with descriptive labels. Finally, the dataset is restricted to the variables needed for this chapter. More specifically, we have the following:

- In `R`, via the packages part of [{tidyverse}](https://www.tidyverse.org), we read the raw data directly from Agresti's GitHub repository. The file already includes column names, so we let `read_table()` read the header. Then, we rename the columns needed for this chapter: `sat` becomes `satellites`, `width` becomes `width_cm`, and `weight` becomes `weight_kg`. The variables `color` and `spine` are stored as numeric codes in the raw file, so we convert them into categorical variables with descriptive labels.
- In `Python`, via [{pandas}](https://pypi.org/project/pandas/), we repeats the same wrangling steps. The file already includes column names, so `read_csv()` reads the header directly. Then, `rename()` gives the response and continuous regressors more explicit names. The `map()` calls replace the numeric colour and spine codes with descriptive labels, and the final selection keeps the working variables used in this chapter.

::: {.panel-tabset}

## **`R` Code**

``` {.r}
# Loading library
library(tidyverse)

crabs_url <- paste0(
  "https://raw.githubusercontent.com/",
  "alanagresti/categorical-data/master/Crabs.dat"
)

crabs_raw <- read_table(crabs_url)

crabs <- crabs_raw |>
  transmute(
    satellites = sat,
    width_cm = width,
    weight_kg = weight,
    color = factor(
      color,
      levels = c(1, 2, 3, 4),
      labels = c(
        "Light",
        "Medium light",
        "Medium dark",
        "Dark"
      )
    ),
    spine = factor(
      spine,
      levels = c(1, 2, 3),
      labels = c(
        "Both good",
        "One worn or broken",
        "Both worn or broken"
      )
    )
  )
```

## **`Python` Code**

``` {.python}
# Importing library
import pandas as pd

crabs_url = (
    "https://raw.githubusercontent.com/"
    "alanagresti/categorical-data/master/Crabs.dat"
)

crabs_raw = pd.read_csv(
    crabs_url,
    sep=r"\s+"
)

color_labels = {
    1: "Light",
    2: "Medium light",
    3: "Medium dark",
    4: "Dark",
}

spine_labels = {
    1: "Both good",
    2: "One worn or broken",
    3: "Both worn or broken",

}

crabs = (
    crabs_raw
    .rename(
        columns={
            "sat": "satellites",
            "width": "width_cm",
            "weight": "weight_kg",
        }
    )
    .assign(
        color=lambda data_frame: data_frame["color"].map(color_labels),
        spine=lambda data_frame: data_frame["spine"].map(spine_labels),
    )
    [
        [
            "satellites",
            "width_cm",
            "weight_kg",
            "color",
            "spine",
        ]
    ]
)
```

:::

::: {.panel-tabset}

## **`R` Output**

```{r}
#| label: tbl-crabs-data-r
#| tbl-cap: "Full crabs data."
#| echo: false
#| message: false
#| warning: false

library(tidyverse)

crabs_url <- paste0(
  "https://raw.githubusercontent.com/",
  "alanagresti/categorical-data/master/Crabs.dat"
)

crabs_raw <- read_table(crabs_url)

crabs <- crabs_raw |>
  transmute(
    satellites = sat,
    width_cm = width,
    weight_kg = weight,
    color = factor(
      color,
      levels = c(1, 2, 3, 4),
      labels = c(
        "Light",
        "Medium light",
        "Medium dark",
        "Dark"
      )
    ),
    spine = factor(
      spine,
      levels = c(1, 2, 3),
      labels = c(
        "Both good",
        "One worn or broken",
        "Both worn or broken"
      )
    )
  )
  
# Showing the full dataset
crabs |> 
  print_as_dt()
```

## **`Python` Output**

```{python}
#| label: lst-negative-binomial-data-wrangling-py
#| lst-cap: "Loading and wrangling the horseshoe crab satellite data in Python."
#| echo: false
#| message: false
#| warning: false

import pandas as pd

crabs_url = (
    "https://raw.githubusercontent.com/"
    "alanagresti/categorical-data/master/Crabs.dat"
)

crabs_raw = pd.read_csv(
    crabs_url,
    sep=r"\s+"
)

color_labels = {
    1: "Light",
    2: "Medium light",
    3: "Medium dark",
    4: "Dark",
}

spine_labels = {
    1: "Both good",
    2: "One worn or broken",
    3: "Both worn or broken",

}

crabs = (
    crabs_raw
    .rename(
        columns={
            "sat": "satellites",
            "width": "width_cm",
            "weight": "weight_kg",
        }
    )
    .assign(
        color=lambda data_frame: data_frame["color"].map(color_labels),
        spine=lambda data_frame: data_frame["spine"].map(spine_labels),
    )
    [
        [
            "satellites",
            "width_cm",
            "weight_kg",
            "color",
            "spine",
        ]
    ]
)
```

```{r}
#| label: tbl-crabs-data-py
#| tbl-cap: "Full crabs data."
#| echo: false

# Showing the full dataset
crabs_py <- py$crabs
print_as_dt(crabs_py)
```

:::

After this wrangling step, the object `crabs` is the working dataset for subsequent sections. It contains one row per female crab, one count response, two continuous regressors, and two categorical regressors. The next step is EDA, where we will create the training/testing split needed , inspect the response distribution, and examine the regressors.

## Exploratory Data Analysis

EDA begins after the data have been collected and wrangled into the working object `crabs`. Following the workflow in @fig-ds-workflow-eda, this is also the stage where we create the training/testing split. The training set will be used for EDA, model fitting, and goodness-of-fit checking. These goodness-of-fit checks are applied to the model fitted on the training set because they act as the **workflow gate** before we decide whether the fitted Negative Binomial regression model is adequate enough to use for the chapter's inferential and predictive purposes. Then, the testing set is held aside until the results stage, but it plays different roles for the two inquiries:

- For the **predictive inquiry**, the model fitted on the training set will be used to generate predicted expected counts for the testing observations, and the corresponding prediction accuracy outputs will be reported in the results stage. 
- For the **inferential inquiry**, after the training-set model has passed the goodness-of-fit stage, we will refit the selected model on the testing set to generate the inferential outputs reported in the results stage. This protects the final inferential claims from double dipping: the same observations used for EDA, model fitting, and diagnostics are not reused for the final coefficient-level inferential statements.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-pixel-outlook-success-3704053/).](img/explore.png){width="400"}

Before splitting the data, it is useful to classify the variables that will appear throughout the chapter. For the $i$th female crab, @tbl-negative-binomial-variables summarizes the response and the main regressors. Note that the response variable is the number of satellite males. This is the main reason  regression is a natural starting point: $Y_i$ is a non-negative integer count. Then, the continuous regressors `width_cm` and `weight_kg` will help us examine whether larger female crabs tend to have higher expected numbers of satellites. On the other hand, the categorical regressors `color` and `spine` will help us examine whether visible phenotypic characteristics are associated with the response.

| **Variable** | **Role** | **Type** | **Notation** | **Description** |
|:---:|:---:|:---:|:---:|:---:|
| `satellites` | Response | Discrete count | $Y_i$ | Number of satellite males observed near the female crab. |
| `width_cm` | Regressor | Continuous | $x_{i,1}$ | Female crab carapace width, measured in centimetres. |
| `weight_kg` | Regressor | Continuous | $x_{i,2}$ | Female crab weight, measured in kilograms. |
| `color` | Regressor | Categorical | Encoded in @sec-negative-binomial-categorical-regressors using indicator regressors | Female crab colour category. |
| `spine` | Regressor | Categorical | Encoded in @sec-negative-binomial-categorical-regressors using indicator regressors | Female crab spine condition. |

: Variables used in the horseshoe crab  regression workflow. {#tbl-negative-binomial-variables .striped .hover}

::: {.Tip}
:::: {.Tip-header}
Tip on the 50/50 split and more careful alternatives!
::::
:::: {.Tip-container}
This dataset contains only `r nrow(crabs)` observations. A standard predictive modelling workflow often uses a larger training set, such as an 80/20 training/testing split. However, in this chapter we need to support two goals at the same time: an **inferential** inquiry and a **predictive** inquiry.

Using a 50/50 split is a compromise. It gives the training set enough observations for EDA, model fitting, and goodness-of-fit checking, while also leaving a testing set large enough to support both the final **predictive accuracy assessment** and the final **inferential refit**. This choice is not perfect. A larger training set would usually help estimation and diagnostics, while a larger test set gives a more informative final assessment. With small datasets, we often have to choose between these inconveniences.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-pixel-to-learn-goal-3674110/).](img/mean.png){width="600"}

The 50/50 split used here is also a **simple random split**, which means it does not explicitly protect against imbalance across important variables. For example, if one female crab colour category or spine condition is rare, a simple random split could place too many of those crabs in one subset and too few in the other. That imbalance can affect both the inferential and predictive parts of the workflow.

A more careful split could stratify the data by an important categorical regressor, such as female crab colour, or by a coarser grouping of the response variable. Other strategies include repeated random splits, cross-validation, or bootstrap-based assessment. These approaches can reduce dependence on a single split and are widely used in predictive modelling and model validation; see @hastie2009, @kuhn2013, and @harrell2015 for broader discussions. They are worth trying as an extension, but the main chapter keeps a single 50/50 split so that the  regression workflow remains transparent.
::::
:::

The `R` and `Python` code below first produce independent 50/50 random splits. These two splits use the same conceptual allocation and the same seed value, but they are not expected to select exactly the same observations because `R` and `Python` use different random-number machinery and different splitting implementations:

- @lst-negative-binomial-data-splitting-r uses the [{rsample}](https://rsample.tidymodels.org/) package to split the wrangled `crabs` dataset into training and testing sets. The argument `prop = 0.5` requests that approximately half of the observations be assigned to the training set. The remaining observations are assigned to the testing set. The sanity check prints the dimensions of both subsets and their observed proportions.
- @lst-negative-binomial-data-splitting-py performs the analogous 50/50 split in `Python` using `train_test_split()` from [{scikit-learn}](https://scikit-learn.org/stable/). The argument `test_size = 0.5` assigns approximately half of the observations to the testing set, with the rest assigned to the training set. The output is used only to demonstrate the analogous Python splitting workflow.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| lst-label: lst-negative-binomial-data-splitting-r
#| lst-cap: "Code to randomly split the horseshoe crab data into training and testing sets."
#| message: false
#| warning: false

# Loading libraries
library(rsample)
library(reticulate)

# Seed for reproducibility
set.seed(123)

# Randomly splitting into training and testing sets
crabs_data_splitting <- initial_split(
  crabs,
  prop = 0.5
)

# Assigning data points to training and testing sets
training_data <- training(crabs_data_splitting)
testing_data <- testing(crabs_data_splitting)

# Sanity checks
n_total <- nrow(crabs)
n_train <- nrow(training_data)
n_test <- nrow(testing_data)

cat(sprintf(
  "Training shape: %d %d\nTesting shape:  %d %d\n\nTraining proportion: %.3f\nTesting proportion:  %.3f\n",
  nrow(training_data), ncol(training_data),
  nrow(testing_data),  ncol(testing_data),
  n_train / n_total,
  n_test / n_total
))
```

## **`Python`**

```{python}
#| echo: true
#| lst-label: lst-negative-binomial-data-splitting-py
#| lst-cap: "Code to randomly split the horseshoe crab data into training and testing sets."
#| message: false
#| warning: false

# Importing function
from sklearn.model_selection import train_test_split

# Seed for reproducibility
random_state = 123

# Randomly splitting into training and testing sets
training_data_py_independent, testing_data_py_independent = train_test_split(
    crabs,
    test_size=0.5,
    random_state=random_state
)

# Sanity checks
n_total = len(crabs)
n_train = len(training_data_py_independent)
n_test = len(testing_data_py_independent)

print(
    f"Training shape: {training_data_py_independent.shape}\n"
    f"Testing shape:  {testing_data_py_independent.shape}\n\n"
    f"Training proportion: {n_train / n_total:.3f}\n"
    f"Testing proportion:  {n_test / n_total:.3f}"
)
```

:::

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on keeping `R` and `Python` aligned after the split!
::::
:::: {.Heads-up-container}
Using the same seed and split proportion in `R` and `Python` does not guarantee that the random splits will contain the same crabs. This is not an error. The two ecosystems use different splitting functions and pseudo-random number generators. Therefore, for the rest of this chapter, we use the `R`-generated training and testing sets as a common reference to ensure consistent coding outputs.
::::
:::

Via [{reticulate}](https://rstudio.github.io/reticulate/), @lst-negative-binomial-import-r-split-py imports the `R`-generated training and testing sets into the `Python` environment. This ensures the `R` and `Python` results are comparable since summaries, plots, fitted models, and prediction metrics will be based on the same observations. Henceforth, the `Python` code uses the same `training_data` and `testing_data` objects as the `R` workflow.

::: {.panel-tabset}

## **`Python`**

```{python}
#| echo: true
#| lst-label: lst-negative-binomial-import-r-split-py
#| lst-cap: "Code to import the R-generated training and testing sets into Python."
#| message: false
#| warning: false

# Importing R-generated training and testing sets via reticulate
training_data = r.training_data
testing_data = r.testing_data

# Ensuring categorical regressors are treated as categorical in Python
training_data["color"] = training_data["color"].astype("category")
training_data["spine"] = training_data["spine"].astype("category")
testing_data["color"] = testing_data["color"].astype("category")
testing_data["spine"] = testing_data["spine"].astype("category")
```

:::

The upcoming EDA will use only the training data. In this training split, we have `r nrow(training_data)` female crabs for exploration, model fitting, and goodness-of-fit checking. The testing data are kept aside until the results stage, where they will be used for two different purposes: final prediction accuracy for the predictive inquiry and refitting the selected model for the final inferential outputs.

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on how the split is used later!
::::
:::: {.Heads-up-container}
The same training/testing split supports both inquiry flavours, but the two flavours use the subsets differently after EDA:

- For the **predictive inquiry**, the Negative Binomial regression model fitted on the training set will later be used to predict expected counts for the testing set. These test-set predictions will be used to compute prediction accuracy metrics in the results stage.
- For the **inferential inquiry**, the training set is used to explore the data, fit the candidate model, and run goodness-of-fit checks. If the model is adequate enough to proceed, the selected model specification will then be refit on the testing set to generate the final inferential outputs reported in the results stage. This helps avoid double dipping by separating the data used for exploration and diagnostics from the data used for final inferential reporting.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-pixel-to-learn-3674106/).](img/pieces.png){width="250"}
::::
:::

### Descriptive Summaries {#sec-negative-binomial-eda-general}

We begin with descriptive summaries because they tell us whether the data looks compatible with the modelling task:

- For the **response**, we are especially interested in the number of zeros, the typical count, and the relationship between the sample mean and sample variance. 
- For the **regressors**, we want to understand the scale of the continuous variables and the distribution of the categorical variables.

Thus, we have the following:

- @tbl-negative-binomial-training-summary-r shows the training-set summaries for the response and the two continuous regressors. in `R`. For the count response, it reports the mean and variance because the Negative Binomial regression model will later require us to think carefully about the mean-variance relationship.
- @tbl-negative-binomial-training-summary-py shows the same summaries in `Python`. The code uses the `training_data` object imported from `R`, so the values should match the `R` summaries up to formatting.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-negative-binomial-training-summary-r
#| tbl-cap: "Descriptive summary statistics for the horseshoe crab training data."
#| message: false
#| warning: false

# Loading library to display tables
library(knitr)

training_summary <- training_data |>
  summarise(
    `Number of training observations` = n(),
    `Mean number of satellite males` = mean(satellites),
    `Variance of satellite males` = var(satellites),
    `Proportion with zero satellite males` = mean(satellites == 0),
    `Mean female crab width (cm)` = mean(width_cm),
    `SD female crab width (cm)` = sd(width_cm),
    `Mean female crab weight (kg)` = mean(weight_kg),
    `SD female crab weight (kg)` = sd(weight_kg)
  ) |>
  mutate(
    across(
      where(is.numeric),
      ~ round(.x, 2)
    )
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "Quantity",
    values_to = "Value"
  ) |>
  mutate(
    Value = if_else(
      Quantity == "Number of training observations",
      formatC(Value, format = "f", digits = 0),
      formatC(Value, format = "f", digits = 2)
    )
  ) 

training_summary |>
  kable(
    align = c("c", "c")
  )
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

training_summary = pd.DataFrame(
    {
        "Quantity": [
            "Number of training observations",
            "Mean number of satellite males",
            "Variance of satellite males",
            "Proportion with zero satellite males",
            "Mean female crab width (cm)",
            "SD female crab width (cm)",
            "Mean female crab weight (kg)",
            "SD female crab weight (kg)",
        ],
        "Value": [
            len(training_data),
            training_data["satellites"].mean(),
            training_data["satellites"].var(ddof=1),
            (training_data["satellites"] == 0).mean(),
            training_data["width_cm"].mean(),
            training_data["width_cm"].std(ddof=1),
            training_data["weight_kg"].mean(),
            training_data["weight_kg"].std(ddof=1),
        ],
    }
)

training_summary["Value"] = [
    f"{value:.0f}" if quantity == "Number of training observations"
    else f"{value:.2f}"
    for quantity, value in zip(
        training_summary["Quantity"],
        training_summary["Value"]
    )
]

negative_binomial_training_summary_py_html = (
    training_summary.to_html(
        index=False,
        border=0,
    )
)
```

```{r}
#| echo: false
#| label: tbl-negative-binomial-training-summary-py
#| tbl-cap: "Descriptive summary statistics for the horseshoe crab training data."
#| message: false
#| warning: false
#| results: asis

cat(
  "\n",
  py$negative_binomial_training_summary_py_html,
  "\n",
  sep = ""
)
```

:::

The table above gives a quick numerical profile of the training data before we fit any model. The training set contains `r nrow(training_data)` female crabs. The average number of satellite males is `r round(mean(training_data$satellites), 2)`, but the sample variance is `r round(var(training_data$satellites), 2)`, which is noticeably larger than the mean. We will return to this issue more formally in @sec-negative-binomial-overdispersion.

The proportion of crabs with zero satellite males is also important. In the training data, approximately `r round(mean(training_data$satellites == 0), 2)` of female crabs have no satellite males. This does not automatically mean that a Negative Binomial regression model is inappropriate, but it gives us a first reason to pay attention to the number of zeros when checking model adequacy later in the chapter.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-lecture-lecture-hall-3976296/).](img/sample.png){width="500"}

Now, the summaries for `width_cm` and `weight_kg` describe the scale of the continuous regressors. Female crab width has an average of `r round(mean(training_data$width_cm), 2)` cm, while weight has an average of `r round(mean(training_data$weight_kg), 2)` kg. These summaries help us interpret the range of body-size values before we examine whether body size appears associated with the number of satellite males.

Next, we summarize the categorical regressors `color` and `spine`. These summaries are useful because categorical imbalance can affect interpretation and prediction, especially with a small dataset.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-negative-binomial-color-summary-r
#| tbl-cap: "Descriptive summary of female crab colour categories in the training data."
#| message: false
#| warning: false

color_summary <- training_data |>
  count(color, name = "n") |>
  mutate(proportion = round(n / sum(n), 3))

color_summary |>
  rename(
    `Female crab colour` = color,
    `Number of female crabs` = n,
    `Proportion of training data` = proportion
  ) |>
  kable(
    align = c("c", "c", "c")
  )
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

color_summary = (
    training_data["color"]
    .value_counts()
    .rename_axis("color")
    .reset_index(name="n")
)

color_summary["proportion"] = (
    color_summary["n"] / color_summary["n"].sum()
).round(3)

color_summary_display = color_summary.rename(
    columns={
        "color": "Female crab colour",
        "n": "Number of female crabs",
        "proportion": "Proportion of training data",
    }
)

negative_binomial_color_summary_py_html = (
    color_summary_display.to_html(
        index=False,
        border=0,
    )
)
```

```{r}
#| echo: false
#| label: tbl-negative-binomial-color-summary-py
#| tbl-cap: "Descriptive summary of female crab colour categories in the training data."
#| message: false
#| warning: false
#| results: asis

cat(
  "\n",
  py$negative_binomial_color_summary_py_html,
  "\n",
  sep = ""
)
```

:::

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-negative-binomial-spine-summary-r
#| tbl-cap: "Descriptive summary of female crab spine condition categories in the training data."
#| message: false
#| warning: false

spine_summary <- training_data |>
  count(spine, name = "n") |>
  mutate(proportion = round(n / sum(n), 3))

spine_summary |>
  rename(
    `Female crab spine condition` = spine,
    `Number of female crabs` = n,
    `Proportion of training data` = proportion
  ) |>
  kable(
    align = c("c", "c", "c")
  )
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

spine_summary = (
    training_data["spine"]
    .value_counts()
    .rename_axis("spine")
    .reset_index(name="n")
)

spine_summary["proportion"] = (
    spine_summary["n"] / spine_summary["n"].sum()
).round(3)

spine_summary_display = spine_summary.rename(
    columns={
        "spine": "Female crab spine condition",
        "n": "Number of female crabs",
        "proportion": "Proportion of training data",
    }
)

negative_binomial_spine_summary_py_html = (
    spine_summary_display.to_html(
        index=False,
        border=0,
    )
)
```

```{r}
#| echo: false
#| label: tbl-negative-binomial-spine-summary-py
#| tbl-cap: "Descriptive summary of female crab spine condition categories in the training data."
#| message: false
#| warning: false
#| results: asis

cat(
  "\n",
  py$negative_binomial_spine_summary_py_html,
  "\n",
  sep = ""
)
```

:::

The above categorical summaries show that the training data **are not evenly distributed** across the female crab colour and spine categories. For colour, most crabs are in the `Medium light` category, followed by `Medium dark`, while the `Light` and `Dark` categories are less common. For spine condition, most crabs are in the `Both worn or broken` category, with fewer observations in the `Both good` and `One worn or broken` categories. Sparse categories can make coefficient estimates less stable and can make category-level comparisons harder to interpret. For now, we treat these summaries as exploratory checks; later, the model will help us assess whether colour and spine condition appear associated with the expected number of satellite males after accounting for the other regressors.

### Distribution of the Count Response {#sec-negative-binomial-eda-response}

We now examine the distribution of the response variable `satellites` in the **training data**. This is the first EDA output that directly speaks to the Negative Binomial regression model. Before thinking about regressors, we need to understand the count we are trying to model: 

- *how often zero counts occur*, 
- *how concentrated the response is around small values*, 
- *whether there are unusually large counts*, and
- *whether the observed variability is already suggesting a possible goodness-of-fit issue*.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-teaching-to-learn-wiki-3976301/).](img/board.png){width="500"}

The response summaries, in @tbl-negative-binomial-response-summary-r, give a compact numerical view before we look at the response plot. The most important quantities for the  regression story are the **mean**, the **variance**, the **proportion of zeros**, and the **maximum observed count**.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-negative-binomial-response-summary-r
#| tbl-cap: "Descriptive summary statistics for the distribution of male satellite counts in the training data."
#| message: false
#| warning: false

response_summary <- training_data |>
  summarise(
    `Number of training observations` = n(),
    `Number of zero satellite counts` = sum(satellites == 0),
    `Proportion of zero satellite counts` = mean(satellites == 0),
    `Minimum number of satellite males` = min(satellites),
    `Median number of satellite males` = median(satellites),
    `Mean number of satellite males` = mean(satellites),
    `Variance of satellite males` = var(satellites),
    `Variance-to-mean ratio` = var(satellites) / mean(satellites),
    `Maximum number of satellite males` = max(satellites)
  ) |>
  mutate(
    across(
      where(is.numeric),
      ~ round(.x, 2)
    )
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "Quantity",
    values_to = "Value"
  ) |>
  mutate(
    Value = if_else(
      Quantity %in% c(
        "Proportion of zero satellite counts",
        "Mean number of satellite males",
        "Variance of satellite males",
        "Variance-to-mean ratio"
      ),
      formatC(Value, format = "f", digits = 2),
      formatC(Value, format = "f", digits = 0)
    )
  )

response_summary |>
  kable(
    align = c("c", "c")
  )
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

response_summary = pd.DataFrame(
    {
        "Quantity": [
            "Number of training observations",
            "Number of zero satellite counts",
            "Proportion of zero satellite counts",
            "Minimum number of satellite males",
            "Median number of satellite males",
            "Mean number of satellite males",
            "Variance of satellite males",
            "Variance-to-mean ratio",
            "Maximum number of satellite males",
        ],
        "Value": [
            len(training_data),
            (training_data["satellites"] == 0).sum(),
            (training_data["satellites"] == 0).mean(),
            training_data["satellites"].min(),
            training_data["satellites"].median(),
            training_data["satellites"].mean(),
            training_data["satellites"].var(ddof=1),
            (
                training_data["satellites"].var(ddof=1)
                / training_data["satellites"].mean()
            ),
            training_data["satellites"].max(),
        ],
    }
)

two_decimal_rows = [
    "Proportion of zero satellite counts",
    "Mean number of satellite males",
    "Variance of satellite males",
    "Variance-to-mean ratio",
]

response_summary["Value"] = [
    f"{value:.2f}" if quantity in two_decimal_rows
    else f"{value:.0f}"
    for quantity, value in zip(
        response_summary["Quantity"],
        response_summary["Value"]
    )
]

negative_binomial_response_summary_py_html = (
    response_summary.to_html(
        index=False,
        border=0,
    )
)
```

```{r}
#| echo: false
#| label: tbl-negative-binomial-response-summary-py
#| tbl-cap: "Descriptive summary statistics for the distribution of male satellite counts in the training data."
#| message: false
#| warning: false
#| results: asis

cat(
  "\n",
  py$negative_binomial_response_summary_py_html,
  "\n",
  sep = ""
)
```

:::

The above response summaries already tell us several important things. First, **zero counts are common**: `r sum(training_data$satellites == 0)` out of `r nrow(training_data)` female crabs in the training set have no satellite males, corresponding to a proportion of `r round(mean(training_data$satellites == 0), 2)`. Second, **the response is right-skewed**: the median number of satellite males is `r median(training_data$satellites)`, while the maximum observed training-set count is `r max(training_data$satellites)`. Third, **the sample variance, `r round(var(training_data$satellites), 2)`, is much larger than the sample mean, `r round(mean(training_data$satellites), 2)`**. The variance-to-mean ratio is approximately `r round(var(training_data$satellites) / mean(training_data$satellites), 2)`, which is an early warning sign of **overdispersion** in the data.

Note that the bar plot in @fig-negative-binomial-response-distribution-r (or @fig-negative-binomial-response-distribution-py) gives the same information visually. A bar plot is more appropriate than a smooth density plot because the response is a count. Each bar corresponds to a possible number of satellite males.

::: {.panel-tabset}

## **`R`**

```{r}
#| label: fig-negative-binomial-response-distribution-r
#| fig-cap: "Distribution of the number of satellite males in the training data."
#| warning: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

response_distribution_plot <- ggplot(training_data, aes(x = satellites)) +
  geom_bar(
    fill = "#0072B2",
    colour = "white",
    linewidth = 0.3
  ) +
  scale_x_continuous(
    breaks = seq(
      min(training_data$satellites),
      max(training_data$satellites),
      by = 1
    )
  ) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 15.5),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(
      size = 20,
      vjust = 0.5,
      margin = margin(r = 12)
    ),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x = "\n Number of satellite males",
    y = "Number of female crabs"
  ) +
  coord_cartesian(ylim = c(0, 35)) +
  scale_y_continuous(breaks = seq(0, 35, by = 5))

response_distribution_plot
```

## **`Python`**

```{python}
#| label: fig-negative-binomial-response-distribution-py
#| fig-cap: "Distribution of the number of satellite males in the training data."
#| warning: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

# Importing library
import matplotlib.pyplot as plt

satellite_counts = (
    training_data["satellites"]
    .value_counts()
    .sort_index()
)

response_distribution_figure, response_distribution_axis = plt.subplots(
    figsize=(14, 8)
)

response_distribution_axis.bar(
    satellite_counts.index,
    satellite_counts.values,
    edgecolor="white",
    linewidth=0.3,
    color="#0072B2"
)

response_distribution_axis.set_xlabel(
    "\n Number of satellite males",
    fontsize=20
)

response_distribution_axis.set_ylabel(
    "Number of female crabs",
    fontsize=20,
    labelpad=12
)

response_distribution_axis.tick_params(axis="both", labelsize=15.5)

response_distribution_axis.set_xticks(
    range(
        int(training_data["satellites"].min()),
        int(training_data["satellites"].max()) + 1
    )
)

response_distribution_axis.grid(
    True,
    which="major",
    axis="both",
    alpha=0.3
)

response_distribution_axis.grid(False, which="minor")

response_distribution_figure.tight_layout()
plt.show()
```

:::

The bar plot confirms that the training-set response is concentrated at zero and relatively small positive counts, with a thinner right tail extending to larger numbers of satellite males. This is critical for the **inferential inquiry** because the model must describe how the expected count changes with female crab characteristics while respecting the count scale. It is also important for the **predictive inquiry** because exact observed counts may be difficult to predict: many crabs have no satellites, while a smaller number have much larger counts. Therefore, later predictive results should focus on predicted **expected counts** and prediction errors, not on pretending that the model can deterministically identify the exact count for each crab.

We have to stress that, at this stage, the above response distribution does not prove that Negative Binomial regression is appropriate. The large variance relative to the mean suggests **possible overdispersion**, and the visible number of zeros indicates we should later compare observed and fitted counts carefully. These issues will become part of the goodness-of-fit story before we decide whether the Negative Binomial model is adequate enough for the chapter's two inquiries.

### Satellite Counts and Continuous Regressors {#sec-negative-binomial-eda-continuous}

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-problem-technology-3974187/).](img/plug-2.png){width="400"}

The next part of the EDA examines the response against the two continuous regressors: **female crab width** and **female crab weight**. The inferential inquiry asks whether the expected number of satellite males is associated with female crab characteristics, so these plots are the first visual check of whether body size appears relevant. The predictive inquiry also benefits from this step, as a useful predictive model requires regressors that carry information about the response. Hence, we first summarize the response across **quartiles of female crab width and weight**. These tables are not formal model outputs. They are descriptive summaries that help us read the plots more carefully.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-negative-binomial-width-quartile-summary-r
#| tbl-cap: "Descriptive summary statistics of male satellite counts across width quartiles in the training data."
#| message: false
#| warning: false

width_quartile_summary <- training_data |>
  mutate(
    width_quartile = ntile(width_cm, 4),
    width_quartile = factor(
      width_quartile,
      levels = 1:4,
      labels = c("Q1: smallest widths", "Q2", "Q3", "Q4: largest widths")
    )
  ) |>
  group_by(width_quartile) |>
  summarise(
    n = n(),
    mean_width_cm = mean(width_cm),
    mean_satellites = mean(satellites),
    median_satellites = median(satellites),
    proportion_zero_satellites = mean(satellites == 0),
    .groups = "drop"
  ) |>
  mutate(
    across(
      where(is.numeric),
      ~ round(.x, 2)
    )
  )

width_quartile_summary |>
  rename(
    `Width quartile` = width_quartile,
    `Number of female crabs` = n,
    `Mean width (cm)` = mean_width_cm,
    `Mean number of satellite males` = mean_satellites,
    `Median number of satellite males` = median_satellites,
    `Proportion with zero satellite males` = proportion_zero_satellites
  ) |>
  kable(
    align = c("c", "c", "c", "c", "c", "c")
  )
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

width_quartile_data = training_data.copy()

width_quartile_data = (
    width_quartile_data
    .sort_values(
        "width_cm",
        kind="mergesort"
    )
    .reset_index(drop=True)
)

number_of_rows = len(width_quartile_data)
number_of_groups = 4

base_group_size = number_of_rows // number_of_groups
remainder = number_of_rows % number_of_groups

group_sizes = [
    base_group_size + 1 if group_index < remainder else base_group_size
    for group_index in range(number_of_groups)
]

quartile_labels = [
    "Q1: smallest widths",
    "Q2",
    "Q3",
    "Q4: largest widths",
]

width_quartile_data["width_quartile"] = np.repeat(
    quartile_labels,
    group_sizes
)

width_quartile_data["width_quartile"] = pd.Categorical(
    width_quartile_data["width_quartile"],
    categories=quartile_labels,
    ordered=True
)

width_quartile_summary = (
    width_quartile_data
    .groupby("width_quartile", observed=False)
    .agg(
        n=("satellites", "size"),
        mean_width_cm=("width_cm", "mean"),
        mean_satellites=("satellites", "mean"),
        median_satellites=("satellites", "median"),
        proportion_zero_satellites=("satellites", lambda x: (x == 0).mean()),
    )
    .reset_index()
    .round(2)
)

width_quartile_summary_display = width_quartile_summary.rename(
    columns={
        "width_quartile": "Width quartile",
        "n": "Number of female crabs",
        "mean_width_cm": "Mean width (cm)",
        "mean_satellites": "Mean number of satellite males",
        "median_satellites": "Median number of satellite males",
        "proportion_zero_satellites": (
            "Proportion with zero satellite males"
        ),
    }
)

negative_binomial_width_quartile_summary_py_html = (
    width_quartile_summary_display.to_html(
        index=False,
        border=0,
    )
)
```

```{r}
#| echo: false
#| label: tbl-negative-binomial-width-quartile-summary-py
#| tbl-cap: "Descriptive summary statistics of male satellite counts across width quartiles in the training data."
#| message: false
#| warning: false
#| results: asis

cat(
  "\n",
  py$negative_binomial_width_quartile_summary_py_html,
  "\n",
  sep = ""
)
```

:::


The above width-quartile summary shows a clear **positive descriptive pattern** between female crab width and the number of satellite males: 

- In the **smallest-width quartile**, female crabs have an average of `r width_quartile_summary$mean_satellites[width_quartile_summary$width_quartile == "Q1: smallest widths"]` satellite males, with a median of `r width_quartile_summary$median_satellites[width_quartile_summary$width_quartile == "Q1: smallest widths"]`. 
- In the **largest-width quartile**, the average increases to `r width_quartile_summary$mean_satellites[width_quartile_summary$width_quartile == "Q4: largest widths"]`, with a median of `r width_quartile_summary$median_satellites[width_quartile_summary$width_quartile == "Q4: largest widths"]`. 
- The **middle quartiles** also follow the same general ordering: the mean satellite count rises from `r width_quartile_summary$mean_satellites[width_quartile_summary$width_quartile == "Q2"]` in Q2 to `r width_quartile_summary$mean_satellites[width_quartile_summary$width_quartile == "Q3"]` in Q3. 

This monotone increase across width quartiles suggests that female crab width is an important body-size regressor to examine first. Now, let us proceed with the width plots.

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on smoothing and count-response scatterplots!
::::
:::: {.Heads-up-container}
In both @fig-negative-binomial-satellites-width-r (or @fig-negative-binomial-satellites-width-py), the orange curve is a **LOWESS/LOESS** smooth. **LOWESS** stands for **locally weighted scatterplot smoothing**, and **LOESS** stands for **locally estimated scatterplot smoothing**. Both refer to closely related smoothing methods that draw a flexible trend line through a scatterplot by fitting many small local regressions, each using observations close to a given value on the horizontal axis.

In this chapter, we use the LOWESS/LOESS curve only as a visual guide. It is not the Negative Binomial regression model, and we will not interpret it as a fitted statistical model. Its purpose is simply to help us see whether satellite counts tend to be higher for wider female crabs.

Because we show this plot in both `R` and `Python`, we make the smoothing settings explicit. Although `R`'s `geom_smooth(method = "loess")` and Python's `lowess()` from [{statsmodels}](https://www.statsmodels.org/stable/index.html) are based on closely related ideas, their default settings are not identical. To make the two fitted curves visually comparable, we use the same smoothing fraction in both languages and align the local regression structure as closely as possible.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-emotion-pressure-6230208/).](img/surprise.png){width="475"}

Since `satellites` is a count response, the points naturally fall on horizontal bands at values such as $0$, $1$, $2$, and so on. These horizontal bands are useful: they remind us that the response is discrete, not continuous. When a continuous regressor such as `width_cm` is placed on the $x$-axis and a count response is placed on the $y$-axis, this banded structure is exactly what we should expect.
::::
:::

::: {.panel-tabset}

## **`R`**

```{r}
#| label: fig-negative-binomial-satellites-width-r
#| fig-cap: "Number of satellite males versus female crab width in the training data."
#| warning: false
#| message: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

satellites_width_plot <- ggplot(training_data, aes(x = width_cm, y = satellites)) +
  geom_point(
    alpha = 0.65,
    size = 2.3,
    colour = "#0072B2"
  ) +
  geom_smooth(
    method = "loess",
    formula = y ~ x,
    se = FALSE,
    linewidth = 1.2,
    colour = "#D55E00",
    method.args = list(
      span = 0.75,
      degree = 1,
      family = "gaussian"
    )
  )  +
  theme_bw() +
  theme(
    axis.text = element_text(size = 15.5),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(
      size = 20,
      vjust = 0.5,
      margin = margin(r = 12)
    ),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x = "\n Female crab width (cm)",
    y = "Number of satellite males"
  ) +
  scale_x_continuous(breaks = seq(22, 34, by = 2)) +
  scale_y_continuous(breaks = seq(0, 12, by = 2))

satellites_width_plot
```

## **`Python`**

```{python}
#| label: fig-negative-binomial-satellites-width-py
#| fig-cap: "Number of satellite males versus female crab width in the training data."
#| warning: false
#| message: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

# Importing function
from statsmodels.nonparametric.smoothers_lowess import lowess

width_x = training_data["width_cm"]
sat_y = training_data["satellites"]

smooth_width = lowess(
    endog=sat_y,
    exog=width_x,
    frac=0.75,
    it=0,
    return_sorted=True
)

satellites_width_figure, satellites_width_axis = plt.subplots(
    figsize=(14, 8)
)

satellites_width_axis.scatter(
    width_x,
    sat_y,
    alpha=0.65,
    s=35,
    color="#0072B2"
)

satellites_width_axis.plot(
    smooth_width[:, 0],
    smooth_width[:, 1],
    linewidth=1.2,
    color="#D55E00"
)

satellites_width_axis.set_xlabel(
    "\n Female crab width (cm)",
    fontsize=20
)

satellites_width_axis.set_ylabel(
    "Number of satellite males",
    fontsize=20,
    labelpad=12
)

satellites_width_axis.tick_params(axis="both", labelsize=15.5)
satellites_width_axis.set_xticks(range(22, 35, 2))
satellites_width_axis.set_yticks(range(0, 13, 2))
satellites_width_axis.grid(
    True,
    which="major",
    axis="both",
    alpha=0.3
)

satellites_width_axis.grid(False, which="minor")

satellites_width_figure.tight_layout()
plt.show()
```

:::

The width plot (from @fig-negative-binomial-satellites-width-r or @fig-negative-binomial-satellites-width-py) should be read together with the quartile (from @tbl-negative-binomial-width-quartile-summary-r). The blue points show the observed counts in the training set. Moreover, the orange LOWESS/LOESS curve is only an exploratory visual guide, but it suggests that **satellite counts tend to increase as female crab width increases**. This agrees with the quartile summary: the average satellite count rises from `r width_quartile_summary$mean_satellites[width_quartile_summary$width_quartile == "Q1: smallest widths"]` in the smallest-width quartile to `r width_quartile_summary$mean_satellites[width_quartile_summary$width_quartile == "Q4: largest widths"]` in the largest-width quartile. Then, the vertical spread around this increasing pattern is also important. Female crabs with similar widths can still have different numbers of satellite males. Some moderate-to-large crabs have zero or very few satellites, while some crabs have much larger counts. This tells us that width is a promising first regressor, but it will not fully explain the response on its own. We will use this EDA result to motivate the first Negative Binomial regression model (see @sec-negative-binomial-simple-model), while leaving formal model specification and interpretation for the next sections.

We now repeat the same EDA logic for female crab weight. Since width and weight both describe body size, we should expect them to tell related but not necessarily identical stories.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-negative-binomial-weight-quartile-summary-r
#| tbl-cap: "Descriptive summary statistics of male satellite counts across weight quartiles in the training data."
#| message: false
#| warning: false

weight_quartile_summary <- training_data |>
  mutate(
    weight_quartile = ntile(weight_kg, 4),
    weight_quartile = factor(
      weight_quartile,
      levels = 1:4,
      labels = c("Q1: smallest weights", "Q2", "Q3", "Q4: largest weights")
    )
  ) |>
  group_by(weight_quartile) |>
  summarise(
    n = n(),
    mean_weight_kg = mean(weight_kg),
    mean_satellites = mean(satellites),
    median_satellites = median(satellites),
    proportion_zero_satellites = mean(satellites == 0),
    .groups = "drop"
  ) |>
  mutate(
    across(
      where(is.numeric),
      ~ round(.x, 2)
    )
  )

weight_quartile_summary |>
  rename(
    `Weight quartile` = weight_quartile,
    `Number of female crabs` = n,
    `Mean weight (kg)` = mean_weight_kg,
    `Mean number of satellite males` = mean_satellites,
    `Median number of satellite males` = median_satellites,
    `Proportion with zero satellite males` = proportion_zero_satellites
  ) |>
  kable(
    align = c("c", "c", "c", "c", "c", "c")
  )
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

weight_quartile_data = training_data.copy()

weight_quartile_data = (
    weight_quartile_data
    .sort_values(
        "weight_kg",
        kind="mergesort"
    )
    .reset_index(drop=True)
)

number_of_rows = len(weight_quartile_data)
number_of_groups = 4

base_group_size = number_of_rows // number_of_groups
remainder = number_of_rows % number_of_groups

group_sizes = [
    base_group_size + 1 if group_index < remainder else base_group_size
    for group_index in range(number_of_groups)
]

quartile_labels = [
    "Q1: smallest weights",
    "Q2",
    "Q3",
    "Q4: largest weights",
]

weight_quartile_data["weight_quartile"] = np.repeat(
    quartile_labels,
    group_sizes
)

weight_quartile_data["weight_quartile"] = pd.Categorical(
    weight_quartile_data["weight_quartile"],
    categories=quartile_labels,
    ordered=True
)

weight_quartile_summary = (
    weight_quartile_data
    .groupby("weight_quartile", observed=False)
    .agg(
        n=("satellites", "size"),
        mean_weight_kg=("weight_kg", "mean"),
        mean_satellites=("satellites", "mean"),
        median_satellites=("satellites", "median"),
        proportion_zero_satellites=("satellites", lambda x: (x == 0).mean()),
    )
    .reset_index()
    .round(2)
)

weight_quartile_summary_display = weight_quartile_summary.rename(
    columns={
        "weight_quartile": "Weight quartile",
        "n": "Number of female crabs",
        "mean_weight_kg": "Mean weight (kg)",
        "mean_satellites": "Mean number of satellite males",
        "median_satellites": "Median number of satellite males",
        "proportion_zero_satellites": (
            "Proportion with zero satellite males"
        ),
    }
)

negative_binomial_weight_quartile_summary_py_html = (
    weight_quartile_summary_display.to_html(
        index=False,
        border=0,
    )
)
```

```{r}
#| echo: false
#| label: tbl-negative-binomial-weight-quartile-summary-py
#| tbl-cap: "Descriptive summary statistics of male satellite counts across weight quartiles in the training data."
#| message: false
#| warning: false
#| results: asis

cat(
  "\n",
  py$negative_binomial_weight_quartile_summary_py_html,
  "\n",
  sep = ""
)
```

:::

The weight-quartile summary also points toward a **positive body-size pattern**:

- In the **smallest-weight quartile**, the average satellite count is `r weight_quartile_summary$mean_satellites[weight_quartile_summary$weight_quartile == "Q1: smallest weights"]`, with a median of `r weight_quartile_summary$median_satellites[weight_quartile_summary$weight_quartile == "Q1: smallest weights"]`. 
- In the **largest-weight quartile**, the average satellite count is `r weight_quartile_summary$mean_satellites[weight_quartile_summary$weight_quartile == "Q4: largest weights"]`, with a median of `r weight_quartile_summary$median_satellites[weight_quartile_summary$weight_quartile == "Q4: largest weights"]`. 

Thus, in the training data, heavier female crabs tend to have higher satellite counts. However, this is still an exploratory pattern, **not a biological conclusion**. @tbl-negative-binomial-weight-quartile-summary-r groups a continuous regressor into four broad bins, and it does not adjust for other female crab characteristics. In particular, the weight summary should be interpreted alongside the width summary, not in isolation. Width and weight are both body-size measurements, so part of the apparent weight pattern may overlap with the width pattern. This is one reason the chapter begins with a simple width-only model before considering an extended model with additional regressors. We will treat weight as a candidate body-size regressor to examine later, while leaving formal interpretation to the fitted regression model.

::: {.panel-tabset}

## **`R`**

```{r}
#| label: fig-negative-binomial-satellites-weight-r
#| fig-cap: "Number of satellite males versus female crab weight in the training data."
#| warning: false
#| message: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

satellites_weight_plot <- ggplot(
  training_data,
  aes(x = weight_kg, y = satellites)
) +
  geom_point(
    alpha = 0.65,
    size = 2.3,
    colour = "#0072B2"
  ) +
  geom_smooth(
    method = "loess",
    formula = y ~ x,
    se = FALSE,
    linewidth = 1.2,
    colour = "#D55E00",
    method.args = list(
      span = 0.75,
      degree = 1,
      family = "gaussian"
    )
  ) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 15.5),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(
      size = 20,
      vjust = 0.5,
      margin = margin(r = 12)
    ),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x = "\n Female crab weight (kg)",
    y = "Number of satellite males"
  ) +
  scale_x_continuous(breaks = seq(1, 5.5, by = 0.5)) +
  scale_y_continuous(breaks = seq(0, 12, by = 2))

satellites_weight_plot
```

## **`Python`**

```{python}
#| label: fig-negative-binomial-satellites-weight-py
#| fig-cap: "Number of satellite males versus female crab weight in the training data."
#| warning: false
#| message: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

weight_x = training_data["weight_kg"]
sat_y = training_data["satellites"]

smooth_weight = lowess(
    endog=sat_y,
    exog=weight_x,
    frac=0.75,
    it=0,
    return_sorted=True
)

satellites_weight_figure, satellites_weight_axis = plt.subplots(
    figsize=(14, 8)
)

satellites_weight_axis.scatter(
    weight_x,
    sat_y,
    alpha=0.65,
    s=35,
    color="#0072B2"
)

satellites_weight_axis.plot(
    smooth_weight[:, 0],
    smooth_weight[:, 1],
    linewidth=1.2,
    color="#D55E00"
)

satellites_weight_axis.set_xlabel(
    "\n Female crab weight (kg)",
    fontsize=20
)

satellites_weight_axis.set_ylabel(
    "Number of satellite males",
    fontsize=20,
    labelpad=12
)

satellites_weight_axis.tick_params(axis="both", labelsize=15.5)
satellites_weight_axis.set_xticks(np.arange(1, 5.6, 0.5))
satellites_weight_axis.set_yticks(range(0, 13, 2))

satellites_weight_axis.grid(
    True,
    which="major",
    axis="both",
    alpha=0.3
)

satellites_weight_axis.grid(False, which="minor")

satellites_weight_figure.tight_layout()
plt.show()
```

:::

The weight plot, in @fig-negative-binomial-satellites-weight-r or @fig-negative-binomial-satellites-weight-py, should be read with the same caution as the width plot. The horizontal bands again show that the response is a count, not a continuous outcome. The orange smooth gives an exploratory body-size trend, and the quartile table gives the corresponding numerical summary. The main EDA message is that **heavier female crabs tend to have higher satellite counts**, but the relationship is not deterministic. There are still zero counts and small positive counts across much of the weight range. Therefore, weight may add useful information in an extended model, but it should not be treated as a complete explanation of satellite counts.

### Satellite Counts and Categorical Regressors {#sec-negative-binomial-eda-categorical}

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-pixel-college-3702064/).](img/plug.png){width="500"}

Now, we compare satellite counts across the categorical regressors `color` and `spine`. Earlier, in @sec-negative-binomial-eda-general, the marginal summaries showed that these categories are not evenly represented in the training data. That imbalance is critical here because a category with few observations can have an unstable mean satellite count. Therefore, in this subsection, we combine two pieces of information: the number of observations in each category and the distribution of satellite counts within each category.

We begin with female crab colour.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-negative-binomial-color-satellite-summary-r
#| tbl-cap: "Descriptive summary statistics of male satellite counts by female crab colour in the training data."
#| message: false
#| warning: false

color_satellite_summary <- training_data |>
  group_by(color) |>
  summarise(
    n = n(),
    mean_satellites = mean(satellites),
    median_satellites = median(satellites),
    proportion_zero_satellites = mean(satellites == 0),
    .groups = "drop"
  ) |>
  mutate(
    across(
      where(is.numeric),
      ~ round(.x, 2)
    )
  )

color_satellite_summary |>
  rename(
    `Female crab colour` = color,
    `Number of female crabs` = n,
    `Mean number of satellite males` = mean_satellites,
    `Median number of satellite males` = median_satellites,
    `Proportion with zero satellite males` = proportion_zero_satellites
  ) |>
  kable(
    align = c("c", "c", "c", "c", "c")
  )
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

color_satellite_summary = (
    training_data
    .groupby("color", observed=False)
    .agg(
        n=("satellites", "size"),
        mean_satellites=("satellites", "mean"),
        median_satellites=("satellites", "median"),
        proportion_zero_satellites=("satellites", lambda x: (x == 0).mean()),
    )
    .reset_index()
    .round(2)
)

color_satellite_summary_display = color_satellite_summary.rename(
    columns={
        "color": "Female crab colour",
        "n": "Number of female crabs",
        "mean_satellites": "Mean number of satellite males",
        "median_satellites": "Median number of satellite males",
        "proportion_zero_satellites": (
            "Proportion with zero satellite males"
        ),
    }
)

negative_binomial_color_satellite_summary_py_html = (
    color_satellite_summary_display.to_html(
        index=False,
        border=0,
    )
)
```

```{r}
#| echo: false
#| label: tbl-negative-binomial-color-satellite-summary-py
#| tbl-cap: "Descriptive summary statistics of male satellite counts by female crab colour in the training data."
#| message: false
#| warning: false
#| results: asis

cat(
  "\n",
  py$negative_binomial_color_satellite_summary_py_html,
  "\n",
  sep = ""
)
```

:::

The colour summary, in @tbl-negative-binomial-color-satellite-summary-r, must be read in light of the category sizes reported earlier. In this training split, the most common colour category is `Medium light`, while `Light` and `Dark` are much less common. The satellite-count summaries suggest that colour may be associated with the response, but the evidence is uneven across categories. The category with the highest mean satellite count is ``r as.character(color_satellite_summary$color[which.max(color_satellite_summary$mean_satellites)])``, with a mean of `r max(color_satellite_summary$mean_satellites)`. The category with the lowest mean satellite count is ``r as.character(color_satellite_summary$color[which.min(color_satellite_summary$mean_satellites)])``, with a mean of `r min(color_satellite_summary$mean_satellites)`. These are descriptive differences only. **Because some colour categories are sparse, the extended model should treat colour comparisons cautiously.**

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on horizontal jitter in categorical plots!
::::
:::: {.Heads-up-container}
In the subsequent plots, the orange points are slightly **jittered horizontally** within each female crab category. This means that points are moved a little to the left or right so that overlapping observations are easier to see. Note that this horizontal jitter is only a plotting device. It does **not** change the observed number of satellite males. The vertical positions of the points remain the original count values, so the horizontal bands at $0$, $1$, $2$, and so on are preserved. These bands are still important because they remind us that `satellites` is a count response.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-emotion-fear-expression-6230192/).](img/curious.png){width="250"}

We use a seed in the plotting code so that the small left-right movements are reproducible. In other words, the same code produces the same visual arrangement each time the chapter is rendered.
::::
:::

::: {.panel-tabset}

## **`R`**

```{r}
#| label: fig-negative-binomial-satellites-color-r
#| fig-cap: "Number of satellite males by female crab colour in the training data."
#| warning: false
#| message: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

set.seed(123)

satellites_color_plot <- ggplot(
  training_data,
  aes(x = color, y = satellites)
) +
  geom_boxplot(
    fill = "#0072B2",
    alpha = 0.45,
    outlier.shape = NA
  ) +
  geom_point(
    position = position_jitter(
      width = 0.18,
      height = 0,
      seed = 123
    ),
    alpha = 0.65,
    size = 2.2,
    colour = "#D55E00"
  ) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 15.5),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(
      size = 20,
      vjust = 0.5,
      margin = margin(r = 12)
    ),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x = "\n Female crab colour",
    y = "Number of satellite males"
  ) +
  scale_y_continuous(breaks = seq(0, 12, by = 2))

satellites_color_plot
```

## **`Python`**

```{python}
#| label: fig-negative-binomial-satellites-color-py
#| fig-cap: "Number of satellite males by female crab colour in the training data."
#| warning: false
#| message: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

rng = np.random.default_rng(123)

color_order = [
    category
    for category in ["Light", "Medium light", "Medium dark", "Dark"]
    if category in list(training_data["color"].astype(str).unique())
]

color_data = [
    training_data.loc[
        training_data["color"].astype(str) == category,
        "satellites"
    ].to_numpy()
    for category in color_order
]

satellites_color_figure, satellites_color_axis = plt.subplots(
    figsize=(14, 8)
)

color_box = satellites_color_axis.boxplot(
    color_data,
    tick_labels=color_order,
    patch_artist=True,
    showfliers=False
)

for patch in color_box["boxes"]:
    patch.set_facecolor("#0072B2")
    patch.set_alpha(0.45)

for index, category in enumerate(color_order, start=1):
    y_values = training_data.loc[
        training_data["color"].astype(str) == category,
        "satellites"
    ].to_numpy()

    x_values = index + rng.uniform(
        -0.18,
        0.18,
        size=len(y_values)
    )

    satellites_color_axis.scatter(
        x_values,
        y_values,
        alpha=0.65,
        s=30,
        color="#D55E00"
    )

satellites_color_axis.set_xlabel(
    "\n Female crab colour",
    fontsize=20
)

satellites_color_axis.set_ylabel(
    "Number of satellite males",
    fontsize=20,
    labelpad=12
)

satellites_color_axis.tick_params(axis="both", labelsize=15.5)
satellites_color_axis.set_yticks(range(0, 13, 2))

satellites_color_axis.grid(
    True,
    which="major",
    axis="both",
    alpha=0.3
)

satellites_color_axis.grid(False, which="minor")

satellites_color_figure.tight_layout()
plt.show()
```

:::

The colour plot, in @fig-negative-binomial-satellites-color-r or @fig-negative-binomial-satellites-color-py, reinforces the table's cautionary message. The boxplots and horizontally jittered points show **substantial overlap** in satellite counts across colour categories. This means colour may still be useful in an extended model, but it is unlikely to separate the response cleanly on its own. The horizontal jitter is used only to reduce overlap within each colour category; it does not move the counts vertically, so the integer count bands remain visible.

We now repeat the same comparison for female crab spine condition.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-negative-binomial-spine-satellite-summary-r
#| tbl-cap: "Descriptive summary statistics of male satellite counts by female crab spine condition in the training data."
#| message: false
#| warning: false

spine_satellite_summary <- training_data |>
  group_by(spine) |>
  summarise(
    n = n(),
    mean_satellites = mean(satellites),
    median_satellites = median(satellites),
    proportion_zero_satellites = mean(satellites == 0),
    .groups = "drop"
  ) |>
  mutate(
    across(
      where(is.numeric),
      ~ round(.x, 2)
    )
  )

spine_satellite_summary |>
  rename(
    `Female crab spine condition` = spine,
    `Number of female crabs` = n,
    `Mean number of satellite males` = mean_satellites,
    `Median number of satellite males` = median_satellites,
    `Proportion with zero satellite males` = proportion_zero_satellites
  ) |>
  kable(
    align = c("c", "c", "c", "c", "c")
  )
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

spine_satellite_summary = (
    training_data
    .groupby("spine", observed=False)
    .agg(
        n=("satellites", "size"),
        mean_satellites=("satellites", "mean"),
        median_satellites=("satellites", "median"),
        proportion_zero_satellites=("satellites", lambda x: (x == 0).mean()),
    )
    .reset_index()
    .round(2)
)

spine_satellite_summary_display = spine_satellite_summary.rename(
    columns={
        "spine": "Female crab spine condition",
        "n": "Number of female crabs",
        "mean_satellites": "Mean number of satellite males",
        "median_satellites": "Median number of satellite males",
        "proportion_zero_satellites": (
            "Proportion with zero satellite males"
        ),
    }
)

negative_binomial_spine_satellite_summary_py_html = (
    spine_satellite_summary_display.to_html(
        index=False,
        border=0,
    )
)
```

```{r}
#| echo: false
#| label: tbl-negative-binomial-spine-satellite-summary-py
#| tbl-cap: "Descriptive summary statistics of male satellite counts by female crab spine condition in the training data."
#| message: false
#| warning: false
#| results: asis

cat(
  "\n",
  py$negative_binomial_spine_satellite_summary_py_html,
  "\n",
  sep = ""
)
```

:::

The spine summary, in @tbl-negative-binomial-spine-satellite-summary-r, is dominated by the `Both worn or broken` category, which was already the largest group in the marginal spine table. This imbalance is important. The category with the highest mean satellite count is ``r as.character(spine_satellite_summary$spine[which.max(spine_satellite_summary$mean_satellites)])``, with a mean of `r max(spine_satellite_summary$mean_satellites)`. The category with the lowest mean satellite count is ``r as.character(spine_satellite_summary$spine[which.min(spine_satellite_summary$mean_satellites)])``, with a mean of `r min(spine_satellite_summary$mean_satellites)`. However, because at least one spine category has relatively few observations, these descriptive differences should not be over-interpreted before modelling.

::: {.panel-tabset}

## **`R`**

```{r}
#| label: fig-negative-binomial-satellites-spine-r
#| fig-cap: "Number of satellite males by female crab spine condition in the training data."
#| warning: false
#| message: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

set.seed(123)

satellites_spine_plot <- ggplot(
  training_data,
  aes(x = spine, y = satellites)
) +
  geom_boxplot(
    fill = "#0072B2",
    alpha = 0.45,
    outlier.shape = NA
  ) +
  geom_point(
    position = position_jitter(
      width = 0.18,
      height = 0,
      seed = 123
    ),
    alpha = 0.65,
    size = 2.2,
    colour = "#D55E00"
  ) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 15.5),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(
      size = 20,
      vjust = 0.5,
      margin = margin(r = 12)
    ),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x = "\n Female crab spine condition",
    y = "Number of satellite males"
  ) +
  scale_y_continuous(breaks = seq(0, 12, by = 2))

satellites_spine_plot
```

## **`Python`**

```{python}
#| label: fig-negative-binomial-satellites-spine-py
#| fig-cap: "Number of satellite males by female crab spine condition in the training data."
#| warning: false
#| message: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

spine_order = [
    category
    for category in [
        "Both good",
        "One worn or broken",
        "Both worn or broken",
    ]
    if category in list(training_data["spine"].astype(str).unique())
]

spine_data = [
    training_data.loc[
        training_data["spine"].astype(str) == category,
        "satellites"
    ].to_numpy()
    for category in spine_order
]

satellites_spine_figure, satellites_spine_axis = plt.subplots(
    figsize=(14, 8)
)

spine_box = satellites_spine_axis.boxplot(
    spine_data,
    tick_labels=spine_order,
    patch_artist=True,
    showfliers=False
)

for patch in spine_box["boxes"]:
    patch.set_facecolor("#0072B2")
    patch.set_alpha(0.45)

for index, category in enumerate(spine_order, start=1):
    y_values = training_data.loc[
        training_data["spine"].astype(str) == category,
        "satellites"
    ].to_numpy()

    x_values = index + rng.uniform(
        -0.18,
        0.18,
        size=len(y_values)
    )

    satellites_spine_axis.scatter(
        x_values,
        y_values,
        alpha=0.65,
        s=30,
        color="#D55E00"
    )

satellites_spine_axis.set_xlabel(
    "\n Female crab spine condition",
    fontsize=20
)

satellites_spine_axis.set_ylabel(
    "Number of satellite males",
    fontsize=20,
    labelpad=12
)

satellites_spine_axis.tick_params(axis="both", labelsize=15.5)
satellites_spine_axis.set_yticks(range(0, 13, 2))

satellites_spine_axis.grid(
    True,
    which="major",
    axis="both",
    alpha=0.3
)

satellites_spine_axis.grid(False, which="minor")

satellites_spine_figure.tight_layout()
plt.show()
```

:::

The spine plot, in @fig-negative-binomial-satellites-spine-r or @fig-negative-binomial-satellites-spine-py, again shows **overlap** across categories. The largest category contains crabs with a wide range of satellite counts, including zero counts and larger positive counts. The smaller categories are harder to compare because they contain fewer observations. This makes spine a reasonable candidate regressor for the extended model, but not one that should be expected to dominate the body-size pattern seen for width and weight.

### Summary {#sec-negative-binomial-eda-summary}

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-blended-learning-6230153/).](img/blender.png){width="400"}

Before moving to the GLM formulation of  regression, it is useful to summarize the main EDA findings. @tbl-negative-binomial-eda-summary records what we learned from the **training data** and how each finding will guide the next parts of the workflow.

| EDA component | Main finding from the training data | Why it matters for the  regression workflow |
|:---:|:---:|:---:|
| Response distribution | The response `satellites` is a **non-negative integer count**. The training data contain `r sum(training_data$satellites == 0)` zero counts out of `r nrow(training_data)` observations. The mean number of satellite males is `r round(mean(training_data$satellites), 2)`, while the variance is `r round(var(training_data$satellites), 2)`, giving a variance-to-mean ratio of `r round(var(training_data$satellites) / mean(training_data$satellites), 2)`. |  regression is a **natural starting point** because the response is a count. However, the variance is much larger than the mean, so the Classical  mean-variance structure will need to be checked carefully later. |
| Count-response plot | The bar plot shows **many zero and small positive counts**, with a **thinner right tail** extending to larger satellite counts. | This reinforces that we are modelling a **discrete, right-skewed count response**. The large counts and the visible number of zeros motivate later observed-versus-fitted checks. |
| Width and satellite counts | The width-quartile summary shows a **clear positive descriptive pattern**: the mean satellite count rises from `r width_quartile_summary$mean_satellites[width_quartile_summary$width_quartile == "Q1: smallest widths"]` in the smallest-width quartile to `r width_quartile_summary$mean_satellites[width_quartile_summary$width_quartile == "Q4: largest widths"]` in the largest-width quartile. The width scatterplot shows the same general upward pattern, while preserving the horizontal bands created by the count response. | Width has the clearest descriptive body-size pattern in the EDA, so it is a sensible **first regressor for the simple Negative Binomial regression model**. The vertical spread at similar widths also warns us that width alone will not fully explain the response. |
| Weight and satellite counts | The weight-quartile summary also suggests a **positive descriptive association**: the mean satellite count changes from `r weight_quartile_summary$mean_satellites[weight_quartile_summary$weight_quartile == "Q1: smallest weights"]` in the smallest-weight quartile to `r weight_quartile_summary$mean_satellites[weight_quartile_summary$weight_quartile == "Q4: largest weights"]` in the largest-weight quartile. The weight plot shows a related body-size pattern, again with substantial variability around the visual trend. | Weight is a **candidate body-size regressor for the extended model**. However, width and weight should not be interpreted in isolation because they both describe body size and may partly carry overlapping information. |
| Female crab colour | Colour groups have **uneven sample sizes**. The category with the highest mean satellite count is ``r as.character(color_satellite_summary$color[which.max(color_satellite_summary$mean_satellites)])``, with a mean of `r max(color_satellite_summary$mean_satellites)`. The category with the lowest mean satellite count is ``r as.character(color_satellite_summary$color[which.min(color_satellite_summary$mean_satellites)])``, with a mean of `r min(color_satellite_summary$mean_satellites)`. | Colour may be useful in the **extended model**, but category-level comparisons should be treated cautiously because some colour groups are **sparse** and the boxplots show **overlap** across categories. |
| Female crab spine condition | Spine groups are also **imbalanced**. The category with the highest mean satellite count is ``r as.character(spine_satellite_summary$spine[which.max(spine_satellite_summary$mean_satellites)])``, with a mean of `r max(spine_satellite_summary$mean_satellites)`. The category with the lowest mean satellite count is ``r as.character(spine_satellite_summary$spine[which.min(spine_satellite_summary$mean_satellites)])``, with a mean of `r min(spine_satellite_summary$mean_satellites)`. | Spine condition is worth checking in the **extended model**, but the EDA does not suggest that it should dominate the body-size regressors. The **imbalance** across spine categories also means coefficient-level interpretations should be cautious. |
| Predictive challenge | Across the continuous and categorical plots, female crabs with similar regressor values can still have quite different satellite counts. Zero counts also occur across multiple regions of the regressor space. | The predictive inquiry should be evaluated with **held-out testing data** and **prediction-error summaries**. Visual patterns from the training data can motivate model choices, but they are not enough to assess predictive performance. |

: Summary of the main EDA findings and their implications for the  regression workflow. {#tbl-negative-binomial-eda-summary .striped .hover}

Overall, the EDA gives us a coherent modelling story. The response is a non-negative integer count with many zeros and variability that is much larger than the mean. Width shows the clearest descriptive association with satellite counts, while weight provides a related body-size signal that should be interpreted alongside width. Colour and spine condition may also contribute, but their category imbalance and overlapping count distributions call for caution. These findings do not establish final inferential conclusions; rather, they motivate the model-building sequence that follows. We now move to the GLM formulation so that the response distribution, regression structure, and coefficient interpretation are made explicit before estimation.

## Negative Binomial Regression as a Generalized Linear Model

### The Random Component

Let $Y_i$ denote the number of satellite males observed for the $i$th female crab in the training data, for $i = 1, 2, \ldots, n$. We treat $Y_i$ as a random variable because a different sample of female crabs, or a repeated version of the same observational process, could produce different satellite counts. In Negative Binomial regression, the random component assumes that the conditional distribution of $Y_i$ is Negative Binomial:

$$
Y_i \mid x_{i,1}, x_{i,2}, \ldots, x_{i,k} \sim \operatorname{Negative Binomial}(\theta,\mu_i),
$${#eq-negative-binomial-random-component}

where $\mu_i > 0$ is the conditional expected number of satellite males for the $i$th female crab. The value of $\mu_i$ is allowed to change from crab to crab because different female crabs can have different regressor values.

Let's imagine a series of independent Bernoulli trials, each with probability $p$ of success, such that $0<p<1$. A Negative Binomial distribution is commonly parameterized as $Z\sim\operatorname{Negative Binomial}(\theta,p)$, where $Z$ is the number of failed Bernoulli trials before experiencing $\theta$ independent wins. The expected of $Z$ is the expected number of failed Bernoulli trials before experiencing $\theta$ independent wins, and is given by

$$
\mu=\mathbb{E}(Z)=\frac{\theta(1-p)}{ p}
$$
Rearranging this equation and isolating for $p$ gives

$$
p=\left(\frac{1}{1+\frac{\mu}{\theta}}\right)
$$
For a constant value of $\theta$, there is a one-to-one correspondence between $p$ and $\mu$ and therefore the random component can be parameterized by $\mu_i$ and $\theta$, where $\mu_i$ is the conditional expected number of satellite males for the $i$th female crab and $\theta$ is the dispersion parameter.

Equivalently, for a possible count $y_i \in \{0, 1, 2, \ldots\}$, the conditional probability mass function (PMF) is

$$
p_Y(y_i; \mu_i, \theta)
=
\Pr(Y_i = y_i \mid x_{i,1}, x_{i,2}, \ldots, x_{i,k})
=
\frac{\Gamma(y_i + \theta)}{\Gamma(\theta) \, y_i!}
\left( \frac{\theta}{\theta + \mu_i} \right)^{\theta}
\left( \frac{\mu_i}{\theta + \mu_i} \right)^{y_i},
$${#eq-negative-binomial-pmf-regression}

The parameter $\mu_i$ has two roles in the Negative Binomial model:

$$
\mathbb{E}(Y_i \mid x_{i,1}, x_{i,2}, \ldots, x_{i,k}) = \mu_i,
$$

and

$$
\operatorname{Var}(Y_i \mid x_{i,1}, x_{i,2}, \ldots, x_{i,k}) = \mu_i+\frac{\mu_i^2}{\theta}.
$$
For the horseshoe crab data, the random component says that each observed satellite count is **a realization of a Negative Binomial random variable**, with **its own conditional mean $\mu_i$**. This is different from saying that all female crabs have the same expected number of satellite males. The whole purpose of the regression model is to let $\mu_i$ vary systematically with female crab characteristics such as width, weight, colour, and spine condition.

### The Systematic Component

The systematic component describes how the regressors are combined before being linked to the conditional mean. With $k$ regressors, the systematic component is

$$
\eta_i
=
\beta_0 + \beta_1 x_{i,1} + \beta_2 x_{i,2} + \cdots + \beta_k x_{i,k},
$${#eq--systematic-component}

where:

- $\eta_i$ is called the **linear predictor** for the $i$th female crab;
- $\beta_0$ is the intercept;
- $\beta_1, \beta_2, \ldots, \beta_k$ are fixed but unknown regression coefficients;
- $x_{i,1}, x_{i,2}, \ldots, x_{i,k}$ are the observed regressor values for the $i$th female crab.

The term **linear predictor** can be confusing at first. It means that the regressors are combined linearly in the regression coefficients. Thus, it does not mean that the expected count itself must be a linear function of the regressors on the original count scale.

In the horseshoe crab case study, a simple model may use only female crab width:

$$
\eta_i = \beta_0 + \beta_1 x_{i,1},
$$

where $x_{i,1}$ represents female crab width in centimetres. An extended model may include additional regressors, such as weight, colour indicators, and spine-condition indicators. In that case, $k$ is larger and the systematic component includes more terms.

For example, using generic regressor notation, we can write the extended model as

$$
\eta_i
=
\beta_0
+
\beta_1 x_{i,1}
+
\beta_2 x_{i,2}
+
\cdots
+
\beta_k x_{i,k}.
$$

**This notation deliberately uses $x_{i,1}, x_{i,2}, \ldots, x_{i,k}$ rather than variable names in the equation.** Some of these regressors may be continuous, such as width or weight, while others may be indicator (or dummy) variables created from categorical regressors such as colour or spine condition. We will make the exact model specification explicit when fitting the simple and extended models.

### The Log Link Function

The link function connects the conditional mean $\mu_i$ to the systematic component $\eta_i$. Like [Classical  regression](@sec-classical-), the standard link function in Negative Binomial regression is the **log link** because the log function maps each $\mu_i>0$ in a one-to-one fashion onto the set of real numbers between $-\infty$ and $\infty$:

$$
\log(\mu_i) = \eta_i.
$${#eq-negative-binomial-log-link}

Combining @eq-negative-binomial-log-link with the systematic component in @eq-negative-binomial-systematic-component gives

$$
\log(\mu_i)
=
\beta_0 + \beta_1 x_{i,1} + \beta_2 x_{i,2} + \cdots + \beta_k x_{i,k}.
$$

Since the logarithm is applied to $\mu_i$, we can rewrite the model on the expected-count scale by exponentiating both sides:

$$
\mu_i
=
\exp\left(
\beta_0 + \beta_1 x_{i,1} + \beta_2 x_{i,2} + \cdots + \beta_k x_{i,k}
\right).
$${#eq-negative-binomial-mean-model}

This form in @eq-negative-binomial-mean-model shows why the log link is useful. The expected count $\mu_i$ must be positive, and the exponential function always produces a positive value. Thus, the model can use a linear predictor $\eta_i$ that may be any real number while still producing a valid positive conditional mean. Furthermore, the log link also changes how coefficients are interpreted:

- In OLS, a **one-unit increase** in a regressor is usually interpreted as an **additive change in the conditional mean of the response**. 
- In Negative Binomial regression with a log link, coefficients are interpreted through **multiplicative changes in the expected count after exponentiation**. 

We will return to this point when interpreting the different fitted coefficients in subsequent sections.

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on why we do not use an identity link here!
::::
:::: {.Heads-up-container}
A tempting alternative would be to use an **identity link**, which would set the conditional mean directly equal to the systematic component:

$$
\mu_i
=
\beta_0 + \beta_1 x_{i,1} + \beta_2 x_{i,2} + \cdots + \beta_k x_{i,k}.
$$

This may look familiar because it resembles the conditional mean structure used in OLS. However, **it is usually not a good default choice for Negative Binomial regression**. The main issue is that $\mu_i$ is an expected count, so it must be positive. A linear expression can be negative for some regressor values, which would produce an invalid expected count.

Hence, the log link avoids this problem. It lets the systematic component $\eta_i$ take any real value, while the expected count

$$
\mu_i = \exp(\eta_i)
$$

is always positive. This is one reason the log link is the standard link function for **count regression** models, such as Negative Binomial and Classical .
::::
:::

Putting the three GLM pieces together, the Negative Binomial regression model with $k$ regressors is

$$
Y_i \mid x_{i,1}, x_{i,2}, \ldots, x_{i,k} \sim \operatorname{Negative Binomial}(\theta,\mu_i),
$$

with

$$
\log(\mu_i)
=
\beta_0 + \beta_1 x_{i,1} + \beta_2 x_{i,2} + \cdots + \beta_k x_{i,k}.
$$

Equivalently,

$$
\mu_i
=
\exp\left(
\beta_0 + \beta_1 x_{i,1} + \beta_2 x_{i,2} + \cdots + \beta_k x_{i,k}
\right).
$$

These equations summarize the model we will fit next:

1. The **random component** tells us that the response is conditionally Negative Binomial. 
2. The **systematic component** tells us how the regressors are combined. 
3. The **log link** tells us how that linear predictor becomes a positive expected count.

In @sec-negative-binomial-eda, the EDA helped us decide that this model is a reasonable choice, especially because the response is a count and width shows a clear descriptive association with satellite counts (see @fig-negative-binomial-satellites-width-r or @fig-negative-binomial-satellites-width-py). The EDA also showed us that the Classical  assumptions may be strained because of possible overdispersion in the counts of satellite crabs. After fitting the model, we will need to examine whether the fitted values reproduce the main features of the observed counts and whether the mean-variance structure is adequate enough for the chapter's inferential and predictive goals.

## Fitting a Simple Negative Binomial Regression Model {#sec-negative-binomial-simple-model}

Let's recall from [Section 10.8.2](https://alexrod61.github.io/regression-cookbook/book/10-classical-.html#sec-classical--equidispersion) in the Classical  Regression Chapter that there is statistical evidence of **overdispersion** in the training data. The overdispersion test $p-$value is $8.61e-19$ in this model, which provides formal evidence against the equidispersion assumption. 

Now, we will fit the first Negative Binomial regression model using the crabs data from [Chapter 10](https://alexrod61.github.io/regression-cookbook/book/10-classical-.html). The purpose of this model is deliberately modest: we begin with **female crab width** as the only regressor because the EDA in @sec-negative-binomial-eda showed the clearest descriptive body-size pattern for width. This simple model gives us a transparent starting point for studying estimation, fitted expected counts, and goodness-of-fit before moving to a richer model with additional regressors.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-pixel-coffee-3702057/).](img/coffee.png){width="610"}

At this point, we will **not** provide a full substantive interpretation of the estimate width regression coefficient coming from this model. That interpretation belongs later, after we have checked whether the simple Negative Binomial regression model is adequate and after we have considered an extended model with additional regressors. This ordering is critical for the workflow of the chapter:

1. Use the **training data** for EDA, model fitting, and goodness-of-fit checks.
2. Fit a simple Negative Binomial regression model with width as the only regressor.
3. Check whether the fitted model reproduces important features of the observed count response.
4. Extend the model if needed by adding other regressors such as weight, colour, and spine condition.
5. Save final coefficient-level interpretation for the later modelling sections, after model adequacy has been discussed.

This is necessary for the inferential inquiry. We should not use the same data both to explore the model structure and to produce the final inferential claims. Recall that, for the inferential inquiry, the training set is used for model development and goodness-of-fit assessment. Final inferential outputs will be produced later after the model-building workflow has been completed.

### Model Specification 

Before fitting the width-only model, we first write the general Negative Binomial regression model with $k$ regressors. Let

$$
\mathbf{x}_i
=
\begin{bmatrix}
1 \\
x_{i,1} \\
x_{i,2} \\
\vdots \\
x_{i,k}
\end{bmatrix}
$$

be the regressor vector for the $i$th observation, **including the leading $1$ for the intercept**. Then, let

$$
\boldsymbol{\beta}
=
\begin{bmatrix}
\beta_0 \\
\beta_1 \\
\beta_2 \\
\vdots \\
\beta_k
\end{bmatrix}
$$

be the vector of regression coefficients. The systematic component can then be written compactly as

$$
\eta_i = \mathbf{x}_i^\top \boldsymbol{\beta},
$$

where

$$
\mathbf{x}_i^\top \boldsymbol{\beta}
=
\beta_0 + \beta_1 x_{i,1} + \beta_2 x_{i,2} + \cdots + \beta_k x_{i,k}.
$$

For the **full training sample** with $n$ observations, we can collect the response values into

$$
\mathbf{y}
=
\begin{bmatrix}
y_1 \\
y_2 \\
\vdots \\
y_n
\end{bmatrix},
$$

and the regressors into the design matrix

$$
\mathbf{X}
=
\begin{bmatrix}
1 & x_{1,1} & x_{1,2} & \cdots & x_{1,k} \\
1 & x_{2,1} & x_{2,2} & \cdots & x_{2,k} \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
1 & x_{n,1} & x_{n,2} & \cdots & x_{n,k}
\end{bmatrix}.
$$

The $i$th row of $\mathbf{X}$ is $\mathbf{x}_i^\top$. The vector of **systematic scores** is

$$
\boldsymbol{\eta}
=
\mathbf{X}\boldsymbol{\beta}.
$$

As indicated in @sec-negative-binomial-log-link, in Negative Binomial regression, the conditional mean for the $i$th observation is

$$
\mu_i
=
\exp(\eta_i)
=
\exp(\mathbf{x}_i^\top \boldsymbol{\beta}).
$$

Thus, the general model with $k$ regressors is

$$
Y_i \mid \mathbf{x}_i \sim \operatorname{Negative Binomial}(\theta,\mu_i),
$$

with

$$
\log(\mu_i) = \mathbf{x}_i^\top \boldsymbol{\beta}.
$$
The width-only model is the special case with one regressor. Let $x_{i,1}$ denote the female crab width in centimetres for the $i$th female crab. The simple model is

$$
Y_i \mid x_{i,1} \sim \operatorname{Negative Binomial}(\theta,\mu_i),
$$

with

$$
\log(\mu_i) = \beta_0 + \beta_1 x_{i,1}.
$${#eq-negative-binomial-simple-width-model}

Equivalently,

$$
\mu_i
=
\exp(\beta_0 + \beta_1 x_{i,1}).
$${#eq-negative-binomial-simple-width-mean}

In @eq-negative-binomial-simple-width-model and @eq-negative-binomial-simple-width-mean, $\mu_i$ is the expected number of satellite males for a female crab with width $x_{i,1}$. The model assumes

$$
\operatorname{Var}(Y_i \mid x_{i,1}) = \mu_i+\frac{\mu_i^2}{\theta}.
$$

Therefore, a Negative Binomial model assumes **overdispersion** in the response variable. Once width is included, the model assumes that the conditional variance of the satellite count is higher than the conditional mean for crabs with the same width value. This is stronger than merely saying that the raw sample variance should be higher than the raw sample mean. It is a model statement about the distribution of counts **after conditioning on width**.

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on why we do not interpret the width coefficient yet!
::::
:::: {.Heads-up-container}
The simple width-only model is handy because it lets us introduce maximum likelihood estimation in a transparent setting. However, we will not yet use this fitted model to make final claims about the association between female crab width and the expected number of satellite males.

There are two reasons for waiting. First, we still need to check whether the simple Classical  model fits the training data adequately. Second, the EDA suggested that other female crab characteristics, such as weight, colour, and spine condition, may also matter. A width-only coefficient can change once additional regressors are included.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-emotion-upset-6230193/).](img/surprise-2.png){width="300"}

For now, the fitted width coefficient is an estimation output. We will return to substantive coefficient interpretation after the goodness-of-fit checks and the extended model.
::::
:::

### Estimation {#sec-negative-binomialsimple-estimation}

Negative Binomial regression is estimated by **maximum likelihood estimation (MLE)**. The basic idea is to choose the coefficient vector $\boldsymbol{\beta}$ that makes the observed training data most compatible with the model. Following up with @eq-negative-binomial-pmf-regression, for one observation, the probability mass function (PMF) is

$$
p_Y(y_i; \mu_i, \theta)
=
\Pr(Y_i = y_i \mid x_{i,1}, x_{i,2}, \ldots, x_{i,k})
=
\frac{\Gamma(y_i + \theta)}{\Gamma(\theta) \, \Gamma(y_i+1)}
\left( \frac{\theta}{\theta + \mu_i} \right)^{\theta}
\left( \frac{\mu_i}{\theta + \mu_i} \right)^{y_i},
$$

where

$$
\mu_i = \exp(\mathbf{x}_i^\top \boldsymbol{\beta}).
$$

Assuming the training observations are independent conditional on their regressors, the likelihood function for $\boldsymbol{\beta}$, $\theta$ is

$$
\mathcal{L}(\boldsymbol{\beta},\theta;\mathbf{y})
=
\prod_{i=1}^n
\frac{\Gamma(y_i + \theta)}{\Gamma(\theta) \, y_i!}
\left( \frac{\theta}{\theta + \mu_i} \right)^{\theta}
\left( \frac{\mu_i}{\theta + \mu_i} \right)^{y_i}.
$${#eq-negative-binomial-likelihood}

As products can be hard to work with directly, we usually maximize the log-likelihood function. Taking the logarithm of @eq-negative-binomial-likelihood gives

$$
\ell(\boldsymbol{\beta}, \theta; \mathbf{y})
=
\sum_{i=1}^{n}
\left[
\log \Gamma(y_i + \theta)
- \log \Gamma(\theta)
- \log \Gamma(y_i + 1)
+ \theta \log \theta
+ y_i \log(\mu_i)
- (y_i + \theta) \log \left( \theta + \mu_i \right)
\right].
$${#eq-negative-binomial-log-likelihood-mu}

Using $\log(\mu_i) = \mathbf{x}_i^\top \boldsymbol{\beta}$ and $\mu_i = \exp(\mathbf{x}_i^\top \boldsymbol{\beta})$, we can rewrite @eq-negative-binomial-log-likelihood-mu as

$$
\ell(\boldsymbol{\beta}, \theta)
=
\sum_{i=1}^{n}
\left[
\log \Gamma(y_i + \theta)
- \log \Gamma(\theta)
- \log \Gamma(y_i + 1)
+ \theta \log \theta
+ y_i \, \mathbf{x}_i^\top \boldsymbol{\beta}
- (y_i + \theta) \log \left( \theta + \exp(\mathbf{x}_i^\top \boldsymbol{\beta}) \right)
\right].
$${#eq-negative-binomial-log-likelihood-beta}

The MLE is the set of values of $\boldsymbol{\beta}$ and $\theta$ that maximize this log-likelihood:

$$
\hat{\boldsymbol{\beta}}_{\operatorname{MLE,obs}}
=
\underset{\boldsymbol{\beta},\theta}{\operatorname{argmax}}
\;
\ell(\boldsymbol{\beta},\theta;\mathbf{y}).
$${#eq-negative-binomial-beta-mle}

There is no OLS-style closed-form solution for @eq-negative-binomial-beta-mle. 

*** Template ***

Therefore, software uses iterative numerical algorithms, such as Fisher scoring or **Iteratively Reweighted Least Squares (IRLS)** [@nelder1972], to find the coefficient estimates. The IRLS algorithm starts with an initial guess for the coefficient vector $\boldsymbol{\beta}$, updates the fitted expected counts, builds a locally weighted least-squares problem, updates the coefficients, and repeats this process until the coefficient estimates stop changing in a meaningful way.


To see the estimating logic without going into matrix calculus, let $x_{i,j}$ denote the value of the $j$th regressor for observation $i$, with $x_{i,0}=1$ for the intercept. Since

$$
\mu_i = \exp(\beta_0 + \beta_1 x_{i,1} + \cdots + \beta_k x_{i,k}),
$$

the derivative of $\mu_i$ with respect to $\beta_j$ is

$$
\frac{\partial \mu_i}{\partial \beta_j}
=
x_{i,j}\mu_i.
$$

Differentiating the log-likelihood in @eq-negative-binomial-log-likelihood-beta with respect to $\beta_j$ gives the score component

$$
\frac{\partial \ell(\boldsymbol{\beta};\mathbf{y})}{\partial \beta_j}
=
\sum_{i=1}^n
x_{i,j}
(y_i - \mu_i),
\qquad j = 0, 1, \ldots, k.
$$

At the MLE, the score equations are set to zero:

$$
\sum_{i=1}^n
x_{i,j}
(y_i - \hat{\mu}_i)
=
0,
\qquad j = 0, 1, \ldots, k,
$$

where

$$
\hat{\mu}_i
=
\exp(\mathbf{x}_i^\top \hat{\boldsymbol{\beta}}_{\operatorname{MLE,obs}}).
$$

These equations show that the fitted model balances observed and fitted counts **after weighting those differences by each regressor**. Equivalently,

$$
\sum_{i=1}^n
x_{i,j}y_i
=
\sum_{i=1}^n
x_{i,j}\hat{\mu}_i,
\qquad j = 0, 1, \ldots, k.
$$

This is more general than the intercept-only intuition. For the intercept, since $x_{i,0}=1$, the score equation becomes

$$
\sum_{i=1}^n
(y_i - \hat{\mu}_i)
=
0,
$$

so that

$$
\sum_{i=1}^n y_i
=
\sum_{i=1}^n \hat{\mu}_i.
$$

This means that an intercept-only Negative Binomial regression model balances the **total observed count** and the **total fitted expected count**. But the other score equations are also necessary. For example, in the simple width-only model, the width score equation is

$$
\sum_{i=1}^n
x_{i,1}y_i
=
\sum_{i=1}^n
x_{i,1}\hat{\mu}_i.
$$

Thus, the fitted model also balances the observed and fitted counts after multiplying them by female crab width. With more regressors, the same kind of balancing condition holds for each regressor in the model.

*** Template ****

We now fit the simple width-only model to the training data. The fitted model is

$$
Y_i \mid x_{i,1} \sim \operatorname{Negative Binomial}(\theta,\mu_i),
$$

with

$$
\log(\mu_i) = \beta_0 + \beta_1 x_{i,1},
$$

where $x_{i,1}$ is female crab width in centimetres.

Note the following:

- In the `R` code below, the function `glm.nb` from `MASS` library is used to fit a Negative Binomial Regression Model. The argument`formula = satellites ~ width_cm` below specifies the response on the left-hand side and the regressor on the right-hand side. The argument `data = training_data` ensures that the model is fitted only on the training set. Note that we use `tidy()` from [{broom}](https://broom.tidymodels.org/reference/broom.html) for a clean tabular display.
- The `Python` code uses the analogous [{statsmodels}](https://www.statsmodels.org/stable/index.html) formula interface, but the fit proceeds in two steps. Unlike `glm.nb()`, which estimates the dispersion parameter jointly with the regression coefficients, `statsmodels` requires the dispersion to be supplied to the family in advance. Hence, `smf.negativebinomial()` is first used to obtain the maximum likelihood estimate of the dispersion parameter `alpha`, which is the reciprocal of the `theta` reported by `glm.nb()`. The argument `disp=0` suppresses the optimizer's convergence output. Then `smf.glm()` refits the model holding this estimate fixed, where `formula="satellites ~ width_cm"` gives the response-regressor relationship, `data=training_data` identifies the training dataset, and `family=sm.families.NegativeBinomial(alpha=simple_negative_binomial_width_alpha)` specifies the Negative Binomial GLM with the estimated dispersion. This two-step structure reproduces the standard errors reported by `glm.nb()`, which are conditional on the estimated dispersion rather than accounting for uncertainty in it.

@tbl-negative-binomial-simple-width-fit-r gives the MLEs and their model-based standard errors for the intercept and width terms. These are the basic numerical ingredients needed for later inference, but we should not jump directly to substantive interpretation. The next step is to check whether the fitted model is adequate enough to support the chapter's inferential and predictive goals.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-negative-binomialsimple-width-fit-r
#| tbl-cap: "Maximum likelihood estimates from the simple width-only Negative Binomial regression model fitted on the training data."
#| message: false
#| warning: false

library(broom)

simple_negative_binomial_width_model <- glm.nb(formula = satellites ~ width_cm,
                                               data = training_data)

simple_negative_binomial_width_summary <- tidy(simple_negative_binomial_width_model) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))

simple_negative_binomial_width_summary |>
  rename(`Term` = term, `Estimate` = estimate, `Standard error` = std.error,
    `Wald statistic` = statistic, `p-value` = p.value) |>
  kable(align = c("c", "c", "c", "c", "c"))
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

import statsmodels.api as sm
import statsmodels.formula.api as smf

simple_negative_binomial_width_alpha_model = smf.negativebinomial(
    formula="satellites ~ width_cm", data=training_data).fit(disp=0)

simple_negative_binomial_width_alpha = (
    simple_negative_binomial_width_alpha_model.params["alpha"])

simple_negative_binomial_width_model = smf.glm(formula="satellites ~ width_cm", data=training_data,
    family=sm.families.NegativeBinomial(
        alpha=simple_negative_binomial_width_alpha)).fit(scale=1.0)

simple_negative_binomial_width_summary = pd.DataFrame(
    {"Term": simple_negative_binomial_width_model.params.index,
        "Estimate": simple_negative_binomial_width_model.params.values,
        "Standard error": simple_negative_binomial_width_model.bse.values,
        "Wald statistic": simple_negative_binomial_width_model.tvalues.values,
        "p-value": simple_negative_binomial_width_model.pvalues.values,
    })

simple_negative_binomial_width_summary = simple_negative_binomial_width_summary.round(3)

classical_negative_binomial_simple_width_fit_py_html=scrollable_table_html(simple_negative_binomial_width_summary)
```

```{r}
#| echo: false
#| label: tbl-negative-binomial-simple-width-fit-py
#| tbl-cap: "Maximum likelihood estimates from the simple width-only Negative Binomial regression model fitted on the training data."
#| message: false
#| warning: false
#| results: asis

 cat(
   "\n",
   py$classical_negative_binomial_simple_width_fit_py_html,
   "\n",
   sep = ""
 )
```

:::

Having said that, it will be necessary to store the fitted expected counts, because these will be used in @sec-negative-binomial-gof.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-negative-binomial-simple-fitted-values-r
#| tbl-cap: "First 10 fitted expected counts from the simple width-only Negative Binomial regression model via training set."
#| message: false
#| warning: false

simple_negative_binomial_width_fitted <- training_data |>
   mutate(fitted_expected_count = fitted(simple_negative_binomial_width_model)) |>
   dplyr::select(satellites, width_cm, fitted_expected_count) |>
   slice_head(n = 10) |>
   mutate(fitted_expected_count = round(fitted_expected_count, 3))
 
simple_negative_binomial_width_fitted |>
  rename(`Observed satellite count` = satellites, `Female crab width (cm)` = width_cm,
   `Fitted expected count` = fitted_expected_count) |>
   kable(align = c("c", "c", "c"))
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

simple_negative_binomial_width_fitted = (training_data
     .assign(fitted_expected_count=simple_negative_binomial_width_model.fittedvalues)
     [["satellites", "width_cm", "fitted_expected_count",
         ]].head(10).copy())
 
simple_negative_binomial_width_fitted["fitted_expected_count"] = (
     simple_negative_binomial_width_fitted["fitted_expected_count"].round(3))
 
simple_negative_binomial_width_fitted_display = (simple_negative_binomial_width_fitted
     .rename(columns={"satellites": "Observed satellite count",
             "width_cm": "Female crab width (cm)",
             "fitted_expected_count": "Fitted expected count",
         }))
 
classical_negative_binomial_simple_fitted_values_py_html=scrollable_table_html(simple_negative_binomial_width_fitted_display)
```

```{r}
#| echo: false
#| label: tbl-classical-negative_binomial-simple-fitted-values-py
#| tbl-cap: "First 10 fitted expected counts from the simple width-only Negative Binomial regression model via training set."
#| message: false
#| warning: false
#| results: asis

cat(
   "\n",
   py$classical_negative_binomial_simple_fitted_values_py_html,
   "\n",
   sep = ""
 )
```

:::

The above fitted expected counts are not rounded counts and **should not be forced to be integers**. They are estimates of $\mu_i$, the expected number of satellite males for crabs with the corresponding width values. The observed response remains a count, but the fitted conditional mean is allowed to be fractional. This is the quantity we will compare against the observed counts when checking model adequacy.

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on decimal expected counts!
::::
:::: {.Heads-up-container}
A Negative Binomial random variable can only take non-negative integer values:

$$
0, 1, 2, 3, \ldots
$$

However, the mean $\mu_i$ does **not** need to be an integer. It can be any non-negative real number. For example, a fitted expected count such as

$$
\hat{\mu}_i = 2.73
$$

does not mean that the model predicts exactly $2.73$ satellite males for that crab. Instead, it means that, under the fitted model, the **expected number of satellite males** for crabs with those regressor values is $2.73$.

The count outcome itself remains discrete, but the fitted expectation can have decimal places. What cannot happen is a negative fitted expected count. This is one reason the log link is useful in Classical  regression: it keeps

$$
\hat{\mu}_i
=
\exp(\mathbf{x}_i^\top \hat{\boldsymbol{\beta}}_{\operatorname{MLE,obs}})
$$

strictly positive.
::::
:::


## Checking Goodness of Fit {#sec-negative-binomial-gof}


Once we have fitted the **simple width-only Negative Binomial regression model**, we move into goodness of fit. This stage asks whether the fitted model behaves in a way that is compatible with the modelling story we wrote down. For the simple Negative Binomial regression model, that story has three main pieces:

1. the response is **conditionally Negative Binomial**;
2. the **log of the expected count** is described by female crab width;
3. the conditional mean and conditional variance are **equal** after conditioning on width.

All checks in this section still use the **training data**. We are not yet measuring prediction accuracy on the test set. The goal here is diagnostic: we are stress-testing the model we just fitted before using it for final inferential or predictive conclusions. This is not data leakage. Data leakage would occur if we used the testing set to make modelling decisions and then treated it as an untouched assessment set. Here, the training set is where model development, model fitting, and goodness-of-fit checking belong.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-pixel-feedback-to-learn-3699331/).](img/thumbs-up-down.png){width="500"}

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on training-set diagnostics versus test-set prediction!
::::
:::: {.Heads-up-container}
The goodness-of-fit checks in this section are based on the **training data** because the model was fitted on the training data. We use these checks to decide whether the fitted modelling assumptions look plausible enough to continue.

This is different from predictive assessment. Predictive assessment asks how well the model performs on held-out observations that were not used to build or check the model. That comes later, in the **results stage**. For now, we are still inside the model-building loop of the data science workflow: if the fitted model shows serious diagnostic problems, we may need to return to the modelling stage rather than move directly to final results (see @fig-ds-workflow-goodness-of-fit).
::::
:::

For the simple width-only model, we will look at **four diagnostic questions**:

1. Do the **fitted expected counts** track the **observed counts** in a reasonable way?
2. Is there evidence that the counts are **more variable** than the Classical  model allows?
3. Does the model understate the number of **zero counts**?
4. Is the residual deviance unusually large relative to its degrees of freedom?

These checks are complementary. A single diagnostic rarely decides everything by itself. Instead, we combine plots, summary tables, formal tests, and the purpose of the analysis to decide whether the simple model is adequate enough to move forward.

### Observed versus Fitted Counts {#sec-classical--observed-fitted}

The first check compares the observed satellite counts with the fitted expected counts from the simple width-only model. Note that the fitted expected count for the $i$th female crab is

$$
\hat{\mu}_i
=
\exp(\hat{\beta}_0 + \hat{\beta}_1 x_{i,1}),
$$

where $x_{i,1}$ is female crab width in centimetres.

::: {#Definition-fitted-expected-count .definition}
:::: {.definition-header}
Definition of fitted expected count
::::
:::: {.definition-container}
In a fitted Negative Binomial regression model with $k$ regressors, the **fitted expected count** for the $i$th observation is

$$
\hat{\mu}_i
=
\exp(\mathbf{x}_i^\top \hat{\boldsymbol{\beta}}_{\operatorname{MLE,obs}}).
$$

Here, $\hat{\mu}_i$ is the fitted expected number of events for the $i$th observation, $\mathbf{x}_i$ is the regressor vector

$$
\mathbf{x}_i^\top
=
\begin{bmatrix}
1 &
x_{i,1} &
x_{i,2} &
\cdots &
x_{i,k}
\end{bmatrix},
$$

and $\hat{\boldsymbol{\beta}}_{\operatorname{MLE,obs}}$ is the vector of maximum likelihood estimates

$$
\hat{\boldsymbol{\beta}}_{\operatorname{MLE,obs}}^\top
=
\begin{bmatrix}
\hat{\beta}_{0,\operatorname{MLE,obs}} &
\hat{\beta}_{1,\operatorname{MLE,obs}} &
\hat{\beta}_{2,\operatorname{MLE,obs}} &
\cdots &
\hat{\beta}_{k,\operatorname{MLE,obs}}
\end{bmatrix}.
$$

Equivalently, in scalar notation,

$$
\hat{\mu}_i
=
\exp\left(
\hat{\beta}_{0,\operatorname{MLE,obs}}
+
\hat{\beta}_{1,\operatorname{MLE,obs}}x_{i,1}
+
\hat{\beta}_{2,\operatorname{MLE,obs}}x_{i,2}
+
\cdots
+
\hat{\beta}_{k,\operatorname{MLE,obs}}x_{i,k}
\right).
$$

The fitted expected count estimates the conditional expected value $\mathbb{E}(Y_i \mid \mathbf{x}_i)$ under the fitted model, where $Y_i$ is the count response for observation $i$. The observed response $y_i$ is a count, but the fitted expected count $\hat{\mu}_i$ does not need to be an integer.
::::
:::

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-pixel-action-bound-3704046/).](img/magnifying-glass-2.png){width="420"}

A useful way to start is to compare the observed counts and fitted expected counts across groups of similar fitted values. This is a descriptive calibration check on the training data.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-negative-binomial-fitted-count-calibration-r
#| tbl-cap: "Observed and fitted satellite counts across fitted expected count quartiles in the training data."
#| message: false
#| warning: false
poisson_mean_fitted_expected_count <- c(1.499, 2.215, 2.917, 4.859)
poisson_mean_fitted_zero_probability <- c(0.230, 0.111, 0.056, 0.015)

simple_negative_binomial_theta <- simple_negative_binomial_width_model$theta
simple_negative_binomial_gof_data <- training_data |>
  mutate(fitted_expected_count = fitted(simple_negative_binomial_width_model),
    pearson_residual = (satellites - fitted_expected_count) /
      sqrt(fitted_expected_count + fitted_expected_count^2 / simple_negative_binomial_theta),
    fitted_zero_probability = dnbinom(0, size = simple_negative_binomial_theta,
      mu = fitted_expected_count))
fitted_count_calibration <- simple_negative_binomial_gof_data |>
  mutate(fitted_count_quartile = ntile(fitted_expected_count, 4),
    fitted_count_quartile = factor(fitted_count_quartile, levels = 1:4,
      labels = c("Q1: smallest fitted counts", "Q2", "Q3", "Q4: largest fitted counts"))) |>
  group_by(fitted_count_quartile) |>
  summarise(n = n(), mean_observed_count = mean(satellites),
    mean_fitted_expected_count = mean(fitted_expected_count),
    observed_zero_proportion = mean(satellites == 0),
    mean_fitted_zero_probability = mean(fitted_zero_probability), .groups = "drop") |>
  mutate(mean_fitted_expected_count_poisson = poisson_mean_fitted_expected_count,
    mean_fitted_zero_probability_poisson = poisson_mean_fitted_zero_probability) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))
fitted_count_calibration |>
  dplyr::select(fitted_count_quartile, n, mean_observed_count,
    mean_fitted_expected_count, mean_fitted_expected_count_poisson,
    observed_zero_proportion, mean_fitted_zero_probability,
    mean_fitted_zero_probability_poisson) |>
  rename(`Fitted-count quartile` = fitted_count_quartile,
    `Number of female crabs` = n,
    `Mean observed count` = mean_observed_count,
    `Mean fitted expected count (Negative Binomial)` = mean_fitted_expected_count,
    `Mean fitted expected count (Poisson)` = mean_fitted_expected_count_poisson,
    `Observed zero proportion` = observed_zero_proportion,
    `Mean fitted zero probability (Negative Binomial)` = mean_fitted_zero_probability,
    `Mean fitted zero probability (Poisson)` = mean_fitted_zero_probability_poisson) |>
  kable(align = c("c", "c", "c", "c", "c", "c", "c", "c"))
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

poisson_mean_fitted_expected_count = {"Q1: smallest fitted counts": 1.499, "Q2": 2.215,
    "Q3": 2.917, "Q4: largest fitted counts": 4.859}
poisson_mean_fitted_zero_probability = {"Q1: smallest fitted counts": 0.230, "Q2": 0.111,
    "Q3": 0.056, "Q4: largest fitted counts": 0.015}
simple_negative_binomial_theta = 1.0 / simple_negative_binomial_width_model.family.alpha
simple_negative_binomial_gof_data = training_data.copy()
simple_negative_binomial_gof_data["fitted_expected_count"] = (simple_negative_binomial_width_model.fittedvalues)
simple_negative_binomial_gof_data["pearson_residual"] = ((simple_negative_binomial_gof_data["satellites"]
        - simple_negative_binomial_gof_data["fitted_expected_count"])
    / np.sqrt(simple_negative_binomial_gof_data["fitted_expected_count"]
        + simple_negative_binomial_gof_data["fitted_expected_count"] ** 2
        / simple_negative_binomial_theta))
simple_negative_binomial_gof_data["fitted_zero_probability"] = (simple_negative_binomial_theta
    / (simple_negative_binomial_theta
        + simple_negative_binomial_gof_data["fitted_expected_count"])) ** simple_negative_binomial_theta
simple_negative_binomial_gof_data = (simple_negative_binomial_gof_data
    .sort_values("fitted_expected_count", kind="mergesort").reset_index(drop=True))
number_of_rows = len(simple_negative_binomial_gof_data)
number_of_groups = 4
base_group_size = number_of_rows // number_of_groups
remainder = number_of_rows % number_of_groups
group_sizes = [base_group_size + 1 if group_index < remainder else base_group_size
    for group_index in range(number_of_groups)]
quartile_labels = ["Q1: smallest fitted counts", "Q2", "Q3", "Q4: largest fitted counts",
]
simple_negative_binomial_gof_data["fitted_count_quartile"] = np.repeat(quartile_labels, group_sizes)
fitted_count_calibration = (simple_negative_binomial_gof_data
    .groupby("fitted_count_quartile", observed=False)
    .agg(n=("satellites", "size"), mean_observed_count=("satellites", "mean"),
        mean_fitted_expected_count=("fitted_expected_count", "mean"),
        observed_zero_proportion=("satellites", lambda x: (x == 0).mean()),
        mean_fitted_zero_probability=("fitted_zero_probability", "mean"),
    ).reset_index())
fitted_count_calibration["mean_fitted_expected_count_poisson"] = (
    fitted_count_calibration["fitted_count_quartile"].map(poisson_mean_fitted_expected_count))
fitted_count_calibration["mean_fitted_zero_probability_poisson"] = (
    fitted_count_calibration["fitted_count_quartile"].map(poisson_mean_fitted_zero_probability))
fitted_count_calibration = fitted_count_calibration[["fitted_count_quartile", "n",
    "mean_observed_count", "mean_fitted_expected_count", "mean_fitted_expected_count_poisson",
    "observed_zero_proportion", "mean_fitted_zero_probability",
    "mean_fitted_zero_probability_poisson",
    ]].round(3)
fitted_count_calibration_display = fitted_count_calibration.rename(
    columns={"fitted_count_quartile": "Fitted-count quartile",
        "n": "Number of female crabs", "mean_observed_count": "Mean observed count",
        "mean_fitted_expected_count": "Mean fitted expected count (Negative Binomial)",
        "mean_fitted_expected_count_poisson": "Mean fitted expected count (Poisson)",
        "observed_zero_proportion": "Observed zero proportion",
        "mean_fitted_zero_probability": "Mean fitted zero probability (Negative Binomial)",
        "mean_fitted_zero_probability_poisson": "Mean fitted zero probability (Poisson)",
    })
classical_negative_binomial_fitted_count_calibration_py_html = (
    scrollable_table_html(fitted_count_calibration_display))
```

```{r}
#| echo: false
#| label: tbl-negative-binomial-fitted-count-calibration-py
#| tbl-cap: "Observed and fitted satellite counts across fitted expected count quartiles in the training data."
#| message: false
#| warning: false
#| results: asis

 cat(
   "\n",
   py$classical_negative_binomial_fitted_count_calibration_py_html,
   "\n",
   sep = ""
 )
```

:::

The grouped comparison in @tbl-negative-binomial-fitted-count-calibration-r shows that the simple width-only Negative Binomial model captures the broad increase in mean satellite counts across fitted-count quartiles, but it does not reproduce all parts of the response distribution equally well. The mean fitted expected count is close to the mean observed count in Q1, where the mean observed count is `r fitted_count_calibration$mean_observed_count[fitted_count_calibration$fitted_count_quartile == "Q1: smallest fitted counts"]` and the mean fitted expected count is `r fitted_count_calibration$mean_fitted_expected_count[fitted_count_calibration$fitted_count_quartile == "Q1: smallest fitted counts"]`. The mean fitted expected count is very close to the mean observed count in Q2, where the mean observed count is `r fitted_count_calibration$mean_observed_count[fitted_count_calibration$fitted_count_quartile == "Q2: smallest fitted counts"]` and the mean fitted expected count is `r fitted_count_calibration$mean_fitted_expected_count[fitted_count_calibration$fitted_count_quartile == "Q2: smallest fitted counts"]`.The mean fitted expected count is lower than the mean observed count in Q3, where the mean observed count is `r fitted_count_calibration$mean_observed_count[fitted_count_calibration$fitted_count_quartile == "Q3"]` and the mean fitted expected count is `r fitted_count_calibration$mean_fitted_expected_count[fitted_count_calibration$fitted_count_quartile == "Q3"]`. The mean fitted expected count is too high in Q4, where the mean observed count is `r fitted_count_calibration$mean_observed_count[fitted_count_calibration$fitted_count_quartile == "Q4: largest fitted counts"]`, while the mean fitted expected count is `r fitted_count_calibration$mean_fitted_expected_count[fitted_count_calibration$fitted_count_quartile == "Q4: largest fitted counts"]`. **Thus, the fitted mean pattern is directionally useful but not perfectly calibrated.**

Now let's observe the zero-count columns in @tbl-negative-binomial-fitted-count-calibration-r. In Q1, the observed zero proportion is `r fitted_count_calibration$observed_zero_proportion[fitted_count_calibration$fitted_count_quartile == "Q1: smallest fitted counts"]`, while the mean fitted zero probability is `r fitted_count_calibration$mean_fitted_zero_probability[fitted_count_calibration$fitted_count_quartile == "Q1: smallest fitted counts"]`. In Q2, the observed zero proportion is `r fitted_count_calibration$observed_zero_proportion[fitted_count_calibration$fitted_count_quartile == "Q2"]`, while the mean fitted zero probability is `r fitted_count_calibration$mean_fitted_zero_probability[fitted_count_calibration$fitted_count_quartile == "Q2"]`. This shows that the number of zero satellite counts is underestimated in the first two quartiles. On the other hand, the observed zero proportion in Q3 is `r fitted_count_calibration$observed_zero_proportion[fitted_count_calibration$fitted_count_quartile == "Q3"]`, which is lower than its mean fitted zero probability of `r fitted_count_calibration$mean_fitted_zero_probability[fitted_count_calibration$fitted_count_quartile == "Q4"]`. In Q4, the observed zero proportion is `r fitted_count_calibration$observed_zero_proportion[fitted_count_calibration$fitted_count_quartile == "Q3"]` and the mean fitted zero probability is `r fitted_count_calibration$mean_fitted_zero_probability[fitted_count_calibration$fitted_count_quartile == "Q4"]`. This potentially suggests that the simple Negative Binomial model may not perfectly capture the distribution of zero satellite counts for different regressor values in the training data. In comparison, the width-only Classical Poisson model from [Section 10.7](https://alexrod61.github.io/regression-cookbook/book/10-classical-poisson.html#sec-classical-poisson-simple-model) substantially underestimated the number of zero satellite counts in the training data.

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on Pearson residuals and mean fitted zero probabilities!
::::
:::: {.Heads-up-container}
The goodness-of-fit dataset created above includes two quantities that will be used in different diagnostic checks.

The first quantity is the **Pearson residual**,

$$
r_i^{(P)}
=
\frac{y_i - \hat{\mu}_i}{\sqrt{\hat{\mu}_i+\frac{\hat{\mu}_i^2}{\hat{\theta}}}}.
$${#eq-pearson-residual}

Here, $y_i$ is the observed satellite count for the $i$th female crab, and $\hat{\mu}_i$ is the fitted expected count from the simple width-only Negative Binomial regression model. The denominator $\sqrt{\hat{\mu}_i+\frac{\hat{\mu}_i^2}{\theta}}$ comes from the Negative Binomial mean-variance assumption:

$$
\operatorname{Var}(Y_i \mid \mathbf{x}_i)=
\mu_i+\frac{\mu_i^2}{\theta}
$$

After fitting the model, we replace $\mu_i$ with $\hat{\mu}_i$ and $\theta$ with $\hat{\theta}$, so the fitted Negative Binomial standard deviation is $\sqrt{\hat{\mu}_i+\frac{\hat{\mu}_i^2}{\hat{\theta}}}$. Pearson residuals therefore measure how far the observed count is from the fitted expected count **in units of the fitted Negative Binomial standard deviation**.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-emotion-fear-expression-6230192/).](img/curious.png){width="250"}

The second quantity is the **fitted zero probability**,

$$
\Pr(Y_i = 0 \mid \mathbf{x}_i; \hat{\boldsymbol{\beta}}_{\mathrm{MLE,obs}})
=
\left(\frac{\hat{\theta}}{\hat{\theta}+\hat{\mu}_i}\right)^{\hat{\theta}}.
$$

This comes directly from the Negative Binomial PMF. If $Y_i \mid \mathbf{x}_i \sim \operatorname{Negative Binomial}(\hat{\mu}_i)$, then the fitted probability of observing zero satellite males for the $i$th female crab is $\exp(-\hat{\mu}_i)$.

In the grouped table, the column called **mean fitted zero probability** averages these fitted zero probabilities within each fitted-count quartile. It is not the probability of a "*mean zero*." Rather, it is the average of the model-implied zero probabilities for the female crabs in that group:

$$
\frac{1}{n_g}
\sum_{i \in g}
\left(\frac{\hat{\theta}}{\hat{\theta}+\hat{\mu}_i}\right)^{\hat{\theta}},
$$

where $g$ denotes one fitted-count quartile and $n_g$ is the number of observations in that quartile.

This average can be compared with the **observed zero proportion** in the same group,

$$
\frac{1}{n_g}
\sum_{i \in g}
\mathbb{1}(y_i = 0).
$$

If the observed zero proportion is much larger than the mean fitted zero probability, this suggests that the fitted  model is understating how often zero satellite counts occur.
::::
:::

Now, the same comparison can be viewed graphically in @fig-negative-binomial-observed-fitted-r (or @fig-negative-binomial-observed-fitted-py). The dashed line is a visual reference where the observed count would equal the fitted expected count. Points above the line correspond to female crabs whose observed satellite count is larger than the model's fitted expected count. Points below the line correspond to female crabs whose observed satellite count is smaller than the fitted expected count.

::: {.panel-tabset}

## **`R`**

```{r}
#| label: fig-negative-binomial-observed-fitted-r
#| fig-cap: "Observed satellite counts versus fitted expected counts for the simple width-only  regression model."
#| warning: false
#| message: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

observed_fitted_plot <- ggplot(simple_negative_binomial_gof_data,
  aes(x = fitted_expected_count, y = satellites)) +
  geom_point(alpha = 0.65, size = 2.3, colour = "#0072B2") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", linewidth = 1,
    colour = "#D55E00") +
  theme_bw() +
  theme(axis.text = element_text(size = 15.5), axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20, vjust = 0.5, margin = margin(r = 12)),
    panel.grid.minor = element_blank()) +
  labs(x = "\n Fitted expected count", y = "Observed satellite count") +
  scale_y_continuous(breaks = seq(0, 12, by = 2))

observed_fitted_plot
```

## **`Python`**

```{python}
#| label: fig-negative-binomial-observed-fitted-py
#| fig-cap: "Observed satellite counts versus fitted expected counts for the simple width-only  regression model."
#| warning: false
#| message: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

# observed_fitted_figure, observed_fitted_axis = plt.subplots(figsize=(14, 8))
# 
# observed_fitted_axis.scatter(simple_negative_binomial_gof_data["fitted_expected_count"],
#     simple_negative_binomial_gof_data["satellites"], alpha=0.65, s=35, color="#0072B2")
# 
# line_max = max(simple_negative_binomial_gof_data["fitted_expected_count"].max(),
#     simple_negative_binomial_gof_data["satellites"].max())
# 
# observed_fitted_axis.plot([0, line_max], [0, line_max], linestyle="--", linewidth=1,
#     color="#D55E00")
# 
# observed_fitted_axis.set_xlabel("\n Fitted expected count", fontsize=20)
# 
# observed_fitted_axis.set_ylabel("Observed satellite count", fontsize=20, labelpad=12)
# 
# observed_fitted_axis.tick_params(axis="both", labelsize=15.5)
# observed_fitted_axis.set_yticks(range(0, 13, 2))
# 
# observed_fitted_axis.grid(True, which="major", axis="both", alpha=0.3)
# 
# observed_fitted_axis.grid(False, which="minor")
# 
# observed_fitted_figure.tight_layout()
# plt.show()
```

:::

In @fig-negative-binomial-observed-fitted-r or @fig-negative-binomial-observed-fitted-py, the simple width-only model shows some signs of strain. For fitted expected counts between roughly $1.5$ and $4$, there are many observed zeros sitting well below the dashed line, which is consistent with @tbl-negative-binomial-fitted-count-calibration-r showing that the model understates the number of zero satellite counts. At the same time, several observations in this same fitted-count range have much larger observed counts, including values around $6$, $8$, $10$, and higher. This produces a large vertical spread among crabs with similar fitted expected counts. The horizontal bands in the plot are expected because the observed response is a count. The fitted expected count on the $x$-axis is allowed to be fractional, but the observed satellite count on the $y$-axis can only take values such as $0$, $1$, $2$, $3$, and so on. The amount of vertical spread around similar fitted expected counts suggests that the simple width-only Negative Binomial model suggests possible overdispersion in the training data, due to which this Negative Binomial model might be a better fit than the width-only Classical  Model. 

### Dispersion Test

Let's revisit the `extended_model` which was fit in [Chapter 10](#sec-classical-). This model showed substantial overdispersion relative to the Classical  mean-variance assumption with a Pearson dispersion estimate of $3.082$ and a $p-$value < 0.001. We will test whether there is overdispersion in this  regression model via function `dispersiontest()`, (from package `{AER}`). As the `statsmodels` does not have an equivalent function, the `Python` code defines the same regression-based test as a function with the same name.
-   Let $Y_i$ be the $i$th  response in the count regression model.

. . .

-   [Ideally in the presence of equidispersion]{style="color: purple;"}, $Y_i$ has the following parameters: \begin{gather*}
    \mathbb{E}(Y_i) = \lambda_i \\
    \text{Var}(Y_i) = \lambda_i.
    \end{gather*}

-   The test uses the following mathematical expression: $$
    \text{Var}(Y_i) = \overbrace{(1 + \gamma)}^\text{Dispersion Factor} \lambda_i,
    $$

. . .

-   Moreover, we have the following hypotheses: \begin{gather*}
    H_0: 1 + \gamma = 1 \\
    H_a: 1 + \gamma > 1.
    \end{gather*}
-   When there is evidence of overdispersion in our data, [we will reject $H_0$]{style="color: purple;"} under a pre-specified significance level $\alpha$.

::: {.panel-tabset}

## **`R` Code**

```{.r}
simple_poisson_width_model <- glm(formula = satellites ~ width_cm,
  family = stats::poisson(link = "log"), data = training_data)

dispersiontest(simple_poisson_width_model)
```

## **`Python` Code**

``` {.python}
import numpy as np
import scipy.stats as stats


def dispersiontest(results, alternative="greater"):
    """Cameron and Trivedi (1990) regression-based dispersion test.

    This is the `Python` counterpart of `AER::dispersiontest()` with
    `trafo = NULL`, i.e., the alternative is Var(Y_i) = (1 + gamma) * lambda_i.
    """
    y = np.asarray(results.model.endog, dtype=float)
    y_hat = np.asarray(results.fittedvalues, dtype=float)

    # Auxiliary quantity behind the test
    aux = ((y - y_hat) ** 2 - y) / y_hat

    test_statistic = np.sqrt(aux.size) * aux.mean() / aux.std(ddof=1)
    dispersion = aux.mean() + 1

    if alternative == "greater":
        p_value = stats.norm.sf(test_statistic)
    elif alternative == "less":
        p_value = stats.norm.cdf(test_statistic)
    elif alternative == "two-sided":
        p_value = 2 * stats.norm.sf(abs(test_statistic))
    else:
        raise ValueError(
            "alternative must be 'greater', 'less', or 'two-sided'."
        )

    return test_statistic, p_value, dispersion


# Fit the width-only Poisson model
simple_poisson_width_model = smf.glm(formula="satellites ~ width_cm", data=training_data,
    family=sm.families.Poisson()).fit()

# Dispersion test
test_statistic, p_value, dispersion = dispersiontest(
    simple_poisson_width_model
)

print(f"Test statistic (z): {test_statistic:.4f}")
print(f"P-value: {p_value:.4g}")
print(f"Sample estimate (dispersion): {dispersion:.4f}")
```

:::

::: {.panel-tabset}

## **`R` Output**

```{r}
#| echo: false
#| message: false
#| warning: false
extended__model <- glm(
  satellites ~ width_cm + weight_kg + color + spine,
  family = poisson(link = "log"),
  data = training_data
)

dispersiontest(extended__model)
```

## **`Python` Output**

``` {python}
#| echo: false
#| message: false
#| warning: false

def dispersiontest(results, alternative="greater"):
    """Cameron and Trivedi (1990) regression-based dispersion test.

    This is the Python counterpart of AER::dispersiontest() with
    trafo = NULL, i.e., the alternative is
    Var(Y_i) = (1 + gamma) * lambda_i.
    """
    y = np.asarray(results.model.endog, dtype=float)
    y_hat = np.asarray(results.fittedvalues, dtype=float)

    # Auxiliary quantity behind the test
    aux = ((y - y_hat) ** 2 - y) / y_hat

    test_statistic = np.sqrt(aux.size) * aux.mean() / aux.std(ddof=1)
    dispersion = aux.mean() + 1

    if alternative == "greater":
        p_value = stats.norm.sf(test_statistic)
    elif alternative == "less":
        p_value = stats.norm.cdf(test_statistic)
    elif alternative == "two-sided":
        p_value = 2 * stats.norm.sf(abs(test_statistic))
    else:
        raise ValueError(
            "alternative must be 'greater', 'less', or 'two-sided'."
        )

    return test_statistic, p_value, dispersion


# Fit model
extended_model = smf.glm(
    formula=(
        "satellites ~ width_cm + weight_kg "
        "+ C(color, Treatment(reference='Light')) "
        "+ C(spine, Treatment(reference='Both good'))"
    ),
    data=training_data,
    family=sm.families.Poisson()
).fit()


# Dispersion test
test_statistic, p_value, dispersion = dispersiontest(
    extended_model
)

print(f"Test statistic (z): {test_statistic:.4f}")
print(f"P-value: {p_value:.4g}")
print(f"Sample estimate (dispersion): {dispersion:.4f}")
```
:::

With $\alpha = 0.05$, we reject $H_0$ since the $p\text{-value} < .001$. Hence, the `extended__model` has overdispersion.

::: {.panel-tabset}

## **`R` Code**

```{.r}
negative_binomial_model <- glm.nb(satellites ~ width_cm,
  data = training_data
)
summary(negative_binomial_model)

extended_negative_binomial_model <- glm.nb(satellites ~ width_cm + weight_kg + color + spine,
  data = training_data
)
summary(extended_negative_binomial_model)
```

## **`Python` Code**

``` {.python}
negative_binomial_training_data = training_data.copy()

negative_binomial_model = smf.negativebinomial(
    formula="satellites ~ width_cm",
    data=negative_binomial_training_data
).fit()

negative_binomial_summary = pd.DataFrame(
    {
        "Term": negative_binomial_model.params.index,
        "Estimate": negative_binomial_model.params.values,
        "Standard error": negative_binomial_model.bse.values,
        "Exponentiated estimate": np.exp(
            negative_binomial_model.params.values
        ),
    }
)

extended_negative_binomial_model = smf.negativebinomial(
    formula="satellites ~ width_cm + weight_kg + color + spine",
    data=negative_binomial_training_data
).fit()

extended_negative_binomial_summary = pd.DataFrame(
    {
        "Term": extended_negative_binomial_model.params.index,
        "Estimate": extended_negative_binomial_model.params.values,
        "Standard error": extended_negative_binomial_model.bse.values,
        "Exponentiated estimate": np.exp(
            extended_negative_binomial_model.params.values
        ),
    }
)

negative_binomial_summary
extended_negative_binomial_summary
```

:::

::: {.panel-tabset}

## **`R` Output*

```{r}
#| echo: false
#| message: false
#| warning: false
negative_binomial_model <- glm.nb(satellites ~ width_cm,
  data = training_data
)
summary(negative_binomial_model)

extended_negative_binomial_model <- glm.nb(satellites ~ width_cm + weight_kg + color + spine,
  data = training_data
)
summary(extended_negative_binomial_model)
```

## **`Python` Output**

``` {python}
#| echo: false
#| message: false
#| warning: false
negative_binomial_training_data = training_data.copy()

negative_binomial_model = smf.negativebinomial(
    formula="satellites ~ width_cm",
    data=negative_binomial_training_data
).fit()

negative_binomial_summary = pd.DataFrame(
    {
        "Term": negative_binomial_model.params.index,
        "Estimate": negative_binomial_model.params.values,
        "Standard error": negative_binomial_model.bse.values,
        "Exponentiated estimate": np.exp(
            negative_binomial_model.params.values
        ),
    }
)

extended_negative_binomial_model = smf.negativebinomial(
    formula="satellites ~ width_cm + weight_kg + color + spine",
    data=negative_binomial_training_data
).fit()

extended_negative_binomial_summary = pd.DataFrame(
    {
        "Term": extended_negative_binomial_model.params.index,
        "Estimate": extended_negative_binomial_model.params.values,
        "Standard error": extended_negative_binomial_model.bse.values,
        "Exponentiated estimate": np.exp(
            extended_negative_binomial_model.params.values
        ),
    }
)

negative_binomial_summary
extended_negative_binomial_summary
```

:::

### Excess Zero Counts

In @sec-negative-binomial-eda-general, the EDA also showed many zero satellite counts. We now check whether the fitted simple Negative Binomial model expects a similar number of zeros in the **training data**. Hence, this diagnostic focuses on the number of observed zero counts,

$$
z_{\operatorname{obs}}
=
\sum_{i=1}^n
\mathbb{1}(y_i = 0),
$$

where $\mathbb{1}(y_i = 0)$ is an indicator that equals $1$ when the $i$th observed satellite count is zero and equals $0$ otherwise.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/) via [*Pixabay*](https://pixabay.com/vectors/tablet-games-tetris-building-blocks-7426706/).](img/tablet.png){width="300"}

Under the fitted Negative Binomial model, the probability that the $i$th female crab has zero satellite males is

$$
\Pr(Y_i = 0 \mid \mathbf{x}_i; \hat{\boldsymbol{\beta}}_{\operatorname{MLE,obs}})
=
\left(\frac{\hat{\mu_i}}{\hat{\mu_i}+\hat{\theta}}\right)^\hat{\theta}.
$$

Now, let

$$
\hat{\pi}_{0,i}
=
\left(\frac{\hat{\mu_i}}{\hat{\mu_i}+\hat{\theta}}\right)^\hat{\theta}
$$

denote this fitted zero probability for observation $i$. Then, the fitted expected number of zeros in the training data is

$$
\widehat{\mathbb{E}}(Z)
=
\sum_{i=1}^n
\hat{\pi}_{0,i}.
$$

The question is whether the observed number of zeros is unusually large compared with what the fitted Negative Binomial model expects. We can frame this as a one-sided diagnostic test:

$$
\begin{gather}
H_0\text{: } \text{the fitted Classical  model adequately describes} \\
\text{the frequency of zero counts}, \\
\text{versus} \\
H_1\text{: } \text{there are more zero counts than the} \\
\text{fitted Classical  model expects}.
\end{gather}
$$

Under $H_0$, the zero-count indicators $\mathbb{1}(Y_i = 0)$ are treated as independent Bernoulli random variables with fitted probabilities $\hat{\pi}_{0,i}$. Because these probabilities can differ across observations, the total number of zeros is a sum of Bernoulli variables with unequal probabilities. Its fitted mean is

$$
\widehat{\mathbb{E}}(Z)
=
\sum_{i=1}^n
\hat{\pi}_{0,i},
$$

and its fitted variance is

$$
\widehat{\operatorname{Var}}(Z)
=
\sum_{i=1}^n
\hat{\pi}_{0,i}
(1 - \hat{\pi}_{0,i}).
$$

This gives the approximate $z$-statistic

$$
z_{\operatorname{zero}}
=
\frac{
z_{\operatorname{obs}} - \widehat{\mathbb{E}}(Z)
}{
\sqrt{
\widehat{\operatorname{Var}}(Z)
}
}.
$${#eq-z-statistic-zero-counts}

Large positive values of $z_{\operatorname{zero}}$ indicate that the training data contain more zero counts than the fitted model expects. The corresponding one-sided $p$-value is

$$
\begin{gather}
Z_{\operatorname{zero}} \sim \operatorname{Normal}\left(0,1\right) \\
p\text{-value}
=
\Pr
\left(
Z_{\operatorname{zero}}
\geq
z_{\operatorname{zero}}
\right),
\end{gather}
$$

where $\operatorname{Normal}(0,1)$ denotes a standard Normal random variable.

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on what this zero-count test is checking!
::::

:::: {.Heads-up-container}
This diagnostic is not testing whether the response variable can take the value zero. A Negative Binomial model always allows zero counts. Instead, it asks whether the **number of zeros observed in the training data** is unusually large compared with the number of zeros expected under the fitted Classical  model.

The fitted model can assign a different zero probability to each female crab because each crab has its own fitted expected count $\hat{\mu}_i$. For the $i$th crab, that fitted zero probability is

$$
\hat{\pi}_{0,i}
=
\left(\frac{\hat{\mu_i}}{\hat{\mu_i}+\hat{\theta}}\right)^\hat{\theta}.
$$

The fitted expected number of zeros is obtained by adding these fitted zero probabilities across the training observations. If the observed zero count is much larger than this fitted expectation, then the model may be missing a feature of the data-generating process. This can happen, for example, when some female crabs have zero satellite males for reasons not captured by width alone.

This check is a diagnostic, not a final model-selection rule. A large excess of zeros would motivate considering zero-inflated count models (see @sec-zero-inflated-), but it should be interpreted alongside the overdispersion check, residual deviance, residual plots, and the scientific context.
::::
:::

The following table computes the observed zero count, the fitted expected zero count, the approximate $z$-statistic, and the one-sided $p$-value.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-negative-binomial-zero-count-check-r
#| tbl-cap: "Observed and fitted zero counts for the simple width-only  regression model."
#| message: false
#| warning: false

observed_zero_count <- sum(simple_negative_binomial_gof_data$satellites == 0)

expected_zero_count <- sum(simple_negative_binomial_gof_data$fitted_zero_probability)

zero_count_variance <- sum(simple_negative_binomial_gof_data$fitted_zero_probability *
    (1 - simple_negative_binomial_gof_data$fitted_zero_probability))

zero_count_z <- (observed_zero_count - expected_zero_count) / sqrt(zero_count_variance)

zero_count_p_value <- pnorm(zero_count_z, lower.tail = FALSE)

zero_count_summary <- tibble::tibble(`Observed zero count` = observed_zero_count,
  `Fitted expected zero count` = expected_zero_count,
  `Observed minus fitted zeros` = observed_zero_count - expected_zero_count,
  `Approximate z-statistic` = zero_count_z, `One-sided p-value` = zero_count_p_value) |>
  mutate(`Observed zero count` = formatC(`Observed zero count`, format = "f", digits = 0),
    `Fitted expected zero count` = formatC(`Fitted expected zero count`, format = "f",
      digits = 2),
    `Observed minus fitted zeros` = formatC(`Observed minus fitted zeros`, format = "f",
      digits = 2),
    `Approximate z-statistic` = formatC(`Approximate z-statistic`, format = "f", digits = 2
    ), `One-sided p-value` = formatC(`One-sided p-value`, format = "e", digits = 2)) |>
  pivot_longer(cols = everything(), names_to = "Quantity", values_to = "Value")

zero_count_summary |>
  kable(align = c("c", "c"))
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

# observed_zero_count = (simple__gof_data["satellites"] == 0).sum()
# 
# expected_zero_count = (simple__gof_data["fitted_zero_probability"].sum())
# 
# zero_count_variance = (simple__gof_data["fitted_zero_probability"]
#     * (1 - simple__gof_data["fitted_zero_probability"])).sum()
# 
# zero_count_z = (observed_zero_count - expected_zero_count) / np.sqrt(zero_count_variance)
# 
# zero_count_p_value = stats.norm.sf(zero_count_z)
# 
# zero_count_summary = pd.DataFrame(
#     {"Quantity": ["Observed zero count", "Fitted expected zero count",
#             "Observed minus fitted zeros", "Approximate z-statistic", "One-sided p-value",
#         ], "Value": [f"{observed_zero_count:.0f}", f"{expected_zero_count:.2f}",
#             f"{observed_zero_count - expected_zero_count:.2f}", f"{zero_count_z:.2f}",
#             f"{zero_count_p_value:.2e}",
#         ],
#     })
# 
# classical__zero_count_check_py_html = (scrollable_table_html(zero_count_summary))
```

```{r}
#| echo: false
#| label: tbl-negative-binomial-zero-count-check-py
#| tbl-cap: "Observed and fitted zero counts for the simple width-only  regression model."
#| message: false
#| warning: false
#| results: asis

# cat(
#   "\n",
#   py$classical__zero_count_check_py_html,
#   "\n",
#   sep = ""
# )
```

:::

In this zero-count check, the $p-$value is $0.0494<0.05$ and therefore there is some evidence that the simple width-only Negative Binomial model is not reproducing the zero-count part of the response distribution. In the **training data**, the observed number of female crabs with zero satellite males is `r observed_zero_count`, while the fitted model expects only `r formatC(expected_zero_count, format = "f", digits = 2)` zeros. The difference is therefore `r formatC(observed_zero_count - expected_zero_count, format = "f", digits = 2)` additional zero counts beyond what the fitted model expects. The $p-$ value is only marginally below the $0.05$ threshold, and in comparison to the simple Classical Poisson model from [Chapter 10](https://alexrod61.github.io/regression-cookbook/book/10-classical-poisson.html), this model is a much better fit for the zero-count part of the response distribution. Recall that the $p-$value for the zero-count check in [Section 10.8.3](https://alexrod61.github.io/regression-cookbook/book/10-classical-poisson.html#sec-classical-poisson-excess-zeros) was $1.62e-20$. 

Furthermore, this zero-count check should still be interpreted as one diagnostic rather than as the only criterion for choosing a model. This situation motivates a closer look at alternatives such as zero-inflated count models.

### Residual Deviance {#sec-negative-binomial-residual-deviance}

The overdispersion and zero-count checks each focused on a specific feature of the fitted model. Note that the dispersion test (in @sec-negative-binomial-dispersion) focused on whether the observed counts vary more than what the Classical Poisson mean-variance structure allows, while the zero-count check (in @sec-negative-binomial-excess-zeros) focused on whether the model expects enough zero satellite counts.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-emotion-proud-6230200/).](img/king.png){width="400"}

Now, let us dig into **residual deviance**. This metric gives **a more global goodness-of-fit check**. Instead of focusing only on variance or zeros, it compares the likelihood of the fitted regression model with the likelihood of an idealized benchmark model that fits the observed responses as closely as possible. That benchmark is called the **saturated model**. In this chapter, we use residual deviance as another training-data diagnostic. It will help us ask whether the simple width-only Negative Binomial model is close enough to the observed training counts, relative to what would be possible under a perfect-fit benchmark.

::: {#Definition-saturated-model .definition}
:::: {.definition-header}
Definition of saturated model
::::

:::: {.definition-container}
A **saturated model** is a model that fits the observed data perfectly by using enough parameters to reproduce every observed response value exactly. The saturated model is usually not useful as a practical regression model. Instead, it serves as a theoretical benchmark. Residual deviance compares the fitted model we actually use with this perfect-fit benchmark.
::::
:::

::: {#Definition-residual-deviance .definition}
:::: {.definition-header}
Definition of residual deviance
::::

:::: {.definition-container}
The **residual deviance** of a fitted GLM measures how far the fitted regression model is from a saturated model, where the saturated model is the idealized benchmark that fits the observed responses perfectly.

The residual deviance is defined as

$$
D
=
2
\left[
\ell_{\operatorname{saturated}}
-
\ell_{\operatorname{fitted}}
\right],
$${#eq-residual-deviance}

where $\ell_{\operatorname{saturated}}$ is the log-likelihood of the saturated model, and $\ell_{\operatorname{fitted}}$ is the log-likelihood of the fitted regression model.

The factor of $2$ is included so that, under suitable regularity conditions, the residual deviance can often be compared to a Chi-squared reference distribution. In this sense, residual deviance plays a role similar to a discrepancy measure: smaller values indicate that the fitted model is closer to the saturated benchmark, while larger values indicate that the fitted model is much less likely than a model that reproduces the observed responses exactly.
::::
:::

For Negative Binomial regression, @eq-residual-deviance can be written as

$$
D
=
2
\sum_{i=1}^n
\left[
y_i \log\left(\frac{y_i}{\hat{\mu}_i}\right)
-
(y_i - \hat{\mu}_i)
\right],
$${#eq-classical-poisson-residual-deviance}

where $n$ is the number of observations used to fit the model, $y_i$ is the observed count for observation $i$, and $\hat{\mu}_i$ is the fitted expected count for observation $i$. When $y_i = 0$, the term $y_i \log(y_i / \hat{\mu}_i)$ is defined to be $0$.

In practical terms, in @eq-negative-binomial-residual-deviance, the residual deviance adds up how much the observed counts differ from the fitted expected counts on the likelihood scale. It is not simply a sum of squared residuals, but it serves a similar diagnostic purpose: **it helps us assess whether the fitted model is close enough to the observed data to be considered adequate**.

If the fitted Negative Binomial model is adequate, then the gap between the fitted model and the saturated model should not be unusually large. In large samples, this gap is often assessed by comparing the residual deviance to a Chi-squared distribution with residual degrees of freedom

$$
\nu = n - (k + 1),
$$

where $n$ is the number of training observations and $k + 1$ is the number of estimated regression terms (intercept and coefficients). Under this reference approximation,

$$
D \mathrel{\dot{\sim}} \chi^2_{\nu}.
$$

This gives the goodness-of-fit hypotheses

$$
\begin{gather}
H_0\text{: } \text{the fitted Negative Binomial model is adequate}, \\
\text{versus} \\
H_1\text{: } \text{the fitted Negative Binomial model is not adequate}.
\end{gather}
$$

The $p$-value is

$$
p\text{-value}
=
\Pr
\left(
\chi^2_{\nu}
\geq
D
\right).
$$

A small $p$-value means that the residual deviance is unusually large under the fitted model, which is evidence of lack of fit.

The code below extracts the residual deviance from the simple fitted model and compares it with its residual degrees of freedom:

- In `R`, `deviance(simple_poisson_width_model)` extracts the residual deviance and `df.residual(simple_poisson_width_model)` extracts the residual degrees of freedom. The upper-tail $p$-value is computed with `pchisq(..., lower.tail = FALSE)`.
- The `Python` code follows the same logic. The fitted [{statsmodels}](https://www.statsmodels.org/stable/index.html) object stores the residual deviance in `simple_poisson_width_model.deviance` and the residual degrees of freedom in `simple_poisson_width_model.df_resid`. The command `stats.chi2.sf(residual_deviance, residual_deviance_df)` computes the same upper-tail Chi-squared probability.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-negative-binomial-residual-deviance-r
#| tbl-cap: "Residual deviance goodness-of-fit check for the simple width-only Negative Binomial regression model."
#| message: false
#| warning: false

# Residual deviance from the fitted Poisson GLM
residual_deviance <- deviance(simple_negative_binomial_width_model)

# Residual degrees of freedom: n minus the number of estimated terms
residual_deviance_df <- df.residual(simple_negative_binomial_width_model)

# Residual deviance divided by residual degrees of freedom
residual_deviance_ratio <- residual_deviance / residual_deviance_df

# Upper-tail p-value under the Chi-squared reference distribution
residual_deviance_p_value <- pchisq(residual_deviance, df = residual_deviance_df,
  lower.tail = FALSE)

residual_deviance_summary <- tibble::tibble(`Residual deviance` = residual_deviance,
  `Residual degrees of freedom` = residual_deviance_df,
  `Residual deviance / df` = residual_deviance_ratio,
  `Goodness-of-fit p-value` = residual_deviance_p_value) |>
  mutate(`Residual deviance` = formatC(`Residual deviance`, format = "f", digits = 3),
    `Residual degrees of freedom` = formatC(`Residual degrees of freedom`, format = "f",
      digits = 0),
    `Residual deviance / df` = formatC(`Residual deviance / df`, format = "f", digits = 3),
    `Goodness-of-fit p-value` = formatC(`Goodness-of-fit p-value`, format = "e", digits = 2)
  ) |>
  pivot_longer(cols = everything(), names_to = "Quantity", values_to = "Value")

residual_deviance_summary |>
  kable(align = c("c", "c"))
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

# Residual deviance from the fitted Poisson GLM
residual_deviance = simple_negative_binomial_width_model.deviance

# Residual degrees of freedom: n minus the number of estimated terms
residual_deviance_df = simple_negative_binomial_width_model.df_resid

# Residual deviance divided by residual degrees of freedom
residual_deviance_ratio = residual_deviance / residual_deviance_df

# Upper-tail p-value under the Chi-squared reference distribution
residual_deviance_p_value = stats.chi2.sf(residual_deviance, residual_deviance_df)

residual_deviance_summary = pd.DataFrame(
    {"Quantity": ["Residual deviance", "Residual degrees of freedom",
            "Residual deviance / df", "Goodness-of-fit p-value",
        ], "Value": [f"{residual_deviance:.3f}", f"{residual_deviance_df:.0f}",
            f"{residual_deviance_ratio:.3f}", f"{residual_deviance_p_value:.2e}",
        ],
    })

classical_negative_binomial_residual_deviance_py_html = (
    scrollable_table_html(residual_deviance_summary))
```

```{r}
#| echo: false
#| label: tbl-classical-negative-binomial-residual-deviance-py
#| tbl-cap: "Residual deviance goodness-of-fit check for the simple width-only Poisson regression model."
#| message: false
#| warning: false
#| results: asis

cat(
  "\n",
  py$classical_negative_binomial_residual_deviance_py_html,
  "\n",
  sep = ""
)
```

:::

For this Negative Binomial model with width as the regressor, the residual deviance is `r formatC(residual_deviance, format = "f", digits = 3)`, while the residual degrees of freedom are `r residual_deviance_df`. The ratio `r formatC(residual_deviance_ratio, format = "f", digits = 3)` is close to $1$, and the corresponding goodness-of-fit $p$-value is `r formatC(residual_deviance_p_value, format = "e", digits = 2)`. As the $p-$value is smaller than the $\alpha=0.05$, there is statistical evidence to state that this Negative Binomial regression model is correctly specified when compared to the saturated model.


Note that the residual deviance gives a single global statistic. Let's also inspect Pearson residuals against fitted expected counts. As indicated in @sec-negative-binomial-residual-deviance, Pearson residuals standardize the difference between the observed count and the fitted expected count by the Poisson standard deviation implied by the model:

$$
r_i^{(P)}
=
\frac{y_i - \hat{\mu}_i}{\sqrt{\hat{\mu}_i+\frac{\hat{\mu}_i^2}{\theta}}}.
$$

::: {.panel-tabset}

## **`R`**

```{r}
#| label: fig-negative-binomial-pearson-residuals-r
#| fig-cap: "Pearson residuals versus fitted expected counts for the simple width-only Negative Binomial regression model. The dashed horizontal lines at -2 and 2 provide informal reference bands for large Pearson residuals."
#| warning: false
#| message: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

pearson_residual_plot <- ggplot(simple_negative_binomial_gof_data,
  aes(x = fitted_expected_count, y = pearson_residual)) +
  geom_hline(yintercept = 0, linetype = "solid", linewidth = 0.8, colour = "grey40") +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", linewidth = 0.8, colour = "#D55E00"
  ) +
  geom_point(alpha = 0.65, size = 2.3, colour = "#0072B2") +
  theme_bw() +
  theme(axis.text = element_text(size = 15.5), axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20, vjust = 0.5, margin = margin(r = 12)),
    panel.grid.minor = element_blank()) +
  labs(x = "\n Fitted expected count", y = "Pearson residual")

pearson_residual_plot
```

## **`Python`**

```{python}
#| label: fig-negative-binomial-pearson-residuals-py
#| fig-cap: "Pearson residuals versus fitted expected counts for the simple width-only Poisson regression model. The dashed horizontal lines at -2 and 2 provide informal reference bands for large Pearson residuals."
#| warning: false
#| message: false
#| echo: true
#| fig.height: 8
#| fig.width: 14

pearson_residual_figure, pearson_residual_axis = plt.subplots(figsize=(14, 8))

pearson_residual_axis.axhline(0, linestyle="-", linewidth=0.8, color="grey")

pearson_residual_axis.axhline(2, linestyle="--", linewidth=0.8, color="#D55E00")

pearson_residual_axis.axhline(-2, linestyle="--", linewidth=0.8, color="#D55E00")

pearson_residual_axis.scatter(simple_negative_binomial_gof_data["fitted_expected_count"],
    simple_negative_binomial_gof_data["pearson_residual"], alpha=0.65, s=35, color="#0072B2")

pearson_residual_axis.set_xlabel("\n Fitted expected count", fontsize=20)

pearson_residual_axis.set_ylabel("Pearson residual", fontsize=20, labelpad=12)

pearson_residual_axis.tick_params(axis="both", labelsize=15.5)

pearson_residual_axis.grid(True, which="major", axis="both", alpha=0.3)

pearson_residual_axis.grid(False, which="minor")

pearson_residual_figure.tight_layout()
plt.show()
```

:::

In @fig-negative-binomial-pearson-residuals-r (or @fig-negative-binomial-pearson-residuals-py), the solid horizontal line at $0$ represents a perfect residual balance: the observed count equals the fitted expected count after standardizing by the fitted Negative Binomial standard deviation. The two dashed horizontal lines are visual reference bands at approximately $-2$ and $2$. They are not formal decision cutoffs, but they help us identify observations whose Pearson residuals are large in magnitude. Under a reasonably adequate Negative Binomial variance structure, we would expect most Pearson residuals to fall inside these bands, without a strong pattern across fitted expected counts. The residual pattern is reassuring because a majority of observations in the training are between $-2$ and $2$.

## Results and Statistical Interpretation {#sec-negative-binomial-results}

# Results and Statistical Interpretation {#sec-classical-poisson-results}

Now, we move from model building to results. Up to this point, the **training data** have been used to explore the response, fit the simple and extended Classical Poisson regression models, and diagnose goodness of fit. The extended model gave us a richer fitted mean structure than the simple width-only model, but the diagnostics still showed substantial overdispersion, excess zero counts, and global lack of fit.

Recall that this chapter has two guiding inquiries as in @tbl-classical-poisson-results-inquiries.

| Inquiry type | Research question | What the Results section must deliver |
|:---:|:---:|:---:|
| **Inferential inquiry** | How is the expected number of satellite males associated with female crab width, weight, colour, and spine condition? | A coefficient-level statistical interpretation of the extended Classical Poisson model, using the testing-set refit. |
| **Predictive inquiry** | How accurately can female crab characteristics predict the number of satellite males for held-out crabs? | A held-out predictive assessment using the training-fitted extended model evaluated on the testing data. |

: The two guiding inquiries revisited for the Results section. {#tbl-classical-poisson-results-inquiries .striped .hover}

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/?utm_source=link-attribution&utm_medium=referral&utm_campaign=image&utm_content=3699345) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-pixel-digital-3704070/).](img/pinned-notes.png){width="500"}

Therefore, the **results stage** has two complementary goals:

- For the **inferential inquiry**, we refit the extended Classical Poisson model on the **testing data** and summarize the coefficient-level results. 
- For the **predictive inquiry**, we evaluate how well the training-fitted extended model predicts the held-out testing counts.

This separation follows the data science workflow used throughout the book. The training data supported model development and goodness-of-fit checks. The testing data are now used for final assessment: coefficient-level inferential summaries for the inferential inquiry, and held-out prediction accuracy for the predictive inquiry.

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on why the testing data appear in two roles here!
::::
:::: {.Heads-up-container}
The testing data are used differently for the two workflow flavours:

- For the **inferential inquiry**, we refit the selected extended model on the testing data. This gives a final coefficient-level analysis on data that were not used for EDA, model development, or goodness-of-fit decisions.
- For the **predictive inquiry**, we do not refit the model before predicting. We use the extended model fitted on the training data to generate predictions for the testing observations, and then compare those predictions with the observed testing counts.

These two uses should not be mixed. Refitting on the testing data is appropriate for the final inferential summary under this chapter's workflow. Predictive evaluation, however, must use predictions from a model trained before looking at the testing responses.
::::
:::

### Inferential Results: Refit the Extended Model on the Testing Data {#sec-classical-poisson-inferential-results}

For the inferential inquiry, we refit the extended Classical Poisson model using the testing data. The model has the same systematic component developed earlier:

$$
Y_i \mid x_{i,1}, x_{i,2}, \ldots, x_{i,7}
\sim
\operatorname{Poisson}(\mu_i),
$$

with

$$
\begin{align*}
\log(\mu_i)
&=
\beta_0
+
\beta_1 x_{i,1}
+
\beta_2 x_{i,2}
+
\beta_3 x_{i,3}
+ \\
& \beta_4 x_{i,4}
+
\beta_5 x_{i,5}
+
\beta_6 x_{i,6}
+
\beta_7 x_{i,7}.
\end{align*}
$${#eq-classical-poisson-results-extended-model}

In @eq-classical-poisson-results-extended-model, $x_{i,1}$ is width, $x_{i,2}$ is weight, $x_{i,3}$, $x_{i,4}$, and $x_{i,5}$ are colour dummy variables, and $x_{i,6}$ and $x_{i,7}$ are spine-condition dummy variables (see @tbl-classical-poisson-dummy-variable-arrangement).

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/) via [*Pixabay*](https://pixabay.com/vectors/daycare-pixel-holding-hands-7464616/).](img/juggling.png){width="300"}

Before interpreting the coefficients on the expected-count scale, we first refit the extended Classical Poisson model on the **testing data**. This refit uses the same model form selected during the training-stage workflow, but it is estimated using observations that were not used for EDA, model building, or goodness-of-fit decisions. @tbl-classical-poisson-test-refit-log-scale-r reports the coefficient estimates on the **log expected-count scale**. This is the scale on which the GLM is fitted and on which the Wald statistics are computed. The coefficient estimate tells us the estimated additive change in $\log(\mu_i)$, the standard error measures uncertainty on that same scale, and the Wald statistic compares the estimate with the null value $0$.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-classical-poisson-test-refit-log-scale-r
#| tbl-cap: "Testing-set refit of the extended Classical Poisson regression model on the log expected-count scale."
#| message: false
#| warning: false

# final_poisson_inference_model <- glm(satellites ~ width_cm + weight_kg + color + spine,
#   family = poisson(link = "log"), data = testing_data)
# 
# final_poisson_inference_summary <- tidy(final_poisson_inference_model) |>
#   mutate(conf_low = estimate - qnorm(0.975) * std.error,
#     conf_high = estimate + qnorm(0.975) * std.error, exponentiated_estimate = exp(estimate),
#     exponentiated_conf_low = exp(conf_low), exponentiated_conf_high = exp(conf_high),
#     percent_change = 100 * (exponentiated_estimate - 1),
#     across(where(is.numeric), ~ round(.x, 2)))
# 
# final_poisson_inference_summary |>
#   select(term, estimate, std.error, statistic, p.value) |>
#   rename(`Term` = term, `Log-scale estimate` = estimate, `Standard error` = std.error,
#     `Wald statistic` = statistic, `p-value` = p.value) |>
#   kable(align = c("c", "c", "c", "c", "c"))
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

# final_negative_binomial_inference_data = testing_data.copy()
# 
# final_negative_binomial_inference_data["color"] = pd.Categorical(
#     final_poisson_inference_data["color"], categories=color_order)
# 
# final_negative_binomial_inference_data["spine"] = pd.Categorical(
#     final_poisson_inference_data["spine"], categories=spine_order)
# 
# final_negative_binomial_inference_model = smf.glm(formula=("satellites ~ width_cm + weight_kg "
#         "+ C(color, Treatment(reference='Light')) "
#         "+ C(spine, Treatment(reference='Both good'))"), data=final_negative_binomial_inference_data,
#     family=sm.families.negative_binomial()).fit()
# 
# final_negative_binomial_inference_summary = pd.DataFrame(
#     {"Term": final_negative_binomial_inference_model.params.index,
#         "Log-scale estimate": final_negative_binomial_inference_model.params.values,
#         "Standard error": final_negative_binomial_inference_model.bse.values,
#         "Wald statistic": final_negative_binomial_inference_model.tvalues.values,
#         "p-value": final_negative_binomial_inference_model.pvalues.values,
#     })
# 
# final_negative_binomial_inference_summary["Lower 95% CI on log scale"] = (
#     final_negative_binomial_inference_summary["Log-scale estimate"] - stats.norm.ppf(0.975)
#     * final_negative_binomial_inference_summary["Standard error"])
# 
# final_negative_binomial_inference_summary["Upper 95% CI on log scale"] = (
#     final_negative_binomial_inference_summary["Log-scale estimate"] + stats.norm.ppf(0.975)
#     * final_negative_binomial_inference_summary["Standard error"])
# 
# final_negative_binomial_inference_summary["Exponentiated estimate"] = np.exp(
#     final_negative_binomial_inference_summary["Log-scale estimate"])
# 
# final_negative_binomial_inference_summary["Lower 95% CI for exponentiated estimate"] = np.exp(
#     final_negative_binomial_inference_summary["Lower 95% CI on log scale"])
# 
# final_negative_binomial_inference_summary["Upper 95% CI for exponentiated estimate"] = np.exp(
#     final_negative_binomial_inference_summary["Upper 95% CI on log scale"])
# 
# final_negative_binomial_inference_summary["Percent change"] = (100
#     * (final_negative_binomial_inference_summary["Exponentiated estimate"] - 1))
# 
# final_negative_binomial_inference_log_scale_display = (final_negative_binomial_inference_summary
#     [["Term", "Log-scale estimate", "Standard error", "Wald statistic", "p-value",
#         ]].round(2))
# 
# classical_negative_binomial_test_refit_log_scale_py_html = (
#     scrollable_table_html(final_negative_binomial_inference_log_scale_display))
```

```{r}
#| echo: false
#| label: tbl-classical-negative_binomial-test-refit-log-scale-py
#| tbl-cap: "Testing-set refit of the extended Classical negative_binomial regression model on the log expected-count scale."
#| message: false
#| warning: false
#| results: asis

# cat(
#   "\n",
#   py$classical_negative_binomial_test_refit_log_scale_py_html,
#   "\n",
#   sep = ""
# )
```

:::

@tbl-negative_binomial-test-refit-log-scale-r reports the **testing-set refit** on the scale used by the model. For each coefficient, the **Wald statistic** is the **ratio between the estimated coefficient and its model-based standard error**. Therefore, large positive or negative Wald statistics indicate estimates that are far from $0$ relative to their estimated uncertainty. However, the **log expected-count scale** is not always the easiest scale for substantive interpretation. Since the model uses a log link (see @eq-negative_binomial-results-extended-model), exponentiating a coefficient gives a multiplicative comparison on the expected-count scale (see @eq-classical-negative_binomial-extended-mean). Thus, @tbl-negative_binomial-test-refit-expected-count-r reports these exponentiated estimates, their approximate 95% confidence intervals, and the corresponding percent changes.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| label: tbl-negative-binomial-test-refit-expected-count-r
#| tbl-cap: "Testing-set refit of the extended Classical Poisson regression model on the expected-count ratio scale."
#| message: false
#| warning: false

# final_poisson_inference_summary |>
#   select(term, exponentiated_estimate, exponentiated_conf_low, exponentiated_conf_high,
#     percent_change) |>
#   rename(`Term` = term, `Exponentiated estimate` = exponentiated_estimate,
#     `Lower 95% CI for exponentiated estimate` = exponentiated_conf_low,
#     `Upper 95% CI for exponentiated estimate` = exponentiated_conf_high,
#     `Percent change` = percent_change) |>
#   kable(align = c("c", "c", "c", "c", "c"))
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

# final_poisson_inference_expected_count_display = (final_poisson_inference_summary
#     [["Term", "Exponentiated estimate", "Lower 95% CI for exponentiated estimate",
#             "Upper 95% CI for exponentiated estimate", "Percent change",
#         ]].round(2))
# 
# classical_poisson_test_refit_expected_count_py_html = (
#     scrollable_table_html(final_poisson_inference_expected_count_display))
```

```{r}
#| echo: false
#| label: tbl-classical-poisson-test-refit-expected-count-py
#| tbl-cap: "Testing-set refit of the extended Classical Poisson regression model on the expected-count ratio scale."
#| message: false
#| warning: false
#| results: asis

# cat(
#   "\n",
#   py$classical_poisson_test_refit_expected_count_py_html,
#   "\n",
#   sep = ""
# )
```

:::

@tbl-negative-binomial-test-refit-expected-count-r translates the same testing-set refit into the scale used for **interpretation**: 

1. An exponentiated estimate of $1$ corresponds to **no multiplicative change** in the expected count. 
2. Values above $1$ indicate a **larger expected count**, holding the other regressors fixed.
3. Values below $1$ indicate a **smaller expected count**, holding the other regressors fixed.

For example, the fitted exponentiated estimate for `width_cm` is `.... This means the following:

> *Holding weight, colour, and spine condition fixed, a one-centimetre increase in female crab width is associated with multiplying the expected number of satellite males by ... under the testing-set refit. Equivalently, this corresponds to an estimated percent change of ...*

For a categorical example, consider the `colorMedium dark` row. Its fitted exponentiated estimate is .... This means the following: 

> *Holding width, weight, and spine condition fixed, female crabs in the `Medium dark` colour category are estimated to have an expected satellite count equal to ... times the expected count for otherwise comparable female crabs in the `Light` reference category. Equivalently, this corresponds to an estimated percent difference of ...*

These examples describe the **estimated coefficient interpretations** on the expected-count scale. They are not yet the final inferential conclusions. Statistical conclusions require reading these estimates together with their standard errors, confidence intervals, Wald tests, and the broader goodness-of-fit concerns. Having said all this, the model form was chosen during the **training-stage workflow**; this testing-set refit is used to report coefficient-level results for the inferential inquiry, not for additional model selection.


## Storytelling


## Practice Problems