<!-- Google tag (gtag.js) -->

```{=html}
<script async src="https://www.googletagmanager.com/gtag/js?id=G-7PRVEBE1EF"></script>
```

```{=html}
<script>
  window.dataLayer = window.dataLayer || [];
  function gtag(){dataLayer.push(arguments);}
  gtag('js', new Date());

  gtag('config', 'G-7PRVEBE1EF');
</script>
```

<!-- MathJax configuration -->

```{=html}
<script>
  MathJax = {
    tex: {
      tags: 'ams',
      useLabelIds: true
    }
  };
</script>
```

# Zero-Inflated Poisson Regression {#sec-zero-inflated-poisson}

```{r}
#| include: false

colourize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color, x)
  } else x
}
```

::::: Warning
::: Warning-header
When to Use and Not Use Zero-Inflated Poisson Regression
:::

::: Warning-container
Zero-Inflated Poisson (ZIP) regression is useful when:

- The outcome variable is **discrete** and **count-type**, taking values $0, 1, 2, \dots$.
- The data contain **more zeros than a standard Poisson model would typically expect**.
- It is scientifically plausible that zeros may arise from **more than one mechanism**.
- We want to model both whether an observation belongs to a **certain-zero group** and, separately, the expected count among observations that remain in the count-generating process.

However, ZIP regression should not be used automatically whenever zeros appear in a dataset. In particular:

- If the outcome is **binary**, use the model introduced in @sec-binary-logistic.
- If the outcome is a count but the data do not exhibit unusual zero inflation, start with the Classical Poisson model introduced in @sec-classical-poisson.
- If the outcome is overdispersed for reasons not primarily driven by excess zeros, compare against the Negative Binomial model introduced in @sec-negative-binomial.
- If there is no substantive reason to believe that a separate zero-generating process exists, a simpler count model is usually easier to interpret.
:::
:::::

::: {#fig-zero-inflated-poisson-regression}
```{mermaid}
mindmap
  root((Regression 
  Analysis)
    Continuous <br/>Outcome Y
      {{Unbounded <br/>Outcome Y}}
        )Chapter 3: <br/>Ordinary <br/>Least Squares <br/>Regression(
          (Normal <br/>Outcome Y)
      {{Nonnegative <br/>Outcome Y}}
        )Chapter 4: <br/>Gamma Regression(
          (Gamma <br/>Outcome Y)
      {{Bounded <br/>Outcome Y <br/> between 0 and 1}}
        )Chapter 5: Beta <br/>Regression(
          (Beta <br/>Outcome Y)
      {{Nonnegative <br/>Survival <br/>Time Y}}
        )Chapter 6: <br/>Parametric <br/> Survival <br/>Regression(
          (Exponential <br/>Outcome Y)
          (Weibull <br/>Outcome Y)
          (Lognormal <br/>Outcome Y)
        )Chapter 7: <br/>Semiparametric <br/>Survival <br/>Regression(
          (Cox Proportional <br/>Hazards Model)
            (Hazard Function <br/>Outcome Y)
    Discrete <br/>Outcome Y
      {{Binary <br/>Outcome Y}}
        {{Ungrouped <br/>Data}}
          )Chapter 8: <br/>Binary Logistic <br/>Regression(
            (Bernoulli <br/>Outcome Y)
        {{Grouped <br/>Data}}
          )Chapter 9: <br/>Binomial Logistic <br/>Regression(
            (Binomial <br/>Outcome Y)
      {{Count <br/>Outcome Y}}
        {{Equidispersed <br/>Data}}
          )Chapter 10: <br/>Classical Poisson <br/>Regression(
            (Poisson <br/>Outcome Y)
        {{Overdispersed <br/>Data}}
          )Chapter 11: <br/>Negative Binomial <br/>Regression(
            (Negative Binomial <br/>Outcome Y)
        {{Overdispersed or <br/>Underdispersed <br/>Data}}
        {{Zero Inflated <br/>Data}}
          )Chapter 12: <br/>Zero Inflated Poisson <br/>Regression(
            (Zero Inflated <br/>Poisson <br/>Outcome Y)
```

Regression analysis mind map depicting where Zero-Inflated Poisson regression fits within the broader family of models introduced in this book.
:::

```{r setup-r, include=FALSE}
# Load required libraries
library(readr)
library(tibble)
library(tidyr)
library(dplyr)
library(purrr)
library(magrittr)
library(ggplot2)
library(scales)
library(knitr)
library(reticulate)
library(DT)

# These packages are used later for model fitting.
# Install them before rendering if they are not already available.
# install.packages(c("pscl", "broom"))
library(pscl)
library(broom)

# Local data path used in the current chapter draft.
# If the cookbook later hosts the CSV in GitHub, replace this local path with the raw CSV URL.
zip_data_path <- "12-zip_files/zero_inflated_poisson.csv"

if (file.exists(zip_data_path)) {
  ZIP_data <- read_csv(zip_data_path, show_col_types = FALSE)
} else {
  ZIP_data <- tibble()
}

data <- ZIP_data
response_var <- "claims"

if (nrow(data) > 0) {
  data <- data |>
    mutate(
      has_claim = as.integer(.data[[response_var]] > 0),
      male = as.integer(male),
      urban_area = as.integer(urban_area),
      prior_claims = as.integer(prior_claims)
    )
}

print_as_dt <- function(x) {
  if (knitr::is_latex_output()) {
    knitr::kable(
      x,
      format = "latex",
      booktabs = TRUE,
      longtable = TRUE
    )
  } else if (knitr::is_html_output()) {
    DT::datatable(
      x,
      rownames = FALSE,
      options = list(
        dom = 'ltipr',
        autoWidth = TRUE,
        columnDefs = list(
          list(className = 'dt-left', targets = "_all"),
          list(width = '200px', targets = "_all")
        )
      )
    )
  } else {
    knitr::kable(x)
  }
}

fmt_sig <- function(x, digits = 4) {
  if (length(x) == 0 || is.null(x) || all(is.na(x))) return("NA")
  x <- x[1]
  if (is.numeric(x) && abs(x - round(x)) < .Machine$double.eps^0.5) {
    return(format(round(x), big.mark = ",", scientific = FALSE, trim = TRUE))
  }
  format(signif(x, digits), scientific = FALSE, trim = TRUE)
}

fmt_table <- function(x, digits = 4) {
  x |>
    mutate(across(where(is.numeric), ~ signif(.x, digits)))
}

zip_has_data <- function() {
  exists("data") && nrow(data) > 0 && response_var %in% names(data)
}

zip_has_train <- function() {
  exists("train_data") && nrow(train_data) > 0 && response_var %in% names(train_data)
}

zip_stat <- function(name) {
  if (exists("zip_eda_stats") && name %in% names(zip_eda_stats)) {
    fmt_sig(zip_eda_stats[[name]])
  } else {
    "NA"
  }
}

zip_theme <- theme_bw() +
  theme(
    axis.text = element_text(size = 15.5),
    axis.title.x = element_text(size = 20),
    axis.title.y = element_text(size = 20, vjust = 0.5),
    strip.text = element_text(size = 20),
    legend.position = "bottom"
  )

py_df_to_tibble <- function(py_object_name) {
  records <- reticulate::py_eval(
    paste0(py_object_name, ".to_dict(orient='records')"),
    convert = TRUE
  )

  dplyr::bind_rows(records)
}

py_require("pandas")
py_require("numpy")
py_require("matplotlib")
py_require("seaborn")
py_require("statsmodels")
py_require("scikit-learn")
```

```{python setup-py, include=FALSE}
from pathlib import Path
import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
import statsmodels.api as sm
from statsmodels.discrete.count_model import ZeroInflatedPoisson

zip_data_path = Path("12-zip_files/zero_inflated_poisson.csv")

if zip_data_path.exists():
    data = pd.read_csv(zip_data_path)
else:
    data = pd.DataFrame()

response_var = "claims"

if not data.empty:
    data = data.copy()
    data["has_claim"] = (data[response_var] > 0).astype(int)
    for col in ["male", "urban_area", "prior_claims"]:
        data[col] = data[col].astype(int)

# Shared visual settings for Python plots.
# This mirrors the white-grid seaborn/matplotlib style used in the Classical Poisson chapter.
sns.set_theme(style="whitegrid")

ZIP_FIGSIZE = (8, 5)
ZIP_WIDE_FIGSIZE = (14, 8)
ZIP_BAR_COLOR = "0.35"
ZIP_POINT_COLOR = "0.10"
ZIP_LINE_COLOR = "0.10"

plt.rcParams.update({
    "figure.figsize": ZIP_FIGSIZE,
    "axes.labelsize": 12,
    "axes.titlesize": 13,
    "xtick.labelsize": 10,
    "ytick.labelsize": 10,
    "legend.fontsize": 10,
})

def zip_clean_axis(ax):
    """Apply a consistent seaborn/matplotlib finishing style."""
    ax.set_axisbelow(True)
    ax.grid(True, axis="both", alpha=0.35)
    sns.despine(ax=ax)
    return ax
```

::::: LO
::: LO-header
Learning Objectives
:::

::: LO-container
By the end of this chapter, you will be able to:

- Explain why Zero-Inflated Poisson (ZIP) regression is useful for count outcomes with **more zeros than a standard Poisson model would typically expect**.
- Describe the difference between **structural zeros** and **ordinary sampling zeros** in a count-data setting.
- Identify the response variable and main explanatory variables in the insurance-claims example used throughout the chapter.
- Use exploratory summaries and plots to assess whether a standard Poisson model may be inadequate.
- Write the two components of a ZIP model: a **zero-inflation model** and a **Poisson count model**.
- Fit and compare Classical Poisson or Logistic Regression with ZIP models using both `R` and `Python`.
- Interpret coefficients from the count component and the zero-inflation component.
:::
:::::

## Introduction

In many applied regression problems, the outcome of interest is a **count**: the number of events, occurrences, or cases observed for an individual or unit over a fixed period of time. Examples include the number of hospital visits, the number of customer complaints, the number of defects on a manufactured item, or the number of insurance claims filed by a policyholder in a year.

A common starting point for this kind of outcome is **Classical Poisson regression**, introduced in @sec-classical-poisson, which is designed for nonnegative integer-valued responses. However, in practice, some count datasets contain **many more zeros than a simple Poisson model would be expected to generate**. In those situations, a standard Poisson regression may fit poorly, especially near zero, and may fail to represent the data-generating process in a meaningful way.

This chapter introduces **Zero-Inflated Poisson (ZIP) regression**, a model designed for count data with **excess zeros**. ZIP regression is **similar to a generalized linear model (GLM)** because each part uses a familiar GLM-style link function. More precisely, though, it is best understood as the combination of two linked regression components: a logistic-style model for whether an observation belongs to a zero-generating state, and a Poisson log-link model for the count-generating state. Thus, ZIP regression connects naturally to @sec-binary-logistic and @sec-classical-poisson respectively, but it is not just a single ordinary GLM.

::::: {#Definition-zip .definition}
::: definition-header
Definition of Zero-Inflated Poisson
:::

::: definition-container
A **Zero-Inflated Poisson (ZIP) model** is a two-component model for a count response with excess zeros. One component models the probability that an observation belongs to a latent group that can only produce zeros. The other component models the expected count among observations that remain in the ordinary Poisson count-generating process.
:::
:::::

Throughout the chapter, we use a **synthetic insurance-claims dataset** as a running example. In this setting, the response of interest is the **number of claims filed by a policyholder during a fixed time window**. This is a useful example because insurance data often contain a large proportion of zeros: many policyholders do not file any claims at all, while a smaller subset files one or more. That practical feature makes the insurance setting a natural context for introducing zero-inflated modelling.

The chapter is organised as follows:

- **Introduction and chapter roadmap:** We introduce Zero-Inflated Poisson (ZIP) regression, explain why it is useful for count data with excess zeros, and position it relative to related cookbook chapters.
- **Running example and data context:** We introduce the cookbook insurance-claims dataset, clarify the response variable, classify the covariates, and explain what a zero means in this setting.
- **Data splitting:** Before EDA or model fitting, we follow the workflow introduced in @sec-ds-workflow and create training and testing sets. The training set is used for EDA and model fitting, while the testing set is held out for later prediction and model assessment.
- **Exploratory Data Analysis (EDA):** Using only the training data, we examine the observed count distribution, the number of zeros and non-zeros, marginal covariate patterns, and the empirical evidence that motivates a zero-inflated model.
- **Why not just use logistic regression?:** We explain what is lost when the count response is reduced to a zero-versus-nonzero outcome and connect this discussion to @sec-binary-logistic.
- **Why not just use Poisson regression?:** We fit and interpret the Classical Poisson baseline, describe its log-link structure, and explain why a single-process count model may not be sufficient; see also @sec-classical-poisson.
- **Intuition behind zero inflation:** We distinguish between structural zeros and ordinary sampling zeros using the insurance example.
- **Overview of the Zero-Inflated Poisson model:** We introduce the two-component ZIP model formally, including the zero-inflation component, the count component, the mean and variance, and the role of the Hessian in standard-error estimation.
- **Interpreting the two components in the running example:** We explain how covariates can affect the probability of belonging to the extra-zero group, the expected claim count, or both.
- **Estimation and model fitting:** We fit the ZIP model numerically and show the full workflow in both R and Python.
- **Interpretation of fitted results:** We read the fitted coefficient tables for both components and translate the results into insurance-claims language.
- **Inference for model parameters:** We discuss standard errors, confidence intervals, p-values, and inferential statements for both the count and zero-inflation components.
- **Hypothesis testing:** We test individual covariates, compare nested ZIP models, and discuss when formal comparisons against simpler alternatives are useful.
- **Prediction with the ZIP model:** We generate predicted expected counts, predicted zero probabilities, probabilities of specific counts, simulated outcomes, and evaluate prediction quality on held-out data.
- **Comparing ZIP to simpler alternatives:** We compare ZIP with logistic regression and Classical Poisson regression, using both statistical summaries and practical interpretation.
- **Model assessment and diagnostics:** We assess whether the fitted model captures the observed zero frequency and count distribution, and whether it improves meaningfully over simpler models.
- **Practical considerations and interpretation cautions:** We discuss common pitfalls, related models such as hurdle and zero-inflated negative binomial models, and the importance of substantive justification.
- **Chapter summary:** We recap the motivation, two-part structure, inferential goals, predictive goals, and practical lessons from the running insurance example.
- **Optional appendices:** We reserve space for additional mathematical details, implementation notes, and extended reproducibility scripts in R and Python.

::::: Heads-up
::: Heads-up-header
Heads-up on further documented use-cases!
:::

::: Heads-up-container
ZIP models are not only teaching examples. Documented applied uses include [manufacturing defects](https://www.stat.cmu.edu/technometrics/90-00/vol-34-01/v3401001.pdf?utm_source=chatgpt.com), [insurance claim counts](file:///C:/Users/hpala/Downloads/55914343.pdf), [public-health and epidemiology count outcomes](https://pmc.ncbi.nlm.nih.gov/articles/PMC3238139/), [health-care utilization](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0291426), among others.
:::
:::::

::::: Tip
::: Tip-header
Tip on model selection for count variables
:::

::: Tip-container
When deciding between @sec-classical-poisson, @sec-negative-binomial, and @sec-zero-inflated-poisson, do not rely only on the number of observed zeros. Combine the empirical evidence with a subject-matter explanation for why some observations may come from a different zero-generating mechanism.
:::
:::::

## Running Example and Data Context

To motivate ZIP regression, we use the insurance-claims data prepared for this chapter of the Regression Cookbook. The current draft loads the file locally from `12-zip_files/zero_inflated_poisson.csv`; this local path can later be replaced by a raw GitHub URL if the cookbook data are hosted centrally. The dataset was provided for the cookbook workflow, and we treat it here as the chapter's teaching dataset rather than as an external real-world insurance portfolio.

Each row in the dataset represents one policyholder, and the main outcome records how many insurance claims that policyholder filed during the observation window. The response variable is `claims`, a nonnegative integer-valued count: $0, 1, 2, \dots$

The dataset contains the following variables.

| **Variable Name** | **Coding Name Alias** | **Type** | **Scale** | **Model Role** | **Notes** |
|:----------:|:----------:|:----------:|:----------:|:----------:|:-----------|
| Number of Insurance Claims (*past year*) | `claims` | Discrete | Nonnegative count; positively unbounded in principle | Response | Zero-inflated count outcome. A value of $0$ means no claims were filed during the observation window. |
| Driver Age (*years*) | `driver_age` | Continuous | Bounded by design: 18 - 75 years | Regressor | Age of the policyholder or driver. |
| Annual Mileage (*km/year*) | `annual_mileage` | Continuous | Bounded by design: 2,000 - 40,000 km/year | Regressor | Total kilometers driven per year. |
| Vehicle Age (*years*) | `vehicle_age` | Discrete | Bounded integer: 0 - 20 years | Regressor | Age of the insured vehicle. |
| Sex Indicator (*female* or *male*) | `male` | Discrete | Bounded binary indicator: $0$ or $1$ | Regressor | $1$ indicates male and $0$ indicates female, according to the dataset coding. |
| Driving Environment (*rural* or *urban*) | `urban_area` | Discrete | Bounded binary indicator: $0$ or $1$ | Regressor | $1$ indicates urban driving environment and $0$ indicates rural driving environment. |
| Prior Claims Indicator (*no* or *yes*) | `prior_claims` | Discrete | Bounded binary indicator: $0$ or $1$ | Regressor | $1$ indicates at least one prior claim and $0$ indicates no prior claims. |

: Variables in the zero-inflated insurance-claims dataset.

A value of **zero** for the response means that the policyholder **filed no claims during the observation period**. Importantly, that zero can be interpreted in more than one way. In some cases, the policyholder may be effectively outside the claims-generating process for that period, or had such limited exposure that a positive claim count was structurally unlikely. In other cases, the policyholder may have been fully at risk of filing a claim but simply happened not to do so. This distinction is central to the two-part structure of the ZIP model, which will be discussed later in the data-modelling section.

### Loading and previewing the data

The following scripts load the cookbook CSV and create the `has_claim` indicator used during the EDA. The two languages use the same column names, but they store objects in language-specific data structures: a tibble in `R` and a pandas DataFrame in `Python`.

::: panel-tabset
## **`R` Code**

``` r
library(readr)
library(dplyr)

zip_data_path <- "12-zip_files/zero_inflated_poisson.csv"

ZIP_data <- read_csv(zip_data_path, show_col_types = FALSE) |>
  mutate(
    has_claim = as.integer(claims > 0),
    male = as.integer(male),
    urban_area = as.integer(urban_area),
    prior_claims = as.integer(prior_claims)
  )

data <- ZIP_data
head(data, 10)
```

## **`Python` Code**

``` python
from pathlib import Path
import pandas as pd

zip_data_path = Path("12-zip_files/zero_inflated_poisson.csv")

data = pd.read_csv(zip_data_path)
data["has_claim"] = (data["claims"] > 0).astype(int)
for col in ["male", "urban_area", "prior_claims"]:
    data[col] = data[col].astype(int)

data.head(10)
```
:::

::: panel-tabset
## **`R` Output**

```{r tbl-load-zip-data-r}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "First 10 rows of the zero-inflated insurance-claims dataset using R."

if (zip_has_data()) {
  head(data, 10) |>
      print_as_dt()
} else {
  tibble(Message = paste0("Data file not found at ", zip_data_path)) |>
      print_as_dt()
}
```

## **`Python` Output**

```{python load-zip-data-py-run}
#| echo: false
#| results: hide

if zip_data_path.exists():
    data = pd.read_csv(zip_data_path)
    data["has_claim"] = (data["claims"] > 0).astype(int)
    for col in ["male", "urban_area", "prior_claims"]:
        data[col] = data[col].astype(int)
else:
    data = pd.DataFrame()

data_preview = data.head(10).copy()
```

```{r tbl-load-zip-data-py}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "First 10 rows of the zero-inflated insurance-claims dataset using Python."

py_records <- reticulate::py_eval(
  "data.head(10).to_dict(orient='records')",
  convert = TRUE
)

py_data_preview <- dplyr::bind_rows(py_records)

if (is.data.frame(py_data_preview) && nrow(py_data_preview) > 0) {
  py_data_preview |>
      print_as_dt()
} else {
  tibble(Message = "Python data preview is empty or unavailable.") |>
      print_as_dt()
}
```
:::

The preview should show seven original variables plus the derived `has_claim` indicator. In the EDA scratchpad for this chapter, the file contains 1,000 rows, the expected seven original columns, and no missing values.

## Data Splitting

Following the data science workflow introduced in @sec-ds-workflow, we split the data before conducting EDA or fitting models. The training set is used for exploratory analysis, model fitting, and coefficient interpretation. The testing set is held out for later prediction and model assessment.

In this chapter, the split is created by **shuffling the row order with seed 123** and then cutting the shuffled data at the 80% threshold for training. We intentionally **do not stratify** by `has_claim`; with a simple random shuffle, the sample itself is expected to carry the excess-zero structure into the training and testing sets.

::::: Heads-up
::: Heads-up-header
Heads-up on `R` and `Python` data splits!
:::

::: Heads-up-container
Even when the same seed value is used, `R` and `Python` may create different training and testing sets because they rely on different pseudo-random number generators and different shuffling implementations. The code below deliberately shows the same conceptual procedure in both languages. It **will produce different row assignments**.
:::
:::::

::: panel-tabset
## **`R` Code**

``` r
set.seed(123)

split_data <- data |>
  mutate(row_id = row_number(), .before = 1)

n <- nrow(split_data)
split_index <- floor(0.8 * n)
shuffled_rows <- sample(seq_len(n), size = n, replace = FALSE)

train_data <- split_data[shuffled_rows[seq_len(split_index)], ]
test_data <- split_data[shuffled_rows[(split_index + 1):n], ]

split_check_r <- tibble(
  split = c("training", "testing"),
  n = c(nrow(train_data), nrow(test_data)),
  proportion = c(nrow(train_data), nrow(test_data)) / nrow(split_data)
)

# Share the canonical R split with Python for all later computations.
py$training_data <- r_to_py(train_data)
py$testing_data <- r_to_py(test_data)
```

## **`Python` Code**

``` python
rng = np.random.default_rng(123)

data_with_row_id = data.copy()
if "row_id" not in data_with_row_id.columns:
    data_with_row_id.insert(0, "row_id", np.arange(1, len(data_with_row_id) + 1))

n = len(data_with_row_id)
split_index = int(np.floor(0.8 * n))
shuffled_rows = rng.permutation(n)

training_data_independent = data_with_row_id.iloc[shuffled_rows[:split_index]].copy()
testing_data_independent = data_with_row_id.iloc[shuffled_rows[split_index:]].copy()

split_check_py_independent = pd.DataFrame({
    "split": ["training", "testing"],
    "n": [len(training_data_independent), len(testing_data_independent)],
    "proportion": [
        len(training_data_independent) / len(data_with_row_id),
        len(testing_data_independent) / len(data_with_row_id),
    ],
})

# The canonical Python objects used later are supplied by R through reticulate:
# training_data and testing_data
```
:::

::: panel-tabset
## **`R` Output**

```{r tbl-split-zip-data-r}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Training and testing split check using the shuffled R split."

if (zip_has_data()) {
  set.seed(123)

  split_data <- data |>
    mutate(row_id = row_number(), .before = 1)

  n <- nrow(split_data)
  split_index <- floor(0.8 * n)
  shuffled_rows <- sample(seq_len(n), size = n, replace = FALSE)

  train_data <- split_data[shuffled_rows[seq_len(split_index)], ]
  test_data <- split_data[shuffled_rows[(split_index + 1):n], ]

  split_check_r <- tibble(
    split = c("training", "testing"),
    n = c(nrow(train_data), nrow(test_data)),
    proportion = c(nrow(train_data), nrow(test_data)) / nrow(split_data)
  )

  py$training_data <- r_to_py(train_data)
  py$testing_data <- r_to_py(test_data)

  split_check_r |>
    fmt_table() |>
      print_as_dt()
} else {
  tibble(Message = paste0("Data file not found at ", zip_data_path)) |>
      print_as_dt()
}
```

## **`Python` Output**

```{python split-zip-data-py-independent-run}
#| echo: false
#| results: hide

split_check_py_independent = pd.DataFrame()

if not data.empty:
    rng = np.random.default_rng(123)

    data_with_row_id = data.copy()
    if "row_id" not in data_with_row_id.columns:
        data_with_row_id.insert(0, "row_id", np.arange(1, len(data_with_row_id) + 1))

    n = len(data_with_row_id)
    split_index = int(np.floor(0.8 * n))
    shuffled_rows = rng.permutation(n)

    training_data_independent = data_with_row_id.iloc[shuffled_rows[:split_index]].copy()
    testing_data_independent = data_with_row_id.iloc[shuffled_rows[split_index:]].copy()

    split_check_py_independent = pd.DataFrame({
        "split": ["training", "testing"],
        "n": [len(training_data_independent), len(testing_data_independent)],
        "proportion": [
            len(training_data_independent) / len(data_with_row_id),
            len(testing_data_independent) / len(data_with_row_id),
        ],
    })
```

```{r tbl-split-zip-data-py}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Training and testing split check using the independently shuffled Python split."

if (reticulate::py_eval("'split_check_py_independent' in globals()", convert = TRUE)) {
  split_check_py_independent_r <- reticulate::py_eval(
    "split_check_py_independent.to_dict(orient='records')",
    convert = TRUE
  ) |>
    dplyr::bind_rows()

  if (nrow(split_check_py_independent_r) > 0) {
    split_check_py_independent_r |>
      fmt_table() |>
        print_as_dt()
  } else {
    tibble(Message = "Independent Python split check is empty.") |>
        print_as_dt()
  }
} else {
  tibble(Message = paste0("Data file not found at ", zip_data_path)) |>
      print_as_dt()
}
```
:::

Although both languages use the same seed and the same 80/20 shuffled-split idea, they are not expected to select the same rows. From this point forward, the EDA and model fitting are based on the canonical `R` training set only. In the `Python` chunks, the objects `training_data` and `testing_data` are copies of the canonical `R` split, passed through `reticulate`.

```{r zip-compute-eda-values, include=FALSE}
if (zip_has_train()) {
  claims_summary_wide <- train_data |>
    summarise(
      `Number of observations` = n(),
      `Number of zeros` = sum(claims == 0, na.rm = TRUE),
      `Proportion of zeros` = mean(claims == 0, na.rm = TRUE),
      `Sample mean` = mean(claims, na.rm = TRUE),
      `Sample variance` = var(claims, na.rm = TRUE),
      `Variance-to-mean ratio` = var(claims, na.rm = TRUE) / mean(claims, na.rm = TRUE),
      `Poisson expected zero rate` = exp(-mean(claims, na.rm = TRUE)),
      `Excess zero rate` = mean(claims == 0, na.rm = TRUE) - exp(-mean(claims, na.rm = TRUE)),
      `Maximum` = max(claims, na.rm = TRUE)
    )

  claims_summary <- claims_summary_wide |>
    pivot_longer(
      cols = everything(),
      names_to = "Summary",
      values_to = "Value"
    )

  zero_group_summary <- train_data |>
    group_by(has_claim) |>
    summarise(
      n = n(),
      mean_driver_age = mean(driver_age, na.rm = TRUE),
      mean_annual_mileage = mean(annual_mileage, na.rm = TRUE),
      mean_vehicle_age = mean(vehicle_age, na.rm = TRUE),
      male_rate = mean(male, na.rm = TRUE),
      urban_rate = mean(urban_area, na.rm = TRUE),
      prior_claims_rate = mean(prior_claims, na.rm = TRUE),
      mean_claims = mean(claims, na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(
      claim_status = if_else(has_claim == 0, "Zero claims", "At least one claim")
    ) |>
    select(claim_status, everything(), -has_claim)

  binary_cols <- c("male", "urban_area", "prior_claims")

  binary_summary <- purrr::map_dfr(binary_cols, function(col) {
    train_data |>
      group_by(value = .data[[col]]) |>
      summarise(
        n = n(),
        mean_claims = mean(claims, na.rm = TRUE),
        var_claims = var(claims, na.rm = TRUE),
        zero_rate = mean(claims == 0, na.rm = TRUE),
        nonzero_rate = mean(claims > 0, na.rm = TRUE),
        max_claims = max(claims, na.rm = TRUE),
        .groups = "drop"
      ) |>
      mutate(variable = col, .before = 1)
  })

  make_binned_summary <- function(df, variable, bins = 6) {
    x <- df[[variable]]
    bin_breaks <- seq(
      min(x, na.rm = TRUE),
      max(x, na.rm = TRUE),
      length.out = bins + 1
    )
  
    df |>
      mutate(
        variable = .env$variable,
        bin_id = cut(
          .data[[variable]],
          breaks = bin_breaks,
          include.lowest = TRUE,
          labels = FALSE
        ),
        bin = cut(
          .data[[variable]],
          breaks = bin_breaks,
          include.lowest = TRUE,
          dig.lab = 6
        ),
        bin_lower = bin_breaks[bin_id],
        bin_upper = bin_breaks[bin_id + 1],
        bin_midpoint = (bin_lower + bin_upper) / 2,
        bin_width = bin_upper - bin_lower
      ) |>
      filter(!is.na(bin_id)) |>
      group_by(
        variable,
        bin_id,
        bin,
        bin_lower,
        bin_upper,
        bin_midpoint,
        bin_width
      ) |>
      summarise(
        n = n(),
        mean_claims = mean(claims, na.rm = TRUE),
        zero_rate = mean(claims == 0, na.rm = TRUE),
        .groups = "drop"
      ) |>
      arrange(variable, bin_midpoint)
  }
  
  binned_summary <- bind_rows(
    make_binned_summary(train_data, "driver_age"),
    make_binned_summary(train_data, "annual_mileage"),
    make_binned_summary(train_data, "vehicle_age")
  )

  correlation_summary <- train_data |>
    select(driver_age, annual_mileage, vehicle_age, male, urban_area, prior_claims, claims, has_claim) |>
    summarise(across(
      c(driver_age, annual_mileage, vehicle_age, male, urban_area, prior_claims),
      list(
        corr_with_claims = ~ cor(.x, claims),
        corr_with_has_claim = ~ cor(.x, has_claim)
      ),
      .names = "{.col}_{.fn}"
    )) |>
    pivot_longer(
      cols = everything(),
      names_to = c("variable", ".value"),
      names_pattern = "(.*)_(corr_with_.*)"
    )

  prior_claims_yes <- binary_summary |> filter(variable == "prior_claims", value == 1)
  prior_claims_no <- binary_summary |> filter(variable == "prior_claims", value == 0)
  urban_yes <- binary_summary |> filter(variable == "urban_area", value == 1)
  urban_no <- binary_summary |> filter(variable == "urban_area", value == 0)
  annual_bins <- binned_summary |> filter(variable == "annual_mileage")
  annual_low <- annual_bins |> arrange(bin_midpoint) |> slice(1)
  annual_high <- annual_bins |> arrange(bin_midpoint) |> slice(n())
  corr_annual <- correlation_summary |> filter(variable == "annual_mileage")
  corr_prior <- correlation_summary |> filter(variable == "prior_claims")

  zip_eda_stats <- list(
    zero_n = sum(train_data$claims == 0, na.rm = TRUE),
    train_n = nrow(train_data),
    max_claims = max(train_data$claims, na.rm = TRUE),
    sample_mean = mean(train_data$claims, na.rm = TRUE),
    sample_variance = var(train_data$claims, na.rm = TRUE),
    variance_to_mean = var(train_data$claims, na.rm = TRUE) / mean(train_data$claims, na.rm = TRUE),
    poisson_zero_rate = exp(-mean(train_data$claims, na.rm = TRUE)),
    observed_zero_rate = mean(train_data$claims == 0, na.rm = TRUE),
    excess_zero_rate = mean(train_data$claims == 0, na.rm = TRUE) - exp(-mean(train_data$claims, na.rm = TRUE)),
    mileage_nonzero = zero_group_summary |> filter(claim_status == "At least one claim") |> pull(mean_annual_mileage),
    mileage_zero = zero_group_summary |> filter(claim_status == "Zero claims") |> pull(mean_annual_mileage),
    prior_rate_nonzero = zero_group_summary |> filter(claim_status == "At least one claim") |> pull(prior_claims_rate),
    prior_rate_zero = zero_group_summary |> filter(claim_status == "Zero claims") |> pull(prior_claims_rate),
    prior_mean_yes = prior_claims_yes$mean_claims,
    prior_zero_yes = prior_claims_yes$zero_rate,
    prior_mean_no = prior_claims_no$mean_claims,
    prior_zero_no = prior_claims_no$zero_rate,
    urban_mean_diff = abs(urban_yes$mean_claims - urban_no$mean_claims),
    urban_zero_diff = abs(urban_yes$zero_rate - urban_no$zero_rate),
    annual_low_mean = annual_low$mean_claims,
    annual_high_mean = annual_high$mean_claims,
    annual_low_zero = annual_low$zero_rate,
    annual_high_zero = annual_high$zero_rate,
    corr_annual_claims = corr_annual$corr_with_claims,
    corr_annual_has_claim = corr_annual$corr_with_has_claim,
    corr_prior_claims = corr_prior$corr_with_claims,
    corr_prior_has_claim = corr_prior$corr_with_has_claim
  )
}
```

```{python validate-canonical-split-py}
#| include: false

numeric_cols = [
    "row_id",
    "driver_age",
    "annual_mileage",
    "vehicle_age",
    "male",
    "urban_area",
    "prior_claims",
    "claims",
    "has_claim",
]

integer_cols = [
    "row_id",
    "vehicle_age",
    "male",
    "urban_area",
    "prior_claims",
    "claims",
    "has_claim",
]

for df_name in ["training_data", "testing_data"]:
    if df_name in globals():
        df = globals()[df_name].copy()

        for col in numeric_cols:
            if col in df.columns:
                df[col] = pd.to_numeric(df[col], errors="coerce")

        for col in integer_cols:
            if col in df.columns:
                df[col] = df[col].astype("Int64")

        globals()[df_name] = df


```

## Exploratory Data Analysis (EDA)

Before fitting any model, we examine whether the empirical distribution of `claims` in the **training data** is consistent with a simple Poisson pattern. The main EDA question is not merely "Are there zeros?" Count data often include zeros. The question is whether there are **more zeros than the Poisson mean-variance structure would lead us to expect**, and whether the covariates show patterns that make a two-component model plausible.

Because the split was performed before EDA, the numerical summaries in this section should be interpreted as **training-set summaries**. The exact values are the same in `R` and `Python` because both languages now use the canonical `R` split.

### Response distribution and zero inflation

A natural first step is to summarise and visualise the count response `claims` in the training data. Because this is a count variable, a bar plot of observed counts is usually more informative than a density plot.

::: panel-tabset
## **`R` Code**

``` r
claim_counts <- train_data |>
  count(claims, name = "n")

ggplot(claim_counts, aes(x = factor(claims), y = n)) +
  geom_col() +
  labs(
    x = "Number of claims",
    y = "Number of policyholders"
  ) +
  zip_theme
```

## **`Python` Code**

``` python
claim_counts = (
    training_data["claims"]
    .astype("int64")
    .value_counts()
    .sort_index()
    .reset_index()
)
claim_counts.columns = ["claims", "n"]

fig, ax = plt.subplots(figsize=ZIP_FIGSIZE)
sns.barplot(
    data=claim_counts,
    x=claim_counts["claims"].astype(str),
    y="n",
    color=ZIP_BAR_COLOR,
    ax=ax
)
ax.set_xlabel("Number of claims")
ax.set_ylabel("Number of policyholders")
zip_clean_axis(ax)
plt.tight_layout()
plt.show()
```
:::

::: panel-tabset
## **`R` Output**

```{r fig-claims-distribution-r}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Observed distribution of the insurance-claims response in the canonical training data using R."
#| eval: !expr zip_has_train()

claim_counts <- train_data |>
  count(claims, name = "n")

ggplot(claim_counts, aes(x = factor(claims), y = n)) +
  geom_col() +
  labs(
    x = "Number of claims",
    y = "Number of policyholders"
  ) +
  zip_theme
```

## **`Python` Output**

```{python fig-claims-distribution-py}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Observed distribution of the insurance-claims response in the canonical training data using Python."
#| eval: !expr reticulate::py_eval("'training_data' in globals() and len(training_data) > 0 and response_var in training_data.columns", convert = TRUE)

claim_counts = (
    training_data["claims"]
    .astype("int64")
    .value_counts()
    .sort_index()
    .reset_index()
)
claim_counts.columns = ["claims", "n"]

fig, ax = plt.subplots(figsize=ZIP_FIGSIZE)
sns.barplot(
    data=claim_counts,
    x=claim_counts["claims"].astype(str),
    y="n",
    color=ZIP_BAR_COLOR,
    ax=ax
)
ax.set_xlabel("Number of claims")
ax.set_ylabel("Number of policyholders")
zip_clean_axis(ax)
plt.tight_layout()
plt.show()
```
:::

As shown in @fig-claims-distribution-r and @fig-claims-distribution-py, the training-data distribution remains strongly concentrated at zero and then declines through the positive counts. In the training set, there are **`r zip_stat("zero_n")`** zero-claim policyholders out of **`r zip_stat("train_n")`** training observations, while the maximum observed count is **`r zip_stat("max_claims")`**. This visual pattern is the first sign that a one-process Poisson model may not be flexible enough.

The next table compares the observed zero rate in the training data with the zero rate implied by a Poisson distribution whose mean equals the training-set sample mean.

::: panel-tabset
## **`R` Code**

``` r
claims_summary <- train_data |>
  summarise(
    `Number of observations` = n(),
    `Number of zeros` = sum(claims == 0, na.rm = TRUE),
    `Proportion of zeros` = mean(claims == 0, na.rm = TRUE),
    `Sample mean` = mean(claims, na.rm = TRUE),
    `Sample variance` = var(claims, na.rm = TRUE),
    `Variance-to-mean ratio` = var(claims, na.rm = TRUE) / mean(claims, na.rm = TRUE),
    `Poisson expected zero rate` = exp(-mean(claims, na.rm = TRUE)),
    `Excess zero rate` = mean(claims == 0, na.rm = TRUE) - exp(-mean(claims, na.rm = TRUE)),
    `Maximum` = max(claims, na.rm = TRUE)
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "Summary",
    values_to = "Value"
  )

claims_summary |>
  fmt_table() |>
    print_as_dt()
```

## **`Python` Code**

``` python
y_train_eda = training_data["claims"]

claims_summary = pd.Series({
    "Number of observations": len(y_train_eda),
    "Number of zeros": int((y_train_eda == 0).sum()),
    "Proportion of zeros": (y_train_eda == 0).mean(),
    "Sample mean": y_train_eda.mean(),
    "Sample variance": y_train_eda.var(ddof=1),
    "Variance-to-mean ratio": y_train_eda.var(ddof=1) / y_train_eda.mean(),
    "Poisson expected zero rate": np.exp(-y_train_eda.mean()),
    "Excess zero rate": (y_train_eda == 0).mean() - np.exp(-y_train_eda.mean()),
    "Maximum": y_train_eda.max()
}).reset_index()
claims_summary.columns = ["Summary", "Value"]
```
:::

::: panel-tabset
## **`R` Output**

```{r tbl-claims-summary-r}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Summary of the insurance-claims response in the canonical training data using R."

if (exists("claims_summary")) {
  claims_summary |>
    fmt_table() |>
      print_as_dt()
} else {
  tibble(Message = "Claims summary is unavailable until the training data are available.") |>
      print_as_dt()
}
```

## **`Python` Output**

```{python claims-summary-py-run}
#| echo: false
#| results: hide

if "training_data" in globals() and len(training_data) > 0:
    y_train_eda = training_data["claims"]

    claims_summary_py = pd.Series({
        "Number of observations": len(y_train_eda),
        "Number of zeros": int((y_train_eda == 0).sum()),
        "Proportion of zeros": (y_train_eda == 0).mean(),
        "Sample mean": y_train_eda.mean(),
        "Sample variance": y_train_eda.var(ddof=1),
        "Variance-to-mean ratio": y_train_eda.var(ddof=1) / y_train_eda.mean(),
        "Poisson expected zero rate": np.exp(-y_train_eda.mean()),
        "Excess zero rate": (y_train_eda == 0).mean() - np.exp(-y_train_eda.mean()),
        "Maximum": y_train_eda.max()
    }).reset_index()
    claims_summary_py.columns = ["Summary", "Value"]
```

```{r tbl-claims-summary-py}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Summary of the insurance-claims response in the canonical training data using Python."

if (reticulate::py_eval("'claims_summary_py' in globals()", convert = TRUE)) {
  claims_summary_py_r <- py_df_to_tibble("claims_summary_py")

  if (nrow(claims_summary_py_r) > 0) {
    claims_summary_py_r |>
      fmt_table() |>
        print_as_dt()
  } else {
    tibble(Message = "Claims summary is empty.") |>
        print_as_dt()
  }
} else {
  tibble(Message = "Claims summary is unavailable until the training data are available.") |>
      print_as_dt()
}
```
:::

This table formalises the visual insight from @fig-claims-distribution-r. In the training data, the sample mean is **`r zip_stat("sample_mean")`**, the sample variance is **`r zip_stat("sample_variance")`**, and the variance-to-mean ratio is **`r zip_stat("variance_to_mean")`**. A Poisson distribution with the same mean would imply an expected zero rate of about **`r zip_stat("poisson_zero_rate")`**, but the observed zero rate is **`r zip_stat("observed_zero_rate")`**. The difference, **`r zip_stat("excess_zero_rate")`**, is a simple empirical measure of excess zeros.

### Zero claims versus nonzero claims

The next comparison keeps the count response intact, but splits the training data into observations with zero claims and observations with at least one claim. This table is especially useful because ZIP regression is motivated by the possibility that the zero and count-generating parts of the data behave differently.

::: panel-tabset
## **`R` Code**

``` r
zero_group_summary <- train_data |>
  group_by(has_claim) |>
  summarise(
    n = n(),
    mean_driver_age = mean(driver_age, na.rm = TRUE),
    mean_annual_mileage = mean(annual_mileage, na.rm = TRUE),
    mean_vehicle_age = mean(vehicle_age, na.rm = TRUE),
    male_rate = mean(male, na.rm = TRUE),
    urban_rate = mean(urban_area, na.rm = TRUE),
    prior_claims_rate = mean(prior_claims, na.rm = TRUE),
    mean_claims = mean(claims, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    claim_status = if_else(has_claim == 0, "Zero claims", "At least one claim")
  ) |>
  select(claim_status, everything(), -has_claim)

zero_group_summary_display <- zero_group_summary |>
  mutate(
    claim_status = factor(
      claim_status,
      levels = c("Zero claims", "At least one claim")
    )
  ) |>
  pivot_longer(
    cols = -claim_status,
    names_to = "Statistic",
    values_to = "Value"
  ) |>
  mutate(
    Statistic = recode(
      Statistic,
      n = "Number of observations",
      mean_driver_age = "Mean driver age",
      mean_annual_mileage = "Mean annual mileage",
      mean_vehicle_age = "Mean vehicle age",
      male_rate = "Male rate",
      urban_rate = "Urban-area rate",
      prior_claims_rate = "Prior-claims rate",
      mean_claims = "Mean claims"
    )
  ) |>
  pivot_wider(
    names_from = claim_status,
    values_from = Value
  )

zero_group_summary_display |>
  fmt_table() |>
    print_as_dt()
```

## **`Python` Code**

``` python
zero_group_summary = (
    training_data.groupby("has_claim")
    .agg(
        n=("claims", "size"),
        mean_driver_age=("driver_age", "mean"),
        mean_annual_mileage=("annual_mileage", "mean"),
        mean_vehicle_age=("vehicle_age", "mean"),
        male_rate=("male", "mean"),
        urban_rate=("urban_area", "mean"),
        prior_claims_rate=("prior_claims", "mean"),
        mean_claims=("claims", "mean")
    )
    .reset_index()
)

zero_group_summary["claim_status"] = zero_group_summary["has_claim"].map({
    0: "Zero claims",
    1: "At least one claim"
})

zero_group_summary = zero_group_summary.drop(columns="has_claim")

statistic_labels = {
    "n": "Number of observations",
    "mean_driver_age": "Mean driver age",
    "mean_annual_mileage": "Mean annual mileage",
    "mean_vehicle_age": "Mean vehicle age",
    "male_rate": "Male rate",
    "urban_rate": "Urban-area rate",
    "prior_claims_rate": "Prior-claims rate",
    "mean_claims": "Mean claims",
}

statistic_order = list(statistic_labels.values())
claim_status_order = ["Zero claims", "At least one claim"]

zero_group_summary_display = zero_group_summary.melt(
    id_vars="claim_status",
    var_name="Statistic",
    value_name="Value"
)
zero_group_summary_display["Statistic"] = zero_group_summary_display["Statistic"].map(statistic_labels)
zero_group_summary_display["Statistic"] = pd.Categorical(
    zero_group_summary_display["Statistic"],
    categories=statistic_order,
    ordered=True
)
zero_group_summary_display["claim_status"] = pd.Categorical(
    zero_group_summary_display["claim_status"],
    categories=claim_status_order,
    ordered=True
)

zero_group_summary_display = (
    zero_group_summary_display
    .pivot(index="Statistic", columns="claim_status", values="Value")
    .reset_index()
)
zero_group_summary_display.columns.name = None
zero_group_summary_display = zero_group_summary_display[
    ["Statistic", "Zero claims", "At least one claim"]
]
```
:::

::: panel-tabset
## **`R` Output**

```{r tbl-zero-vs-nonzero-r}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "covariate comparison between zero-claim and nonzero-claim policyholders in the canonical training data using R."

if (exists("zero_group_summary")) {
  zero_group_summary_display <- zero_group_summary |>
    mutate(
      claim_status = factor(
        claim_status,
        levels = c("Zero claims", "At least one claim")
      )
    ) |>
    pivot_longer(
      cols = -claim_status,
      names_to = "Statistic",
      values_to = "Value"
    ) |>
    mutate(
      Statistic = recode(
        Statistic,
        n = "Number of observations",
        mean_driver_age = "Mean driver age",
        mean_annual_mileage = "Mean annual mileage",
        mean_vehicle_age = "Mean vehicle age",
        male_rate = "Male rate",
        urban_rate = "Urban-area rate",
        prior_claims_rate = "Prior-claims rate",
        mean_claims = "Mean claims"
      )
    ) |>
    pivot_wider(
      names_from = claim_status,
      values_from = Value
    )

  zero_group_summary_display |>
    fmt_table() |>
      print_as_dt()
} else {
  tibble(Message = "Zero-versus-nonzero summary is unavailable until the training data are available.") |>
      print_as_dt()
}
```

## **`Python` Output**

```{python zero-vs-nonzero-py-run}
#| echo: false
#| results: hide

if "training_data" in globals() and len(training_data) > 0:
    zero_group_summary_py = (
        training_data.groupby("has_claim")
        .agg(
            n=("claims", "size"),
            mean_driver_age=("driver_age", "mean"),
            mean_annual_mileage=("annual_mileage", "mean"),
            mean_vehicle_age=("vehicle_age", "mean"),
            male_rate=("male", "mean"),
            urban_rate=("urban_area", "mean"),
            prior_claims_rate=("prior_claims", "mean"),
            mean_claims=("claims", "mean")
        )
        .reset_index()
    )

    zero_group_summary_py["claim_status"] = zero_group_summary_py["has_claim"].map({
        0: "Zero claims",
        1: "At least one claim"
    })

    zero_group_summary_py = zero_group_summary_py.drop(columns="has_claim")

    statistic_labels = {
        "n": "Number of observations",
        "mean_driver_age": "Mean driver age",
        "mean_annual_mileage": "Mean annual mileage",
        "mean_vehicle_age": "Mean vehicle age",
        "male_rate": "Male rate",
        "urban_rate": "Urban-area rate",
        "prior_claims_rate": "Prior-claims rate",
        "mean_claims": "Mean claims",
    }

    statistic_order = list(statistic_labels.values())
    claim_status_order = ["Zero claims", "At least one claim"]

    zero_group_summary_py_display = zero_group_summary_py.melt(
        id_vars="claim_status",
        var_name="Statistic",
        value_name="Value"
    )
    zero_group_summary_py_display["Statistic"] = zero_group_summary_py_display["Statistic"].map(statistic_labels)
    zero_group_summary_py_display["Statistic"] = pd.Categorical(
        zero_group_summary_py_display["Statistic"],
        categories=statistic_order,
        ordered=True
    )
    zero_group_summary_py_display["claim_status"] = pd.Categorical(
        zero_group_summary_py_display["claim_status"],
        categories=claim_status_order,
        ordered=True
    )

    zero_group_summary_py_display = (
        zero_group_summary_py_display
        .pivot(index="Statistic", columns="claim_status", values="Value")
        .reset_index()
    )
    zero_group_summary_py_display.columns.name = None
    zero_group_summary_py_display = zero_group_summary_py_display[
        ["Statistic", "Zero claims", "At least one claim"]
    ]
```

```{r tbl-zero-vs-nonzero-py}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "covariate comparison between zero-claim and nonzero-claim policyholders in the canonical training data using Python."

if (reticulate::py_eval("'zero_group_summary_py_display' in globals()", convert = TRUE)) {
  zero_group_summary_py_display_r <- py_df_to_tibble("zero_group_summary_py_display")

  if (nrow(zero_group_summary_py_display_r) > 0) {
    zero_group_summary_py_display_r |>
      fmt_table() |>
        print_as_dt()
  } else {
    tibble(Message = "Zero-versus-nonzero summary is empty.") |>
        print_as_dt()
  }
} else {
  tibble(Message = "Zero-versus-nonzero summary is unavailable until the training data are available.") |>
      print_as_dt()
}
```
:::

The training-data EDA shows that the nonzero-claim group drives more miles on average: **`r zip_stat("mileage_nonzero")`** annual miles compared with **`r zip_stat("mileage_zero")`** for the zero-claim group. The prior-claims rate is also higher among policyholders with at least one claim during the study period: **`r zip_stat("prior_rate_nonzero")`** compared with **`r zip_stat("prior_rate_zero")`**. This does not prove a causal mechanism, but it suggests that mileage and prior claims may help distinguish the zero and nonzero portions of the response.

### Binary covariates

The binary covariates provide a compact way to inspect whether group membership is associated with different average counts and different zero rates in the training data.

::: panel-tabset
## **`R` Code**

``` r
binary_cols <- c("male", "urban_area", "prior_claims")

binary_summary <- purrr::map_dfr(binary_cols, function(col) {
  train_data |>
    group_by(value = .data[[col]]) |>
    summarise(
      n = n(),
      mean_claims = mean(claims, na.rm = TRUE),
      var_claims = var(claims, na.rm = TRUE),
      zero_rate = mean(claims == 0, na.rm = TRUE),
      nonzero_rate = mean(claims > 0, na.rm = TRUE),
      max_claims = max(claims, na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(variable = col, .before = 1)
})

binary_summary |>
  fmt_table() |>
    print_as_dt()
```

## **`Python` Code**

``` python
binary_cols = ["male", "urban_area", "prior_claims"]
binary_summary = []

for col in binary_cols:
    summary = (
        training_data.groupby(col)
        .agg(
            n=("claims", "size"),
            mean_claims=("claims", "mean"),
            var_claims=("claims", "var"),
            zero_rate=("claims", lambda x: (x == 0).mean()),
            nonzero_rate=("claims", lambda x: (x > 0).mean()),
            max_claims=("claims", "max")
        )
        .reset_index()
        .rename(columns={col: "value"})
    )
    summary.insert(0, "variable", col)
    binary_summary.append(summary)

binary_summary = pd.concat(binary_summary, ignore_index=True)
```
:::

::: panel-tabset
## **`R` Output**

```{r tbl-binary-summary-r}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Claims summary by binary covariates in the canonical training data using R."

if (exists("binary_summary")) {
  binary_summary |>
    fmt_table() |>
      print_as_dt()
} else {
  tibble(Message = "Binary-covariate summary is unavailable until the training data are available.") |>
      print_as_dt()
}
```

## **`Python` Output**

```{python binary-summary-py-run}
#| echo: false
#| results: hide

if "training_data" in globals() and len(training_data) > 0:
    binary_cols = ["male", "urban_area", "prior_claims"]
    binary_summary_py = []

    for col in binary_cols:
        summary = (
            training_data.groupby(col)
            .agg(
                n=("claims", "size"),
                mean_claims=("claims", "mean"),
                var_claims=("claims", "var"),
                zero_rate=("claims", lambda x: (x == 0).mean()),
                nonzero_rate=("claims", lambda x: (x > 0).mean()),
                max_claims=("claims", "max")
            )
            .reset_index()
            .rename(columns={col: "value"})
        )
        summary.insert(0, "variable", col)
        binary_summary_py.append(summary)

    binary_summary_py = pd.concat(binary_summary_py, ignore_index=True)
```

```{r tbl-binary-summary-py}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Claims summary by binary covariates in the canonical training data using Python."

if (reticulate::py_eval("'binary_summary_py' in globals()", convert = TRUE)) {
  binary_summary_py_r <- py_df_to_tibble("binary_summary_py")

  if (nrow(binary_summary_py_r) > 0) {
    binary_summary_py_r |>
      fmt_table() |>
        print_as_dt()
  } else {
    tibble(Message = "Binary-covariate summary is empty.") |>
        print_as_dt()
  }
} else {
  tibble(Message = "Binary-covariate summary is unavailable until the training data are available.") |>
      print_as_dt()
}
```
:::

The strongest binary-covariate contrast in the training-data EDA is expected to remain `prior_claims`. Policyholders with prior claims have a mean of about **`r zip_stat("prior_mean_yes")`** claims and a zero rate of about **`r zip_stat("prior_zero_yes")`**; those without prior claims have a mean of about **`r zip_stat("prior_mean_no")`** claims and a zero rate of about **`r zip_stat("prior_zero_no")`**. By contrast, the `urban_area` groups differ by about **`r zip_stat("urban_mean_diff")`** in mean claims and **`r zip_stat("urban_zero_diff")`** in zero rate in this training set.

### Continuous covariates and visual trends

The continuous covariates can be inspected through scatterplots and binned summaries. Because the response is a discrete count, points stack horizontally at integer values.

::: panel-tabset
## **`R` Code**

``` r
continuous_data <- train_data |>
  select(claims, driver_age, annual_mileage, vehicle_age) |>
  pivot_longer(
    cols = c(driver_age, annual_mileage, vehicle_age),
    names_to = "covariate",
    values_to = "value"
  )

ggplot(continuous_data, aes(x = value, y = claims)) +
  geom_point(alpha = 0.35) +
  facet_wrap(~ covariate, scales = "free_x") +
  labs(
    x = "Covariate value",
    y = "Number of claims"
  ) +
  zip_theme
```

## **`Python` Code**

``` python
continuous_cols = ["driver_age", "annual_mileage", "vehicle_age"]
covariate_labels = {
    "driver_age": "Driver age",
    "annual_mileage": "Annual mileage",
    "vehicle_age": "Vehicle age",
}

plt.close("all")

fig, axes = plt.subplots(
    nrows=1,
    ncols=len(continuous_cols),
    figsize=(14, 4.5),
    sharey=True
)

for ax, col in zip(axes, continuous_cols):
    plot_data = training_data[[col, "claims"]].copy()
    plot_data[col] = pd.to_numeric(plot_data[col], errors="coerce")
    plot_data["claims"] = pd.to_numeric(plot_data["claims"], errors="coerce")

    sns.scatterplot(
        data=plot_data,
        x=col,
        y="claims",
        color=ZIP_POINT_COLOR,
        alpha=0.35,
        edgecolor=None,
        s=28,
        ax=ax
    )

    ax.set_title(covariate_labels[col])
    ax.set_xlabel("Covariate value")
    ax.set_ylabel("Number of claims" if col == continuous_cols[0] else "")
    zip_clean_axis(ax)

plt.tight_layout()
plt.show()
plt.close(fig)
```
:::

::: panel-tabset
## **`R` Output**

```{r fig-claims-vs-continuous-r}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Claims versus continuous covariates in the canonical training data using R."
#| eval: !expr zip_has_train()

continuous_data <- train_data |>
  select(claims, driver_age, annual_mileage, vehicle_age) |>
  pivot_longer(
    cols = c(driver_age, annual_mileage, vehicle_age),
    names_to = "covariate",
    values_to = "value"
  )

ggplot(continuous_data, aes(x = value, y = claims)) +
  geom_point(alpha = 0.35) +
  facet_wrap(~ covariate, scales = "free_x") +
  labs(
    x = "Covariate value",
    y = "Number of claims"
  ) +
  zip_theme
```

## **`Python` Output**

```{python fig-claims-vs-continuous-py}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Claims versus continuous covariates in the canonical training data using Python."
#| eval: !expr reticulate::py_eval("'training_data' in globals() and len(training_data) > 0 and response_var in training_data.columns", convert = TRUE)

continuous_cols = ["driver_age", "annual_mileage", "vehicle_age"]
covariate_labels = {
    "driver_age": "Driver age",
    "annual_mileage": "Annual mileage",
    "vehicle_age": "Vehicle age",
}

plt.close("all")

fig, axes = plt.subplots(
    nrows=1,
    ncols=len(continuous_cols),
    figsize=(14, 4.5),
    sharey=True
)

for ax, col in zip(axes, continuous_cols):
    plot_data = training_data[[col, "claims"]].copy()
    plot_data[col] = pd.to_numeric(plot_data[col], errors="coerce")
    plot_data["claims"] = pd.to_numeric(plot_data["claims"], errors="coerce")

    sns.scatterplot(
        data=plot_data,
        x=col,
        y="claims",
        color=ZIP_POINT_COLOR,
        alpha=0.35,
        edgecolor=None,
        s=28,
        ax=ax
    )

    ax.set_title(covariate_labels[col])
    ax.set_xlabel("Covariate value")
    ax.set_ylabel("Number of claims" if col == continuous_cols[0] else "")
    zip_clean_axis(ax)

plt.tight_layout()
plt.show()
plt.close(fig)
```
:::

As shown in @fig-claims-vs-continuous-r and @fig-claims-vs-continuous-py, annual mileage displays the clearest positive marginal pattern: higher-mileage policyholders are more likely to appear among the larger claim counts. Vehicle age shows a weaker positive pattern, while driver age has a more modest relationship.

Binned summaries make these patterns easier to read.

::: panel-tabset
## **`R` Code**

``` r
make_binned_summary <- function(df, variable, bins = 6) {
  x <- df[[variable]]
  bin_breaks <- seq(
    min(x, na.rm = TRUE),
    max(x, na.rm = TRUE),
    length.out = bins + 1
  )

  df |>
    mutate(
      variable = .env$variable,
      bin_id = cut(
        .data[[variable]],
        breaks = bin_breaks,
        include.lowest = TRUE,
        labels = FALSE
      ),
      bin = cut(
        .data[[variable]],
        breaks = bin_breaks,
        include.lowest = TRUE,
        dig.lab = 6
      ),
      bin_lower = bin_breaks[bin_id],
      bin_upper = bin_breaks[bin_id + 1],
      bin_midpoint = (bin_lower + bin_upper) / 2,
      bin_width = bin_upper - bin_lower
    ) |>
    filter(!is.na(bin_id)) |>
    group_by(
      variable,
      bin_id,
      bin,
      bin_lower,
      bin_upper,
      bin_midpoint,
      bin_width
    ) |>
    summarise(
      n = n(),
      mean_claims = mean(claims, na.rm = TRUE),
      zero_rate = mean(claims == 0, na.rm = TRUE),
      .groups = "drop"
    ) |>
    arrange(variable, bin_midpoint)
}

binned_summary <- bind_rows(
  make_binned_summary(train_data, "driver_age"),
  make_binned_summary(train_data, "annual_mileage"),
  make_binned_summary(train_data, "vehicle_age")
)

binned_summary |>
  select(variable, bin, bin_midpoint, n, mean_claims, zero_rate) |>
  mutate(bin = as.character(bin)) |>
  fmt_table() |>
    print_as_dt()
```

## **`Python` Code**

``` python
continuous_cols = ["driver_age", "annual_mileage", "vehicle_age"]
binned_summaries = []

def fmt_axis_value(x):
    if abs(x) >= 1000:
        return f"{x:,.0f}"
    if abs(x - round(x)) < 1e-8:
        return f"{x:.0f}"
    return f"{x:.1f}"

def fmt_interval(left, right, first=False):
    left_bracket = "[" if first else "("
    return f"{left_bracket}{fmt_axis_value(left)}, {fmt_axis_value(right)}]"

for col in continuous_cols:
    temp = training_data.copy()
    temp[col] = pd.to_numeric(temp[col], errors="coerce")
    temp["claims"] = pd.to_numeric(temp["claims"], errors="coerce")

    bin_edges = np.linspace(temp[col].min(), temp[col].max(), 7)

    temp["bin_id"] = (
        pd.cut(
            temp[col],
            bins=bin_edges,
            include_lowest=True,
            labels=False
        )
        .astype("Int64") + 1
    )

    summary = (
        temp.dropna(subset=["bin_id"])
        .groupby("bin_id", observed=True)
        .agg(
            n=("claims", "size"),
            mean_claims=("claims", "mean"),
            zero_rate=("claims", lambda x: (x == 0).mean())
        )
        .reset_index()
    )

    summary.insert(0, "variable", col)
    summary["bin_lower"] = summary["bin_id"].apply(lambda i: bin_edges[int(i) - 1])
    summary["bin_upper"] = summary["bin_id"].apply(lambda i: bin_edges[int(i)])
    summary["bin_midpoint"] = (summary["bin_lower"] + summary["bin_upper"]) / 2
    summary["bin_width"] = summary["bin_upper"] - summary["bin_lower"]

    summary["bin"] = summary.apply(
        lambda row: fmt_interval(
            row["bin_lower"],
            row["bin_upper"],
            first=int(row["bin_id"]) == 1
        ),
        axis=1
    )

    binned_summaries.append(summary)

binned_summary = pd.concat(binned_summaries, ignore_index=True)
```
:::

::: panel-tabset
## **`R` Output**

```{r tbl-binned-summary-r}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Mean claims and zero rates by binned continuous covariates in the canonical training data using R."

if (exists("binned_summary")) {
  binned_summary |>
    select(variable, bin, bin_midpoint, n, mean_claims, zero_rate) |>
    mutate(bin = as.character(bin)) |>
    fmt_table() |>
      print_as_dt()
} else {
  tibble(Message = "Binned summary is unavailable until the training data are available.") |>
      print_as_dt()
}
```

## **`Python` Output**

```{python binned-summary-py-run}
#| echo: false
#| results: hide

continuous_cols = ["driver_age", "annual_mileage", "vehicle_age"]
binned_summaries = []

def fmt_axis_value(x):
    if abs(x) >= 1000:
        return f"{x:,.0f}"
    if abs(x - round(x)) < 1e-8:
        return f"{x:.0f}"
    return f"{x:.1f}"

def fmt_interval(left, right, first=False):
    left_bracket = "[" if first else "("
    return f"{left_bracket}{fmt_axis_value(left)}, {fmt_axis_value(right)}]"

for col in continuous_cols:
    temp = training_data.copy()
    temp[col] = pd.to_numeric(temp[col], errors="coerce")
    temp["claims"] = pd.to_numeric(temp["claims"], errors="coerce")

    bin_edges = np.linspace(temp[col].min(), temp[col].max(), 7)

    temp["bin_id"] = (
        pd.cut(
            temp[col],
            bins=bin_edges,
            include_lowest=True,
            labels=False
        )
        .astype("Int64") + 1
    )

    summary = (
        temp.dropna(subset=["bin_id"])
        .groupby("bin_id", observed=True)
        .agg(
            n=("claims", "size"),
            mean_claims=("claims", "mean"),
            zero_rate=("claims", lambda x: (x == 0).mean())
        )
        .reset_index()
    )

    summary.insert(0, "variable", col)
    summary["bin_lower"] = summary["bin_id"].apply(lambda i: bin_edges[int(i) - 1])
    summary["bin_upper"] = summary["bin_id"].apply(lambda i: bin_edges[int(i)])
    summary["bin_midpoint"] = (summary["bin_lower"] + summary["bin_upper"]) / 2
    summary["bin_width"] = summary["bin_upper"] - summary["bin_lower"]

    summary["bin"] = summary.apply(
        lambda row: fmt_interval(
            row["bin_lower"],
            row["bin_upper"],
            first=int(row["bin_id"]) == 1
        ),
        axis=1
    )

    binned_summaries.append(summary)

binned_summary_py = pd.concat(binned_summaries, ignore_index=True)
```

```{r tbl-binned-summary-py}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Mean claims and zero rates by binned continuous covariates in the canonical training data using Python."

if (reticulate::py_eval("'binned_summary_py' in globals()", convert = TRUE)) {
  binned_summary_py_r <- py_df_to_tibble("binned_summary_py")

  if (nrow(binned_summary_py_r) > 0) {
    binned_summary_py_r |>
      select(variable, bin, bin_midpoint, n, mean_claims, zero_rate) |>
      fmt_table() |>
        print_as_dt()
  } else {
    tibble(Message = "Binned summary is empty.") |>
        print_as_dt()
  }
} else {
  tibble(Message = "Binned summary is unavailable until the training data are available.") |>
      print_as_dt()
}
```
:::

The binned EDA indicates that annual mileage is the most pronounced continuous covariate. In the training data, the mean number of claims rises from about **`r zip_stat("annual_low_mean")`** in the lowest mileage bin to about **`r zip_stat("annual_high_mean")`** in the highest mileage bin, while the zero-claim rate falls from about **`r zip_stat("annual_low_zero")`** to about **`r zip_stat("annual_high_zero")`**. This pattern supports including `annual_mileage` in both the count and zero-inflation components of the model.

The following plot visualizes the same binned summaries from @tbl-binned-summary-r, making it easier to compare the direction of the mean-count and zero-rate patterns across the continuous covariates.

::: panel-tabset
## **`R` Code**

``` r
binned_plot_data <- binned_summary |>
  mutate(
    variable = factor(
      variable,
      levels = c("driver_age", "annual_mileage", "vehicle_age"),
      labels = c("Driver age", "Annual mileage", "Vehicle age")
    )
  ) |>
  pivot_longer(
    cols = c(mean_claims, zero_rate),
    names_to = "metric",
    values_to = "value"
  ) |>
  mutate(
    metric = recode(
      metric,
      mean_claims = "Mean claims",
      zero_rate = "Zero-claim rate"
    ),
    metric = factor(metric, levels = c("Mean claims", "Zero-claim rate"))
  )

midpoint_breaks <- sort(unique(binned_plot_data$bin_midpoint))

ggplot(binned_plot_data) +
  geom_rect(
    aes(
      xmin = bin_lower,
      xmax = bin_upper,
      ymin = 0,
      ymax = value
    ),
    fill = "grey35",
    color = "white",
    linewidth = 0.25
  ) +
  facet_grid(metric ~ variable, scales = "free") +
  scale_x_continuous(
    breaks = function(lims) {
      midpoint_breaks[midpoint_breaks >= lims[1] & midpoint_breaks <= lims[2]]
    },
    labels = function(x) {
      format(signif(x, 4), big.mark = ",", scientific = FALSE, trim = TRUE)
    },
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  labs(
    x = "Class midpoint",
    y = "Training-set summary"
  ) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14, vjust = 0.5),
    strip.text = element_text(size = 12),
    legend.position = "bottom",
    panel.spacing.x = unit(0.45, "lines"),
    panel.spacing.y = unit(0.45, "lines")
  )
```

## **`Python` Code**

``` python
plt.close("all")

binned_plot_data = binned_summary.copy()

variable_order = ["driver_age", "annual_mileage", "vehicle_age"]
variable_labels = {
    "driver_age": "Driver age",
    "annual_mileage": "Annual mileage",
    "vehicle_age": "Vehicle age",
}
metric_order = ["Mean claims", "Zero-claim rate"]

binned_plot_data = binned_plot_data.melt(
    id_vars=[
        "variable",
        "bin",
        "bin_id",
        "bin_lower",
        "bin_upper",
        "bin_midpoint",
        "bin_width"
    ],
    value_vars=["mean_claims", "zero_rate"],
    var_name="metric",
    value_name="value"
)

binned_plot_data["metric"] = binned_plot_data["metric"].map({
    "mean_claims": "Mean claims",
    "zero_rate": "Zero-claim rate"
})

fig, axes = plt.subplots(
    nrows=len(metric_order),
    ncols=len(variable_order),
    figsize=ZIP_WIDE_FIGSIZE,
    sharey=False
)

for row_idx, metric in enumerate(metric_order):
    for col_idx, variable in enumerate(variable_order):
        ax = axes[row_idx, col_idx]

        plot_data = (
            binned_plot_data[
                (binned_plot_data["metric"] == metric)
                & (binned_plot_data["variable"] == variable)
            ]
            .sort_values("bin_midpoint")
            .copy()
        )

        ax.bar(
            x=plot_data["bin_midpoint"],
            height=plot_data["value"],
            width=plot_data["bin_width"] * 0.95,
            align="center",
            color=ZIP_BAR_COLOR,
            edgecolor="white",
            linewidth=0.25
        )

        ax.set_xticks(plot_data["bin_midpoint"])
        ax.set_xticklabels(
            [fmt_axis_value(x) for x in plot_data["bin_midpoint"]],
            rotation=45,
            ha="right"
        )

        ax.set_xlim(
            plot_data["bin_lower"].min(),
            plot_data["bin_upper"].max()
        )

        ax.set_title(f"{metric} | {variable_labels[variable]}")
        ax.set_xlabel("Class midpoint" if row_idx == len(metric_order) - 1 else "")
        ax.set_ylabel("Training-set summary" if col_idx == 0 else "")

        zip_clean_axis(ax)

plt.subplots_adjust(wspace=0.22, hspace=0.55)
plt.tight_layout()
plt.show()
plt.close(fig)
```
:::

::: panel-tabset
## **`R` Output**

```{r fig-binned-trends-r}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Binned mean claims and zero-claim rates for continuous covariates in the canonical training data using R."
#| fig.width: 14
#| fig.height: 8
#| eval: !expr exists("binned_summary") && nrow(binned_summary) > 0

binned_plot_data <- binned_summary |>
  mutate(
    variable = factor(
      variable,
      levels = c("driver_age", "annual_mileage", "vehicle_age"),
      labels = c("Driver age", "Annual mileage", "Vehicle age")
    )
  ) |>
  pivot_longer(
    cols = c(mean_claims, zero_rate),
    names_to = "metric",
    values_to = "value"
  ) |>
  mutate(
    metric = recode(
      metric,
      mean_claims = "Mean claims",
      zero_rate = "Zero-claim rate"
    ),
    metric = factor(metric, levels = c("Mean claims", "Zero-claim rate"))
  )

midpoint_breaks <- sort(unique(binned_plot_data$bin_midpoint))

ggplot(binned_plot_data) +
  geom_rect(
    aes(
      xmin = bin_lower,
      xmax = bin_upper,
      ymin = 0,
      ymax = value
    ),
    fill = "grey35",
    color = "white",
    linewidth = 0.25
  ) +
  facet_grid(metric ~ variable, scales = "free") +
  scale_x_continuous(
    breaks = function(lims) {
      midpoint_breaks[midpoint_breaks >= lims[1] & midpoint_breaks <= lims[2]]
    },
    labels = function(x) {
      format(signif(x, 4), big.mark = ",", scientific = FALSE, trim = TRUE)
    },
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  labs(
    x = "Class midpoint",
    y = "Training-set summary"
  ) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14, vjust = 0.5),
    strip.text = element_text(size = 12),
    legend.position = "bottom",
    panel.spacing.x = unit(0.45, "lines"),
    panel.spacing.y = unit(0.45, "lines")
  )
```

## **`Python` Output**

```{python fig-binned-trends-py}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Binned mean claims and zero-claim rates for continuous covariates in the canonical training data using Python."
#| eval: !expr reticulate::py_eval("'binned_summary_py' in globals() and len(binned_summary_py) > 0", convert = TRUE)

plt.close("all")

binned_plot_data = binned_summary_py.copy()

variable_order = ["driver_age", "annual_mileage", "vehicle_age"]
variable_labels = {
    "driver_age": "Driver age",
    "annual_mileage": "Annual mileage",
    "vehicle_age": "Vehicle age",
}
metric_order = ["Mean claims", "Zero-claim rate"]

binned_plot_data = binned_plot_data.melt(
    id_vars=[
        "variable",
        "bin",
        "bin_id",
        "bin_lower",
        "bin_upper",
        "bin_midpoint",
        "bin_width"
    ],
    value_vars=["mean_claims", "zero_rate"],
    var_name="metric",
    value_name="value"
)

binned_plot_data["metric"] = binned_plot_data["metric"].map({
    "mean_claims": "Mean claims",
    "zero_rate": "Zero-claim rate"
})

fig, axes = plt.subplots(
    nrows=len(metric_order),
    ncols=len(variable_order),
    figsize=ZIP_WIDE_FIGSIZE,
    sharey=False
)

for row_idx, metric in enumerate(metric_order):
    for col_idx, variable in enumerate(variable_order):
        ax = axes[row_idx, col_idx]

        plot_data = (
            binned_plot_data[
                (binned_plot_data["metric"] == metric)
                & (binned_plot_data["variable"] == variable)
            ]
            .sort_values("bin_midpoint")
            .copy()
        )

        ax.bar(
            x=plot_data["bin_midpoint"],
            height=plot_data["value"],
            width=plot_data["bin_width"] * 0.95,
            align="center",
            color=ZIP_BAR_COLOR,
            edgecolor="white",
            linewidth=0.25
        )

        ax.set_xticks(plot_data["bin_midpoint"])
        ax.set_xticklabels(
            [fmt_axis_value(x) for x in plot_data["bin_midpoint"]],
            rotation=45,
            ha="right"
        )

        ax.set_xlim(
            plot_data["bin_lower"].min(),
            plot_data["bin_upper"].max()
        )

        ax.set_title(f"{metric} | {variable_labels[variable]}")
        ax.set_xlabel("Class midpoint" if row_idx == len(metric_order) - 1 else "")
        ax.set_ylabel("Training-set summary" if col_idx == 0 else "")

        zip_clean_axis(ax)

plt.subplots_adjust(wspace=0.22, hspace=0.55)
plt.tight_layout()
plt.show()
plt.close(fig)
```
:::

### Correlation summary

The final EDA table gives a compact numerical summary of the marginal associations between each covariate, the original count response, and the binary `has_claim` indicator.

::: panel-tabset
## **`R` Code**

``` r
correlation_summary <- train_data |>
  select(driver_age, annual_mileage, vehicle_age, male, urban_area, prior_claims, claims, has_claim) |>
  summarise(across(
    c(driver_age, annual_mileage, vehicle_age, male, urban_area, prior_claims),
    list(
      corr_with_claims = ~ cor(.x, claims),
      corr_with_has_claim = ~ cor(.x, has_claim)
    ),
    .names = "{.col}_{.fn}"
  )) |>
  pivot_longer(
    cols = everything(),
    names_to = c("variable", ".value"),
    names_pattern = "(.*)_(corr_with_.*)"
  )

correlation_summary |>
  fmt_table() |>
    print_as_dt()
```

## **`Python` Code**

``` python
corr = training_data[[
    "driver_age", "annual_mileage", "vehicle_age", "male",
    "urban_area", "prior_claims", "claims", "has_claim"
]].corr()

correlation_summary = pd.DataFrame({
    "corr_with_claims": corr["claims"].drop(["claims", "has_claim"]),
    "corr_with_has_claim": corr["has_claim"].drop(["claims", "has_claim"])
}).reset_index(names="variable")
```
:::

::: panel-tabset
## **`R` Output**

```{r tbl-correlation-summary-r}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Correlation of covariates with claims and has_claim in the canonical training data using R."

if (exists("correlation_summary")) {
  correlation_summary |>
    fmt_table() |>
      print_as_dt()
} else {
  tibble(Message = "Correlation summary is unavailable until the training data are available.") |>
      print_as_dt()
}
```

## **`Python` Output**

```{python correlation-summary-py-run}
#| echo: false
#| results: hide

if "training_data" in globals() and len(training_data) > 0:
    corr = training_data[[
        "driver_age", "annual_mileage", "vehicle_age", "male",
        "urban_area", "prior_claims", "claims", "has_claim"
    ]].corr()

    correlation_summary_py = pd.DataFrame({
        "corr_with_claims": corr["claims"].drop(["claims", "has_claim"]),
        "corr_with_has_claim": corr["has_claim"].drop(["claims", "has_claim"])
    }).reset_index(names="variable")
```

```{r tbl-correlation-summary-py}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Correlation of covariates with claims and has_claim in the canonical training data using Python."

if (reticulate::py_eval("'correlation_summary_py' in globals()", convert = TRUE)) {
  correlation_summary_py_r <- py_df_to_tibble("correlation_summary_py")

  if (nrow(correlation_summary_py_r) > 0) {
    correlation_summary_py_r |>
      fmt_table() |>
        print_as_dt()
  } else {
    tibble(Message = "Correlation summary is empty.") |>
        print_as_dt()
  }
} else {
  tibble(Message = "Correlation summary is unavailable until the training data are available.") |>
      print_as_dt()
}
```
:::

The correlation summary reinforces the graphical EDA. In the training data, `annual_mileage` has the strongest correlation with `claims`, about **`r zip_stat("corr_annual_claims")`**, and with `has_claim`, about **`r zip_stat("corr_annual_has_claim")`**. `prior_claims` is also notable, with correlations of about **`r zip_stat("corr_prior_claims")`** with `claims` and **`r zip_stat("corr_prior_has_claim")`** with `has_claim`. These correlations do not establish causality, but they help explain why these covariates are sensible choices for the model components.

### EDA summary

The exploratory analysis points to two connected conclusions. First, `claims` is a count response with a large concentration at zero. Second, the covariates do not all show the same marginal relationship with the response, so the later ZIP model should be allowed to separate the zero-generating mechanism from the count-generating mechanism.

| **EDA Theme** | **TLDR** | **Dynamic Evidence** | **Modelling Implication** |
|:------------------:|:-----------------|:-----------------|:-----------------|
| Response shape | The response is heavily concentrated at zero, with a smaller right tail of positive claim counts. | There are **`r zip_stat("zero_n")`** zero-claim observations out of **`r zip_stat("train_n")`** training observations; the largest observed count is **`r zip_stat("max_claims")`**. | A count model is needed, but the zero pile-up should be modelled explicitly. |
| Mean - variance pattern | The response is more variable than a simple equidispersed Poisson pattern would suggest. | The sample mean is **`r zip_stat("sample_mean")`**, the sample variance is **`r zip_stat("sample_variance")`**, and the variance-to-mean ratio is **`r zip_stat("variance_to_mean")`**. | A one-process Classical Poisson model may be too restrictive. |
| Zero inflation check | The observed zero rate is larger than the zero rate implied by a Poisson model with the same mean. | The observed zero rate is **`r zip_stat("observed_zero_rate")`**, while the Poisson-expected zero rate is **`r zip_stat("poisson_zero_rate")`**; the excess zero rate is **`r zip_stat("excess_zero_rate")`**. | This is direct empirical motivation for a zero-inflated model. |
| Zero versus nonzero profiles | Policyholders with at least one claim tend to have higher mileage and a higher prior-claims rate. | Mean annual mileage is **`r zip_stat("mileage_nonzero")`** for nonzero-claim policyholders versus **`r zip_stat("mileage_zero")`** for zero-claim policyholders; the prior-claims rate is **`r zip_stat("prior_rate_nonzero")`** versus **`r zip_stat("prior_rate_zero")`**. | `annual_mileage` and `prior_claims` are plausible regressors for both claim frequency and the zero/nonzero distinction. |
| Binary covariates | `prior_claims` shows the clearest binary-covariate contrast; `urban_area` is weaker in this training split. | With prior claims, mean claims are **`r zip_stat("prior_mean_yes")`** and the zero rate is **`r zip_stat("prior_zero_yes")`**; without prior claims, mean claims are **`r zip_stat("prior_mean_no")`** and the zero rate is **`r zip_stat("prior_zero_no")`**. The urban groups differ by about **`r zip_stat("urban_mean_diff")`** in mean claims and **`r zip_stat("urban_zero_diff")`** in zero rate. | `prior_claims` should be treated as an important candidate covariate; `urban_area` may still be retained for substantive insurance interpretation. |
| Continuous covariates | `annual_mileage` has the clearest positive marginal association with the response. | Mean claims rise from **`r zip_stat("annual_low_mean")`** in the lowest mileage bin to **`r zip_stat("annual_high_mean")`** in the highest mileage bin; the zero rate falls from **`r zip_stat("annual_low_zero")`** to **`r zip_stat("annual_high_zero")`**. Its correlations are **`r zip_stat("corr_annual_claims")`** with `claims` and **`r zip_stat("corr_annual_has_claim")`** with `has_claim`. | `annual_mileage` is a natural candidate for the count component and the zero-inflation component. |

: TLDR summary of the EDA findings for the zero-inflated insurance-claims training data.

## Preparing for Modelling

The data schema confirms that the response is a zero-inflated count and that the available regressors are either continuous, integer-valued, or binary. There are no multi-category nominal covariates in this dataset, so we do not need to create a baseline category plus a collection of dummy variables. The binary covariates are already coded as $0/1$ indicators, and the continuous or integer-valued regressors can enter the model directly unless a later modelling choice calls for centering, scaling, or nonlinear transformations.

For policyholder $i$, define the response as

$$Y_i = \texttt{claims}_i,$$

where $Y_i$ is the number of claims filed during the observation window. The candidate covariates are indexed in the same order as the variable table above, excluding the response.

| **Symbol** | **Coding Name Alias** | **Variable Meaning** | **Type** | **Binary Coding** |
|:-------------:|:-------------:|:--------------|:-------------:|:--------------|
| $x_{1,i}$ | `driver_age` | Age of policyholder $i$, measured in years. | Continuous | Not binary. |
| $x_{2,i}$ | `annual_mileage` | Total kilometers driven per year by policyholder $i$. | Continuous | Not binary. |
| $x_{3,i}$ | `vehicle_age` | Age of the insured vehicle for policyholder $i$, measured in years. | Discrete | Not binary; integer-valued from 0 to 20. |
| $x_{4,i}$ | `male` | Sex indicator for policyholder $i$. | Discrete | $1$ means male; $0$ means female. |
| $x_{5,i}$ | `urban_area` | Primary driving-environment indicator for policyholder $i$. | Discrete | $1$ means urban; $0$ means rural. |
| $x_{6,i}$ | `prior_claims` | Prior-claims-history indicator for policyholder $i$. | Discrete | $1$ means the driver filed claims in previous years; $0$ means no prior claims. |

: Covariate notation for the zero-inflated insurance-claims model.

Using this notation, a compact covariate vector for the full set of candidate regressors is

$$
\mathbf{x}_i = \left(1, x_{1,i}, x_{2,i}, x_{3,i}, x_{4,i}, x_{5,i}, x_{6,i}\right)
$$

Where the leading $1$ represents the intercept. In the ZIP model, this covariate vector can be used to build two linked regression components: a count component for the expected number of claims among policyholders in the count-generating process, and a zero-inflation component for the probability of belonging to the extra-zero group. Later sections can choose whether both components use the same covariates or whether one component uses a smaller, substantively motivated subset.

## Comparing to Other Approaches

Before introducing the full Zero-Inflated Poisson (ZIP) model, it is useful to compare it with two simpler approaches that each captures only part of the modelling problem. The goal of this comparison is not only to fit two baselines, but also to see what each baseline reveals about the insurance-claims data.

First, we can turn the count response into a binary response. Define $H_i = \mathbb{I}(Y_i > 0)$ where $Y_i$ is the number of claims for policyholder $i$. A Binary Logistic Regression model for `has_claim` then describes whether a policyholder filed at least one claim. This is directly connected to the large zero-versus-nonzero split observed in @fig-claims-distribution-r and @tbl-claims-summary-r, but it does not distinguish between one claim and several claims.

Second, we can keep the original count response and fit a Classical Poisson Regression model for `claims`. This model describes the expected number of claims directly, so it keeps the count scale used in @fig-claims-vs-continuous-r and @tbl-binned-summary-r. However, it treats all observations as coming from a single count-generating process. In particular, it ties the zero rate to the same mean parameter that governs the positive counts.

These two models are not replacements for ZIP regression in this chapter. Instead, they give us useful comparison points. Logistic Regression focuses on the zero-versus-nonzero distinction. Classical Poisson Regression focuses on the count scale. ZIP regression combines these ideas by allowing a separate zero-generating component and a count-generating component.


```{r lr-cpr-setup-r, include=FALSE}
if (zip_has_train()) {
  train_data <- train_data |>
    mutate(
      has_claim = as.integer(claims > 0),
      male = as.integer(male),
      urban_area = as.integer(urban_area),
      prior_claims = as.integer(prior_claims)
    )

  zip_model_covariates <- c(
    "driver_age",
    "annual_mileage",
    "vehicle_age",
    "male",
    "urban_area",
    "prior_claims"
  )

  zip_covariate_labels <- c(
    driver_age = "Driver age",
    annual_mileage = "Annual mileage",
    vehicle_age = "Vehicle age",
    male = "Male indicator",
    urban_area = "Urban-area indicator",
    prior_claims = "Prior-claims indicator"
  )

  zip_binary_covariates <- c("male", "urban_area", "prior_claims")
  zip_continuous_covariates <- c("driver_age", "annual_mileage", "vehicle_age")

  zip_mode_binary <- function(x) {
    as.integer(names(sort(table(x), decreasing = TRUE))[1])
  }

  zip_reference_values <- tibble(
    driver_age = median(train_data$driver_age, na.rm = TRUE),
    annual_mileage = median(train_data$annual_mileage, na.rm = TRUE),
    vehicle_age = median(train_data$vehicle_age, na.rm = TRUE),
    male = zip_mode_binary(train_data$male),
    urban_area = zip_mode_binary(train_data$urban_area),
    prior_claims = zip_mode_binary(train_data$prior_claims)
  )

  zip_interpret_scales <- tibble(
    term = zip_model_covariates,
    comparison = c(
      "+10 years of driver age",
      "+10,000 km of annual mileage",
      "+5 years of vehicle age",
      "male (1) versus female (0)",
      "urban (1) versus rural (0)",
      "prior claims (1) versus none (0)"
    ),
    change = c(10, 10000, 5, 1, 1, 1)
  )

  lr_model_r <- glm(
    has_claim ~ driver_age + annual_mileage + vehicle_age + male + urban_area + prior_claims,
    data = train_data,
    family = binomial(link = "logit")
  )

  cpr_model_r <- glm(
    claims ~ driver_age + annual_mileage + vehicle_age + male + urban_area + prior_claims,
    data = train_data,
    family = poisson(link = "log")
  )

  lr_coef_r <- broom::tidy(lr_model_r) |>
    mutate(model = "Binary Logistic Regression", .before = 1)

  cpr_coef_r <- broom::tidy(cpr_model_r) |>
    mutate(model = "Classical Poisson Regression", .before = 1)

  make_scaled_effects <- function(model, scale_tbl, effect_name) {
    coef_tbl <- broom::tidy(model)
    ci_tbl <- as.data.frame(confint.default(model, level = 0.95)) |>
      tibble::rownames_to_column("term") |>
      rename(conf.low = `2.5 %`, conf.high = `97.5 %`)

    coef_tbl |>
      left_join(ci_tbl, by = "term") |>
      inner_join(scale_tbl, by = "term") |>
      transmute(
        covariate = term,
        comparison,
        log_scale_change = estimate * change,
        !!effect_name := exp(estimate * change),
        lower_95 = exp(conf.low * change),
        upper_95 = exp(conf.high * change),
        percent_change = 100 * (exp(estimate * change) - 1),
        p_value = p.value
      )
  }

  lr_effects_r <- make_scaled_effects(
    lr_model_r,
    zip_interpret_scales,
    "odds_ratio"
  )

  cpr_effects_r <- make_scaled_effects(
    cpr_model_r,
    zip_interpret_scales,
    "rate_ratio"
  )

  make_reference_grid <- function(df, focal, values) {
    grid <- zip_reference_values[rep(1, length(values)), ]
    grid[[focal]] <- values
    grid
  }

  make_fitted_curve <- function(model, df, focal, outcome_type) {
    if (focal %in% zip_binary_covariates) {
      values <- c(0, 1)
    } else {
      values <- seq(
        min(df[[focal]], na.rm = TRUE),
        max(df[[focal]], na.rm = TRUE),
        length.out = 100
      )
    }

    new_data <- make_reference_grid(df, focal, values)

    tibble(
      covariate = focal,
      covariate_label = unname(zip_covariate_labels[focal]),
      x = values,
      fitted = predict(model, newdata = new_data, type = "response"),
      outcome_type = outcome_type
    )
  }

  make_observed_continuous <- function(df, focal, outcome, bins = 6) {
    breaks <- seq(
      min(df[[focal]], na.rm = TRUE),
      max(df[[focal]], na.rm = TRUE),
      length.out = bins + 1
    )

    df |>
      mutate(
        bin_id = cut(
          .data[[focal]],
          breaks = breaks,
          include.lowest = TRUE,
          labels = FALSE
        )
      ) |>
      filter(!is.na(bin_id)) |>
      group_by(bin_id) |>
      summarise(
        x = mean(.data[[focal]], na.rm = TRUE),
        observed = mean(.data[[outcome]], na.rm = TRUE),
        n = n(),
        .groups = "drop"
      ) |>
      mutate(
        covariate = focal,
        covariate_label = unname(zip_covariate_labels[focal])
      )
  }

  make_observed_binary <- function(df, focal, outcome) {
    df |>
      group_by(x = .data[[focal]]) |>
      summarise(
        observed = mean(.data[[outcome]], na.rm = TRUE),
        n = n(),
        .groups = "drop"
      ) |>
      mutate(
        covariate = focal,
        covariate_label = unname(zip_covariate_labels[focal])
      )
  }

  lr_fitted_r <- purrr::map_dfr(
    zip_model_covariates,
    ~ make_fitted_curve(lr_model_r, train_data, .x, "Fitted probability")
  )

  cpr_fitted_r <- purrr::map_dfr(
    zip_model_covariates,
    ~ make_fitted_curve(cpr_model_r, train_data, .x, "Fitted mean")
  )

  lr_observed_r <- bind_rows(
    purrr::map_dfr(
      zip_continuous_covariates,
      ~ make_observed_continuous(train_data, .x, "has_claim")
    ),
    purrr::map_dfr(
      zip_binary_covariates,
      ~ make_observed_binary(train_data, .x, "has_claim")
    )
  )

  cpr_observed_r <- bind_rows(
    purrr::map_dfr(
      zip_continuous_covariates,
      ~ make_observed_continuous(train_data, .x, "claims")
    ),
    purrr::map_dfr(
      zip_binary_covariates,
      ~ make_observed_binary(train_data, .x, "claims")
    )
  )
}
```

```{python lr-cpr-setup-py, include=FALSE}
if "training_data" in globals() and len(training_data) > 0:
    training_data = training_data.copy()

    zip_model_covariates_py = [
        "driver_age",
        "annual_mileage",
        "vehicle_age",
        "male",
        "urban_area",
        "prior_claims",
    ]

    # Create the binary response explicitly.
    training_data["claims"] = pd.to_numeric(
        training_data["claims"],
        errors="coerce"
    )
    training_data["has_claim"] = (training_data["claims"] > 0).astype(int)

    # Force all model columns to ordinary numeric values.
    model_cols = zip_model_covariates_py + ["claims", "has_claim"]

    for col in model_cols:
        training_data[col] = pd.to_numeric(
            training_data[col],
            errors="coerce"
        )

    training_data = training_data.dropna(subset=model_cols).copy()

    # IMPORTANT:
    # statsmodels can fail when pandas nullable Int64 columns are passed through
    # reticulate, because they may arrive as object dtype.
    # Use ordinary float arrays/data frames for the design matrix.
    X_train_zip = sm.add_constant(
        training_data[zip_model_covariates_py].astype(float),
        has_constant="add"
    )

    H_train_zip = training_data["has_claim"].astype(int)
    Y_train_zip = training_data["claims"].astype(float)

    lr_model_py = sm.Logit(H_train_zip, X_train_zip).fit(disp=False)

    cpr_model_py = sm.GLM(
        Y_train_zip,
        X_train_zip,
        family=sm.families.Poisson()
    ).fit()

    def model_table(model, model_name):
        ci = pd.DataFrame(
            np.asarray(model.conf_int()),
            index=model.params.index,
            columns=["conf_low", "conf_high"]
        )

        out = pd.DataFrame({
            "model": model_name,
            "term": model.params.index,
            "estimate": model.params.values,
            "std.error": model.bse.values,
            "statistic": model.tvalues.values,
            "p.value": model.pvalues.values,
            "conf.low": ci["conf_low"].values,
            "conf.high": ci["conf_high"].values,
        })

        return out

    lr_coef_py = model_table(
        lr_model_py,
        "Binary Logistic Regression"
    )

    cpr_coef_py = model_table(
        cpr_model_py,
        "Classical Poisson Regression"
    )

    scale_rows_py = pd.DataFrame({
        "term": zip_model_covariates_py,
        "comparison": [
            "+10 years of driver age",
            "+10,000 km of annual mileage",
            "+5 years of vehicle age",
            "male ($1$) versus female ($0$)",
            "urban ($1$) versus rural ($0$)",
            "prior claims ($1$) versus none ($0$)",
        ],
        "change": [10, 10000, 5, 1, 1, 1],
    })

    def scaled_effects(model, effect_name):
        ci = pd.DataFrame(
            np.asarray(model.conf_int()),
            index=model.params.index,
            columns=["conf_low", "conf_high"]
        )

        coef = pd.DataFrame({
            "term": model.params.index,
            "estimate": model.params.values,
            "conf.low": ci["conf_low"].values,
            "conf.high": ci["conf_high"].values,
            "p_value": model.pvalues.values,
        })

        out = coef.merge(scale_rows_py, on="term", how="inner")
        out["log_scale_change"] = out["estimate"] * out["change"]
        out[effect_name] = np.exp(out["log_scale_change"])
        out["lower_95"] = np.exp(out["conf.low"] * out["change"])
        out["upper_95"] = np.exp(out["conf.high"] * out["change"])
        out["percent_change"] = 100 * (out[effect_name] - 1)

        return out[[
            "term",
            "comparison",
            "log_scale_change",
            effect_name,
            "lower_95",
            "upper_95",
            "percent_change",
            "p_value",
        ]].rename(columns={"term": "covariate"})

    lr_effects_py = scaled_effects(lr_model_py, "odds_ratio")
    cpr_effects_py = scaled_effects(cpr_model_py, "rate_ratio")
```


```{r lr-cpr-insight-stats-r, include=FALSE}
# Dynamic quantities used to interpret the Logistic and Classical Poisson
# comparison models without hardcoding rendered values in the prose.
lr_cpr_stat <- function(name, digits = 4) {
  if (exists("lr_cpr_stats") && name %in% names(lr_cpr_stats)) {
    fmt_sig(lr_cpr_stats[[name]], digits = digits)
  } else {
    "NA"
  }
}

lr_cpr_pct <- function(name, digits = 4) {
  if (exists("lr_cpr_stats") && name %in% names(lr_cpr_stats)) {
    paste0(fmt_sig(100 * lr_cpr_stats[[name]], digits = digits), "%")
  } else {
    "NA"
  }
}

pull_effect_value <- function(effect_tbl, covariate_name, column_name) {
  effect_tbl |>
    filter(covariate == covariate_name) |>
    pull(.data[[column_name]]) |>
    first()
}

pull_fitted_endpoint <- function(fitted_tbl, covariate_name, which_end = c("low", "high")) {
  which_end <- match.arg(which_end)

  out <- fitted_tbl |>
    filter(covariate == covariate_name) |>
    arrange(x)

  if (which_end == "low") {
    out |> slice(1) |> pull(fitted) |> first()
  } else {
    out |> slice(n()) |> pull(fitted) |> first()
  }
}

if (
  zip_has_train() &&
  exists("lr_effects_r") &&
  exists("cpr_effects_r") &&
  exists("lr_fitted_r") &&
  exists("cpr_fitted_r") &&
  exists("cpr_model_r")
) {
  cpr_fitted_mu_r <- predict(cpr_model_r, type = "response")
  cpr_fitted_zero_prob_r <- exp(-cpr_fitted_mu_r)

  cpr_zero_check_r <- tibble(
    Summary = c(
      "Observed zero rate",
      "Average fitted Poisson zero probability",
      "Observed minus fitted zero rate",
      "Observed number of zeros",
      "Expected number of zeros under fitted Poisson",
      "Pearson dispersion ratio"
    ),
    Value = c(
      mean(train_data$claims == 0, na.rm = TRUE),
      mean(cpr_fitted_zero_prob_r, na.rm = TRUE),
      mean(train_data$claims == 0, na.rm = TRUE) - mean(cpr_fitted_zero_prob_r, na.rm = TRUE),
      sum(train_data$claims == 0, na.rm = TRUE),
      sum(cpr_fitted_zero_prob_r, na.rm = TRUE),
      sum(residuals(cpr_model_r, type = "pearson")^2, na.rm = TRUE) / df.residual(cpr_model_r)
    )
  )

  lr_cpr_stats <- list(
    lr_driver_or = pull_effect_value(lr_effects_r, "driver_age", "odds_ratio"),
    lr_driver_p = pull_effect_value(lr_effects_r, "driver_age", "p_value"),
    lr_annual_or = pull_effect_value(lr_effects_r, "annual_mileage", "odds_ratio"),
    lr_annual_p = pull_effect_value(lr_effects_r, "annual_mileage", "p_value"),
    lr_vehicle_or = pull_effect_value(lr_effects_r, "vehicle_age", "odds_ratio"),
    lr_vehicle_p = pull_effect_value(lr_effects_r, "vehicle_age", "p_value"),
    lr_male_or = pull_effect_value(lr_effects_r, "male", "odds_ratio"),
    lr_urban_or = pull_effect_value(lr_effects_r, "urban_area", "odds_ratio"),
    lr_prior_or = pull_effect_value(lr_effects_r, "prior_claims", "odds_ratio"),
    lr_prior_p = pull_effect_value(lr_effects_r, "prior_claims", "p_value"),
    lr_annual_prob_low = pull_fitted_endpoint(lr_fitted_r, "annual_mileage", "low"),
    lr_annual_prob_high = pull_fitted_endpoint(lr_fitted_r, "annual_mileage", "high"),
    lr_prior_prob_no = pull_fitted_endpoint(lr_fitted_r, "prior_claims", "low"),
    lr_prior_prob_yes = pull_fitted_endpoint(lr_fitted_r, "prior_claims", "high"),
    positive_mean_claims = mean(train_data$claims[train_data$claims > 0], na.rm = TRUE),
    multiple_claim_rate_given_positive = mean(train_data$claims[train_data$claims > 0] >= 2, na.rm = TRUE),
    positive_max_claims = max(train_data$claims[train_data$claims > 0], na.rm = TRUE),
    cpr_driver_rr = pull_effect_value(cpr_effects_r, "driver_age", "rate_ratio"),
    cpr_driver_p = pull_effect_value(cpr_effects_r, "driver_age", "p_value"),
    cpr_annual_rr = pull_effect_value(cpr_effects_r, "annual_mileage", "rate_ratio"),
    cpr_annual_p = pull_effect_value(cpr_effects_r, "annual_mileage", "p_value"),
    cpr_vehicle_rr = pull_effect_value(cpr_effects_r, "vehicle_age", "rate_ratio"),
    cpr_vehicle_p = pull_effect_value(cpr_effects_r, "vehicle_age", "p_value"),
    cpr_male_rr = pull_effect_value(cpr_effects_r, "male", "rate_ratio"),
    cpr_male_p = pull_effect_value(cpr_effects_r, "male", "p_value"),
    cpr_urban_rr = pull_effect_value(cpr_effects_r, "urban_area", "rate_ratio"),
    cpr_urban_p = pull_effect_value(cpr_effects_r, "urban_area", "p_value"),
    cpr_prior_rr = pull_effect_value(cpr_effects_r, "prior_claims", "rate_ratio"),
    cpr_prior_p = pull_effect_value(cpr_effects_r, "prior_claims", "p_value"),
    cpr_annual_mean_low = pull_fitted_endpoint(cpr_fitted_r, "annual_mileage", "low"),
    cpr_annual_mean_high = pull_fitted_endpoint(cpr_fitted_r, "annual_mileage", "high"),
    cpr_vehicle_mean_low = pull_fitted_endpoint(cpr_fitted_r, "vehicle_age", "low"),
    cpr_vehicle_mean_high = pull_fitted_endpoint(cpr_fitted_r, "vehicle_age", "high"),
    cpr_prior_mean_no = pull_fitted_endpoint(cpr_fitted_r, "prior_claims", "low"),
    cpr_prior_mean_yes = pull_fitted_endpoint(cpr_fitted_r, "prior_claims", "high"),
    cpr_fitted_zero_rate = mean(cpr_fitted_zero_prob_r, na.rm = TRUE),
    cpr_expected_zero_n = sum(cpr_fitted_zero_prob_r, na.rm = TRUE),
    cpr_zero_gap = mean(train_data$claims == 0, na.rm = TRUE) - mean(cpr_fitted_zero_prob_r, na.rm = TRUE),
    cpr_pearson_dispersion = sum(residuals(cpr_model_r, type = "pearson")^2, na.rm = TRUE) / df.residual(cpr_model_r),
    observed_zero_rate = mean(train_data$claims == 0, na.rm = TRUE),
    observed_zero_n = sum(train_data$claims == 0, na.rm = TRUE)
  )
}
```

### Binary Logistic Regression for `has_claim`

#### Model definition

The Logistic Regression alternative uses the derived binary response in @eq-zip-lr-binary-response.

$$
H_i =
\begin{cases}
1, & \text{if policyholder } i \text{ filed at least one claim}, \\
0, & \text{if policyholder } i \text{ filed no claims}.
\end{cases}
$${#eq-zip-lr-binary-response}

The probabilistic assumption is given in @eq-zip-lr-bernoulli, where $\pi_i = P(H_i = 1 \mid \mathbf{x}_i)$ is the probability that policyholder $i$ filed at least one claim.

$$
H_i \mid \mathbf{x}_i \sim \operatorname{Bernoulli}(\pi_i)
$${#eq-zip-lr-bernoulli}

The link function is the logit link in @eq-zip-lr-logit-link.

$$
\operatorname{logit}(\pi_i)
=
\log\left(\frac{\pi_i}{1 - \pi_i}\right)
$${#eq-zip-lr-logit-link}

Using the six covariates from the running example, the systematic component is shown in @eq-zip-lr-systematic.

$$
\log\left(\frac{\pi_i}{1 - \pi_i}\right)
=
\alpha_0
+
\alpha_1 x_{1,i}
+
\alpha_2 x_{2,i}
+
\alpha_3 x_{3,i}
+
\alpha_4 x_{4,i}
+
\alpha_5 x_{5,i}
+
\alpha_6 x_{6,i}
$${#eq-zip-lr-systematic}

In @eq-zip-lr-systematic, there are $7$ coefficients, $\alpha_0, \alpha_1, \ldots, \alpha_6$, which change the log-odds of having at least one claim. The intercept $\alpha_0$ is the baseline log-odds when all covariates are equal to zero, while $\alpha_j$ for $j \in \{1, 2, \ldots, 6\}$ represents the change in log-odds for a one-unit increase in the corresponding regressor $x_{j,i}$, holding the other covariates fixed.

In this formulation, the model describes the probability of at least one claim. It intentionally ignores the difference between one claim and multiple claims.

::::: Heads-up
::: Heads-up-header
Heads-up on the connection to Chapter 8!
:::

::: Heads-up-container
The logit link, Bernoulli assumption, and odds-ratio interpretation are introduced in @sec-binary-logistic. Here we use them only as a comparison point for the ZIP model, not as a full reintroduction of Binary Logistic Regression.
:::
:::::

#### Fitting the model

For this comparison, we fit the model using the canonical training data created earlier in the chapter. The binary response `has_claim` is created explicitly in both languages so that the transformation from `claims` to $H_i$ is visible.

::: panel-tabset
## **`R` Code**

``` r
train_data <- train_data |>
  mutate(has_claim = as.integer(claims > 0))

lr_model_r <- glm(
  has_claim ~ driver_age + annual_mileage + vehicle_age + male + urban_area + prior_claims,
  data = train_data,
  family = binomial(link = "logit")
)

broom::tidy(lr_model_r)
```

## **`Python` Code**

``` python
training_data = training_data.copy()
training_data["has_claim"] = (training_data["claims"] > 0).astype(int)

covariates = [
    "driver_age",
    "annual_mileage",
    "vehicle_age",
    "male",
    "urban_area",
    "prior_claims",
]

X = sm.add_constant(training_data[covariates], has_constant="add")
H = training_data["has_claim"]

lr_model_py = sm.Logit(H, X).fit(disp=False)

lr_coef_py = pd.DataFrame({
    "term": lr_model_py.params.index,
    "estimate": lr_model_py.params.values,
    "std.error": lr_model_py.bse.values,
    "statistic": lr_model_py.tvalues.values,
    "p.value": lr_model_py.pvalues.values,
})
```
:::

::: panel-tabset
## **`R` Output**

```{r tbl-lr-fit-r}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Binary Logistic Regression coefficient table for the canonical training data using R."

if (exists("lr_coef_r")) {
  lr_coef_r |>
    fmt_table() |>
      print_as_dt()
} else {
  tibble(Message = "Binary Logistic Regression output is unavailable until the training data are available.") |>
      print_as_dt()
}
```

## **`Python` Output**

```{python lr-fit-py-run}
#| echo: false
#| results: hide
#| message: false
#| warning: false

import pandas as pd
import statsmodels.api as sm

# Use the canonical R split if the Python object is not already available.
if "training_data" not in globals():
    training_data = r.train_data.copy()

training_data = training_data.copy()
training_data["has_claim"] = (training_data["claims"] > 0).astype(int)

covariates = [
    "driver_age",
    "annual_mileage",
    "vehicle_age",
    "male",
    "urban_area",
    "prior_claims",
]

for col in covariates + ["claims", "has_claim"]:
    training_data[col] = pd.to_numeric(training_data[col], errors="coerce")

training_data = training_data.dropna(subset=covariates + ["has_claim"]).copy()

X = sm.add_constant(training_data[covariates].astype(float), has_constant="add")
H = training_data["has_claim"].astype(int)

lr_model_py = sm.Logit(H, X).fit(disp=False)

lr_coef_py = pd.DataFrame({
    "term": lr_model_py.params.index,
    "estimate": lr_model_py.params.values,
    "std.error": lr_model_py.bse.values,
    "statistic": lr_model_py.tvalues.values,
    "p.value": lr_model_py.pvalues.values,
})
```

```{r tbl-lr-fit-py}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Binary Logistic Regression coefficient table for the canonical training data using Python."

if (reticulate::py_eval("'lr_coef_py' in globals()", convert = TRUE)) {
  py_df_to_tibble("lr_coef_py") |>
    fmt_table() |>
      print_as_dt()
} else {
  tibble(Message = "Binary Logistic Regression output is unavailable until the training data are available.") |>
      print_as_dt()
}
```
:::

The two outputs may not be formatted identically. `R` and `Python` use different summary conventions, and their numerical optimizers can differ in small implementation details. Because both chunks use the same canonical training split, the fitted coefficients should nevertheless be within similar ranges.

The coefficient table in @tbl-lr-fit-r already points to the main event-probability story. The strongest positive fitted associations with `has_claim` are attached to `annual_mileage` and `prior_claims`, matching the EDA pattern in which higher-mileage policyholders and policyholders with prior claims were less concentrated at zero. `driver_age` is also positive in this binary model, but the fitted evidence is weaker than for mileage and prior claims. By contrast, `vehicle_age`, `male`, and `urban_area` are much flatter on the zero-versus-nonzero scale in this fit.

#### Interpreting the coefficients

The raw Logistic Regression coefficients are on the log-odds scale. To make them easier to read, the next table exponentiates scaled coefficient changes. The continuous covariates are rescaled to more meaningful comparisons: 10 years for `driver_age`, 10,000 km for `annual_mileage`, and 5 years for `vehicle_age`. Binary covariates are interpreted as the change from $0$ to $1$.

::: panel-tabset
## **`R` Code**

``` r
zip_interpret_scales <- tibble(
  term = c("driver_age", "annual_mileage", "vehicle_age", "male", "urban_area", "prior_claims"),
  comparison = c(
    "+10 years of driver age",
    "+10,000 km of annual mileage",
    "+5 years of vehicle age",
    "male (1) versus female (0)",
    "urban (1) versus rural (0)",
    "prior claims (1) versus none (0)"
  ),
  change = c(10, 10000, 5, 1, 1, 1)
)

coef_tbl <- broom::tidy(lr_model_r)
ci_tbl <- confint.default(lr_model_r) |>
  as.data.frame() |>
  tibble::rownames_to_column("term") |>
  rename(conf.low = `2.5 %`, conf.high = `97.5 %`)

lr_effects_r <- coef_tbl |>
  left_join(ci_tbl, by = "term") |>
  inner_join(zip_interpret_scales, by = "term") |>
  transmute(
    covariate = term,
    comparison,
    log_scale_change = estimate * change,
    odds_ratio = exp(estimate * change),
    lower_95 = exp(conf.low * change),
    upper_95 = exp(conf.high * change),
    percent_change = 100 * (exp(estimate * change) - 1),
    p_value = p.value
  )
```

## **`Python` Code**

``` python
scale_rows = pd.DataFrame({
    "term": covariates,
    "comparison": [
        "+10 years of driver age",
        "+10,000 km of annual mileage",
        "+5 years of vehicle age",
        "male (1) versus female (0)",
        "urban (1) versus rural (0)",
        "prior claims (1) versus none (0)",
    ],
    "change": [10, 10000, 5, 1, 1, 1],
})

ci = lr_model_py.conf_int()
ci.columns = ["conf.low", "conf.high"]
coef = pd.DataFrame({
    "term": lr_model_py.params.index,
    "estimate": lr_model_py.params.values,
    "conf.low": ci["conf.low"].values,
    "conf.high": ci["conf.high"].values,
    "p_value": lr_model_py.pvalues.values,
})

lr_effects_py = coef.merge(scale_rows, on="term", how="inner")
lr_effects_py["log_scale_change"] = lr_effects_py["estimate"] * lr_effects_py["change"]
lr_effects_py["odds_ratio"] = np.exp(lr_effects_py["log_scale_change"])
lr_effects_py["lower_95"] = np.exp(lr_effects_py["conf.low"] * lr_effects_py["change"])
lr_effects_py["upper_95"] = np.exp(lr_effects_py["conf.high"] * lr_effects_py["change"])
lr_effects_py["percent_change"] = 100 * (lr_effects_py["odds_ratio"] - 1)
```
:::

::: panel-tabset
## **`R` Output**

```{r tbl-lr-effects-r}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Scaled odds-ratio interpretation for the Binary Logistic Regression model using R."

if (exists("lr_effects_r")) {
  lr_effects_r |>
    fmt_table() |>
      print_as_dt()
} else {
  tibble(Message = "Scaled Logistic Regression effects are unavailable until the model is fitted.") |>
      print_as_dt()
}
```

## **`Python` Output**

```{python lr-effects-py-run}
#| echo: false
#| results: hide
#| message: false
#| warning: false

import numpy as np
import pandas as pd
import statsmodels.api as sm

# Recreate the Logistic Regression model if it is not already available.
if "lr_model_py" not in globals():
    if "training_data" not in globals():
        training_data = r.train_data.copy()

    training_data = training_data.copy()
    training_data["has_claim"] = (training_data["claims"] > 0).astype(int)

    covariates = [
        "driver_age",
        "annual_mileage",
        "vehicle_age",
        "male",
        "urban_area",
        "prior_claims",
    ]

    for col in covariates + ["claims", "has_claim"]:
        training_data[col] = pd.to_numeric(training_data[col], errors="coerce")

    training_data = training_data.dropna(subset=covariates + ["has_claim"]).copy()

    X = sm.add_constant(training_data[covariates].astype(float), has_constant="add")
    H = training_data["has_claim"].astype(int)

    lr_model_py = sm.Logit(H, X).fit(disp=False)

scale_rows = pd.DataFrame({
    "term": [
        "driver_age",
        "annual_mileage",
        "vehicle_age",
        "male",
        "urban_area",
        "prior_claims",
    ],
    "comparison": [
        "+10 years of driver age",
        "+10,000 km of annual mileage",
        "+5 years of vehicle age",
        "male (1) versus female (0)",
        "urban (1) versus rural (0)",
        "prior claims (1) versus none (0)",
    ],
    "change": [10, 10000, 5, 1, 1, 1],
})

ci = pd.DataFrame(
    np.asarray(lr_model_py.conf_int()),
    index=lr_model_py.params.index,
    columns=["conf.low", "conf.high"]
)

coef = pd.DataFrame({
    "term": lr_model_py.params.index,
    "estimate": lr_model_py.params.values,
    "conf.low": ci["conf.low"].values,
    "conf.high": ci["conf.high"].values,
    "p_value": lr_model_py.pvalues.values,
})

lr_effects_py = coef.merge(scale_rows, on="term", how="inner")
lr_effects_py["log_scale_change"] = lr_effects_py["estimate"] * lr_effects_py["change"]
lr_effects_py["odds_ratio"] = np.exp(lr_effects_py["log_scale_change"])
lr_effects_py["lower_95"] = np.exp(lr_effects_py["conf.low"] * lr_effects_py["change"])
lr_effects_py["upper_95"] = np.exp(lr_effects_py["conf.high"] * lr_effects_py["change"])
lr_effects_py["percent_change"] = 100 * (lr_effects_py["odds_ratio"] - 1)

lr_effects_py = lr_effects_py[[
    "term",
    "comparison",
    "log_scale_change",
    "odds_ratio",
    "lower_95",
    "upper_95",
    "percent_change",
    "p_value",
]].rename(columns={"term": "covariate"})
```

```{r tbl-lr-effects-py}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Scaled odds-ratio interpretation for the Binary Logistic Regression model using Python."

if (reticulate::py_eval("'lr_effects_py' in globals()", convert = TRUE)) {
  py_df_to_tibble("lr_effects_py") |>
    fmt_table() |>
      print_as_dt()
} else {
  tibble(Message = "Scaled Logistic Regression effects are unavailable until the model is fitted.") |>
      print_as_dt()
}
```
:::

An odds ratio greater than $1$ indicates higher odds of at least one claim for the stated comparison, holding the other covariates fixed. An odds ratio below $1$ indicates lower odds. These are odds-scale statements, not direct probability differences.

The scaled effects in @tbl-lr-effects-r make the size of the binary-response patterns clearer. Holding the other covariates fixed, a +10,000 km comparison in `annual_mileage` multiplies the fitted odds of at least one claim by about **`r lr_cpr_stat("lr_annual_or")`**, while moving from no prior claims to prior claims multiplies the fitted odds by about **`r lr_cpr_stat("lr_prior_or")`**. The +10 year `driver_age` comparison is positive but smaller, with an odds ratio of about **`r lr_cpr_stat("lr_driver_or")`**. The remaining comparisons are much closer to the neutral value of $1$ in this Logistic Regression model.

#### Visualizing the fitted Logistic Regression patterns

The next figures overlay empirical training-set summaries with fitted probabilities from the Logistic Regression model. For continuous covariates, the empirical points are binned averages. For binary covariates, the empirical points are group averages. The fitted curves or fitted line segments vary one covariate at a time while holding the remaining covariates at typical training-set values.

::: panel-tabset
## **`R` Code**

``` r
# Fitted probabilities were computed by varying one covariate at a time
# while holding the remaining covariates at typical training-set values.

ggplot() +
  geom_point(
    data = lr_observed_r,
    aes(x = x, y = observed),
    alpha = 0.85,
    size = 2
  ) +
  geom_line(
    data = lr_fitted_r,
    aes(x = x, y = fitted),
    linewidth = 1.1
  ) +
  facet_wrap(~ covariate_label, scales = "free_x") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(
    x = "Covariate value",
    y = "Probability of at least one claim"
  ) +
  zip_theme
```

## **`Python` Code**

``` python
# Fitted probabilities are computed by varying one covariate at a time
# while holding the remaining covariates at typical training-set values.

continuous_covariates = ["driver_age", "annual_mileage", "vehicle_age"]
binary_covariates = ["male", "urban_area", "prior_claims"]
labels = {
    "driver_age": "Driver age",
    "annual_mileage": "Annual mileage",
    "vehicle_age": "Vehicle age",
    "male": "Male indicator",
    "urban_area": "Urban-area indicator",
    "prior_claims": "Prior-claims indicator",
}

reference = {
    "driver_age": training_data["driver_age"].median(),
    "annual_mileage": training_data["annual_mileage"].median(),
    "vehicle_age": training_data["vehicle_age"].median(),
    "male": int(training_data["male"].mode().iloc[0]),
    "urban_area": int(training_data["urban_area"].mode().iloc[0]),
    "prior_claims": int(training_data["prior_claims"].mode().iloc[0]),
}

fig, axes = plt.subplots(2, 3, figsize=ZIP_WIDE_FIGSIZE, sharey=True)
axes = axes.flatten()

for ax, covariate in zip(axes, continuous_covariates + binary_covariates):
    if covariate in binary_covariates:
        grid_values = np.array([0, 1])
        observed = (
            training_data.groupby(covariate)["has_claim"]
            .mean()
            .reset_index()
            .rename(columns={covariate: "x", "has_claim": "observed"})
        )
    else:
        grid_values = np.linspace(training_data[covariate].min(), training_data[covariate].max(), 100)
        temp = training_data[[covariate, "has_claim"]].copy()
        temp["bin"] = pd.cut(temp[covariate], bins=6, include_lowest=True)
        observed = (
            temp.groupby("bin", observed=True)
            .agg(x=(covariate, "mean"), observed=("has_claim", "mean"))
            .reset_index(drop=True)
        )

    grid = pd.DataFrame([reference] * len(grid_values))
    grid[covariate] = grid_values
    X_grid = sm.add_constant(grid[zip_model_covariates_py], has_constant="add")
    fitted = lr_model_py.predict(X_grid)

    ax.scatter(observed["x"], observed["observed"], alpha=0.85, s=35)
    ax.plot(grid_values, fitted, linewidth=1.8)
    ax.set_title(labels[covariate])
    ax.set_xlabel("Covariate value")
    ax.set_ylabel("Probability of at least one claim")
    ax.set_ylim(0, 1)
    zip_clean_axis(ax)

plt.tight_layout()
plt.show()
```
:::

::: panel-tabset
## **`R` Output**

```{r fig-lr-fitted-patterns-r}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Observed event rates and fitted Logistic Regression probabilities by covariate using R."
#| fig.width: 14
#| fig.height: 8
#| eval: !expr exists("lr_observed_r") && exists("lr_fitted_r")

ggplot() +
  geom_point(
    data = lr_observed_r,
    aes(x = x, y = observed),
    alpha = 0.85,
    size = 2
  ) +
  geom_line(
    data = lr_fitted_r,
    aes(x = x, y = fitted),
    linewidth = 1.1
  ) +
  facet_wrap(~ covariate_label, scales = "free_x") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(
    x = "Covariate value",
    y = "Probability of at least one claim"
  ) +
  zip_theme
```

## **`Python` Output**

```{python fig-lr-fitted-patterns-py}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Observed event rates and fitted Logistic Regression probabilities by covariate using Python."
#| eval: !expr reticulate::py_eval("'lr_model_py' in globals()", convert = TRUE)

continuous_covariates = ["driver_age", "annual_mileage", "vehicle_age"]
binary_covariates = ["male", "urban_area", "prior_claims"]

labels = {
    "driver_age": "Driver age",
    "annual_mileage": "Annual mileage",
    "vehicle_age": "Vehicle age",
    "male": "Male indicator",
    "urban_area": "Urban-area indicator",
    "prior_claims": "Prior-claims indicator",
}

reference = {
    "driver_age": training_data["driver_age"].median(),
    "annual_mileage": training_data["annual_mileage"].median(),
    "vehicle_age": training_data["vehicle_age"].median(),
    "male": int(training_data["male"].mode().iloc[0]),
    "urban_area": int(training_data["urban_area"].mode().iloc[0]),
    "prior_claims": int(training_data["prior_claims"].mode().iloc[0]),
}

fig, axes = plt.subplots(2, 3, figsize=ZIP_WIDE_FIGSIZE, sharey=True)
axes = axes.flatten()

for ax, covariate in zip(axes, continuous_covariates + binary_covariates):
    if covariate in binary_covariates:
        grid_values = np.array([0, 1])

        observed = (
            training_data.groupby(covariate)["has_claim"]
            .mean()
            .reset_index()
            .rename(columns={covariate: "x", "has_claim": "observed"})
        )
    else:
        grid_values = np.linspace(
            training_data[covariate].min(),
            training_data[covariate].max(),
            100
        )

        temp = training_data[[covariate, "has_claim"]].copy()
        temp["bin"] = pd.cut(temp[covariate], bins=6, include_lowest=True)

        observed = (
            temp.groupby("bin", observed=True)
            .agg(
                x=(covariate, "mean"),
                observed=("has_claim", "mean")
            )
            .reset_index(drop=True)
        )

    grid = pd.DataFrame([reference] * len(grid_values))
    grid[covariate] = grid_values

    X_grid = sm.add_constant(
        grid[zip_model_covariates_py],
        has_constant="add"
    )

    fitted = lr_model_py.predict(X_grid)

    ax.scatter(
        observed["x"],
        observed["observed"],
        color=ZIP_POINT_COLOR,
        alpha=0.85,
        s=35,
        label="Observed"
    )

    ax.plot(
        grid_values,
        fitted,
        color=ZIP_LINE_COLOR,
        linewidth=1.8,
        label="Fitted"
    )

    ax.set_title(labels[covariate])
    ax.set_xlabel("Covariate value")
    ax.set_ylabel("Probability of at least one claim")
    ax.set_ylim(0, 1)
    zip_clean_axis(ax)

plt.tight_layout()
plt.show()
```
:::

The fitted patterns in @fig-lr-fitted-patterns-r show the same message visually. When other covariates are held at typical training-set values, the fitted probability of at least one claim rises from about **`r lr_cpr_pct("lr_annual_prob_low")`** at the low end of the mileage range to about **`r lr_cpr_pct("lr_annual_prob_high")`** at the high end. The fitted prior-claims contrast is also visible: the fitted probability changes from about **`r lr_cpr_pct("lr_prior_prob_no")`** for policyholders without prior claims to about **`r lr_cpr_pct("lr_prior_prob_yes")`** for policyholders with prior claims. The other panels are comparatively flatter, which is consistent with the smaller odds-ratio contrasts in @tbl-lr-effects-r.

This is exactly what Logistic Regression is meant to do: describe the probability that a policyholder enters the nonzero group. Its limitation is also visible from the original response distribution. Among training observations with at least one claim, the average positive count is **`r lr_cpr_stat("positive_mean_claims")`**, the maximum positive count is **`r lr_cpr_stat("positive_max_claims")`**, and about **`r lr_cpr_pct("multiple_claim_rate_given_positive")`** of nonzero-claim policyholders have more than one claim. After converting `claims` to `has_claim`, all of that frequency information is collapsed into $H_i = 1$.

### Classical Poisson Regression for `claims`

#### Model definition

The Classical Poisson Regression alternative keeps the original count response, defined in @eq-zip-cpr-response.

$$
Y_i = \text{number of claims filed by policyholder } i
$${#eq-zip-cpr-response}

The probabilistic assumption is shown in @eq-zip-cpr-distribution, where $\lambda_i = \mathbb{E}(Y_i \mid \mathbf{x}_i)$ is the expected number of claims for policyholder $i$.

$$
Y_i \mid \mathbf{x}_i \sim \operatorname{Poisson}(\lambda_i)
$${#eq-zip-cpr-distribution}

The link function is the log link in @eq-zip-cpr-log-link.

$$
g(\lambda_i) = \log(\lambda_i)
$${#eq-zip-cpr-log-link}

Using the same six covariates, the systematic component is given in @eq-zip-cpr-systematic.

$$
\log(\lambda_i)
=
\beta_0
+
\beta_1 x_{1,i}
+
\beta_2 x_{2,i}
+
\beta_3 x_{3,i}
+
\beta_4 x_{4,i}
+
\beta_5 x_{5,i}
+
\beta_6 x_{6,i}
$${#eq-zip-cpr-systematic}

In @eq-zip-cpr-systematic, there are $7$ coefficients, $\beta_0, \beta_1, \ldots, \beta_6$, which change the log-mean number of claims. The intercept $\beta_0$ is the baseline log-mean when all covariates are equal to zero, while $\beta_j$ for $j \in \{1, 2, \ldots, 6\}$ represents the change in log-mean claims for a one-unit increase in the corresponding regressor $x_{j,i}$, holding the other covariates fixed.

This model works directly on the count scale through $\lambda_i$, but it also imposes the Poisson mean-variance relationship in @eq-zip-cpr-mean-variance.

$$
\mathbb{E}(Y_i \mid \mathbf{x}_i)
=
\operatorname{Var}(Y_i \mid \mathbf{x}_i)
=
\lambda_i
$${#eq-zip-cpr-mean-variance}

::::: Heads-up
::: Heads-up-header
Heads-up on the connection to Chapter 10!
:::

::: Heads-up-container
The Poisson probability model, log link, and rate-ratio interpretation are introduced in @sec-classical-poisson. Here we use Classical Poisson Regression as a baseline count model before moving to ZIP regression.
:::
:::::

#### Fitting the model

::: panel-tabset
## **`R` Code**

``` r
cpr_model_r <- glm(
  claims ~ driver_age + annual_mileage + vehicle_age + male + urban_area + prior_claims,
  data = train_data,
  family = poisson(link = "log")
)

broom::tidy(cpr_model_r)
```

## **`Python` Code**

``` python
X = sm.add_constant(training_data[covariates], has_constant="add")
Y = training_data["claims"]

cpr_model_py = sm.GLM(
    Y,
    X,
    family=sm.families.Poisson()
).fit()

cpr_coef_py = pd.DataFrame({
    "term": cpr_model_py.params.index,
    "estimate": cpr_model_py.params.values,
    "std.error": cpr_model_py.bse.values,
    "statistic": cpr_model_py.tvalues.values,
    "p.value": cpr_model_py.pvalues.values,
})
```
:::

::: panel-tabset
## **`R` Output**

```{r tbl-cpr-fit-r}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Classical Poisson Regression coefficient table for the canonical training data using R."

if (exists("cpr_coef_r")) {
  cpr_coef_r |>
    fmt_table() |>
      print_as_dt()
} else {
  tibble(Message = "Classical Poisson Regression output is unavailable until the training data are available.") |>
      print_as_dt()
}
```

## **`Python` Output**

```{python cpr-fit-py-run}
#| echo: false
#| results: hide
#| message: false
#| warning: false

import pandas as pd
import statsmodels.api as sm

# Use the canonical R split if the Python object is not already available.
if "training_data" not in globals():
    training_data = r.train_data.copy()

training_data = training_data.copy()

zip_model_covariates_py = [
    "driver_age",
    "annual_mileage",
    "vehicle_age",
    "male",
    "urban_area",
    "prior_claims",
]

for col in zip_model_covariates_py + ["claims"]:
    training_data[col] = pd.to_numeric(training_data[col], errors="coerce")

training_data = training_data.dropna(
    subset=zip_model_covariates_py + ["claims"]
).copy()

X_cpr = sm.add_constant(
    training_data[zip_model_covariates_py].astype(float),
    has_constant="add"
)

Y = training_data["claims"].astype(float)

cpr_model_py = sm.GLM(
    Y,
    X_cpr,
    family=sm.families.Poisson()
).fit()

cpr_coef_py = pd.DataFrame({
    "term": cpr_model_py.params.index,
    "estimate": cpr_model_py.params.values,
    "std.error": cpr_model_py.bse.values,
    "statistic": cpr_model_py.tvalues.values,
    "p.value": cpr_model_py.pvalues.values,
})
```

```{r tbl-cpr-fit-py}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Classical Poisson Regression coefficient table for the canonical training data using Python."

if (reticulate::py_eval("'cpr_coef_py' in globals()", convert = TRUE)) {
  py_df_to_tibble("cpr_coef_py") |>
    fmt_table() |>
      print_as_dt()
} else {
  tibble(Message = "Classical Poisson Regression output is unavailable until the training data are available.") |>
      print_as_dt()
}
```
:::

As with the Logistic Regression output, the `R` and `Python` summaries may differ in formatting and small numerical details. The key comparison is substantive: the signs, approximate magnitudes, and interpretation on the exponential scale.

The coefficient table in @tbl-cpr-fit-r tells a count-scale story rather than a zero-versus-nonzero story. `annual_mileage` and `prior_claims` remain strong positive predictors, but `vehicle_age` and `male` now also show clearer positive fitted associations with the expected number of claims. This is an important difference from the Logistic Regression fit: a covariate may be weak for predicting whether any claim occurs, but still matter for the expected number of claims once the count scale is preserved.

#### Interpreting the coefficients

Poisson Regression coefficients are on the log-mean scale. Exponentiating a coefficient gives a multiplicative effect on the expected count. The next table uses the same scaling choices as above so that the coefficient interpretation is easier to compare across the two approaches.

::: panel-tabset
## **`R` Code**

``` r
coef_tbl <- broom::tidy(cpr_model_r)
ci_tbl <- confint.default(cpr_model_r) |>
  as.data.frame() |>
  tibble::rownames_to_column("term") |>
  rename(conf.low = `2.5 %`, conf.high = `97.5 %`)

cpr_effects_r <- coef_tbl |>
  left_join(ci_tbl, by = "term") |>
  inner_join(zip_interpret_scales, by = "term") |>
  transmute(
    covariate = term,
    comparison,
    log_scale_change = estimate * change,
    rate_ratio = exp(estimate * change),
    lower_95 = exp(conf.low * change),
    upper_95 = exp(conf.high * change),
    percent_change = 100 * (exp(estimate * change) - 1),
    p_value = p.value
  )
```

## **`Python` Code**

``` python
ci = cpr_model_py.conf_int()
ci.columns = ["conf.low", "conf.high"]
coef = pd.DataFrame({
    "term": cpr_model_py.params.index,
    "estimate": cpr_model_py.params.values,
    "conf.low": ci["conf.low"].values,
    "conf.high": ci["conf.high"].values,
    "p_value": cpr_model_py.pvalues.values,
})

cpr_effects_py = coef.merge(scale_rows, on="term", how="inner")
cpr_effects_py["log_scale_change"] = cpr_effects_py["estimate"] * cpr_effects_py["change"]
cpr_effects_py["rate_ratio"] = np.exp(cpr_effects_py["log_scale_change"])
cpr_effects_py["lower_95"] = np.exp(cpr_effects_py["conf.low"] * cpr_effects_py["change"])
cpr_effects_py["upper_95"] = np.exp(cpr_effects_py["conf.high"] * cpr_effects_py["change"])
cpr_effects_py["percent_change"] = 100 * (cpr_effects_py["rate_ratio"] - 1)
```
:::

::: panel-tabset
## **`R` Output**

```{r tbl-cpr-effects-r}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Scaled rate-ratio interpretation for the Classical Poisson Regression model using R."

if (exists("cpr_effects_r")) {
  cpr_effects_r |>
    fmt_table() |>
      print_as_dt()
} else {
  tibble(Message = "Scaled Poisson Regression effects are unavailable until the model is fitted.") |>
      print_as_dt()
}
```

## **`Python` Output**

```{python cpr-effects-py-run}
#| echo: false
#| results: hide
#| message: false
#| warning: false

import numpy as np
import pandas as pd
import statsmodels.api as sm

# Recreate the Classical Poisson Regression model if it is not already available.
if "cpr_model_py" not in globals():
    if "training_data" not in globals():
        training_data = r.train_data.copy()

    training_data = training_data.copy()

    zip_model_covariates_py = [
        "driver_age",
        "annual_mileage",
        "vehicle_age",
        "male",
        "urban_area",
        "prior_claims",
    ]

    for col in zip_model_covariates_py + ["claims"]:
        training_data[col] = pd.to_numeric(training_data[col], errors="coerce")

    training_data = training_data.dropna(
        subset=zip_model_covariates_py + ["claims"]
    ).copy()

    X_cpr = sm.add_constant(
        training_data[zip_model_covariates_py].astype(float),
        has_constant="add"
    )

    Y = training_data["claims"].astype(float)

    cpr_model_py = sm.GLM(
        Y,
        X_cpr,
        family=sm.families.Poisson()
    ).fit()

scale_rows = pd.DataFrame({
    "term": [
        "driver_age",
        "annual_mileage",
        "vehicle_age",
        "male",
        "urban_area",
        "prior_claims",
    ],
    "comparison": [
        "+10 years of driver age",
        "+10,000 km of annual mileage",
        "+5 years of vehicle age",
        "male (1) versus female (0)",
        "urban (1) versus rural (0)",
        "prior claims (1) versus none (0)",
    ],
    "change": [10, 10000, 5, 1, 1, 1],
})

ci = pd.DataFrame(
    np.asarray(cpr_model_py.conf_int()),
    index=cpr_model_py.params.index,
    columns=["conf.low", "conf.high"]
)

coef = pd.DataFrame({
    "term": cpr_model_py.params.index,
    "estimate": cpr_model_py.params.values,
    "conf.low": ci["conf.low"].values,
    "conf.high": ci["conf.high"].values,
    "p_value": cpr_model_py.pvalues.values,
})

cpr_effects_py = coef.merge(scale_rows, on="term", how="inner")
cpr_effects_py["log_scale_change"] = (
    cpr_effects_py["estimate"] * cpr_effects_py["change"]
)
cpr_effects_py["rate_ratio"] = np.exp(cpr_effects_py["log_scale_change"])
cpr_effects_py["lower_95"] = np.exp(
    cpr_effects_py["conf.low"] * cpr_effects_py["change"]
)
cpr_effects_py["upper_95"] = np.exp(
    cpr_effects_py["conf.high"] * cpr_effects_py["change"]
)
cpr_effects_py["percent_change"] = 100 * (cpr_effects_py["rate_ratio"] - 1)

cpr_effects_py = cpr_effects_py[[
    "term",
    "comparison",
    "log_scale_change",
    "rate_ratio",
    "lower_95",
    "upper_95",
    "percent_change",
    "p_value",
]].rename(columns={"term": "covariate"})
```

```{r tbl-cpr-effects-py}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Scaled rate-ratio interpretation for the Classical Poisson Regression model using Python."

if (reticulate::py_eval("'cpr_effects_py' in globals()", convert = TRUE)) {
  py_df_to_tibble("cpr_effects_py") |>
    fmt_table() |>
      print_as_dt()
} else {
  tibble(Message = "Scaled Poisson Regression effects are unavailable until the model is fitted.") |>
      print_as_dt()
}
```

:::

A rate ratio greater than $1$ means that the expected number of claims is higher for the stated comparison, holding the other covariates fixed. A rate ratio below $1$ means that the expected number of claims is lower.

The scaled rate ratios in @tbl-cpr-effects-r show that the largest count-scale comparison is again `annual_mileage`: a +10,000 km comparison multiplies the fitted expected number of claims by about **`r lr_cpr_stat("cpr_annual_rr")`**. The fitted rate ratio for `prior_claims` is about **`r lr_cpr_stat("cpr_prior_rr")`**, and the +5 year `vehicle_age` comparison has a fitted rate ratio of about **`r lr_cpr_stat("cpr_vehicle_rr")`**. These CPR effects are not identical to the Logistic Regression effects because they answer a different question: how the expected count changes, not only whether the count is positive.

#### Visualizing the fitted Poisson Regression patterns

The next figures overlay empirical mean claims with fitted Poisson means. As before, continuous covariates use binned empirical summaries, binary covariates use group summaries, and the fitted curves vary one covariate at a time while holding the remaining covariates fixed at typical training-set values.

::: panel-tabset
## **`R` Code**

``` r
ggplot() +
  geom_point(
    data = cpr_observed_r,
    aes(x = x, y = observed),
    alpha = 0.85,
    size = 2
  ) +
  geom_line(
    data = cpr_fitted_r,
    aes(x = x, y = fitted),
    linewidth = 1.1
  ) +
  facet_wrap(~ covariate_label, scales = "free_x") +
  labs(
    x = "Covariate value",
    y = "Mean number of claims"
  ) +
  zip_theme
```

## **`Python` Code**

``` python
fig, axes = plt.subplots(2, 3, figsize=ZIP_WIDE_FIGSIZE, sharey=False)
axes = axes.flatten()

for ax, covariate in zip(axes, continuous_covariates + binary_covariates):
    if covariate in binary_covariates:
        grid_values = np.array([0, 1])
        observed = (
            training_data.groupby(covariate)["claims"]
            .mean()
            .reset_index()
            .rename(columns={covariate: "x", "claims": "observed"})
        )
    else:
        grid_values = np.linspace(training_data[covariate].min(), training_data[covariate].max(), 100)
        temp = training_data[[covariate, "claims"]].copy()
        temp["bin"] = pd.cut(temp[covariate], bins=6, include_lowest=True)
        observed = (
            temp.groupby("bin", observed=True)
            .agg(x=(covariate, "mean"), observed=("claims", "mean"))
            .reset_index(drop=True)
        )

    grid = pd.DataFrame([reference] * len(grid_values))
    grid[covariate] = grid_values
    X_grid = sm.add_constant(grid[zip_model_covariates_py], has_constant="add")
    fitted = cpr_model_py.predict(X_grid)

    ax.scatter(observed["x"], observed["observed"], alpha=0.85, s=35)
    ax.plot(grid_values, fitted, linewidth=1.8)
    ax.set_title(labels[covariate])
    ax.set_xlabel("Covariate value")
    ax.set_ylabel("Mean number of claims")
    zip_clean_axis(ax)

plt.tight_layout()
plt.show()
```
:::

::: panel-tabset
## **`R` Output**

```{r fig-cpr-fitted-patterns-r}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Observed mean claims and fitted Classical Poisson Regression means by covariate using R."
#| fig.width: 14
#| fig.height: 8
#| eval: !expr exists("cpr_observed_r") && exists("cpr_fitted_r")

ggplot() +
  geom_point(
    data = cpr_observed_r,
    aes(x = x, y = observed),
    alpha = 0.85,
    size = 2
  ) +
  geom_line(
    data = cpr_fitted_r,
    aes(x = x, y = fitted),
    linewidth = 1.1
  ) +
  facet_wrap(~ covariate_label, scales = "free_x") +
  labs(
    x = "Covariate value",
    y = "Mean number of claims"
  ) +
  zip_theme
```

## **`Python` Output**

```{python fig-cpr-fitted-patterns-py}
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Observed mean claims and fitted Classical Poisson Regression means by covariate using Python."
#| eval: !expr reticulate::py_eval("'cpr_model_py' in globals()", convert = TRUE)

fig, axes = plt.subplots(2, 3, figsize=ZIP_WIDE_FIGSIZE, sharey=False)
axes = axes.flatten()

for ax, covariate in zip(axes, continuous_covariates + binary_covariates):
    if covariate in binary_covariates:
        grid_values = np.array([0, 1])

        observed = (
            training_data.groupby(covariate)["claims"]
            .mean()
            .reset_index()
            .rename(columns={covariate: "x", "claims": "observed"})
        )
    else:
        grid_values = np.linspace(
            training_data[covariate].min(),
            training_data[covariate].max(),
            100
        )

        temp = training_data[[covariate, "claims"]].copy()
        temp["bin"] = pd.cut(temp[covariate], bins=6, include_lowest=True)

        observed = (
            temp.groupby("bin", observed=True)
            .agg(
                x=(covariate, "mean"),
                observed=("claims", "mean")
            )
            .reset_index(drop=True)
        )

    grid = pd.DataFrame([reference] * len(grid_values))
    grid[covariate] = grid_values

    X_grid = sm.add_constant(
        grid[zip_model_covariates_py],
        has_constant="add"
    )

    fitted = cpr_model_py.predict(X_grid)

    ax.scatter(
        observed["x"],
        observed["observed"],
        color=ZIP_POINT_COLOR,
        alpha=0.85,
        s=35,
        label="Observed"
    )

    ax.plot(
        grid_values,
        fitted,
        color=ZIP_LINE_COLOR,
        linewidth=1.8,
        label="Fitted"
    )

    ax.set_title(labels[covariate])
    ax.set_xlabel("Covariate value")
    ax.set_ylabel("Mean number of claims")
    zip_clean_axis(ax)

plt.tight_layout()
plt.show()
```

:::

The fitted patterns in @fig-cpr-fitted-patterns-r show why the Poisson baseline is informative. The fitted mean count increases from about **`r lr_cpr_stat("cpr_annual_mean_low")`** to about **`r lr_cpr_stat("cpr_annual_mean_high")`** across the mileage range, and from about **`r lr_cpr_stat("cpr_vehicle_mean_low")`** to about **`r lr_cpr_stat("cpr_vehicle_mean_high")`** across the vehicle-age range, with other covariates held at typical training-set values. The prior-claims panel also shows a clear fitted increase, from about **`r lr_cpr_stat("cpr_prior_mean_no")`** to about **`r lr_cpr_stat("cpr_prior_mean_yes")`**.

Classical Poisson Regression therefore preserves the count response and gives a direct model for claim frequency. However, it uses one count-generating process for all observations. If many zeros arise from a separate mechanism, a single Poisson mean may have difficulty representing both the large mass at zero and the positive-count distribution. The next check makes this limitation explicit by comparing the observed zero frequency with the zero frequency implied by the fitted Poisson means.

The calculation is straightforward once the fitted Poisson means are available. For each policyholder, the fitted Classical Poisson Regression model gives a fitted mean $\widehat{\lambda}_i$. Under the Poisson model in @eq-zip-cpr-distribution, the probability of observing zero claims for that same policyholder is shown in @eq-zip-cpr-zero-probability.

$$
\widehat{p}^{\text{CPR}}_{0,i}
=
P(Y_i = 0 \mid \widehat{\lambda}_i)
=
\exp(-\widehat{\lambda}_i)
$${#eq-zip-cpr-zero-probability}

Averaging the fitted zero probabilities in @eq-zip-cpr-zero-probability gives the model-implied zero rate, and summing them gives the expected number of zero-claim policyholders under the fitted Poisson model, as shown in @eq-zip-cpr-zero-frequency.

$$
\widehat{r}^{\text{CPR}}_0
=
\frac{1}{n}\sum_{i=1}^{n}\widehat{p}^{\text{CPR}}_{0,i},
\qquad
\widehat{N}^{\text{CPR}}_0
=
\sum_{i=1}^{n}\widehat{p}^{\text{CPR}}_{0,i}
$${#eq-zip-cpr-zero-frequency}

The same table also reports the Pearson dispersion ratio in @eq-zip-cpr-pearson-dispersion, where $p$ is the number of estimated regression coefficients in the fitted Classical Poisson model.

$$
\widehat{\phi}_{\text{Pearson}}
=
\frac{\sum_{i=1}^{n}\frac{(y_i - \widehat{\lambda}_i)^2}{\widehat{\lambda}_i}}{n - p}
$${#eq-zip-cpr-pearson-dispersion}

::: panel-tabset
## **`R` Code**

``` r
# Step 1: extract the fitted Poisson means for the training rows.
cpr_fitted_mu_r <- predict(cpr_model_r, type = "response")

# Step 2: convert each fitted mean into its implied zero probability.
cpr_fitted_zero_prob_r <- exp(-cpr_fitted_mu_r)

# Step 3: compare the observed and model-implied zero frequencies.
cpr_zero_check_r <- tibble(
  Summary = c(
    "Observed zero rate",
    "Average fitted Poisson zero probability",
    "Observed minus fitted zero rate",
    "Observed number of zeros",
    "Expected number of zeros under fitted Poisson",
    "Pearson dispersion ratio"
  ),
  Value = c(
    mean(train_data$claims == 0, na.rm = TRUE),
    mean(cpr_fitted_zero_prob_r, na.rm = TRUE),
    mean(train_data$claims == 0, na.rm = TRUE) -
      mean(cpr_fitted_zero_prob_r, na.rm = TRUE),
    sum(train_data$claims == 0, na.rm = TRUE),
    sum(cpr_fitted_zero_prob_r, na.rm = TRUE),
    sum(residuals(cpr_model_r, type = "pearson")^2, na.rm = TRUE) /
      df.residual(cpr_model_r)
  )
)

cpr_zero_check_r |>
  fmt_table() |>
    print_as_dt()
```

## **`Python` Code**

``` python
# Step 1: recreate the design matrix used by the fitted Poisson model.
X_cpr = sm.add_constant(
    training_data[zip_model_covariates_py].astype(float),
    has_constant="add"
)

# Step 2: extract the fitted Poisson means for the training rows.
cpr_fitted_mu_py = cpr_model_py.predict(X_cpr)

# Step 3: convert each fitted mean into its implied zero probability.
cpr_fitted_zero_prob_py = np.exp(-cpr_fitted_mu_py)

# Step 4: compare the observed and model-implied zero frequencies.
observed_zero_rate_py = (training_data["claims"] == 0).mean()
observed_zero_n_py = int((training_data["claims"] == 0).sum())
pearson_dispersion_py = cpr_model_py.pearson_chi2 / cpr_model_py.df_resid

cpr_zero_check_py = pd.DataFrame({
    "Summary": [
        "Observed zero rate",
        "Average fitted Poisson zero probability",
        "Observed minus fitted zero rate",
        "Observed number of zeros",
        "Expected number of zeros under fitted Poisson",
        "Pearson dispersion ratio",
    ],
    "Value": [
        observed_zero_rate_py,
        cpr_fitted_zero_prob_py.mean(),
        observed_zero_rate_py - cpr_fitted_zero_prob_py.mean(),
        observed_zero_n_py,
        cpr_fitted_zero_prob_py.sum(),
        pearson_dispersion_py,
    ],
})
```
:::

::: panel-tabset
## **`R` Output**

```{r tbl-cpr-zero-check-r}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Observed zero frequency compared with the zero frequency implied by the fitted Classical Poisson Regression model using R."

if (exists("cpr_model_r") && zip_has_train()) {
  cpr_fitted_mu_r <- predict(cpr_model_r, type = "response")
  cpr_fitted_zero_prob_r <- exp(-cpr_fitted_mu_r)

  cpr_zero_check_r <- tibble(
    Summary = c(
      "Observed zero rate",
      "Average fitted Poisson zero probability",
      "Observed minus fitted zero rate",
      "Observed number of zeros",
      "Expected number of zeros under fitted Poisson",
      "Pearson dispersion ratio"
    ),
    Value = c(
      mean(train_data$claims == 0, na.rm = TRUE),
      mean(cpr_fitted_zero_prob_r, na.rm = TRUE),
      mean(train_data$claims == 0, na.rm = TRUE) -
        mean(cpr_fitted_zero_prob_r, na.rm = TRUE),
      sum(train_data$claims == 0, na.rm = TRUE),
      sum(cpr_fitted_zero_prob_r, na.rm = TRUE),
      sum(residuals(cpr_model_r, type = "pearson")^2, na.rm = TRUE) /
        df.residual(cpr_model_r)
    )
  )

  cpr_zero_check_r |>
    fmt_table() |>
      print_as_dt()
} else {
  tibble(Message = "Poisson zero-frequency check is unavailable until the Classical Poisson model is fitted.") |>
      print_as_dt()
}
```

## **`Python` Output**

```{python cpr-zero-check-py-run}
#| echo: false
#| results: hide
#| message: false
#| warning: false

import numpy as np
import pandas as pd
import statsmodels.api as sm

# Recreate the Classical Poisson Regression model if it is not already available.
if "cpr_model_py" not in globals():
    if "training_data" not in globals():
        training_data = r.train_data.copy()

    training_data = training_data.copy()

    zip_model_covariates_py = [
        "driver_age",
        "annual_mileage",
        "vehicle_age",
        "male",
        "urban_area",
        "prior_claims",
    ]

    for col in zip_model_covariates_py + ["claims"]:
        training_data[col] = pd.to_numeric(training_data[col], errors="coerce")

    training_data = training_data.dropna(
        subset=zip_model_covariates_py + ["claims"]
    ).copy()

    X_cpr = sm.add_constant(
        training_data[zip_model_covariates_py].astype(float),
        has_constant="add"
    )

    Y = training_data["claims"].astype(float)

    cpr_model_py = sm.GLM(
        Y,
        X_cpr,
        family=sm.families.Poisson()
    ).fit()
else:
    X_cpr = sm.add_constant(
        training_data[zip_model_covariates_py].astype(float),
        has_constant="add"
    )

cpr_fitted_mu_py = cpr_model_py.predict(X_cpr)
cpr_fitted_zero_prob_py = np.exp(-cpr_fitted_mu_py)

observed_zero_rate_py = (training_data["claims"] == 0).mean()
observed_zero_n_py = int((training_data["claims"] == 0).sum())
pearson_dispersion_py = cpr_model_py.pearson_chi2 / cpr_model_py.df_resid

cpr_zero_check_py = pd.DataFrame({
    "Summary": [
        "Observed zero rate",
        "Average fitted Poisson zero probability",
        "Observed minus fitted zero rate",
        "Observed number of zeros",
        "Expected number of zeros under fitted Poisson",
        "Pearson dispersion ratio",
    ],
    "Value": [
        observed_zero_rate_py,
        cpr_fitted_zero_prob_py.mean(),
        observed_zero_rate_py - cpr_fitted_zero_prob_py.mean(),
        observed_zero_n_py,
        cpr_fitted_zero_prob_py.sum(),
        pearson_dispersion_py,
    ],
})
```

```{r tbl-cpr-zero-check-py}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Observed zero frequency compared with the zero frequency implied by the fitted Classical Poisson Regression model using Python."

if (reticulate::py_eval("'cpr_zero_check_py' in globals()", convert = TRUE)) {
  py_df_to_tibble("cpr_zero_check_py") |>
    fmt_table() |>
      print_as_dt()
} else {
  tibble(Message = "Python Poisson zero-frequency check is unavailable until the Classical Poisson model is fitted.") |>
      print_as_dt()
}
```
:::

In @tbl-cpr-zero-check-r and @tbl-cpr-zero-check-py, the observed zero rate is **`r lr_cpr_pct("observed_zero_rate")`**, while the average zero probability implied by the fitted Classical Poisson model is about **`r lr_cpr_pct("cpr_fitted_zero_rate")`**. The gap is about **`r lr_cpr_pct("cpr_zero_gap")`**, or roughly **`r lr_cpr_stat("observed_zero_n")`** observed zeros versus **`r lr_cpr_stat("cpr_expected_zero_n")`** expected zeros under the fitted Poisson means. This does not mean the Poisson model is useless; it captures important count-scale gradients. It does mean that the model has no separate mechanism for the extra zeros.

### Why ZIP can be useful here

The two simpler approaches highlight the motivation for ZIP regression.

| **Approach** | **Response Used** | **What the fitted comparison shows here** | **Main Limitation in This Example** |
|:-----------------|:-----------------|:-----------------|:-----------------|
| Binary Logistic Regression | $H_i = \mathbb{I}(Y_i > 0)$, implemented as `has_claim` | @tbl-lr-effects-r and @fig-lr-fitted-patterns-r show a strong zero-versus-nonzero signal for `annual_mileage` and `prior_claims`; the fitted odds ratios are about **`r lr_cpr_stat("lr_annual_or")`** and **`r lr_cpr_stat("lr_prior_or")`** for the scaled comparisons. | It collapses all positive counts into one category. In the training data, positive counts average **`r lr_cpr_stat("positive_mean_claims")`** and can reach **`r lr_cpr_stat("positive_max_claims")`**, so a binary response discards meaningful claim-frequency information. |
| Classical Poisson Regression | $Y_i$, implemented as `claims` | @tbl-cpr-effects-r and @fig-cpr-fitted-patterns-r show clear count-scale gradients, especially for `annual_mileage`, `prior_claims`, and `vehicle_age`; their scaled fitted rate ratios are about **`r lr_cpr_stat("cpr_annual_rr")`**, **`r lr_cpr_stat("cpr_prior_rr")`**, and **`r lr_cpr_stat("cpr_vehicle_rr")`**. | It assumes a single Poisson count-generating process. As shown in @tbl-cpr-zero-check-r and @tbl-cpr-zero-check-py, its fitted zero probability is about **`r lr_cpr_pct("cpr_fitted_zero_rate")`**, below the observed zero rate of **`r lr_cpr_pct("observed_zero_rate")`**. |
| Zero-Inflated Poisson Regression | $Y_i$, implemented as `claims` | The EDA and the two baseline fits point to two simultaneous features: a strong zero-versus-nonzero process and meaningful variation in the positive count scale. | It is more complex, so it should be used when the EDA and subject-matter context justify a separate zero-generating mechanism. |

In the insurance-claims setting, the EDA suggested a large zero-claim group and a positive-count pattern among the remaining observations. Logistic Regression describes the first piece: who is likely to file at least one claim. Classical Poisson Regression describes the second piece: how expected claim counts change with the covariates. ZIP regression is useful because it gives these two pieces separate but connected roles. One component models extra zeros, while the other models claim counts among observations that remain in the count-generating process.
