<!-- Google tag (gtag.js) -->
<script async src="https://www.googletagmanager.com/gtag/js?id=G-7PRVEBE1EF"></script>
<script>
  window.dataLayer = window.dataLayer || [];
  function gtag(){dataLayer.push(arguments);}
  gtag('js', new Date());
  gtag('config', 'G-7PRVEBE1EF');
</script>

<!-- MathJax configuration -->
<script>
  MathJax = {
    tex: {
      tags: 'ams',
      useLabelIds: true
    }
  };
</script>

# Generalized Poisson Regression {#sec-generalized-poisson}

```{r}
#| include: false
colourize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color, x)
  } else x
}
```

::: {.Warning}
::::{.Warning-header}
When to Use and Not Use Generalized Poisson Regression
::::
::::{.Warning-container}
Generalized Poisson regression is a type of generalized linear model that is appropriately used under the following conditions:

- The outcome variable is **discrete** and **count-type** as shown in @fig-generalied-poisson-regression. Examples include number of doctor visits, number of web clicks, number of accidents, etc.
- Outcomes are statistically independent from one to the next.
- The data shows **overdispersion** (variance exceeds mean) **or underdispersion** (variance is less than mean). See @sec-classical-poisson for the equidispersion case.
- Counts occur over fixed and known exposures (such as time intervals or geographic areas).
- We intend to use the **logarithm of the mean** of the outcome as a **link function**, which ensures that our predictions are always positive.

However, Generalized Poisson regression should not be used in the following scenarios:

- The outcome is not a count (e.g., binary outcomes, proportions, or continuous variables).
- The dispersion parameter $\delta = 0$ is strongly supported - use [Classical Poisson](10-classical-poisson.qmd) (Chapter 10) instead.
- The outcome variable includes a large number of zeros - use [Zero-Inflated Poisson](12-zero-inflated-poisson.qmd) (Chapter 12) instead.
::::
:::

::: {#fig-generalied-poisson-regression}
```{mermaid}
mindmap
  root((Regression 
  Analysis)
    Continuous <br/>Outcome Y
      {{Unbounded <br/>Outcome Y}}
        )Chapter 3: <br/>Ordinary <br/>Least Squares <br/>Regression(
          (Normal <br/>Outcome Y)
      {{Nonnegative <br/>Outcome Y}}
        )Chapter 4: <br/>Gamma Regression(
          (Gamma <br/>Outcome Y)
      {{Bounded <br/>Outcome Y <br/> between 0 and 1}}
        )Chapter 5: <br/>Beta <br/>Regression(
          (Beta <br/>Outcome Y)
      {{Nonnegative <br/>Survival <br/>Time Y}}
        )Chapter 6: <br/>Parametric <br/> Survival <br/>Regression(
          (Exponential <br/>Outcome Y)
          (Weibull <br/>Outcome Y)
          (Lognormal <br/>Outcome Y)
        )Chapter 7: <br/>Semiparametric <br/>Survival <br/>Regression(
          (Cox Proportional <br/>Hazards Model)
            (Hazard Function <br/>Outcome Y)
    Discrete <br/>Outcome Y
      {{Binary <br/>Outcome Y}}
        {{Ungrouped <br/>Data}}
          )Chapter 8: <br/>Binary Logistic <br/>Regression(
            (Bernoulli <br/>Outcome Y)
        {{Grouped <br/>Data}}
          )Chapter 9: <br/>Binomial Logistic <br/>Regression(
            (Binomial <br/>Outcome Y)
      {{Count <br/>Outcome Y}}
        {{Equidispersed <br/>Data}}
          )Chapter 10: <br/>Classical Poisson <br/>Regression(
            (Poisson <br/>Outcome Y)
        {{Overdispersed <br/>Data}}
          )Chapter 11: <br/>Negative Binomial <br/>Regression(
            (Negative Binomial <br/>Outcome Y)
        {{Zero Inflated <br/>Data}}
          )Chapter 12: <br/>Zero Inflated <br/>Poisson <br/>Regression(
            (Zero Inflated <br/>Poisson <br/>Outcome Y)
        {{Overdispersed or <br/>Underdispersed <br/>Data}}
          )Chapter 13: <br/>Generalized <br/>Poisson <br/>Regression(
            (Generalized <br/>Poisson <br/>Outcome Y)
```

Regression analysis mind map depicting all modelling techniques explored in this book. Depending on the type of outcome $Y$, these techniques are split into two large zones: *discrete* and *continuous*.
:::

```{r setup-r, include=FALSE}
# TODO: Add required libraries (VGAM, tidyverse, broom, asbio)
```

```{python setup-py, include=FALSE}
# TODO: Add required Python imports and custom GP likelihood class
```

::: {.LO}
::::{.LO-header}
Learning Objectives
::::
::::{.LO-container}
By the end of this chapter, you will be able to:

- Explain why Classical Poisson regression underperforms when the equidispersion assumption is violated.
- Determine when Generalized Poisson regression is preferred over Classical Poisson and Negative Binomial regression.
- Write down the Generalized Poisson PMF and its mean-variance relationship.
- Describe the role of the dispersion parameter $\delta$ and interpret its sign and magnitude.
- Perform and interpret a dispersion test to detect over- or underdispersion.
- Fit a Generalized Poisson regression model in R and Python.
- Interpret the regression coefficients (IRR) and the dispersion parameter $\delta$.
- Evaluate model performance and compare results against Classical Poisson regression.
::::
:::

## Introduction

This chapter extends [Chapter 10](10-classical-poisson.qmd) by introducing a more flexible count regression model. Classical Poisson regression enforces **equidispersion**: the mean and variance of the count outcome must be equal. In practice, this assumption is frequently violated. When the variance *exceeds* the mean, the data is **overdispersed**; when the variance falls *below* the mean, it is **underdispersed**. Under overdispersion, Classical Poisson underestimates standard errors, inflating the risk of Type I error. Under underdispersion, it overestimates standard errors, making real associations harder to detect. Classical Poisson has no mechanism to handle underdispersion at all.

The Generalized Poisson (GP) distribution, introduced by Consul & Jain (1973), relaxes this constraint by adding a single dispersion parameter $\delta$. When $\delta = 0$, GP reduces exactly to Classical Poisson. When $\delta > 0$ it accommodates overdispersion; when $\delta < 0$ it accommodates underdispersion - making it the only standard count model that handles both directions. Both GP and Negative Binomial regression (@sec-negative-binomial, Chapter 11) address overdispersion, but only GP handles underdispersion, and GP nests Classical Poisson as a special case, making model comparison straightforward via a likelihood ratio test.

Some research questions you might explore using Generalized Poisson regression:

- How does the number of satellite male horseshoe crabs around a female nest vary by shell width and color, accounting for the overdispersion present in the data?
- How does the number of workplace injury claims per employee vary by industry sector and firm size, where variance in claims far exceeds the mean?

::: {.callout-note appearance="simple"}
**TODO:** Add an underdispersion research question once the Case Study 2 dataset is confirmed.
:::

### Generalized Poisson Regression Assumptions

The assumptions of Generalized Poisson regression parallel those of Classical Poisson regression, with one key modification:

- **Independence**: Each count response $Y_i$ is independent of all others, with its own mean $\mu_i > 0$.
- **Log-linearity**: The log of the mean is modelled as a linear combination of the predictors: $\log(\mu_i) = \beta_0 + \beta_1 X_{i,1} + \dots + \beta_k X_{i,k}$.
- **Dispersion flexibility**: Unlike Classical Poisson, GP allows $\text{Var}(Y_i) \neq \mathbb{E}(Y_i)$. Specifically, $\text{Var}(Y_i) = \mu_i / (1 - \delta)^2$, where $\delta$ controls the direction and degree of dispersion.

### A Graphical Look

::: {.callout-note appearance="simple"}
**TODO:** Side-by-side figure - left panel: Classical Poisson fit on overdispersed data; right panel: GP fit on same data. Adapt simulation code from Chapter 10.
:::

## Case Study 1: Horseshoe Crab Satellites (Overdispersion)

We return to the horseshoe crab dataset from Chapter 10. The outcome variable `satell` records the number of male satellites around each female nest. The sample mean of `satell` is approximately 2.92 and the sample variance is approximately 9.91, giving a variance-to-mean ratio of roughly 3.4 - strong preliminary evidence of overdispersion that Classical Poisson was not equipped to handle.

## Data Collection and Wrangling

::: {.callout-note appearance="simple"}
**TODO:** Confirm how much wrangling to repeat here vs. reference Chapter 10. Current plan: brief reload and mean/variance check only.
:::

The `crabs` dataset is described in full in Chapter 10. We reload it here using the same 70/30 train/test split and random seed for direct comparability.

::: {.panel-tabset}
## R Code
```{.r}
# TODO: Reload crabs, repeat train/test split (seed = 1046)
# Recompute mean and variance of satell to confirm overdispersion
```

## Python Code
```{.python}
# TODO: Reload crabs dataframe
# Recompute mean and variance of satell to confirm overdispersion
```
:::

## Data Modelling

The Generalized Poisson (GP) distribution (Consul & Jain, 1973) extends Classical Poisson by introducing a dispersion parameter $\delta$. A random variable $Y_i$ follows a GP distribution if its probability mass function is:

$$
P(Y_i = y_i \mid \theta_i, \delta) = \frac{\theta_i(\theta_i + \delta y_i)^{y_i - 1} e^{-(\theta_i + \delta y_i)}}{y_i!}, \quad y_i = 0, 1, 2, \ldots
$$

where $\theta_i > 0$ and $\max(-1, -\theta_i/m) \leq \delta \leq 1$ for $m \geq 4$. The mean and variance are:

$$
\mathbb{E}(Y_i) = \frac{\theta_i}{1 - \delta} = \mu_i, \qquad \text{Var}(Y_i) = \frac{\mu_i}{(1 - \delta)^2}.
$$

The dispersion ratio $\text{Var}(Y_i) / \mathbb{E}(Y_i) = 1/(1-\delta)^2$ summarizes departure from equidispersion: equal to 1 when $\delta = 0$ (Classical Poisson), greater than 1 when $\delta > 0$ (overdispersion), and less than 1 when $\delta < 0$ (underdispersion).

We use the same log link as Chapter 10:

$$
\log(\mu_i) = \beta_0 + \beta_1 X_{i,1} + \dots + \beta_k X_{i,k}.
$$

The interpretation of $\beta$ coefficients and their exponentiated IRRs is identical to Chapter 10. The only addition is $\delta$, estimated alongside $\boldsymbol{\beta}$.

## Estimation

Parameters are estimated via joint maximum likelihood for $(\boldsymbol{\beta}, \delta)$. The log-likelihood is:

$$
\ell(\boldsymbol{\beta}, \delta) = \sum_{i=1}^n \left[ \log\theta_i + (y_i - 1)\log(\theta_i + \delta y_i) - (\theta_i + \delta y_i) - \log(y_i!) \right]
$$

where $\theta_i = (1-\delta)\exp(X_i^\top \boldsymbol{\beta})$. Unlike Classical Poisson where only $\boldsymbol{\beta}$ is estimated via IRLS, here both parameters are estimated jointly via Newton-Raphson / BFGS, initialized from a Classical Poisson fit with $\delta^{(0)} = 0$.

::: {.callout-note appearance="simple"}
**TODO:** Confirm preferred depth for this section - light treatment vs. same depth as the Estimation section in Chapter 10.
:::

::: {.panel-tabset}
## R Code
```{.r}
# TODO: Fit GP model using VGAM::vglm() with genpoisson1() family
```

## Python Code
```{.python}
# TODO: Fit GP model using custom GenericLikelihoodModel
```
:::

## Goodness of Fit

Goodness of fit uses two components. First, residual deviance follows the same formula as in @goodness:

$$
\text{Residual Deviance} = 2(\ell_{\text{saturated}} - \ell_{\text{fitted}}).
$$

Second, a dispersion test formally tests $H_0: \delta = 0$ via a likelihood ratio test:

$$
\Lambda = -2\left(\ell_{\text{Poisson}} - \ell_{\text{GP}}\right) \sim \chi^2_1.
$$

A small $p$-value confirms that GP significantly improves on Classical Poisson.

::: {.callout-note appearance="simple"}
**TODO:** Confirm whether to also include the score test for overdispersion from the DSCI 562 notes alongside the LR test.
:::

::: {.panel-tabset}
## R Code
```{.r}
# TODO: Residual deviance + LR dispersion test (Classical Poisson vs GP)
```

## Python Code
```{.python}
# TODO: Residual deviance + LR dispersion test (Classical Poisson vs GP)
```
:::

## Results

### Univariate Generalized Poisson Model

We first fit a univariate GP model with `width` as the sole predictor:

$$
\log(\mu_i) = \beta_0 + \beta_1 X_{i,\text{width}}.
$$

::: {.panel-tabset}
## R Code
```{.r}
# TODO: Fit univariate GP model (width only)
# Report coefficients, SEs, Wald test, 95% CIs
```

## Python Code
```{.python}
# TODO: Fit univariate GP model (width only)
# Report coefficients, SEs, Wald test, 95% CIs
```
:::

### Interpreting the Dispersion Parameter

The dispersion parameter $\hat{\delta}$ is new in this chapter:

- If $\hat{\delta} > 0$ and the 95% CI excludes 0: overdispersion is confirmed; GP is preferred over Classical Poisson.
- The implied dispersion ratio is $1 / (1 - \hat{\delta})^2$, quantifying how much the variance exceeds the mean.
- If $\hat{\delta} \approx 0$ and the CI contains 0: Classical Poisson was sufficient.

::: {.callout-note appearance="simple"}
**TODO:** Fill in actual $\hat{\delta}$ and dispersion ratio once model is fit.
:::

### Multivariate Generalized Poisson Model

We now add `color` as a second predictor:

$$
\log(\mu_i) = \beta_0 + \beta_1 X_{i,\text{width}} + \beta_2 X_{i,\text{color}_2} + \beta_3 X_{i,\text{color}_3} + \beta_4 X_{i,\text{color}_4}.
$$

The IRR interpretation for both continuous and categorical predictors follows the same logic as Chapter 10.

::: {.panel-tabset}
## R Code
```{.r}
# TODO: Fit multivariate GP model (width + color)
# Report full coefficient table with IRRs
```

## Python Code
```{.python}
# TODO: Fit multivariate GP model (width + color)
# Report full coefficient table with IRRs
```
:::

### Comparing Classical Poisson vs. Generalized Poisson

::: {.callout-note appearance="simple"}
**TODO:** Side-by-side table of coefficients, SEs, and p-values from both models. Key point: Classical Poisson underestimates SEs under overdispersion, inflating Type I error risk.
:::

::: {.panel-tabset}
## R Code
```{.r}
# TODO: Side-by-side comparison table: Classical Poisson vs GP
```

## Python Code
```{.python}
# TODO: Side-by-side comparison table: Classical Poisson vs GP
```
:::

### Prediction Using Test Data

Using the same KL divergence metric from Chapter 10:

$$
\overline{D}_{\mathrm{KL}} = \frac{1}{n}\sum_{i=1}^n \left[ y_i \log\left(\frac{y_i}{\hat{\mu}_i}\right) - y_i + \hat{\mu}_i \right].
$$

We compare average KL divergence of the GP model against Classical Poisson on the test set. A lower value confirms improved fit.

::: {.panel-tabset}
## R Code
```{.r}
# TODO: Predict on test set, compute and compare KL divergence vs Classical Poisson
```

## Python Code
```{.python}
# TODO: Predict on test set, compute and compare KL divergence vs Classical Poisson
```
:::

## Case Study 2: [TBD Dataset] (Underdispersion)

::: {.callout-note appearance="simple"}
**TODO:** Confirm underdispersion dataset. Full workflow mirrors Case Study 1 with $\delta < 0$.
:::

## Storytelling

::: {.callout-note appearance="simple"}
**TODO:** Complete once models are fit. Cover: why GP was needed, what $\delta$ tells us, key predictor findings, and practical guidance for choosing between Poisson, NB, and GP.
:::

## Practice Problems

::: {.quiz}

### Conceptual Questions

::: {.callout-note appearance="simple"}
**TODO:** 10-12 questions covering role of $\delta$, dispersion test, GP vs NB, underdispersion intuition, and consequences of ignoring overdispersion.
:::

:::

### Coding Question

::: {.callout-note appearance="simple"}
**TODO:** Full A-J problem set parallel to Chapter 10. Dataset TBD - pending decision on underdispersion dataset.
:::
