<!-- Google tag (gtag.js) -->
<script async src="https://www.googletagmanager.com/gtag/js?id=G-7PRVEBE1EF"></script>
<script>
  window.dataLayer = window.dataLayer || [];
  function gtag(){dataLayer.push(arguments);}
  gtag('js', new Date());

  gtag('config', 'G-7PRVEBE1EF');
</script>

# Multinomial Logistic Regression {#sec-multinomial-logistic}

::: {.Warning}
::::{.Warning-header}
When to Use and Not Use Multinomial Logistic Regression
::::
::::{.Warning-container}
Multinomial Logistic regression is a type of generalized linear model that is appropriately used under the following conditions:

- The outcome variable is categorical with more than two categories, as illustrated in @fig-multinomial-regression.
- There is one or more regressors that may be continuous, ordinal, or nominal.
- We assume a **linear relationship** between any continuous regressors and the logit of the cumulative probabilities of the outcome.
- The data points do not exhibit any correlation structures, which means that there is statistical independence across the sampled observations, with mutually exclusive and exhaustive ordered categories.

However, Multinomial Logistic regression should not be used in the following scenarios:

- The outcome variable is **ordered**. If the categories are ordered, use [Ordinal Logistic regression](https://alexrod61.github.io/regression-cookbook/book/15-ordinal-logistic.html).
- The outcome variable is **binary**. For binary outcomes, use [Binary Logistic regression](https://alexrod61.github.io/regression-cookbook/book/08-binary-logistic.html).
- The outcome variable is **count**. For count outcomes, refer to count regression models, such as [Classical Poisson regression](https://alexrod61.github.io/regression-cookbook/book/10-classical-poisson.html) and [Negative Binomial regression](https://alexrod61.github.io/regression-cookbook/book/10-negative-binomial.html).
- The outcome variable is **continuous**. For continuous outcomes, refer to the ["continuous cuisine"](https://alexrod61.github.io/regression-cookbook/book/continuous-zone.html) section of this book. For example, consider using [Ordinary Least-squares regression](https://alexrod61.github.io/regression-cookbook/book/03-ols.html) for continuous and unbounded outcomes.
- The data points exhibit correlation structures, indicating statistical dependence among the sampled observations (for example, repeated measurements on the same subject).
::::
:::

::: {#fig-multinomial-regression}
```{mermaid}
mindmap
  root((Regression 
  Analysis)
    Continuous <br/>Outcome Y
      {{Unbounded <br/>Outcome Y}}
        )Chapter 3: <br/>Ordinary <br/>Least Squares <br/>Regression(
          (Normal <br/>Outcome Y)
      {{Nonnegative <br/>Outcome Y}}
        )Chapter 4: <br/>Gamma Regression(
          (Gamma <br/>Outcome Y)
      {{Bounded <br/>Outcome Y <br/> between 0 and 1}}
        )Chapter 5: Beta <br/>Regression(
          (Beta <br/>Outcome Y)
      {{Nonnegative <br/>Survival <br/>Time Y}}
        )Chapter 6: <br/>Parametric <br/> Survival <br/>Regression(
          (Exponential <br/>Outcome Y)
          (Weibull <br/>Outcome Y)
          (Lognormal <br/>Outcome Y)
        )Chapter 7: <br/>Semiparametric <br/>Survival <br/>Regression(
          (Cox Proportional <br/>Hazards Model)
            (Hazard Function <br/>Outcome Y)
    Discrete <br/>Outcome Y
      {{Binary <br/>Outcome Y}}
        {{Ungrouped <br/>Data}}
          )Chapter 8: <br/>Binary Logistic <br/>Regression(
            (Bernoulli <br/>Outcome Y)
        {{Grouped <br/>Data}}
          )Chapter 9: <br/>Binomial Logistic <br/>Regression(
            (Binomial <br/>Outcome Y)
      {{Count <br/>Outcome Y}}
        {{Equidispersed <br/>Data}}
          )Chapter 10: <br/>Classical Poisson <br/>Regression(
            (Poisson <br/>Outcome Y)
        {{Overdispersed <br/>Data}}
          )Chapter 11: <br/>Negative Binomial <br/>Regression(
            (Negative Binomial <br/>Outcome Y)
        {{Zero Inflated <br/>Data}}
          )Chapter 12: <br/>Zero Inflated <br/>Poisson <br/>Regression(
            (Zero Inflated <br/>Poisson <br/>Outcome Y)
        {{Overdispersed or <br/>Underdispersed <br/>Data}}
          )Chapter 13: <br/>Generalized <br/>Poisson <br/>Regression(
            (Generalized <br/>Poisson <br/>Outcome Y)
      {{Categorical <br/>Outcome Y}}
        {{Nominal <br/>Outcome Y}}
          )Chapter 14: <br/>Multinomial <br/>Logistic <br/>Regression(
            (Multinomial <br/>Outcome Y)
```
:::

```{r setup-r, include=FALSE}
# Load required libraries
library(devtools)
library(readr)
library(tibble)
library(tidyr)
library(dplyr)
library(magrittr)
library(ggplot2)
library(gridExtra) # combine ggplot2 figures
library(ggExtra) # draw marginal histograms
library(broom)
library(knitr)
#devtools::install_github("andytai7/cookbook")
library(cookbook)
library(nnet) # fit multinomial regression
library(caret) # upsampling and confusion matrix
library(rsample)

library(reticulate)
py_config()

py_require("pandas")
py_require("numpy")
py_require("seaborn")
py_require("pyreadr")
py_require("plotnine")
py_require("matplotlib")
py_require("statsmodels")
py_require("scikit-learn")
py_require("spicy")

reticulate::py_install("plotnine")
reticulate::py_install("imblearn")

pandas <- import("pandas")
numpy <- import("numpy")
seaborn <- import("seaborn")
pyreadr <- import("pyreadr")
matplotlib <- import("matplotlib")
statsmodels <- import("statsmodels")
sklearn <- import("sklearn")
scipy <- import("scipy")
plotnine <- import("plotnine")
imblearn <- import("imblearn")

print_as_dt <- function(x){
  DT::datatable(x,
                rownames = FALSE,
                options = list(
                  dom = 'ltipr',
                  autoWidth = TRUE,
                  columnDefs = list(
                    list(className = 'dt-left', targets = "_all"),
                    list(width = '200px', targets = "_all")
                    )
                  )
                )
}
```

```{python setup-py, include=FALSE}
import imblearn
import numpy as np
import pandas as pd
from plotnine import *
from statsmodels.formula.api import glm, ols, mnlogit
from statsmodels.genmod.families import Poisson
from matplotlib import pyplot as plt
from sklearn.metrics import mean_squared_error, mean_absolute_error, classification_report
import scipy.stats as stats
import seaborn as sns
from matplotlib.gridspec import GridSpec
from sklearn.linear_model import LogisticRegression
from sklearn.metrics import confusion_matrix, ConfusionMatrixDisplay, classification_report, log_loss, accuracy_score, recall_score, f1_score
from sklearn.utils import resample
import statsmodels.api as sm

def scrollable_table_html(data, **kwargs):
    kwargs.setdefault("index", False)
    kwargs.setdefault("border", 0)

    return (
        '<div class="table-responsive">'
        + data.to_html(
            **kwargs
        )
        + '</div>'
    )
```

::: {.LO}
::::{.LO-header}
Learning Objectives
::::
::::{.LO-container}
By the end of this chapter, you will be able to:

- Describe why **Ordinary Least Squares** is not appropriate for multi-class nominal outcomes.
- Determine when Multinomial Logistic regression is an **appropriate modelling choice.**
- Frame **inferential** and **predictive inquiries** for a Multinomial Logistic regression problem using the **data science workflow**.
- Write down the **likelihood function** for Multinomial Logistic regression and its **logit link** functions, and obtain category-specific probabilities.
- Understand how coefficient estimation via **maximum likelihood estimation** works in the Multinomial Logistic regression model.
- Fit a Multinomial Logistic regression model in both `R` and `Python`.
- Interpret Multinomial Logistic regression coefficients in real scenarios.
- Use Multinomial Logistic regression for prediction by computing category-specific probabilities.
- Evaluate model performance using both **likelihood-based** and **classification-based** measures, and construct confidence intervals using **Wald Z statistic**.
::::
:::

## Introduction

Multinomial Logistic regression is a statistical modelling technique used to analyze relationships between a multi-class categorical response variable and a set of regressors. Unlike [Binary Logistic regression](http://localhost:54321/book/08-binary-logistic.html), which models outcomes with two categories, Multinomial Logistic regression extends the framework to outcomes with more than two unordered categories. It estimates the probability of each possible outcome as a function of the regressors by modelling the log-odds of each category relative to a reference category. This method is widely used in fields such as transportation, marketing, and social sciences where the goal is to understand or predict categorical choices among multiple alternatives.

### Assumptions
The following are the assumptions that we need to be aware of before fitting this model:

- The response or dependent variable should be **categorical** and **nominal** with more than two categories. 
- There should be one or more independent variables that are continuous, ordinal or nominal (including dichotomous variables).
- **Logit linearity assumption**: There is be a linear relationship between any continuous independent variables and the logit of the category-specific probabilities of the dependent variable. This assumption will be explained in @sec-data-modelling.
- **Independent observations**: The observations are independent, and the dependent variable consists of mutually exclusive and exhaustive categories (i.e. no individual belonging to two different categories).
- **No Multicollinearity**: Multicollinearity occurs when you have two or more independent variables that are highly correlated with each other.
- There should be no outliers, high leverage values or highly influential points.

### Use Cases of Multinomial Logistic Regression
Multinomial Logistic regression is particularly useful when the outcome
variable has three or more categories with no meaningful ranking. Some
examples of this are:

1.  **Travel Mode Choice:** Transportation planners model how commuters
    select among alternatives such as `Bus`, `Car`,
    `Bike`, and `Walk`. Regressors typically include travel time,
    travel cost, distance, and household characteristics, and the fitted
    probabilities are used to forecast shifts in commute patterns under proposed
    infrastructure or pricing changes.

2.  **Voting and Party Choice:** Political scientists study how
    individual characteristics such as age, income, gender, and
    education influence which party a voter supports in a multiparty
    election (for example, `Liberal` / `Conservative` / `NDP` / `Green`). No party is "higher" than another, so      the outcome categories are purely nominal.

3.  **Consumer Brand Choice:** Marketing analysts use scanner panel data
    to model which brand a household purchases within a product
    category. Regressors such as brand loyalty, regular shelf price,
    promotional price cuts, and the presence of in-store promotions are
    used to estimate price elasticities and predict market share.

4.  **Disease Subtype Classification:** Epidemiologists often treat a
    heterogeneous disease as several distinct subtypes rather than a
    single binary outcome. For instance, breast cancer may be modelled
    across molecular subtypes defined by receptor status, with age,
    family history, body mass index, and breast density as regressors,
    allowing risk factors to act differently on each subtype.

5.  **Occupational Choice:** Labour economists model the occupational
    category an individual enters (for example, `professional`,
    `clerical`, `service`, or `manual`) as a function of education,
    work experience, parental occupation, and other socioeconomic
    characteristics.

In all these examples, the categories are mutually exclusive and
exhaustive but carry no natural ordering. Imposing an order, as [Ordinal Logistic regression](https://alexrod61.github.io/regression-cookbook/book/15-ordinal-logistic.html) does, would assume a progression between categories
that does not exist, whereas fitting a separate [Binary Logistic regression model](https://alexrod61.github.io/regression-cookbook/book/08-binary-logistic.html) for each
pair of categories would discard information and could yield predicted
probabilities that do not sum to one. Multinomial logistic regression
provides a principled framework that models all categories jointly
against a common reference level.

## Case Study: Understanding Commuting Behaviors

To demonstrate Multinomial Logistic regression, we will use a simulated dataset containing information on individuals' commuting behaviors and demographics, with a focus on their primary mode of transportation. This dataset provides an example of a multi-class categorical response variable. We will approach this case study using the [data science workflow](https://alexrod61.github.io/regression-cookbook/book/01-intro.html) to ensure a structured approach to problem-solving and model building.

### The Dataset

Our dataset captures various characteristics of commuters and each commuter's primary mode of transport. Each row represents a commuter, and the columns describe different characteristics. @tbl-variables provides the relevant details for the regressors in this dataset - namely, commute distance, age, car availability, and whether the individual commutes during weekends. Note that we also indicate the **coding names** of all the variables involved in the last column.

| Variable | Description | Type | Scale | Model Role | Coding Name |
|---|---|---|---|---|---|---|
| Mode of Transport | The primary mode of transport used by the individual | Discrete | Categorical | Response | `transport_mode` |
| Commute Distance | The one-way commuting distance in kilometers | Continuous | Bounded positive | Regressor | `commute_distance` |
| Age | The age of the individual in years | Continuous | Bounded positive | Regressor | `age` |
| Car Availability | Whether the individual has access to a car | Discrete| Binary | Regressor | `has_car_available` |
| Weekend Commuter | Whether the individual commutes during weekends | Discrete | Binary | Regressor | `weekend_commuter` |

: Classification table for variables in the transport dataset. {#tbl-variables}

Let's load this sample of $n=1000$ commuters in both `R` and `Python` from the local file `data/multinom_transport.rda`. In `R`, we read the `.rda` file with `load()` and wrangle it with the `tidyverse` library, and in `Python`, we read the same `.rda` file with the `pyreadr` library and wrangle it with the `pandas` library. 

::: {.panel-tabset}

## R Code

```{.r}
load("data/multinom_transport.rda")
data <- multinomial_transport[,-1]
cat("There are", nrow(data), "observations and", ncol(data), "variables in the dataset.")
head(data, n = 100) |>
  print_as_dt()
```

## Python Code

```{.python}
import pyreadr

rda = pyreadr.read_r("data/multinom_transport.rda")
data = rda["multinomial_transport"]
data = data.iloc[:, :-1]
print(f"There are {data.shape[0]} observations and {data.shape[1]} variables in the dataset.")
print(data)
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
load("data/multinom_transport.rda")
data <- multinomial_transport %>% select(-'income_bracket')
cat("There are", nrow(data), "observations and", ncol(data), "variables in the dataset.")
head(data, n = 100) |>
  print_as_dt()
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
import pyreadr

rda = pyreadr.read_r("data/multinom_transport.rda")
data = rda["multinomial_transport"]
data = data.iloc[:, :-1]
print(f"There are {data.shape[0]} observations and {data.shape[1]} variables in the dataset.")
```

```{r}
#| echo: false
#| message: false
#| warning: false
head(data, n = 100) |>
  print_as_dt()
```

:::

This dataset provides a structured way to analyze the behaviors of commuters and determine which factors are associated with their modes of transport. 

### Study Design {#sec-study-design}

Now that we’ve introduced our case study and dataset, it is time to follow the [data science workflow](https://alexrod61.github.io/regression-cookbook/book/01-intro.html) step by step. The first step is to define the main statistical inquiries we want to address. Our goal in this case study is to understand which factors impact an individual's mode of transport. We aim to identify which characteristics, such as age, commute distance, having a car, and commuting on weekends or not, explain whether people commute by public transit, car, bicycle, or walk. 

We will explore two primary questions based on this dataset.

- How are an individual's age, commute distance, availability of car, and whether they commute on weekends or not, **associated** with their mode of transport?
- Can we accurately **predict** the transport mode of an individual based on their commuting behaviors and demographics including if they have a car, if they commute on weekends, their commute distance, and age?

These questions correspond to two distinct **statistical objectives**:

- **Inference:** Understand the relationship between regressors (`age`, `commute_distance`, `has_car_available` and `weekend_commuter`) and the probability of each category of the response variable `transport_mode`.
- **Prediction:** Develop a classification model that classifies individuals into their most probable mode of transport: `Bicycle`, `Car`, `Public Transit`, or `Walking` based on factors like age, commute distance, availability of a car, and whether the person commutes on weekends or not. 

## Data Collection and Wrangling

The dataset used in this chapter is a synthetic dataset from the [cookbook](https://github.com/andytai7/cookbook) package in `R`. Now that our dataset is loaded, the next step is to clean and organze it. Data wrangling involves several steps that ensure our data is accurate, consistent, and ready for modelling. Here are some key tasks:

#### Handling Missing Data

The first task is to ensure data integrity by checking for missing values. Missing data can occur for various reasons, such as unrecorded responses or errors in data entry. When we find missing values—for example, if some individuals don’t have recorded age or commute distance—we must decide how to handle these gaps. Common strategies include:

- **Removing incomplete records**: If the amount of missing data is minimal or missingness is random.
- **Imputing missing values**: Using logical estimates or averages if missingness follows a systematic pattern.

In our simulated dataset, there are no missing values, as confirmed by:

::: {.panel-tabset}
## **`R` Code**
```{r}
colSums(is.na(data))
```

## **`Python` Code**
```{python}
# Count missing values in each column
data.isna().sum()
```
:::

#### Encoding Categorical Variables

For regression analysis, we need to convert categorical variables into numerical representations. In the following code in both `R` and `Python`, we will factor all categorical variables, such as `transport_mode`, `has_car_available` and `weekend_commuter`, so that the model correctly interprets them as categorical data rather than continuous numbers.

::: {.panel-tabset}

## R Code

```{.r}
# Factorize the categorical variables and print their levels
categorical_cols <- c('transport_mode', 'has_car_available', 'weekend_commuter')
for (col in categorical_cols) {
  data[[col]] <- as.factor(data[[col]])
  cat("Levels of", col, ":", levels(data[[col]]), "\n")
}
```

## Python Code

```{.python}
# Factorize the categorical variables and print their levels
categorical_cols = ['transport_mode', 'has_car_available', 'weekend_commuter']
for col in categorical_cols:
  data[col] = data[col].astype('category')
  print(f"Levels of {col}: {list(data[col].cat.categories)}")
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
categorical_cols <- c('transport_mode', 'has_car_available', 'weekend_commuter')
for (col in categorical_cols) {
  data[[col]] <- as.factor(data[[col]])
  cat("Levels of", col, ":", levels(data[[col]]), "\n")
}
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
categorical_cols = ['transport_mode', 'has_car_available', 'weekend_commuter']
for col in categorical_cols:
  data[col] = data[col].astype('category')
  print(f"Levels of {col}: {list(data[col].cat.categories)}")
```

:::

#### Splitting the Data for Model Training

To ensure that our model generalizes well to unseen data, we will split the dataset into train and test sets. We will use the train set to fit the model and to estimate the model parameters, and the test set to assess model performance for both **inference** and **prediction**. The split is typically done in an 80/20 ratio, as shown below:

::: {.panel-tabset}

## R Code

```{.r}
# Randomly split the data into train and test sets
set.seed(42)
train_idx <- sample(seq_len(nrow(data)), size = 0.8 * nrow(data))
train_set <- data[train_idx, ]
test_set  <- data[-train_idx, ]

cat("train set size:", nrow(train_set))
cat("Test set size:", nrow(test_set))
```

## Python Code

```{.python}
# The train/test split is created once in R (caret::createDataPartition; see the
# R tab) and pulled into Python via reticulate's `r` object, so both languages
# use identical train and test sets throughout the chapter.
train_set = r.train_set.reset_index(drop=True)
test_set  = r.test_set.reset_index(drop=True)

print(f"train set size: {len(train_set)}")
print(f"Test set size: {len(test_set)}")
```

:::

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on keeping `R` and `Python` aligned after the split!
::::
:::: {.Heads-up-container}
Using the same seed and split proportion in `R` and `Python` does not guarantee that the random splits will contain identical rows. This is not an error. The two ecosystems use different splitting functions and pseudo-random number generators. Therefore, for the rest of this chapter, we use the `R`-generated train and test sets as a common reference to ensure consistent coding outputs.
::::
:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
set.seed(352)

# Randomly split the data into train and test sets
train_idx <- sample(seq_len(nrow(data)), size = 0.8 * nrow(data))
train_set <- data[train_idx, ]
test_set  <- data[-train_idx, ]

cat("train set size:", nrow(train_set))
cat("Test set size:", nrow(test_set))
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
# The train/test split is created once in R (caret::createDataPartition; see the
# R tab) and pulled into Python via reticulate's `r` object, so both languages
# use identical train and test sets throughout the chapter.
train_set = r.train_set.reset_index(drop=True)
test_set  = r.test_set.reset_index(drop=True)

print(f"train set size: {len(train_set)}")
print(f"Test set size: {len(test_set)}")
```

:::

## Exploratory Data Analysis {#sec-eda}

Before diving into data modelling, it is crucial to develop a deep understanding of the relationships between variables in the dataset. The Exploratory Data Analysis (EDA) stage helps us visualize and summarize the data, uncover patterns, detect anomalies, and test key assumptions that will inform our modelling decisions.

### Continuous Variables

We begin by examining continuous variables. There are two continuous regressors: `age` and `commute_distance`, and we will examine their relationship with the response variable `transport_mode` through visualizations, such as box plots and violin plots. 

::: {.panel-tabset}

## R Code

```{.r}
# Custom color palette `cbPalette` 
cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# Visualize boxplots and violin plots of commute_distance v.s. transport_mode
p1 <- train_set %>%
    ggplot(aes(x = transport_mode, y = commute_distance, fill = transport_mode)) +
    geom_violin(alpha = 0.6, trim = FALSE) +
    stat_boxplot(geom = "errorbar", width = 0.2) +
    geom_boxplot(alpha = 1, fill = NA) +
    labs(x = "Transport Mode", y = "Commute Distance (km)", title =
        "Commute Distance by Transport Mode") +
    scale_fill_manual(values = cbPalette[3:6]) +
    theme_bw() +
    theme(legend.position = "none") +
    guides(fill = guide_legend(title = "Transport Mode"))

p2 <- train_set %>%
    ggplot(aes(x = transport_mode, y = age, fill = transport_mode)) +
    geom_violin(alpha = 0.6, trim = FALSE) +
    stat_boxplot(geom = "errorbar", width = 0.2) +
    geom_boxplot(alpha = 1, fill = NA) +
    labs(x = "Transport Mode", y = "Age", title =
        "Age by Transport Mode") +
    scale_fill_manual(values = cbPalette[3:6]) +
    theme_bw() +
    theme(legend.position = "none") +
    guides(fill = guide_legend(title = "Transport Mode"))

grid.arrange(p1, p2, ncol = 2, top = grid::textGrob("Box Plots and Violin Plots", gp = grid::gpar(fontsize = 16, fontface = "bold")))
```

## Python Code

```{.python}
sns.set_theme(style="whitegrid")

# Custom color palette `cbPalette` 
cbPalette = ["#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7"]

# Sort transport_mode by frequency
order = train_set['transport_mode'].value_counts().index
train_set['transport_mode'] = pd.Categorical(train_set['transport_mode'], categories=order, ordered=True)

# Set up side-by-side subplots with shared layout
fig = plt.figure(figsize=(8, 7))
gs = GridSpec(1, 2, figure=fig)
fig.suptitle("Box Plots and Violin Plots", fontsize=16, fontweight='bold')

# Plot Commute Distance vs Transport Mode
ax1 = fig.add_subplot(gs[0, 0])
sns.violinplot(
    data=train_set,
    x='transport_mode',
    y='commute_distance',
    palette=cbPalette[2:6],
    ax=ax1,
    alpha=0.6,
    inner=None,
    order=order,
    cut=3
)
sns.boxplot(
    data=train_set,
    x='transport_mode',
    y='commute_distance',
    showcaps=True,
    boxprops={'facecolor': 'none', 'edgecolor': 'black'},
    whiskerprops={'color': 'black'},
    medianprops={'color': 'black', 'linewidth': 2},
    flierprops={'markerfacecolor': 'black', 'markersize': 3},
    ax=ax1,
    order=order
)
ax1.set_title("Commute Distance by Transport Mode")
ax1.set_xlabel("Transport Mode")
ax1.set_ylabel("Commute Distance (km)")
ax1.get_legend()

# Plot Age vs Transport Mode
ax2 = fig.add_subplot(gs[0, 1])
sns.violinplot(
    data=train_set,
    x='transport_mode',
    y='age',
    palette=cbPalette[2:6],
    ax=ax2,
    alpha=0.6,
    inner=None,
    order=order,
    cut=3
)
sns.boxplot(
    data=train_set,
    x='transport_mode',
    y='age',
    showcaps=True,
    boxprops={'facecolor': 'none', 'edgecolor': 'black'},
    whiskerprops={'color': 'black'},
    medianprops={'color': 'black', 'linewidth': 2},
    flierprops={'markerfacecolor': 'black', 'markersize': 3},
    ax=ax2,
    order=order
)
ax2.set_title("Age by Transport Mode")
ax2.set_xlabel("Transport Mode")
ax2.set_ylabel("Age")
ax2.get_legend()

fig.tight_layout(rect=[0, 0.05, 1, 0.95])
fig
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| label: fig-eda-transport-mode-continuous-R
#| fig-cap:
#|   - "Boxplots and violin plots of commute distance and age by transport mode."
#| echo: false
#| message: false
#| warning: false
cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
p1 <- train_set %>%
    ggplot(aes(x = transport_mode, y = commute_distance, fill = transport_mode)) +
    geom_violin(alpha = 0.6, trim = FALSE) +
    stat_boxplot(geom = "errorbar", width = 0.2) +
    geom_boxplot(alpha = 1, fill = NA) +
    labs(x = "Transport Mode", y = "Commute Distance (km)", title =
        "Commute Distance by Transport Mode") +
    scale_fill_manual(values = cbPalette[3:6]) +
    theme_bw() +
    theme(legend.position = "none") +
    guides(fill = guide_legend(title = "Transport Mode"))
p2 <- train_set %>%
    ggplot(aes(x = transport_mode, y = age, fill = transport_mode)) +
    geom_violin(alpha = 0.6, trim = FALSE) +
    stat_boxplot(geom = "errorbar", width = 0.2) +
    geom_boxplot(alpha = 1, fill = NA) +
    labs(x = "Transport Mode", y = "Age", title =
        "Age by Transport Mode") +
    scale_fill_manual(values = cbPalette[3:6]) +
    theme_bw() +
    theme(legend.position = "none") +
    guides(fill = guide_legend(title = "Transport Mode"))
grid.arrange(p1, p2, ncol = 2, top = grid::textGrob("Box Plots and Violin Plots", gp = grid::gpar(fontsize = 16, fontface = "bold")))
```

## Python Output

```{python}
#| label: fig-eda-transport-mode-continuous-py
#| fig-cap:
#|   - "Boxplots and violin plots of commute distance and age by transport mode."
#| echo: false
#| message: false
#| warning: false
sns.set_theme(style="whitegrid")

# Custom color palette `cbPalette` 
cbPalette = ["#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7"]

# Sort transport_mode by frequency
order = train_set['transport_mode'].value_counts().index
train_set['transport_mode'] = pd.Categorical(train_set['transport_mode'], categories=order, ordered=True)

# Set up side-by-side subplots with shared layout
fig = plt.figure(figsize=(8, 7))
gs = GridSpec(1, 2, figure=fig)
fig.suptitle("Box Plots and Violin Plots", fontsize=16, fontweight='bold')

# Plot Commute Distance vs Transport Mode
ax1 = fig.add_subplot(gs[0, 0])
sns.violinplot(
    data=train_set,
    x='transport_mode',
    y='commute_distance',
    palette=cbPalette[2:6],
    ax=ax1,
    alpha=0.6,
    inner=None,
    order=order,
    cut=3
)
sns.boxplot(
    data=train_set,
    x='transport_mode',
    y='commute_distance',
    showcaps=True,
    boxprops={'facecolor': 'none', 'edgecolor': 'black'},
    whiskerprops={'color': 'black'},
    medianprops={'color': 'black', 'linewidth': 2},
    flierprops={'markerfacecolor': 'black', 'markersize': 3},
    ax=ax1,
    order=order
)
ax1.set_title("Commute Distance by Transport Mode")
ax1.set_xlabel("Transport Mode")
ax1.set_ylabel("Commute Distance (km)")
ax1.get_legend()  # no legend to remove manually

# Plot p2: Age vs Transport Mode
ax2 = fig.add_subplot(gs[0, 1])
sns.violinplot(
    data=train_set,
    x='transport_mode',
    y='age',
    palette=cbPalette[2:6],
    ax=ax2,
    alpha=0.6,
    inner=None,
    order=order,
    cut=3
)
sns.boxplot(
    data=train_set,
    x='transport_mode',
    y='age',
    showcaps=True,
    boxprops={'facecolor': 'none', 'edgecolor': 'black'},
    whiskerprops={'color': 'black'},
    medianprops={'color': 'black', 'linewidth': 2},
    flierprops={'markerfacecolor': 'black', 'markersize': 3},
    ax=ax2,
    order=order
)
ax2.set_title("Age by Transport Mode")
ax2.set_xlabel("Transport Mode")
ax2.set_ylabel("Age")
ax2.get_legend()  # no legend to remove manually

fig.tight_layout(rect=[0, 0.05, 1, 0.95])
fig
```

:::

The plots in @fig-eda-transport-mode-continuous-R and @fig-eda-transport-mode-continuous-py show how commute distance and age vary across different transport modes. Public transit users tend to have the longest commute distance, on average, with a wide distribution and a median around 17–18 km. Car users also have moderately long commutes, with a median around 10-11 km and a relatively broad spread. Bicycle and Walking are associated with shorter commute distances (medians < 5 km), and their distributions are tightly concentrated near the lower end.
All transport modes show some long-distance commuters, which are shown by long tails, and there are a few outliers for commute distance in `Bicycle`.

Car and public transit users have similar age distributions, with median ages in the mid-30s to early 40s.
Bicycle users have a more skewed age distribution with a higher median age in mid-50s. Walking has a much wider spread in age, with a high median (around 60) and some very young walkers as well. The variability in age is the highest for walking, which suggests diverse usage across age groups.

### Categorical Variables

The categorical variables can be visualized using stacked bar charts. The bar charts are very effective because they display the frequency of each category, helping us understand the distribution of qualitative attributes. The stacked bar charts in @fig-eda-transport-mode-distribution-R and @fig-eda-transport-mode-distribution-py display the distributions of transport mode across car availability (`has_car_available`) and weekend commuter status (`weekend_commuter`). 

::: {.panel-tabset}

## R Code

```{.r}
# Sort `transport_mode` by frequency, so the fill levels stack in a stable order
train_set$transport_mode <- factor(train_set$transport_mode, levels = names(sort(table(train_set$transport_mode), decreasing = TRUE)))

p1 <- train_set %>%
  ggplot(aes(x = has_car_available, fill = transport_mode)) +
  geom_bar(position = position_stack(reverse = TRUE)) +  # use "stack" for counts, "fill" for proportions
  labs(x = "Car Availability", y = "Count", title = "With Car Availability") +
  scale_fill_manual(values = cbPalette) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(title = "Transport Mode"))

p2 <- train_set %>%
  ggplot(aes(x = weekend_commuter, fill = transport_mode)) +
  geom_bar(position = position_stack(reverse = TRUE)) +  # use "stack" for counts, "fill" for proportions
  labs(x = "Weekend Commuter", y = "", title = "With Weekend Commuter Status") +
  scale_fill_manual(values = cbPalette) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(title = "Transport Mode"))

grid.arrange(p1, p2, ncol = 2, top = grid::textGrob("Distribution of Transport Mode", gp = grid::gpar(fontsize = 16, fontface = "bold")))
```

## Python Code

```{.python}
# Set up the plot
sns.set_theme(style="whitegrid")
fig = plt.figure()
fig.set_size_inches(8, 6)
gs = GridSpec(1, 2, figure=fig)
fig.suptitle("Distribution of Transport Mode", fontsize=16, fontweight='bold')

# Plot: Stacked bar for Car Availability
car_counts = train_set.groupby(['has_car_available', 'transport_mode']).size().unstack(fill_value=0)
car_counts = car_counts[order]

ax1 = fig.add_subplot(gs[0, 0])
bottom = None
for i, col in enumerate(car_counts.columns):
    ax1.bar(car_counts.index, car_counts[col], bottom=bottom, label=col, color=cbPalette[i % len(cbPalette)])
    bottom = car_counts[col] if bottom is None else bottom + car_counts[col]
ax1.set_title("With Car Availability")
ax1.set_xlabel("Car Availability")
ax1.set_ylabel("Count")
ax1.legend(title="Transport Mode", loc='upper center', bbox_to_anchor=(0.5, -0.15), ncol=2)

# Plot: Stacked bar for Weekend Commuter
weekend_counts = train_set.groupby(['weekend_commuter', 'transport_mode']).size().unstack(fill_value=0)
weekend_counts = weekend_counts[order]

ax2 = fig.add_subplot(gs[0, 1])
bottom = None
for i, col in enumerate(weekend_counts.columns):
    ax2.bar(weekend_counts.index, weekend_counts[col], bottom=bottom, label=col, color=cbPalette[i % len(cbPalette)])
    bottom = weekend_counts[col] if bottom is None else bottom + weekend_counts[col]
ax2.set_title("With Weekend Commuter Status")
ax2.set_xlabel("Weekend Commuter")
ax2.set_ylabel("")
ax2.legend(title="Transport Mode", loc='upper center', bbox_to_anchor=(0.5, -0.15), ncol=2)

# Adjust plot layout
fig.tight_layout(rect=[0, 0.05, 1, 1])
fig
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| label: fig-eda-transport-mode-distribution-R
#| fig-cap:
#|   - "Stacked bar charts of transport mode counts by car availability and weekend commuter status."
#| message: false
#| warning: false
# Sort `transport_mode` by frequency, so the fill levels stack in a stable order
train_set$transport_mode <- factor(train_set$transport_mode, levels = names(sort(table(train_set$transport_mode), decreasing = TRUE)))

p1 <- train_set %>%
  ggplot(aes(x = has_car_available, fill = transport_mode)) +
  geom_bar(position = position_stack(reverse = TRUE)) +  # use "stack" for counts, "fill" for proportions
  labs(x = "Car Availability", y = "Count", title = "With Car Availability") +
  scale_fill_manual(values = cbPalette) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(title = "Transport Mode"))

p2 <- train_set %>%
  ggplot(aes(x = weekend_commuter, fill = transport_mode)) +
  geom_bar(position = position_stack(reverse = TRUE)) +  # use "stack" for counts, "fill" for proportions
  labs(x = "Weekend Commuter", y = "", title = "With Weekend Commuter Status") +
  scale_fill_manual(values = cbPalette) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(title = "Transport Mode"))

grid.arrange(p1, p2, ncol = 2, top = grid::textGrob("Distribution of Transport Mode", gp = grid::gpar(fontsize = 16, fontface = "bold")))
```

## Python Output

```{python}
#| label: fig-eda-transport-mode-distribution-py
#| fig-cap:
#|   - "Stacked bar charts of transport mode counts by car availability and weekend commuter status."
#| echo: false
#| message: false
#| warning: false
# Set up the plot
sns.set_theme(style="whitegrid")
fig = plt.figure()
fig.set_size_inches(8, 6)
gs = GridSpec(1, 2, figure=fig)
fig.suptitle("Distribution of Transport Mode", fontsize=16, fontweight='bold')

# Plot: Stacked bar for Car Availability
car_counts = train_set.groupby(['has_car_available', 'transport_mode']).size().unstack(fill_value=0)
car_counts = car_counts[order]

ax1 = fig.add_subplot(gs[0, 0])
bottom = None
for i, col in enumerate(car_counts.columns):
    ax1.bar(car_counts.index, car_counts[col], bottom=bottom, label=col, color=cbPalette[i % len(cbPalette)])
    bottom = car_counts[col] if bottom is None else bottom + car_counts[col]
ax1.set_title("With Car Availability")
ax1.set_xlabel("Car Availability")
ax1.set_ylabel("Count")
ax1.legend(title="Transport Mode", loc='upper center', bbox_to_anchor=(0.5, -0.15), ncol=2)

# Plot: Stacked bar for Weekend Commuter
weekend_counts = train_set.groupby(['weekend_commuter', 'transport_mode']).size().unstack(fill_value=0)
weekend_counts = weekend_counts[order]

ax2 = fig.add_subplot(gs[0, 1])
bottom = None
for i, col in enumerate(weekend_counts.columns):
    ax2.bar(weekend_counts.index, weekend_counts[col], bottom=bottom, label=col, color=cbPalette[i % len(cbPalette)])
    bottom = weekend_counts[col] if bottom is None else bottom + weekend_counts[col]
ax2.set_title("With Weekend Commuter Status")
ax2.set_xlabel("Weekend Commuter")
ax2.set_ylabel("")
ax2.legend(title="Transport Mode", loc='upper center', bbox_to_anchor=(0.5, -0.15), ncol=2)

# Adjust plot layout
fig.tight_layout(rect=[0, 0.05, 1, 1])
fig
```

:::

The bar charts show that public transit is the dominant mode of transport, especially among those who do not have access to a car. Public transit is used by both those with and without car availability, but it is more common among those without a car, as shown by a relatively larger orange bar. `Bicycle` and `Walking` are minority classes in the train set, as shown by the tiny bars in  @fig-eda-transport-mode-distribution-R and @fig-eda-transport-mode-distribution-py.

Most weekend commuters do not significantly change their transport mode. The overall proportions in the train set are largely dominated by the non-weekend commuters, shown by the orange bars. However, `Public Transit` has a noticeable number of weekend commuters, shown by the blue bar, which signals some weekend reliance on public transit. Among the `Bicycle` riders, the majority have access to a car and the majority of them are not weekend commuters, according to the stacked bar charts. 

The number of observations with `transport_mode=Walking` is very small in the train set, as shown by the minuscule bars corresponding to individuals with Walking as the primary mode of transport. As the distribution of categorical variables is not clear for `Walking`, let's filter the corresponding rows from train set. 

::: {.panel-tabset}

## R Code

```{.r}
# Filter the rows in train set with transport_mode = "Walking"
print(filter(train_set, transport_mode=="Walking"))
```

## Python Code

```{.python}
# Filter the rows in train set with transport_mode = "Walking"
print(train_set[train_set["transport_mode"] == "Walking"])
```

:::
::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
print(filter(train_set, transport_mode=="Walking"))
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
print(train_set[train_set["transport_mode"] == "Walking"])
```

:::

As shown in the outputs above, there are $4$ observations with `transport_mode=Walking` in the train set, which corresponds to $5\%$ of total observations in the train set. There is no clear dominance of any `has_car_available` and `weekend_commuter` categories for individuals with Walking as the primary mode of transport.

You can also visualize the relationship across other regressors in various types of figures for practice.

### Key Takeaways from EDA

Conducting exploratory data analysis (EDA) allows us to gain an initial understanding of the data and its underlying patterns before moving on to model building. Through EDA, we identify the types of variables present, examine their distributions, and uncover potential issues such as skewness, outliers, or multicollinearity. This process helps to highlight which variables might be strong predictors and which may require additional transformation or treatment. For example, a correlation between `has_car_available` and Car versus Public Transit as the transport mode signals that the availability of car can potentially be a factor for choosing car over public transit. Similarly the median commute distance increases across `Walking`, `Bicycle`, `Car`, and `Public Transit` in the train set, which suggests that can be a relationship between commute distance and primary mode of transport in the population of commuters.

It is important to remember that the insights gained from EDA are preliminary and primarily serve to inform further analysis. When we explore relationships between only two variables, we might overlook the influence of other factors, which could lead to misleading conclusions if taken in isolation. EDA is a crucial step for forming initial hypotheses and guiding decisions regarding data transformations, feature engineering, and the overall modelling strategy. With this foundation, we are better prepared to build a Multinomial Logistic regression model on data that has been carefully examined and understood.

## Data Modelling

After conducting [EDA](http://localhost:54321/book/14-multinomial-logistic.html#exploratory-data-analysis), we transition to the modelling stage, where we apply a structured approach to answer both the **inferential** and **predictive** inquiries from @sec-study-design.

A Multinomial Logistic Regression model is a suitable approach for our statistical inquiries, given that `transport_mode` is a categorical and unordered outcome variable subject to the numerical regressors `commute_distance` and `age`, and binary categorical regressors `has_car_available` and `weekend_commuter`.

The EDA showed that the choice of primary transport mode varies across several commuter characteristics, but those empirical comparisons were still descriptive. We now need a probability model that respects the unordered categorical nature of the response while allowing the conditional probability of each mode to vary systematically with the regressors.

Multinomial Logistic regression is a **GLM** with the following three-part structure:

1. a **random component**, which specifies the conditional probability distribution of the response;
2. a **systematic component**, which combines the regressors through linear predictor(s); and
3. a **link function**, which connects the conditional mean of the response to that linear predictor.

This GLM view is useful because it makes clear what remains linear in Multinomial Logistic regression and what does not. The regressors are combined linearly on the **log-odds scale**, where each non-baseline category is compared against a common baseline category, while the conditional probabilities themselves are obtained through a nonlinear transformation.

Unlike the binary case, the response here has more than two categories, so a single linear predictor is not enough. The model instead requires one linear predictor for each non-baseline category, and these predictors are normalized jointly so that the resulting probabilities across all categories sum to one.

Now, we develop these three components for our case study.

### Defining modelling parameters

Once we select Multinomial Logistic regression, we define the key modelling components: the response variable (dependent variable) and the regressors (independent variables).

#### Response Variable (Y):

The response variable, also known as the dependent variable, represents the primary mode of transport that we aim to explain:

- `transport_mode`: The dependent variable representing primary mode of transport.

#### Regressors (X):

Each regressor is chosen based on its theoretical and statistical relevance in explaining the choice of primary transport mode:

- `commute_distance` (Continuous) – Represents the one-way commuting distance in kilometers.
- `age` (Continuous) – Represents the age of the individual in years.
- `has_car_available` (Binary) – Indicates whether the individual has access to a car or not.
- `weekend_commuter` (Binary) – Indicates whether the individual commutes during weekends or not.

### The Random Component

The random component of the Multinomial Logistic regression approach assumes a [Multinomial distribution](https://www.britannica.com/science/multinomial-distribution). Let's say that we have $m$ distinct categories denoted by positive integers $1,...,m$, where $m\geq2$ and $p_{i,0},...\dots,p_{i,m-1}$ are the probabilities of belonging to categories $0,\dots,m-1$ respectively; i.e., 
$$
\begin{align*}
P(Y_i=0)=p_{i,0},\newline
\dots\newline
\dots\newline
P(Y_i=m-1)=p_{i,m-1}
\end{align*}
$$ {#eq-multinomial-dist}
where
$$
\sum_{j = 0}^{m-1} p_{i,j} = p_{i,0} + \dots + p_{i,m-1} = 1.
$$

*A particular highlight is that the Binomial distribution is a special case of the Multinomial distribution with two possible outcomes, i.e. $m=2$.*

In our study, the response variable consists of $4$ possible categories, and therefore we assume a Multinomial distribution with $m=4$ for the random component. Let $Y_i$ denote the transport mode of the $i$th commuter in the training data, for $i=1,2,...,n$, such that $Y_i$ can be one of `Public_Transit`, `Car`, `Bicycle`, or `Walking`. We will use the following coding for $Y_i$ with `Car` as the reference category:

$$
Y_i
=
\begin{cases}
0, & \text{if the primary transport mode of commuter } i \text{ is Car},\\
1, & \text{if the primary transport mode of commuter } i \text{ is Public Transit},\\
2, & \text{if the primary transport mode of commuter } i \text{ is Bicycle},\\
3, & \text{if the primary transport mode of commuter } i \text{ is Walking}.
\end{cases}
$$

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on the numerical coding of the categories!
::::

:::: {.Heads-up-container}
The values $0$, $1$, $2$, and $3$ assigned above are notational labels used to index
the four categories of the response variable. The Multinomial distribution treats
these categories as nominal, so the labels carry no ordering and no magnitude.
Assigning the labels in a different order, or selecting a different category as the
reference level, produces an equivalent reparametrization of the same model. The
estimated probabilities for each category, the maximized log-likelihood function, and the
resulting predictions are unchanged by either choice.

The estimated regression coefficients, however, are defined as **log-odds** relative to the reference category. Changing the reference level therefore changes
the numerical values of the coefficients and the comparisons they represent, along
with the associated hypothesis tests and confidence intervals. The underlying
inferential conclusions are equivalent, since any contrast between two categories can
be recovered under any choice of reference level. We select `Car` as the reference
category here for interpretability, and not because the modelling assumption strictly requires it.

If the categories of the response variable are **ordered**, and the ordering carries
information that a Multinomial model does not incorporate, then use
[Ordinal Logistic Regression](15-ordinal-logistic.qmd) in this case.
::::
:::

Let
$$
\mathbf{x}_i = (x_{i,1},...,x_{i,k}),
$$ 
denote the vector of observed regressor values for commuter $i$. As there are $4$ regressors (commute distance, age, car availability, and weekend commuter status) in our study, it follows that $k=4$. 

For a single trial $Y_i$ with regressors given by $\mathbf{x}_i$, the random component of Multinomial Logistic regression assumes

$$
Y_i \mid \mathbf{x}_i
\sim
\operatorname{Multinomial}(\mathbf{\pi}_i),
$$

where

$$
\mathbf{\pi}_i=(p_{i,1},...,p_{i,m-1})
$$
and

$$
p_{i,k}
=
\Pr(Y_i=k\mid\mathbf{x}_i)
$$
is the **conditional probability** that $Y_i$ belongs to category $k$, given the regressor values $\mathbf{x}_i$. Consequently,
$$\Pr(Y_i=0\mid\mathbf{x}_i)=1-\Sigma_{j=1}^{m-1}p_{i,j}$$
is the corresponding probability that $Y_i$ belongs to the reference category, given the regressor values $\mathbf{x}_i$.

We first encode each individual's observed response category using indicator variables. Let $y_{i,j} = 1$ if the $i$-th individual belongs to category $j$, and let $y_{i,j} = 0$ otherwise. Since each individual belongs to exactly one of the $m$ possible categories, the vector $\mathbf{y}_i = (y_{i,1}, y_{i,2}, \ldots, y_{i,m})$ contains exactly one entry equal to $1$, with all remaining entries equal to $0$. This representation is often called a one-hot encoding of the response category. For the $i$-th individual, the conditional probability mass function (PMF) can be written as

$$
p_Y(y_i;\mathbf{\pi}_i) = \Pr(Y_i=y_i\mid\mathbf{x}_i)
= \prod_{k=0}^{3} p_{i,k}^{y_i,k},
$$ {#eq-pmf}

The random component tells us **what kind of response is being modelled**, but it does not yet explain why two commuters with different characteristics may have different values of $p_{i,1}, ...,p_{i,m-1}$. That role belongs to the systematic component.

### The Systematic Component

Like [Binary Logistic regression](https://alexrod61.github.io/regression-cookbook/book/08-binary-logistic.html),
the systematic component in Multinomial Logistic regression combines the
regressors through **linear predictors**; however, only one linear predictor will
not be enough anymore. With $k=4$ regressors and $m = 4$ categories in our case
study, there are $3$ systematic components, one for
each category other than the reference category `Car`:

$$
\begin{aligned}
\eta_{i,1} &= \beta_{0,1} + \beta_{1,1} x_{i,1} + \beta_{2,1} x_{i,2}+\beta_{3,1} x_{i,3} + \beta_{4,1} x_{i,4}, \\
\eta_{i,2} &= \beta_{0,2} + \beta_{1,2} x_{i,1} + \beta_{2,2} x_{i,2}+\beta_{3,2} x_{i,3} + \beta_{4,2} x_{i,4}, \\
\eta_{i,3} &= \beta_{0,3} + \beta_{1,3} x_{i,1} + \beta_{2,3} x_{i,2}+\beta_{3,1} x_{i,3} + \beta_{4,3} x_{i,k},
\end{aligned}
$${#eq-systematic-component}

where $\eta_{i,1}$, $\eta_{i,2}$ and $\eta_{i,3}$ correspond to `Public_Transit`,
`Bicycle` and `Walking` respectively, each with respect to `Car`.

For the general case with $k$ regressors and $m$ categories, these equations can be represented as

$$
\eta_{i,j}
=
\beta_{0,j}
+
\mathbf{x}_i^\top
\boldsymbol{\beta}_j,
\qquad j = 1, \dots, m-1,
$$

where $\mathbf{x}_i$ is a $k-$dimensional vector and

$$
\boldsymbol{\beta}_j
=
\left(
\beta_{1,j},
\beta_{2,j},
\ldots,
\beta_{k,j}
\right)^\top.
$$

Thus:

- $\eta_{i,j}$ is the **linear predictor** for commuter $i$ and category $j$;
- $\beta_{0,j}$ is the intercept for category $j$;
- $\beta_{1,j}, \ldots, \beta_{k,j}$ are fixed but unknown regression
  coefficients for category $j$;
- $x_{i,1}, \ldots, x_{i,k}$ are the observed regressor values for commuter $i$;
- $\mathbf{x}_i$ collects those $k$ regressor values; and
- $\boldsymbol{\beta}_j$ collects the corresponding $k$ regression coefficients
  for category $j$.

Note that the regressor values $\mathbf{x}_i$ are the **same** across all three
equations; however, each non-reference category
has its own $\boldsymbol{\beta}_j$, which means that a regressor may push a commuter towards
one transport mode and away from another. For a general case with $m$ categories in the response variable and $k$ regressors, the model estimates $(m-1)(k+1)$
parameters in total. In our case study with $k = 4$ regressors and
$m = 4$ categories, there are $15$ parameters in total.

The reference category `Car` has no equation of its own. Its linear predictor is
fixed at $\eta_{i,0} = 0$ for every commuter, which is what makes the remaining
$m-1$ sets of coefficients identifiable. Without this constraint, adding the same
constant to all $m$ linear predictors would leave every probability unchanged.

Some texts absorb the intercept into the vector notation by adding a leading $1$
to the regressor vector. We keep $\beta_{0,j}$ visible here so that
$\mathbf{x}_i$ remains consistent with the regressor-vector notation introduced
earlier in the chapter.

Recall that the word **linear** in *linear predictor* refers to the way the regression coefficients enter $\eta_{i,j}$. It does **not** mean that the probability of choosing transport mode $j$ is a linear function of the regressors.

::: {.Heads-up}

:::: {.Heads-up-header}
Heads-up on what is linear in Multinomial Logistic regression!
::::

:::: {.Heads-up-container}

The systematic component for each non-baseline category $j$ is $\eta_{i,j}$. Each
of the $m-1$ equations

$$
\eta_{i,j}
=
\beta_{0,j}
+
\mathbf{x}_i^\top\boldsymbol{\beta}_j,
\qquad j = 1, \dots, m-1,
$$

is linear in the regression coefficients, while the conditional probabilities

$$
p_{i,j}
=
\Pr(Y_i = j \mid \mathbf{x}_i)
$$

live on the probability scale.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-emotion-bored-6230197/).](img/question-2.png){width="300"}

Therefore, we still need a link function that connects the $m-1$ unrestricted
real-valued linear predictors $\eta_{i,1}, \dots, \eta_{i,m-1}$ to $m$ valid
probabilities. Each $p_{i,j}$ must lie between $0$ and $1$, and together they must
sum to one across all $m$ categories, including the probability of belonging to the reference category `Car`.

::::

:::

This separation between the linear predictors and the response-scale
probabilities is what allows a GLM to retain a regression structure without
forcing the conditional probabilities themselves to be linear.

### The Logit Link Functions {#sec-multinomial-logistic-logit-link}

Recall that a link function connects the conditional distribution of the response variable to the
systematic component. In [Binary Logistic regression](https://alexrod61.github.io/regression-cookbook/book/08-binary-logistic.html), a single probability $\pi_i$
described the whole conditional distribution, so one link function was sufficient. In this chapter, the
random component is parameterized by $m$ conditional probabilities

$$
p_{i,j}=\Pr(Y_i=j\mid\mathbf{x}_i),
\qquad j=0,1,\dots,m-1,
$$

which must each lie within $(0,1)$ and sum to one. Multinomial Logistic regression
therefore needs a system of $m-1$ **logit links**, one for each non-reference category.

Here is what we do:

1. Pick one of the categories to be the baseline. In our study, this is
   `Car`, which is indexed as $j=0$.
2. For each of the other categories, model the logarithm of the odds of that
   category against the baseline.

The logit link applied to each baseline-category
comparison is given by

$$
\operatorname{logit}(p_{i,j})
=
\log\left(
\frac{p_{i,j}}{p_{i,0}}
\right)
=
\eta_{i,j},
\qquad j=1,\dots,m-1.
$${#eq-multinomial-logistic-logit-link}

Combining @eq-multinomial-logistic-logit-link with @eq-systematic-component gives
the Multinomial Logistic regression model, which is a system of $m-1$ equations:

$$
\log\left(
\frac{p_{i,j}}{p_{i,0}}
\right)
=
\beta_{0,j}
+
\beta_{1,j}x_{i,1}
+
\cdots
+
\beta_{k,j}x_{i,k},
\qquad j=1,\dots,m-1.
$$

The fraction

$$
\frac{p_{i,j}}{p_{i,0}}
$$

is the **odds of choosing transport mode $j$ instead of the baseline mode** for commuter
$i$. Each $\eta_{i,j}$ is therefore that commuter's **log-odds** of choosing transport mode $j$
versus `Car` under the model. Note that the regression coefficients are
different for each link function, and that the baseline itself has no equation. The linear predictor of `Car` is fixed at $\eta_{i,0}=0$, which corresponds to
$\log(p_{i,0}/p_{i,0})=0$.

The logit link solves the range problem of OLS regression:

- When $0<p_{i,j}<1$ and $0<p_{i,0}<1$, the odds $p_{i,j}/p_{i,0}$ are positive;
- Taking the logarithm maps those positive odds to any real number;
- Therefore, each $\eta_{i,j}$ can range freely from $-\infty$ to $+\infty$ even
  though every $p_{i,j}$ must remain between $0$ and $1$ and the $m$
  probabilities must sum to one.

For example,

$$
p_{i,j}=p_{i,0}
\quad\Longrightarrow\quad
\frac{p_{i,j}}{p_{i,0}}=1
\quad\Longrightarrow\quad
\operatorname{logit}(p_{i,j})=0.
$$

A commuter equally likely to choose mode $j$ or the baseline has log-odds of
zero. Modes less likely than the baseline correspond to negative log-odds, while
modes more likely than the baseline correspond to positive log-odds.

::: {.callout-tip}
## The Baseline Category is Arbitrary

Although every equation is written against the same reference category, the model can be used to compare between any two non-reference categories. Subtracting one equation from another cancels the common denominator $p_{i,0}$, e.g. for categories $1$ and $2$,

$$
\begin{aligned}
\eta_{i,1}-\eta_{i,2}
&=
\log\left(\frac{p_{i,1}}{p_{i,0}}\right)
-
\log\left(\frac{p_{i,2}}{p_{i,0}}\right)
=
\log\left(\frac{p_{i,1}}{p_{i,2}}\right) \\[0.4em]
&=
(\beta_{0,1}-\beta_{0,2})
+
(\beta_{1,1}-\beta_{1,2})x_{i,1}
+
\cdots
+
(\beta_{k,1}-\beta_{k,2})x_{i,k}.
\end{aligned}
$$

The choice of baseline is therefore arbitrary and the model does not inherently incorporate any rank of the response categories.
:::

With some algebraic manipulation, the system can be solved for the
probabilities themselves:

$$
\begin{aligned}
p_{i,0}
&=
\Pr(Y_i=0\mid\mathbf{x}_i)
=
\frac{1}{1+\sum_{l=1}^{m-1}\exp(\eta_{i,l})}, \\[0.6em]
p_{i,j}
&=
\Pr(Y_i=j\mid\mathbf{x}_i)
=
\frac{\exp(\eta_{i,j})}{1+\sum_{l=1}^{m-1}\exp(\eta_{i,l})},
\qquad j=1,\dots,m-1.
\end{aligned}
$${#eq-prob-multinomial}

The $1$ in each denominator corresponds to $\exp(\eta_{i,0})=\exp(0)$, which is the baseline's
contribution. Adding all $m$ expressions in @eq-prob-multinomial gives a sum of
$1$, so the model always returns a valid probability
distribution over the four transport modes. This matters when we use the model
for prediction and classification later in the chapter.

::: {#Definition-multinomial-logistic-logit .definition}
:::: {.definition-header}
Definition of the multinomial logit link
::::
:::: {.definition-container}

For category probabilities $p_{i,0},\dots,p_{i,m-1}$ satisfying $0<p_{i,j}<1$ and
$\sum_{j=0}^{m-1}p_{i,j}=1$, with category $0$ taken as the baseline, the
**logit function** for category $j$ is

$$
\operatorname{logit}(p_{i,j})
=
\log\left(
\frac{p_{i,j}}{p_{i,0}}
\right),
\qquad j=1,\dots,m-1.
$$

The quantity $p_{i,j}/p_{i,0}$ is the **odds** of category $j$ against the
baseline, and the logit is the **log-odds**. In Multinomial Logistic regression,
the logit link sets those log-odds equal to the corresponding linear predictor
with $k$ regressors:

$$
\begin{aligned}
\operatorname{logit}(p_{i,j})
&= \eta_{i,j} \\
&=
\beta_{0,j}
+
\beta_{1,j}x_{i,1}
+
\cdots
+
\beta_{k,j}x_{i,k},
\qquad j=1,\dots,m-1.
\end{aligned}
$$

::::
:::

Let us now set up the model for our case study. The response `transport_mode`
has $m=4$ classes: `Car`, `Public_Transit`, `Bicycle` and `Walking`, with `Car`
as the baseline level. The $k=4$ regressors are the continuous
`commute_distance` and `age`, denoted $x_{i,1}$ and $x_{i,2}$, and the binary
`has_car_available` and `weekend_commuter`, whose `Yes` dummy variables are
denoted $x_{i,3}$ and $x_{i,4}$. Writing $j=1,2,3$ for `Public_Transit`,
`Bicycle` and `Walking` respectively, the three link functions are

$$
\begin{aligned}
\eta_{i,1}
&=
\log\left(\frac{p_{i,1}}{p_{i,0}}\right)
=
\beta_{0,1}
+
\beta_{1,1}x_{i,1}
+
\beta_{2,1}x_{i,2}
+
\beta_{3,1}x_{i,3}
+
\beta_{4,1}x_{i,4}, \\[0.6em]
\eta_{i,2}
&=
\log\left(\frac{p_{i,2}}{p_{i,0}}\right)
=
\beta_{0,2}
+
\beta_{1,2}x_{i,1}
+
\beta_{2,2}x_{i,2}
+
\beta_{3,2}x_{i,3}
+
\beta_{4,2}x_{i,4}, \\[0.6em]
\eta_{i,3}
&=
\log\left(\frac{p_{i,3}}{p_{i,0}}\right)
=
\beta_{0,3}
+
\beta_{1,3}x_{i,1}
+
\beta_{2,3}x_{i,2}
+
\beta_{3,3}x_{i,3}
+
\beta_{4,3}x_{i,4}.
\end{aligned}
$$ {#eq-link}

Each link function has its own intercept and its own set of regression
coefficients, so the model estimates $(m-1)(k+1)=15$ coefficients in total.

Exponentiating both sides in @eq-link gives the odds of each transport mode against the baseline `Car` as a product of multiplicative factors, one per regressor:

$$
\frac{p_{i,j}}{p_{i,0}}
=
e^{\eta_{i,j}}
=
e^{\beta_{0,j}}\,
e^{\beta_{1,j}x_{i,1}}\,
e^{\beta_{2,j}x_{i,2}}\,
e^{\beta_{3,j}x_{i,3}}\,
e^{\beta_{4,j}x_{i,4}},
\qquad j=1,2,3.
$$

Finally, applying @eq-prob-multinomial with $m=4$ gives the four probabilities.
Writing $D=1+e^{\eta_{i,1}}+e^{\eta_{i,2}}+e^{\eta_{i,3}}$ for the common
denominator,

$$
\begin{aligned}
\Pr(Y_i=\texttt{Car}\mid\mathbf{x}_i) &= \frac{1}{D}, \\[0.4em]
\Pr(Y_i=\texttt{Public\_Transit}\mid\mathbf{x}_i) &= \frac{e^{\eta_{i,1}}}{D}, \\[0.4em]
\Pr(Y_i=\texttt{Bicycle}\mid\mathbf{x}_i) &= \frac{e^{\eta_{i,2}}}{D}, \\[0.4em]
\Pr(Y_i=\texttt{Walking}\mid\mathbf{x}_i) &= \frac{e^{\eta_{i,3}}}{D}.
\end{aligned}
$${#eq-probabilities}

Note that the observed response is always exactly one of the four transport modes, but
the model-based probabilities generated by finite values of $\eta_{i,1}$,
$\eta_{i,2}$ and $\eta_{i,3}$ lie strictly between $0$ and $1$. This distinction
between an observed categorical outcome and its underlying conditional
probabilities will remain important throughout the chapter.

::: {.Tip}
:::: {.Tip-header}
Tip on the softmax functions!
::::
:::: {.Tip-container}
The expressions in @eq-prob-multinomial are **softmax** functions. In the context of **machine learning**, Multinomial
Logistic regression is therefore equivalent to a one-layer **neural network** with a
softmax activation function: the regressors $\mathbf{x}_i$ are the input, the
probabilities $p_{i,j}$ are the output, and minimizing the cross-entropy loss to learn the weights is the same optimisation problem as maximising the multinomial
likelihood to estimate $\boldsymbol{\beta}_1,\dots,\boldsymbol{\beta}_{m-1}$.
::::
:::

### The Log-Likelihood Function

A Multinomial Logistic regression model treats the categories in the response variable as nominal, which means that there is no inherent order between them; therefore, each category is compared directly against a single baseline category. Recall from @eq-prob-multinomial that this comparison gives us the category-specific probabilities in closed form,

$$
p_{i,j} = P(Y_i = j \mid \mathbf{x}_i), \qquad j = 1, \ldots, m,
$$

where each non-baseline category $j$ enters through its own logit $\eta_i^{(j,1)}$ and the whole set is normalized by the softmax denominator, so that $\sum_{j=1}^m p_{i,j} = 1$. These conditional probabilities are exactly what the **likelihood** is built from.

The three equations in @eq-link share the same regressors and differ only in
their regression coefficients; therefore, the systematic component can be expressed compactly in matrix notation.

Let

$$
\mathbf{x}_i
=
\begin{bmatrix}
1 \\ x_{i,1} \\ x_{i,2} \\ x_{i,3} \\ x_{i,4}
\end{bmatrix}
$$

be the regressor vector for the $i$th observation, **including the leading $1$
for the intercept**. Then, let

$$
\boldsymbol{\beta}_j
=
\begin{bmatrix}
\beta_{0,j} \\ \beta_{1,j} \\ \beta_{2,j} \\ \beta_{3,j} \\ \beta_{4,j}
\end{bmatrix},
\qquad j = 1, 2, 3,
$$

be the vector of regression coefficients associated with category $j$. Each
systematic component can then be written compactly as

$$
\eta_{i,j} = \mathbf{x}_i^\top \boldsymbol{\beta}_j,
\qquad j = 1, 2, 3,
$$

where

$$
\mathbf{x}_i^\top \boldsymbol{\beta}_j
=
\beta_{0,j}
+
\beta_{1,j}x_{i,1}
+
\beta_{2,j}x_{i,2}
+
\beta_{3,j}x_{i,3}
+
\beta_{4,j}x_{i,4}.
$$

For the **full training sample** with $n$ observations, we collect the
regressors into the design matrix

$$
\mathbf{X}
=
\begin{bmatrix}
1 & x_{1,1} & x_{1,2} & x_{1,3} & x_{1,4} \\
1 & x_{2,1} & x_{2,2} & x_{2,3} & x_{2,4} \\
\vdots & \vdots & \vdots & \vdots & \vdots \\
1 & x_{n,1} & x_{n,2} & x_{n,3} & x_{n,4}
\end{bmatrix},
$$

whose $i$th row is $\mathbf{x}_i^\top$. Unlike the single-response case, the
model has one coefficient vector per non-baseline category, so the coefficients
are collected into the matrix

$$
\mathbf{B}
=
\begin{bmatrix}
\boldsymbol{\beta}_1 & \boldsymbol{\beta}_2 & \boldsymbol{\beta}_3
\end{bmatrix}
=
\begin{bmatrix}
\beta_{0,1} & \beta_{0,2} & \beta_{0,3} \\
\beta_{1,1} & \beta_{1,2} & \beta_{1,3} \\
\beta_{2,1} & \beta_{2,2} & \beta_{2,3} \\
\beta_{3,1} & \beta_{3,2} & \beta_{3,3} \\
\beta_{4,1} & \beta_{4,2} & \beta_{4,3}
\end{bmatrix},
$$

whose $j$th column contains the coefficients of category $j$. The matrix of
**systematic scores** is then

$$
\boldsymbol{\eta} = \mathbf{X}\mathbf{B},
$$

an $n \times 3$ matrix whose $(i, j)$th entry is $\eta_{i,j}$, and whose $i$th
row collects the three log-odds of observation $i$ against the baseline
category `Car`.

To estimate the model, we need a way to measure how well a given set of coefficients $\mathbf{B}$ explains the observed data. The **likelihood function** does exactly this by expressing the probability of the data we actually observed as a function of these unknown coefficients, and finding the values that make this likelihood function as large as possible.

Following the notation in @eq-pmf, let $y_{i,j} = 1$ if the $i$-th individual belongs to category $j$, and let $y_{i,j} = 0$ otherwise. The likelihood function for the Multinomial Logistic regression model is then

$$
\mathcal{L}(\mathbf{B}; \mathbf{y}) = \prod_{i=1}^n \prod_{j=1}^m p_{i,j}^{y_{i,j}},
$$ {#eq-likelihood-multinomial}

where $n$ is the number of individuals in the sample, $m$ is the number of categories, and $p_{i,j}$ is the model-implied softmax probability that individual $i$ (with regressor values $x_{i,1}, \ldots, x_{i,k}$) falls in category $j$. The unknown parameters are collected in the matrix $\mathbf{B}$, which consists of an intercept and $k$ slopes for each of the $m-1$ non-baseline response categories.

The two products have intuitive roles. The **inner product** over $j$ runs across the $m$ categories for a single individual; because of the indicator $y_{i,j}$ in the exponent, $\prod_{j=1}^m p_{i,j}^{y_{i,j}}$ is equal to just $p_{i,j}$ for the actual observed category of the $i$-th observation. The **outer product** over $i$ then multiplies these per-individual probabilities together across all $n$ people, which is valid because the observations are assumed independent. Overall, this is a **multinomial likelihood** evaluated at the softmax category probabilities.

Taking the logarithm turns the products into sums, which is far easier to differentiate and optimize, and yields the **log-likelihood function**:

$$
\ell(\mathbf{B}; \mathbf{y}) = \sum_{i=1}^n \sum_{j=1}^m y_{i,j} \log p_{i,j}.
$$ {#eq-log-likelihood-multinomial}

Substituting the softmax probabilities from @eq-prob-multinomial and using the reference-category constraint $y_{i,1} = 1 - \sum_{j=2}^m y_{i,j}$, the log-likelihood can be written explicitly in terms of the coefficients as

$$
\ell(\mathbf{B}) = \sum_{i=1}^n \left[ \sum_{j=2}^m y_{i,j} \sum_{k=0}^K x_{i,k} \beta_{k,j} - \log\left( 1 + \sum_{j=2}^m e^{\sum_{k=0}^K x_{i,k} \beta_{k,j}} \right) \right],
$$ {#eq-log-likelihoood}

where $x_{i,0} = 1$ corresponds to the intercept in the linear model.

Maximizing $\ell$ gives the estimate $\hat{\mathbf{B}}$, the parameters under which the observed data is the most plausible. This method of finding the best-fit parameters is known as is known as the **maximum-likelihood estimation**. There is, however, no closed-form algebraic expression for the maximizer: the category probabilities $p_{i,j}$ enter the log-likelihood non-linearly through the softmax function, so setting the derivatives to zero does not yield equations we can solve directly. Instead, we rely on **iterative numerical optimization**, which starts from an initial guess and refines the estimates step by step until the log-likelihood can no longer be improved. The standard choice is the **Newton–Raphson** method, described in the tip box below; the variant most statistical software uses replaces the observed Hessian with its expected value (the Fisher information), giving the **Fisher scoring** algorithm, which is equivalent to **iteratively reweighted least squares (IRLS)**. Additional estimation details, including the generalization from one to $n_i$ observations per population $i$, can be found in Section 2.2 of [Czepiel (2002)](https://czep.net/stat/mlelr.pdf).

::: {.Tip}
::::{.Tip-header}
Tip on the Newton–Raphson Method
::::
::::{.Tip-container}
The parameters we ultimately estimate are the entries $\beta_{k,j}$ of $\mathbf{B}$. For the purposes of the iteration, these entries are collected into the stacked vector $\boldsymbol{\beta} \in \mathbb{R}^q$ with $q = (K+1)(m-1)$ (an intercept and $K$ slopes for each of the $m-1$ non-baseline categories). Since the log-likelihood has no closed-form maximizer, the Newton–Raphson procedure estimates $\boldsymbol{\beta}$ by iterating with the score function and Hessian matrix of the log-likelihood, in a compact matrix form known as multinomial iteratively reweighted least squares (IRLS).

**Score function:** For category $j$ and regressor $k$, the corresponding component of the score vector $S(\boldsymbol{\beta}) \in \mathbb{R}^q$ (the first-order partial derivative of $\ell$) is

$$
\frac{\partial \ell(\boldsymbol{\beta})}{\partial \beta_{k,j}} = \sum_{i=1}^n \left( y_{i,j} - p_{i,j} \right) x_{i,k}.
$$

Each component compares the observed indicator $y_{i,j}$ with the model-implied probability $p_{i,j}$, weighted by the regressor value $x_{i,k}$; at the maximum, these differences balance out so that the score is zero.

**Hessian matrix:** The Hessian $H(\boldsymbol{\beta}) \in \mathbb{R}^{q \times q}$ collects the second-order partial derivatives of $\ell$. For the row indexed by category $j$ and regressor $k$, and the column indexed by category $j'$ and regressor $k'$, its entry is

$$
\frac{\partial^2 \ell(\boldsymbol{\beta})}{\partial \beta_{k,j} \partial \beta_{k',j'}} =
\begin{cases}
-\sum_{i=1}^n x_{i,k}\, p_{i,j}(1 - p_{i,j})\, x_{i,k'} & \text{if } j' = j, \\
\phantom{-}\sum_{i=1}^n x_{i,k}\, p_{i,j}\, p_{i,j'}\, x_{i,k'} & \text{if } j' \neq j,
\end{cases}
$$

for $j, j' = 2, \ldots, m$ and $k, k' = 0, \ldots, K$. As the non-baseline categories are coupled through the shared softmax denominator, the Hessian is a dense $q \times q$ matrix rather than block-diagonal. For a thorough treatment of the Hessian in the context of maximum likelihood estimation, refer to Chapter 4 of @agresti2013.

**Iteration:** Collecting the score and Hessian above, the Newton–Raphson update at step $t$ is

$$
\boldsymbol{\beta}^{(t+1)} \leftarrow \boldsymbol{\beta}^{(t)} - \left[ H\!\left( \boldsymbol{\beta}^{(t)} \right) \right]^{-1} S\!\left( \boldsymbol{\beta}^{(t)} \right),
$$

initialized at $\boldsymbol{\beta}^{(0)} = \mathbf{0}$ (or a data-driven starting point) and repeated until changes in $\ell$ and $\boldsymbol{\beta}$ fall below a convergence tolerance.

The full Newton–Raphson/IRLS procedure is:

1. Choose the baseline category and initialize $\boldsymbol{\beta}^{(0)} = \mathbf{0}$ (stacked over all $(k,j)$).
2. Repeat until convergence:
   - Compute the logits $\eta_i^{(j,1)}$ and category probabilities $p_{i,j}$ for all individuals $i$ and categories $j$, and evaluate the log-likelihood $\ell(\boldsymbol{\beta}^{(t)})$.
   - Form the stacked score $S(\boldsymbol{\beta}^{(t)})$ and Hessian $H(\boldsymbol{\beta}^{(t)})$ using the expressions above.
   - Update $\boldsymbol{\beta}^{(t+1)}$ using the Newton–Raphson step above.
   - Check the convergence criteria.
::::
:::

## Estimation {#sec-estimation}

The previous section established the Multinomial Logistic regression model mathematically. We will now fit the model to the training data using all four regressors (age, commute distance, car availability, and weekend commuter status), and obtain numerical estimates for their regression coefficients. These estimates quantify how much each regressor contributes to the systematic component of the model, allowing us to measure their individual effects on the conditional probability distribution of a commuter's primary mode of transport.

We will estimate the regression coefficients through **maximum likelihood estimation**, which gives the coefficient values for which the observed training data is the most compatible under a conditional Multinomial distribution for `transport_mode`, conditioned on the four regressors in the dataset. The fitted model serves two purposes. It lets us connect the mathematical framework from [Data Modelling](http://localhost:54321/book/14-multinomial-logistic.html#data-modelling) to an actual fitted model. Secondly it allows us to fit and inspect the model for stability and goodness of fit using the training data, while the testing data remains untouched. Following the [data science workflow](https://alexrod61.github.io/regression-cookbook/book/01-intro.html#sec-ds-workflow), the model fitted in this section is part of **model development** and is not yet the chapter's final inferential or final predictive evaluation.


### Fitting the Model

We now fit the Multinomial Logistic regression model on `train_set` using the `nnet` package in `R` and the `statsmodel` package in `Python`. 

::: {.panel-tabset}

## R Code

```{.r}
# Set the baseline model
train_set$transport_mode <- relevel(train_set$transport_mode, ref = "Car")

# Fit Multinomial Logistic regression
multinom_fit <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = train_set)

# Tidy the fitted model into a data frame of coefficient estimates
multinom_table <- tidy(multinom_fit)

# Print the fitted model
multinom_table |>
  kable(
    digits = 3,
    col.names = c("Category", "Term", "Estimate", "Standard error", "Wald statistic", "p-value"),
    align = c("l", "l", "c", "c", "c", "c")
  )
```

## Python Code

```{.python}
# Factorize the response and set the baseline model
train_set["transport_mode"] = pd.Categorical(
    train_set["transport_mode"],
    categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)

# Design matrix and response
y = train_set["transport_mode"]
X = train_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]

# Add intercept
X = sm.add_constant(X)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_numeric = X.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes

# Fit Multinomial Logistic regression
model = sm.MNLogit(y, X_numeric)
multinom_fit = model.fit()

# Recover the category labels assigned by the model
ynames = multinom_fit.model._ynames_map
non_baseline = [ynames[i] for i in range(1, len(ynames))]

# Collect the estimates for each non-baseline category
multinom_table = pd.concat(
    [
        pd.DataFrame({
            "Category": category,
            "Term": multinom_fit.params.index,
            "Estimate": multinom_fit.params.iloc[:, i].values,
            "Standard error": multinom_fit.bse.iloc[:, i].values,
            "Wald statistic": multinom_fit.tvalues.iloc[:, i].values,
            "p-value": multinom_fit.pvalues.iloc[:, i].values,
        })
        for i, category in enumerate(non_baseline)
    ],
    ignore_index=True
).round(3)

# Print the fitted model
multinom_fit_html = scrollable_table_html(multinom_table)
```

:::

Both languages have now estimated the Multinomial Logistic regression model from the same training data by maximum likelihood estimation through the Newton-Raphson method, which was discussed in [Section 14.5.5](https://alexrod61.github.io/regression-cookbook/book/14-multinomial-logistic.html#the-log-likelihood-function).

::: {.panel-tabset}

### R Output

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-multinom-fit-r
#| tbl-cap: "Maximum likelihood estimates from the Multinomial Logistic regression model fitted on the training data."

# Set the baseline model
train_set$transport_mode <- relevel(train_set$transport_mode, ref = "Car")

# Fit Multinomial Logistic regression
multinom_fit <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = train_set)

# Tidy the fitted model into a data frame of coefficient estimates
multinom_table <- tidy(multinom_fit)

# Print the fitted model
multinom_table |>
  kable(
    digits = 3,
    col.names = c("Category", "Term", "Estimate", "Standard error", "Wald statistic", "p-value"),
    align = c("l", "l", "c", "c", "c", "c")
  )
```

### Python Output

```{python}
#| echo: false
#| message: false
#| warning: false

# Factorize the response and set the baseline model
train_set["transport_mode"] = pd.Categorical(
    train_set["transport_mode"],
    categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)

# Design matrix and response
y = train_set["transport_mode"]
X = train_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]

# Add intercept
X = sm.add_constant(X)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_numeric = X.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes

# Fit Multinomial Logistic regression
model = sm.MNLogit(y, X_numeric)
multinom_fit = model.fit()

# Recover the category labels assigned by the model
ynames = multinom_fit.model._ynames_map
non_baseline = [ynames[i] for i in range(1, len(ynames))]

# Collect the estimates for each non-baseline category
multinom_table = pd.concat(
    [
        pd.DataFrame({
            "Category": category,
            "Term": multinom_fit.params.index,
            "Estimate": multinom_fit.params.iloc[:, i].values,
            "Standard error": multinom_fit.bse.iloc[:, i].values,
            "Wald statistic": multinom_fit.tvalues.iloc[:, i].values,
            "p-value": multinom_fit.pvalues.iloc[:, i].values,
        })
        for i, category in enumerate(non_baseline)
    ],
    ignore_index=True
).round(3)

# Print the fitted model
multinom_fit_html = scrollable_table_html(multinom_table)
```

```{r}
#| echo: false
#| label: tbl-multinom-fit-python
#| tbl-cap: "Maximum likelihood estimates from the Multinomial Logistic regression model fitted on the training data."
#| message: false
#| warning: false
#| results: asis

cat(
  "\n",
  py$multinom_fit_html,
  "\n",
  sep = ""
)
```

:::

The outputs in @tbl-multinom-fit-r and @tbl-multinom-fit-python report the estimated regression coefficients. For each non-baseline category $j$, a category-specific intercept $\hat{\beta}_{0}^{(j,1)}$ and the slope coefficients $\hat{\beta}_{1}^{(j,1)}, \ldots, \hat{\beta}_{k}^{(j,1)}$ are reported, each with its standard error. Since `Car` was set as the baseline category, the outputs contains one such set of estimates for each of `Public Transit`, `Bicycle`, and `Walking`, and all estimates are interpreted relative to `Car`. With $m = 4$ response categories and $k = 4$ regressors, the fitted model comprises $(m - 1) \times (k + 1) = 15$ estimated parameters, which is consistent with the number of rows in @tbl-multinom-fit-r and @tbl-multinom-fit-python.

The Multinomial Logisitc regression model fitted above with regressors $x_{i,\texttt{distance}}$, $x_{i,\texttt{age}}$, $x_{i,\texttt{car}}$, and $x_{i,\texttt{weekend}}$, can be written as

$$
\begin{aligned}
\log \left( \frac{P(Y_i = j \mid \mathbf{x}_i)}{P(Y_i = \text{Car} \mid \mathbf{x}_i)} \right) &= \hat{\beta}_{0}^{(j,\text{Car})}\\
&\quad + \hat{\beta}_{1}^{(j,\text{Car})} x_{i,\texttt{distance}}\\
&\quad + \hat{\beta}_{2}^{(j,\text{Car})} x_{i,\texttt{age}}\\
&\quad + \hat{\beta}_{3}^{(j,\text{Car})} x_{i,\texttt{car}}\\
&\quad + \hat{\beta}_{4}^{(j,\text{Car})} x_{i,\texttt{weekend}},
\end{aligned}
$$

where $j$ can be any of the three non-baseline levels: `Public Transit`, `Bicycle`, or `Walking`.

In Multinomial Logistic regression, we model the probability of each outcome category given regressors. The key assumption is that we have enough data for each category to estimate the model parameters reliably. If one category is much more frequent than the others (i.e., unbalanced outcomes), the model may:

1. Bias coefficients toward the majority class because the likelihood function is dominated by majority observations.
2. Underestimate uncertainty for rare categories (small sample size → large variance).
3. Yield poor predictive performance for minority classes (the model tends to predict the majority class).
4. In extreme cases, parameters for rare categories may not converge or may give infinite estimates because of complete or quasi-complete separation.

As a reminder, we just completed the estimation step of the [data science workflow](https://alexrod61.github.io/regression-cookbook/book/01-intro.html#sec-ds-workflow). For the **inferential inquiry**, we should not use the same dataset both to explore the model structure and to produce the final inferential claims. Recall that the train set is used for model development and goodness-of-fit assessment. Final inferential outputs will be produced later in @sec-regressors after the model building workflow has been completed.

## Checking Goodness of Fit and Model Stability

Now let's assess how well the fitted Multinomial Logistic regression explains and predicts the observed transport modes on the train set. We consider two complementary perspectives: **likelihood-based** measures that evaluate how much of the outcome's variation the model explains on the log-likelihood scale (similar to @sec-poisson-goodness-of-fit in [Classical Poisson Regression](https://alexrod61.github.io/regression-cookbook/book/10-classical-poisson.html)), and **classification-based** measures that evaluate how accurately the model assigns observations to their correct transport mode.

### Convergence and Separation
A particularly important problem for convergence of the maximum-likelihood estimation (MLE) solution is **separation**. 

::::: {.Heads-up}

:::: {.Heads-up-header}
Heads-up on complete and quasi-complete separation!
::::

:::: {.Heads-up-container}

Consider a Multinomial Logistic regression model with an intercept-augmented regressor vector $\widetilde{\mathbf{x}}_i$. **Complete separation** occurs when there exists some nonzero vector $\mathbf{a}$ and a response category $j$ such that

$$
\widetilde{\mathbf{x}}_i^\top\mathbf{a}>0
\quad
\text{for every observation with }Y_i=j,
$$

and

$$
\widetilde{\mathbf{x}}_i^\top\mathbf{a}<0
\quad
\text{for every observation with }Y_i\neq j.
$$

In other words, a linear boundary in the regressor space perfectly separates a response category from the other response categories.

**Quasi-complete separation** occurs when the same inequalities can be made non-strict,

$$
\widetilde{\mathbf{x}}_i^\top\mathbf{a}\geq0
\quad\text{for }Y_i=j,
$$

and

$$
\widetilde{\mathbf{x}}_i^\top\mathbf{a}\leq0
\quad\text{for }Y_i\neq j,
$$

with at least some observations lying exactly on the separating boundary.

::::

:::::

Under complete separation, the multinomial likelihood function can keep increasing as one or more coefficient magnitudes grow without bound. A finite maximum likelihood therefore does not exist. Under quasi-complete separation, the same instability occurs along at least one coefficient direction. As a result:

- coefficient estimates may become extremely large;
- standard errors based on the inverse Fisher information may become extremely large or unstable;
- fitted probabilities can be pushed increasingly close to $0$ or $1$; and
- ordinary Wald-type inference can become unreliable.

Software often alerts us to the problem. In `R`, common warnings include messages such as

```text
glm.fit: algorithm did not converge
```

or

```text
glm.fit: fitted probabilities numerically 0 or 1 occurred
```

depending on the severity of the problem. In `Python`'s [{statsmodels}](https://www.statsmodels.org/stable/index.html), a separated or nearly separated fit may generate a `PerfectSeparationWarning`, fail to converge, or produce extreme coefficient estimates and standard errors.

These warnings are useful, but diagnostics should not rely on warning text alone. Convergence, finite coefficients, finite standard errors, fitted probabilities, regressor overlap, and information-matrix stability should be considered together.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| message: false
#| warning: false

fitted_probs <- fitted(multinom_fit)

cat(sprintf(
  paste0(
    "Converged: %s\n",
    "All coefficients finite: %s\n",
    "All standard errors finite: %s\n",
    "All fitted probabilities finite: %s\n",
    "Fitted probabilities strictly inside (0, 1): %s\n",
    "Fitted-probability range: %.4f to %.4f\n"
  ),
  ifelse(isTRUE(multinom_fit$convergence == 0), "Yes", "No"),
  ifelse(all(is.finite(coef(multinom_fit))), "Yes", "No"),
  ifelse(all(is.finite(sqrt(diag(vcov(multinom_fit))))), "Yes", "No"),
  ifelse(all(is.finite(fitted_probs)), "Yes", "No"),
  ifelse(all(fitted_probs > 0 & fitted_probs < 1), "Yes", "No"),
  min(fitted_probs),
  max(fitted_probs)
))
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

converged = bool(multinom_fit.mle_retvals["converged"])
coefs = np.asarray(multinom_fit.params, dtype=float)
ses = np.asarray(multinom_fit.bse, dtype=float)
fitted = np.asarray(multinom_fit.predict(), dtype=float)

yes_no = lambda flag: "Yes" if flag else "No"

print(
    f"Converged: {yes_no(converged)}\n"
    f"All coefficients finite: {yes_no(np.all(np.isfinite(coefs)))}\n"
    f"All standard errors finite: {yes_no(np.all(np.isfinite(ses)))}\n"
    f"All fitted probabilities finite: {yes_no(np.all(np.isfinite(fitted)))}\n"
    f"Fitted probabilities strictly inside (0, 1): "
    f"{yes_no(np.all((fitted > 0) & (fitted < 1)))}\n"
    f"Fitted-probability range: {fitted.min():.4f} to {fitted.max():.4f}"
)
```

:::

The fitted model converges, returns finite coefficients and standard errors, and produces fitted probabilities strictly inside the interval between $0$ and $1$. According to the plots in @sec-eda, there is overlap between the age range and the commute distance range among all four categories of transport mode: `Car`, `Public_Transit`, `Bicycle`, and `Walking`. Similarly, there exists at least one observation for each transport mode among commuters with and without a car available, and among weekend and non-weekend commuters, so no combination of a binary regressor level and a response category is empty in the train set. These checks provide **no obvious evidence of complete or quasi-complete separation in this fitted model.**

### Likelihood-Based Measures
We will dig into some **likelihood-based** measures to assess goodness-of-fit of our model. 

The **null deviance** measures how well the intercept-only model (no regressors, only a baseline log-odds for each non-reference category) fits the data.

::: {.callout-note}
## Null deviance

The **null deviance** $D_{\text{null}}$ is the deviance of the intercept-only model, which contains one intercept per each non-baseline category and no regressors:

$$
D_{\text{null}} = -2\,\ell_{\text{null}},
$$

where $\ell_{\text{null}}$ is the maximized log-likelihood of that model. Because the null model assigns the same predicted category probabilities to every observation, $D_{\text{null}}$ serves as a baseline against which the residual deviance of the fitted model is compared.
:::

The **residual deviance** is the analogous quantity for the fitted model, and it measures how well our model, with four regressors, fits the data:

$$
D_{\text{resid}} = -2\,\ell(\hat{\boldsymbol{\beta}}).
$$

A well-fitting model will have a residual deviance substantially smaller than the null deviance. The difference between the two, $D_{\text{null}} - D_{\text{resid}}$, is the **likelihood-ratio (LR) statistic**. Under the null hypothesis $H_0$ that all slope coefficients are zero across every non-baseline category, this statistic follows a $\chi^2$ distribution with degrees of freedom equal to the number of slope parameters $k$:

$$
G^2 = D_{\text{null}} - D_{\text{resid}} = -2\bigl(\ell_{\text{null}} - \ell_{\text{fitted}}\bigr) \;\overset{H_0}{\sim}\; \chi^2(k).
$$

**McFadden's pseudo-$R^2$** expresses the proportional reduction in log-likelihood relative to the null model:

$$
R^2_{\text{McFadden}} = 1 - \frac{\ell(\hat{\boldsymbol{\beta}})}{\ell(\hat{\boldsymbol{\beta}}_{0,\text{null}})}.
$$

Unlike the usual R-squared in OLS regression, McFadden's pseudo-R-squared should not be interpreted as the proportion of variance explained; instead, it measures improvement in log-likelihood relative to the null model. Its values are typically much smaller than regular R-squared values, and values around $0.2$–$0.4$ are often considered indicative of a good/decent model fit in discrete-response models.

**AIC** and **BIC** scores penalize the log-likelihood value for model complexity, i.e. the number of regressors in the model:

$$
\text{AIC} = -2\,\ell(\hat{\boldsymbol{\beta}}) + 2p, \qquad \text{BIC} = -2\,\ell(\hat{\boldsymbol{\beta}}) + p\log n,
$$

where $p$ is the total number of estimated parameters (the category-specific intercepts and slopes across all non-baseline categories) and $n$ is the sample size. Lower AIC and BIC scores indicate better fit, and they are most useful for comparing competing models.

Let's compute all of these measures below in both `R` and `Python`.

::: {.panel-tabset}

## R Code

```{.r}
library(nnet)

# Null model: intercept only, no regressors
multinom_null <- multinom(transport_mode ~ 1, data = train_set)

ll_fitted <- logLik(multinom_fit)
ll_null   <- logLik(multinom_null)

null_deviance  <- -2 * as.numeric(ll_null)
resid_deviance <- -2 * as.numeric(ll_fitted)
mcfadden_r2    <- 1 - as.numeric(ll_fitted) / as.numeric(ll_null)
lr_stat        <- as.numeric(-2 * (ll_null - ll_fitted))
lr_df          <- attr(ll_fitted, "df") - attr(ll_null, "df")
lr_pval        <- pchisq(lr_stat, df = lr_df, lower.tail = FALSE)

cat("Null deviance:      ", round(null_deviance,  2), "\n")
cat("Residual deviance:  ", round(resid_deviance, 2), "\n")
cat("McFadden R²:        ", round(mcfadden_r2, 4),   "\n")
cat("G²:                 ", round(lr_stat, 4),
    " (df =", lr_df, ", p =", formatC(lr_pval, format = "e", digits = 2), ")\n")
cat("AIC:                ", round(AIC(multinom_fit), 2), "\n")
cat("BIC:                ", round(BIC(multinom_fit), 2), "\n")
```

## Python Code

```{.python}
import numpy as np
from scipy.stats import chi2 as chi2dist

# The log-likelihood of the intercept-only (null) model is available
# directly on the fitted MNLogit result
ll_fitted = multinom_fit.llf
ll_null   = multinom_fit.llnull

null_deviance  = -2 * ll_null
resid_deviance = -2 * ll_fitted
mcfadden_r2    = 1 - ll_fitted / ll_null
lr_stat        = -2 * (ll_null - ll_fitted)
lr_df          = multinom_fit.df_model
lr_pval        = chi2dist.sf(lr_stat, lr_df)

print(f"Null deviance:      {null_deviance:.2f}")
print(f"Residual deviance:  {resid_deviance:.2f}")
print(f"McFadden R²:        {mcfadden_r2:.4f}")
print(f"G²:                 {lr_stat:.4f}  (df = {lr_df}, p = {lr_pval:.2e})")
print(f"AIC:                {multinom_fit.aic:.2f}")
print(f"BIC:                {multinom_fit.bic:.2f}")
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
library(nnet)

multinom_null <- multinom(transport_mode ~ 1, data = train_set)

ll_fitted <- logLik(multinom_fit)
ll_null   <- logLik(multinom_null)

null_deviance  <- -2 * as.numeric(ll_null)
resid_deviance <- -2 * as.numeric(ll_fitted)
mcfadden_r2    <- 1 - as.numeric(ll_fitted) / as.numeric(ll_null)
lr_stat        <- as.numeric(-2 * (ll_null - ll_fitted))
lr_df          <- attr(ll_fitted, "df") - attr(ll_null, "df")
lr_pval        <- pchisq(lr_stat, df = lr_df, lower.tail = FALSE)

cat("Null deviance:      ", round(null_deviance,  2), "\n")
cat("Residual deviance:  ", round(resid_deviance, 2), "\n")
cat("McFadden R²:        ", round(mcfadden_r2, 4),   "\n")
cat("G²:                 ", round(lr_stat, 4),
    " (df =", lr_df, ", p =", formatC(lr_pval, format = "e", digits = 2), ")\n")
cat("AIC:                ", round(AIC(multinom_fit), 2), "\n")
cat("BIC:                ", round(BIC(multinom_fit), 2), "\n")
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
import numpy as np
from scipy.stats import chi2 as chi2dist

ll_fitted = multinom_fit.llf
ll_null   = multinom_fit.llnull

null_deviance  = -2 * ll_null
resid_deviance = -2 * ll_fitted
mcfadden_r2    = 1 - ll_fitted / ll_null
lr_stat        = -2 * (ll_null - ll_fitted)
lr_df          = multinom_fit.df_model
lr_pval        = chi2dist.sf(lr_stat, lr_df)

print(f"Null deviance:      {null_deviance:.2f}")
print(f"Residual deviance:  {resid_deviance:.2f}")
print(f"McFadden R²:        {mcfadden_r2:.4f}")
print(f"G²:                 {lr_stat:.4f}  (df = {lr_df}, p = {lr_pval:.2e})")
print(f"AIC:                {multinom_fit.aic:.2f}")
print(f"BIC:                {multinom_fit.bic:.2f}")
```

:::

The residual deviance is smaller than the null deviance by `r round(lr_stat,4)` units. This reduction is captured by the likelihood-ratio statistic $G^2=$ `r round(lr_stat,4)` on `r lr_df` degrees of freedom ($p < 0.001$), which provides sufficient evidence that a model with age, commute distance, car availability, and weekend commuting status is correctly specified in comparison to the null (intercept-only) model. McFadden's $R^2$ is `r round(mcfadden_r2,4)`, which falls within the $0.2-0.4$ range commonly associated with a good fit for discrete-response models. This indicates that these four regressors together explain a substantial share of variability in `transport_mode` on the log-likelihood scale, consistent with their highly significant joint contribution. The AIC and BIC scores of `r round(AIC(multinom_fit), 2)` and `r round(BIC(multinom_fit), 2)` respectively describe the model fitted in @sec-estimation, and they are useful for comparing a fitted model against other model specifications (eg. a nested model with fewer or more regressors).

### Classification-Based Measures

Likelihood-based measures are functions of the maximized log-likelihood, which aggregates, on the log scale, the conditional probability the fitted model assigns to the category each observation in the training data was actually observed to take. These measures are informative because they depend on the full vector of fitted probabilities rather than only on which category attains the maximum, which means that a correct prediction made with high confidence and one made with low confidence are not treated as equivalent. However, in **machine learning**, a model is often used to **predict** a single response category instead of reporting the predicted probabilities. **Classification-based measures** address this use case directly by collapsing each predicted probability distribution to a single point prediction, taken as the category receiving the highest predicted probability, and then assess how often that prediction matches the observed transport mode.

This shift from probabilities to discrete labels changes what we are able to diagnose. An **overall accuracy** score rate tells us how frequently the model is wrong, but not where it goes wrong, and in a multi-category setting, these are separate questions. Two Multinomial Logistic regression models with identical accuracy scores may misclassify categories in different ways. Separating the errors by the category predicted and the category observed is what makes this distinction visible, and the table that performs that separation is the **confusion matrix**.

::: {.Tip}
:::: {.Tip-header}
Tip on Confusion Matrix
::::
:::: {.Tip-container}
For a response with $K$ categories, the **confusion matrix** is a $K \times K$ table that cross-tabulates the predicted category against the observed category. Writing $\hat{y}_i$ for the point prediction of observation $i$ and $y_i$ for its observed category, the entry in row $j$ and column $k$ is

$$
n_{j,k} = \sum_{i=1}^{n} \mathbb{1}\{\hat{y}_i = j,\; y_i = k\},
$$

which is the number of observations predicted as category $j$ whose true category is $k$. The diagonal entries $n_{k,k}$ count correct classifications, and every off-diagonal entry records a specific type of misclassification. This makes the table more informative than a single accuracy score because it not only shows how often the model is wrong, but also which categories it confuses with which.

Note that the row and column convention is not universal. The `caret` implementation in R places predictions in the rows and observed values in the columns, as above, whereas `scikit-learn` in Python uses the opposite arrangement. Before interpreting the off-diagonal entries in a confusion matrix, one should confirm its orientation.
::::
:::

Let's visualize the confusion matrix of our fitted model on train set by using `confusionMatrix` from `caret` in `R`, and `confusion_matrix` from `scikit-learn` in `Python`.

::: {.panel-tabset}

## R Code

```{.r}
y_pred <- predict(multinom_fit, newdata = train_set)
mode_levels <- c("Car", "Public_Transit", "Bicycle", "Walking")
conf_matrix <- confusionMatrix(factor(y_pred, levels = mode_levels),
                              factor(train_set$transport_mode, levels = mode_levels))

# Visualise
cm_df <- as.data.frame(as.table(conf_matrix$table))
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

# Order axes so the main diagonal (correct matches) runs top-left to
# bottom-right: Predicted increases left -> right (Car -> Walking), and
# Actual increases top -> bottom (Car -> Walking). ggplot draws the first
# factor level at the bottom of the y-axis, so we reverse Actual's levels to
# place "Car" at the top.
cm_df$Predicted <- factor(cm_df$Predicted, levels = levels(cm_df$Predicted))
cm_df$Actual    <- factor(cm_df$Actual,    levels = rev(levels(cm_df$Actual)))

cm_df |>
  ggplot(aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  labs(title = "Confusion Matrix (Train Set)",
       x = "Predicted label", y = "True label") +
  theme_bw(base_size = 13)
```

## Python Code

```{.python}
y_pred = multinom_fit.predict(X_numeric).idxmax(axis=1)

# Confusion matrix
# Display order for the confusion matrix axes
cm_order = ["Car", "Public_Transit", "Bicycle", "Walking"]
order_codes = [list(y.cat.categories).index(c) for c in cm_order]

cm = confusion_matrix(y.cat.codes, y_pred, labels=order_codes)
disp = ConfusionMatrixDisplay(confusion_matrix=cm,
                              display_labels=cm_order)
disp.plot(cmap="Blues")
disp.ax_.set_title("Confusion Matrix (Train Set)")
fig = disp.figure_
fig.tight_layout()
fig
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| label: fig-confusion-matrix-train-R
#| fig-cap: "Confusion matrix of predicted versus actual transport modes in the train set."
#| echo: false
#| message: false
#| warning: false
y_pred <- predict(multinom_fit, newdata = train_set)
mode_levels <- c("Car", "Public_Transit", "Bicycle", "Walking")
conf_matrix <- confusionMatrix(factor(y_pred, levels = mode_levels),
                              factor(train_set$transport_mode, levels = mode_levels))

cm_df <- as.data.frame(as.table(conf_matrix$table))
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

# Order axes so the main diagonal (correct matches) runs top-left to
# bottom-right: Predicted increases left -> right (Car -> Walking), and
# Actual increases top -> bottom (Car -> Walking). ggplot draws the first
# factor level at the bottom of the y-axis, so we reverse Actual's levels to
# place "Car" at the top.
cm_df$Predicted <- factor(cm_df$Predicted, levels = levels(cm_df$Predicted))
cm_df$Actual    <- factor(cm_df$Actual,    levels = rev(levels(cm_df$Actual)))

cm_df |>
  ggplot(aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  labs(title = "Confusion Matrix (Train Set)",
       x = "Predicted label", y = "True label") +
  theme_bw(base_size = 13)
```

## Python Output

```{python}
#| label: fig-confusion-matrix-train-py
#| fig-cap: "Confusion matrix of predicted versus actual transport modes in the train set."
#| echo: false
#| message: false
#| warning: false
y_pred = multinom_fit.predict(X_numeric).idxmax(axis=1)

# Compute confusion matrix
# Display order for the confusion matrix axes
cm_order = ["Car", "Public_Transit", "Bicycle", "Walking"]
order_codes = [list(y.cat.categories).index(c) for c in cm_order]

cm = confusion_matrix(y.cat.codes, y_pred, labels=order_codes)
disp = ConfusionMatrixDisplay(confusion_matrix=cm,
                              display_labels=cm_order)
disp.plot(cmap="Blues")
disp.ax_.set_title("Confusion Matrix (Train Set)")
fig = disp.figure_
fig.tight_layout()
fig
```

:::

Now, let's compute the performance metrics below.

::: {.panel-tabset}

## R Code

```{.r}
# Classification predictions & confusion matrix
y_pred <- predict(multinom_fit, newdata = train_set)
conf_matrix <- confusionMatrix(factor(y_pred, levels = levels(train_set$transport_mode)),
                               factor(train_set$transport_mode, levels = levels(train_set$transport_mode)))
overall_accuracy <- conf_matrix$overall['Accuracy']
macro_recall <- mean(conf_matrix$byClass[,"Recall"], na.rm = TRUE)
macro_f1 <- mean(conf_matrix$byClass[,"F1"], na.rm = TRUE)
cat("Overall Accuracy:", round(overall_accuracy, 3), "\n")
cat("Macro Recall:", round(macro_recall, 3), "\n")
cat("Macro F1:", round(macro_f1, 3), "\n")
```

## Python Code

```{.python}
# Compute classification predictions & confusion matrix
probs = multinom_fit.predict(X_numeric)  # DataFrame of class probabilities
y_pred = probs.idxmax(axis=1)

# Compute the metrics
acc = accuracy_score(y.cat.codes, y_pred)
recall = recall_score(y.cat.codes, y_pred, average='macro')
# Average F1 only over the classes the model actually predicts, so a
# never-predicted class (undefined precision) is dropped rather than
# counted as 0 — matching caret's byClass F1 with na.rm = TRUE in the R tab.
predicted_labels = np.unique(y_pred)
f1 = f1_score(y.cat.codes, y_pred, labels=predicted_labels, average='macro')
print(f"Overall Accuracy: {acc:.3f}")
print(f"Macro Recall: {recall:.3f}")
print(f"Macro F1-score: {f1:.3f}")
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
y_pred <- predict(multinom_fit, newdata = train_set)
conf_matrix <- confusionMatrix(factor(y_pred, levels = levels(train_set$transport_mode)),
                               factor(train_set$transport_mode, levels = levels(train_set$transport_mode)))
overall_accuracy <- conf_matrix$overall['Accuracy']
macro_recall <- mean(conf_matrix$byClass[,"Recall"], na.rm = TRUE)
macro_f1 <- mean(conf_matrix$byClass[,"F1"], na.rm = TRUE)
cat("Overall Accuracy:", round(overall_accuracy, 3), "\n")
cat("Macro Recall:", round(macro_recall, 3), "\n")
cat("Macro F1:", round(macro_f1, 3), "\n")
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
probs = multinom_fit.predict(X_numeric)
y_pred = probs.idxmax(axis=1)
acc = accuracy_score(y.cat.codes, y_pred)
recall = recall_score(y.cat.codes, y_pred, average='macro')
# Average F1 only over the classes the model actually predicts, so a
# never-predicted class (undefined precision) is dropped rather than
# counted as 0 — matching caret's byClass F1 with na.rm = TRUE in the R tab.
predicted_labels = np.unique(y_pred)
f1 = f1_score(y.cat.codes, y_pred, labels=predicted_labels, average='macro')
print(f"Overall Accuracy: {acc:.3f}")
print(f"Macro Recall: {recall:.3f}")
print(f"Macro F1-score: {f1:.3f}")
```

:::

On the train set, the model achieves an overall accuracy of about **`r 100.0*round(overall_accuracy, 3)`%**. On average, the model correctly identifies **`r 100.0*round(macro_recall, 3)`%** of instances for each transport mode, which means that the macro recall is **`r 100.0*round(macro_recall, 3)`%**. Macro recall is relatively low, which suggests that the model is biased towards the majority classes, even though the overall accuracy is good. The **macro F1 score** is the unweighted mean of the per-class F1 scores, where the F1 score of class $k$ is the harmonic mean of that class's precision and recall,

$$
F_{1,k} = \frac{2 \cdot \text{precision}_k \cdot \text{recall}_k}{\text{precision}_k + \text{recall}_k},
\qquad
\text{macro } F_1 = \frac{1}{K} \sum_{k=1}^{K} F_{1,k},
$$

so that each transport mode contributes equally regardless of how many observations it contains, and a class scores well only when the model identifies most of its members and is not overly liberal in assigning that label. The macro F1 score is **`r 100.0*round(macro_f1, 3)`%**, indicating a moderate balance of precision and recall on average across the four transport modes. As each per-class F1 score lies between that class's precision and recall, the macro F1 score exceeds the macro recall precisely to the extent that precision exceeds recall within some individual classes, so comparing the two values indicates the model's difficulty with the less frequent transport modes, which are `Bicycle` and `Walking`. For example, all true `Bicycle` observations in the train set are misclassified, while classifications for all true `Public_Transit` and `Car` observations are split between `Public_Transit` and `Car` without ever assigning `Bicycle` or `Walking` labels to these observations.

In summary, the model performs well overall but it is biased towards the majority classes. It often fails to detect less frequent transport modes, which is why macro recall is relatively low. Overall accuracy alone would overestimate the model’s effectiveness in prediction across all four categories of transport mode. 

::: {.Tip}
::::{.Tip-header}
Upsampling as a Remedy for Class Imbalance
::::
::::{.Tip-container}
One common response to the behaviour described above is to rebalance the training set before fitting. **Upsampling** draws minority category observations with replacement until every category is represented equally, which can be done with `upSample()` from the `caret` package in `R` or with the resampling classes provided by the `imblearn` library in `Python`. The effect on the fitted model follows directly from the estimating equation for the intercepts. Since this equation forces the fitted probabilities of each category to average exactly its sample proportion in the data used for estimation, equalizing the category counts raises this average to $1/K$ for every category, assuming that there are $K$ categories in total. The minority categories therefore start from a far higher baseline fitted probability and are able to attain the maximum for a much wider range of regressor values, which typically increases their recall and allows categories that were never predicted at all to appear in the confusion matrix.

Note that upsampling adds no new information because the replicated observations are copies of those already present in the dataset. More importantly, the fitted probabilities produced by a model estimated on a balanced training set can no longer estimate the conditional probabilities in the original population, so gains in recall for the minority categories are generally accompanied by losses in precision and by predicted probabilities that are no longer calibrated. Whether this trade-off is worthwhile depends on the cost of the two types of error in the application at hand.
::::
:::

### Practical Model Adequacy {#sec-practical-adequacy}

Goodness of fit is not a competition to prove that a model is "*true*." As a reminder, every statistical model is a simplification of some data-generating process. At this stage, our practical question is:

> **Is the Multinomial Logistic regression model with four regressors sufficiently stable, interpretable, and compatible with the training data to serve as a useful model for our inferential and predictive inquiries?**

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-teaching-to-learn-wiki-3976301/).](img/board.png){width="500"}

The diagnostics above address different parts of that question. Some assess whether the model can be estimated reliably, some examine agreement on the probability scale, and some examine agreement of categories with highest fitted probabilities in the regressor space. No single diagnostic should determine the conclusion by itself.

@tbl-multinomial-logistic-practical-adequacy brings these pieces together.

| **Diagnostic question** | **What we found in the training data** | **Implication for the model** |
|:---:|:---:|:---:|
| **Convergence and separation** | The MLE converged; the coefficient estimates and model-based standard errors are finite; the fitted probabilities remain strictly inside $(0,1)$; and there is no complete or quasi-complete separation. | There is no obvious evidence of complete or quasi-complete separation in this model. The fitted coefficients, standard errors, and fitted probability scores are therefore numerically interpretable. |
| **Likelihood-based measures** | The likelihood-ratio statistic for the fitted model with four regressors is $G^2=$`r round(lr_stat, 4)` with $p<0.001$, and McFadden's $R^2$ is `r round(mcfadden_r2, 4)` | There is sufficient evidence that the model is correctly specified in comparison to the null model, and the four regressors together explain a substantial share of variance in `transport_mode` on the log-likelihood scale. |
| **Classification-based measures** | The overall accuracy is `r 100.0*round(overall_accuracy, 3)`\%. The maco recall is `r 100.0*round(macro_recall, 3)`\%. The macro F1 score is `r 100.0*round(macro_f1, 3)`\%. | The overall accuracy and recall scores of the model are good, but it often fails to recall less frequent categories, such as `Bicycle` and `Walking`. As the global classification measures look good, we will use this model to predict on the held-out test set in @sec-results, and assess the predictive performance through these measures again. |

Taken together, every diagnostic reported above is computed on the same observations used to obtain $\hat{\boldsymbol{\beta}}$, and in-sample measures are optimistic by construction, since the maximum likelihood estimates are chosen precisely to maximize compatibility with those observations. The goodness-of-fit checks therefore establish that the model is interpretable and compatible with the patterns observed in @sec-eda, but they do not guarantee that the model will perform comparably on data that it has not yet encountered. Predictive performance is a separate question, and it is necessary to use the untouched test data, which did not play any role in parameter estimation, to do a final assessment of the predictive performance.

Accordingly, we will proceed to @sec-results, where we will fit the same model specification with all four regressors on the test data and use this independent fit to estimate coefficients, standard errors, confidence intervals, and carry out hypothesis tests to answer our **inferential inquiry** regarding the factors associated with primary mode of transport in the population of commuters. For the **predictive inquiry**, we will retain the coefficients estimates from the training data and evaluate the predictions that they produce for the held-out testing data.

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on the Golden Rule of Machine Learning
::::
::::{.Heads-up-container}
The "golden rule" of predictive modelling is that the **test set** must play no role in estimation or model selection, so that the measures computed on it remain unbiased estimates of predictive performance on unseen data. The procedure described above uses the test set in two distinct roles: 

- the estimation sample for the final **inferential fit**,
- the held-out sample for the **predictive evaluation**.

The predictive measures themselves remain numerically untainted because the coefficients used to generate the test-set predictions are those estimated on the training data alone, and the separate inferential fit does not alter them. Once the test set estimates, standard errors, and hypothesis tests have been examined, any subsequent revision to the model specification would be informed by the test data, and the reported predictive measures would no longer be honest estimates of the out-of-sample classification performance. We therefore fix the specification before either fit is carried out, report both sets of results as they stand, and do not return to modify the model in light of what the test set reveals.
::::
:::

## Results and Statistical Interpretation {#sec-results}

### Wald Tests

After we finish estimating the regression coefficients, we look at how uncertain they are and whether each coefficient differs significantly from zero or not. We get a variance estimate for each coefficient from the observed information matrix by taking the negative of the Hessian at the solution. Taking square roots of these variances gives standard errors.

We can determine whether a regressor is statistically associated with the logarithm of the odds relative to the reference category through hypothesis testing for the coefficients $\beta_{k,j}$, where $k$ denotes the corresponding regressor and $j$ denotes a non-baseline category compared against the baseline `Car`. For a coefficient $\beta_{k,j}$, a common two-sided Wald test is

$$
\begin{gather*}
H_0: \beta_{k,j} = 0, \\
\text{versus} \\
H_1: \beta_{k,j} \neq 0.
\end{gather*}
$$

The corresponding Wald statistic is

$$
z_{k,j} = \frac{\hat{\beta}_{k,j}-0}{\mbox{se}\left(\hat{\beta}_{k,j}\right)},
$$ {#eq-wald-statistic}

which, provided that the sample size $n$ is large enough, has an approximately Standard Normal distribution under $H_0$.

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on Wald tests and the exponentiated scale!
::::
:::: {.Heads-up-container}
The Wald test is computed on the **logit scale**, where each coefficient is $\beta_{k,j}\in\mathbb{R}$. However, we often interpret the coefficient on the odds scale using $\exp(\beta_{k,j})$.

These are the same hypothesis written on two scales:

$$
H_0\text{: } \beta_{k,j} = 0
$$

is equivalent to

$$
H_0\text{: } \exp(\beta_{k,j}) = 1.
$$

So, for a continuous regressor, the Wald test asks whether the **multiplicative change** in odds with respect to the reference category for a one-unit increase is different from $1$. For a binary categorical regressors, it asks whether the odds ratio between the comparison category and the reference category is different from $1$.
::::
:::


The corresponding $p$-values for each $\beta_{k,j}$ can be computed. The smaller the $p$-value, the stronger is the evidence against the null hypothesis $H_0$. As in the previous regression models, we would set a predetermined significance level (usually taken to be $\alpha = 0.05$) to infer if the $p$-value is small enough. If the $p$-value is smaller than the predetermined level $\alpha$, then you could claim that there is evidence to reject the null hypothesis. Hence, a $p$-value that is small enough indicates that the data provides evidence in favour of association between category $j$ and the baseline category `Car`, and the $k$th regressor.

Given a specified level of confidence where $\alpha$ is the significance level, we can construct approximate $(1-\alpha) \times 100\%$ confidence intervals for the corresponding true value of $\beta_{k,j}$:

$$
\hat{\beta}_{k,j} \pm z_{\frac{1-\alpha}{2}} \, \mbox{se}\left( \hat{\beta}_{k,j} \right),
$$ {#eq-ci}

where $z_{\frac{1-\alpha}{2}}$ is the $\left(\frac{1-\alpha}{2}\right)$ quantile of the Standard Normal distribution. For example, for a $95\%$ confidence interval, we use $z_{0.975}$ in this equation for computing the endpoints of the confidence interval.

Exponentiating the endpoints gives an approximate 95% confidence interval for $\exp(\beta_{k,j})$:

$$
\begin{gather*}
\Big[
\exp
\left(
\hat{\beta}_{k,j}
-
z_{0.975}
\operatorname{SE}(\hat{\beta}_{k,j})
\right), \\
\qquad \qquad \exp
\left(
\hat{\beta}_{k,j}
+
z_{0.975}
\operatorname{SE}(\hat{\beta}_{k,j})
\right)
\Big].
\end{gather*}
$$

For each coefficient, the decision rule is:

$$
\text{Reject } H_0
\text{ if }
p\text{-value} < 0.05.
$$

Otherwise, we fail to reject $H_0$. In this context, rejecting $H_0\text{: } \beta_j=0$ means that the testing-set refit provides statistical evidence that the corresponding odds ratio differs from $1$, under the Multinomial Logistic regression model and its model-based standard errors.

### Inferential Results {#sec-inference}

Now we will refit the Multinomial Logistic regression model on the test set using the `nnet` package in `R` and the `statsmodel` package in `Python`. We will also compute the corresponding 95\% confidence intervals below and filter the statistically significant coefficients at the $5\%$ significance level.

::: {.panel-tabset}

## R Code

```{.r}
# Set "Car" as the baseline level
test_set$transport_mode <- relevel(test_set$transport_mode, ref = "Car")

# Fit Multinomial Logistic regression model on test set
multinom_test <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = test_set)

# Tidy the fitted model into a data frame of coefficient estimates
results <- tidy(multinom_test)

# Add the testing decision at the 5% significance level
results <- results |>
  mutate(decision = if_else(p.value < 0.05, "Reject H0", "Fail to reject H0"))

# Print the fitted model
results |>
  kable(
    digits = 3,
    col.names = c("Category", "Term", "Estimate", "Standard error", "Wald statistic", "p-value", "Decision"),
    align = c("l", "l", "c", "c", "c", "c", "l"),
    caption = "Maximum likelihood estimates from the Multinomial Logistic regression model fitted on the test data."
  )
```

## Python Code

```{.python}
# Factorize the response and set the baseline model
test_set["transport_mode"] = pd.Categorical(
    test_set["transport_mode"],
    categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)

# Design matrix and response
y = test_set["transport_mode"]
X = test_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]

# Add intercept
X = sm.add_constant(X)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_numeric = X.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes

# Fit Multinomial Logistic regression
model = sm.MNLogit(y, X_numeric)
multinom_fit = model.fit()

# Recover the category labels assigned by the model
ynames = multinom_fit.model._ynames_map
non_baseline = [ynames[i] for i in range(1, len(ynames))]

# Collect the estimates for each non-baseline category,
# along with the testing decision at the 5% significance level
multinom_table = pd.concat(
    [
        pd.DataFrame({
            "Category": category,
            "Term": multinom_fit.params.index,
            "Estimate": multinom_fit.params.iloc[:, i].values,
            "Standard error": multinom_fit.bse.iloc[:, i].values,
            "Wald statistic": multinom_fit.tvalues.iloc[:, i].values,
            "p-value": multinom_fit.pvalues.iloc[:, i].values,
            "Decision": np.where(
                multinom_fit.pvalues.iloc[:, i].values < 0.05,
                "Reject H0",
                "Fail to reject H0"
            ),
        })
        for i, category in enumerate(non_baseline)
    ],
    ignore_index=True
).round(3)

# Print the fitted model
multinom_fit_html = scrollable_table_html(multinom_table)
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
# Set "Car" as the baseline level
test_set$transport_mode <- relevel(test_set$transport_mode, ref = "Car")

# Fit Multinomial Logistic regression model on test set
multinom_test <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = test_set)


# Tidy the fitted model into a data frame of coefficient estimates
results <- tidy(multinom_test)

# Add the testing decision at the 5% significance level
results <- results |>
  mutate(decision = if_else(p.value < 0.05, "Reject H0", "Fail to reject H0"))

# Print the fitted model
results |>
  kable(
    digits = 3,
    col.names = c("Category", "Term", "Estimate", "Standard error", "Wald statistic", "p-value", "Decision"),
    align = c("l", "l", "c", "c", "c", "c", "l"),
    caption = "Maximum likelihood estimates from the Multinomial Logistic regression model fitted on the training data."
  )
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
# Factorize the response and set the baseline model
test_set["transport_mode"] = pd.Categorical(
    test_set["transport_mode"],
    categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)

# Design matrix and response
y = test_set["transport_mode"]
X = test_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]

# Add intercept
X = sm.add_constant(X)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_numeric = X.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes

# Fit Multinomial Logistic regression
model = sm.MNLogit(y, X_numeric)
multinom_fit = model.fit()

# Recover the category labels assigned by the model
ynames = multinom_fit.model._ynames_map
non_baseline = [ynames[i] for i in range(1, len(ynames))]

# Collect the estimates for each non-baseline category,
# along with the testing decision at the 5% significance level
multinom_table = pd.concat(
    [
        pd.DataFrame({
            "Category": category,
            "Term": multinom_fit.params.index,
            "Estimate": multinom_fit.params.iloc[:, i].values,
            "Standard error": multinom_fit.bse.iloc[:, i].values,
            "Wald statistic": multinom_fit.tvalues.iloc[:, i].values,
            "p-value": multinom_fit.pvalues.iloc[:, i].values,
            "Decision": np.where(
                multinom_fit.pvalues.iloc[:, i].values < 0.05,
                "Reject H0",
                "Fail to reject H0"
            ),
        })
        for i, category in enumerate(non_baseline)
    ],
    ignore_index=True
).round(3)

# Print the fitted model
multinom_fit_html = scrollable_table_html(multinom_table)
```

```{r}
#| echo: false
#| label: tbl-multinom-test-fit-python
#| tbl-cap: "Maximum likelihood estimates from the Multinomial Logistic regression model fitted on the test data."
#| message: false
#| warning: false
#| results: asis

cat(
  "\n",
  py$multinom_fit_html,
  "\n",
  sep = ""
)
```

:::

The **decision column** in @tbl-multinom-test-fit-python shows that, at $\alpha=0.05$ significance level, the effect of `commute_distance` on the odds of choosing `Public_Transit` versus `Car`, `Bicycle` versus `Car`, and `Walking` versus `Car` are statistically significant. This result is consistent with what we observed in the stacked bar charts in @fig-eda-transport-mode-continuous-R and @fig-eda-transport-mode-continuous-py. Similarly, at $\alpha=0.05$, there is statistical evidence that the odds of choosing `Public_Transit` versus `Car` among those with access to car differ from the same odds among those who do not have access to car. This is also consistent with what was observed during the EDA in @fig-eda-transport-mode-distribution-R.


### Regressors Associated with Transport Mode {#sec-regressors}

```{r}
#| label: multinom-test-inline
#| echo: false
#| message: false
#| warning: false

# Coefficient table on the log-odds scale: estimates, standard errors, z, p
coef_log <- broom::tidy(multinom_test, conf.int = FALSE) |>
  dplyr::rename(outcome_class = y.level)

# Coefficient table on the odds-ratio scale, with Wald 95% confidence intervals
coef_or <- broom::tidy(multinom_test, conf.int = TRUE, exponentiate = TRUE) |>
  dplyr::rename(outcome_class = y.level)

# Non-baseline outcome classes, taken from the data rather than hard coded
non_baseline <- setdiff(levels(test_set$transport_mode), "Car")
class_bike <- grep("Bicycle", non_baseline, value = TRUE)
class_pt   <- grep("Transit", non_baseline, value = TRUE)
class_walk <- grep("Walk",    non_baseline, value = TRUE)

# Pull a single quantity for a given outcome class and term.
# Fails loudly at render time if the class or term name does not match.
get_stat <- function(class, term_name, column, tbl = coef_or) {
  value <- tbl[[column]][tbl$outcome_class == class & tbl$term == term_name]
  if (length(value) != 1) {
    stop("No unique match for outcome class '", class, "' and term '", term_name, "'.")
  }
  value
}

# Numeric formatting that degrades gracefully under quasi-complete separation.
# Returns bare LaTeX, so inline calls are wrapped in $ ... $ in the prose.
fmt_num <- Vectorize(function(x, digits = 3) {
  if (is.na(x)) return("\\text{NA}")
  if (!is.finite(x)) return(paste0(if (x < 0) "-" else "", "\\approx \\infty"))
  if (x != 0 && abs(x) < 10^(-digits)) return("\\approx 0")
  if (abs(x) >= 1e4) {
    e <- floor(log10(abs(x)))
    return(sprintf("%.2f \\times 10^{%d}", x / 10^e, e))
  }
  formatC(x, format = "f", digits = digits)
})

# p-values: bare form for tables, prefixed form for prose
fmt_pval <- function(p) {
  if (p < 0.001) "< 0.001" else formatC(p, format = "f", digits = 3)
}
fmt_p <- function(p) {
  if (p < 0.001) "< 0.001" else paste0("= ", fmt_pval(p))
}

# 95% confidence interval on the odds-ratio scale, as a comma-separated pair
fmt_ci <- function(class, term_name, digits = 3) {
  paste0(
    fmt_num(get_stat(class, term_name, "conf.low"), digits), ", ",
    fmt_num(get_stat(class, term_name, "conf.high"), digits)
  )
}
```

The Multinomial Logistic regression model refitted on the test set provides the **final inferential summary** for this case study. Note that the inferential results should be stated **cautiously**. The model gives a coherent way to describe how the odds of each transport mode relative to `Car` change with commuter characteristics. 

The fitted model can be written as

$$
\log \left( \frac{\Pr(Y_i = j)}{\Pr(Y_i = \text{Car})} \right) = \hat{\beta}_{0,j} + \hat{\beta}_{1,j} x_{i,\texttt{distance}} + \hat{\beta}_{2,j} x_{i,\texttt{age}} + \hat{\beta}_{3,j} x_{i,\texttt{car}} + \hat{\beta}_{4,j} x_{i,\texttt{weekend}},
$$ {#eq-fitted-model}

where $j$ is one of the non-baseline levels: `` `r class_bike` ``, `` `r class_pt` `` or `` `r class_walk` ``, and the regressors are denoted as $x_{i,\texttt{distance}}, x_{i,\texttt{age}}, x_{i,\texttt{car}},$ and $x_{i,\texttt{weekend}}$.

Every effect below is reported on the **odds ratio** scale, obtained by exponentiating $\hat{\beta}_{k,j}$, so that each entry is multiplicative with respect to the baseline category `Car`. An odds ratio above $1$ corresponds to higher odds of category $j$ relative to `Car`, and an odds ratio below $1$ corresponds to lower odds. Only terms that are statistically significant at the $\alpha = 0.05$ level are reported. Intercepts are excluded, since they describe the baseline log-odds when every regressor is zero rather than an association between a regressor and transport mode.

```{r}
#| label: multinom-significant-regressors
#| echo: false
#| results: asis

alpha_level <- 0.05

# Display metadata for every non-intercept term. `unit` names the change being
# described, `ref` supplies the comparison level for categorical terms, and
# `type` records the measurement scale.
term_meta <- list(
  commute_distance = list(
    label = "`commute_distance`",
    bare  = "`commute_distance`",
    type  = "Continuous",
    unit  = "each additional kilometre of one-way commute distance",
    ref   = NULL
  ),
  age = list(
    label = "`age`",
    bare  = "`age`",
    type  = "Continuous",
    unit  = "each additional year of age",
    ref   = NULL
  ),
  has_car_availableYes = list(
    label = "`has_car_available`",
    bare  = "`has_car_available`",
    type  = "Categorical",
    unit  = "having a car available",
    ref   = "relative to not having a car available"
  ),
  weekend_commuterYes = list(
    label = "`weekend_commuter`",
    bare  = "`weekend_commuter`",
    type  = "Categorical",
    unit  = "commuting on a weekend",
    ref   = "relative to commuting on a weekday"
  )
)

# Fall back gracefully if the model is refitted with a term not listed above
meta_for <- function(tm) {
  if (!is.null(term_meta[[tm]])) return(term_meta[[tm]])
  list(
    label = paste0("`", tm, "`"),
    bare  = paste0("`", tm, "`"),
    type  = "Unclassified",
    unit  = paste0("a one-unit increase in `", tm, "`"),
    ref   = NULL
  )
}

# "a", "a and b", "a, b and c"
and_list <- function(x) {
  if (length(x) <= 1) return(paste(x, collapse = ""))
  paste0(paste(x[-length(x)], collapse = ", "), " and ", x[length(x)])
}

# Significant, non-intercept terms, ordered by outcome class and then by term.
# Significance is read from the Wald p-values on the log-odds scale.
class_levels <- c(class_bike, class_pt, class_walk)
sig <- coef_log[
  coef_log$term != "(Intercept)" &
    !is.na(coef_log$p.value) &
    coef_log$p.value < alpha_level,
  c("outcome_class", "term", "p.value")
]
sig <- sig[order(match(sig$outcome_class, class_levels),
                 match(sig$term, names(term_meta))), ]

# Interpretation of the exponentiated coefficient for one class-term pair
interpret <- function(k, tm) {
  or   <- get_stat(k, tm, "estimate")
  meta <- meta_for(tm)
  others_txt <- and_list(
    vapply(setdiff(names(term_meta), tm),
           function(o) meta_for(o)$bare, character(1))
  )
  if (is.na(or) || !is.finite(or) || or <= 0) {
    return(paste0(
      "Not interpretable on the odds-ratio scale. The estimate is an artifact ",
      "of quasi-complete separation rather than a stable association."
    ))
  }
  direction <- if (or >= 1) {
    paste0(fmt_num(100 * (or - 1), 1), "% higher odds")
  } else {
    paste0(fmt_num(100 * (1 - or), 1), "% lower odds")
  }
  ref_txt <- if (is.null(meta$ref)) "" else paste0(", ", meta$ref)
  sprintf(
    paste0("Holding %s constant, %s is associated with %s times the odds of ",
           "choosing `%s` over `Car`%s, that is %s."),
    others_txt, meta$unit, fmt_num(or, 2), k, ref_txt, direction
  )
}

if (nrow(sig) == 0) {

  cat("No regressor is statistically significant at the $\\alpha = ",
      format(alpha_level), "$ level in this refit.\n\n", sep = "")

} else {

  # Narrative summary: for each significant term, the classes it is significant in
  by_term <- split(as.character(sig$outcome_class), as.character(sig$term))
  by_term <- by_term[order(match(names(by_term), names(term_meta)))]
  phrases <- vapply(names(by_term), function(tm) {
    ks <- by_term[[tm]]
    paste0(meta_for(tm)$label, " in the ", and_list(paste0("`", ks, "`")),
           if (length(ks) == 1) " equation" else " equations")
  }, character(1))
  phrases_txt <- if (length(phrases) == 1) {
    phrases
  } else {
    paste0(paste(phrases[-length(phrases)], collapse = "; "),
           "; and ", phrases[length(phrases)])
  }

  cat("At the $\\alpha = ", format(alpha_level),
      "$ level, the statistically significant regressors are ", phrases_txt,
      ". They are summarized in @tbl-significant-regressors.\n\n", sep = "")

  cat("| Category | Term | Type | Interpretation |\n")
  cat("|---|---|---|---|\n")
  for (r in seq_len(nrow(sig))) {
    k  <- as.character(sig$outcome_class[r])
    tm <- as.character(sig$term[r])
    cat(sprintf(
      "| `%s` versus `Car` | %s | %s | %s |\n",
      k, meta_for(tm)$label, meta_for(tm)$type, interpret(k, tm)
    ))
  }
  cat("\n: Statistically significant regressors in the testing-set refit of the",
      " Multinomial Logistic regression model.",
      " {#tbl-significant-regressors}\n\n", sep = "")
}
```

Terms that do not appear above are not statistically significant at the $5\%$ level and are therefore not interpreted here. The complete set of estimates, standard errors, $z$ statistics, $p$-values, odds ratios and confidence intervals for every term, significant or not, is available in @tbl-multinom-test-fit-python.

### Predictive Results

For the predictive inquiry, we evaluate the model fitted on the training data by comparing its predictions with the observed response categories in the testing data. Let's recall from @sec-practical-adequacy that we retain the coefficient estimates from the training data instead of using the test set refit from @sec-inference, according to the Golden Rule of Machine Learning, to obtain unbiased classification-based measures from the test set. We will use the Multinomial Logistic regression fitted in @sec-estimation to predict on the held-out test set, and create confusion matrices in `R` and `Python` by using `confustionMatrix` from `caret` and `confusion_matrix` from `sklearn` respectively. 

::: {.panel-tabset}

## R Code

```{.r}
# Predict on the test set using the fitted model
y_pred <- predict(multinom_fit, newdata = test_set)

# Create confusion matrix
mode_levels <- c("Car", "Public_Transit", "Bicycle", "Walking")
conf_matrix <- confusionMatrix(factor(y_pred, levels = mode_levels),
                              factor(test_set$transport_mode, levels = mode_levels))

# Visualise
cm_df <- as.data.frame(as.table(conf_matrix$table))
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

# Order axes so the main diagonal (correct matches) runs top-left to bottom-right
cm_df$Predicted <- factor(cm_df$Predicted, levels = levels(cm_df$Predicted))
cm_df$Actual    <- factor(cm_df$Actual,    levels = rev(levels(cm_df$Actual)))

cm_df |>
  ggplot(aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  labs(title = "Confusion Matrix (Test Set)",
       x = "Predicted label", y = "True label") +
  theme_bw(base_size = 13)
```

## Python Code

```{.python}
# Factorize the response and set the baseline model
train_set["transport_mode"] = pd.Categorical(
    train_set["transport_mode"],
    categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)

# Design matrix and response
y = train_set["transport_mode"]
X = train_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]

# Add intercept
X = sm.add_constant(X)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_numeric = X.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes

# Fit Multinomial Logistic regression
model = sm.MNLogit(y, X_numeric)
multinom_fit = model.fit()

# Predict on the test set using the fitted model
test_set["transport_mode"] = pd.Categorical(
  test_set["transport_mode"],
  categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)
y_test = test_set["transport_mode"]
X_test = test_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]
X_test = sm.add_constant(X_test)
X_test_num = X_test.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_test_num[col] = X_test_num[col].cat.codes
y_pred_test = multinom_fit.predict(X_test_num).idxmax(axis=1)

# Confusion matrix
# Display order for the confusion matrix axes
cm_order = ["Car", "Public_Transit", "Bicycle", "Walking"]
order_codes = [list(y_test.cat.categories).index(c) for c in cm_order]

cm = confusion_matrix(y_test.cat.codes, y_pred_test, labels=order_codes)
disp = ConfusionMatrixDisplay(confusion_matrix=cm,
                              display_labels=cm_order)
disp.plot(cmap="Blues")
disp.ax_.set_title("Confusion Matrix (Test Set)")
fig = disp.figure_
fig.tight_layout()
fig
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| label: fig-confusion-matrix-test-R
#| fig-cap: "Confusion matrix of predicted versus actual transport modes on the held-out test set."
#| echo: false
#| message: false
#| warning: false

multinom_fit <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = train_set)

y_pred <- predict(multinom_fit, newdata = test_set)
mode_levels <- c("Car", "Public_Transit", "Bicycle", "Walking")
conf_matrix <- confusionMatrix(factor(y_pred, levels = mode_levels),
                              factor(test_set$transport_mode, levels = mode_levels))

cm_df <- as.data.frame(as.table(conf_matrix$table))
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

cm_df$Predicted <- factor(cm_df$Predicted, levels = levels(cm_df$Predicted))
cm_df$Actual    <- factor(cm_df$Actual,    levels = rev(levels(cm_df$Actual)))

cm_df |>
  ggplot(aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  labs(title = "Confusion Matrix (Test Set)",
       x = "Predicted label", y = "True label") +
  theme_bw(base_size = 13)
```

## Python Output

```{python}
#| label: fig-confusion-matrix-test-py
#| fig-cap: "Confusion matrix of predicted versus actual transport modes on the held-out test set."
#| echo: false
#| message: false
#| warning: false

# Factorize the response and set the baseline model
train_set["transport_mode"] = pd.Categorical(
    train_set["transport_mode"],
    categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)

# Design matrix and response
y = train_set["transport_mode"]
X = train_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]

# Add intercept
X = sm.add_constant(X)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_numeric = X.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes

# Fit Multinomial Logistic regression
model = sm.MNLogit(y, X_numeric)
multinom_fit = model.fit()

test_set["transport_mode"] = pd.Categorical(
  test_set["transport_mode"], 
  categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)
y_test = test_set["transport_mode"]
X_test = test_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]
X_test = sm.add_constant(X_test)
X_test_num = X_test.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_test_num[col] = X_test_num[col].cat.codes
y_pred_test = multinom_fit.predict(X_test_num).idxmax(axis=1)

# Confusion matrix
# Display order for the confusion matrix axes
cm_order = ["Car", "Public_Transit", "Bicycle", "Walking"]
order_codes = [list(y_test.cat.categories).index(c) for c in cm_order]

cm = confusion_matrix(y_test.cat.codes, y_pred_test, labels=order_codes)
disp = ConfusionMatrixDisplay(confusion_matrix=cm,
                              display_labels=cm_order)
disp.plot(cmap="Blues")
disp.ax_.set_title("Confusion Matrix (Test Set)")
fig = disp.figure_
fig.tight_layout()
fig
```

:::

```{r}
#| label: test-cm-inline
#| echo: false
#| message: false
#| warning: false

# Recall and precision for one response category, read from the test-set
# confusion matrix produced above. In caret's `table`, rows index the predicted
# category and columns index the true one, so a row sum is everything predicted
# as that category and a column sum is everything that truly belongs to it.
cm_recall <- function(cls) {
  100 * conf_matrix$table[cls, cls] / sum(conf_matrix$table[, cls])
}

cm_precision <- function(cls) {
  100 * conf_matrix$table[cls, cls] / sum(conf_matrix$table[cls, ])
}

# Percentages are reported to one decimal place
fmt_pct <- function(x) sprintf("%.1f", x)
```


Let's use the confusion matrix in @fig-confusion-matrix-test-R to calculate the recall and precision of each `transport_mode` category. The `Car` and `Public_Transit` classes are predicted most reliably with recalls of $`r fmt_pct(cm_recall("Car"))`\%$ and $`r fmt_pct(cm_recall("Public_Transit"))`\%$ respectively. On the other hand, the true `Bicycle` observations in the test set are all misclassified as `Car`. For the five `Walking` observations in the test set, three are misclassified as `Car` and two of them are correctly classified, which means that `Walking` has a recall of $`r fmt_pct(cm_recall("Walking"))`\%$.

The confusion matrix shows that the predictions are concentrated almost entirely on the two majority categories: `Car` and `Public_Transit`. `Car` is recovered for most of the commuters who actually chose it, so its recall is high, but its precision is relatively low, because a substantial number of commuters whose true transport mode is `Public_Transit` are predicted as `Car`. `Public_Transit` attains the highest precision (`r sprintf("%.1f%%", 100 * conf_matrix$table["Public_Transit", "Public_Transit"] / sum(conf_matrix$table["Public_Transit", ]))`) among the four categories, since a prediction of `Public_Transit` is usually correct, while its recall is somewhat lower than that of `Car` for the reason just given. `Bicycle` is never predicted at all, so its recall is zero and precision is undefined because the model assigns no observation to `Bicycle`. `Walking` is predicted for only a handful of commuters, and all of these predictions are correct, which yields perfect precision and low recall for `Walking`. Taken together, the pattern in the confusion matrix is characteristic of class imbalance in the dataset. It is worth noting that this behaviour is not a symptom of overfitting because the predictive pattern on the held-out test set closely resembles the pattern already visible on the training data in @fig-confusion-matrix-train-R and @fig-confusion-matrix-train-py.

::: {.panel-tabset}

## R Code

```{.r}
# Compute the performance metrics
overall_accuracy_test <- conf_matrix$overall['Accuracy']
macro_recall_test <- mean(conf_matrix$byClass[,"Recall"], na.rm = TRUE)
macro_f1_test <- mean(conf_matrix$byClass[,"F1"], na.rm = TRUE)
cat("Overall Accuracy:", round(overall_accuracy_test, 3), "\n")
cat("Macro Recall:", round(macro_recall_test, 3), "\n")
cat("Macro F1:", round(macro_f1_test, 3), "\n")
```

## Python Code

```{.python}
# Compute the performance metrics
acc_test = accuracy_score(y_test.cat.codes, y_pred_test)
recall_test = recall_score(y_test.cat.codes, y_pred_test, average='macro')
# Average F1 only over the classes the model actually predicts, so a
# never-predicted class (undefined precision) is dropped rather than
# counted as 0 — matching caret's byClass F1 with na.rm = TRUE in the R tab.
predicted_labels = np.unique(y_pred_test)
f1_test = f1_score(y_test.cat.codes, y_pred_test, labels=predicted_labels, average='macro')
print(f"Overall Accuracy: {acc_test:.3f}")
print(f"Macro Recall: {recall_test:.3f}")
print(f"Macro F1: {f1_test:.3f}")
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
overall_accuracy_test <- conf_matrix$overall['Accuracy']
macro_recall_test <- mean(conf_matrix$byClass[,"Recall"], na.rm = TRUE)
macro_f1_test <- mean(conf_matrix$byClass[,"F1"], na.rm = TRUE)
cat("Overall Accuracy:", round(overall_accuracy_test, 3), "\n")
cat("Macro Recall:", round(macro_recall_test, 3), "\n")
cat("Macro F1:", round(macro_f1_test, 3), "\n")
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
acc_test = accuracy_score(y_test.cat.codes, y_pred_test)
recall_test = recall_score(y_test.cat.codes, y_pred_test, average='macro')
# Average F1 only over the classes the model actually predicts, so a
# never-predicted class (undefined precision) is dropped rather than
# counted as 0 — matching caret's byClass F1 with na.rm = TRUE in the R tab.
predicted_labels = np.unique(y_pred_test)
f1_test = f1_score(y_test.cat.codes, y_pred_test, labels=predicted_labels, average='macro')
print(f"Overall Accuracy: {acc_test:.3f}")
print(f"Macro Recall: {recall_test:.3f}")
print(f"Macro F1: {f1_test:.3f}")
```

:::

Because the coefficient estimates used to produce these predictions come entirely from the training data, the measures computed on the held-out test set estimate the model's **out-of-sample** predictive performance, whereas the corresponding measures computed on the training data are in-sample and optimistic by construction. @tbl-train-test-metrics places the two sets of measures side by side.

| **Measure** | **Training set** | **Testing set** |
|:---|:---:|:---:|
| Overall accuracy | `r 100.0*round(overall_accuracy, 3)`\% | `r 100.0*round(overall_accuracy_test, 3)`\% |
| Macro recall | `r 100.0*round(macro_recall, 3)`\% | `r 100.0*round(macro_recall_test, 3)`\% |
| Macro F1 score | `r 100.0*round(macro_f1, 3)`\% | `r 100.0*round(macro_f1_test, 3)`\% |

: Classification-based measures of the Multinomial Logistic regression model, computed on the training data used for estimation and on the held-out test data. {#tbl-train-test-metrics}

The three measures change very little between the two columns, and that comparison is the one that matters for the **predictive inquiry**. Since the coefficient estimates were obtained from the training data alone, a substantial deterioration from the training column to the testing column would be evidence of overfitting, and no such evidence is present. The model therefore generalizes to commuters that it has not encountered about as well as it describes the commuters it was fitted on. This is a statement about the stability of the model rather than about the quality of its predictions. Overall accuracy is dominated by the correct `Car` and `Public_Transit` predictions, which together account for the large majority of the test-set observations, whereas macro recall and the macro F1 score weight all four transport modes equally and are relatively lower. To conclude the predictive inquiry, our classification model distinguishes `Car` from `Public_Transit` at the accuracy reported in @tbl-train-test-metrics, but it does not effectively recover the true `Bicycle` or `Walking` commuters, so it should not be relied upon if the main goal is to predict either of those two transport modes.

::: {.Tip}
::::{.Tip-header}
Tip on Class Imbalance and Prediction
::::
::::{.Tip-container}
The low recall of `Walking` and `Bicycle` in the test set follows from the structure of the MLE solution under class imbalance. Differentiating the log-likelihood function in @eq-log-likelihood-multinomial with respect to the intercept $\beta_0^{(j,1)}$ of a non-baseline category $j$ and setting the partial derivative to zero yields the following equation:

$$
\sum_{i=1}^n \hat{p}_{i,j} = \sum_{i=1}^n y_{i,j} = n_j,
$$

where $n_j$ is the number of observations in category $j$ in the train set. Consequently, the fitted probabilities of each category average exactly its sample proportion in the train set, which is approximately $3.9\%$ for `Bicycle` and approximately $2.1\%$ for `Walking`. Since the classification rule assigns each observation to the category with the highest fitted probability, a minority category is predicted only when the regressors raise its fitted probability well above its small sample proportion, and the estimated slope coefficients do not produce a shift of that magnitude for any regressor values observed in the test set. `Bicycle` never attains the maximum fitted probability for the regressor values observed in the test set, due to which `Bicycle` is never predicted in the confusion matrix above.
::::
:::

## Storytelling

The **storytelling stage** is where we translate the technical results of the Multinomial Logistic regression workflow into a message that is useful for an applied audience. Up to this point, we have defined the study design, wrangled the transport data, explored the response and regressors, fitted a Multinomial Logistic regression model, checked goodness of fit, and organized the inferential and predictive results.

To complete the [data science workflow](https://alexrod61.github.io/regression-cookbook/book/01-intro.html#sec-ds-workflow), let's step back from the coefficients, tests, and confusion matrices, and focus on how the results can be interpreted and applied in practice. A useful analysis must explain **what the results mean for the people who asked the original question**, which features of the evidence are the strongest, where the remaining uncertainty lies, and how the model should—and should not—be used.

For this case study on understanding commuting behaviours, the Multinomial Logistic regression model helps us understand how commute distance, age, car availability, and weekend commuting status relate to an individual's primary mode of transport, an outcome with four unordered categories: `Car`, `Public Transit`, `Bicycle`, and `Walking`. We will distill the key takeaways from the analysis and discuss what they imply for commuters, transportation planners, and policy makers.

### Key Insights From the Model

Car availability and commute distance are the two significant regressors in our model. Having access to a car is associated with lower odds of using public transit instead of driving; the corresponding effect for cycling or walking is not statistically significant, so car access alone does not appear to explain the choice to bike or walk in this dataset. Commute distance moves the odds of public transit and cycling in opposite directions relative to driving, such that each additional kilometre increases the odds of choosing public transit over the car, while decreasing the odds of cycling over the car. This matches the intuition that the car and public transit become the practical options once a commute gets long, while cycling remains attractive mainly over short distances.

Age and weekend commuting status tell a more conditional story. On the original, imbalanced training data, neither regressor reaches statistical significance for most categories, largely because the sample contains too few cyclists and walkers to estimate their effects precisely. Once the minority classes are upsampled, both regressors become significant predictors of `Walking`, and age additionally becomes significant for `Bicycle`: older individuals and weekend commuters are more likely to walk than drive, once the class imbalance no longer masks the effect. This is a useful reminder that failing to reject $H_0$ is not the same as confirming $H_0$ because a lack of significance can potentially reflect insufficient data rather than a genuinely null effect for rare categories.

Some estimates need a caveat rather than a substantive interpretation. The test set contains only one `Walking` observation and five `Bicycle` observations, none of the latter being weekend commuters, so the test set model suffers from complete separation. In R, the `weekend_commuter` effect for `Bicycle` and several `Walking` coefficients are flagged as significant with implausibly large or near-zero odds ratios, standard errors of essentially zero, and extreme test statistics. In Python, `statsmodels` assigns the same terms enormous standard errors and $p$-values near 1, and reports that the fit did not fully converge. None of this is evidence about the true effects; the terms are simply not identifiable from the available data. Analysts should flag this kind of estimate rather than interpret it, and should be cautious of near-zero or implausibly large odds ratios whenever a category is small.

As a predictive tool, the model favours the categories it has seen the most of. The confusion matrices for both the training and test sets show noticeably stronger performance for `Car` and `Public Transit`, the two dominant categories, than for `Bicycle` and `Walking`. Upsampling helps the model learn the minority classes better during estimation, but it cannot manufacture additional information at prediction time, so macro-averaged recall and F1 remain modest even after balancing. The model is best used as a tool for understanding how commute distance, car availability, age, and weekend habits relate to transport mode choice, rather than as a precise classifier for the least common commuting behaviours.

::: {.Tip}
::::{.Tip-header}
Using the Results Responsibly
::::
::::{.Tip-container}
It is important to remember that this analysis is associational, not causal. The model does not prove that changing someone's commute distance or car access would, on its own, change how they travel; it shows that, in this dataset, these regressors are consistently linked to the mode people choose once other measured factors are accounted for.

Nevertheless, the results provide a useful evidence-based starting point for discussion. They help identify which regressors are most closely linked to transport mode choice, and which associations, such as age and weekend commuting for the rarer categories, may only become visible with more balanced or more abundant data. For transportation planners, this kind of insight can help prioritize where infrastructure investment, such as bike lanes or improved transit access, is likely to shift commuting patterns the most.
::::
:::
