<!-- Google tag (gtag.js) -->
<script async src="https://www.googletagmanager.com/gtag/js?id=G-7PRVEBE1EF"></script>
<script>
  window.dataLayer = window.dataLayer || [];
  function gtag(){dataLayer.push(arguments);}
  gtag('js', new Date());

  gtag('config', 'G-7PRVEBE1EF');
</script>

# Multinomial Logistic Regression {#sec-multinomial-logistic}

::: {.Warning}
::::{.Warning-header}
When to Use and Not Use Classical Poisson Regression
::::
::::{.Warning-container}
Multinomial Logistic regression is a type of generalized linear model that is appropriately used under the following conditions:


However, Multinomial Logistic regression should not be used in the following scenarios:

::::
:::

::: {#fig-multinomial-regression}
```{mermaid}
mindmap
  root((Regression 
  Analysis)
    Continuous <br/>Outcome Y
      {{Unbounded <br/>Outcome Y}}
        )Chapter 3: <br/>Ordinary <br/>Least Squares <br/>Regression(
          (Normal <br/>Outcome Y)
      {{Nonnegative <br/>Outcome Y}}
        )Chapter 4: <br/>Gamma Regression(
          (Gamma <br/>Outcome Y)
      {{Bounded <br/>Outcome Y <br/> between 0 and 1}}
        )Chapter 5: Beta <br/>Regression(
          (Beta <br/>Outcome Y)
      {{Nonnegative <br/>Survival <br/>Time Y}}
        )Chapter 6: <br/>Parametric <br/> Survival <br/>Regression(
          (Exponential <br/>Outcome Y)
          (Weibull <br/>Outcome Y)
          (Lognormal <br/>Outcome Y)
        )Chapter 7: <br/>Semiparametric <br/>Survival <br/>Regression(
          (Cox Proportional <br/>Hazards Model)
            (Hazard Function <br/>Outcome Y)
    Discrete <br/>Outcome Y
      {{Binary <br/>Outcome Y}}
        {{Ungrouped <br/>Data}}
          )Chapter 8: <br/>Binary Logistic <br/>Regression(
            (Bernoulli <br/>Outcome Y)
        {{Grouped <br/>Data}}
          )Chapter 9: <br/>Binomial Logistic <br/>Regression(
            (Binomial <br/>Outcome Y)
      {{Count <br/>Outcome Y}}
        {{Equidispersed <br/>Data}}
          )Chapter 10: <br/>Classical Poisson <br/>Regression(
            (Poisson <br/>Outcome Y)
        {{Overdispersed <br/>Data}}
          )Chapter 11: <br/>Negative Binomial <br/>Regression(
            (Negative Binomial <br/>Outcome Y)
        {{Zero Inflated <br/>Data}}
          )Chapter 12: <br/>Zero Inflated <br/>Poisson <br/>Regression(
            (Zero Inflated <br/>Poisson <br/>Outcome Y)
        {{Overdispersed or <br/>Underdispersed <br/>Data}}
          )Chapter 13: <br/>Generalized <br/>Poisson <br/>Regression(
            (Generalized <br/>Poisson <br/>Outcome Y)
      {{Categorical <br/>Outcome Y}}
        {{Nominal <br/>Outcome Y}}
          )Chapter 14: <br/>Multinomial <br/>Logistic <br/>Regression(
            (Multinomial <br/>Outcome Y)
```
:::

```{r setup-r, include=FALSE}
# Load required libraries
library(devtools)
library(readr)
library(tibble)
library(tidyr)
library(dplyr)
library(magrittr)
library(ggplot2)
library(gridExtra) # combine ggplot2 figures
library(ggExtra) # draw marginal histograms
library(broom)
library(knitr)
#devtools::install_github("andytai7/cookbook")
library(cookbook)
library(nnet) # fit multinomial regression
library(caret) # upsampling and confusion matrix

library(reticulate)
py_config()

py_require("pandas")
py_require("numpy")
py_require("seaborn")
py_require("pyreadr")
py_require("plotnine")
py_require("matplotlib")
py_require("statsmodels")
py_require("scikit-learn")
py_require("spicy")

reticulate::py_install("plotnine")

pandas <- import("pandas")
numpy <- import("numpy")
seaborn <- import("seaborn")
pyreadr <- import("pyreadr")
matplotlib <- import("matplotlib")
statsmodels <- import("statsmodels")
sklearn <- import("sklearn")
scipy <- import("scipy")
plotnine <- import("plotnine")
```

```{python setup-py, include=FALSE}
import numpy as np
import pandas as pd
from plotnine import *
from statsmodels.formula.api import glm, ols, mnlogit
from statsmodels.genmod.families import Poisson
from matplotlib import pyplot as plt
from sklearn.metrics import mean_squared_error, mean_absolute_error, classification_report
import scipy.stats as stats
import seaborn as sns
from matplotlib.gridspec import GridSpec
from sklearn.linear_model import LogisticRegression
from sklearn.metrics import confusion_matrix, ConfusionMatrixDisplay, classification_report, log_loss, accuracy_score, recall_score, f1_score
from sklearn.utils import resample
import statsmodels.api as sm
```

::: {.LO}
::::{.LO-header}
Learning Objectives
::::
::::{.LO-container}
By the end of this chapter, you will be able to:

- Describe why ordinary linear models are inappropriate for multi-class nominal outcomes.
- Determine when multinomial logistic regression is an appropriate modeling choice (unordered categorical response with >2 categories; suitable predictors; independent observations).
- Write down the multinomial likelihood and the system of logit link functions relative to a reference category; express class probabilities that sum to one.
- Understand the computation procedure for coefficient estimation via maximum likelihood.
- Interpret multinomial logistic regression coefficients in real scenarios (log-odds vs. baseline, odds ratios, contrasts between non-baseline categories, predicted probabilities).
- Evaluate model performance and construct confidence intervals (confusion matrix, accuracy, log-loss, cross-validation; Wald/LR tests and CIs).
::::
:::

## Introduction

Multinomial logistic regression is a statistical modeling technique used to analyze relationships between a multi-class categorical response variable and a set of explanatory variables. Unlike binary logistic regression, which models outcomes with two categories, multinomial logistic regression extends the framework to outcomes with more than two unordered categories. It estimates the probability of each possible outcome as a function of the predictors by modeling the log-odds of each category relative to a reference category. This method is widely used in fields such as transportation, marketing, and social sciences where the goal is to understand or predict categorical choices among multiple alternatives.


### Multinomial Logistic Regression Assumptions

- The response or dependent variable should be measured at the nominal level with more than two values. 
- There should be one or more independent variables that are continuous, ordinal or nominal (including dichotomous variables).
- **Logit linearity assumption**: There needs to be a linear relationship between any continuous independent variables and the logit transformation of the dependent variable.
- **Independent observations**: The observations should be independent and the dependent variable should have mutually exclusive and exhaustive categories (i.e. no individual belonging to two different categories).
- **No Multicollinearity**: Multicollinearity occurs when you have two or more independent variables that are highly correlated with each other.
- There should be no outliers, high leverage values or highly influential points.

## Case Study

We will explore a dataset containing information on individuals' commuting behaviors and demographics, with a focus on their primary mode of transportation. We explore a couple of research problems based on the dataset. 

- How is the transportation mode of people (driving own cars or taking public transportation specifically) **associated** by their commuting distances, given the individuals' commuting behaviors and demographics including if they have a car, if they commute on weekends, and ages?
- Can we accurately **predict** the transportation mode of an individual based on their commuting behaviors and demographics including if they have a car, if they commute on weekends, their commute distance, and age?

The two research problems correspond to **inferential** and **predictive** **statistical objectives** respectively:

- To estimate the **association** between the explanatory variable---commute distance---and the probability of each category of the response---using each transport mode (car v.s. public transportation specifically), given the confounding effects of age, car availability, and weekend commuter status. 
- To develop a **predictive** model that classifies individuals into their most likely transport mode using the predictors commute distance, age, car availability, and weekend commuting status. 

## Data Collection and Wrangling

The dataset includes the following five variables:

- `transport_mode` (categorical, 4 levels):
  - The primary mode of transport used by the individual. Categories: "Bicycle", "Car", "Public Transit", "Walking". This variable is used as the response in data modelling.

- `commute_distance` (numeric):
  - The one-way commuting distance in kilometers.

- `age` (numeric):
   - The age of the individual in years.

- `has_car_available` (binary categorical):
  - Indicates whether the individual has access to a car. Levels: "Yes" or "No".

- `weekend_commuter` (binary categorical):
  - Indicates whether the individual commutes during weekends. Levels: "Yes" or "No".

::: {.panel-tabset}

## R Code

```{.r}
data(multinomial_transport)
data <- multinomial_transport[,-1]
cat("There are", nrow(data), "observations and", ncol(data), "variables in the dataset.")
print(data)
```

## Python Code

```{.python}
url = "https://raw.githubusercontent.com/andytai7/cookbook/refs/heads/main/raw-data/multinomial_transport.csv"
data = pd.read_csv(url)
data = data.iloc[:, :-1]
print(f"There are {data.shape[0]} observations and {data.shape[1]} variables in the dataset.")
print(data)
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
data(multinomial_transport)
data <- multinomial_transport %>% select(-'income_bracket')
cat("There are", nrow(data), "observations and", ncol(data), "variables in the dataset.")
print(data)
```

## Python Output

```{python}
#| echo: true
#| message: false
#| warning: false
url = "https://raw.githubusercontent.com/andytai7/cookbook/refs/heads/main/raw-data/multinomial_transport.csv"
data = pd.read_csv(url)
data = data.iloc[:, :-1]
print(f"There are {data.shape[0]} observations and {data.shape[1]} variables in the dataset.")
print(data)
```

:::

Factorize the categorical variables. There are four types of transport mode which are by bicycle, car, public transit, and walking. 

::: {.panel-tabset}

## R Code

```{.r}
# Factorize the categorical variables and print their levels
categorical_cols <- c('transport_mode', 'has_car_available', 'weekend_commuter')
for (col in categorical_cols) {
  data[[col]] <- as.factor(data[[col]])
  cat("Levels of", col, ":", levels(data[[col]]), "\n")
}
```

## Python Code

```{.python}
# Factorize the categorical variables and print their levels
categorical_cols = ['transport_mode', 'has_car_available', 'weekend_commuter']
for col in categorical_cols:
  data[col] = data[col].astype('category')
  print(f"Levels of {col}: {list(data[col].cat.categories)}")
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
categorical_cols <- c('transport_mode', 'has_car_available', 'weekend_commuter')
for (col in categorical_cols) {
  data[[col]] <- as.factor(data[[col]])
  cat("Levels of", col, ":", levels(data[[col]]), "\n")
}
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
categorical_cols = ['transport_mode', 'has_car_available', 'weekend_commuter']
for col in categorical_cols:
  data[col] = data[col].astype('category')
  print(f"Levels of {col}: {list(data[col].cat.categories)}")
```

:::

Now, we split the dataset into training and test sets, use training set to fit the model, and use test set to assess the model performance. 

::: {.panel-tabset}

## R Code

```{.r}
# Split the data: first 90% for training, last 10% for test
n <- nrow(data)
split_idx <- floor(0.9 * n)
train_set <- data[1:split_idx, ]
test_set <- data[(split_idx + 1):n, ]

cat("Training set size:", nrow(train_set))
cat("Test set size:", nrow(test_set))
```

## Python Code

```{.python}
# Split the data: first 90% for training, last 10% for test
split_idx = int(len(data) * 0.9)
train_set = data.iloc[:split_idx].reset_index(drop=True)
test_set = data.iloc[split_idx:].reset_index(drop=True)

print(f"Training set size: {len(train_set)}")
print(f"Test set size: {len(test_set)}")
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
n <- nrow(data)
split_idx <- floor(0.9 * n)
train_set <- data[1:split_idx, ]
test_set <- data[(split_idx + 1):n, ]
cat("Training set size:", nrow(train_set))
cat("Test set size:", nrow(test_set))
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
split_idx = int(len(data) * 0.9)
train_set = data.iloc[:split_idx].reset_index(drop=True)
test_set = data.iloc[split_idx:].reset_index(drop=True)
print(f"Training set size: {len(train_set)}")
print(f"Test set size: {len(test_set)}")
```

:::

## Exploratory Data Analysis

Let's visualize the response `transport_mode` and its relationship with the continuous or categorical(binary) explanatory variables (predictors).

::: {.panel-tabset}

## R Code

```{.r}
# Custom color palette `cbPalette` 
cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
# Sort `transport_mode` by frequency
train_set$transport_mode <- factor(train_set$transport_mode, levels = names(sort(table(train_set$transport_mode), decreasing = TRUE)))

p1 <- train_set %>%
  ggplot(aes(x = transport_mode, fill = has_car_available)) +
  geom_bar(position = "stack") +  # use "stack" for counts, "fill" for proportions
  labs(y = "Count", title = "With Car Availability") +
  scale_fill_manual(values=cbPalette) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(title = "Car Availability"))

p2 <- train_set %>%
  ggplot(aes(x = transport_mode, fill = weekend_commuter)) +
  geom_bar(position = "stack") +  # use "stack" for counts, "fill" for proportions
  labs(x = "Transport Mode", y = "Count", title = "With Weekend Commuter Status") +
  scale_fill_manual(values=cbPalette) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(title = "Weekend Commuter"))

grid.arrange(p1, p2, ncol = 2, top = grid::textGrob("Distribution of Transport Mode", gp = grid::gpar(fontsize = 16, fontface = "bold")))
```

## Python Code

```{.python}
# Custom color palette `cbPalette` 
cbPalette = ["#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7"]

# Sort transport_mode by frequency
order = train_set['transport_mode'].value_counts().index
train_set['transport_mode'] = pd.Categorical(train_set['transport_mode'], categories=order, ordered=True)

# Set up the plot
sns.set_theme(style="whitegrid")
fig = plt.figure()
fig.set_size_inches(8, 6)
gs = GridSpec(1, 2, figure=fig)
_ = fig.suptitle("Distribution of Transport Mode", fontsize=16, fontweight='bold')

# Plot: Stacked bar for Car Availability
car_counts = train_set.groupby(['transport_mode', 'has_car_available']).size().unstack(fill_value=0)
car_counts = car_counts.loc[order] 

ax1 = fig.add_subplot(gs[0, 0])
bottom = None
for i, col in enumerate(car_counts.columns):
    ax1.bar(car_counts.index, car_counts[col], bottom=bottom, label=col, color=cbPalette[i % len(cbPalette)])
    bottom = car_counts[col] if bottom is None else bottom + car_counts[col]
ax1.set_title("With Car Availability")
ax1.set_xlabel("Transport Mode")
ax1.set_ylabel("Count")
ax1.legend(title="Car Availability", loc='upper center', bbox_to_anchor=(0.5, -0.15), ncol=2)

# Plot: Stacked bar for Weekend Commuter
weekend_counts = train_set.groupby(['transport_mode', 'weekend_commuter']).size().unstack(fill_value=0)
weekend_counts = weekend_counts.loc[order]

ax2 = fig.add_subplot(gs[0, 1])
bottom = None
for i, col in enumerate(weekend_counts.columns):
    ax2.bar(weekend_counts.index, weekend_counts[col], bottom=bottom, label=col, color=cbPalette[i % len(cbPalette)])
    bottom = weekend_counts[col] if bottom is None else bottom + weekend_counts[col]
ax2.set_title("With Weekend Commuter Status")
ax2.set_xlabel("Transport Mode")
ax2.set_ylabel("")
ax2.legend(title="Weekend Commuter", loc='upper center', bbox_to_anchor=(0.5, -0.15), ncol=2)

# Adjust plot layout
plt.tight_layout(rect=[0, 0.05, 1, 1])
plt.show()
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
train_set$transport_mode <- factor(train_set$transport_mode, levels = names(sort(table(train_set$transport_mode), decreasing = TRUE)))
p1 <- train_set %>%
  ggplot(aes(x = transport_mode, fill = has_car_available)) +
  geom_bar(position = "stack") +  # use "stack" for counts, "fill" for proportions
  labs(x = "Transport Mode", y = "Count", title = "With Car Availability") +
  scale_fill_manual(values=cbPalette) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(title = "Car Availability"))
p2 <- train_set %>%
  ggplot(aes(x = transport_mode, fill = weekend_commuter)) +
  geom_bar(position = "stack") +  # use "stack" for counts, "fill" for proportions
  labs(x = "Transport Mode", y = "", title = "With Weekend Commuter Status") +
  scale_fill_manual(values=cbPalette) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(title = "Weekend Commuter"))
grid.arrange(p1, p2, ncol = 2, top = grid::textGrob("Distribution of Transport Mode", gp = grid::gpar(fontsize = 16, fontface = "bold")))
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
cbPalette = ["#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7"]
order = train_set['transport_mode'].value_counts().index
train_set['transport_mode'] = pd.Categorical(train_set['transport_mode'], categories=order, ordered=True)

sns.set_theme(style="whitegrid")
fig = plt.figure()
fig.set_size_inches(8, 6)
gs = GridSpec(1, 2, figure=fig)
_ = fig.suptitle("Distribution of Transport Mode", fontsize=16, fontweight='bold')

# Plot 1: Stacked bar for Car Availability
car_counts = train_set.groupby(['transport_mode', 'has_car_available']).size().unstack(fill_value=0)
car_counts = car_counts.loc[order] 
ax1 = fig.add_subplot(gs[0, 0])
bottom = None
for i, col in enumerate(car_counts.columns):
    ax1.bar(car_counts.index, car_counts[col], bottom=bottom, label=col, color=cbPalette[i % len(cbPalette)])
    bottom = car_counts[col] if bottom is None else bottom + car_counts[col]
ax1.set_title("With Car Availability")
ax1.set_xlabel("Transport Mode")
ax1.set_ylabel("Count")
ax1.legend(title="Car Availability", loc='upper center', bbox_to_anchor=(0.5, -0.15), ncol=2)

# Plot 2: Stacked bar for Weekend Commuter
weekend_counts = train_set.groupby(['transport_mode', 'weekend_commuter']).size().unstack(fill_value=0)
weekend_counts = weekend_counts.loc[order]
ax2 = fig.add_subplot(gs[0, 1])
bottom = None
for i, col in enumerate(weekend_counts.columns):
    ax2.bar(weekend_counts.index, weekend_counts[col], bottom=bottom, label=col, color=cbPalette[i % len(cbPalette)])
    bottom = weekend_counts[col] if bottom is None else bottom + weekend_counts[col]
ax2.set_title("With Weekend Commuter Status")
ax2.set_xlabel("Transport Mode")
ax2.set_ylabel("")
ax2.legend(title="Weekend Commuter", loc='upper center', bbox_to_anchor=(0.5, -0.15), ncol=2)

# Adjust layout
plt.tight_layout(rect=[0, 0.05, 1, 1])
plt.show()
```

:::

The bar plots displays the distribution of transport mode across car availability and weekend commute status. 

Car is the dominant mode of transport, especially among those who have access to a car (blue bar is much taller than orange).
Public Transit is used by both those with and without car availability, but is more common among those without a car (orange section is relatively larger). 
Bicycle and Walking are rare overall but tend to occur more when no car is available. 

Most weekend commuters do not significantly change their transport mode: the proportions are largely dominated by the “No” (non-weekend commuter) group (orange).
However, Public Transit has a noticeable number of weekend commuters (blue), indicating some weekend reliance on public transit.
Walking and Bicycling show little weekend commuting activity.

Visualize the relationship between `transport_mode` and continuous variables `commute_distance` and `age`.

::: {.panel-tabset}

## R Code

```{.r}
# visualize boxplots and violin plots of commute_distance v.s. transport_mode
p3 <- train_set %>%
    ggplot(aes(x = transport_mode, y = commute_distance, fill = transport_mode)) +
    geom_violin(alpha = 0.6, trim = FALSE) +
    geom_boxplot(alpha = 1, fill = NA) +
    labs(x = "Transport Mode", y = "Commute Distance (km)", title =
        "Commute Distance by Transport Mode") +
    scale_fill_manual(values = cbPalette[3:6]) +
    theme_bw() +
    theme(legend.position = "none") +
    guides(fill = guide_legend(title = "Transport Mode"))

p4 <- train_set %>%
    ggplot(aes(x = transport_mode, y = age, fill = transport_mode)) +
    geom_violin(alpha = 0.6, trim = FALSE) +
    geom_boxplot(alpha = 1, fill = NA) +
    labs(x = "Transport Mode", y = "Age", title =
        "Age by Transport Mode") +
    scale_fill_manual(values = cbPalette[3:6]) +
    theme_bw() +
    theme(legend.position = "none") +
    guides(fill = guide_legend(title = "Transport Mode"))

grid.arrange(p3, p4, ncol = 2, top = grid::textGrob("Box Plots and Violin Plots", gp = grid::gpar(fontsize = 16, fontface = "bold")))
```

## Python Code

```{.python}
sns.set_theme(style="whitegrid")

# Set up side-by-side subplots with shared layout
fig = plt.figure(figsize=(8, 7))
gs = GridSpec(1, 2, figure=fig)
fig.suptitle("Box Plots and Violin Plots", fontsize=16, fontweight='bold')

# Plot Commute Distance vs Transport Mode
ax1 = fig.add_subplot(gs[0, 0])
sns.violinplot(
    data=train_set,
    x='transport_mode',
    y='commute_distance',
    palette=cbPalette[2:6],
    ax=ax1,
    alpha=0.6,
    inner=None,
    order=order,
    cut=3
)
sns.boxplot(
    data=train_set,
    x='transport_mode',
    y='commute_distance',
    showcaps=True,
    boxprops={'facecolor': 'none', 'edgecolor': 'black'},
    whiskerprops={'color': 'black'},
    flierprops={'markerfacecolor': 'black', 'markersize': 3},
    ax=ax1,
    order=order
)
ax1.set_title("Commute Distance by Transport Mode")
ax1.set_xlabel("Transport Mode")
ax1.set_ylabel("Commute Distance (km)")
ax1.get_legend()

# Plot Age vs Transport Mode
ax2 = fig.add_subplot(gs[0, 1])
sns.violinplot(
    data=train_set,
    x='transport_mode',
    y='age',
    palette=cbPalette[2:6],
    ax=ax2,
    alpha=0.6,
    inner=None,
    order=order,
    cut=3
)
sns.boxplot(
    data=train_set,
    x='transport_mode',
    y='age',
    showcaps=True,
    boxprops={'facecolor': 'none', 'edgecolor': 'black'},
    whiskerprops={'color': 'black'},
    flierprops={'markerfacecolor': 'black', 'markersize': 3},
    ax=ax2,
    order=order
)
ax2.set_title("Age by Transport Mode")
ax2.set_xlabel("Transport Mode")
ax2.set_ylabel("Age")
ax2.get_legend()

plt.tight_layout(rect=[0, 0.05, 1, 0.95])
plt.show()
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
p3 <- train_set %>%
    ggplot(aes(x = transport_mode, y = commute_distance, fill = transport_mode)) +
    geom_violin(alpha = 0.6, trim = FALSE) +
    geom_boxplot(alpha = 1, fill = NA) +
    labs(x = "Transport Mode", y = "Commute Distance (km)", title =
        "Commute Distance by Transport Mode") +
    scale_fill_manual(values = cbPalette[3:6]) +
    theme_bw() +
    theme(legend.position = "none") +
    guides(fill = guide_legend(title = "Transport Mode"))
p4 <- train_set %>%
    ggplot(aes(x = transport_mode, y = age, fill = transport_mode)) +
    geom_violin(alpha = 0.6, trim = FALSE) +
    geom_boxplot(alpha = 1, fill = NA) +
    labs(x = "Transport Mode", y = "Age", title =
        "Age by Transport Mode") +
    scale_fill_manual(values = cbPalette[3:6]) +
    theme_bw() +
    theme(legend.position = "none") +
    guides(fill = guide_legend(title = "Transport Mode"))
grid.arrange(p3, p4, ncol = 2, top = grid::textGrob("Box Plots and Violin Plots", gp = grid::gpar(fontsize = 16, fontface = "bold")))
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
sns.set_theme(style="whitegrid")

# Set up side-by-side subplots with shared layout
fig = plt.figure(figsize=(8, 7))
gs = GridSpec(1, 2, figure=fig)
fig.suptitle("Box Plots and Violin Plots", fontsize=16, fontweight='bold')

# Plot Commute Distance vs Transport Mode
ax1 = fig.add_subplot(gs[0, 0])
sns.violinplot(
    data=train_set,
    x='transport_mode',
    y='commute_distance',
    palette=cbPalette[2:6],
    ax=ax1,
    alpha=0.6,
    inner=None,
    order=order,
    cut=3
)
sns.boxplot(
    data=train_set,
    x='transport_mode',
    y='commute_distance',
    showcaps=True,
    boxprops={'facecolor': 'none', 'edgecolor': 'black'},
    whiskerprops={'color': 'black'},
    flierprops={'markerfacecolor': 'black', 'markersize': 3},
    ax=ax1,
    order=order
)
ax1.set_title("Commute Distance by Transport Mode")
ax1.set_xlabel("Transport Mode")
ax1.set_ylabel("Commute Distance (km)")
ax1.get_legend()  # no legend to remove manually

# Plot p4: Age vs Transport Mode
ax2 = fig.add_subplot(gs[0, 1])
sns.violinplot(
    data=train_set,
    x='transport_mode',
    y='age',
    palette=cbPalette[2:6],
    ax=ax2,
    alpha=0.6,
    inner=None,
    order=order,
    cut=3
)
sns.boxplot(
    data=train_set,
    x='transport_mode',
    y='age',
    showcaps=True,
    boxprops={'facecolor': 'none', 'edgecolor': 'black'},
    whiskerprops={'color': 'black'},
    flierprops={'markerfacecolor': 'black', 'markersize': 3},
    ax=ax2,
    order=order
)
ax2.set_title("Age by Transport Mode")
ax2.set_xlabel("Transport Mode")
ax2.set_ylabel("Age")
ax2.get_legend()  # no legend to remove manually

plt.tight_layout(rect=[0, 0.05, 1, 0.95])
plt.show()
```

:::

Box and violin plots present that how commute distance and age vary across different transport modes.

Public transit users tend to have the longest commutes on average, with a wide distribution and a median around 17–18 km.
Car users also have moderately long commutes, with a median around 10-11 km and a relatively broad spread.
Bicycle and Walking are associated with shorter commute distances (medians < 5 km), and their distributions are tightly concentrated near the lower end.
All modes show some long-distance commuters (long tails), especially public transit and car.

Car and public transit users have similar age distributions, with median ages in the mid-30s to early 40s.
Bicycle users skew slightly younger, but with a long tail of older users.
Walking has a much wider spread in age, with a high median (around 60) and some very young walkers as well.
The variability in age is greatest for walking, suggesting diverse usage across age groups.

You can also visualize the relationship across other explanatory variables or predictors in various types of figures for practice.

## Data Modelling

A Multinomial Logistic Regression model is a suitable approach to our statistical inquiries given that `transport_mode` is categorical and nominal (our response of interest) subject to the numerical regressors `commute_distance` and `age` and binary regressors `has_car_available` and `weekend_commuter`. Moreover, its corresponding regression estimates will allow us to measure variable association.

This regression approach assumes a [Multinomial distribution](https://www.sciencedirect.com/topics/mathematics/multinomial-distribution) where $p_{i,1},p_{i,2},\dots,p_{i,m}$ are the probabilities that will belong to categories $1,2,\dots,m$ respectively; i.e., 
$$P(Y_i=1)=p_{i,1}, P(Y_i=2)=p_{i,2}, \dots, P(Y_i=m)=p_{i,m}$$
where
$$
\sum_{j = 1}^m p_{i,j} = p_{i,1} + p_{i,2} + \dots + p_{i,m} = 1.
$$

*A particular highlight is that the Binomial distriution is a special Multinomial distribution when $m=2$.*

The Multinomial Logistic regression **also models the logarithm of the odds**. However, only one logarithm of the odds (or **logit**) will not be enough anymore. Recall we can capture the odds between two categories with a single logit function. **What about adding some other ones?**

Here is what we can do:

1. Pick one of the categories to be the **baseline**. For example, the category "$1$".
2. For each of the **other** categories, we model the logarithm of the odds to the baseline category.

Now, **what is the math for the general case with $m$ response categories and $K$ regressors?** For the $i$th observation, we end up with a system of $m - 1$ linK functions in the Multinomial Logistic regression model as follows:

$$
\begin{gather*} \label{eq:multinomial-model}
\eta_i^{(2,1)} = \log\left[\frac{P(Y_i = 2\mid X_{i,1}, \ldots, X_{i,K})}{P(Y_i = 1 \mid X_{i,1}, \ldots, X_{i,K})}\right] = \beta_0^{(2,1)} + \beta_1^{(2,1)} X_{i, 1} + \beta_2^{(2,1)} X_{i, 2} + \ldots + \beta_K^{(2,1)} X_{i, K} \\
\eta_i^{(3,1)} = \log\left[\frac{P(Y_i = 3\mid X_{i,1}, \ldots, X_{i,K})}{P(Y_i = 1 \mid X_{i,1}, \ldots, X_{i,K})}\right] = \beta_0^{(3,1)} + \beta_1^{(3,1)} X_{i, 1} + \beta_2^{(3,1)} X_{i, 2} + \ldots + \beta_K^{(3,1)} X_{i, K} \\
\vdots \\
\eta_i^{(m,1)} = \log\left[\frac{P(Y_i = m\mid X_{i,1}, \ldots, X_{i,K})}{P(Y_i = 1 \mid X_{i,1}, \ldots, X_{i,K})}\right] = \beta_0^{(m,1)} + \beta_1^{(m,1)} X_{i, 1} + \beta_2^{(m,1)} X_{i, 2} + \ldots + \beta_K^{(m,1)} X_{i, K}.
\end{gather*}
$$

***Note that the superscript $(j, 1)$ in @eq:multinomial-model indicates that the equation is on level $j$ (for $j = 2, \dots, m$) with respect to level $1$. Furthermore, **the regression coefficients are different for each link function**.***

Each of the logit-linear functions in @eq:multinomial-model writes the log of odds between each category $j=2,\dots,m$ and the baseline category $j=1$ as a linear combination of the regressors. To compare between the categories $j=2,\dots,m$, we can simply deduct one equation be another, e.g.,
$$
\begin{gather*} \label{eq:multinomial-deduct}
\eta_i^{(2,1)} - \eta_i^{(3,1)} = \log\left[\frac{P(Y_i = 2\mid X_{i,1}, \ldots, X_{i,K})}{P(Y_i = 1 \mid X_{i,1}, \ldots, X_{i,K})}\right] - \log\left[\frac{P(Y_i = 3\mid X_{i,1}, \ldots, X_{i,K})}{P(Y_i = 1 \mid X_{i,1}, \ldots, X_{i,K})}\right] \\
= \log\left[\frac{P(Y_i = 3\mid X_{i,1}, \ldots, X_{i,K})}{P(Y_i = 2 \mid X_{i,1}, \ldots, X_{i,K})}\right] 
= (\beta_0^{(2,1)}-\beta_0^{(3,1)}) + (\beta_1^{(2,1)}-\beta_1^{(3,1)}) X_{i, 1} + (\beta_2^{(2,1)}-\beta_2^{(3,1)}) X_{i, 2} + \ldots
\end{gather*}
$$

With some algebraic manipulation, we can show that the probabilities $p_{i,1}, p_{i,2}, \dots, p_{i,m}$ of $Y_i$ belonging to categories $1, 2, \dots, m$ are:

$$ 
\begin{gather*} \label{eq:prob-multinomial}
p_{i,1} = P(Y_i = 1 \mid X_{i,1}, \ldots, X_{i,K}) = \frac{1}{1 + \sum_{j = 2}^m \exp \big( \eta_i^{(j,1)} \big)} \\
p_{i,2} = P(Y_i = 2 \mid X_{i,1}, \ldots, X_{i,K}) = \frac{\exp \big( \eta_i^{(2,1)} \big)}{1 + \sum_{j = 2}^m \exp \big( \eta_i^{(j,1)} \big)} \\
\vdots \\
p_{i,m} = P(Y_i = m \mid X_{i,1}, \ldots, X_{i,K}) = \frac{\exp \big( \eta_i^{(m,1)} \big)}{1 + \sum_{j = 2}^m \exp \big( \eta_i^{(j,1)} \big)}.
\end{gather*}
$$

***Note: The equations above are Softmax functions. The multinomial logistic regression is theoretically equivalent to a one-layer neural network with Softmax activation function, where we input $X$, output probability $p_{i,j}$, and minimize the cross-entropy loss function to train the neural network to learn parameters $\beta$.***

***If we sum all $m$ probabilities in @eq:prob-multinomial, the sum will be equal to $1$ for the $i$th observation. This is particularly important when we want to use this model for making predictions in classification matters.***

Goting back to our data example, let us set the Multinomial logistic regression model with `transport_mode` as the response with four classes: `Car`, `Public_Transit`, `Bicycle`, and `Walking` (denoted as `Car`, `Public`, `Bike`, `Walk` respectively) subject to the continuous regressors `commute_distance` and `age`, denoted as $X_{\texttt{distance}}$ and $X_{\texttt{age}}$ respectively, and binary regressors `has_car_available` and `weekend_commuter`, with the dummy variables of class `Yes` denoted as $Z_{\texttt{car}}$ and $Z_{\texttt{weekend}}$ respectively. Use `Car` class as the baseline model.

$$
\begin{align*}
\eta_i^{(\texttt{Public},\texttt{Car})} &= \log\left[\frac{P(Y_i = \texttt{Public} \mid X_{i, \texttt{distance}}, X_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}{P(Y_i = \texttt{Car} \mid X_{i, \texttt{distance}}, X_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}\right] \\
&= \beta_0^{(\texttt{Public},\texttt{Car})} + \beta_1^{(\texttt{Public},\texttt{Car})} X_{i, \texttt{distance}} + \beta_2^{(\texttt{Public},\texttt{Car})} X_{i, \texttt{age}} + \beta_3^{(\texttt{Public},\texttt{Car})} Z_{i, \texttt{car}} + \beta_4^{(\texttt{Public},\texttt{Car})} Z_{i,\texttt{weekend}}, \\
\eta_i^{(\texttt{Bike},\texttt{Car})} &= \log\left[\frac{P(Y_i = \texttt{Bike} \mid X_{i, \texttt{distance}}, X_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}{P(Y_i = \texttt{Car} \mid X_{i, \texttt{distance}}, X_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}\right] \\
&= \beta_0^{(\texttt{Bike},\texttt{Car})} + \beta_1^{(\texttt{Bike},\texttt{Car})} X_{i, \texttt{distance}} + \beta_2^{(\texttt{Bike},\texttt{Car})} X_{i, \texttt{age}} + \beta_3^{(\texttt{Bike},\texttt{Car})} Z_{i, \texttt{car}} + \beta_4^{(\texttt{Bike},\texttt{Car})} Z_{i,\texttt{weekend}}, \\
\eta_i^{(\texttt{Walk},\texttt{Car})} &= \log\left[\frac{P(Y_i = \texttt{Walk} \mid X_{i, \texttt{distance}}, X_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}{P(Y_i = \texttt{Car} \mid X_{i, \texttt{distance}}, X_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}\right] \\
&= \beta_0^{(\texttt{Walk},\texttt{Car})} + \beta_1^{(\texttt{Walk},\texttt{Car})} X_{i, \texttt{distance}} + \beta_2^{(\texttt{Walk},\texttt{Car})} X_{i, \texttt{age}} + \beta_3^{(\texttt{Walk},\texttt{Car})} Z_{i, \texttt{car}} + \beta_4^{(\texttt{Walk},\texttt{Car})} Z_{i,\texttt{weekend}} .
\end{align*}
$$

***In a Multinomial Logistic regression model, each link function has its own intercept and regression coefficients.***

Taking exponential on both sides of the model equations gives the ratio of probability of each category `Public_Transit`, `Bicycle`, and `Walking` over the probability of the baseline level `Car`:

$$
\begin{align*}
&\frac{P(Y_i = \texttt{Public} \mid X_{i, \texttt{distance}}, X_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}{P(Y_i = \texttt{Car} \mid X_{i, \texttt{distance}}, X_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})} = \exp\left[\eta_i^{(\texttt{Public},\texttt{Car})} \right]\\
&= \exp\left[\beta_0^{(\texttt{Public},\texttt{Car})}\right] \exp\left[\beta_1^{(\texttt{Public},\texttt{Car})} X_{i,\texttt{distance}}\right] \exp\left[\beta_2^{(\texttt{Public},\texttt{Car})} X_{i,\texttt{age}}\right] \exp\left[ \beta_3^{(\texttt{Public},\texttt{Car})} Z_{i, \texttt{car}}\right] \exp\left[ \beta_4^{(\texttt{Public},\texttt{Car})} Z_{i,\texttt{weekend}}\right]\\
&\frac{P(Y_i = \texttt{Bike} \mid X_{i, \texttt{distance}}, X_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}{P(Y_i = \texttt{Car} \mid X_{i, \texttt{distance}}, X_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})} = \exp\left[\eta_i^{(\texttt{Bike},\texttt{Car})} \right]\\
&= \exp\left[\beta_0^{(\texttt{Bike},\texttt{Car})}\right] \exp\left[\beta_1^{(\texttt{Bike},\texttt{Car})} X_{i,\texttt{distance}}\right] \exp\left[\beta_2^{(\texttt{Bike},\texttt{Car})} X_{i,\texttt{age}}\right] \exp\left[ \beta_3^{(\texttt{Bike},\texttt{Car})} Z_{i, \texttt{car}}\right] \exp\left[ \beta_4^{(\texttt{Bike},\texttt{Car})} Z_{i,\texttt{weekend}}\right]\\
&\frac{P(Y_i = \texttt{Walk} \mid X_{i, \texttt{distance}}, X_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}{P(Y_i = \texttt{Car} \mid X_{i, \texttt{distance}}, X_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})} = \exp\left[\eta_i^{(\texttt{Walk},\texttt{Car})} \right]\\
&= \exp\left[\beta_0^{(\texttt{Walk},\texttt{Car})}\right] \exp\left[\beta_1^{(\texttt{Walk},\texttt{Car})} X_{i,\texttt{distance}}\right] \exp\left[\beta_2^{(\texttt{Walk},\texttt{Car})} X_{i,\texttt{age}}\right] \exp\left[ \beta_3^{(\texttt{Walk},\texttt{Car})} Z_{i, \texttt{car}}\right] \exp\left[ \beta_4^{(\texttt{Walk},\texttt{Car})} Z_{i,\texttt{weekend}}\right]
\end{align*}
$$

Finally, the probability of $Y_i$ belonging to categories `Car`, `Public_Transit`, `Bicycle`, and `Walking` are:

$$
\begin{align*}
P(Y_i = \texttt{Car} \mid X_{i, \texttt{distance}}, X_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}}) &= \frac{1}{1+\exp\left[\eta_i^{(\texttt{Public},\texttt{Car})} \right]+\exp\left[\eta_i^{(\texttt{Bike},\texttt{Car})}\right]+\exp\left[\eta_i^{(\texttt{Walk},\texttt{Car})} \right]} \\
P(Y_i = \texttt{Public} \mid X_{i, \texttt{distance}}, X_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}}) &= \frac{\exp\left[\eta_i^{(\texttt{Public},\texttt{Car})} \right]}{1+\exp\left[\eta_i^{(\texttt{Public},\texttt{Car})} \right]+\exp\left[\eta_i^{(\texttt{Bike},\texttt{Car})}\right]+\exp\left[\eta_i^{(\texttt{Walk},\texttt{Car})} \right]} \\
P(Y_i = \texttt{Bike} \mid X_{i, \texttt{distance}}, X_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}}) &= \frac{\exp\left[\eta_i^{(\texttt{Bike},\texttt{Car})}\right]}{1+\exp\left[\eta_i^{(\texttt{Public},\texttt{Car})} \right]+\exp\left[\eta_i^{(\texttt{Bike},\texttt{Car})}\right]+\exp\left[\eta_i^{(\texttt{Walk},\texttt{Car})} \right]}\\
P(Y_i = \texttt{Walk} \mid X_{i, \texttt{distance}}, X_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}}) &= \frac{\exp\left[\eta_i^{(\texttt{Walk},\texttt{Car})} \right]}{1+\exp\left[\eta_i^{(\texttt{Public},\texttt{Car})} \right]+\exp\left[\eta_i^{(\texttt{Bike},\texttt{Car})}\right]+\exp\left[\eta_i^{(\texttt{Walk},\texttt{Car})} \right]}
\end{align*}
$$

***Note: Changing the baseline category, e.g., switching from `Car` to `Public` won't change the estimated probability for each category for each individual.***

## Estimation

We summarise the key estimation steps in math below. We consider one observation at each population $i$.
More estimation details and generalization from one to $n_i$ observations for population $i$ can be found in Section 2.2 Parameter Estimation of [Czepiel, S. A. (2002).](https://czep.net/stat/mlelr.pdf). 

Let's follow the notations in @eq:multinomial-model and @eq:prob-multinomial. Define $y_{i,j} = 1$ if $i$th individual is observed as category $j$ and $y_{i,j} = 0$ otherwise. 
Define the kernel of the log likelihood function for multinomial logistic regression models is 
$$
\mathcal{L}(\beta \mid y) \simeq \prod_{i=1}^n \prod_{j=1}^m p_{i,j}^{y_{i,j}}.
$$

Replacing $y_{i,1}$ by $1-\sum_{j=2}^{m} y_{i,j}$ and rearranging the function gives
$$
\mathcal{L}(\beta \mid y) \simeq \prod_{i=1}^n \prod_{j=2}^{m} \left(\frac{p_{i,j}}{p_{i,1}} \right)^{y_{i,j}} p_{i,1}.
$$
Substituting the probabilities by the function of linear combination of regressors as in @eq:prob-multinomial and taking the logarithm of the likelihood function yields the log likelihood function, which is the loss function to minimize:
$$
\ell(\beta) = \sum_{i=1}^n \sum_{j=2}^m \left(y_{i,j}\sum_{k=0}^K X_{i,k} \beta_k^{(j,1)}\right) - \log\left(1+\sum_{j=2}^m e^{\sum_{k=0}^K X_{i,k}\beta_k^{(j,1)}}\right),
$$
where $X_{i,0}=1$ corresponds to the intercept in the linear model. We solve the minimization problem by the Newton-Raphson method as in Section 10.6 in the Poisson Regression Section.

### The Newton–Raphson Method

To carry out maximum-likelihood estimation, we iterate Newton–Raphson in a compact matrix form (multinomial iteratively reweighted least squares (IRLS)). 
Let $q=(K+1)(m-1)$ be the total number of parameters (an intercept and $K$ slopes for each of the $m-1$ non-baseline classes). Then denote

- $\beta \in \mathbb{R}^q$ (stack of all $\beta_k^{(j,1)}$),
- $S(\beta) \in \mathbb{R}^q$ (stacked Score function components),
- $H(\beta) \in \mathbb{R}^{q\times q}$ (stacked second derivatives).

For model (category) $j$ and regressor $k$, the component of the score function (first-order partial detivative) $S(\beta)$ is 
$$
\frac{\partial \ell(\beta)}{\partial \beta_k^{(j,1)}} = \sum_i^n y_{i,j}X_{i,k} - p_{i,j}X_{i,k}.
$$
For model $j$ and regressor $k$ on the row, the component for category $j'$ and regressor $k'$ on the column of the Hessian matrix (second-order partial derivative) $H(\beta)$ is 
$$
\frac{\partial \ell^2(\beta)}{\partial \beta_k^{(j,1)} \partial \beta_{k'}^{(j',1)}} = 
\begin{cases}
-\sum_{i=1^n} X_{i,k}p_{i,j}(1-p_{i,j})X_{i,k'} & \text{ if } j'=j, \\
\sum_{i=1}^n X_{i,k}p_{i,j}p_{i,j'}X_{i,k'} & \text{ if } j' \neq j,
\end{cases}
$$
where $j,j'=1,\dots,m$ and $k,k'=0,\dots,K$. 

The Newton-Raphson/IRLS iteration procedure is as follows.

1. Choose baseline class; initialize $\beta^{(0)}=\mathbf{0}$ (stacked over all $(j,k)$).
2. Repeat until convergence:
   - Compute the log-odds $\eta_i^{(j,1)}$, probability $p_{i,j}$ for all individual $i$ and category $j$ across all regressors $k=1,\dots,K$, and the log-likelihood function $\ell(\beta^{(t)})$.
   - Form the stacked score $S(\beta^{(t)})$ and Hessian functions $H(\beta^{(t)})$ using the expressions above.
   - Update $\beta^{(t+1)} \leftarrow \beta^{(t)}+\alpha (H(\beta^{(t)}))^{-1}S(\beta^{(t)})$, with step size $\alpha\in(0,1]$.
   - Check convergence criteria.

### Fitting the Model

Now, we fit the multinomial regression model using the `nnet` package in R or `statsmodel` package in Python. To handle the imbalanced groups, we resample the training sets to balance out the groups to eliminate the bias in modelling. 

::: {.panel-tabset}

## R Code

```{.r}
# Set the baseline model
train_set$transport_mode <- relevel(train_set$transport_mode, ref = "Car")

# Fit multinomial logistic regression
multinom_fit <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = train_set)

# Print the fitted model
print(summary(multinom_fit))
```

## Python Code

```{.python}
# Factorize the response and set the baseline model
train_set["transport_mode"] = pd.Categorical(
    train_set["transport_mode"],
    categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)

# Design matrix and response
y = train_set["transport_mode"]
X = train_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]

# Add intercept
X = sm.add_constant(X)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_numeric = X.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes

# Fit multinomial logistic regression
model = sm.MNLogit(y, X_numeric)
multinom_fit = model.fit()

# Print the fitted model
print(multinom_fit.summary())
```

:::

::: {.panel-tabset}

### R Output

```{r}
#| echo: false
#| message: false
#| warning: false
train_set$transport_mode <- relevel(train_set$transport_mode, ref = "Car")
multinom_fit <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = train_set)
#print(t(round(summary(multinom_fit)$coefficients, 3)))
print(summary(multinom_fit))
```

### Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
train_set["transport_mode"] = pd.Categorical(
    train_set["transport_mode"],
    categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)
y = train_set["transport_mode"]
X = train_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]
X = sm.add_constant(X)
X_numeric = X.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes
model = sm.MNLogit(y, X_numeric)
multinom_fit = model.fit()
print(multinom_fit.summary())
```

:::

In multinomial logistic regression, we model the probability of each outcome category given predictors. The key assumption is that we have enough data for each category to estimate the model parameters reliably.
If one category is much more frequent than the others (i.e., unbalanced outcomes), the model may:

1. Bias coefficients toward the majority class because the likelihood function is dominated by majority observations.
2. Underestimate uncertainty for rare categories (small sample size → large variance).
3. Yield poor predictive performance for minority classes (the model tends to predict the majority class).
4. In extreme cases, parameters for rare categories may not converge or may give infinite estimates (complete separation).

The intuition behind is if 95\% are "car", then the model has little incentive to adjust parameters for "walk" and "bus," so estimates for those categories can be unstable. There are two common approaches: 

1. Downsampling: Reduce the majority class size so that each class is roughly balanced. This risks throwing away data, but helps balance influence.
2. Upsampling: Duplicate (or synthetically generate) minority class cases until classes are balanced. This risks overfitting, but improves minority estimation.

Check the class counts of the outcome below.

::: {.panel-tabset}

## R Code

```{.r}
cat("Class counts before balancing:", "\n")
print(table(train_set$transport_mode))
```

## Python Code

```{.python}
print("Class counts before balancing:", "\n")
print(y.value_counts())
```

:::

::: {.panel-tabset}

## R Code

```{r}
#| echo: false
#| message: false
#| warning: false
cat("Class counts before balancing:", "\n")
print(table(train_set$transport_mode))
```

## Python Code

```{python}
#| echo: false
#| message: false
#| warning: false
print("Class counts before balancing:", "\n")
print(y.value_counts())
```

:::

The fitted model on upsampled training set is below.

::: {.panel-tabset}

## R Code

```{.r}
# Upsample the outcome classes
set.seed(352) # set a seed for sampling
train_set_balanced <- upSample(x = train_set[, c("age","commute_distance")],
                  y = train_set$transport_mode,
                  yname = "transport_mode")
# Fit the multinomial logit regression
multinom_fit_balanced <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = train_set_balanced)
print(summary(multinom_fit_balanced))
```

## Python Code

```{.python}
# Upsample the outcome classes
balanced_df = pd.DataFrame()
for label in train_set["transport_mode"].unique():
    subset = train_set[train_set["transport_mode"] == label]
    balanced_subset = resample(
        subset,
        replace=True,  # sample with replacement
        n_samples=train_set["transport_mode"].value_counts().max(),  # match majority class size
        random_state=352
    )
    balanced_df = pd.concat([balanced_df, balanced_subset])
X_bal = balanced_df[["age", "commute_distance", "has_car_available", "weekend_commuter"]]
y_bal = balanced_df["transport_mode"]
X_bal = sm.add_constant(X_bal)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_bal_num = X_bal.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_bal_num[col] = X_bal_num[col].cat.codes

# Fit the multinomial logit regression
model_balanced = sm.MNLogit(y_bal, X_bal_num)
multinom_fit_balanced = model_balanced.fit()
print(multinom_fit_balanced.summary())
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
set.seed(352)
train_set_balanced <- upSample(x = train_set[, c("age","commute_distance","has_car_available", "weekend_commuter")],
                  y = train_set$transport_mode,
                  yname = "transport_mode")
multinom_fit_balanced <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = train_set_balanced)
print(summary(multinom_fit_balanced))
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
balanced_df = pd.DataFrame()
for label in train_set["transport_mode"].unique():
    subset = train_set[train_set["transport_mode"] == label]
    balanced_subset = resample(
        subset,
        replace=True,
        n_samples=train_set["transport_mode"].value_counts().max(), 
        random_state=352
    )
    balanced_df = pd.concat([balanced_df, balanced_subset])
X_bal = balanced_df[["age", "commute_distance", "has_car_available", "weekend_commuter"]]
y_bal = balanced_df["transport_mode"]
X_bal = sm.add_constant(X_bal)

X_bal_num = X_bal.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_bal_num[col] = X_bal_num[col].cat.codes

model_balanced = sm.MNLogit(y_bal, X_bal_num)
multinom_fit_balanced = model_balanced.fit()
print(multinom_fit_balanced.summary())
```

:::

## Goodness of Fit

Let us assess how well the fitted multinomial logistic regression explains and predicts the observed transport modes. We consider the residual deviance (similar as @goodness in the Classical Poisson Regression chapter) and the classification performance (confusion matrix, accuracy, per-class precision, recall, F1 score). 

Let's visualize and summarize goodness of fit for the training set. 

::: {.panel-tabset}

## R code

```{.r}
y_pred <- predict(multinom_fit, newdata = train_set)
conf_matrix <- confusionMatrix(factor(y_pred, levels = levels(train_set$transport_mode)),
                              factor(train_set$transport_mode, levels = levels(train_set$transport_mode)))

cm_table <- as.table(conf_matrix$table)
cm_df <- as.data.frame(cm_table)
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

ggplot(cm_df, aes(x = Actual, y = Predicted, fill = Freq)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "white", high = "#0072B2") +
    geom_text(aes(label = Freq), color = "black", size = 4) +
    labs(title = "Confusion Matrix Heatmap", x = "Actual", y = "Predicted") +
    theme_bw()
```

## Python Code

```{.python}
y_pred = multinom_fit.predict(X_numeric).idxmax(axis=1)

# Compute confusion matrix
cm = confusion_matrix(y.cat.codes, y_pred)

disp = ConfusionMatrixDisplay(confusion_matrix=cm, display_labels=y.cat.categories)
disp.plot(cmap='Blues')
plt.title("Confusion Matrix Heatmap")
plt.xlabel("Actual")
plt.ylabel("Predicted")

plt.show()
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
y_pred <- predict(multinom_fit, newdata = train_set)
conf_matrix <- confusionMatrix(factor(y_pred, levels = levels(train_set$transport_mode)),
                              factor(train_set$transport_mode, levels = levels(train_set$transport_mode)))

cm_table <- as.table(conf_matrix$table)
cm_df <- as.data.frame(cm_table)
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

ggplot(cm_df, aes(x = Actual, y = Predicted, fill = Freq)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "white", high = "#0072B2") +
    geom_text(aes(label = Freq), color = "black", size = 4) +
    labs(title = "Confusion Matrix Heatmap", x = "Actual", y = "Predicted") +
    theme_bw()
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
y_pred = multinom_fit.predict(X_numeric).idxmax(axis=1)

# Compute confusion matrix
cm = confusion_matrix(y.cat.codes, y_pred)

disp = ConfusionMatrixDisplay(confusion_matrix=cm, display_labels=y.cat.categories)
disp.plot(cmap='Blues')
plt.title("Confusion Matrix Heatmap")
plt.xlabel("Actual")
plt.ylabel("Predicted")

plt.show()
```

:::

You're encouraged to repeat these diagnostics on the balanced (upsampled) model and on the held‑out test set to compare whether balancing improves minority class recall without overly harming overall calibration.

Let's compute the metrics.

::: {.panel-tabset}

## R Code

```{.r}
# Classification predictions & confusion matrix
y_pred <- predict(multinom_fit, newdata = train_set)
conf_matrix <- confusionMatrix(factor(y_pred, levels = levels(train_set$transport_mode)),
                               factor(train_set$transport_mode, levels = levels(train_set$transport_mode)))
overall_accuracy <- conf_matrix$overall['Accuracy']
macro_recall <- mean(conf_matrix$byClass[,"Recall"], na.rm = TRUE)
macro_f1 <- mean(conf_matrix$byClass[,"F1"], na.rm = TRUE)
cat("Overall Accuracy:", round(overall_accuracy, 3), "\n")
cat("Macro Recall:", round(macro_recall, 3), "\n")
cat("Macro F1:", round(macro_f1, 3), "\n")
```

## Python Code

```{.python}
# Compute classification predictions & confusion matrix
probs = multinom_fit.predict(X_numeric)  # DataFrame of class probabilities
y_pred = probs.idxmax(axis=1)

# Compute the metrics
acc = accuracy_score(y.cat.codes, y_pred)
recall = recall_score(y.cat.codes, y_pred, average='macro')
f1 = f1_score(y.cat.codes, y_pred, average='macro')
print(f"Overall Accuracy: {acc:.3f}")
print(f"Macro Recall: {recall:.3f}")
print(f"Macro F1-score: {f1:.3f}")
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
y_pred <- predict(multinom_fit, newdata = train_set)
conf_matrix <- confusionMatrix(factor(y_pred, levels = levels(train_set$transport_mode)),
                               factor(train_set$transport_mode, levels = levels(train_set$transport_mode)))
overall_accuracy <- conf_matrix$overall['Accuracy']
macro_recall <- mean(conf_matrix$byClass[,"Recall"], na.rm = TRUE)
macro_f1 <- mean(conf_matrix$byClass[,"F1"], na.rm = TRUE)
cat("Overall Accuracy:", round(overall_accuracy, 3), "\n")
cat("Macro Recall:", round(macro_recall, 3), "\n")
cat("Macro F1:", round(macro_f1, 3), "\n")
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
probs = multinom_fit.predict(X_numeric)
y_pred = probs.idxmax(axis=1)
acc = accuracy_score(y.cat.codes, y_pred)
recall = recall_score(y.cat.codes, y_pred, average='macro')
f1 = f1_score(y.cat.codes, y_pred, average='macro')
print(f"Overall Accuracy: {acc:.3f}")
print(f"Macro Recall: {recall:.3f}")
print(f"Macro F1-score: {f1:.3f}")
```

:::

About 69.6\% of all individual transportation modes were classified correctly. On average, the model correctly identifies 34.4\% of instances for each transport mode. Recall is relatively low, which suggests the model is biased to the majority classes, even though accuracy is decent. 
Macro F1 score is 0.682, meaning the model has a moderate balance of precision and recall on average per class. Macro F1 is much higher than macro recall — this can happen if precision is high for the majority class, but recall for minority classes is low.
In summary, the model performs okay overall but is biased toward the dominant class(es). It often fails to detect less frequent transport modes, which is why macro recall is low. Accuracy alone would overestimate the model’s effectiveness across all classes.
You are encouraged to explore the performance of the fitted model on the balanced sets. 

## Inference

After we finish estimating the coefficients, we look at how uncertain they are and whether they differ from zero. We get a variance estimate for each coefficient from the observed information matrix (the negative of the Hessian at the solution). Taking square roots of those variances gives standard errors. 
Similar as in Poisson regression, each coefficient divided by its standard error gives a (asymptotically normal) **Wald Z statistic** to test if the true effect is zero. Using the same standard errors we build confidence intervals: estimate $\pm$ (critical value $\times$ standard error). To make results easier to interpret, we often exponentiate a coefficient (or its confidence limits) to get an odds ratio—how the odds of choosing a category change for a one‑unit increase in the predictor, holding other variables fixed.

We can determine whether a regressor is statistically associated with the logarithm of the odds through hypothesis testing for the parameters $\beta^{(u,v)}_j$ by link function. Define the Wald statistic $z^{(u,v)}_j$ as:
$$
z_j^{(u, v)} = \frac{\hat{\beta}_j^{(u, v)}}{\mbox{se}\left(\hat{\beta}_j^{(u, v)}\right)}
$$
and we test the hypotheses: 
$$
\begin{gather*}
H_0: \beta_j^{(u, v)} = 0\\
H_a: \beta_j^{(u, v)} \neq 0.
\end{gather*}
$$
Provided the sample size $n$ is large enough, $z_j$ has an approximately Standard Normal distribution under $H_0$.

The corresponding $p$-values for each $\beta^{(u,v)}_j$ can be computed. The smaller the $p$-value, the stronger the evidence against the null hypothesis $H_0$. As in the previous regression models, we would set a predetermined significance level  (usually taken to be 0.05) to infer if the $p$-value is small enough. If the $p$-value is smaller than the predetermined level $\alpha$, then you could claim that there is evidence to reject the null hypothesis. Hence, $p$-values that are small enough indicate that the data provides evidence in favour of association (or causation in the case of an experimental study!) between the response variable and the $j$th regressor.

Given a specified level of confidence where $\alpha$ is the significance level, we can construct approximate $(1-\alpha)\times 100\%$ confidence intervals for the corresponding true value of $\beta^{(u,v)}_j$:
$$
\begin{equation*}
\hat{\beta}_j^{(u, v)} \pm z_{\alpha/2}\mbox{se} \left( \hat{\beta}_j^{(u, v)} \right),
\end{equation*}
$$
where $z_{\alpha/2}$ is the upper $\alpha/2$ quantile of the Standard Normal distribution.

We compute the 95\% confidence intervals below and filter the signicant coefficients under the 5\% significance level.

::: {.panel-tabset}

## R Code

```{.r}
broom::tidy(multinom_fit, conf.int = TRUE, exponentiate = TRUE) |>
  dplyr::mutate_if(is.numeric, round, 3) |>
  dplyr::filter(p.value < 0.05) |>
  rename(outcome_class = y.level)
```

## Python Code

```{.python}
data_dict = {
    "estimate": np.exp(multinom_fit.params),      # odds ratios
    "pvalue": multinom_fit.pvalues,
    "std_err": multinom_fit.bse,
    "statistic": multinom_fit.tvalues
}

# Outcome class mapping
class_map = {0: "Bicycle", 1: "Public_Transit", 2: "Walking"}

# Container for processed tidy tables
tidy_dict = {}

for name, df in data_dict.items():
    tidy = (
        df.stack()               # long form
        .swaplevel()             # variable first, outcome second
        .reset_index()           # convert MultiIndex to columns
        .rename(columns={"level_0": "outcome_class", 
                         "level_1": "variable", 
                         0: name})
    )
    tidy["outcome_class"] = tidy["outcome_class"].replace(class_map)
    tidy = tidy.sort_values(by="outcome_class").reset_index(drop=True)
    
    tidy_dict[name] = tidy

# CIs
conf = multinom_fit.conf_int()
or_conf = np.exp(conf)
ci_lower_stacked = or_conf["lower"].reset_index()
ci_lower_stacked.columns = ['outcome_class', 'term', 'ci_lower']
ci_upper_stacked = or_conf["upper"].reset_index()
ci_upper_stacked.columns = ['outcome_class', 'term', 'ci_upper']

# Merge the result tables
ref_cols = ["outcome_class", "term"]
results = tidy_dict['estimate'] \
    .merge(tidy_dict['pvalue'], on=ref_cols) \
    .merge(tidy_dict['std_err'], on=ref_cols) \
    .merge(tidy_dict['statistic'], on=ref_cols) \
    .merge(ci_lower_stacked, on=ref_cols) \
    .merge(ci_upper_stacked, on=ref_cols)

# Filter the rows with significant p-values
results_sig = results[results["pvalue"] < 0.05].round(3)
print(results_sig)
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
broom::tidy(multinom_fit, conf.int = TRUE, exponentiate = TRUE) |>
  mutate_if(is.numeric, round, 3) |>
  filter(p.value < 0.05) |>
  rename(outcome_class = y.level)
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
data_dict = {
    "estimate": np.exp(multinom_fit.params),      # odds ratios
    "pvalue": multinom_fit.pvalues,
    "std_err": multinom_fit.bse,
    "statistic": multinom_fit.tvalues
}
class_map = {0: "Bicycle", 1: "Public_Transit", 2: "Walking"}
tidy_dict = {}
for name, df in data_dict.items():
    tidy = (
        df.stack()
        .swaplevel()
        .reset_index()
        .rename(columns={"level_0": "outcome_class", 
                         "level_1": "term", 
                         0: name})
    )
    tidy["outcome_class"] = tidy["outcome_class"].replace(class_map)
    tidy = tidy.sort_values(by="outcome_class").reset_index(drop=True)
    tidy_dict[name] = tidy

# CIs
conf = multinom_fit.conf_int()
or_conf = np.exp(conf)
ci_lower_stacked = or_conf["lower"].reset_index()
ci_lower_stacked.columns = ['outcome_class', 'term', 'ci_lower']
ci_upper_stacked = or_conf["upper"].reset_index()
ci_upper_stacked.columns = ['outcome_class', 'term', 'ci_upper']

ref_cols = ["outcome_class", "term"]
results = tidy_dict['estimate'] \
    .merge(tidy_dict['pvalue'], on=ref_cols) \
    .merge(tidy_dict['std_err'], on=ref_cols) \
    .merge(tidy_dict['statistic'], on=ref_cols) \
    .merge(ci_lower_stacked, on=ref_cols) \
    .merge(ci_upper_stacked, on=ref_cols)
results_sig = results[results["pvalue"] < 0.05].round(3)
print(results_sig)
```

:::

Whether people choose driving or taking public transits varys significantly on average. The commute distance and the availability of cars both have significant effects on the choices of transportation mode.

## Results

The predicted model can be written as 
$$
\log \big[ \frac{\Pr(Y_i = k)}{\Pr(Y_i = \text{Car})} \big] = \hat{\beta}_{0k} + \hat{\beta}_{1k} X_{1i} + \hat{\beta}_{2k} X_{2i} + \hat{\beta}_{3k} X_{3i} + \hat{\beta}_{4k} X_{4i},
$$
where $k\in\{$ Bicycle, Public Transit, Walking $\}$ and $X_1,X_2,X_3,X_4$ represent age, commute distance, car availability, and weekend commuter indicator respectively.

### Prediction 

We predict the model on test set and visualize the prediction results in the confusion matrix below. 

::: {.panel-tabset}

## R Code

```{.r}
# Predict on the test set using the fitted model
y_pred <- predict(multinom_fit, newdata = test_set)

# Compute confusion matrix
conf_matrix <- confusionMatrix(factor(y_pred, levels = levels(test_set$transport_mode)),
                              factor(test_set$transport_mode, levels = levels(test_set$transport_mode)))

# Visualize the confusion matrix
cm_table <- as.table(conf_matrix$table)
cm_df <- as.data.frame(cm_table)
colnames(cm_df) <- c("Predicted", "Actual", "Freq")
cm_df %>%
  ggplot(aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 4) +
  scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  labs(title = "Confusion Matrix (Test Set)", x = "Predicted", y = "Actual") +
  theme_bw()
```

## Python Code

```{.python}
# Predict on the test set using the fitted model
test_set["transport_mode"] = pd.Categorical(
  test_set["transport_mode"],  # align with earlier label spelling
  categories=["Car", "Bicycle", "Public Transit", "Walking"],
    ordered=False
)
y_test = test_set["transport_mode"]
X_test = test_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]
X_test = sm.add_constant(X_test)
X_test_num = X_test.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_test_num[col] = X_test_num[col].cat.codes
y_pred_test = multinom_fit.predict(X_test_num).idxmax(axis=1)

# Compute the confusion matrix
cm = confusion_matrix(y_test.cat.codes, y_pred_test)

# Print confusion matrix
labels = y_test.cat.categories
cm = confusion_matrix(y_test.cat.codes, y_pred_test, labels=range(len(labels)))
disp = ConfusionMatrixDisplay(confusion_matrix=cm, display_labels=y_test.cat.categories)
disp.plot(cmap='Blues')
plt.title("Confusion Matrix")
plt.xlabel("Actual")
plt.ylabel("Predicted")
plt.show()
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
y_pred <- predict(multinom_fit, newdata = test_set)
conf_matrix <- confusionMatrix(factor(y_pred, levels = levels(test_set$transport_mode)),
                              factor(test_set$transport_mode, levels = levels(test_set$transport_mode)))

cm_table <- as.table(conf_matrix$table)
cm_df <- as.data.frame(cm_table)
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

cm_df %>%
  ggplot(aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 4) +
  scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  labs(title = "Confusion Matrix (Test Set)", x = "Predicted", y = "Actual") +
  theme_bw()
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
test_set["transport_mode"] = pd.Categorical(
  test_set["transport_mode"], 
  categories=["Car", "Bicycle", "Public Transit", "Walking"],
    ordered=False
)
y_test = test_set["transport_mode"]
X_test = test_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]
X_test = sm.add_constant(X_test)
X_test_num = X_test.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_test_num[col] = X_test_num[col].cat.codes
y_pred_test = multinom_fit.predict(X_test_num).idxmax(axis=1)
cm = confusion_matrix(y_test.cat.codes, y_pred_test)
labels = y_test.cat.categories
cm = confusion_matrix(y_test.cat.codes, y_pred_test, labels=range(len(labels)))

import matplotlib.pyplot as plt
disp = ConfusionMatrixDisplay(confusion_matrix=cm, display_labels=y_test.cat.categories)
disp.plot(cmap='Blues')
plt.title("Confusion Matrix")
plt.xlabel("Actual")
plt.ylabel("Predicted")
plt.show()
```

:::

The corresponding metrics are computed below.

::: {.panel-tabset}

## R Code

```{.r}
# Compute the metrics
overall_accuracy <- conf_matrix$overall['Accuracy']
macro_recall <- mean(conf_matrix$byClass[,"Recall"], na.rm = TRUE)
macro_f1 <- mean(conf_matrix$byClass[,"F1"], na.rm = TRUE)
cat("Overall Accuracy:", round(overall_accuracy, 3), "\n")
cat("Macro Recall:", round(macro_recall, 3), "\n")
cat("Macro F1:", round(macro_f1, 3), "\n")
```

## Python Code

```{.python}
# Compute the metrics
acc = accuracy_score(y_test.cat.codes, y_pred_test)
recall = recall_score(y_test.cat.codes, y_pred_test, average='macro')
f1 = f1_score(y_test.cat.codes, y_pred_test, average='macro')
print(f"Overall Accuracy: {acc:.3f}")
print(f"Macro Recall: {recall:.3f}")
print(f"Macro F1-score: {f1:.3f}")
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
overall_accuracy <- conf_matrix$overall['Accuracy']
macro_recall <- mean(conf_matrix$byClass[,"Recall"], na.rm = TRUE)
macro_f1 <- mean(conf_matrix$byClass[,"F1"], na.rm = TRUE)
cat("Overall Accuracy:", round(overall_accuracy, 3), "\n")
cat("Macro Recall:", round(macro_recall, 3), "\n")
cat("Macro F1:", round(macro_f1, 3), "\n")
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
acc = accuracy_score(y_test.cat.codes, y_pred_test)
recall = recall_score(y_test.cat.codes, y_pred_test, average='macro')
f1 = f1_score(y_test.cat.codes, y_pred_test, average='macro')
print(f"Overall Accuracy: {acc:.3f}")
print(f"Macro Recall: {recall:.3f}")
print(f"Macro F1-score: {f1:.3f}")
```

:::

On the test set, the model’s overall accuracy is slightly lower than training. The higher macro recall suggests the model is now better at detecting less frequent transport modes, though overall prediction quality per class (macro F1) is still moderate. This indicates the model generalizes reasonably well, but there is room to improve detection of all classes evenly.


## Storytelling

Commuting behaviors and demographic factors are both associated with transportation mode choice and useful for predicting the choice between choosing cars or public transportation. In general, prediction is stronger for more frequent modes. 
This inferential question provides a comprehensive understanding of how individual characteristics shape transportation decisions and the predictive question can guide planning or personalized recommendations. You may choose one perspective of either inferential or predictive depending on your research goal. 

## Practice Problems

In this final section, you can test your understanding with the following conceptual interactive exercises, and then proceed to work through an inferential analysis of a simple dataset.

::: {.quiz}

### Conceptual Questions

::: {.Question}
::::{.Question-header}
Question 14.1
::::
::::{.Question-container}
**Multiple Choice**

Which response variable would be most appropriate for a Multinomial Logistic regression model? Select the correct option:

**A.** The number of doctor visits a patient has in one year.

**B.** A student's exam score out of 100.

**C.** A person's primary transport mode, with categories `Car`, `Bicycle`, `Public_Transit`, and `Walking`.

**D.** Whether a person owns a car, coded as `Yes` or `No`.
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 14.1
::::
::::{.Answer-container}
<details>

<summary><strong>Click here to reveal the answer!</strong></summary>

**Correct answer:** C.

**Rationale:**

Multinomial Logistic regression is used for a **nominal categorical response** with more than two mutually exclusive categories. A count response would motivate a count regression model, a continuous score would motivate a continuous-outcome model, and a binary response would motivate binary logistic regression.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 14.2
::::
::::{.Question-container}
**Multiple Choice**

Why is Multinomial Logistic regression more appropriate than Ordinal Logistic regression for the transportation example in this chapter? Select the correct option:

**A.** `transport_mode` has categories that are naturally ordered from low to high.

**B.** `transport_mode` has unordered categories, so there is no single ranking such as `Car` $<$ `Bicycle` $<$ `Public_Transit` $<$ `Walking`.

**C.** `transport_mode` is a continuous variable.

**D.** Ordinal Logistic regression can only be used with binary outcomes.
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 14.2
::::
::::{.Answer-container}
<details>

<summary><strong>Click here to reveal the answer!</strong></summary>

**Correct answer:** B.

**Rationale:**

The categories in `transport_mode` are **nominal**, not ordinal. Multinomial Logistic regression is designed for unordered response categories, whereas Ordinal Logistic regression is used when the response levels have a meaningful order.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 14.3
::::
::::{.Question-container}
**Open-ended Question**

**In two to four sentences**, explain why an ordinary linear regression model would not be appropriate for predicting `transport_mode` from `commute_distance`, `age`, `has_car_available`, and `weekend_commuter`.
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 14.3
::::
::::{.Answer-container}
<details>

<summary><strong>Click here to reveal the answer!</strong></summary>

**Answer:**

A correct answer would note that `transport_mode` is not continuous, so a linear model's fitted values would not naturally represent valid category probabilities. Ordinary linear regression can also produce predictions outside the probability scale and does not enforce that the predicted probabilities across transport modes sum to 1. Multinomial Logistic regression handles this by modelling category probabilities through baseline-category logits and the Softmax form.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 14.4
::::
::::{.Question-container}
**Short Answer**

If a Multinomial Logistic regression model has $m$ response categories and one category is chosen as the baseline, how many logit link functions must be estimated? In the chapter's transportation example with four categories and `Car` as the baseline, how many link functions are there?
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 14.4
::::
::::{.Answer-container}
<details>

<summary><strong>Click here to reveal the answer!</strong></summary>

**Rationale:**

The model estimates $m - 1$ logit link functions, one for each non-baseline category compared to the baseline. With four transport categories and `Car` as the baseline, there are three link functions: `Public_Transit` versus `Car`, `Bicycle` versus `Car`, and `Walking` versus `Car`.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 14.5
::::
::::{.Question-container}
**Multiple Choice**

In the chapter's notation, what does

$$
\eta_i^{(\texttt{Public},\texttt{Car})}
= \log\left[
\frac{P(Y_i = \texttt{Public} \mid \mathbf{x}_i)}
{P(Y_i = \texttt{Car} \mid \mathbf{x}_i)}
\right]
$$

represent? Select the correct option:

**A.** The probability that individual $i$ chooses public transit.

**B.** The log odds of choosing `Public_Transit` relative to choosing `Car`, conditional on the predictors.

**C.** The difference between the probability of choosing `Public_Transit` and the probability of choosing `Car`.

**D.** The residual for individual $i$.
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 14.5
::::
::::{.Answer-container}
<details>

<summary><strong>Click here to reveal the answer!</strong></summary>

**Correct answer:** B.

**Rationale:**

Each linear predictor in Multinomial Logistic regression is a **log odds comparison** between one non-baseline category and the baseline category. It is not itself a probability, a probability difference, or a residual.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 14.6
::::
::::{.Question-container}
**Short Answer**

Why does the chapter use the Softmax form to convert the logit-linear functions into category probabilities? In practice, what problem would occur if we tried to interpret the separate logit-linear scores as probabilities without this final conversion?
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 14.6
::::
::::{.Answer-container}
<details>

<summary><strong>Click here to reveal the answer!</strong></summary>

**Rationale:**

The Softmax form converts the $m - 1$ baseline-category logits into probabilities for all $m$ categories. In practical terms, the model first produces one relative score for each non-baseline category; Softmax then turns those scores into a complete probability distribution. This matters because a person cannot have separate category probabilities that add up to more than 1 or less than 1 across mutually exclusive choices such as `Car`, `Bicycle`, `Public_Transit`, and `Walking`. Softmax guarantees that each probability is between 0 and 1 and that the probabilities across all categories sum to 1 for each observation. This is also why Softmax appears in many multi-class machine learning models, including the output layer of some [neural networks](https://en.wikipedia.org/wiki/Softmax_function#Applications).

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 14.7
::::
::::{.Question-container}
**Short Answer**

The chapter notes that changing the baseline category, for example from `Car` to `Public_Transit`, does not change the estimated probability for each category. If the probabilities do not change, how does changing the baseline affect the coefficients and their interpretation?
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 14.7
::::
::::{.Answer-container}
<details>

<summary><strong>Click here to reveal the answer!</strong></summary>

**Rationale:**

Changing the baseline changes the **parameterization** of the model: the intercepts, coefficients, and log-odds comparisons are rewritten relative to a different reference category. If `Car` is the baseline, a coefficient describes a change in the log odds of some other category versus `Car`. If `Public_Transit` becomes the baseline, the same model is now described through comparisons versus `Public_Transit`. The fitted category probabilities remain the same, but the wording and numerical values of the coefficients change because the reference comparison has changed.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 14.8
::::
::::{.Question-container}
**Short Answer**

Suppose the coefficient of `commute_distance` in the `Public_Transit` versus `Car` link function is positive. Interpret this coefficient in words, using the phrase "*holding other variables fixed*".
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 14.8
::::
::::{.Answer-container}
<details>

<summary><strong>Click here to reveal the answer!</strong></summary>

**Answer:**

A positive coefficient means that longer commute distance is associated with higher log odds of choosing `Public_Transit` rather than `Car`, **holding other variables fixed**. Equivalently, for a one-unit increase in commute distance, the odds of `Public_Transit` relative to `Car` increase by a multiplicative factor of $\exp(\hat{\beta})$.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 14.9
::::
::::{.Question-container}
**Multiple Choice**

In Multinomial Logistic regression, why do we often exponentiate a coefficient $\hat{\beta}$? Select the correct option:

**A.** To turn a log-odds coefficient into an odds ratio.

**B.** To turn a category probability into a residual.

**C.** To make the response variable continuous.

**D.** To remove the need for a baseline category.
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 14.9
::::
::::{.Answer-container}
<details>

<summary><strong>Click here to reveal the answer!</strong></summary>

**Correct answer:** A.

**Rationale:**

The raw coefficient is on the **log-odds** scale. Exponentiating it gives an **odds ratio**, which describes the multiplicative change in the odds of a non-baseline category relative to the baseline for a one-unit increase in the predictor, holding other variables fixed.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 14.10
::::
::::{.Question-container}
**Short Answer**

In the transportation model, both `Bicycle` and `Walking` are compared to the baseline category `Car`. How could you compare `Walking` directly to `Bicycle` using the fitted model?
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 14.10
::::
::::{.Answer-container}
<details>

<summary><strong>Click here to reveal the answer!</strong></summary>

**Rationale:**

You can subtract the two baseline-category logit equations. If `Car` is the baseline, then

$$
\log\left[
\frac{P(Y_i = \texttt{Walking} \mid \mathbf{x}_i)}
{P(Y_i = \texttt{Car} \mid \mathbf{x}_i)}
\right]
= \eta_i^{(\texttt{Walking}, \texttt{Car})}
$$

and

$$
\log\left[
\frac{P(Y_i = \texttt{Bicycle} \mid \mathbf{x}_i)}
{P(Y_i = \texttt{Car} \mid \mathbf{x}_i)}
\right]
= \eta_i^{(\texttt{Bicycle}, \texttt{Car})}.
$$

Therefore,

$$
\log\left[
\frac{P(Y_i = \texttt{Walking} \mid \mathbf{x}_i)}
{P(Y_i = \texttt{Bicycle} \mid \mathbf{x}_i)}
\right]
=
\eta_i^{(\texttt{Walking}, \texttt{Car})}
-
\eta_i^{(\texttt{Bicycle}, \texttt{Car})}.
$$

This also means subtracting the corresponding coefficient vectors.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 14.11
::::
::::{.Question-container}
**Multiple Choice**

Which statement best describes the **independence** assumption in the chapter's transportation example? Select the correct option:

**A.** Every person must have exactly the same predicted probability for each transport mode.

**B.** Each observation should represent an independently sampled individual's transport choice, with response categories that are mutually exclusive and exhaustive.

**C.** The predictors must be independent of the response variable.

**D.** The model must have independent coefficients for every category and every predictor.
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 14.11
::::
::::{.Answer-container}
<details>

<summary><strong>Click here to reveal the answer!</strong></summary>

**Correct answer:** B.

**Rationale:**

The independence assumption concerns the sampled observations and the structure of the response categories. Each person should contribute one independent observation, and each response should fall into exactly one of the available categories.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 14.12
::::
::::{.Question-container}
**Short Answer**

Why do we factorize variables such as `transport_mode`, `has_car_available`, and `weekend_commuter` before fitting the model in R or Python?
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 14.12
::::
::::{.Answer-container}
<details>

<summary><strong>Click here to reveal the answer!</strong></summary>

**Rationale:**

Factorizing these variables tells the software to treat them as **categorical** rather than continuous numeric variables. This is especially important for `transport_mode`, which defines the response categories, and for binary predictors such as `has_car_available` and `weekend_commuter`, which should be represented through category indicators or codes rather than interpreted as arbitrary continuous measurements. In coding terms, factors in `R` determine the available levels and the reference category used in the model matrix. In `Python`, converting to categorical variables or creating dummy variables with `pd.get_dummies()` serves the same purpose: it makes the category labels explicit and controls which indicator columns are included in the model. Without this step, software may treat categories as raw numbers or may choose an unintended baseline.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 14.13
::::
::::{.Question-container}
**Multiple Choice**

How are the coefficients in a Multinomial Logistic regression model estimated in this chapter? Select the correct option:

**A.** By minimizing the sum of squared residuals, exactly as in OLS.

**B.** By maximum likelihood estimation, using an iterative procedure such as Newton-Raphson or IRLS.

**C.** By choosing the coefficients that make every predicted class correct.

**D.** By computing the sample mean of each predictor within each class.
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 14.13
::::
::::{.Answer-container}
<details>

<summary><strong>Click here to reveal the answer!</strong></summary>

**Correct answer:** B.

**Rationale:**

Multinomial Logistic regression uses **maximum likelihood estimation**. Since the likelihood equations do not usually have a simple closed-form solution, the chapter describes iterative estimation through Newton-Raphson or IRLS.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 14.14
::::
::::{.Question-container}
**Short Answer**

The chapter discusses imbalanced outcome classes. Why can high overall accuracy be misleading when some transport modes are much less frequent than others?
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 14.14
::::
::::{.Answer-container}
<details>

<summary><strong>Click here to reveal the answer!</strong></summary>

**Rationale:**

If one or two classes dominate the dataset, a model can achieve decent overall accuracy by mostly predicting the majority classes. This can hide poor performance for minority classes, which is why the chapter also examines per-class metrics such as recall and macro F1.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 14.15
::::
::::{.Question-container}
**Short Answer**

What is the difference between downsampling and upsampling as strategies for dealing with imbalanced response categories?
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 14.15
::::
::::{.Answer-container}
<details>

<summary><strong>Click here to reveal the answer!</strong></summary>

**Rationale:**

Downsampling reduces the number of observations in the majority class so the classes are more balanced, but it can throw away useful data. Upsampling increases the number of minority-class observations, often by resampling with replacement, but it can increase the risk of overfitting to repeated minority cases. In practice, this choice depends on the goal of the analysis. If a transit planner mostly cares about overall accuracy, keeping the observed class distribution may be reasonable. If the planner also needs to identify less common choices such as walking or bicycling, balancing methods may help the model pay more attention to those smaller groups, even if overall accuracy decreases.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 14.16
::::
::::{.Question-container}
**Short Answer**

In the chapter's training-set evaluation, accuracy is around 69.6% but macro recall is much lower. What does this pattern suggest about the model's classification performance?
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 14.16
::::
::::{.Answer-container}
<details>

<summary><strong>Click here to reveal the answer!</strong></summary>

**Rationale:**

This suggests the model performs reasonably well overall but does not identify all classes equally well. In particular, the lower macro recall indicates weak detection of less frequent transport modes, so accuracy alone overstates how useful the classifier is across all categories.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 14.17
::::
::::{.Question-container}
**Multiple Choice**

For inference, the chapter uses the Wald statistic

$$
z_j^{(u, v)} = \frac{\hat{\beta}_j^{(u, v)}}{\mbox{se}\left(\hat{\beta}_j^{(u, v)}\right)}.
$$

What null hypothesis is being tested for a particular coefficient? Select the correct option:

**A.** $H_0: \beta_j^{(u, v)} = 0$.

**B.** $H_0: p_{i,j} = 1$.

**C.** $H_0: \hat{y}_i = y_i$ for every observation.

**D.** $H_0: \mbox{accuracy} = 1$.
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 14.17
::::
::::{.Answer-container}
<details>

<summary><strong>Click here to reveal the answer!</strong></summary>

**Correct answer:** A.

**Rationale:**

The Wald test asks whether a particular true coefficient is zero in a particular logit comparison. A small p-value provides evidence that the predictor is associated with the log odds for that category comparison, given the model.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 14.18
::::
::::{.Question-container}
**Short Answer**

Suppose an exponentiated 95% confidence interval for an odds ratio is $(0.82, 1.24)$ for the `Bicycle` versus `Car` comparison, where `Car` is the baseline category. At the 5% significance level, would this provide strong evidence that the predictor changes the odds of choosing `Bicycle` rather than `Car`? Why or why not?
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 14.18
::::
::::{.Answer-container}
<details>

<summary><strong>Click here to reveal the answer!</strong></summary>

**Rationale:**

No. Because the confidence interval includes 1, the data are compatible with no multiplicative change in the odds for that category comparison at the 5% significance level. In this context, the interval says that the predictor could plausibly decrease the odds of choosing the non-baseline category `Bicycle` rather than the baseline category `Car`, increase those odds, or have essentially no effect. This would usually correspond to a p-value greater than 0.05 for the associated coefficient.

</details>
::::
:::

::: {.Question}
::::{.Question-header}
Question 14.19
::::
::::{.Question-container}
**Short Answer**

The chapter distinguishes an inferential objective from a predictive objective. Give one example of each in a setting different from the transportation case study.
::::
:::

::: {.Answer}
::::{.Answer-header}
Answer 14.19
::::
::::{.Answer-container}
<details>

<summary><strong>Click here to reveal the answer!</strong></summary>

**Rationale:**

For a university dining study, an inferential objective could be to estimate how `weekly_food_budget` or `has_meal_plan` is associated with the odds of choosing `Food_Delivery`, `Packed_Lunch`, or `Fast_Food` relative to `Cafeteria`. A predictive objective could be to use a student's budget, schedule, distance from home, meal-plan status, and dietary restriction status to classify their most likely lunch choice, then evaluate the classifier with a confusion matrix, accuracy, macro recall, and macro F1.

</details>
::::
:::

### Coding Question

### Problem 1 {.unnumbered}

In this problem, you will investigate a dataset of surveyed students and their usual lunch choices on campus.
Each student was asked about their weekly food budget, the amount of time they usually have between classes, their distance from home, whether they have a campus meal plan, whether they have a dietary restriction, and which lunch option they usually choose. These variables are stored as `weekly_food_budget`, `time_between_classes`, `distance_from_home`, `has_meal_plan`, `dietary_restriction`, and `lunch_choice`, respectively. Below is a full breakdown of the variables:

| **Variable Name** | **Description** |
|-------------------|-----------------|
| **weekly_food_budget** | The student's approximate weekly food budget in dollars. Float (e.g. 62.50, 118.20, etc.). |
| **time_between_classes** | The number of minutes the student usually has between classes around lunch time. Float (e.g. 15.5, 84.0, etc.). |
| **distance_from_home** | The student's distance from home to campus in kilometers. Float (e.g. 2.4, 18.7, etc.). |
| **has_meal_plan** | Whether the student has a campus meal plan. Categorical: `Yes` or `No`. |
| **dietary_restriction** | Whether the student has a dietary restriction. Categorical: `Yes` or `No`. |
| **lunch_choice** | The student's usual lunch choice. Categorical: `Cafeteria`, `Packed_Lunch`, `Food_Delivery`, or `Fast_Food`. |

A campus dining office wants to understand how student schedules, budgets, and constraints shape lunch choices. They are also interested in whether these variables can be used to predict a student's most likely lunch option.

To answer this, the questions below will guide you through a Multinomial Logistic regression analysis on this data. Specifically, by the end of these questions, you will be able to answer the following concrete questions:
1. How are `weekly_food_budget`, `has_meal_plan`, and `dietary_restriction` associated with lunch choice?
2. Can we predict whether a student is most likely to choose `Cafeteria`, `Packed_Lunch`, `Food_Delivery`, or `Fast_Food`?
3. How should we think about prediction quality when one lunch category is much more common than the others?

### A. Data Collection and Wrangling {.unnumbered}

1. Read in the data from `problem-1-data.csv`, checking for any missing entries.
2. Convert the categorical variables to factors in `R` or categorical variables in `Python`.
3. Drop rows with missing entries, then split the dataset into training and test sets.
4. Check the training-set class counts for `lunch_choice`. Is the response balanced?

<details>
<summary><strong>Click to reveal answer</strong></summary>

::: {.panel-tabset}
## **`R` Code**
```{r}
library(readr)
library(tidyverse)
library(nnet)
library(caret)
library(broom)

df <- read_csv("./14-multinomial-logistic_files/simulations/problem-1-data.csv")

print(colSums(is.na(df)))

df <- df %>%
  drop_na() %>%
  mutate(
    lunch_choice = factor(lunch_choice,
                          levels = c("Cafeteria", "Packed_Lunch",
                                     "Food_Delivery", "Fast_Food")),
    has_meal_plan = factor(has_meal_plan, levels = c("No", "Yes")),
    dietary_restriction = factor(dietary_restriction, levels = c("No", "Yes"))
  )

set.seed(1411)
train_id <- sample(seq_len(nrow(df)), size = floor(0.8 * nrow(df)))
train_set <- df[train_id, ]
test_set <- df[-train_id, ]

table(train_set$lunch_choice)
```

## **`Python` Code**
```{python}
import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
import statsmodels.api as sm
from sklearn.metrics import confusion_matrix, accuracy_score, recall_score, f1_score, log_loss
from sklearn.utils import resample

df = pd.read_csv("./14-multinomial-logistic_files/simulations/problem-1-data.csv")

print(df.isnull().sum())

levels = ["Cafeteria", "Packed_Lunch", "Food_Delivery", "Fast_Food"]
df = df.dropna().copy()
df["lunch_choice"] = pd.Categorical(df["lunch_choice"], categories=levels)
df["has_meal_plan"] = pd.Categorical(df["has_meal_plan"], categories=["No", "Yes"])
df["dietary_restriction"] = pd.Categorical(df["dietary_restriction"], categories=["No", "Yes"])

train_set = df.sample(frac=0.8, random_state=1411)
test_set = df.drop(train_set.index)

print(train_set["lunch_choice"].value_counts())
```

:::
There are a small number of missing values in `weekly_food_budget` and `time_between_classes`, so dropping incomplete rows is reasonable here. The training set is not balanced: `Cafeteria` is the most common choice, while `Fast_Food`, `Packed_Lunch`, and `Food_Delivery` are less frequent.
</details>

### B. Exploratory Data Analysis {.unnumbered}

1. Plot the distribution of `lunch_choice`.
2. Plot `weekly_food_budget` by `lunch_choice`. Which choices tend to have higher budgets?
3. Plot lunch choice proportions by `has_meal_plan`.
4. Plot lunch choice proportions across binned values of `time_between_classes`. Do the patterns look perfectly linear?

<details>
<summary><strong>Click to reveal answer</strong></summary>

::: {.panel-tabset}
## **`R` Code**
```{r}
ggplot(train_set, aes(x = lunch_choice, fill = lunch_choice)) +
  geom_bar() +
  theme_minimal() +
  labs(x = "Lunch choice", y = "Count", title = "Distribution of lunch choice")

ggplot(train_set, aes(x = lunch_choice, y = weekly_food_budget, fill = lunch_choice)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(x = "Lunch choice", y = "Weekly food budget")

ggplot(train_set, aes(x = has_meal_plan, fill = lunch_choice)) +
  geom_bar(position = "fill") +
  theme_minimal() +
  labs(x = "Has meal plan", y = "Proportion", fill = "Lunch choice")

train_set %>%
  mutate(time_bin = cut(time_between_classes, breaks = seq(0, 160, by = 20))) %>%
  count(time_bin, lunch_choice) %>%
  group_by(time_bin) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = time_bin, y = prop, color = lunch_choice, group = lunch_choice)) +
  geom_point() +
  geom_line() +
  theme_minimal() +
  labs(x = "Time between classes", y = "Proportion", color = "Lunch choice")
```

## **`Python` Code**
```{python}
counts = train_set["lunch_choice"].value_counts().reindex(levels)
plt.figure()
plt.bar(counts.index, counts.values)
plt.title("Distribution of lunch choice")
plt.xlabel("Lunch choice")
plt.ylabel("Count")
plt.xticks(rotation=20)
plt.show()

budget_groups = [
    train_set.loc[train_set["lunch_choice"] == level, "weekly_food_budget"]
    for level in levels
]
plt.figure()
plt.boxplot(budget_groups)
plt.xticks(range(1, len(levels) + 1), levels, rotation=20)
plt.xlabel("Lunch choice")
plt.ylabel("Weekly food budget")
plt.show()

meal_props = pd.crosstab(train_set["has_meal_plan"], train_set["lunch_choice"], normalize="index")
x = np.arange(len(meal_props.index))
bottom = np.zeros(len(meal_props.index))
plt.figure()
for level in levels:
    plt.bar(x, meal_props[level], bottom=bottom, label=level)
    bottom += meal_props[level].values
plt.xticks(x, meal_props.index)
plt.xlabel("Has meal plan")
plt.ylabel("Proportion")
plt.legend(title="Lunch choice", loc="upper center", bbox_to_anchor=(0.5, 1.18), ncol=2)
plt.tight_layout()
plt.show()

time_bins = pd.cut(train_set["time_between_classes"], bins=np.arange(0, 161, 20))
time_props = pd.crosstab(time_bins, train_set["lunch_choice"], normalize="index")
plt.figure()
for level in levels:
    plt.plot(time_props.index.astype(str), time_props[level], marker="o", label=level)
plt.xlabel("Time between classes")
plt.ylabel("Proportion")
plt.xticks(rotation=45)
plt.legend(title="Lunch choice", loc="upper center", bbox_to_anchor=(0.5, 1.18), ncol=2)
plt.tight_layout()
plt.show()
```

:::
`Cafeteria` is the most common lunch choice. Students choosing `Food_Delivery` and `Fast_Food` tend to have higher food budgets, while `Packed_Lunch` is more common among students with lower budgets or dietary restrictions. Meal-plan students are much more likely to choose `Cafeteria`. The relationship with `time_between_classes` is not perfectly linear, which is common in behavioural data.
</details>

### C. Data Modelling {.unnumbered}

1. Fit a Multinomial Logistic regression model predicting `lunch_choice` using all other variables. Use `Cafeteria` as the baseline class.
2. Display a summary of the fitted model.
3. Exponentiate the coefficients to obtain odds ratios.
4. Since the outcome classes are imbalanced, create an upsampled training set and fit a second Multinomial Logistic regression model.

<details>
<summary><strong>Click to reveal answer</strong></summary>

::: {.panel-tabset}
## **`R` Code**
```{r}
model1 <- multinom(
  lunch_choice ~ weekly_food_budget + time_between_classes + distance_from_home +
    has_meal_plan + dietary_restriction,
  data = train_set,
  trace = FALSE
)

summary(model1)
exp(coef(model1))

set.seed(1411)
train_balanced <- upSample(
  x = train_set[, c("weekly_food_budget", "time_between_classes",
                    "distance_from_home", "has_meal_plan", "dietary_restriction")],
  y = train_set$lunch_choice,
  yname = "lunch_choice"
)

train_balanced$lunch_choice <- relevel(train_balanced$lunch_choice, ref = "Cafeteria")

model_balanced <- multinom(
  lunch_choice ~ weekly_food_budget + time_between_classes + distance_from_home +
    has_meal_plan + dietary_restriction,
  data = train_balanced,
  trace = FALSE
)

summary(model_balanced)
```

## **`Python` Code**
```{python}
predictors = [
    "weekly_food_budget", "time_between_classes", "distance_from_home",
    "has_meal_plan", "dietary_restriction"
]

X_train = pd.get_dummies(train_set[predictors], drop_first=True, dtype=float)
X_train = sm.add_constant(X_train, has_constant="add")
y_train = train_set["lunch_choice"].cat.codes

model1 = sm.MNLogit(y_train, X_train).fit(method="newton", maxiter=100, disp=False)
print(model1.summary())

odds_ratios = np.exp(model1.params)
odds_ratios.columns = levels[1:]
print(odds_ratios.round(3))

max_n = train_set["lunch_choice"].value_counts().max()
balanced_parts = []
for level in levels:
    subset = train_set[train_set["lunch_choice"] == level]
    balanced_parts.append(resample(subset, replace=True, n_samples=max_n, random_state=1411))

train_balanced = pd.concat(balanced_parts)

X_bal = pd.get_dummies(train_balanced[predictors], drop_first=True, dtype=float)
X_bal = sm.add_constant(X_bal, has_constant="add")
y_bal = train_balanced["lunch_choice"].cat.codes

model_balanced = sm.MNLogit(y_bal, X_bal).fit(method="newton", maxiter=100, disp=False)
print(model_balanced.summary())
```

:::
The original model is the better choice for interpreting associations because it uses the observed class distribution. The balanced model is useful as a prediction comparison because it gives the minority classes more influence during fitting.
</details>

### D. Goodness of Fit {.unnumbered}

1. Use the test set to compute predictions from both the original and balanced models.
2. Create a confusion matrix for each model.
3. Compare accuracy, macro recall, macro F1, and log-loss.
4. Which model performs better if the main goal is overall accuracy? Which model performs better if the main goal is recognizing less common lunch choices?

<details>
<summary><strong>Click to reveal answer</strong></summary>

::: {.panel-tabset}
## **`R` Code**
```{r}
pred1 <- predict(model1, newdata = test_set)
cm1 <- confusionMatrix(
  factor(pred1, levels = levels(test_set$lunch_choice)),
  factor(test_set$lunch_choice, levels = levels(test_set$lunch_choice))
)

pred_bal <- predict(model_balanced, newdata = test_set)
cm_bal <- confusionMatrix(
  factor(pred_bal, levels = levels(test_set$lunch_choice)),
  factor(test_set$lunch_choice, levels = levels(test_set$lunch_choice))
)

print(cm1$table)
print(cm_bal$table)

probs1 <- predict(model1, newdata = test_set, type = "probs")
probs_bal <- predict(model_balanced, newdata = test_set, type = "probs")

log_loss <- function(probs, actual) {
  p <- probs[cbind(seq_along(actual), match(actual, colnames(probs)))]
  -mean(log(pmax(p, 1e-15)))
}

metrics <- tibble(
  Model = c("Original", "Balanced"),
  Accuracy = c(cm1$overall["Accuracy"], cm_bal$overall["Accuracy"]),
  Macro_Recall = c(mean(cm1$byClass[, "Recall"], na.rm = TRUE),
                   mean(cm_bal$byClass[, "Recall"], na.rm = TRUE)),
  Macro_F1 = c(mean(cm1$byClass[, "F1"], na.rm = TRUE),
               mean(cm_bal$byClass[, "F1"], na.rm = TRUE)),
  Log_Loss = c(log_loss(probs1, test_set$lunch_choice),
               log_loss(probs_bal, test_set$lunch_choice))
)

print(metrics)
```

## **`Python` Code**
```{python}
def prepare_x(data, columns):
    x = pd.get_dummies(data[predictors], drop_first=True, dtype=float)
    x = sm.add_constant(x, has_constant="add")
    return x.reindex(columns=columns, fill_value=0)

def summarize_predictions(probs, actual, model_name):
    pred = pd.Categorical([levels[i] for i in probs.idxmax(axis=1)], categories=levels)
    cm = pd.DataFrame(confusion_matrix(actual, pred, labels=levels),
                      index=levels, columns=levels)
    metrics = pd.DataFrame({
        "Model": [model_name],
        "Accuracy": [accuracy_score(actual, pred)],
        "Macro_Recall": [recall_score(actual, pred, labels=levels, average="macro", zero_division=0)],
        "Macro_F1": [f1_score(actual, pred, labels=levels, average="macro", zero_division=0)],
        "Log_Loss": [log_loss(actual, probs.values, labels=levels)]
    })
    return cm, metrics

X_test1 = prepare_x(test_set, X_train.columns)
probs1 = model1.predict(X_test1)
cm1, metrics1 = summarize_predictions(probs1, test_set["lunch_choice"], "Original")

X_test_bal = prepare_x(test_set, X_bal.columns)
probs_bal = model_balanced.predict(X_test_bal)
cm_bal, metrics_bal = summarize_predictions(probs_bal, test_set["lunch_choice"], "Balanced")

print(cm1)
print(cm_bal)
print(pd.concat([metrics1, metrics_bal]).round(3))
```

:::
The original model usually has higher overall accuracy and better log-loss because it respects the true class frequencies. The balanced model often improves macro recall, meaning it is somewhat better at noticing less common lunch choices, but it may do worse at predicting the majority class.
</details>

### E. Inference {.unnumbered}

1. Using the original model, identify statistically significant predictors for each non-baseline category.
2. Interpret the odds ratio for `weekly_food_budget` in the `Food_Delivery` versus `Cafeteria` comparison.
3. Interpret the odds ratio for `has_meal_planYes` in one non-baseline comparison.
4. Explain why these interpretations are associations rather than causal effects.

<details>
<summary><strong>Click to reveal answer</strong></summary>

::: {.panel-tabset}
## **`R` Code**
```{r}
broom::tidy(model1, conf.int = TRUE, exponentiate = TRUE) %>%
  mutate_if(is.numeric, round, 3) %>%
  filter(p.value < 0.05) %>%
  rename(outcome_class = y.level)
```

## **`Python` Code**
```{python}
results = []
for j, level in enumerate(levels[1:]):
    results.append(pd.DataFrame({
        "outcome_class": level,
        "term": model1.params.index,
        "odds_ratio": np.exp(model1.params[j]),
        "pvalue": model1.pvalues[j]
    }))

results = pd.concat(results)
print(results[results["pvalue"] < 0.05].round(3))
```

:::
For `Food_Delivery` versus `Cafeteria`, the odds ratio for `weekly_food_budget` is greater than 1. This means that higher weekly food budgets are associated with higher odds of choosing `Food_Delivery` rather than `Cafeteria`, holding other variables fixed. For `has_meal_planYes`, the odds ratios for non-cafeteria choices are below 1, so having a meal plan is associated with lower odds of choosing those options over `Cafeteria`. These are associations because students were not randomly assigned budgets, meal plans, or dietary restrictions.
</details>

### F. Prediction and Final Evaluation {.unnumbered}

1. Suppose a student has a weekly food budget of 95 dollars, 25 minutes between classes, lives 6 km from campus, does not have a meal plan, and does not have a dietary restriction. Use both models to predict this student's lunch-choice probabilities.
2. Which lunch choice is most likely under each model?
3. Returning to the original campus dining question: what have we learned about understanding and predicting student lunch choice?
4. Write a short storytelling summary for a campus dining office. Your summary should explain the main finding, the prediction limitation, and one practical implication in plain language.

<details>
<summary><strong>Click to reveal answer</strong></summary>

::: {.panel-tabset}
## **`R` Code**
```{r}
new_student <- data.frame(
  weekly_food_budget = 95,
  time_between_classes = 25,
  distance_from_home = 6,
  has_meal_plan = factor("No", levels = levels(train_set$has_meal_plan)),
  dietary_restriction = factor("No", levels = levels(train_set$dietary_restriction))
)

predict(model1, newdata = new_student, type = "probs")
predict(model_balanced, newdata = new_student, type = "probs")

predict(model1, newdata = new_student)
predict(model_balanced, newdata = new_student)
```

## **`Python` Code**
```{python}
new_student = pd.DataFrame({
    "weekly_food_budget": [95],
    "time_between_classes": [25],
    "distance_from_home": [6],
    "has_meal_plan": ["No"],
    "dietary_restriction": ["No"]
})

new_student["has_meal_plan"] = pd.Categorical(new_student["has_meal_plan"], categories=["No", "Yes"])
new_student["dietary_restriction"] = pd.Categorical(new_student["dietary_restriction"], categories=["No", "Yes"])

X_new1 = prepare_x(new_student, X_train.columns)
new_probs1 = model1.predict(X_new1)
new_probs1.columns = levels

X_new_bal = prepare_x(new_student, X_bal.columns)
new_probs_bal = model_balanced.predict(X_new_bal)
new_probs_bal.columns = levels

print(new_probs1.round(3))
print(new_probs_bal.round(3))
print(new_probs1.idxmax(axis=1))
print(new_probs_bal.idxmax(axis=1))
```

:::
The original model may still lean toward `Cafeteria`, because that is the majority class in the observed data. The balanced model may assign more probability to minority classes such as `Fast_Food` or `Food_Delivery` for this profile. Overall, the predictors are useful for understanding lunch-choice patterns, especially the roles of food budget and meal plan status, but prediction is only moderate. Accuracy alone is not enough here, because the model can look good while missing less common lunch choices.

A short storytelling summary could be: student lunch choice is shaped by practical constraints such as food budget, meal-plan status, dietary restrictions, and time between classes. The model can identify broad patterns, such as meal-plan students being more likely to use the cafeteria and higher-budget students being more likely to buy food elsewhere, but it should not be treated as a perfect individual-level prediction tool. For a campus dining office, the main implication is to use the model as a planning aid for understanding demand across lunch options, while still looking carefully at minority choices that overall accuracy may hide.
</details>

:::
