<!-- Google tag (gtag.js) -->
<script async src="https://www.googletagmanager.com/gtag/js?id=G-7PRVEBE1EF"></script>
<script>
  window.dataLayer = window.dataLayer || [];
  function gtag(){dataLayer.push(arguments);}
  gtag('js', new Date());

  gtag('config', 'G-7PRVEBE1EF');
</script>

# Multinomial Logistic Regression {#sec-multinomial-logistic}

::: {.Warning}
::::{.Warning-header}
When to Use and Not Use Multinomial Logistic Regression
::::
::::{.Warning-container}
Multinomial Logistic regression is a type of generalized linear model that is appropriately used under the following conditions:

- The outcome variable is categorical with more than two categories, as illustrated in @fig-multinomial-regression.
- There is one or more regressors that may be continuous, ordinal, or nominal.
- We assume a **linear relationship** between any continuous regressors and the logit of the cumulative probabilities of the outcome.
- The data points do not exhibit any correlation structures, which means that there is statistical independence across the sampled observations, with mutually exclusive and exhaustive ordered categories.

However, Multinomial Logistic regression should not be used in the following scenarios:

- The outcome variable is **ordered**. If the categories are ordered, use [Ordinal Logistic regression](https://alexrod61.github.io/regression-cookbook/book/15-ordinal-logistic.html).
- The outcome variable is **binary**. For binary outcomes, use [Binary Logistic regression](https://alexrod61.github.io/regression-cookbook/book/08-binary-logistic.html).
- The outcome variable is **count**. For count outcomes, refer to count regression models, such as [Classical Poisson regression](https://alexrod61.github.io/regression-cookbook/book/10-classical-poisson.html) and [Negative Binomial regression](https://alexrod61.github.io/regression-cookbook/book/10-negative-binomial.html).
- The outcome variable is **continuous**. For continuous outcomes, refer to the ["continuous cuisine"](https://alexrod61.github.io/regression-cookbook/book/continuous-zone.html) section of this book. For example, consider using [Ordinary Least-squares regression](https://alexrod61.github.io/regression-cookbook/book/03-ols.html) for continuous and unbounded outcomes.
- The data points exhibit correlation structures, indicating statistical dependence among the sampled observations (for example, repeated measurements on the same subject).
::::
:::

::: {#fig-multinomial-regression}
```{mermaid}
mindmap
  root((Regression 
  Analysis)
    Continuous <br/>Outcome Y
      {{Unbounded <br/>Outcome Y}}
        )Chapter 3: <br/>Ordinary <br/>Least Squares <br/>Regression(
          (Normal <br/>Outcome Y)
      {{Nonnegative <br/>Outcome Y}}
        )Chapter 4: <br/>Gamma Regression(
          (Gamma <br/>Outcome Y)
      {{Bounded <br/>Outcome Y <br/> between 0 and 1}}
        )Chapter 5: Beta <br/>Regression(
          (Beta <br/>Outcome Y)
      {{Nonnegative <br/>Survival <br/>Time Y}}
        )Chapter 6: <br/>Parametric <br/> Survival <br/>Regression(
          (Exponential <br/>Outcome Y)
          (Weibull <br/>Outcome Y)
          (Lognormal <br/>Outcome Y)
        )Chapter 7: <br/>Semiparametric <br/>Survival <br/>Regression(
          (Cox Proportional <br/>Hazards Model)
            (Hazard Function <br/>Outcome Y)
    Discrete <br/>Outcome Y
      {{Binary <br/>Outcome Y}}
        {{Ungrouped <br/>Data}}
          )Chapter 8: <br/>Binary Logistic <br/>Regression(
            (Bernoulli <br/>Outcome Y)
        {{Grouped <br/>Data}}
          )Chapter 9: <br/>Binomial Logistic <br/>Regression(
            (Binomial <br/>Outcome Y)
      {{Count <br/>Outcome Y}}
        {{Equidispersed <br/>Data}}
          )Chapter 10: <br/>Classical Poisson <br/>Regression(
            (Poisson <br/>Outcome Y)
        {{Overdispersed <br/>Data}}
          )Chapter 11: <br/>Negative Binomial <br/>Regression(
            (Negative Binomial <br/>Outcome Y)
        {{Zero Inflated <br/>Data}}
          )Chapter 12: <br/>Zero Inflated <br/>Poisson <br/>Regression(
            (Zero Inflated <br/>Poisson <br/>Outcome Y)
        {{Overdispersed or <br/>Underdispersed <br/>Data}}
          )Chapter 13: <br/>Generalized <br/>Poisson <br/>Regression(
            (Generalized <br/>Poisson <br/>Outcome Y)
      {{Categorical <br/>Outcome Y}}
        {{Nominal <br/>Outcome Y}}
          )Chapter 14: <br/>Multinomial <br/>Logistic <br/>Regression(
            (Multinomial <br/>Outcome Y)
```
:::

```{r setup-r, include=FALSE}
# Load required libraries
library(devtools)
library(readr)
library(tibble)
library(tidyr)
library(dplyr)
library(magrittr)
library(ggplot2)
library(gridExtra) # combine ggplot2 figures
library(ggExtra) # draw marginal histograms
library(broom)
library(knitr)
#devtools::install_github("andytai7/cookbook")
library(cookbook)
library(nnet) # fit multinomial regression
library(caret) # upsampling and confusion matrix
library(rsample)

library(reticulate)
py_config()

py_require("pandas")
py_require("numpy")
py_require("seaborn")
py_require("pyreadr")
py_require("plotnine")
py_require("matplotlib")
py_require("statsmodels")
py_require("scikit-learn")
py_require("spicy")

reticulate::py_install("plotnine")
reticulate::py_install("imblearn")

pandas <- import("pandas")
numpy <- import("numpy")
seaborn <- import("seaborn")
pyreadr <- import("pyreadr")
matplotlib <- import("matplotlib")
statsmodels <- import("statsmodels")
sklearn <- import("sklearn")
scipy <- import("scipy")
plotnine <- import("plotnine")
imblearn <- import("imblearn")

print_as_dt <- function(x){
  DT::datatable(x,
                rownames = FALSE,
                options = list(
                  dom = 'ltipr',
                  autoWidth = TRUE,
                  columnDefs = list(
                    list(className = 'dt-left', targets = "_all"),
                    list(width = '200px', targets = "_all")
                    )
                  )
                )
}
```

```{python setup-py, include=FALSE}
import imblearn
import numpy as np
import pandas as pd
from plotnine import *
from statsmodels.formula.api import glm, ols, mnlogit
from statsmodels.genmod.families import Poisson
from matplotlib import pyplot as plt
from sklearn.metrics import mean_squared_error, mean_absolute_error, classification_report
import scipy.stats as stats
import seaborn as sns
from matplotlib.gridspec import GridSpec
from sklearn.linear_model import LogisticRegression
from sklearn.metrics import confusion_matrix, ConfusionMatrixDisplay, classification_report, log_loss, accuracy_score, recall_score, f1_score
from sklearn.utils import resample
import statsmodels.api as sm
```

::: {.LO}
::::{.LO-header}
Learning Objectives
::::
::::{.LO-container}
By the end of this chapter, you will be able to:

- Describe why **Ordinary Least Squares** is not appropriate for multi-class nominal outcomes.
- Determine when Multinomial Logistic regression is an **appropriate modelling choice.**
- Frame **inferential** and **predictive inquiries** for a Multinomial Logistic regression problem using the **data science workflow**.
- Write down the **likelihood function** for Multinomial Logistic regression and its **logit link** functions, and obtain category-specific probabilities.
- Understand how coefficient estimation via **maximum likelihood estimation** works in the Multinomial Logistic regression model.
- Fit a Multinomial Logistic regression model in both `R` and `Python`.
- Interpret Multinomial Logistic regression coefficients in real scenarios.
- Use Multinomial Logistic regression for prediction by computing category-specific probabilities.
- Evaluate model performance using both **likelihood-based** and **classification-based** measures, and construct confidence intervals using **Wald Z statistic**.
::::
:::

## Introduction

Multinomial Logistic regression is a statistical modelling technique used to analyze relationships between a multi-class categorical response variable and a set of regressors. Unlike [Binary Logistic regression](http://localhost:54321/book/08-binary-logistic.html), which models outcomes with two categories, Multinomial Logistic regression extends the framework to outcomes with more than two unordered categories. It estimates the probability of each possible outcome as a function of the regressors by modelling the log-odds of each category relative to a reference category. This method is widely used in fields such as transportation, marketing, and social sciences where the goal is to understand or predict categorical choices among multiple alternatives.

### Assumptions
The following are the assumptions that we need to be aware of before fitting this model:

- The response or dependent variable should be **categorical** and **nominal** with more than two categories. 
- There should be one or more independent variables that are continuous, ordinal or nominal (including dichotomous variables).
- **Logit linearity assumption**: There is be a linear relationship between any continuous independent variables and the logit of the category-specific probabilities of the dependent variable. This assumption will be explained in @sec-data-modelling.
- **Independent observations**: The observations are independent, and the dependent variable consists of mutually exclusive and exhaustive categories (i.e. no individual belonging to two different categories).
- **No Multicollinearity**: Multicollinearity occurs when you have two or more independent variables that are highly correlated with each other.
- There should be no outliers, high leverage values or highly influential points.

### Use Cases of Multinomial Logistic Regression
Multinomial Logistic regression is particularly useful when the outcome
variable has three or more categories with no meaningful ranking. Some
examples of this are:

1.  **Travel Mode Choice:** Transportation planners model how commuters
    select among alternatives such as `Bus`, `Car`,
    `Bike`, and `Walk`. Regressors typically include travel time,
    travel cost, distance, and household characteristics, and the fitted
    probabilities are used to forecast shifts in commute patterns under proposed
    infrastructure or pricing changes.

2.  **Voting and Party Choice:** Political scientists study how
    individual characteristics such as age, income, gender, and
    education influence which party a voter supports in a multiparty
    election (for example, `Liberal` / `Conservative` / `NDP` / `Green`). No party is "higher" than another, so      the outcome categories are purely nominal.

3.  **Consumer Brand Choice:** Marketing analysts use scanner panel data
    to model which brand a household purchases within a product
    category. Regressors such as brand loyalty, regular shelf price,
    promotional price cuts, and the presence of in-store promotions are
    used to estimate price elasticities and predict market share.

4.  **Disease Subtype Classification:** Epidemiologists often treat a
    heterogeneous disease as several distinct subtypes rather than a
    single binary outcome. For instance, breast cancer may be modelled
    across molecular subtypes defined by receptor status, with age,
    family history, body mass index, and breast density as regressors,
    allowing risk factors to act differently on each subtype.

5.  **Occupational Choice:** Labour economists model the occupational
    category an individual enters (for example, `professional`,
    `clerical`, `service`, or `manual`) as a function of education,
    work experience, parental occupation, and other socioeconomic
    characteristics.

In all these examples, the categories are mutually exclusive and
exhaustive but carry no natural ordering. Imposing an order, as [Ordinal Logistic regression](https://alexrod61.github.io/regression-cookbook/book/15-ordinal-logistic.html) does, would assume a progression between categories
that does not exist, whereas fitting a separate [Binary Logistic regression model](https://alexrod61.github.io/regression-cookbook/book/08-binary-logistic.html) for each
pair of categories would discard information and could yield predicted
probabilities that do not sum to one. Multinomial logistic regression
provides a principled framework that models all categories jointly
against a common reference level.

## Case Study: Understanding Commuting Behaviors

To demonstrate Multinomial Logistic regression, we will use a simulated dataset containing information on individuals' commuting behaviors and demographics, with a focus on their primary mode of transportation. This dataset provides an example of a multi-class categorical response variable. We will approach this case study using the [data science workflow](https://alexrod61.github.io/regression-cookbook/book/01-intro.html) to ensure a structured approach to problem-solving and model building.

### The Dataset

Our dataset captures various characteristics of commuters and each commuter's primary mode of transport. Each row represents a commuter, and the columns describe different characteristics. @tbl-variables provides the relevant details for the regressors in this dataset - namely, commute distance, age, car availability, and whether the individual commutes during weekends. Note that we also indicate the **coding names** of all the variables involved in the last column.

| Variable | Description | Type | Scale | Model Role | Coding Name |
|---|---|---|---|---|---|---|
| Mode of Transport | The primary mode of transport used by the individual | Discrete | Categorical | Response | `transport_mode` |
| Commute Distance | The one-way commuting distance in kilometers | Continuous | Bounded positive | Regressor | `commute_distance` |
| Age | The age of the individual in years | Continuous | Bounded positive | Regressor | `age` |
| Car Availability | Whether the individual has access to a car | Discrete| Binary | Regressor | `has_car_available` |
| Weekend Commuter | Whether the individual commutes during weekends | Discrete | Binary | Regressor | `weekend_commuter` |

: Classification table for variables in the transport dataset. {#tbl-variables}

Let's load this sample of $n=1000$ commuters in both `R` and `Python`. In `R`, we use the `tidyverse` library, and in `Python`, we use the `pandas` library. 

::: {.panel-tabset}

## R Code

```{.r}
data(multinomial_transport)
data <- multinomial_transport[,-1]
cat("There are", nrow(data), "observations and", ncol(data), "variables in the dataset.")
head(data, n = 100) |>
  print_as_dt()
```

## Python Code

```{.python}
url = "https://raw.githubusercontent.com/andytai7/cookbook/refs/heads/main/raw-data/multinomial_transport.csv"
data = pd.read_csv(url)
data = data.iloc[:, :-1]
print(f"There are {data.shape[0]} observations and {data.shape[1]} variables in the dataset.")
print(data)
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
data(multinomial_transport)
data <- multinomial_transport %>% select(-'income_bracket')
cat("There are", nrow(data), "observations and", ncol(data), "variables in the dataset.")
head(data, n = 100) |>
  print_as_dt()
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
url = "https://raw.githubusercontent.com/andytai7/cookbook/refs/heads/main/raw-data/multinomial_transport.csv"
data = pd.read_csv(url)
data = data.iloc[:, :-1]
print(f"There are {data.shape[0]} observations and {data.shape[1]} variables in the dataset.")
```

```{r}
#| echo: false
#| message: false
#| warning: false
head(data, n = 100) |>
  print_as_dt()
```

:::

This dataset provides a structured way to analyze the behaviors of commuters and determine which factors are associated with their modes of transport. 

### Study Design

Now that we’ve introduced our case study and dataset, it’s time to follow the [data science workflow](https://alexrod61.github.io/regression-cookbook/book/01-intro.html) step by step. The first step is to define the main statistical inquiries we want to address. Our goal in this case study is to understand which factors impact an individual's mode of transport. We aim to identify which characteristics, such as age, commute distance, having a car, and commuting on weekends or not, explain whether people commute by public transit, car, bicycle, or walk. 

We will explore two primary questions based on this dataset.

- How are an individual's age, commute distance, availability of car, and whether they commute on weekends or not, **associated** with their mode of transport?
- Can we accurately **predict** the transport mode of an individual based on their commuting behaviors and demographics including if they have a car, if they commute on weekends, their commute distance, and age?

These questions correspond to two distinct **statistical objectives**:

- **Inference:** Understand the relationship between regressors (`age`, `commute_distance`, `has_car_available` and `weekend_commuter`) and the probability of each category of the response variable `transport_mode`.
- **Prediction:** Develop a classification model that classifies individuals into their most probable mode of transport: `Bicycle`, `Car`, `Public Transit`, or `Walking` based on factors like age, commute distance, availability of a car, and whether the person commutes on weekends or not. 

## Data Collection and Wrangling

The dataset used in this chapter is a synthetic dataset from the [cookbook](https://github.com/andytai7/cookbook) package in `R`. Now that our dataset is loaded, the next step is to clean and organze it. Data wrangling involves several steps that ensure our data is accurate, consistent, and ready for modelling. Here are some key tasks:

#### Handling Missing Data

The first task is to ensure data integrity by checking for missing values. Missing data can occur for various reasons, such as unrecorded responses or errors in data entry. When we find missing values—for example, if some individuals don’t have recorded age or commute distance—we must decide how to handle these gaps. Common strategies include:

- **Removing incomplete records**: If the amount of missing data is minimal or missingness is random.
- **Imputing missing values**: Using logical estimates or averages if missingness follows a systematic pattern.

In our simulated dataset, there are no missing values, as confirmed by:

::: {.panel-tabset}
## **`R` Code**
```{r}
colSums(is.na(data))
```

## **`Python` Code**
```{python}
# Count missing values in each column
data.isna().sum()
```
:::

#### Encoding Categorical Variables

For regression analysis, we need to convert categorical variables into numerical representations. In the following code in both `R` and `Python`, we will factor all categorical variables, such as `transport_mode`, `has_car_available` and `weekend_commuter`, so that the model correctly interprets them as categorical data rather than continuous numbers.

::: {.panel-tabset}

## R Code

```{.r}
# Factorize the categorical variables and print their levels
categorical_cols <- c('transport_mode', 'has_car_available', 'weekend_commuter')
for (col in categorical_cols) {
  data[[col]] <- as.factor(data[[col]])
  cat("Levels of", col, ":", levels(data[[col]]), "\n")
}
```

## Python Code

```{.python}
# Factorize the categorical variables and print their levels
categorical_cols = ['transport_mode', 'has_car_available', 'weekend_commuter']
for col in categorical_cols:
  data[col] = data[col].astype('category')
  print(f"Levels of {col}: {list(data[col].cat.categories)}")
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
categorical_cols <- c('transport_mode', 'has_car_available', 'weekend_commuter')
for (col in categorical_cols) {
  data[[col]] <- as.factor(data[[col]])
  cat("Levels of", col, ":", levels(data[[col]]), "\n")
}
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
categorical_cols = ['transport_mode', 'has_car_available', 'weekend_commuter']
for col in categorical_cols:
  data[col] = data[col].astype('category')
  print(f"Levels of {col}: {list(data[col].cat.categories)}")
```

:::

#### Splitting the Data for Model Training

To ensure that our model generalizes well to unseen data, we will split the dataset into train and test sets. We will use the train set to fit the model and to estimate the model parameters, and the test set to assess model performance for both **inference** and **prediction**. The split is typically done in an 80/20 ratio, as shown below:

::: {.panel-tabset}

## R Code

```{.r}
# Randomly split the data into train and test sets
set.seed(352)
train_idx <- sample(seq_len(nrow(data)), size = 0.8 * nrow(data))
train_set <- data[train_idx, ]
test_set  <- data[-train_idx, ]

cat("train set size:", nrow(train_set))
cat("Test set size:", nrow(test_set))
```

## Python Code

```{.python}
# The train/test split is created once in R (caret::createDataPartition; see the
# R tab) and pulled into Python via reticulate's `r` object, so both languages
# use identical train and test sets throughout the chapter.
train_set = r.train_set.reset_index(drop=True)
test_set  = r.test_set.reset_index(drop=True)

print(f"train set size: {len(train_set)}")
print(f"Test set size: {len(test_set)}")
```

:::

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on keeping `R` and `Python` aligned after the split!
::::
:::: {.Heads-up-container}
Using the same seed and split proportion in `R` and `Python` does not guarantee that the random splits will contain identical rows. This is not an error. The two ecosystems use different splitting functions and pseudo-random number generators. Therefore, for the rest of this chapter, we use the `R`-generated train and test sets as a common reference to ensure consistent coding outputs.
::::
:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
set.seed(352)

# Randomly split the data into train and test sets
train_idx <- sample(seq_len(nrow(data)), size = 0.8 * nrow(data))
train_set <- data[train_idx, ]
test_set  <- data[-train_idx, ]

cat("train set size:", nrow(train_set))
cat("Test set size:", nrow(test_set))
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
# The train/test split is created once in R (caret::createDataPartition; see the
# R tab) and pulled into Python via reticulate's `r` object, so both languages
# use identical train and test sets throughout the chapter.
train_set = r.train_set.reset_index(drop=True)
test_set  = r.test_set.reset_index(drop=True)

print(f"train set size: {len(train_set)}")
print(f"Test set size: {len(test_set)}")
```

:::

## Exploratory Data Analysis {#sec-eda}

Before diving into data modelling, it is crucial to develop a deep understanding of the relationships between variables in the dataset. The Exploratory Data Analysis (EDA) stage helps us visualize and summarize the data, uncover patterns, detect anomalies, and test key assumptions that will inform our modelling decisions.

### Continuous Variables

We begin by examining continuous variables. There are two continuous regressors: `age` and `commute_distance`, and we will examine their relationship with the response variable `transport_mode` through visualizations, such as box plots and violin plots. 

::: {.panel-tabset}

## R Code

```{.r}
# Custom color palette `cbPalette` 
cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# Visualize boxplots and violin plots of commute_distance v.s. transport_mode
p1 <- train_set %>%
    ggplot(aes(x = transport_mode, y = commute_distance, fill = transport_mode)) +
    geom_violin(alpha = 0.6, trim = FALSE) +
    stat_boxplot(geom = "errorbar", width = 0.2) +
    geom_boxplot(alpha = 1, fill = NA) +
    labs(x = "Transport Mode", y = "Commute Distance (km)", title =
        "Commute Distance by Transport Mode") +
    scale_fill_manual(values = cbPalette[3:6]) +
    theme_bw() +
    theme(legend.position = "none") +
    guides(fill = guide_legend(title = "Transport Mode"))

p2 <- train_set %>%
    ggplot(aes(x = transport_mode, y = age, fill = transport_mode)) +
    geom_violin(alpha = 0.6, trim = FALSE) +
    stat_boxplot(geom = "errorbar", width = 0.2) +
    geom_boxplot(alpha = 1, fill = NA) +
    labs(x = "Transport Mode", y = "Age", title =
        "Age by Transport Mode") +
    scale_fill_manual(values = cbPalette[3:6]) +
    theme_bw() +
    theme(legend.position = "none") +
    guides(fill = guide_legend(title = "Transport Mode"))

grid.arrange(p1, p2, ncol = 2, top = grid::textGrob("Box Plots and Violin Plots", gp = grid::gpar(fontsize = 16, fontface = "bold")))
```

## Python Code

```{.python}
sns.set_theme(style="whitegrid")

# Custom color palette `cbPalette` 
cbPalette = ["#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7"]

# Sort transport_mode by frequency
order = train_set['transport_mode'].value_counts().index
train_set['transport_mode'] = pd.Categorical(train_set['transport_mode'], categories=order, ordered=True)

# Set up side-by-side subplots with shared layout
fig = plt.figure(figsize=(8, 7))
gs = GridSpec(1, 2, figure=fig)
fig.suptitle("Box Plots and Violin Plots", fontsize=16, fontweight='bold')

# Plot Commute Distance vs Transport Mode
ax1 = fig.add_subplot(gs[0, 0])
sns.violinplot(
    data=train_set,
    x='transport_mode',
    y='commute_distance',
    palette=cbPalette[2:6],
    ax=ax1,
    alpha=0.6,
    inner=None,
    order=order,
    cut=3
)
sns.boxplot(
    data=train_set,
    x='transport_mode',
    y='commute_distance',
    showcaps=True,
    boxprops={'facecolor': 'none', 'edgecolor': 'black'},
    whiskerprops={'color': 'black'},
    medianprops={'color': 'black', 'linewidth': 2},
    flierprops={'markerfacecolor': 'black', 'markersize': 3},
    ax=ax1,
    order=order
)
ax1.set_title("Commute Distance by Transport Mode")
ax1.set_xlabel("Transport Mode")
ax1.set_ylabel("Commute Distance (km)")
ax1.get_legend()

# Plot Age vs Transport Mode
ax2 = fig.add_subplot(gs[0, 1])
sns.violinplot(
    data=train_set,
    x='transport_mode',
    y='age',
    palette=cbPalette[2:6],
    ax=ax2,
    alpha=0.6,
    inner=None,
    order=order,
    cut=3
)
sns.boxplot(
    data=train_set,
    x='transport_mode',
    y='age',
    showcaps=True,
    boxprops={'facecolor': 'none', 'edgecolor': 'black'},
    whiskerprops={'color': 'black'},
    medianprops={'color': 'black', 'linewidth': 2},
    flierprops={'markerfacecolor': 'black', 'markersize': 3},
    ax=ax2,
    order=order
)
ax2.set_title("Age by Transport Mode")
ax2.set_xlabel("Transport Mode")
ax2.set_ylabel("Age")
ax2.get_legend()

fig.tight_layout(rect=[0, 0.05, 1, 0.95])
fig
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| label: fig-eda-transport-mode-continuous-R
#| fig-cap:
#|   - "Boxplots and violin plots of commute distance and age by transport mode."
#| echo: false
#| message: false
#| warning: false
cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
p1 <- train_set %>%
    ggplot(aes(x = transport_mode, y = commute_distance, fill = transport_mode)) +
    geom_violin(alpha = 0.6, trim = FALSE) +
    stat_boxplot(geom = "errorbar", width = 0.2) +
    geom_boxplot(alpha = 1, fill = NA) +
    labs(x = "Transport Mode", y = "Commute Distance (km)", title =
        "Commute Distance by Transport Mode") +
    scale_fill_manual(values = cbPalette[3:6]) +
    theme_bw() +
    theme(legend.position = "none") +
    guides(fill = guide_legend(title = "Transport Mode"))
p2 <- train_set %>%
    ggplot(aes(x = transport_mode, y = age, fill = transport_mode)) +
    geom_violin(alpha = 0.6, trim = FALSE) +
    stat_boxplot(geom = "errorbar", width = 0.2) +
    geom_boxplot(alpha = 1, fill = NA) +
    labs(x = "Transport Mode", y = "Age", title =
        "Age by Transport Mode") +
    scale_fill_manual(values = cbPalette[3:6]) +
    theme_bw() +
    theme(legend.position = "none") +
    guides(fill = guide_legend(title = "Transport Mode"))
grid.arrange(p1, p2, ncol = 2, top = grid::textGrob("Box Plots and Violin Plots", gp = grid::gpar(fontsize = 16, fontface = "bold")))
```

## Python Output

```{python}
#| label: fig-eda-transport-mode-continuous-py
#| fig-cap:
#|   - "Boxplots and violin plots of commute distance and age by transport mode."
#| echo: false
#| message: false
#| warning: false
sns.set_theme(style="whitegrid")

# Custom color palette `cbPalette` 
cbPalette = ["#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7"]

# Sort transport_mode by frequency
order = train_set['transport_mode'].value_counts().index
train_set['transport_mode'] = pd.Categorical(train_set['transport_mode'], categories=order, ordered=True)

# Set up side-by-side subplots with shared layout
fig = plt.figure(figsize=(8, 7))
gs = GridSpec(1, 2, figure=fig)
fig.suptitle("Box Plots and Violin Plots", fontsize=16, fontweight='bold')

# Plot Commute Distance vs Transport Mode
ax1 = fig.add_subplot(gs[0, 0])
sns.violinplot(
    data=train_set,
    x='transport_mode',
    y='commute_distance',
    palette=cbPalette[2:6],
    ax=ax1,
    alpha=0.6,
    inner=None,
    order=order,
    cut=3
)
sns.boxplot(
    data=train_set,
    x='transport_mode',
    y='commute_distance',
    showcaps=True,
    boxprops={'facecolor': 'none', 'edgecolor': 'black'},
    whiskerprops={'color': 'black'},
    medianprops={'color': 'black', 'linewidth': 2},
    flierprops={'markerfacecolor': 'black', 'markersize': 3},
    ax=ax1,
    order=order
)
ax1.set_title("Commute Distance by Transport Mode")
ax1.set_xlabel("Transport Mode")
ax1.set_ylabel("Commute Distance (km)")
ax1.get_legend()  # no legend to remove manually

# Plot p2: Age vs Transport Mode
ax2 = fig.add_subplot(gs[0, 1])
sns.violinplot(
    data=train_set,
    x='transport_mode',
    y='age',
    palette=cbPalette[2:6],
    ax=ax2,
    alpha=0.6,
    inner=None,
    order=order,
    cut=3
)
sns.boxplot(
    data=train_set,
    x='transport_mode',
    y='age',
    showcaps=True,
    boxprops={'facecolor': 'none', 'edgecolor': 'black'},
    whiskerprops={'color': 'black'},
    medianprops={'color': 'black', 'linewidth': 2},
    flierprops={'markerfacecolor': 'black', 'markersize': 3},
    ax=ax2,
    order=order
)
ax2.set_title("Age by Transport Mode")
ax2.set_xlabel("Transport Mode")
ax2.set_ylabel("Age")
ax2.get_legend()  # no legend to remove manually

fig.tight_layout(rect=[0, 0.05, 1, 0.95])
fig
```

:::

The plots in @fig-eda-transport-mode-continuous-R and @fig-eda-transport-mode-continuous-py show how commute distance and age vary across different transport modes. Public transit users tend to have the longest commute distance, on average, with a wide distribution and a median around 17–18 km. Car users also have moderately long commutes, with a median around 10-11 km and a relatively broad spread. Bicycle and Walking are associated with shorter commute distances (medians < 5 km), and their distributions are tightly concentrated near the lower end.
All transport modes show some long-distance commuters, which are shown by long tails, and there are a few outliers for commute distance in `Bicycle`.

Car and public transit users have similar age distributions, with median ages in the mid-30s to early 40s.
Bicycle users have a more skewed age distribution with a higher median age in mid-50s. Walking has a much wider spread in age, with a high median (around 60) and some very young walkers as well. The variability in age is the highest for walking, which suggests diverse usage across age groups.

### Categorical Variables

The categorical variables can be visualized using stacked bar charts. The bar charts are very effective because they display the frequency of each category, helping us understand the distribution of qualitative attributes. The stacked bar charts in @fig-eda-transport-mode-distribution-R and @fig-eda-transport-mode-distribution-py display the distributions of transport mode across car availability (`has_car_available`) and weekend commuter status (`weekend_commuter`). 

::: {.panel-tabset}

## R Code

```{.r}
# Sort `transport_mode` by frequency
train_set$transport_mode <- factor(train_set$transport_mode, levels = names(sort(table(train_set$transport_mode), decreasing = TRUE)))

p1 <- train_set %>%
  ggplot(aes(x = transport_mode, fill = has_car_available)) +
  geom_bar(position = position_stack(reverse = TRUE)) +  # use "stack" for counts, "fill" for proportions
  labs(y = "Count", title = "With Car Availability") +
  scale_fill_manual(values=cbPalette) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(title = "Car Availability"))

p2 <- train_set %>%
  ggplot(aes(x = transport_mode, fill = weekend_commuter)) +
  geom_bar(position = position_stack(reverse = TRUE)) +  # use "stack" for counts, "fill" for proportions
  labs(x = "Transport Mode", y = "Count", title = "With Weekend Commuter Status") +
  scale_fill_manual(values=cbPalette) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(title = "Weekend Commuter"))

grid.arrange(p1, p2, ncol = 2, top = grid::textGrob("Distribution of Transport Mode", gp = grid::gpar(fontsize = 16, fontface = "bold")))
```

## Python Code

```{.python}
# Set up the plot
sns.set_theme(style="whitegrid")
fig = plt.figure()
fig.set_size_inches(8, 6)
gs = GridSpec(1, 2, figure=fig)
fig.suptitle("Distribution of Transport Mode", fontsize=16, fontweight='bold')

# Plot: Stacked bar for Car Availability
car_counts = train_set.groupby(['transport_mode', 'has_car_available']).size().unstack(fill_value=0)
car_counts = car_counts.loc[order] 

ax1 = fig.add_subplot(gs[0, 0])
bottom = None
for i, col in enumerate(car_counts.columns):
    ax1.bar(car_counts.index, car_counts[col], bottom=bottom, label=col, color=cbPalette[i % len(cbPalette)])
    bottom = car_counts[col] if bottom is None else bottom + car_counts[col]
ax1.set_title("With Car Availability")
ax1.set_xlabel("Transport Mode")
ax1.set_ylabel("Count")
ax1.legend(title="Car Availability", loc='upper center', bbox_to_anchor=(0.5, -0.15), ncol=2)

# Plot: Stacked bar for Weekend Commuter
weekend_counts = train_set.groupby(['transport_mode', 'weekend_commuter']).size().unstack(fill_value=0)
weekend_counts = weekend_counts.loc[order]

ax2 = fig.add_subplot(gs[0, 1])
bottom = None
for i, col in enumerate(weekend_counts.columns):
    ax2.bar(weekend_counts.index, weekend_counts[col], bottom=bottom, label=col, color=cbPalette[i % len(cbPalette)])
    bottom = weekend_counts[col] if bottom is None else bottom + weekend_counts[col]
ax2.set_title("With Weekend Commuter Status")
ax2.set_xlabel("Transport Mode")
ax2.set_ylabel("")
ax2.legend(title="Weekend Commuter", loc='upper center', bbox_to_anchor=(0.5, -0.15), ncol=2)

# Adjust plot layout
fig.tight_layout(rect=[0, 0.05, 1, 1])
fig
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| label: fig-eda-transport-mode-distribution-R
#| fig-cap:
#|   - "Stacked bar charts of transport mode counts by car availability and weekend commuter status."
#| message: false
#| warning: false
train_set$transport_mode <- factor(train_set$transport_mode, levels = names(sort(table(train_set$transport_mode), decreasing = TRUE)))
p1 <- train_set %>%
  ggplot(aes(x = transport_mode, fill = has_car_available)) +
  geom_bar(position = position_stack(reverse = TRUE)) +  # use "stack" for counts, "fill" for proportions
  labs(x = "Transport Mode", y = "Count", title = "With Car Availability") +
  scale_fill_manual(values=cbPalette) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(title = "Car Availability"))
p2 <- train_set %>%
  ggplot(aes(x = transport_mode, fill = weekend_commuter)) +
  geom_bar(position = position_stack(reverse = TRUE)) +  # use "stack" for counts, "fill" for proportions
  labs(x = "Transport Mode", y = "", title = "With Weekend Commuter Status") +
  scale_fill_manual(values=cbPalette) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(title = "Weekend Commuter"))
grid.arrange(p1, p2, ncol = 2, top = grid::textGrob("Distribution of Transport Mode", gp = grid::gpar(fontsize = 16, fontface = "bold")))
```

## Python Output

```{python}
#| label: fig-eda-transport-mode-distribution-py
#| fig-cap:
#|   - "Stacked bar charts of transport mode counts by car availability and weekend commuter status."
#| echo: false
#| message: false
#| warning: false
sns.set_theme(style="whitegrid")
fig = plt.figure()
fig.set_size_inches(8, 6)
gs = GridSpec(1, 2, figure=fig)
fig.suptitle("Distribution of Transport Mode", fontsize=16, fontweight='bold')

# Plot 1: Stacked bar for Car Availability
car_counts = train_set.groupby(['transport_mode', 'has_car_available']).size().unstack(fill_value=0)
car_counts = car_counts.loc[order] 
ax1 = fig.add_subplot(gs[0, 0])
bottom = None
for i, col in enumerate(car_counts.columns):
    ax1.bar(car_counts.index, car_counts[col], bottom=bottom, label=col, color=cbPalette[i % len(cbPalette)])
    bottom = car_counts[col] if bottom is None else bottom + car_counts[col]
ax1.set_title("With Car Availability")
ax1.set_xlabel("Transport Mode")
ax1.set_ylabel("Count")
ax1.legend(title="Car Availability", loc='upper center', bbox_to_anchor=(0.5, -0.15), ncol=2)

# Plot 2: Stacked bar for Weekend Commuter
weekend_counts = train_set.groupby(['transport_mode', 'weekend_commuter']).size().unstack(fill_value=0)
weekend_counts = weekend_counts.loc[order]
ax2 = fig.add_subplot(gs[0, 1])
bottom = None
for i, col in enumerate(weekend_counts.columns):
    ax2.bar(weekend_counts.index, weekend_counts[col], bottom=bottom, label=col, color=cbPalette[i % len(cbPalette)])
    bottom = weekend_counts[col] if bottom is None else bottom + weekend_counts[col]
ax2.set_title("With Weekend Commuter Status")
ax2.set_xlabel("Transport Mode")
ax2.set_ylabel("")
ax2.legend(title="Weekend Commuter", loc='upper center', bbox_to_anchor=(0.5, -0.15), ncol=2)

# Adjust layout
fig.tight_layout(rect=[0, 0.05, 1, 1])
fig
```

:::

The bar charts show that car is the dominant mode of transport, especially among those who have access to a car. Public transit is used by both those with and without car availability, but it is more common among those without a car, as shown by a relatively larger orange bar. `Bicycle` and `Walking` are minority classes in the train set, as shown by the tiny bars in  @fig-eda-transport-mode-distribution-R and @fig-eda-transport-mode-distribution-py.

Most weekend commuters do not significantly change their transport mode. The overall proportions in the train set are largely dominated by the non-weekend commuters, shown by the orange bars. However, `Public Transit` has a noticeable number of weekend commuters, shown by the blue bar, which signals some weekend reliance on public transit. Among the `Bicycle` riders, the majority have access to a car and the majority of them are not weekend commuters, according to the stacked bar charts. 

The number of observations with `transport_mode=Walking` is very small in the train set, as shown by the minuscule bars corresponding to individuals with Walking as the primary mode of transport. As the distribution of categorical variables is not clear for `Walking`, let's filter the corresponding rows from train set. 

::: {.panel-tabset}

## R Code

```{.r}
# Filter the rows in train set with transport_mode = "Walking"
print(filter(train_set, transport_mode=="Walking"))
```

## Python Code

```{.python}
# Filter the rows in train set with transport_mode = "Walking"
print(train_set[train_set["transport_mode"] == "Walking"])
```

:::
::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
print(filter(train_set, transport_mode=="Walking"))
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
print(train_set[train_set["transport_mode"] == "Walking"])
```

:::

As shown in the outputs above, there are $4$ observations with `transport_mode=Walking` in the train set, which corresponds to $5\%$ of total observations in the train set. There is no clear dominance of any `has_car_available` and `weekend_commuter` categories for individuals with Walking as the primary mode of transport.

You can also visualize the relationship across other regressors in various types of figures for practice.

### Key Takeaways from EDA

Conducting exploratory data analysis (EDA) allows us to gain an initial understanding of the data and its underlying patterns before moving on to model building. Through EDA, we identify the types of variables present, examine their distributions, and uncover potential issues such as skewness, outliers, or multicollinearity. This process helps to highlight which variables might be strong predictors and which may require additional transformation or treatment. For example, a correlation between `has_car_available` and Car versus Public Transit as the transport mode signals that the availability of car might be a factor for choosing car over public transit. Similarly the median commute distance increases across `Bicycle`, `Car`, and `Public Transit` in the train set, which suggests that there might be an association between commute distance and primary mode of transport in the population of commuters.

It is important to remember that the insights gained from EDA are preliminary and primarily serve to inform further analysis. When we explore relationships between only two variables, we might overlook the influence of other factors, which could lead to misleading conclusions if taken in isolation. EDA is a crucial step for forming initial hypotheses and guiding decisions regarding data transformations, feature engineering, and the overall modelling strategy. With this foundation, we are better prepared to build a Multinomial Logistic regression model on data that has been carefully examined and understood.

## Data Modelling

After conducting [EDA](http://localhost:54321/book/14-multinomial-logistic.html#exploratory-data-analysis), we transition to the modelling stage, where we apply a structured approach to uncover relationships between variables and predict outcomes. In this section, we focus on Ordinary Least Squares (OLS) regression, a widely used statistical technique for modelling linear relationships.

A Multinomial Logistic Regression model is a suitable approach for our statistical inquiries, given that `transport_mode` is a categorical and unordered outcome variable subject to the numerical regressors `commute_distance` and `age`, and binary categorical regressors `has_car_available` and `weekend_commuter`. 

### Defining modelling parameters

Once we select Multinomial Logistic regression, we define the key modelling components: the response variable (dependent variable) and the regressors (independent variables).

#### Response Variable (Y):

The response variable, also known as the dependent variable, represents the financial outcome we aim to explain:

- `transport_mode`: The dependent variable representing primary mode of transport.

#### Regressors (X):

Each regressor is chosen based on its theoretical and statistical relevance in explaining the choice of primary transport mode:

- `commute_distance` (Continuous) – Represents the one-way commuting distance in kilometers.
- `age` (Continuous) – Represents the age of the individual in years.
- `has_car_available` (Binary) – Indicates whether the individual has access to a car
- `weekend_commuter` (Binary) – Indicates whether the individual commutes during weekends.

### The Random Component

The random component of Multinomial Logistic regression approach assumes a [Multinoulli distribution](https://www.britannica.com/science/multinomial-distribution) where $p_{i,1},p_{i,2},\dots,p_{i,m}$ are the probabilities of belonging to categories $1,2,\dots,m$ respectively; i.e., 
$$
\begin{align*}
P(Y_i=1)=p_{i,1},\newline
P(Y_i=2)=p_{i,2}, \newline
\dots\newline
\dots\newline
P(Y_i=m)=p_{i,m}
\end{align*}
$$ {#eq-multinomial-dist}
where
$$
\sum_{j = 1}^m p_{i,j} = p_{i,1} + p_{i,2} + \dots + p_{i,m} = 1.
$$

*A particular highlight is that the Bernoulli distribution is a special case of the Multinoulli distribution with two possible outcomes, i.e. $m=2$.*

Let $Y_i$ denote the transport mode of the $i$th commuter in the training data, for $i=1,2,...,n$, such that $Y_i$ can be one of `Public_Transit`, `Car`, `Bicycle`, or `Walking`. 

Let
$$
\mathbf{x}_i = (x_{i,1},...,x_{i,k}),
$$ 
denote the vector of observed regressor values for commuter $i$. For the largest candidate specification in this chapter, $k=4$, corresponding to commute distance, age, car availability, and weekend commuter status. 

The random component of Multinomial Logistic regression assumes

$$
Y_i \mid \mathbf{x}_i
\sim
\operatorname{Multinoulli}(p_{i,1}, ...,p_{i,m-1}),
$$

where

$$
p_{i,k}
=
\Pr(Y_i=k\mid\mathbf{x}_i)
$$
is the **conditional probability** that $Y_i$ belongs to category $k$, given the regressor values $\mathbf{x}_i$. Consequently,
$$\Pr(Y_i=m\mid\mathbf{x}_i)=1-\Sigma_{j=1}^{m-1}p_{i,j}$$
As $0\leq p_{i,k}\leq1$ and $\Sigma_{j=1}^mp_{i,j}=1$, it follows that the sum of parameters, $\Sigma_{j=1}^{m-1}p_{i,j}$ is a real number between $0$ and $1$. 

The random component tells us what kind of response is being modelled, but it does not yet explain why two commuters with different characteristics may have different values of $p_{i,1}, ...,p_{i,m-1}$.That role belongs to the systematic component.

### The Systematic Component

Like [Binary Logistic regression](https://alexrod61.github.io/regression-cookbook/book/08-binary-logistic.html), the systematic component combines the regressors through a **linear predictor**. With $k$ regressors,

$$
\eta_i^{(j,1)}
=
\beta_0
+
\beta_1x_{i,1}
+
\beta_2x_{i,2}
+
\cdots
+
\beta_kx_{i,k}.
$${#eq-systematic-component}

Equivalently,

$$
\eta_i
=
\beta_0
+
\mathbf{x}_i^\top
\boldsymbol{\beta},
$$

where

$$
\boldsymbol{\beta}
=
\left(
\beta_1,
\beta_2,
\ldots,
\beta_k
\right)^\top.
$$

Thus:

- $\eta_i$ is the **linear predictor** for commuter $i$;
- $\beta_0$ is the intercept;
- $\beta_1,\ldots,\beta_k$ are fixed but unknown regression coefficients;
- $x_{i,1},\ldots,x_{i,k}$ are the observed regressor values for commuter $i$;
- $\mathbf{x}_i$ collects those $k$ regressor values; and
- $\boldsymbol{\beta}$ collects the corresponding $k$ regression coefficients.

Some texts absorb the intercept into the vector notation by adding a leading $1$ to the regressor vector. We keep $\beta_0$ visible here so that $\mathbf{x}_i$ remains consistent with the regressor-vector notation introduced earlier in the chapter. 

The word **linear** in *linear predictor* refers to the way the regression coefficients enter $\eta_i$. It does **not** mean that the default probability itself is a straight-line function of the regressors.

::: {.Heads-up}
:::: {.Heads-up-header}
Heads-up on what is linear in Multinomial Logistic regression!
::::
:::: {.Heads-up-container}
The systematic component for each non-baseline category $j$ is $\eta_i^{(j,1)}$. In other words,

$$
\eta_i
=
\beta_0
+
\mathbf{x}_i^\top\boldsymbol{\beta}
$$

is linear in the regression coefficients, while

$$
\pi_i
=
\Pr(Y_i=1\mid\mathbf{x}_i)
$$

lives on the probability scale.

![Image by [*Manfred Steger*](https://pixabay.com/users/manfredsteger-1848497/) via [*Pixabay*](https://pixabay.com/vectors/pixel-cells-emotion-bored-6230197/).](img/question-2.png){width="300"}

Therefore, we still need a link function that connects the unrestricted real-valued linear predictor $\eta_i$ to a valid probability between $0$ and $1$.
::::
:::

This separation between the linear predictor and the response-scale probability is what allows a GLM to retain a regression structure without forcing the conditional probability itself to be linear.


### The Logit Link Functions

Like [Binary Logistic regression](https://alexrod61.github.io/regression-cookbook/book/08-binary-logistic.html), the Multinomial Logistic regression also models the **logits**, which are logarithm of the odds. However, only one logit will not be enough anymore. Recall that we can capture the odds between two categories with a single logit function. What about adding some other ones?

Here is what we can do:

1. Pick one of the categories to be the baseline.
2. For each of the other categories, we model the logarithm of the odds to the baseline category.

**Notation:** To avoid lengthy formulas, we introduce compact notation that will be used consistently throughout this chapter. In expressions such as $P(Y_i = 1 \mid x_{i,1}, ..., x_{i,K})$, the conditioning is on all regressors. Instead of repeatedly writing all regressors, we define:  

$$
\mathbf{x}_i = (x_{i,1},...,x_{i,K}),
$$ 

where $\mathbf{x}_i$ is a vector of size $k$ that bundles all $k$ regressors into a single object, so we can refer to them all at once instead of listing each one. For example,

$$
P(Y_i = 1 \mid \mathbf{x}_i).
$$
This notation simplifies expressions and improves readability.

Now, **what is the mathematics for the general case with $m$ response categories and $K$ regressors?** For the $i$th observation, we end up with a system of $m - 1$ **logit link** functions:

$$
\begin{gathered}
\eta_i^{(2,1)} = \log\left[\frac{P(Y_i = 2\mid \mathbf{x}_i)}{P(Y_i = 1 \mid \mathbf{x}_i)}\right] = \beta_0^{(2,1)} + \beta_1^{(2,1)} x_{i, 1} + \beta_2^{(2,1)} x_{i, 2} + \ldots + \beta_K^{(2,1)} x_{i, k} \\
\eta_i^{(3,1)} = \log\left[\frac{P(Y_i = 3\mid \mathbf{x}_i)}{P(Y_i = 1 \mid \mathbf{x}_i)}\right] = \beta_0^{(3,1)} + \beta_1^{(3,1)} x_{i, 1} + \beta_2^{(3,1)} x_{i, 2} + \ldots + \beta_K^{(3,1)} x_{i, k} \\
\vdots \\
\eta_i^{(m,1)} = \log\left[\frac{P(Y_i = m\mid \mathbf{x}_i)}{P(Y_i = 1 \mid \mathbf{x}_i)}\right] = \beta_0^{(m,1)} + \beta_1^{(m,1)} x_{i, 1} + \beta_2^{(m,1)} x_{i, 2} + \ldots + \beta_K^{(m,1)} x_{i, k}.
\end{gathered}
$$ {#eq-multinomial}

Note that the superscript $(j, 1)$ in @eq-multinomial indicates that the equation is on level $j$ (for $j = 2, \dots, m$) with respect to level $1$. Furthermore, the regression coefficients are different for each link function.

Each of the logit-linear functions in @eq-multinomial writes the log of odds between each category $j=2,\dots,m$ and the baseline category $j=1$ as a linear combination of the regressors. To compare between the categories $j=2,\dots,m$, we can simply deduct one equation from another, e.g.,
$$
\begin{gather*} \label{eq:multinomial-deduct}
\eta_i^{(2,1)} - \eta_i^{(3,1)} = \log\left[\frac{P(Y_i = 2\mid \mathbf{x}_i)}{P(Y_i = 1 \mid \mathbf{x}_i)}\right] - \log\left[\frac{P(Y_i = 3\mid \mathbf{x}_i)}{P(Y_i = 1 \mid \mathbf{x}_i)}\right] \\
= \log\left[\frac{P(Y_i = 3\mid \mathbf{x}_i)}{P(Y_i = 2 \mid \mathbf{x}_i)}\right] 
= (\beta_0^{(2,1)}-\beta_0^{(3,1)}) + (\beta_1^{(2,1)}-\beta_1^{(3,1)}) x_{i, 1} + (\beta_2^{(2,1)}-\beta_2^{(3,1)}) x_{i, 2} + \ldots
\end{gather*}
$$

With some algebraic manipulation, we can show that the probabilities $p_{i,1}, p_{i,2}, \dots, p_{i,m}$ of $Y_i$ belonging to categories $1, 2, \dots, m$ are:

$$ 
\begin{gathered}
p_{i,1} = P(Y_i = 1 \mid \mathbf{x}_i) = \frac{1}{1 + \sum_{j = 2}^m \exp \big( \eta_i^{(j,1)} \big)} \\
p_{i,2} = P(Y_i = 2 \mid \mathbf{x}_i) = \frac{\exp \big( \eta_i^{(2,1)} \big)}{1 + \sum_{j = 2}^m \exp \big( \eta_i^{(j,1)} \big)} \\
\vdots \\
p_{i,m} = P(Y_i = m \mid \mathbf{x}_i) = \frac{\exp \big( \eta_i^{(m,1)} \big)}{1 + \sum_{j = 2}^m \exp \big( \eta_i^{(j,1)} \big)}.
\end{gathered}
$$ {#eq-prob-multinomial}

***Note: The equations above are Softmax functions. The Multinomial Logistic regression is theoretically equivalent to a one-layer neural network with a Softmax activation function, where we input $X$, output probability $p_{i,j}$, and minimize the cross-entropy loss function to train the neural network to learn parameters $\beta$.***

***If we add all $m$ probabilities in @eq-prob-multinomial, the sum is equal to $1$ for the $i$th observation. This is particularly important when we want to use this model for making predictions in classification matters.***

Let's set up a Multinomial Logistic regression model with `transport_mode` as the response with four classes: `Car`, `Public Transit`, `Bicycle`, and `Walking`, subject to the continuous regressors `commute_distance` and `age`, denoted as $x_{i,\texttt{distance}}$ and $x_{i,\texttt{age}}$ respectively, and binary regressors `has_car_available` and `weekend_commuter`, with the dummy variables of class `Yes` denoted as $Z_{i,\texttt{car}}$ and $Z_{i,\texttt{weekend}}$ respectively. We will use `Car` as the baseline level for `transport_mode`. The equation of each link function for this multinomial model is given below, such that each link function has its own intercept and a set of regression coefficients.

$$
\begin{align*}
&\eta_i^{(\texttt{Public},\texttt{Car})} = \log\left[\frac{P(Y_i = \texttt{Public} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}{P(Y_i = \texttt{Car} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}\right] \\
&= \beta_0^{(\texttt{Public},\texttt{Car})} + \beta_1^{(\texttt{Public},\texttt{Car})} x_{i, \texttt{distance}} + \beta_2^{(\texttt{Public},\texttt{Car})} x_{i, \texttt{age}} + \beta_3^{(\texttt{Public},\texttt{Car})} Z_{i, \texttt{car}} + \beta_4^{(\texttt{Public},\texttt{Car})} Z_{i,\texttt{weekend}}, \\[0.6em]
&\eta_i^{(\texttt{Bike},\texttt{Car})} = \log\left[\frac{P(Y_i = \texttt{Bike} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}{P(Y_i = \texttt{Car} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}\right] \\
&= \beta_0^{(\texttt{Bike},\texttt{Car})} + \beta_1^{(\texttt{Bike},\texttt{Car})} x_{i, \texttt{distance}} + \beta_2^{(\texttt{Bike},\texttt{Car})} x_{i, \texttt{age}} + \beta_3^{(\texttt{Bike},\texttt{Car})} Z_{i, \texttt{car}} + \beta_4^{(\texttt{Bike},\texttt{Car})} Z_{i,\texttt{weekend}}, \\[0.6em]
&\eta_i^{(\texttt{Walk},\texttt{Car})} = \log\left[\frac{P(Y_i = \texttt{Walk} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}{P(Y_i = \texttt{Car} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}\right] \\
&= \beta_0^{(\texttt{Walk},\texttt{Car})} + \beta_1^{(\texttt{Walk},\texttt{Car})} x_{i, \texttt{distance}} + \beta_2^{(\texttt{Walk},\texttt{Car})} x_{i, \texttt{age}} + \beta_3^{(\texttt{Walk},\texttt{Car})} Z_{i, \texttt{car}} + \beta_4^{(\texttt{Walk},\texttt{Car})} Z_{i,\texttt{weekend}} .
\end{align*}
$$

Taking exponential on both sides of the model equations gives the ratio of the probability of each category `Public Transit`, `Bicycle`, and `Walking` over the probability of the baseline level `Car`:

$$
\begin{align*}
&\frac{P(Y_i = \texttt{Public} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}{P(Y_i = \texttt{Car} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})} = e^{\eta_i^{(\texttt{Public},\texttt{Car})}}\\
&= e^{\beta_0^{(\texttt{Public},\texttt{Car})}} \, e^{\beta_1^{(\texttt{Public},\texttt{Car})} x_{i,\texttt{distance}}} \, e^{\beta_2^{(\texttt{Public},\texttt{Car})} x_{i,\texttt{age}}} \, e^{\beta_3^{(\texttt{Public},\texttt{Car})} Z_{i, \texttt{car}}} \, e^{\beta_4^{(\texttt{Public},\texttt{Car})} Z_{i,\texttt{weekend}}}\\[12pt]
&\frac{P(Y_i = \texttt{Bike} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}{P(Y_i = \texttt{Car} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})} = e^{\eta_i^{(\texttt{Bike},\texttt{Car})}}\\
&= e^{\beta_0^{(\texttt{Bike},\texttt{Car})}} \, e^{\beta_1^{(\texttt{Bike},\texttt{Car})} x_{i,\texttt{distance}}} \, e^{\beta_2^{(\texttt{Bike},\texttt{Car})} x_{i,\texttt{age}}} \, e^{\beta_3^{(\texttt{Bike},\texttt{Car})} Z_{i, \texttt{car}}} \, e^{\beta_4^{(\texttt{Bike},\texttt{Car})} Z_{i,\texttt{weekend}}}\\[12pt]
&\frac{P(Y_i = \texttt{Walk} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}{P(Y_i = \texttt{Car} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})} = e^{\eta_i^{(\texttt{Walk},\texttt{Car})}}\\
&= e^{\beta_0^{(\texttt{Walk},\texttt{Car})}} \, e^{\beta_1^{(\texttt{Walk},\texttt{Car})} x_{i,\texttt{distance}}} \, e^{\beta_2^{(\texttt{Walk},\texttt{Car})} x_{i,\texttt{age}}} \, e^{\beta_3^{(\texttt{Walk},\texttt{Car})} Z_{i, \texttt{car}}} \, e^{\beta_4^{(\texttt{Walk},\texttt{Car})} Z_{i,\texttt{weekend}}}
\end{align*}
$$

Finally, the probabilities of $Y_i$ belonging to categories `Car`, `Public Transit`, `Bicycle`, and `Walking` respectively are given by the following expressions.

$$
\begin{align*}
P(Y_i = \texttt{Car} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}}) &= \frac{1}{1 + e^{\eta_i^{(\texttt{Public},\texttt{Car})}} + e^{\eta_i^{(\texttt{Bike},\texttt{Car})}} + e^{\eta_i^{(\texttt{Walk},\texttt{Car})}}} \\
P(Y_i = \texttt{Public} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}}) &= \frac{e^{\eta_i^{(\texttt{Public},\texttt{Car})}}}{1 + e^{\eta_i^{(\texttt{Public},\texttt{Car})}} + e^{\eta_i^{(\texttt{Bike},\texttt{Car})}} + e^{\eta_i^{(\texttt{Walk},\texttt{Car})}}} \\
P(Y_i = \texttt{Bike} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}}) &= \frac{e^{\eta_i^{(\texttt{Bike},\texttt{Car})}}}{1 + e^{\eta_i^{(\texttt{Public},\texttt{Car})}} + e^{\eta_i^{(\texttt{Bike},\texttt{Car})}} + e^{\eta_i^{(\texttt{Walk},\texttt{Car})}}} \\
P(Y_i = \texttt{Walk} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}}) &= \frac{e^{\eta_i^{(\texttt{Walk},\texttt{Car})}}}{1 + e^{\eta_i^{(\texttt{Public},\texttt{Car})}} + e^{\eta_i^{(\texttt{Bike},\texttt{Car})}} + e^{\eta_i^{(\texttt{Walk},\texttt{Car})}}}
\end{align*}
$$ {#eq-probabilities}

### The Log-Likelihood Function

A Multinomial Logistic regression model treats the categories in the response variable as nominal, which means that there is no inherent ranking or ordering among them; therefore, each category is compared directly against a single baseline category. Recall from @eq-prob-multinomial that this comparison gives us the category-specific probabilities in closed form,

$$
p_{i,j} = P(Y_i = j \mid \mathbf{x}_i), \qquad j = 1, \ldots, m,
$$

where each non-baseline category $j$ enters through its own logit $\eta_i^{(j,1)}$ and the whole set is normalized by the softmax denominator, so that $\sum_{j=1}^m p_{i,j} = 1$. These probabilities are exactly what the likelihood is built from.

***Note: Changing the baseline category, e.g., switching from `Car` to `Public` would not change the estimated probability for each category for each individual.***

To estimate the model, we need a way to measure how well a given set of coefficients explains the observed data. The **likelihood function** does exactly this by expressing the probability of the data we actually observed as a function of the unknown coefficients, and finding the coefficient values that make this probability as large as possible.

We first encode each individual's observed response category using indicator variables. Following the notation in @eq-multinomial and @eq-prob-multinomial, let $y_{i,j} = 1$ if the $i$-th individual belongs to category $j$, and let $y_{i,j} = 0$ otherwise. Since each individual belongs to exactly one of the $m$ possible categories, the vector $\mathbf{y}_i = (y_{i,1}, y_{i,2}, \ldots, y_{i,m})$ contains exactly one entry equal to $1$, with all remaining entries equal to $0$. This representation is often called a one-hot encoding of the response category. The likelihood function for the Multinomial Logistic regression model is then

$$
\mathcal{L}(\boldsymbol{\beta}; \mathbf{y}) = \prod_{i=1}^n \prod_{j=1}^m p_{i,j}^{y_{i,j}},
$$ {#eq-likelihood-multinomial}

where $n$ is the number of individuals in the sample, $m$ is the number of categories, and $p_{i,j}$ is the model-implied softmax probability that individual $i$ (with regressor values $x_{i,1}, \ldots, x_{i,K}$) falls in category $j$. The unknown parameters are collected in $\boldsymbol{\beta}$, the stack of all reference-category coefficients $\beta_k^{(j,1)}$ for $j = 2, \ldots, m$ and $k = 0, 1, \ldots, K$, such that it consists of an intercept and $K$ slopes for each of the $m-1$ non-baseline categories.

The two products have intuitive roles. The **inner product** over $j$ runs across the $m$ categories for a single individual; because of the indicator $y_{i,j}$ in the exponent, $\prod_{j=1}^m p_{i,j}^{y_{i,j}}$ collapses to just $p_{i,j}$ for the actual observed category of $i$th observation. The **outer product** over $i$ then multiplies these per-individual probabilities together across all $n$ people, which is valid because the observations are assumed independent. Overall, this is a **multinomial likelihood** evaluated at the softmax category probabilities.

Taking the logarithm turns the products into sums, which is far easier to differentiate and optimize, and yields the **log-likelihood function**:

$$
\ell(\boldsymbol{\beta}; \mathbf{y}) = \sum_{i=1}^n \sum_{j=1}^m y_{i,j} \log p_{i,j}.
$$ {#eq-log-likelihood-multinomial}

Substituting the softmax probabilities from @eq-prob-multinomial and using the reference-category constraint $y_{i,1} = 1 - \sum_{j=2}^m y_{i,j}$, the log-likelihood can be written explicitly in terms of the coefficients as

$$
\ell(\boldsymbol{\beta}) = \sum_{i=1}^n \left[ \sum_{j=2}^m y_{i,j} \sum_{k=0}^K x_{i,k} \beta_k^{(j,1)} - \log\left( 1 + \sum_{j=2}^m e^{\sum_{k=0}^K x_{i,k} \beta_k^{(j,1)}} \right) \right],
$$ {#eq-log-likelihoood}

where $x_{i,0} = 1$ corresponds to the intercept in the linear model.

Maximizing $\ell$ gives the estimates $\hat{\boldsymbol{\beta}}$, the parameters under which the observed data is the most plausible. This method of finding the best-fit parameters is known as is known as the **maximum-likelihood estimation**. There is, however, no closed-form algebraic expression for the maximizer: the category probabilities $p_{i,j}$ enter the log-likelihood non-linearly through the softmax function, so setting the derivatives to zero does not yield equations we can solve directly. Instead, we rely on **iterative numerical optimization**, which starts from an initial guess and refines the estimates step by step until the log-likelihood can no longer be improved. The standard choice is the **Newton–Raphson** method, described in the tip box below; the variant most statistical software uses replaces the observed Hessian with its expected value (the Fisher information), giving the **Fisher scoring** algorithm, which is equivalent to **iteratively reweighted least squares (IRLS)**. Additional estimation details, including the generalization from one to $n_i$ observations per population $i$, can be found in Section 2.2 of [Czepiel (2002)](https://czep.net/stat/mlelr.pdf).

::: {.Tip}
::::{.Tip-header}
Tip on the Newton–Raphson Method
::::
::::{.Tip-container}
The parameters we ultimately estimate are the reference-category coefficients, collected into the stacked vector $\boldsymbol{\beta} \in \mathbb{R}^q$ with $q = (K+1)(m-1)$ (an intercept and $K$ slopes for each of the $m-1$ non-baseline categories). Since the log-likelihood has no closed-form maximizer, the Newton–Raphson procedure estimates $\boldsymbol{\beta}$ by iterating with the score function and Hessian matrix of the log-likelihood, in a compact matrix form known as multinomial iteratively reweighted least squares (IRLS).

**Score function:** For category $j$ and regressor $k$, the corresponding component of the score vector $S(\boldsymbol{\beta}) \in \mathbb{R}^q$ (the first-order partial derivative of $\ell$) is

$$
\frac{\partial \ell(\boldsymbol{\beta})}{\partial \beta_k^{(j,1)}} = \sum_{i=1}^n \left( y_{i,j} - p_{i,j} \right) x_{i,k}.
$$

Each component compares the observed indicator $y_{i,j}$ with the model-implied probability $p_{i,j}$, weighted by the regressor value $x_{i,k}$; at the maximum, these differences balance out so that the score is zero.

**Hessian matrix:** The Hessian $H(\boldsymbol{\beta}) \in \mathbb{R}^{q \times q}$ collects the second-order partial derivatives of $\ell$. For the row indexed by category $j$ and regressor $k$, and the column indexed by category $j'$ and regressor $k'$, its entry is

$$
\frac{\partial^2 \ell(\boldsymbol{\beta})}{\partial \beta_k^{(j,1)} \partial \beta_{k'}^{(j',1)}} =
\begin{cases}
-\sum_{i=1}^n x_{i,k}\, p_{i,j}(1 - p_{i,j})\, x_{i,k'} & \text{if } j' = j, \\
\phantom{-}\sum_{i=1}^n x_{i,k}\, p_{i,j}\, p_{i,j'}\, x_{i,k'} & \text{if } j' \neq j,
\end{cases}
$$

for $j, j' = 2, \ldots, m$ and $k, k' = 0, \ldots, K$. As the non-baseline categories are coupled through the shared softmax denominator, the Hessian is a dense $q \times q$ matrix rather than block-diagonal. For a thorough treatment of the Hessian in the context of maximum likelihood estimation, refer to Chapter 4 of @agresti2013.

**Iteration:** Collecting the score and Hessian above, the Newton–Raphson update at step $t$ is

$$
\boldsymbol{\beta}^{(t+1)} \leftarrow \boldsymbol{\beta}^{(t)} - \left[ H\!\left( \boldsymbol{\beta}^{(t)} \right) \right]^{-1} S\!\left( \boldsymbol{\beta}^{(t)} \right),
$$

initialized at $\boldsymbol{\beta}^{(0)} = \mathbf{0}$ (or a data-driven starting point) and repeated until changes in $\ell$ and $\boldsymbol{\beta}$ fall below a convergence tolerance.

The full Newton–Raphson/IRLS procedure is:

1. Choose the baseline category and initialize $\boldsymbol{\beta}^{(0)} = \mathbf{0}$ (stacked over all $(j,k)$).
2. Repeat until convergence:
   - Compute the logits $\eta_i^{(j,1)}$ and category probabilities $p_{i,j}$ for all individuals $i$ and categories $j$, and evaluate the log-likelihood $\ell(\boldsymbol{\beta}^{(t)})$.
   - Form the stacked score $S(\boldsymbol{\beta}^{(t)})$ and Hessian $H(\boldsymbol{\beta}^{(t)})$ using the expressions above.
   - Update $\boldsymbol{\beta}^{(t+1)}$ using the Newton–Raphson step above.
   - Check the convergence criteria.
::::
:::

## Estimation {#sec-estimation}

With the data modelling stage completed, we now move to estimation, where we fit the Multinomial Logistic regression model to the data and obtain numerical estimates for the regression coefficients. These estimates quantify how much each regressor contributes to the response variable, allowing us to measure their individual effects on primary mode of transport.

The goal of estimation is to determine the coefficient estimates through **maximum likelihood estimation**, which gives the coefficients estimates for which the observed training data is the most probable under a **conditional Multinomial distribution** for `transport_mode`. This provides a mathematical framework for analyzing the commuting behaviors and primary modes of transport. 

### Fitting the Model

We now turn to the question of how the parameters of the Multinomial Logistic regression model are estimated from data. We summarize the key estimation steps in mathematical terms below, working with one observation at a time before generalizing to the full sample.

Recall that the model for a response $Y_i$ with $m$ categories and $K$ regressors is given by

$$
\begin{aligned}
\log \left( \frac{P(Y_i = j \mid \mathbf{x}_i)}{P(Y_i = 1 \mid \mathbf{x}_i)} \right) &= \beta_{0}^{(j,1)}\\
&\quad+\beta_{1}^{(j,1)}x_{i,1}\\
&\quad+\beta_{2}^{(j,1)} x_{i,2}\\
&\quad+ ... \\
&\quad+ \beta_{k}^{(j,1)} x_{i,k},\,\,\,\,j=2,...,m
\end{aligned}
$$ {#eq-multinomial-model}

where $\beta_{0}^{(2,1)}$, ..., $\beta_{0}^{(m,1)}$ are category-specific intercepts, and $\beta_{1}^{(k,1)}$, $\beta_{2}^{(j,1)}$, ..., $\beta_{k}^{(j,1)}$ are the slope coefficients corresponding to the $j$-th category of $Y$.

The model therefore comprises $(m-1)\times (k+1)$ parameters: $m-1$ intercepts and $(m-1)\times k$ regression coefficients. 

We now fit the Multinomial Logistic regression model on `train_set` using the `nnet` package in `R` and the `statsmodel` package in `Python`. 

::: {.panel-tabset}

## R Code

```{.r}
# Set the baseline model
train_set$transport_mode <- relevel(train_set$transport_mode, ref = "Car")

# Fit Multinomial Logistic regression
multinom_fit <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = train_set)

# Print the fitted model
print(summary(multinom_fit))
```

## Python Code

```{.python}
# Factorize the response and set the baseline model
train_set["transport_mode"] = pd.Categorical(
    train_set["transport_mode"],
    categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)

# Design matrix and response
y = train_set["transport_mode"]
X = train_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]

# Add intercept
X = sm.add_constant(X)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_numeric = X.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes

# Fit Multinomial Logistic regression
model = sm.MNLogit(y, X_numeric)
multinom_fit = model.fit()

# Print the fitted model
print(multinom_fit.summary())
```

:::

::: {.panel-tabset}

### R Output

```{r}
#| echo: false
#| message: false
#| warning: false
train_set$transport_mode <- relevel(train_set$transport_mode, ref = "Car")
multinom_fit <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = train_set)
#print(t(round(summary(multinom_fit)$coefficients, 3)))
print(summary(multinom_fit))
```

### Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
train_set["transport_mode"] = pd.Categorical(
    train_set["transport_mode"],
    categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)
y = train_set["transport_mode"]
X = train_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]
X = sm.add_constant(X)
X_numeric = X.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes
model = sm.MNLogit(y, X_numeric)
multinom_fit = model.fit()
print(multinom_fit.summary())
```

:::

The output reports the estimated parameters. For each non-baseline category $j$, a category-specific intercept $\hat{\beta}_{0}^{(j,1)}$ and the slope coefficients $\hat{\beta}_{1}^{(j,1)}, \ldots, \hat{\beta}_{k}^{(j,1)}$ (one per regressor) are reported, each with its standard error. Since `Car` was set as the baseline category, the output contains one such set of estimates for each of `Public Transit`, `Bicycle`, and `Walking`, and all estimates are interpreted relative to `Car`; the coefficients of the baseline category itself are fixed at zero for identifiability and therefore do not appear in the output. With $m = 4$ response categories and $K = 4$ regressors, the fitted model comprises $(m - 1) \times (k + 1) = 15$ estimated parameters, which is consistent with the parameter count derived above.

In the case study model with regressors $x_{i,\texttt{distance}}$, $x_{i,\texttt{age}}$, $Z_{i,\texttt{car}}$, and $Z_{i,\texttt{weekend}}$, the fitted model can be written as

$$
\begin{aligned}
\log \left( \frac{P(Y_i = j \mid \mathbf{x}_i)}{P(Y_i = \text{Car} \mid \mathbf{x}_i)} \right) &= \hat{\beta}_{0}^{(k,\text{Car})}\\
&\quad + \hat{\beta}_{1}^{(j,\text{Car})} x_{i,\texttt{distance}}\\
&\quad + \hat{\beta}_{2}^{(j,\text{Car})} x_{i,\texttt{age}}\\
&\quad + \hat{\beta}_{3}^{(j,\text{Car})} Z_{i,\texttt{car}}\\
&\quad + \hat{\beta}_{4}^{(j,\text{Car})} Z_{i,\texttt{weekend}},
\end{aligned}
$$

where $k$ can be one of the non-baseline levels: `Public Transit`, `Bicycle`, or `Walking`.

In Multinomial Logistic regression, we model the probability of each outcome category given regressors. The key assumption is that we have enough data for each category to estimate the model parameters reliably. 
If one category is much more frequent than the others (i.e., unbalanced outcomes), the model may:

1. Bias coefficients toward the majority class because the likelihood function is dominated by majority observations.
2. Underestimate uncertainty for rare categories (small sample size → large variance).
3. Yield poor predictive performance for minority classes (the model tends to predict the majority class).
4. In extreme cases, parameters for rare categories may not converge or may give infinite estimates (complete separation).

For example, in our training data, only $5\%$ of the observations are `Walking`, so the model has little incentive to adjust parameters for `Walking`. There are two common approaches: 

1. Downsampling: Reduce the majority class size so that each class is roughly balanced. This risks throwing away data, but helps balance influence.
2. Upsampling: Duplicate (or synthetically generate) minority class cases until classes are balanced. This risks overfitting, but improves minority estimation.

Let's check the class counts of the outcome below.

::: {.panel-tabset}

## R Code

```{.r}
cat("Class counts before balancing:", "\n")
print(table(train_set$transport_mode))
```

## Python Code

```{.python}
print("Class counts before balancing:", "\n")
print(y.value_counts())
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
cat("Class counts before balancing:", "\n")
print(table(train_set$transport_mode))
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
print("Class counts before balancing:", "\n")
print(y.value_counts())
```

:::

To handle the imbalanced groups, we resample the train sets to balance out the groups to eliminate the bias in modelling. The fitted model on upsampled train set is below. The upsampling is performed once in `R` with `caret::upSample`, which retains every original training row and tops up only the classes below the majority size with duplicated rows sampled with replacement. The resulting `train_set_balanced` is then shared with `Python` through `reticulate`, so that both languages fit the model on the identical balanced train set and produce matching estimates. The upsampling can be performed in `Python` with `RandomOverSampler` from the imbalanced-learn (`imblearn`) package. For more details on the `RandomOverSampler` function, refer to its [documentation](https://imbalanced-learn.org/stable/references/generated/imblearn.over_sampling.RandomOverSampler.html).

::: {.panel-tabset}

## R Code

```{.r}
# Upsample the outcome classes
set.seed(352) # set a seed for sampling
train_set_balanced <- upSample(x = train_set[, c("age","commute_distance","has_car_available", "weekend_commuter")],
                  y = train_set$transport_mode,
                  yname = "transport_mode")
# Fit the multinomial logit regression
multinom_fit_balanced <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = train_set_balanced)
print(summary(multinom_fit_balanced))
```

## Python Code

```{.python}
# The upsampled train set is created once in R (caret::upSample; see the
# R tab) and pulled into Python via reticulate's `r` object, so both languages
# fit the model on the identical balanced train set.
train_set_balanced = r.train_set_balanced.reset_index(drop=True)

X_bal = train_set_balanced[["age", "commute_distance", "has_car_available", "weekend_commuter"]]

y_bal = pd.Series(
    pd.Categorical(
        train_set_balanced["transport_mode"],
        categories=["Car", "Bicycle", "Public_Transit", "Walking"],
        ordered=False
    ),
    name="transport_mode"
)
X_bal = sm.add_constant(X_bal)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_bal_num = X_bal.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_bal_num[col] = X_bal_num[col].cat.codes

# Fit the multinomial logit regression
model_balanced = sm.MNLogit(y_bal, X_bal_num)
multinom_fit_balanced = model_balanced.fit()
print(multinom_fit_balanced.summary())
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
set.seed(352)
train_set_balanced <- upSample(x = train_set[, c("age","commute_distance","has_car_available", "weekend_commuter")],
                  y = train_set$transport_mode,
                  yname = "transport_mode")
multinom_fit_balanced <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = train_set_balanced)
print(summary(multinom_fit_balanced))
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
# The upsampled train set is created once in R (caret::upSample; see the
# R tab) and pulled into Python via reticulate's `r` object, so both languages
# fit the model on the identical balanced train set.
train_set_balanced = r.train_set_balanced.reset_index(drop=True)

X_bal = train_set_balanced[["age", "commute_distance", "has_car_available", "weekend_commuter"]]

y_bal = pd.Series(
    pd.Categorical(
        train_set_balanced["transport_mode"],
        categories=["Car", "Bicycle", "Public_Transit", "Walking"],
        ordered=False
    ),
    name="transport_mode"
)
X_bal = sm.add_constant(X_bal)

X_bal_num = X_bal.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_bal_num[col] = X_bal_num[col].cat.codes

model_balanced = sm.MNLogit(y_bal, X_bal_num)
multinom_fit_balanced = model_balanced.fit()
print(multinom_fit_balanced.summary())
```

:::

As a reminder, we just completed the estimation step of the [data science workflow](https://alexrod61.github.io/regression-cookbook/book/01-intro.html#sec-ds-workflow). For the **inferential inquiry**, we should not use the same dataset both to explore the model structure and to produce the final inferential claims. Recall that the train set is used for model development and goodness-of-fit assessment. Final inferential outputs will be produced later in @sec-regressors after the model-building workflow has been completed.

## Checking Goodness of Fit and Model Stability

Now let's assess how well the fitted Multinomial Logistic regression explains and predicts the observed transport modes on the train set. We consider two complementary perspectives: **likelihood-based** measures that evaluate how much of the outcome's variation the model explains on the log-likelihood scale (similar to @sec-poisson-goodness-of-fit in [Classical Poisson Regression](https://alexrod61.github.io/regression-cookbook/book/10-classical-poisson.html)), and **classification-based** measures that evaluate how accurately the model assigns observations to their correct transport mode.

### Convergence and Separation
A particularly important problem for convergence of the MLE solution is **separation**. 

::::: {.Heads-up}

:::: {.Heads-up-header}
Heads-up on complete and quasi-complete separation!
::::

:::: {.Heads-up-container}

Consider a Binary Logistic regression model with an intercept-augmented regressor vector $\widetilde{\mathbf{x}}_i$. **Complete separation** occurs when there is some nonzero vector $\mathbf{a}$ such that

$$
\widetilde{\mathbf{x}}_i^\top\mathbf{a}>0
\quad
\text{for every observation with }Y_i=1,
$$

and

$$
\widetilde{\mathbf{x}}_i^\top\mathbf{a}<0
\quad
\text{for every observation with }Y_i=0.
$$

In other words, a linear boundary in the regressor space perfectly separates all events from all non-events.

**Quasi-complete separation** occurs when the same inequalities can be made non-strict,

$$
\widetilde{\mathbf{x}}_i^\top\mathbf{a}\geq0
\quad\text{for }Y_i=1,
$$

and

$$
\widetilde{\mathbf{x}}_i^\top\mathbf{a}\leq0
\quad\text{for }Y_i=0,
$$

with at least some observations lying exactly on the separating boundary.

::::

:::::

Under complete separation, the Multinomial likelihood function can keep increasing as one or more coefficient magnitudes grow without bound. A finite ordinary MLE therefore does not exist. Under quasi-complete separation, the same instability occurs along at least one coefficient direction. As a result:

- coefficient estimates may become extremely large;
- standard errors based on the inverse Fisher information may become extremely large or unstable;
- fitted probabilities can be pushed increasingly close to $0$ or $1$;
- the Fisher-information matrix can become nearly singular; and
- ordinary Wald-type inference can become unreliable.

Software often alerts us to the problem. In `R`, common warnings include messages such as

```text
glm.fit: algorithm did not converge
```

or

```text
glm.fit: fitted probabilities numerically 0 or 1 occurred
```

depending on the severity of the problem. In `Python`'s [{statsmodels}](https://www.statsmodels.org/stable/index.html), a separated or nearly separated fit may generate a `PerfectSeparationWarning`, fail to converge, or produce extreme coefficient estimates and standard errors.

These warnings are useful, but diagnostics should not rely on warning text alone. Convergence, finite coefficients, finite standard errors, fitted probabilities, regressor overlap, and information-matrix stability should be considered together.

::: {.panel-tabset}

## **`R`**

```{r}
#| echo: true
#| message: false
#| warning: false

fitted_probs <- fitted(multinom_fit)

cat(sprintf(
  paste0(
    "Converged: %s\n",
    "All coefficients finite: %s\n",
    "All standard errors finite: %s\n",
    "All fitted probabilities finite: %s\n",
    "Fitted probabilities strictly inside (0, 1): %s\n",
    "Fitted-probability range: %.4f to %.4f\n"
  ),
  ifelse(isTRUE(multinom_fit$convergence == 0), "Yes", "No"),
  ifelse(all(is.finite(coef(multinom_fit))), "Yes", "No"),
  ifelse(all(is.finite(sqrt(diag(vcov(multinom_fit))))), "Yes", "No"),
  ifelse(all(is.finite(fitted_probs)), "Yes", "No"),
  ifelse(all(fitted_probs > 0 & fitted_probs < 1), "Yes", "No"),
  min(fitted_probs),
  max(fitted_probs)
))
```

## **`Python`**

```{python}
#| echo: true
#| message: false
#| warning: false

converged = bool(multinom_fit.mle_retvals["converged"])
coefs = np.asarray(multinom_fit.params, dtype=float)
ses = np.asarray(multinom_fit.bse, dtype=float)
fitted = np.asarray(multinom_fit.predict(), dtype=float)

yes_no = lambda flag: "Yes" if flag else "No"

print(
    f"Converged: {yes_no(converged)}\n"
    f"All coefficients finite: {yes_no(np.all(np.isfinite(coefs)))}\n"
    f"All standard errors finite: {yes_no(np.all(np.isfinite(ses)))}\n"
    f"All fitted probabilities finite: {yes_no(np.all(np.isfinite(fitted)))}\n"
    f"Fitted probabilities strictly inside (0, 1): "
    f"{yes_no(np.all((fitted > 0) & (fitted < 1)))}\n"
    f"Fitted-probability range: {fitted.min():.4f} to {fitted.max():.4f}"
)
```

:::

The fitted model converges, returns finite coefficients and standard errors, and produces fitted probabilities strictly inside $(0,1)$. According to the plots in @sec-eda, there is overlap between the age range and the commute distance range among all four categories of transport mode: `Car`, `Public_Transit`, `Bicycle`, and `Walking`. These checks provide **no obvious evidence of complete or quasi-complete separation in this fitted model.**

### Likelihood-Based Measures

The **null deviance** measures how well the intercept-only model (no regressors, only a baseline log-odds for each non-reference category) fits the data:

$$
D_{\text{null}} = -2\,\ell(\hat{\boldsymbol{\beta}}_{0,\text{null}}),
$$

where $\hat{\boldsymbol{\beta}}_{0,\text{null}}$ are the category-specific intercepts estimated without any regressor.

The **residual deviance** is the analogous quantity for the fitted model:

$$
D_{\text{resid}} = -2\,\ell(\hat{\boldsymbol{\beta}}).
$$

A well-fitting model should have a substantially smaller residual deviance than the null deviance. The reduction $D_{\text{null}} - D_{\text{resid}}$ is the likelihood-ratio (LR) statistic, which under $H_0$ (all slope coefficients across all non-baseline categories are zero) follows a $\chi^2$ distribution with degrees of freedom equal to the number of slope parameters $k$:

$$
G^2 = D_{\text{null}} - D_{\text{resid}} = -2\bigl(\ell_{\text{null}} - \ell_{\text{fitted}}\bigr) \;\overset{H_0}{\sim}\; \chi^2(k).
$$

**McFadden's pseudo-$R^2$** expresses the proportional reduction in log-likelihood relative to the null model:

$$
R^2_{\text{McFadden}} = 1 - \frac{\ell(\hat{\boldsymbol{\beta}})}{\ell(\hat{\boldsymbol{\beta}}_{0,\text{null}})}.
$$

Unlike the usual R-squared in OLS regression, McFadden's pseudo-R-squared should not be interpreted as the proportion of variance explained. Instead, it measures improvement in log-likelihood relative to the null model. Its values are typically much smaller than regular R-squared values, and values around $0.2$–$0.4$ are often considered indicative of a good/decent model fit in discrete-response models.

**AIC** and **BIC** scores penalize the log-likelihood for model complexity:

$$
\text{AIC} = -2\,\ell(\hat{\boldsymbol{\beta}}) + 2p, \qquad \text{BIC} = -2\,\ell(\hat{\boldsymbol{\beta}}) + p\log n,
$$

where $p$ is the total number of estimated parameters (the category-specific intercepts and slopes across all non-baseline categories) and $n$ is the sample size. Lower AIC and BIC scores indicate better fit, and they are most useful for comparing competing models (e.g. this imbalanced fit versus the upsampled fit).

We compute all of these measures below.

::: {.panel-tabset}

## R Code

```{.r}
library(nnet)

# Null model: intercept only, no regressors
multinom_null <- multinom(transport_mode ~ 1, data = train_set)

ll_fitted <- logLik(multinom_fit)
ll_null   <- logLik(multinom_null)

null_deviance  <- -2 * as.numeric(ll_null)
resid_deviance <- -2 * as.numeric(ll_fitted)
mcfadden_r2    <- 1 - as.numeric(ll_fitted) / as.numeric(ll_null)
lr_stat        <- as.numeric(-2 * (ll_null - ll_fitted))
lr_df          <- attr(ll_fitted, "df") - attr(ll_null, "df")
lr_pval        <- pchisq(lr_stat, df = lr_df, lower.tail = FALSE)

cat("Null deviance:      ", round(null_deviance,  2), "\n")
cat("Residual deviance:  ", round(resid_deviance, 2), "\n")
cat("McFadden R²:        ", round(mcfadden_r2, 4),   "\n")
cat("G²:                 ", round(lr_stat, 4),
    " (df =", lr_df, ", p =", formatC(lr_pval, format = "e", digits = 2), ")\n")
cat("AIC:                ", round(AIC(multinom_fit), 2), "\n")
cat("BIC:                ", round(BIC(multinom_fit), 2), "\n")
```

## Python Code

```{.python}
import numpy as np
from scipy.stats import chi2 as chi2dist

# The log-likelihood of the intercept-only (null) model is available
# directly on the fitted MNLogit result
ll_fitted = multinom_fit.llf
ll_null   = multinom_fit.llnull

null_deviance  = -2 * ll_null
resid_deviance = -2 * ll_fitted
mcfadden_r2    = 1 - ll_fitted / ll_null
lr_stat        = -2 * (ll_null - ll_fitted)
lr_df          = multinom_fit.df_model
lr_pval        = chi2dist.sf(lr_stat, lr_df)

print(f"Null deviance:      {null_deviance:.2f}")
print(f"Residual deviance:  {resid_deviance:.2f}")
print(f"McFadden R²:        {mcfadden_r2:.4f}")
print(f"G²:                 {lr_stat:.4f}  (df = {lr_df}, p = {lr_pval:.2e})")
print(f"AIC:                {multinom_fit.aic:.2f}")
print(f"BIC:                {multinom_fit.bic:.2f}")
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
library(nnet)

multinom_null <- multinom(transport_mode ~ 1, data = train_set)

ll_fitted <- logLik(multinom_fit)
ll_null   <- logLik(multinom_null)

null_deviance  <- -2 * as.numeric(ll_null)
resid_deviance <- -2 * as.numeric(ll_fitted)
mcfadden_r2    <- 1 - as.numeric(ll_fitted) / as.numeric(ll_null)
lr_stat        <- as.numeric(-2 * (ll_null - ll_fitted))
lr_df          <- attr(ll_fitted, "df") - attr(ll_null, "df")
lr_pval        <- pchisq(lr_stat, df = lr_df, lower.tail = FALSE)

cat("Null deviance:      ", round(null_deviance,  2), "\n")
cat("Residual deviance:  ", round(resid_deviance, 2), "\n")
cat("McFadden R²:        ", round(mcfadden_r2, 4),   "\n")
cat("G²:                 ", round(lr_stat, 4),
    " (df =", lr_df, ", p =", formatC(lr_pval, format = "e", digits = 2), ")\n")
cat("AIC:                ", round(AIC(multinom_fit), 2), "\n")
cat("BIC:                ", round(BIC(multinom_fit), 2), "\n")
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
import numpy as np
from scipy.stats import chi2 as chi2dist

ll_fitted = multinom_fit.llf
ll_null   = multinom_fit.llnull

null_deviance  = -2 * ll_null
resid_deviance = -2 * ll_fitted
mcfadden_r2    = 1 - ll_fitted / ll_null
lr_stat        = -2 * (ll_null - ll_fitted)
lr_df          = multinom_fit.df_model
lr_pval        = chi2dist.sf(lr_stat, lr_df)

print(f"Null deviance:      {null_deviance:.2f}")
print(f"Residual deviance:  {resid_deviance:.2f}")
print(f"McFadden R²:        {mcfadden_r2:.4f}")
print(f"G²:                 {lr_stat:.4f}  (df = {lr_df}, p = {lr_pval:.2e})")
print(f"AIC:                {multinom_fit.aic:.2f}")
print(f"BIC:                {multinom_fit.bic:.2f}")
```

:::

The residual deviance is smaller than the null deviance by `r round(lr_stat,4)` units. This reduction is captured by the likelihood-ratio statistic $G^2=$ `r round(lr_stat,4)` on `r lr_df` degrees of freedom ($p < 0.001$), providing statistical evidence that a model with age, commute distance, car availability, and weekend commuting status fits the training data better than the null (intercept-only) model. McFadden's $R^2$ (`r round(mcfadden_r2,4)`) is below the 0.2–0.4 range commonly associated with a good fit for discrete-response models, which indicates that these four regressors together explain only a modest share of variability in `transport_mode` on the log-likelihood scale, even though their joint contribution is highly significant. The AIC and BIC scores of `r round(AIC(multinom_fit), 2)` and `r round(BIC(multinom_fit), 2)` respectively describe the imbalanced training model, and they are useful later for comparing this fit against the upsampled model or other model specifications.

### Classification-Based Measures

While likelihood-based measures evaluate the model on the probability scale, classification-based measures assess how well the model assigns each observation to its predicted transport mode by taking the category with the highest predicted probability as the point prediction.

Let's visualize and summarize goodness of fit for the train set. 

::: {.panel-tabset}

## R Code

```{.r}
y_pred <- predict(multinom_fit, newdata = train_set)
mode_levels <- c("Car", "Bicycle", "Public_Transit", "Walking")
conf_matrix <- confusionMatrix(factor(y_pred, levels = mode_levels),
                              factor(train_set$transport_mode, levels = mode_levels))

# Visualise
cm_df <- as.data.frame(as.table(conf_matrix$table))
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

# Order axes so the main diagonal (correct matches) runs top-left to
# bottom-right: Predicted increases left -> right (Car -> Walking), and
# Actual increases top -> bottom (Car -> Walking). ggplot draws the first
# factor level at the bottom of the y-axis, so we reverse Actual's levels to
# place "Car" at the top.
cm_df$Predicted <- factor(cm_df$Predicted, levels = levels(cm_df$Predicted))
cm_df$Actual    <- factor(cm_df$Actual,    levels = rev(levels(cm_df$Actual)))

cm_df |>
  ggplot(aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  labs(title = "Confusion Matrix (Train Set)",
       x = "Predicted label", y = "True label") +
  theme_bw(base_size = 13)
```

## Python Code

```{.python}
y_pred = multinom_fit.predict(X_numeric).idxmax(axis=1)

# Confusion matrix
cm = confusion_matrix(y.cat.codes, y_pred, labels=range(len(y.cat.categories)))
disp = ConfusionMatrixDisplay(confusion_matrix=cm,
                              display_labels=y.cat.categories)
disp.plot(cmap="Blues")
disp.ax_.set_title("Confusion Matrix (Train Set)")
fig = disp.figure_
fig.tight_layout()
fig
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| label: fig-confusion-matrix-train-R
#| fig-cap: "Confusion matrix of predicted versus actual transport modes on the train set."
#| echo: false
#| message: false
#| warning: false
y_pred <- predict(multinom_fit, newdata = train_set)
mode_levels <- c("Car", "Bicycle", "Public_Transit", "Walking")
conf_matrix <- confusionMatrix(factor(y_pred, levels = mode_levels),
                              factor(train_set$transport_mode, levels = mode_levels))

cm_df <- as.data.frame(as.table(conf_matrix$table))
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

# Order axes so the main diagonal (correct matches) runs top-left to
# bottom-right: Predicted increases left -> right (Car -> Walking), and
# Actual increases top -> bottom (Car -> Walking). ggplot draws the first
# factor level at the bottom of the y-axis, so we reverse Actual's levels to
# place "Car" at the top.
cm_df$Predicted <- factor(cm_df$Predicted, levels = levels(cm_df$Predicted))
cm_df$Actual    <- factor(cm_df$Actual,    levels = rev(levels(cm_df$Actual)))

cm_df |>
  ggplot(aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  labs(title = "Confusion Matrix (Train Set)",
       x = "Predicted label", y = "True label") +
  theme_bw(base_size = 13)
```

## Python Output

```{python}
#| label: fig-confusion-matrix-train-py
#| fig-cap: "Confusion matrix of predicted versus actual transport modes on the train set."
#| echo: false
#| message: false
#| warning: false
y_pred = multinom_fit.predict(X_numeric).idxmax(axis=1)

# Compute confusion matrix
cm = confusion_matrix(y.cat.codes, y_pred, labels=range(len(y.cat.categories)))
disp = ConfusionMatrixDisplay(confusion_matrix=cm,
                              display_labels=y.cat.categories)
disp.plot(cmap="Blues")
disp.ax_.set_title("Confusion Matrix (Train Set)")
fig = disp.figure_
fig.tight_layout()
fig
```

:::

You are encouraged to repeat these diagnostics on the balanced (upsampled) model to determine if balancing improves minority class recall without overly harming the overall calibration.

Let's compute the performance metrics.

::: {.panel-tabset}

## R Code

```{.r}
# Classification predictions & confusion matrix
y_pred <- predict(multinom_fit, newdata = train_set)
conf_matrix <- confusionMatrix(factor(y_pred, levels = levels(train_set$transport_mode)),
                               factor(train_set$transport_mode, levels = levels(train_set$transport_mode)))
overall_accuracy <- conf_matrix$overall['Accuracy']
macro_recall <- mean(conf_matrix$byClass[,"Recall"], na.rm = TRUE)
macro_f1 <- mean(conf_matrix$byClass[,"F1"], na.rm = TRUE)
cat("Overall Accuracy:", round(overall_accuracy, 3), "\n")
cat("Macro Recall:", round(macro_recall, 3), "\n")
cat("Macro F1:", round(macro_f1, 3), "\n")
```

## Python Code

```{.python}
# Compute classification predictions & confusion matrix
probs = multinom_fit.predict(X_numeric)  # DataFrame of class probabilities
y_pred = probs.idxmax(axis=1)

# Compute the metrics
acc = accuracy_score(y.cat.codes, y_pred)
recall = recall_score(y.cat.codes, y_pred, average='macro')
# Average F1 only over the classes the model actually predicts, so a
# never-predicted class (undefined precision) is dropped rather than
# counted as 0 — matching caret's byClass F1 with na.rm = TRUE in the R tab.
predicted_labels = np.unique(y_pred)
f1 = f1_score(y.cat.codes, y_pred, labels=predicted_labels, average='macro')
print(f"Overall Accuracy: {acc:.3f}")
print(f"Macro Recall: {recall:.3f}")
print(f"Macro F1-score: {f1:.3f}")
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
y_pred <- predict(multinom_fit, newdata = train_set)
conf_matrix <- confusionMatrix(factor(y_pred, levels = levels(train_set$transport_mode)),
                               factor(train_set$transport_mode, levels = levels(train_set$transport_mode)))
overall_accuracy <- conf_matrix$overall['Accuracy']
macro_recall <- mean(conf_matrix$byClass[,"Recall"], na.rm = TRUE)
macro_f1 <- mean(conf_matrix$byClass[,"F1"], na.rm = TRUE)
cat("Overall Accuracy:", round(overall_accuracy, 3), "\n")
cat("Macro Recall:", round(macro_recall, 3), "\n")
cat("Macro F1:", round(macro_f1, 3), "\n")
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
probs = multinom_fit.predict(X_numeric)
y_pred = probs.idxmax(axis=1)
acc = accuracy_score(y.cat.codes, y_pred)
recall = recall_score(y.cat.codes, y_pred, average='macro')
# Average F1 only over the classes the model actually predicts, so a
# never-predicted class (undefined precision) is dropped rather than
# counted as 0 — matching caret's byClass F1 with na.rm = TRUE in the R tab.
predicted_labels = np.unique(y_pred)
f1 = f1_score(y.cat.codes, y_pred, labels=predicted_labels, average='macro')
print(f"Overall Accuracy: {acc:.3f}")
print(f"Macro Recall: {recall:.3f}")
print(f"Macro F1-score: {f1:.3f}")
```

:::

On the train set, the model achieves an overall accuracy of about **`r 100.0*round(overall_accuracy, 3)`%**. On average, the model correctly identifies **`r 100.0*round(macro_recall, 3)`%** of instances for each transport mode, which means that the macro recall is **`r 100.0*round(macro_recall, 3)`%**. Macro recall is relatively low, which suggests that the model is biased towards the majority classes, even though the overall accuracy is decent. Macro F1 score is **`r 100.0*round(macro_f1, 3)`%**, which means the model has a moderate balance of precision and recall on average per class. Macro F1 is much higher than macro recall, and this can happen if precision is high for the majority classes, but recall for the minority classes is low.

In summary, the model performs okay overall but is biased towards the dominant classes. It often fails to detect less frequent transport modes, which is why macro recall is low. Accuracy alone would overestimate the model’s effectiveness across all classes. For comparison, you are encouraged to explore the performance of the fitted model on the balanced sets. 

## Inference {#sec-inference}

After we finish estimating the coefficients, we look at how uncertain they are and whether they differ from zero. We get a variance estimate for each coefficient from the observed information matrix (the negative of the Hessian at the solution). Taking square roots of those variances gives standard errors. 
Similar as in [Classical Poisson regression](https://alexrod61.github.io/regression-cookbook/book/10-classical-poisson.html), each coefficient divided by its standard error gives a (asymptotically normal) **Wald Z statistic** to test if the true effect is zero. Using the same standard errors we build confidence intervals: estimate $\pm$ (critical value $\times$ standard error). To make results easier to interpret, we often exponentiate a coefficient (or its confidence limits) to get an odds ratio—how the odds of choosing a category change for a one‑unit increase in the regressor, holding other variables fixed.

We can determine whether a regressor is statistically associated with the logarithm of the odds through hypothesis testing for the parameters $\beta^{(u,v)}_j$ by link function. Define the Wald statistic $z^{(u,v)}_j$ as:
$$
z_j^{(u, v)} = \frac{\hat{\beta}_j^{(u, v)}}{\mbox{se}\left(\hat{\beta}_j^{(u, v)}\right)}
$$ {#eq-wald-statistic}
and we will test the following hypotheses: 
$$
\begin{gather*}
H_0: \beta_j^{(u, v)} = 0\\
H_a: \beta_j^{(u, v)} \neq 0.
\end{gather*}
$$
Provided that the sample size $n$ is large enough, $z_j$ has an approximately Standard Normal distribution under $H_0$.

The corresponding $p$-values for each $\beta^{(u,v)}_j$ can be computed. The smaller the $p$-value, the stronger is the evidence against the null hypothesis $H_0$. As in the previous regression models, we would set a predetermined significance level  (usually taken to be $\alpha=0.05$) to infer if the $p$-value is small enough. If the $p$-value is smaller than the predetermined level $\alpha$, then you could claim that there is evidence to reject the null hypothesis. Hence, $p$-values that are small enough indicate that the data provides evidence in favour of association (or causation in the case of an experimental study!) between the response variable and the $j$th regressor.

Given a specified level of confidence where $\alpha$ is the significance level, we can construct approximate $(1-\alpha)\times 100\%$ confidence intervals for the corresponding true value of $\beta^{(u,v)}_j$:
$$
\begin{equation*}
\hat{\beta}_j^{(u, v)} \pm z_{\alpha/2}\mbox{se} \left( \hat{\beta}_j^{(u, v)} \right),
\end{equation*}
$$ {#eq-ci}
where $z_{\alpha/2}$ is the upper $\alpha/2$ quantile of the Standard Normal distribution.

### Fitting the Model

Now we will fit the model on the test set using the `nnet` package in `R` or the `statsmodel` package in `Python`. We will also compute the corresponding 95\% confidence intervals below and filter the statistically significant coefficients at the $5\%$ significance level.

::: {.panel-tabset}

## R Code

```{.r}
# Set "Car" as the baseline level
test_set$transport_mode <- relevel(test_set$transport_mode, ref = "Car")

# Fit Multinomial Logistic regression model on test set
multinom_test <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = test_set)

# Print the fitted model
broom::tidy(multinom_test, conf.int = TRUE, exponentiate = TRUE) |>
  dplyr::rename(outcome_class = y.level) |>
  dplyr::mutate_if(is.numeric, round, 3) |>
  print(n = Inf, width = Inf)

# Significant coefficients only
cat("\nSignificant coefficients (5% threshold):\n")

results_df <- broom::tidy(multinom_test, conf.int = TRUE, exponentiate = TRUE) |>
  dplyr::mutate_if(is.numeric, round, 3) |>
  dplyr::filter(p.value < 0.05 & term != "(Intercept)") |>          # keep only significant coefficients
  rename(outcome_class = y.level) |>
  print(n = Inf, width = Inf)               # show full term names without truncation

# Display the results
results_df
```

## Python Code

```{.python}
# Factorize the response and set the baseline model
test_set["transport_mode"] = pd.Categorical(
    test_set["transport_mode"],
    categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)

# Design matrix and response
y = test_set["transport_mode"]
X = test_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]

# Add intercept
X = sm.add_constant(X)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_numeric = X.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes

# Fit Multinomial Logistic regression
model = sm.MNLogit(y, X_numeric)
multinom_fit = model.fit()

# Print the fitted model
print(multinom_fit.summary())
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
# Set "Car" as the baseline level
test_set$transport_mode <- relevel(test_set$transport_mode, ref = "Car")

# Fit Multinomial Logistic regression model on test set
multinom_test <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = test_set)

# Print the fitted model
print(summary(multinom_test))

# Significant coefficients only
cat("\nSignificant coefficients (5% threshold):\n")

results_df <-broom::tidy(multinom_test, conf.int = TRUE, exponentiate = TRUE) |>
  dplyr::mutate_if(is.numeric, round, 3) |>
  dplyr::filter(p.value < 0.05 & term != "(Intercept)") |>          # keep only significant coefficients
  rename(outcome_class = y.level) |>
  print(n = Inf, width = Inf)               # show full term names without truncation

# Display the results
results_df
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
# Factorize the response and set the baseline model
test_set["transport_mode"] = pd.Categorical(
    test_set["transport_mode"],
    categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)

# Design matrix and response
y = test_set["transport_mode"]
X = test_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]

# Add intercept
X = sm.add_constant(X)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_numeric = X.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes

# Fit Multinomial Logistic regression
model = sm.MNLogit(y, X_numeric)
multinom_fit = model.fit()

# Print the fitted model
print(multinom_fit.summary())
```

:::

On the test set, commute distance and the availability of cars are significant predictors of choosing public transit over driving at a significance threshold of $5\%$. The remaining rows in the table of significant coefficients in the `R` output should be read with caution. As shown in the outputs below, the test set contains only two `Walking` observations and seven `Bicycle` observations. None of these nine observations are weekend commuters and both the `Walking` observations do not have car available; therefore, the `Walking` and `Bicycle` equations suffer from quasi-complete separation. 

::: {.panel-tabset}

## R Code
```{.r}
test_set |> filter(transport_mode == "Walking" | transport_mode == "Bicycle")
```

## Python Code
```{.python}
test_set[(test_set["transport_mode"] == "Walking") |
         (test_set["transport_mode"] == "Bicycle")]
```

:::

::: {.panel-tabset}

## R Output
```{r}
#| echo: false
#| message: false
#| warning: false
test_set |> filter(transport_mode == "Walking" | transport_mode == "Bicycle")
```

## Python Output
```{python}
#| echo: false
#| message: false
#| warning: false
test_set[(test_set["transport_mode"] == "Walking") |
         (test_set["transport_mode"] == "Bicycle")]
```

:::

The `R` output consequently flags the `weekend_commuter` coefficients as significant for both `Bicycle` and `Walking` with standard errors of essentially zero and extreme test statistics. On the other hand, `statsmodels` handles the separation differently by assigning the separated terms very large standard errors, and yielding $p$-values close to $1$, so that they are correctly excluded from the significant coefficients in the `Python` output.  

To handle the imbalanced groups in the test set, let's upsample the test set by using the `caret::upSample` function in `R` and the `RandomOverSampler` function in `Python`.

::: {.panel-tabset}

## R Code
```{.r}
# Upsample the outcome classes
set.seed(352) # set a seed for sampling
test_set_balanced <- upSample(x = test_set[, c("age","commute_distance","has_car_available", "weekend_commuter")],
                  y = test_set$transport_mode,
                  yname = "transport_mode")
# Fit the multinomial logit regression
multinom_fit <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = test_set_balanced)
print(summary(multinom_fit))
```

## Python Code
```{.python}
import numpy as np
import pandas as pd
import statsmodels.api as sm
from imblearn.over_sampling import RandomOverSampler

ros = RandomOverSampler(random_state=352)
X_resampled, y_resampled = ros.fit_resample(
    test_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]],
    test_set["transport_mode"]
)

# Add intercept
X_resampled = sm.add_constant(X_resampled)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_numeric = X_resampled.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes

# Fit Multinomial Logistic regression
model_resampled = sm.MNLogit(y_resampled, X_numeric)
multinom_fit = model_resampled.fit()

# Print the fitted model
print(multinom_fit.summary())
```

:::

::: {.panel-tabset}

## R Output
```{r}
#| echo: false
#| message: false
#| warning: false

# Upsample the outcome classes
set.seed(352) # set a seed for sampling
test_set_balanced <- upSample(x = test_set[, c("age","commute_distance","has_car_available", "weekend_commuter")],
                  y = test_set$transport_mode,
                  yname = "transport_mode")
# Fit the multinomial logit regression
multinom_fit <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = test_set_balanced)
print(summary(multinom_fit))
```

## Python Output
```{python}
#| echo: false
#| message: false
#| warning: false

import numpy as np
import pandas as pd
import statsmodels.api as sm
from imblearn.over_sampling import RandomOverSampler

ros = RandomOverSampler(random_state=352)
X_resampled, y_resampled = ros.fit_resample(
    test_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]],
    test_set["transport_mode"]
)

# Add intercept
X_resampled = sm.add_constant(X_resampled)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_numeric = X_resampled.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes

# Fit Multinomial Logistic regression
model_resampled = sm.MNLogit(y_resampled, X_numeric)
multinom_fit = model_resampled.fit()

# Print the fitted model
print(multinom_fit.summary())
```

:::

### Regressors Associated with Transport Mode {#sec-regressors}

```{r}
#| label: multinom-test-inline
#| echo: false
#| message: false
#| warning: false

# Coefficient table on the log-odds scale: estimates, standard errors, z, p
coef_log <- broom::tidy(multinom_test, conf.int = FALSE) |>
  dplyr::rename(outcome_class = y.level)

# Coefficient table on the odds-ratio scale, with Wald 95% confidence intervals
coef_or <- broom::tidy(multinom_test, conf.int = TRUE, exponentiate = TRUE) |>
  dplyr::rename(outcome_class = y.level)

# Non-baseline outcome classes, taken from the data rather than hard coded
non_baseline <- setdiff(levels(test_set$transport_mode), "Car")
class_bike <- grep("Bicycle", non_baseline, value = TRUE)
class_pt   <- grep("Transit", non_baseline, value = TRUE)
class_walk <- grep("Walk",    non_baseline, value = TRUE)

# Pull a single quantity for a given outcome class and term.
# Fails loudly at render time if the class or term name does not match.
get_stat <- function(class, term_name, column, tbl = coef_or) {
  value <- tbl[[column]][tbl$outcome_class == class & tbl$term == term_name]
  if (length(value) != 1) {
    stop("No unique match for outcome class '", class, "' and term '", term_name, "'.")
  }
  value
}

# Numeric formatting that degrades gracefully under quasi-complete separation.
# Returns bare LaTeX, so inline calls are wrapped in $ ... $ in the prose.
fmt_num <- Vectorize(function(x, digits = 3) {
  if (is.na(x)) return("\\text{NA}")
  if (!is.finite(x)) return(paste0(if (x < 0) "-" else "", "\\approx \\infty"))
  if (x != 0 && abs(x) < 10^(-digits)) return("\\approx 0")
  if (abs(x) >= 1e4) {
    e <- floor(log10(abs(x)))
    return(sprintf("%.2f \\times 10^{%d}", x / 10^e, e))
  }
  formatC(x, format = "f", digits = digits)
})

# p-values: bare form for tables, prefixed form for prose
fmt_pval <- function(p) {
  if (p < 0.001) "< 0.001" else formatC(p, format = "f", digits = 3)
}
fmt_p <- function(p) {
  if (p < 0.001) "< 0.001" else paste0("= ", fmt_pval(p))
}

# 95% confidence interval on the odds-ratio scale, as a comma-separated pair
fmt_ci <- function(class, term_name, digits = 3) {
  paste0(
    fmt_num(get_stat(class, term_name, "conf.low"), digits), ", ",
    fmt_num(get_stat(class, term_name, "conf.high"), digits)
  )
}
```

The Multinomial Logistic regression model fitted on the test set yields the estimated coefficients above. Note that the estimates and confidence limits in the table of significant coefficients are exponentiated, so that each entry is an odds ratio relative to the baseline category `Car`. A positive value of $\hat{\beta}_j^{(k,\texttt{Car})}$ corresponds to higher values of regressor $j$ being associated with a higher odds of category $k$ relative to `Car`, whereas a negative value of $\hat{\beta}_j^{(k,\texttt{Car})}$ corresponds to lower odds of category $k$ relative to `Car` for higher values of regressor $j$. The significant regression coefficients belong to the `` `r class_pt` `` equation, and we will interpret them below.

**Commute distance:** Commute distance is a significant predictor of choosing public transit over the car ($\text{OR} = `r fmt_num(get_stat(class_pt, "commute_distance", "estimate"), 2)`$, $z = `r fmt_num(get_stat(class_pt, "commute_distance", "statistic", coef_log), 2)`$, $p `r fmt_p(get_stat(class_pt, "commute_distance", "p.value"))`$, 95% CI: $`r fmt_ci(class_pt, "commute_distance", 2)`$). It has the following interpretation.

> *Holding age, car availability, and weekend commuting status constant, each additional kilometre of one-way commute distance is associated with `r fmt_num(get_stat(class_pt, "commute_distance", "estimate"), 2)` times higher odds of choosing public transit over car.*

Although the per-kilometre effect appears modest, it compounds over larger differences in commute length. For example, a commute that is $10$ kilometres longer multiplies the odds of choosing public transit over the car by approximately $`r fmt_num(get_stat(class_pt, "commute_distance", "estimate"), 2)`^{10} \approx `r fmt_num(get_stat(class_pt, "commute_distance", "estimate")^10, 1)`$, assuming that age, car availability, and weekend commuting status are held constant.

**Car availability:** Car availability is a significant predictor of choosing public transit over the car ($\text{OR} = `r fmt_num(get_stat(class_pt, "has_car_availableYes", "estimate"))`$, $z = `r fmt_num(get_stat(class_pt, "has_car_availableYes", "statistic", coef_log), 2)`$, $p `r fmt_p(get_stat(class_pt, "has_car_availableYes", "p.value"))`$, 95% CI: $`r fmt_ci(class_pt, "has_car_availableYes")`$). It has the following interpretation.

> *Holding commute distance, age, and weekend commuting status constant, individuals with a car available have `r fmt_num(get_stat(class_pt, "has_car_availableYes", "estimate"))` times the odds of choosing public transit over the car, relative to individuals without a car available. In other words, car availability is associated with a reduction of approximately `r fmt_num(100 * (1 - get_stat(class_pt, "has_car_availableYes", "estimate")), 1)`% in the odds of choosing public transit over car.*

The remaining rows of the significant coefficients table in the `R` output are artifacts of the quasi-complete separation discussed above and are therefore not interpreted.

## Results

The fitted model can be written as
$$
\log \left( \frac{\Pr(Y_i = k)}{\Pr(Y_i = \text{Car})} \right) = \hat{\beta}_{0}^{(k,\text{Car})} + \hat{\beta}_{1}^{(k,\text{Car})} x_{i,\texttt{distance}} + \hat{\beta}_{2}^{(k,\text{Car})} x_{i,\texttt{age}} + \hat{\beta}_{3}^{(k,\text{Car})}Z_{i,\texttt{car}} + \hat{\beta}_{4}^{(k,\text{Car})} Z_{i,\texttt{weekend}},
$$ {#eq-fitted-model}
where $k$ is one of the non-baseline levels: `` `r class_bike` ``, `` `r class_pt` ``, or `` `r class_walk` ``, and the regressors are denoted as $x_{i,\texttt{distance}}, x_{i,\texttt{age}}, Z_{i,\texttt{car}},$ and $Z_{i,\texttt{weekend}}$.

The estimated intercepts and slope coefficients of the Multinomial Logistic regression model fitted on the test set in @sec-inference are summarized in the three tables below. Every estimate is interpreted relative to the baseline category `Car`, and the $z$ statistics, $p$-values, odds ratios, and $95\%$ Wald confidence intervals are taken directly from the `R` output:

```{r}
#| label: multinom-coef-tables
#| echo: false
#| results: asis

term_order <- c("(Intercept)", "commute_distance", "age",
                "has_car_availableYes", "weekend_commuterYes")
param_labels <- c("(intercept)", "`commute_distance`", "`age`",
                  "`has_car_availableYes`", "`weekend_commuterYes`")

for (k in c(class_bike, class_pt, class_walk)) {

  # Escape underscores and spaces for use inside \text{ } in the parameter symbols
  k_math <- gsub("[ _]", "\\\\_", k)

  cat("**`", k, "` versus `Car`:**\n\n", sep = "")
  cat("| Parameter | Estimate | Std. Error | $z$ | $p$-value | OR | 95% CI |\n")
  cat("|-----------|----------|------------|-----|-----------|-----|--------|\n")

  for (i in seq_along(term_order)) {
    tm <- term_order[i]
    cat(sprintf(
      "| $\\hat{\\beta}_{%d}^{(\\text{%s},\\,\\text{Car})}$ %s | $%s$ | $%s$ | $%s$ | $%s$ | $%s$ | $(%s)$ |\n",
      i - 1L, k_math, param_labels[i],
      fmt_num(get_stat(k, tm, "estimate",   coef_log)),
      fmt_num(get_stat(k, tm, "std.error",  coef_log)),
      fmt_num(get_stat(k, tm, "statistic",  coef_log), 2),
      fmt_pval(get_stat(k, tm, "p.value",   coef_log)),
      fmt_num(get_stat(k, tm, "estimate")),
      fmt_ci(k, tm)
    ))
  }

  cat("\n")
}
```

The `` `r class_bike` `` and `` `r class_pt` `` blocks are estimated on adequate numbers of observations, and within these blocks, `commute_distance` and `has_car_availableYes` in the `` `r class_pt` `` equation are significant at the $5\%$ level. As discussed in @sec-inference, the `` `r class_walk` `` equation and the `weekend_commuterYes` term in the `` `r class_bike` `` equation suffer from quasi-complete separation; the resulting extreme coefficient magnitudes paired with astronomical standard errors drive the associated $z$-statistics toward zero and the $p$-values toward one, which is a symptom of a maximum likelihood estimate that does not exist rather than genuine null effects.

### Prediction 

Now we will fit the Multinomial Logistic regression model on the train set, and predict on the held-out test set. The confusion matrix and performance metrics are reported below.

::: {.panel-tabset}

## R Code

```{.r}
# Fit the model on the train set
multinom_fit <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = train_set)

# Predict on the test set using the fitted model
y_pred <- predict(multinom_fit, newdata = test_set)

# Compute confusion matrix
mode_levels <- c("Car", "Bicycle", "Public_Transit", "Walking")
conf_matrix <- confusionMatrix(factor(y_pred, levels = mode_levels),
                              factor(test_set$transport_mode, levels = mode_levels))

# Visualise
cm_df <- as.data.frame(as.table(conf_matrix$table))
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

# Order axes so the main diagonal (correct matches) runs top-left to bottom-right
cm_df$Predicted <- factor(cm_df$Predicted, levels = levels(cm_df$Predicted))
cm_df$Actual    <- factor(cm_df$Actual,    levels = rev(levels(cm_df$Actual)))

cm_df |>
  ggplot(aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  labs(title = "Confusion Matrix (Test Set)",
       x = "Predicted label", y = "True label") +
  theme_bw(base_size = 13)
```

## Python Code

```{.python}
# Factorize the response and set the baseline model
train_set["transport_mode"] = pd.Categorical(
    train_set["transport_mode"],
    categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)

# Design matrix and response
y = train_set["transport_mode"]
X = train_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]

# Add intercept
X = sm.add_constant(X)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_numeric = X.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes

# Fit Multinomial Logistic regression
model = sm.MNLogit(y, X_numeric)
multinom_fit = model.fit()

# Predict on the test set using the fitted model
test_set["transport_mode"] = pd.Categorical(
  test_set["transport_mode"],
  categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)
y_test = test_set["transport_mode"]
X_test = test_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]
X_test = sm.add_constant(X_test)
X_test_num = X_test.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_test_num[col] = X_test_num[col].cat.codes
y_pred_test = multinom_fit.predict(X_test_num).idxmax(axis=1)

# Confusion matrix
labels = y_test.cat.categories
cm = confusion_matrix(y_test.cat.codes, y_pred_test, labels=range(len(labels)))
disp = ConfusionMatrixDisplay(confusion_matrix=cm,
                              display_labels=y_test.cat.categories)
disp.plot(cmap="Blues")
disp.ax_.set_title("Confusion Matrix (Test Set)")
fig = disp.figure_
fig.tight_layout()
fig
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| label: fig-confusion-matrix-test-R
#| fig-cap: "Confusion matrix of predicted versus actual transport modes on the held-out test set."
#| echo: false
#| message: false
#| warning: false

multinom_fit <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = train_set)

y_pred <- predict(multinom_fit, newdata = test_set)
mode_levels <- c("Car", "Bicycle", "Public_Transit", "Walking")
conf_matrix <- confusionMatrix(factor(y_pred, levels = mode_levels),
                              factor(test_set$transport_mode, levels = mode_levels))

cm_df <- as.data.frame(as.table(conf_matrix$table))
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

cm_df$Predicted <- factor(cm_df$Predicted, levels = levels(cm_df$Predicted))
cm_df$Actual    <- factor(cm_df$Actual,    levels = rev(levels(cm_df$Actual)))

cm_df |>
  ggplot(aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  labs(title = "Confusion Matrix (Test Set)",
       x = "Predicted label", y = "True label") +
  theme_bw(base_size = 13)
```

## Python Output

```{python}
#| label: fig-confusion-matrix-test-py
#| fig-cap: "Confusion matrix of predicted versus actual transport modes on the held-out test set."
#| echo: false
#| message: false
#| warning: false

# Factorize the response and set the baseline model
train_set["transport_mode"] = pd.Categorical(
    train_set["transport_mode"],
    categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)

# Design matrix and response
y = train_set["transport_mode"]
X = train_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]

# Add intercept
X = sm.add_constant(X)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_numeric = X.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes

# Fit Multinomial Logistic regression
model = sm.MNLogit(y, X_numeric)
multinom_fit = model.fit()

test_set["transport_mode"] = pd.Categorical(
  test_set["transport_mode"], 
  categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)
y_test = test_set["transport_mode"]
X_test = test_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]
X_test = sm.add_constant(X_test)
X_test_num = X_test.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_test_num[col] = X_test_num[col].cat.codes
y_pred_test = multinom_fit.predict(X_test_num).idxmax(axis=1)

# Confusion matrix
labels = y_test.cat.categories
cm = confusion_matrix(y_test.cat.codes, y_pred_test, labels=range(len(labels)))
disp = ConfusionMatrixDisplay(confusion_matrix=cm,
                              display_labels=y_test.cat.categories)
disp.plot(cmap="Blues")
disp.ax_.set_title("Confusion Matrix (Test Set)")
fig = disp.figure_
fig.tight_layout()
fig
```

:::

::: {.panel-tabset}

## R Code

```{.r}
# Compute the performance metrics
overall_accuracy <- conf_matrix$overall['Accuracy']
macro_recall <- mean(conf_matrix$byClass[,"Recall"], na.rm = TRUE)
macro_f1 <- mean(conf_matrix$byClass[,"F1"], na.rm = TRUE)
cat("Overall Accuracy:", round(overall_accuracy, 3), "\n")
cat("Macro Recall:", round(macro_recall, 3), "\n")
cat("Macro F1:", round(macro_f1, 3), "\n")
```

## Python Code

```{.python}
# Compute the performance metrics
acc = accuracy_score(y_test.cat.codes, y_pred_test)
recall = recall_score(y_test.cat.codes, y_pred_test, average='macro')
# Average F1 only over the classes the model actually predicts, so a
# never-predicted class (undefined precision) is dropped rather than
# counted as 0 — matching caret's byClass F1 with na.rm = TRUE in the R tab.
predicted_labels = np.unique(y_pred_test)
f1 = f1_score(y_test.cat.codes, y_pred_test, labels=predicted_labels, average='macro')
print(f"Overall Accuracy: {acc:.3f}")
print(f"Macro Recall: {recall:.3f}")
print(f"Macro F1-score: {f1:.3f}")
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
overall_accuracy <- conf_matrix$overall['Accuracy']
macro_recall <- mean(conf_matrix$byClass[,"Recall"], na.rm = TRUE)
macro_f1 <- mean(conf_matrix$byClass[,"F1"], na.rm = TRUE)
cat("Overall Accuracy:", round(overall_accuracy, 3), "\n")
cat("Macro Recall:", round(macro_recall, 3), "\n")
cat("Macro F1:", round(macro_f1, 3), "\n")
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
acc = accuracy_score(y_test.cat.codes, y_pred_test)
recall = recall_score(y_test.cat.codes, y_pred_test, average='macro')
# Average F1 only over the classes the model actually predicts, so a
# never-predicted class (undefined precision) is dropped rather than
# counted as 0 — matching caret's byClass F1 with na.rm = TRUE in the R tab.
predicted_labels = np.unique(y_pred_test)
f1 = f1_score(y_test.cat.codes, y_pred_test, labels=predicted_labels, average='macro')
print(f"Overall Accuracy: {acc:.3f}")
print(f"Macro Recall: {recall:.3f}")
print(f"Macro F1-score: {f1:.3f}")
```

:::

On the test set, the model achieves an overall accuracy of **`r 100.0*round(overall_accuracy, 3)`%**, a macro recall of **`r 100.0*round(macro_recall, 3)`%**, and a macro F1 score of **`r 100.0*round(macro_f1, 3)`%**. 

The `Car` and `Public Transit` classes are predicted most reliably with recalls of $82.9\%$ and $64.7\%$ respectively. On the other hand, the observations corresponding to minority classes (`Bicycle` and `Walking`) in the test set are all misclassified as `Car` or `Public Transit`. Among the seven `Bicycle` observations in the test set, six are misclassified as `Car` and one is misclassified as `Public_Transit`. For the two `Walking` observations in the test set, one is misclassified as `Car` and one is misclassified as `Public_Transit`. 

::: {.Tip}
::::{.Tip-header}
Class Imbalance and Prediction
::::
::::{.Tip-container}
The absence of `Bicycle` and `Walking` predictions follows from the structure of the maximum likelihood solution under class imbalance. Differentiating the log-likelihood function in @eq-log-likelihood-multinomial with respect to the intercept $\beta_0^{(j,1)}$ of a non-baseline category $j$ and setting the partial derivative to zero yields the following equation:

$$
\sum_{i=1}^n \hat{p}_{i,j} = \sum_{i=1}^n y_{i,j} = n_j,
$$

where $n_j$ is the number of observations in category $j$ in the train set. Consequently, the fitted probabilities of each category average exactly its sample proportion in the train set, which is roughly $2.4\%$ for `Bicycle` and less than $1\%$ for `Walking`. Since the classification rule assigns each observation to the category with the highest fitted probability, a minority category is predicted only when the regressors raise its fitted probability well above its small sample proportion, and the estimated slope coefficients do not produce a shift of that magnitude for any regressor values observed in the test set. Neither `Bicycle` nor `Walking` therefore ever attains the maximum fitted probability, due to which these minority classes are not predicted at all in the test set.
::::
:::

Now let's use the upsampled model (`multinom_fit_balanced`) from @sec-estimation to predict on the held-out test set. The confusion matrix and performance metrics are reported below.

::: {.panel-tabset}

## R Code
```{.r}
# Predict on the test set using the fitted multinom_fit_balanced model
y_pred <- predict(multinom_fit_balanced, newdata = test_set)

# Compute confusion matrix
mode_levels <- c("Car", "Bicycle", "Public_Transit", "Walking")
conf_matrix <- confusionMatrix(factor(y_pred, levels = mode_levels),
                              factor(test_set$transport_mode, levels = mode_levels))

# Visualise
cm_df <- as.data.frame(as.table(conf_matrix$table))
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

# Order axes so the main diagonal (correct matches) runs top-left to
# bottom-right: Predicted increases left -> right (Car -> Walking), and
# Actual increases top -> bottom (Car -> Walking). ggplot draws the first
# factor level at the bottom of the y-axis, so we reverse Actual's levels to
# place "Car" at the top.
cm_df$Predicted <- factor(cm_df$Predicted, levels = levels(cm_df$Predicted))
cm_df$Actual    <- factor(cm_df$Actual,    levels = rev(levels(cm_df$Actual)))

cm_df |>
  ggplot(aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  labs(title = "Confusion Matrix (Test Set)",
       x = "Predicted label", y = "True label") +
  theme_bw(base_size = 13)
```

## Python Code
```{.python}
# Predict on the test set using the fitted model
y_pred_test = multinom_fit_balanced.predict(X_test_num).idxmax(axis=1)

# Confusion matrix
labels = y_test.cat.categories
cm = confusion_matrix(y_test.cat.codes, y_pred_test, labels=range(len(labels)))
disp = ConfusionMatrixDisplay(confusion_matrix=cm,
                              display_labels=y_test.cat.categories)
disp.plot(cmap="Blues")
disp.ax_.set_title("Confusion Matrix (Test Set)")
fig = disp.figure_
fig.tight_layout()
fig
```

:::

::: {.panel-tabset}

## R Output
```{r}
#| echo: false
#| message: false
#| warning: false
# Predict on the test set using the fitted multinom_fit_balanced model
y_pred <- predict(multinom_fit_balanced, newdata = test_set)

# Compute confusion matrix
mode_levels <- c("Car", "Bicycle", "Public_Transit", "Walking")
conf_matrix <- confusionMatrix(factor(y_pred, levels = mode_levels),
                              factor(test_set$transport_mode, levels = mode_levels))

# Visualise
cm_df <- as.data.frame(as.table(conf_matrix$table))
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

# Order axes so the main diagonal (correct matches) runs top-left to
# bottom-right: Predicted increases left -> right (Car -> Walking), and
# Actual increases top -> bottom (Car -> Walking). ggplot draws the first
# factor level at the bottom of the y-axis, so we reverse Actual's levels to
# place "Car" at the top.
cm_df$Predicted <- factor(cm_df$Predicted, levels = levels(cm_df$Predicted))
cm_df$Actual    <- factor(cm_df$Actual,    levels = rev(levels(cm_df$Actual)))

cm_df |>
  ggplot(aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  labs(title = "Confusion Matrix (Test Set)",
       x = "Predicted label", y = "True label") +
  theme_bw(base_size = 13)
```

## Python Output
```{python}
#| echo: false
#| message: false
#| warning: false
# Predict on the test set using the fitted model
y_pred_test = multinom_fit_balanced.predict(X_test_num).idxmax(axis=1)

# Confusion matrix
labels = y_test.cat.categories
cm = confusion_matrix(y_test.cat.codes, y_pred_test, labels=range(len(labels)))
disp = ConfusionMatrixDisplay(confusion_matrix=cm,
                              display_labels=y_test.cat.categories)
disp.plot(cmap="Blues")
disp.ax_.set_title("Confusion Matrix (Test Set)")
fig = disp.figure_
fig.tight_layout()
fig
```
:::

::: {.panel-tabset}

## R Code

```{.r}
# Compute the performance metrics
overall_accuracy <- conf_matrix$overall['Accuracy']
macro_recall <- mean(conf_matrix$byClass[,"Recall"], na.rm = TRUE)
macro_f1 <- mean(conf_matrix$byClass[,"F1"], na.rm = TRUE)
cat("Overall Accuracy:", round(overall_accuracy, 3), "\n")
cat("Macro Recall:", round(macro_recall, 3), "\n")
cat("Macro F1:", round(macro_f1, 3), "\n")
```

## Python Code

```{.python}
# Compute the performance metrics
acc = accuracy_score(y_test.cat.codes, y_pred_test)
recall = recall_score(y_test.cat.codes, y_pred_test, average='macro')
# Average F1 only over the classes the model actually predicts, so a
# never-predicted class (undefined precision) is dropped rather than
# counted as 0 — matching caret's byClass F1 with na.rm = TRUE in the R tab.
predicted_labels = np.unique(y_pred_test)
f1 = f1_score(y_test.cat.codes, y_pred_test, labels=predicted_labels, average='macro')
print(f"Overall Accuracy: {acc:.3f}")
print(f"Macro Recall: {recall:.3f}")
print(f"Macro F1-score: {f1:.3f}")
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
overall_accuracy <- conf_matrix$overall['Accuracy']
macro_recall <- mean(conf_matrix$byClass[,"Recall"], na.rm = TRUE)
macro_f1 <- mean(conf_matrix$byClass[,"F1"], na.rm = TRUE)
cat("Overall Accuracy:", round(overall_accuracy, 3), "\n")
cat("Macro Recall:", round(macro_recall, 3), "\n")
cat("Macro F1:", round(macro_f1, 3), "\n")
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
acc = accuracy_score(y_test.cat.codes, y_pred_test)
recall = recall_score(y_test.cat.codes, y_pred_test, average='macro')
# Average F1 only over the classes the model actually predicts, so a
# never-predicted class (undefined precision) is dropped rather than
# counted as 0 — matching caret's byClass F1 with na.rm = TRUE in the R tab.
predicted_labels = np.unique(y_pred_test)
f1 = f1_score(y_test.cat.codes, y_pred_test, labels=predicted_labels, average='macro')
print(f"Overall Accuracy: {acc:.3f}")
print(f"Macro Recall: {recall:.3f}")
print(f"Macro F1-score: {f1:.3f}")
```

:::

On the test set, the upsampled model achieves an overall accuracy of **`r 100.0*round(overall_accuracy, 3)`%**, a macro recall of **`r 100.0*round(macro_recall, 3)`%**,, and a macro F1 score of **`r 100.0*round(macro_f1, 3)`%**. Overall this upsampled model has a lower accuracy on the test set than the original `multinom_fit` model; however, it correctly classifies some of the `Bicycle` and `Walking` observations in the test set with recalls of roughly $66.7\%$ and $50\%$ respectively. As discussed in @sec-estimation, there is evidence that the upsampled model improves recall of minority classes but the trade-off is that the overall accuracy is lower, meaning that it is an overfit model. 

## Storytelling

To complete the data science workflow, let's step back from the coefficients, tests, and confusion matrices, and focus on how the results can be interpreted and applied in practice. This final stage, storytelling, is where statistical output is translated into insights that support real world decision making.

In this case study, the Multinomial Logistic regression model helps us understand how commute distance, age, car availability, and weekend commuting status relate to an individual's primary mode of transport, an outcome with four unordered categories: `Car`, `Public Transit`, `Bicycle`, and `Walking`. We will distill the key takeaways from the analysis and discuss what they imply for commuters, transportation planners, and policy makers.

### Key Insights From the Model

Car availability and commute distance are the two most consistently significant regressors across both the imbalanced and upsampled models. Having access to a car is associated with lower odds of using public transit instead of driving; the corresponding effect for cycling is not statistically significant, so car access alone does not appear to explain the choice to bike in this dataset. Commute distance moves the odds of public transit and cycling in opposite directions relative to driving, such that each additional kilometre increases the odds of choosing public transit over the car, while decreasing the odds of cycling over the car. This matches the intuition that the car and public transit become the practical options once a commute gets long, while cycling remains attractive mainly over short distances.

Age and weekend commuting status tell a more conditional story. On the original, imbalanced training data, neither regressor reaches statistical significance for most categories, largely because the sample contains too few cyclists and walkers to estimate their effects precisely. Once the minority classes are upsampled, both regressors become significant predictors of `Walking`, and age additionally becomes significant for `Bicycle`: older individuals and weekend commuters are more likely to walk than drive, once the class imbalance no longer masks the effect. This is a useful reminder that failing to reject $H_0$ is not the same as confirming $H_0$ because a lack of significance can potentially reflect insufficient data rather than a genuinely null effect for rare categories.

Some estimates need a caveat rather than a substantive interpretation. The test set contains only one `Walking` observation and five `Bicycle` observations, none of the latter being weekend commuters, so the test set model suffers from complete separation. In R, the `weekend_commuter` effect for `Bicycle` and several `Walking` coefficients are flagged as significant with implausibly large or near-zero odds ratios, standard errors of essentially zero, and extreme test statistics. In Python, `statsmodels` assigns the same terms enormous standard errors and $p$-values near 1, and reports that the fit did not fully converge. None of this is evidence about the true effects; the terms are simply not identifiable from the available data. Analysts should flag this kind of estimate rather than interpret it, and should be cautious of near-zero or implausibly large odds ratios whenever a category is small.

As a predictive tool, the model favours the categories it has seen the most of. The confusion matrices for both the training and test sets show noticeably stronger performance for `Car` and `Public Transit`, the two dominant categories, than for `Bicycle` and `Walking`. Upsampling helps the model learn the minority classes better during estimation, but it cannot manufacture additional information at prediction time, so macro-averaged recall and F1 remain modest even after balancing. The model is best used as a tool for understanding how commute distance, car availability, age, and weekend habits relate to travel mode choice, rather than as a precise classifier for the least common commuting behaviours.

::: {.Tip}
::::{.Tip-header}
Using the Results Responsibly
::::
::::{.Tip-container}
It is important to remember that this analysis is associational, not causal. The model does not prove that changing someone's commute distance or car access would, on its own, change how they travel; it shows that, in this dataset, these regressors are consistently linked to the mode people choose once other measured factors are accounted for.

Nevertheless, the results provide a useful evidence-based starting point for discussion. They help identify which regressors are most closely linked to travel mode choice, and which associations, such as age and weekend commuting for the rarer categories, may only become visible with more balanced or more abundant data. For transportation planners, this kind of insight can help prioritize where infrastructure investment, such as bike lanes or improved transit access, is likely to shift commuting patterns the most.
::::
:::
