<!-- Google tag (gtag.js) -->
<script async src="https://www.googletagmanager.com/gtag/js?id=G-7PRVEBE1EF"></script>
<script>
  window.dataLayer = window.dataLayer || [];
  function gtag(){dataLayer.push(arguments);}
  gtag('js', new Date());

  gtag('config', 'G-7PRVEBE1EF');
</script>

# Multinomial Logistic Regression {#sec-multinomial-logistic}

::: {.Warning}
::::{.Warning-header}
When to Use and Not Use Multinomial Logistic Regression
::::
::::{.Warning-container}
Multinomial Logistic regression is a type of generalized linear model that is appropriately used under the following conditions:

- The outcome variable is categorical with more than two categories, as illustrated in @fig-multinomial-regression.
- There is one or more regressors that may be continuous, ordinal, or nominal.
- We assume a **linear relationship** between any continuous regressors and the logit of the cumulative probabilities of the outcome.
- The data points do not exhibit any correlation structures, which means that there is statistical independence across the sampled observations, with mutually exclusive and exhaustive ordered categories.

However, Multinomial Logistic regression should not be used in the following scenarios:

- The outcome variable is ordered. If the categories are ordered, refer to the chapter on Ordinal Logistic regression (@sec-ordinal-logistic).
- The outcome variable is binary with two categories. For binary outcomes, refer to the chapter on Binary Logistic regression (@sec-binary-logistic).
- The outcome variable is continuous or a count. For continuous and unbounded outcomes, refer to Ordinary Least-squares regression (@sec-ols). For count outcomes, refer to count regression models, such as Classical Poisson regression (@sec-classical-poisson).
- The data points exhibit correlation structures, indicating statistical dependence among the sampled observations (for example, repeated measurements on the same subject).
::::
:::

::: {#fig-multinomial-regression}
```{mermaid}
mindmap
  root((Regression 
  Analysis)
    Continuous <br/>Outcome Y
      {{Unbounded <br/>Outcome Y}}
        )Chapter 3: <br/>Ordinary <br/>Least Squares <br/>Regression(
          (Normal <br/>Outcome Y)
      {{Nonnegative <br/>Outcome Y}}
        )Chapter 4: <br/>Gamma Regression(
          (Gamma <br/>Outcome Y)
      {{Bounded <br/>Outcome Y <br/> between 0 and 1}}
        )Chapter 5: Beta <br/>Regression(
          (Beta <br/>Outcome Y)
      {{Nonnegative <br/>Survival <br/>Time Y}}
        )Chapter 6: <br/>Parametric <br/> Survival <br/>Regression(
          (Exponential <br/>Outcome Y)
          (Weibull <br/>Outcome Y)
          (Lognormal <br/>Outcome Y)
        )Chapter 7: <br/>Semiparametric <br/>Survival <br/>Regression(
          (Cox Proportional <br/>Hazards Model)
            (Hazard Function <br/>Outcome Y)
    Discrete <br/>Outcome Y
      {{Binary <br/>Outcome Y}}
        {{Ungrouped <br/>Data}}
          )Chapter 8: <br/>Binary Logistic <br/>Regression(
            (Bernoulli <br/>Outcome Y)
        {{Grouped <br/>Data}}
          )Chapter 9: <br/>Binomial Logistic <br/>Regression(
            (Binomial <br/>Outcome Y)
      {{Count <br/>Outcome Y}}
        {{Equidispersed <br/>Data}}
          )Chapter 10: <br/>Classical Poisson <br/>Regression(
            (Poisson <br/>Outcome Y)
        {{Overdispersed <br/>Data}}
          )Chapter 11: <br/>Negative Binomial <br/>Regression(
            (Negative Binomial <br/>Outcome Y)
        {{Zero Inflated <br/>Data}}
          )Chapter 12: <br/>Zero Inflated <br/>Poisson <br/>Regression(
            (Zero Inflated <br/>Poisson <br/>Outcome Y)
        {{Overdispersed or <br/>Underdispersed <br/>Data}}
          )Chapter 13: <br/>Generalized <br/>Poisson <br/>Regression(
            (Generalized <br/>Poisson <br/>Outcome Y)
      {{Categorical <br/>Outcome Y}}
        {{Nominal <br/>Outcome Y}}
          )Chapter 14: <br/>Multinomial <br/>Logistic <br/>Regression(
            (Multinomial <br/>Outcome Y)
```
:::

```{r setup-r, include=FALSE}
# Load required libraries
library(devtools)
library(readr)
library(tibble)
library(tidyr)
library(dplyr)
library(magrittr)
library(ggplot2)
library(gridExtra) # combine ggplot2 figures
library(ggExtra) # draw marginal histograms
library(broom)
library(knitr)
#devtools::install_github("andytai7/cookbook")
library(cookbook)
library(nnet) # fit multinomial regression
library(caret) # upsampling and confusion matrix
library(rsample)

library(reticulate)
py_config()

py_require("pandas")
py_require("numpy")
py_require("seaborn")
py_require("pyreadr")
py_require("plotnine")
py_require("matplotlib")
py_require("statsmodels")
py_require("scikit-learn")
py_require("spicy")

reticulate::py_install("plotnine")

pandas <- import("pandas")
numpy <- import("numpy")
seaborn <- import("seaborn")
pyreadr <- import("pyreadr")
matplotlib <- import("matplotlib")
statsmodels <- import("statsmodels")
sklearn <- import("sklearn")
scipy <- import("scipy")
plotnine <- import("plotnine")

print_as_dt <- function(x){
  DT::datatable(x,
                rownames = FALSE,
                options = list(
                  dom = 'ltipr',
                  autoWidth = TRUE,
                  columnDefs = list(
                    list(className = 'dt-left', targets = "_all"),
                    list(width = '200px', targets = "_all")
                    )
                  )
                )
}
```

```{python setup-py, include=FALSE}
import numpy as np
import pandas as pd
from plotnine import *
from statsmodels.formula.api import glm, ols, mnlogit
from statsmodels.genmod.families import Poisson
from matplotlib import pyplot as plt
from sklearn.metrics import mean_squared_error, mean_absolute_error, classification_report
import scipy.stats as stats
import seaborn as sns
from matplotlib.gridspec import GridSpec
from sklearn.linear_model import LogisticRegression
from sklearn.metrics import confusion_matrix, ConfusionMatrixDisplay, classification_report, log_loss, accuracy_score, recall_score, f1_score
from sklearn.utils import resample
import statsmodels.api as sm
```

::: {.LO}
::::{.LO-header}
Learning Objectives
::::
::::{.LO-container}
By the end of this chapter, you will be able to:

- Describe why **Ordinary Least Squares** is not appropriate for multi-class nominal outcomes.
- Determine when **Multinomial Logistic Regression** is an appropriate modelling choice.
- Write down the **likelihood function** for multinomial logistic regression and its **logit link** functions, and obtain category-specific probabilities.
- Understand how coefficient estimation via **maximum likelihood estimation** works in the multinomial logistic regression model.
- Fit a multinomial logistic regression model in both `R` and `Python`.
- Interpret multinomial logistic regression coefficients in real scenarios.
- Use multinomial logistic regression for prediction by computing category-specific probabilities.
- Evaluate model performance using both **likelihood-based** and **classification-based** measures, and construct confidence intervals using **Wald Z statistic**.
::::
:::

## Introduction

Multinomial logistic regression is a statistical modelling technique used to analyze relationships between a multi-class categorical response variable and a set of regressors. Unlike binary logistic regression, which models outcomes with two categories, multinomial logistic regression extends the framework to outcomes with more than two unordered categories. It estimates the probability of each possible outcome as a function of the regressors by modelling the log-odds of each category relative to a reference category. This method is widely used in fields such as transportation, marketing, and social sciences where the goal is to understand or predict categorical choices among multiple alternatives.


### Assumptions

- The response or dependent variable should be measured at the nominal level with more than two values. 
- There should be one or more independent variables that are continuous, ordinal or nominal (including dichotomous variables).
- **Logit linearity assumption**: There needs to be a linear relationship between any continuous independent variables and the logit transformation of the dependent variable.
- **Independent observations**: The observations should be independent and the dependent variable should have mutually exclusive and exhaustive categories (i.e. no individual belonging to two different categories).
- **No Multicollinearity**: Multicollinearity occurs when you have two or more independent variables that are highly correlated with each other.
- There should be no outliers, high leverage values or highly influential points.

## Case Study

We will explore a dataset containing information on individuals' commuting behaviors and demographics, with a focus on their primary mode of transportation. We explore a couple of research problems based on the dataset. 

- How is the transportation mode of people (driving own cars or taking public transportation specifically) **associated** with their commuting distances, given the individuals' commuting behaviors and demographics including if they have a car, if they commute on weekends, their commute distance, and age?
- Can we accurately **predict** the transportation mode of an individual based on their commuting behaviors and demographics including if they have a car, if they commute on weekends, their commute distance, and age?

These questions correspond to two distinct **statistical objectives**:

- **Inference:** Understand the relationship between the regressor `commute_distance` and the probability of each category of the response variable `transport_mode`, given the confounding effects of age, car availability, and weekend commuter status. 
- **Prediction:** Develop a classification model that classifies individuals into their most probable transport mode: `Bicycle`, `Car`, `Public Transit`, or `Walking`.

## Data Collection and Wrangling

The dataset contains the following variables:

| Variable | Description | Scale | Model Role | Variable Name |
|---|---|---|---|---|---|
| Mode of Transport | The primary mode of transport used by the individual | Nominal | Response | `transport_mode` |
| Commute Distance | The one-way commuting distance in kilometers | Numeric | Regressor | `commute_distance` |
| Age | The age of the individual in years | Numeric | Regressor | `age` |
| Car Availability | Whether the individual has access to a car | Binary Categorical | Regressor | `has_car_available` |
| Weekend Commuter | Whether the individual commutes during weekends | Binary Categorical | Regressor | `weekend_commuter` |

: Classification table for variables in the dataset. {#tbl-variables}

::: {.panel-tabset}

## R Code

```{.r}
data(multinomial_transport)
data <- multinomial_transport[,-1]
cat("There are", nrow(data), "observations and", ncol(data), "variables in the dataset.")
head(data, n = 100) |>
  print_as_dt()
```

## Python Code

```{.python}
url = "https://raw.githubusercontent.com/andytai7/cookbook/refs/heads/main/raw-data/multinomial_transport.csv"
data = pd.read_csv(url)
data = data.iloc[:, :-1]
print(f"There are {data.shape[0]} observations and {data.shape[1]} variables in the dataset.")
print(data)
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
data(multinomial_transport)
data <- multinomial_transport %>% select(-'income_bracket')
cat("There are", nrow(data), "observations and", ncol(data), "variables in the dataset.")
head(data, n = 100) |>
  print_as_dt()
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
url = "https://raw.githubusercontent.com/andytai7/cookbook/refs/heads/main/raw-data/multinomial_transport.csv"
data = pd.read_csv(url)
data = data.iloc[:, :-1]
print(f"There are {data.shape[0]} observations and {data.shape[1]} variables in the dataset.")
```

```{r}
#| echo: false
#| message: false
#| warning: false
head(data, n = 100) |>
  print_as_dt()
```

:::

Let's begin by factorizing the categorical variables.

::: {.panel-tabset}

## R Code

```{.r}
# Factorize the categorical variables and print their levels
categorical_cols <- c('transport_mode', 'has_car_available', 'weekend_commuter')
for (col in categorical_cols) {
  data[[col]] <- as.factor(data[[col]])
  cat("Levels of", col, ":", levels(data[[col]]), "\n")
}
```

## Python Code

```{.python}
# Factorize the categorical variables and print their levels
categorical_cols = ['transport_mode', 'has_car_available', 'weekend_commuter']
for col in categorical_cols:
  data[col] = data[col].astype('category')
  print(f"Levels of {col}: {list(data[col].cat.categories)}")
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
categorical_cols <- c('transport_mode', 'has_car_available', 'weekend_commuter')
for (col in categorical_cols) {
  data[[col]] <- as.factor(data[[col]])
  cat("Levels of", col, ":", levels(data[[col]]), "\n")
}
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
categorical_cols = ['transport_mode', 'has_car_available', 'weekend_commuter']
for col in categorical_cols:
  data[col] = data[col].astype('category')
  print(f"Levels of {col}: {list(data[col].cat.categories)}")
```

:::

Now, we split the dataset into train and test sets, use train set to fit the model, and use test set to assess the model performance. 

::: {.panel-tabset}

## R Code

```{.r}
set.seed(42)

# Randomly splitting into training and testing sets by transport_mode
data_splitting <- initial_split(data,
  prop = 0.8, strata = transport_mode
)

# Assigning data points to training and testing sets
train_set <- training(data_splitting)
test_set <- testing(data_splitting)

cat("Training set size:", nrow(train_set))
cat("Test set size:", nrow(test_set))
```

## Python Code

```{.python}
# The train/test split is created once in R (caret::createDataPartition; see the
# R tab) and pulled into Python via reticulate's `r` object, so both languages
# use identical train and test sets throughout the chapter.
train_set = r.train_set.reset_index(drop=True)
test_set  = r.test_set.reset_index(drop=True)

print(f"Training set size: {len(train_set)}")
print(f"Test set size: {len(test_set)}")
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
set.seed(42)

# Randomly splitting into training and testing sets by transport_mode
data_splitting <- initial_split(data,
  prop = 0.8, strata = transport_mode
)

# Assigning data points to training and testing sets
train_set <- training(data_splitting)
test_set <- testing(data_splitting)

cat("Training set size:", nrow(train_set))
cat("Test set size:", nrow(test_set))
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
# The train/test split is created once in R (caret::createDataPartition; see the
# R tab) and pulled into Python via reticulate's `r` object, so both languages
# use identical train and test sets throughout the chapter.
train_set = r.train_set.reset_index(drop=True)
test_set  = r.test_set.reset_index(drop=True)

print(f"Training set size: {len(train_set)}")
print(f"Test set size: {len(test_set)}")
```

:::

## Exploratory Data Analysis

Let's visualize the response `transport_mode` and its relationship with the continuous or categorical regressors.

::: {.panel-tabset}

## R Code

```{.r}
# Custom color palette `cbPalette` 
cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
# Sort `transport_mode` by frequency
train_set$transport_mode <- factor(train_set$transport_mode, levels = names(sort(table(train_set$transport_mode), decreasing = TRUE)))

p1 <- train_set %>%
  ggplot(aes(x = transport_mode, fill = has_car_available)) +
  geom_bar(position = position_stack(reverse = TRUE)) +  # use "stack" for counts, "fill" for proportions
  labs(y = "Count", title = "With Car Availability") +
  scale_fill_manual(values=cbPalette) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(title = "Car Availability"))

p2 <- train_set %>%
  ggplot(aes(x = transport_mode, fill = weekend_commuter)) +
  geom_bar(position = position_stack(reverse = TRUE)) +  # use "stack" for counts, "fill" for proportions
  labs(x = "Transport Mode", y = "Count", title = "With Weekend Commuter Status") +
  scale_fill_manual(values=cbPalette) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(title = "Weekend Commuter"))

grid.arrange(p1, p2, ncol = 2, top = grid::textGrob("Distribution of Transport Mode", gp = grid::gpar(fontsize = 16, fontface = "bold")))
```

## Python Code

```{.python}
# Custom color palette `cbPalette` 
cbPalette = ["#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7"]

# Sort transport_mode by frequency
order = train_set['transport_mode'].value_counts().index
train_set['transport_mode'] = pd.Categorical(train_set['transport_mode'], categories=order, ordered=True)

# Set up the plot
sns.set_theme(style="whitegrid")
fig = plt.figure()
fig.set_size_inches(8, 6)
gs = GridSpec(1, 2, figure=fig)
fig.suptitle("Distribution of Transport Mode", fontsize=16, fontweight='bold')

# Plot: Stacked bar for Car Availability
car_counts = train_set.groupby(['transport_mode', 'has_car_available']).size().unstack(fill_value=0)
car_counts = car_counts.loc[order] 

ax1 = fig.add_subplot(gs[0, 0])
bottom = None
for i, col in enumerate(car_counts.columns):
    ax1.bar(car_counts.index, car_counts[col], bottom=bottom, label=col, color=cbPalette[i % len(cbPalette)])
    bottom = car_counts[col] if bottom is None else bottom + car_counts[col]
ax1.set_title("With Car Availability")
ax1.set_xlabel("Transport Mode")
ax1.set_ylabel("Count")
ax1.legend(title="Car Availability", loc='upper center', bbox_to_anchor=(0.5, -0.15), ncol=2)

# Plot: Stacked bar for Weekend Commuter
weekend_counts = train_set.groupby(['transport_mode', 'weekend_commuter']).size().unstack(fill_value=0)
weekend_counts = weekend_counts.loc[order]

ax2 = fig.add_subplot(gs[0, 1])
bottom = None
for i, col in enumerate(weekend_counts.columns):
    ax2.bar(weekend_counts.index, weekend_counts[col], bottom=bottom, label=col, color=cbPalette[i % len(cbPalette)])
    bottom = weekend_counts[col] if bottom is None else bottom + weekend_counts[col]
ax2.set_title("With Weekend Commuter Status")
ax2.set_xlabel("Transport Mode")
ax2.set_ylabel("")
ax2.legend(title="Weekend Commuter", loc='upper center', bbox_to_anchor=(0.5, -0.15), ncol=2)

# Adjust plot layout
fig.tight_layout(rect=[0, 0.05, 1, 1])
fig
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| label: fig-eda-transport-mode-distribution-R
#| fig-cap:
#|   - "Stacked bar chart of transport mode counts by car availability."
#|   - "Stacked bar chart of transport mode counts by weekend commuter status."
#| echo: false
#| message: false
#| warning: false
cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
train_set$transport_mode <- factor(train_set$transport_mode, levels = names(sort(table(train_set$transport_mode), decreasing = TRUE)))
p1 <- train_set %>%
  ggplot(aes(x = transport_mode, fill = has_car_available)) +
  geom_bar(position = position_stack(reverse = TRUE)) +  # use "stack" for counts, "fill" for proportions
  labs(x = "Transport Mode", y = "Count", title = "With Car Availability") +
  scale_fill_manual(values=cbPalette) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(title = "Car Availability"))
p2 <- train_set %>%
  ggplot(aes(x = transport_mode, fill = weekend_commuter)) +
  geom_bar(position = position_stack(reverse = TRUE)) +  # use "stack" for counts, "fill" for proportions
  labs(x = "Transport Mode", y = "", title = "With Weekend Commuter Status") +
  scale_fill_manual(values=cbPalette) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(title = "Weekend Commuter"))
grid.arrange(p1, p2, ncol = 2, top = grid::textGrob("Distribution of Transport Mode", gp = grid::gpar(fontsize = 16, fontface = "bold")))
```

## Python Output

```{python}
#| label: fig-eda-transport-mode-distribution-py
#| fig-cap:
#|   - "Stacked bar chart of transport mode counts by car availability."
#|   - "Stacked bar chart of transport mode counts by weekend commuter status."
#| echo: false
#| message: false
#| warning: false
cbPalette = ["#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7"]
order = train_set['transport_mode'].value_counts().index
train_set['transport_mode'] = pd.Categorical(train_set['transport_mode'], categories=order, ordered=True)

sns.set_theme(style="whitegrid")
fig = plt.figure()
fig.set_size_inches(8, 6)
gs = GridSpec(1, 2, figure=fig)
fig.suptitle("Distribution of Transport Mode", fontsize=16, fontweight='bold')

# Plot 1: Stacked bar for Car Availability
car_counts = train_set.groupby(['transport_mode', 'has_car_available']).size().unstack(fill_value=0)
car_counts = car_counts.loc[order] 
ax1 = fig.add_subplot(gs[0, 0])
bottom = None
for i, col in enumerate(car_counts.columns):
    ax1.bar(car_counts.index, car_counts[col], bottom=bottom, label=col, color=cbPalette[i % len(cbPalette)])
    bottom = car_counts[col] if bottom is None else bottom + car_counts[col]
ax1.set_title("With Car Availability")
ax1.set_xlabel("Transport Mode")
ax1.set_ylabel("Count")
ax1.legend(title="Car Availability", loc='upper center', bbox_to_anchor=(0.5, -0.15), ncol=2)

# Plot 2: Stacked bar for Weekend Commuter
weekend_counts = train_set.groupby(['transport_mode', 'weekend_commuter']).size().unstack(fill_value=0)
weekend_counts = weekend_counts.loc[order]
ax2 = fig.add_subplot(gs[0, 1])
bottom = None
for i, col in enumerate(weekend_counts.columns):
    ax2.bar(weekend_counts.index, weekend_counts[col], bottom=bottom, label=col, color=cbPalette[i % len(cbPalette)])
    bottom = weekend_counts[col] if bottom is None else bottom + weekend_counts[col]
ax2.set_title("With Weekend Commuter Status")
ax2.set_xlabel("Transport Mode")
ax2.set_ylabel("")
ax2.legend(title="Weekend Commuter", loc='upper center', bbox_to_anchor=(0.5, -0.15), ncol=2)

# Adjust layout
fig.tight_layout(rect=[0, 0.05, 1, 1])
fig
```

:::

The bar plots display the distribution of transport mode across car availability and weekend commute status. 

The blue bar is much taller than the orange bar, which means that car is the dominant mode of transport, especially among those who have access to a car. Public Transit is used by both those with and without car availability, but is more common among those without a car (orange section is relatively larger). 
Bicycle and Walking are rare overall but tend to occur more when no car is available. 

Most weekend commuters do not significantly change their transport mode: the proportions are largely dominated by the non-weekend commuter group (orange).
However, Public Transit has a noticeable number of weekend commuters (blue), which indicates some weekend reliance on public transit.
Walking and Bicycling show little weekend commuting activity.

Now let's visualize the relationship between `transport_mode` and continuous regressors `commute_distance` and `age`.

::: {.panel-tabset}

## R Code

```{.r}
# visualize boxplots and violin plots of commute_distance v.s. transport_mode
p3 <- train_set %>%
    ggplot(aes(x = transport_mode, y = commute_distance, fill = transport_mode)) +
    geom_violin(alpha = 0.6, trim = FALSE) +
    stat_boxplot(geom = "errorbar", width = 0.2) +
    geom_boxplot(alpha = 1, fill = NA) +
    labs(x = "Transport Mode", y = "Commute Distance (km)", title =
        "Commute Distance by Transport Mode") +
    scale_fill_manual(values = cbPalette[3:6]) +
    theme_bw() +
    theme(legend.position = "none") +
    guides(fill = guide_legend(title = "Transport Mode"))

p4 <- train_set %>%
    ggplot(aes(x = transport_mode, y = age, fill = transport_mode)) +
    geom_violin(alpha = 0.6, trim = FALSE) +
    stat_boxplot(geom = "errorbar", width = 0.2) +
    geom_boxplot(alpha = 1, fill = NA) +
    labs(x = "Transport Mode", y = "Age", title =
        "Age by Transport Mode") +
    scale_fill_manual(values = cbPalette[3:6]) +
    theme_bw() +
    theme(legend.position = "none") +
    guides(fill = guide_legend(title = "Transport Mode"))

grid.arrange(p3, p4, ncol = 2, top = grid::textGrob("Box Plots and Violin Plots", gp = grid::gpar(fontsize = 16, fontface = "bold")))
```

## Python Code

```{.python}
sns.set_theme(style="whitegrid")

# Set up side-by-side subplots with shared layout
fig = plt.figure(figsize=(8, 7))
gs = GridSpec(1, 2, figure=fig)
fig.suptitle("Box Plots and Violin Plots", fontsize=16, fontweight='bold')

# Plot Commute Distance vs Transport Mode
ax1 = fig.add_subplot(gs[0, 0])
sns.violinplot(
    data=train_set,
    x='transport_mode',
    y='commute_distance',
    palette=cbPalette[2:6],
    ax=ax1,
    alpha=0.6,
    inner=None,
    order=order,
    cut=3
)
sns.boxplot(
    data=train_set,
    x='transport_mode',
    y='commute_distance',
    showcaps=True,
    boxprops={'facecolor': 'none', 'edgecolor': 'black'},
    whiskerprops={'color': 'black'},
    medianprops={'color': 'black', 'linewidth': 2},
    flierprops={'markerfacecolor': 'black', 'markersize': 3},
    ax=ax1,
    order=order
)
ax1.set_title("Commute Distance by Transport Mode")
ax1.set_xlabel("Transport Mode")
ax1.set_ylabel("Commute Distance (km)")
ax1.get_legend()

# Plot Age vs Transport Mode
ax2 = fig.add_subplot(gs[0, 1])
sns.violinplot(
    data=train_set,
    x='transport_mode',
    y='age',
    palette=cbPalette[2:6],
    ax=ax2,
    alpha=0.6,
    inner=None,
    order=order,
    cut=3
)
sns.boxplot(
    data=train_set,
    x='transport_mode',
    y='age',
    showcaps=True,
    boxprops={'facecolor': 'none', 'edgecolor': 'black'},
    whiskerprops={'color': 'black'},
    medianprops={'color': 'black', 'linewidth': 2},
    flierprops={'markerfacecolor': 'black', 'markersize': 3},
    ax=ax2,
    order=order
)
ax2.set_title("Age by Transport Mode")
ax2.set_xlabel("Transport Mode")
ax2.set_ylabel("Age")
ax2.get_legend()

fig.tight_layout(rect=[0, 0.05, 1, 0.95])
fig
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| label: fig-eda-transport-mode-continuous-R
#| fig-cap:
#|   - "Boxplots and violin plots of commute distance and age by transport mode."
#| echo: false
#| message: false
#| warning: false
p3 <- train_set %>%
    ggplot(aes(x = transport_mode, y = commute_distance, fill = transport_mode)) +
    geom_violin(alpha = 0.6, trim = FALSE) +
    stat_boxplot(geom = "errorbar", width = 0.2) +
    geom_boxplot(alpha = 1, fill = NA) +
    labs(x = "Transport Mode", y = "Commute Distance (km)", title =
        "Commute Distance by Transport Mode") +
    scale_fill_manual(values = cbPalette[3:6]) +
    theme_bw() +
    theme(legend.position = "none") +
    guides(fill = guide_legend(title = "Transport Mode"))
p4 <- train_set %>%
    ggplot(aes(x = transport_mode, y = age, fill = transport_mode)) +
    geom_violin(alpha = 0.6, trim = FALSE) +
    stat_boxplot(geom = "errorbar", width = 0.2) +
    geom_boxplot(alpha = 1, fill = NA) +
    labs(x = "Transport Mode", y = "Age", title =
        "Age by Transport Mode") +
    scale_fill_manual(values = cbPalette[3:6]) +
    theme_bw() +
    theme(legend.position = "none") +
    guides(fill = guide_legend(title = "Transport Mode"))
grid.arrange(p3, p4, ncol = 2, top = grid::textGrob("Box Plots and Violin Plots", gp = grid::gpar(fontsize = 16, fontface = "bold")))
```

## Python Output

```{python}
#| label: fig-eda-transport-mode-continuous-py
#| fig-cap:
#|   - "Boxplots and violin plots of commute distance and age by transport mode."
#| echo: false
#| message: false
#| warning: false
sns.set_theme(style="whitegrid")

# Set up side-by-side subplots with shared layout
fig = plt.figure(figsize=(8, 7))
gs = GridSpec(1, 2, figure=fig)
fig.suptitle("Box Plots and Violin Plots", fontsize=16, fontweight='bold')

# Plot Commute Distance vs Transport Mode
ax1 = fig.add_subplot(gs[0, 0])
sns.violinplot(
    data=train_set,
    x='transport_mode',
    y='commute_distance',
    palette=cbPalette[2:6],
    ax=ax1,
    alpha=0.6,
    inner=None,
    order=order,
    cut=3
)
sns.boxplot(
    data=train_set,
    x='transport_mode',
    y='commute_distance',
    showcaps=True,
    boxprops={'facecolor': 'none', 'edgecolor': 'black'},
    whiskerprops={'color': 'black'},
    medianprops={'color': 'black', 'linewidth': 2},
    flierprops={'markerfacecolor': 'black', 'markersize': 3},
    ax=ax1,
    order=order
)
ax1.set_title("Commute Distance by Transport Mode")
ax1.set_xlabel("Transport Mode")
ax1.set_ylabel("Commute Distance (km)")
ax1.get_legend()  # no legend to remove manually

# Plot p4: Age vs Transport Mode
ax2 = fig.add_subplot(gs[0, 1])
sns.violinplot(
    data=train_set,
    x='transport_mode',
    y='age',
    palette=cbPalette[2:6],
    ax=ax2,
    alpha=0.6,
    inner=None,
    order=order,
    cut=3
)
sns.boxplot(
    data=train_set,
    x='transport_mode',
    y='age',
    showcaps=True,
    boxprops={'facecolor': 'none', 'edgecolor': 'black'},
    whiskerprops={'color': 'black'},
    medianprops={'color': 'black', 'linewidth': 2},
    flierprops={'markerfacecolor': 'black', 'markersize': 3},
    ax=ax2,
    order=order
)
ax2.set_title("Age by Transport Mode")
ax2.set_xlabel("Transport Mode")
ax2.set_ylabel("Age")
ax2.get_legend()  # no legend to remove manually

fig.tight_layout(rect=[0, 0.05, 1, 0.95])
fig
```

:::

Box and violin plots show how commute distance and age vary across different transport modes.

Public transit users tend to have the longest commute distance, on average, with a wide distribution and a median around 17–18 km.
Car users also have moderately long commutes, with a median around 10-11 km and a relatively broad spread.
Bicycle and Walking are associated with shorter commute distances (medians < 5 km), and their distributions are tightly concentrated near the lower end.
All modes show some long-distance commuters (long tails), especially public transit and car.

Car and public transit users have similar age distributions, with median ages in the mid-30s to early 40s.
Bicycle users skew slightly younger, but with a long tail of older users.
Walking has a much wider spread in age, with a high median (around 60) and some very young walkers as well.
The variability in age is greatest for walking, suggesting diverse usage across age groups.

You can also visualize the relationship across other regressors in various types of figures for practice.

## Data Modelling

A Multinomial Logistic Regression model is a suitable approach to our statistical inquiries given that `transport_mode` is categorical and nominal (our response of interest) subject to the numerical regressors `commute_distance` and `age`, and binary regressors `has_car_available` and `weekend_commuter`. 

This regression approach assumes a [Multinomial distribution](https://www.britannica.com/science/multinomial-distribution) where $p_{i,1},p_{i,2},\dots,p_{i,m}$ are the probabilities that will belong to categories $1,2,\dots,m$ respectively; i.e., 
$$P(Y_i=1)=p_{i,1}, P(Y_i=2)=p_{i,2}, \dots, P(Y_i=m)=p_{i,m}$$ {#eq-multinomial}
where
$$
\sum_{j = 1}^m p_{i,j} = p_{i,1} + p_{i,2} + \dots + p_{i,m} = 1.
$$

*A particular highlight is that the Binomial distribution is a special case of the Multinomial distribution with $m=2$.*

The Multinomial Logistic regression also models the **logits**, which are logarithm of the odds. However, only one logit will not be enough anymore. Recall that we can capture the odds between two categories with a single logit function. What about adding some other ones?

Here is what we can do:

1. Pick one of the categories to be the **baseline**.
2. For each of the other categories, we model the logarithm of the odds to the baseline category.

Now, **what is the math for the general case with $m$ response categories and $K$ regressors?** For the $i$th observation, we end up with a system of $m - 1$ **logit link** functions:

$$
\begin{gathered}
\eta_i^{(2,1)} = \log\left[\frac{P(Y_i = 2\mid x_{i,1}, \ldots, x_{i,K})}{P(Y_i = 1 \mid x_{i,1}, \ldots, x_{i,K})}\right] = \beta_0^{(2,1)} + \beta_1^{(2,1)} x_{i, 1} + \beta_2^{(2,1)} x_{i, 2} + \ldots + \beta_K^{(2,1)} x_{i, k} \\
\eta_i^{(3,1)} = \log\left[\frac{P(Y_i = 3\mid x_{i,1}, \ldots, x_{i,K})}{P(Y_i = 1 \mid x_{i,1}, \ldots, x_{i,K})}\right] = \beta_0^{(3,1)} + \beta_1^{(3,1)} x_{i, 1} + \beta_2^{(3,1)} x_{i, 2} + \ldots + \beta_K^{(3,1)} x_{i, k} \\
\vdots \\
\eta_i^{(m,1)} = \log\left[\frac{P(Y_i = m\mid x_{i,1}, \ldots, x_{i,K})}{P(Y_i = 1 \mid x_{i,1}, \ldots, x_{i,K})}\right] = \beta_0^{(m,1)} + \beta_1^{(m,1)} x_{i, 1} + \beta_2^{(m,1)} x_{i, 2} + \ldots + \beta_K^{(m,1)} x_{i, k}.
\end{gathered}
$$ {#eq-multinomial-model}

Note that the superscript $(j, 1)$ in @eq-multinomial-model indicates that the equation is on level $j$ (for $j = 2, \dots, m$) with respect to level $1$. Furthermore, the regression coefficients are different for each link function.

Each of the logit-linear functions in @eq-multinomial-model writes the log of odds between each category $j=2,\dots,m$ and the baseline category $j=1$ as a linear combination of the regressors. To compare between the categories $j=2,\dots,m$, we can simply deduct one equation be another, e.g.,
$$
\begin{gather*} \label{eq:multinomial-deduct}
\eta_i^{(2,1)} - \eta_i^{(3,1)} = \log\left[\frac{P(Y_i = 2\mid x_{i,1}, \ldots, x_{i,K})}{P(Y_i = 1 \mid x_{i,1}, \ldots, x_{i,K})}\right] - \log\left[\frac{P(Y_i = 3\mid x_{i,1}, \ldots, x_{i,K})}{P(Y_i = 1 \mid x_{i,1}, \ldots, x_{i,K})}\right] \\
= \log\left[\frac{P(Y_i = 3\mid x_{i,1}, \ldots, x_{i,K})}{P(Y_i = 2 \mid x_{i,1}, \ldots, x_{i,K})}\right] 
= (\beta_0^{(2,1)}-\beta_0^{(3,1)}) + (\beta_1^{(2,1)}-\beta_1^{(3,1)}) x_{i, 1} + (\beta_2^{(2,1)}-\beta_2^{(3,1)}) x_{i, 2} + \ldots
\end{gather*}
$$

With some algebraic manipulation, we can show that the probabilities $p_{i,1}, p_{i,2}, \dots, p_{i,m}$ of $Y_i$ belonging to categories $1, 2, \dots, m$ are:

$$ 
\begin{gathered}
p_{i,1} = P(Y_i = 1 \mid x_{i,1}, \ldots, x_{i,K}) = \frac{1}{1 + \sum_{j = 2}^m \exp \big( \eta_i^{(j,1)} \big)} \\
p_{i,2} = P(Y_i = 2 \mid x_{i,1}, \ldots, x_{i,K}) = \frac{\exp \big( \eta_i^{(2,1)} \big)}{1 + \sum_{j = 2}^m \exp \big( \eta_i^{(j,1)} \big)} \\
\vdots \\
p_{i,m} = P(Y_i = m \mid x_{i,1}, \ldots, x_{i,K}) = \frac{\exp \big( \eta_i^{(m,1)} \big)}{1 + \sum_{j = 2}^m \exp \big( \eta_i^{(j,1)} \big)}.
\end{gathered}
$$ {#eq-prob-multinomial}

***Note: The equations above are Softmax functions. The multinomial logistic regression is theoretically equivalent to a one-layer neural network with a Softmax activation function, where we input $X$, output probability $p_{i,j}$, and minimize the cross-entropy loss function to train the neural network to learn parameters $\beta$.***

***If we add all $m$ probabilities in @eq-prob-multinomial, the sum is equal to $1$ for the $i$th observation. This is particularly important when we want to use this model for making predictions in classification matters.***

Let's set up a multinomial logistic regression model with `transport_mode` as the response with four classes: `Car`, `Public Transit`, `Bicycle`, and `Walking`, subject to the continuous regressors `commute_distance` and `age`, denoted as $x_{i,\texttt{distance}}$ and $x_{i,\texttt{age}}$ respectively, and binary regressors `has_car_available` and `weekend_commuter`, with the dummy variables of class `Yes` denoted as $Z_{i,\texttt{car}}$ and $Z_{i,\texttt{weekend}}$ respectively. We will use `Car` as the baseline level for `transport_mode`. The equation of each link function for this multinomial model is given below, such that each link function has its own intercept and a set of regression coefficients.

$$
\begin{align*}
\eta_i^{(\texttt{Public},\texttt{Car})} &= \log\left[\frac{P(Y_i = \texttt{Public} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}{P(Y_i = \texttt{Car} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}\right] \\
&= \beta_0^{(\texttt{Public},\texttt{Car})} + \beta_1^{(\texttt{Public},\texttt{Car})} x_{i, \texttt{distance}} + \beta_2^{(\texttt{Public},\texttt{Car})} x_{i, \texttt{age}} + \beta_3^{(\texttt{Public},\texttt{Car})} Z_{i, \texttt{car}} + \beta_4^{(\texttt{Public},\texttt{Car})} Z_{i,\texttt{weekend}}, \\
\eta_i^{(\texttt{Bike},\texttt{Car})} &= \log\left[\frac{P(Y_i = \texttt{Bike} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}{P(Y_i = \texttt{Car} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}\right] \\
&= \beta_0^{(\texttt{Bike},\texttt{Car})} + \beta_1^{(\texttt{Bike},\texttt{Car})} x_{i, \texttt{distance}} + \beta_2^{(\texttt{Bike},\texttt{Car})} x_{i, \texttt{age}} + \beta_3^{(\texttt{Bike},\texttt{Car})} Z_{i, \texttt{car}} + \beta_4^{(\texttt{Bike},\texttt{Car})} Z_{i,\texttt{weekend}}, \\
\eta_i^{(\texttt{Walk},\texttt{Car})} &= \log\left[\frac{P(Y_i = \texttt{Walk} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}{P(Y_i = \texttt{Car} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}\right] \\
&= \beta_0^{(\texttt{Walk},\texttt{Car})} + \beta_1^{(\texttt{Walk},\texttt{Car})} x_{i, \texttt{distance}} + \beta_2^{(\texttt{Walk},\texttt{Car})} x_{i, \texttt{age}} + \beta_3^{(\texttt{Walk},\texttt{Car})} Z_{i, \texttt{car}} + \beta_4^{(\texttt{Walk},\texttt{Car})} Z_{i,\texttt{weekend}} .
\end{align*}
$$

Taking exponential on both sides of the model equations gives the ratio of probability of each category `Public Transit`, `Bicycle`, and `Walking` over the probability of the baseline level `Car`:

$$
\begin{align*}
&\frac{P(Y_i = \texttt{Public} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}{P(Y_i = \texttt{Car} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})} = \exp\left[\eta_i^{(\texttt{Public},\texttt{Car})} \right]\\
&= \exp\left[\beta_0^{(\texttt{Public},\texttt{Car})}\right] \exp\left[\beta_1^{(\texttt{Public},\texttt{Car})} X_{i,\texttt{distance}}\right] \exp\left[\beta_2^{(\texttt{Public},\texttt{Car})} X_{i,\texttt{age}}\right] \exp\left[ \beta_3^{(\texttt{Public},\texttt{Car})} Z_{i, \texttt{car}}\right] \exp\left[ \beta_4^{(\texttt{Public},\texttt{Car})} Z_{i,\texttt{weekend}}\right]\\
&\frac{P(Y_i = \texttt{Bike} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}{P(Y_i = \texttt{Car} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})} = \exp\left[\eta_i^{(\texttt{Bike},\texttt{Car})} \right]\\
&= \exp\left[\beta_0^{(\texttt{Bike},\texttt{Car})}\right] \exp\left[\beta_1^{(\texttt{Bike},\texttt{Car})} X_{i,\texttt{distance}}\right] \exp\left[\beta_2^{(\texttt{Bike},\texttt{Car})} X_{i,\texttt{age}}\right] \exp\left[ \beta_3^{(\texttt{Bike},\texttt{Car})} Z_{i, \texttt{car}}\right] \exp\left[ \beta_4^{(\texttt{Bike},\texttt{Car})} Z_{i,\texttt{weekend}}\right]\\
&\frac{P(Y_i = \texttt{Walk} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})}{P(Y_i = \texttt{Car} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}})} = \exp\left[\eta_i^{(\texttt{Walk},\texttt{Car})} \right]\\
&= \exp\left[\beta_0^{(\texttt{Walk},\texttt{Car})}\right] \exp\left[\beta_1^{(\texttt{Walk},\texttt{Car})} X_{i,\texttt{distance}}\right] \exp\left[\beta_2^{(\texttt{Walk},\texttt{Car})} X_{i,\texttt{age}}\right] \exp\left[ \beta_3^{(\texttt{Walk},\texttt{Car})} Z_{i, \texttt{car}}\right] \exp\left[ \beta_4^{(\texttt{Walk},\texttt{Car})} Z_{i,\texttt{weekend}}\right]
\end{align*}
$$

Finally, the probability of $Y_i$ belonging to categories `Car`, `Public Transit`, `Bicycle`, and `Walking` respectively are given by the following expressions.

$$
\begin{align*}
P(Y_i = \texttt{Car} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}}) &= \frac{1}{1+\exp\left[\eta_i^{(\texttt{Public},\texttt{Car})} \right]+\exp\left[\eta_i^{(\texttt{Bike},\texttt{Car})}\right]+\exp\left[\eta_i^{(\texttt{Walk},\texttt{Car})} \right]} \\
P(Y_i = \texttt{Public} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}}) &= \frac{\exp\left[\eta_i^{(\texttt{Public},\texttt{Car})} \right]}{1+\exp\left[\eta_i^{(\texttt{Public},\texttt{Car})} \right]+\exp\left[\eta_i^{(\texttt{Bike},\texttt{Car})}\right]+\exp\left[\eta_i^{(\texttt{Walk},\texttt{Car})} \right]} \\
P(Y_i = \texttt{Bike} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}}) &= \frac{\exp\left[\eta_i^{(\texttt{Bike},\texttt{Car})}\right]}{1+\exp\left[\eta_i^{(\texttt{Public},\texttt{Car})} \right]+\exp\left[\eta_i^{(\texttt{Bike},\texttt{Car})}\right]+\exp\left[\eta_i^{(\texttt{Walk},\texttt{Car})} \right]}\\
P(Y_i = \texttt{Walk} \mid x_{i, \texttt{distance}}, x_{i, \texttt{age}}, Z_{i, \texttt{car}}, Z_{i,\texttt{weekend}}) &= \frac{\exp\left[\eta_i^{(\texttt{Walk},\texttt{Car})} \right]}{1+\exp\left[\eta_i^{(\texttt{Public},\texttt{Car})} \right]+\exp\left[\eta_i^{(\texttt{Bike},\texttt{Car})}\right]+\exp\left[\eta_i^{(\texttt{Walk},\texttt{Car})} \right]}
\end{align*}
$$ {#eq-probabilities}


### The Log-Likelihood Function

A multinomial logistic regression model treats the categories in the response variable as nominal, which means that there is no inherent ranking or ordering among them; therefore, each category is compared directly against a single baseline category. Recall from @eq-prob-multinomial that this comparison gives us the category-specific probabilities in closed form,

$$
p_{i,j} = P(Y_i = j \mid x_{i,1}, \ldots, x_{i,K}), \qquad j = 1, \ldots, m,
$$

where each non-baseline category $j$ enters through its own logit $\eta_i^{(j,1)}$ and the whole set is normalized by the softmax denominator, so that $\sum_{j=1}^m p_{i,j} = 1$. These probabilities are exactly what the likelihood is built from.

***Note: Changing the baseline category, e.g., switching from `Car` to `Public` would not change the estimated probability for each category for each individual.***

To estimate the model, we need a way to measure how well a given set of coefficients explains the observed data. The **likelihood function** does exactly this by expressing the probability of the data we actually observed as a function of the unknown coefficients, and finding the coefficient values that make this probability as large as possible.

We first encode each individual's observed response category using indicator variables. Following the notation in @eq-multinomial-model and @eq-prob-multinomial, let $y_{i,j} = 1$ if the $i$-th individual belongs to category $j$, and let $y_{i,j} = 0$ otherwise. Since each individual belongs to exactly one of the $m$ possible categories, the vector $\mathbf{y}_i = (y_{i,1}, y_{i,2}, \ldots, y_{i,m})$ contains exactly one entry equal to $1$, with all remaining entries equal to $0$. This representation is often called a one-hot encoding of the response category. The likelihood function for the multinomial logistic regression model is then

$$
\mathcal{L}(\boldsymbol{\beta}; \mathbf{y}) = \prod_{i=1}^n \prod_{j=1}^m p_{i,j}^{y_{i,j}},
$$ {#eq-likelihood-multinomial}

where $n$ is the number of individuals in the sample, $m$ is the number of categories, and $p_{i,j}$ is the model-implied softmax probability that individual $i$ (with regressor values $x_{i,1}, \ldots, x_{i,K}$) falls in category $j$. The unknown parameters are collected in $\boldsymbol{\beta}$, the stack of all reference-category coefficients $\beta_k^{(j,1)}$ for $j = 2, \ldots, m$ and $k = 0, 1, \ldots, K$, such that it consists of an intercept and $K$ slopes for each of the $m-1$ non-baseline categories.

The two products have intuitive roles. The **inner product** over $j$ runs across the $m$ categories for a single individual: because of the indicator $y_{i,j}$ in the exponent, every term equals $1$ except the one matching the individual's true category, so the inner product collapses to just $p_{i,j}$ for that observed category. The **outer product** over $i$ then multiplies these per-individual probabilities together across all $n$ people, which is valid because the observations are assumed independent. Overall, this is a **multinomial likelihood** evaluated at the softmax category probabilities.

Taking the logarithm turns the products into sums, which is far easier to differentiate and optimize, and yields the **log-likelihood function**:

$$
\ell(\boldsymbol{\beta}; \mathbf{y}) = \sum_{i=1}^n \sum_{j=1}^m y_{i,j} \log p_{i,j}.
$$ {#eq-log-likelihood-multinomial}

Substituting the softmax probabilities from @eq-prob-multinomial and using the reference-category constraint $y_{i,1} = 1 - \sum_{j=2}^m y_{i,j}$, the log-likelihood can be written explicitly in terms of the coefficients as

$$
\ell(\boldsymbol{\beta}) = \sum_{i=1}^n \left[ \sum_{j=2}^m y_{i,j} \sum_{k=0}^K X_{i,k} \beta_k^{(j,1)} - \log\left( 1 + \sum_{j=2}^m e^{\sum_{k=0}^K X_{i,k} \beta_k^{(j,1)}} \right) \right],
$$

where $X_{i,0} = 1$ corresponds to the intercept in the linear model.

Maximizing $\ell$ gives the estimates $\hat{\boldsymbol{\beta}}$, the coefficient values under which the observed data are most probable. There is, however, no closed-form algebraic expression for the maximizer: the category probabilities $p_{i,j}$ enter the log-likelihood non-linearly through the softmax function, so setting the derivatives to zero does not yield equations we can solve directly. Instead, we rely on **iterative numerical optimization**, which starts from an initial guess and refines the estimates step by step until the log-likelihood can no longer be improved. The standard choice is the **Newton–Raphson** method, described in the tip box below; the variant most statistical software uses replaces the observed Hessian with its expected value (the Fisher information), giving the **Fisher scoring** algorithm, which is equivalent to **iteratively reweighted least squares (IRLS)**. Additional estimation details, including the generalization from one to $n_i$ observations per population $i$, can be found in Section 2.2 of [Czepiel (2002)](https://czep.net/stat/mlelr.pdf).

::: {.Tip}
::::{.Tip-header}
Tip on the Newton–Raphson Method
::::
::::{.Tip-container}
The parameters we ultimately estimate are the reference-category coefficients, collected into the stacked vector $\boldsymbol{\beta} \in \mathbb{R}^q$ with $q = (K+1)(m-1)$ (an intercept and $K$ slopes for each of the $m-1$ non-baseline categories). Since the log-likelihood has no closed-form maximizer, the Newton–Raphson procedure estimates $\boldsymbol{\beta}$ by iterating with the score function and Hessian matrix of the log-likelihood, in a compact matrix form known as multinomial iteratively reweighted least squares (IRLS).

**Score function:** For category $j$ and regressor $k$, the corresponding component of the score vector $S(\boldsymbol{\beta}) \in \mathbb{R}^q$ (the first-order partial derivative of $\ell$) is

$$
\frac{\partial \ell(\boldsymbol{\beta})}{\partial \beta_k^{(j,1)}} = \sum_{i=1}^n \left( y_{i,j} - p_{i,j} \right) X_{i,k}.
$$

Each component compares the observed indicator $y_{i,j}$ with the model-implied probability $p_{i,j}$, weighted by the regressor value $X_{i,k}$; at the maximum, these differences balance out so that the score is zero.

**Hessian matrix:** The Hessian $H(\boldsymbol{\beta}) \in \mathbb{R}^{q \times q}$ collects the second-order partial derivatives of $\ell$. For the row indexed by category $j$ and regressor $k$, and the column indexed by category $j'$ and regressor $k'$, its entry is

$$
\frac{\partial^2 \ell(\boldsymbol{\beta})}{\partial \beta_k^{(j,1)} \partial \beta_{k'}^{(j',1)}} =
\begin{cases}
-\sum_{i=1}^n X_{i,k}\, p_{i,j}(1 - p_{i,j})\, X_{i,k'} & \text{if } j' = j, \\
\phantom{-}\sum_{i=1}^n X_{i,k}\, p_{i,j}\, p_{i,j'}\, X_{i,k'} & \text{if } j' \neq j,
\end{cases}
$$

for $j, j' = 2, \ldots, m$ and $k, k' = 0, \ldots, K$. As the non-baseline categories are coupled through the shared softmax denominator, the Hessian is a dense $q \times q$ matrix rather than block-diagonal. For a thorough treatment of the Hessian in the context of maximum likelihood estimation, refer to Chapter 4 of @agresti2013.

**Iteration:** Collecting the score and Hessian above, the Newton–Raphson update at step $t$ is

$$
\boldsymbol{\beta}^{(t+1)} \leftarrow \boldsymbol{\beta}^{(t)} - \left[ H\!\left( \boldsymbol{\beta}^{(t)} \right) \right]^{-1} S\!\left( \boldsymbol{\beta}^{(t)} \right),
$$

initialized at $\boldsymbol{\beta}^{(0)} = \mathbf{0}$ (or a data-driven starting point) and repeated until changes in $\ell$ and $\boldsymbol{\beta}$ fall below a convergence tolerance.

The full Newton–Raphson/IRLS procedure is:

1. Choose the baseline category and initialize $\boldsymbol{\beta}^{(0)} = \mathbf{0}$ (stacked over all $(j,k)$).
2. Repeat until convergence:
   - Compute the logits $\eta_i^{(j,1)}$ and category probabilities $p_{i,j}$ for all individuals $i$ and categories $j$, and evaluate the log-likelihood $\ell(\boldsymbol{\beta}^{(t)})$.
   - Form the stacked score $S(\boldsymbol{\beta}^{(t)})$ and Hessian $H(\boldsymbol{\beta}^{(t)})$ using the expressions above.
   - Update $\boldsymbol{\beta}^{(t+1)}$ using the Newton–Raphson step above.
   - Check the convergence criteria.
::::
:::

## Estimation

### Fitting the Model

We now turn to the question of how the parameters of the multinomial logistic regression model are estimated from data. We summarize the key estimation steps in mathematical terms below, working with one observation at a time before generalizing to the full sample.

Recall that the model for a response $Y$ with $m$ categories and $K$ regressors is given by

$$
\log \left( \frac{\Pr(Y_i = k)}{\Pr(Y_i = 1)} \right) = \beta_{0}^{(k,1)} + \beta_{1}^{(k,1)} x_{i,1} + \beta_{2}^{(k,1)} x_{i,2} + ... + \beta_{K}^{(k,1)} x_{i,K},\,\,\,\,k=2,...,m
$$ {#eq-multinomial-model}

where $\beta_{0}^{(2,1)}$, ..., $\beta_{0}^{(m,1)}$ are category-specific intercepts, and $\beta_{1}^{(k,1)}$, $\beta_{2}^{(k,1)}$, ..., $\beta_{K}^{(k,1)}$ are the slope coefficients corresponding to the $k$-th category of $Y$.

The model therefore comprises $(m-1)\times (K+1)$ parameters: $m-1$ intercepts and $(m-1)\times K$ regression coefficients. 

We now fit the multinomial logistic regression model using the `nnet` package in `R` or the `statsmodel` package in `Python`. 

::: {.panel-tabset}

## R Code

```{.r}
# Set the baseline model
train_set$transport_mode <- relevel(train_set$transport_mode, ref = "Car")

# Fit multinomial logistic regression
multinom_fit <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = train_set)

# Print the fitted model
print(summary(multinom_fit))
```

## Python Code

```{.python}
# Factorize the response and set the baseline model
train_set["transport_mode"] = pd.Categorical(
    train_set["transport_mode"],
    categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)

# Design matrix and response
y = train_set["transport_mode"]
X = train_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]

# Add intercept
X = sm.add_constant(X)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_numeric = X.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes

# Fit multinomial logistic regression
model = sm.MNLogit(y, X_numeric)
multinom_fit = model.fit()

# Print the fitted model
print(multinom_fit.summary())
```

:::

::: {.panel-tabset}

### R Output

```{r}
#| echo: false
#| message: false
#| warning: false
train_set$transport_mode <- relevel(train_set$transport_mode, ref = "Car")
multinom_fit <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = train_set)
#print(t(round(summary(multinom_fit)$coefficients, 3)))
print(summary(multinom_fit))
```

### Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
train_set["transport_mode"] = pd.Categorical(
    train_set["transport_mode"],
    categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)
y = train_set["transport_mode"]
X = train_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]
X = sm.add_constant(X)
X_numeric = X.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes
model = sm.MNLogit(y, X_numeric)
multinom_fit = model.fit()
print(multinom_fit.summary())
```

:::

The output reports the estimated parameters: for each non-baseline category $k$, a category-specific intercept $\hat{\beta}_{0}^{(k,1)}$ together with the slope coefficients $\hat{\beta}_{1}^{(k,1)}, \ldots, \hat{\beta}_{K}^{(k,1)}$ (one per regressor), each with its standard error. Since `Car` was set as the baseline category, the output contains one such set of estimates for each of `Public Transit`, `Bicycle`, and `Walking`, and all estimates are interpreted relative to `Car`; the coefficients of the baseline category itself are fixed at zero for identifiability and therefore do not appear in the output. With $m = 4$ response categories and $K = 4$ regressors, the fitted model comprises $(m - 1) \times (K + 1) = 15$ estimated parameters, which is consistent with the parameter count derived above.

In the case study model with regressors $x_{i,\texttt{distance}}$, $x_{i,\texttt{age}}$, $Z_{i,\texttt{car}}$, and $Z_{i,\texttt{weekend}}$, the fitted model can be written as

$$
\log \left( \frac{\Pr(Y_i = k)}{\Pr(Y_i = \text{Car})} \right) = \hat{\beta}_{0}^{(k,\text{Car})} + \hat{\beta}_{1}^{(k,\text{Car})} x_{i,\texttt{distance}} + \hat{\beta}_{2}^{(k,\text{Car})} x_{i,\texttt{age}} + \hat{\beta}_{3}^{(k,\text{Car})} Z_{i,\texttt{car}} + \hat{\beta}_{4}^{(k,\text{Car})}Z_{i,\texttt{weekend}},
$$
where $k$ can be one of the non-baseline levels: `Public Transit`, `Bicycle`, or `Walking`.

In multinomial logistic regression, we model the probability of each outcome category given regressors. The key assumption is that we have enough data for each category to estimate the model parameters reliably. 
If one category is much more frequent than the others (i.e., unbalanced outcomes), the model may:

1. Bias coefficients toward the majority class because the likelihood function is dominated by majority observations.
2. Underestimate uncertainty for rare categories (small sample size → large variance).
3. Yield poor predictive performance for minority classes (the model tends to predict the majority class).
4. In extreme cases, parameters for rare categories may not converge or may give infinite estimates (complete separation).

For example, if 95\% of the observations are `Car`, then the model has little incentive to adjust parameters for `Walk` and `Bus`, so estimates for those categories can be unstable. There are two common approaches: 

1. Downsampling: Reduce the majority class size so that each class is roughly balanced. This risks throwing away data, but helps balance influence.
2. Upsampling: Duplicate (or synthetically generate) minority class cases until classes are balanced. This risks overfitting, but improves minority estimation.

Let's check the class counts of the outcome below.

::: {.panel-tabset}

## R Code

```{.r}
cat("Class counts before balancing:", "\n")
print(table(train_set$transport_mode))
```

## Python Code

```{.python}
print("Class counts before balancing:", "\n")
print(y.value_counts())
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
cat("Class counts before balancing:", "\n")
print(table(train_set$transport_mode))
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
print("Class counts before balancing:", "\n")
print(y.value_counts())
```

:::

To handle the imbalanced groups, we resample the training sets to balance out the groups to eliminate the bias in modelling. The fitted model on upsampled training set is below. The upsampling is performed once in `R` with `caret::upSample`, which retains every original training row and tops up only the classes below the majority size with duplicated rows sampled with replacement. The resulting `train_set_balanced` is then shared with `Python` through `reticulate`, so that both languages fit the model on the identical balanced training set and produce matching estimates. The upsampling can be performed in `Python` with `RandomOverSampler` from the imbalanced-learn (`imblearn`) package. For more details on the `RandomOverSampler` function, refer to its [documentation](https://imbalanced-learn.org/stable/references/generated/imblearn.over_sampling.RandomOverSampler.html).

::: {.panel-tabset}

## R Code

```{.r}
# Upsample the outcome classes
set.seed(352) # set a seed for sampling
train_set_balanced <- upSample(x = train_set[, c("age","commute_distance","has_car_available", "weekend_commuter")],
                  y = train_set$transport_mode,
                  yname = "transport_mode")
# Fit the multinomial logit regression
multinom_fit_balanced <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = train_set_balanced)
print(summary(multinom_fit_balanced))
```

## Python Code

```{.python}
# The upsampled training set is created once in R (caret::upSample; see the
# R tab) and pulled into Python via reticulate's `r` object, so both languages
# fit the model on the identical balanced training set.
train_set_balanced = r.train_set_balanced.reset_index(drop=True)

X_bal = train_set_balanced[["age", "commute_distance", "has_car_available", "weekend_commuter"]]
y_bal = train_set_balanced["transport_mode"]
X_bal = sm.add_constant(X_bal)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_bal_num = X_bal.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_bal_num[col] = X_bal_num[col].cat.codes

# Fit the multinomial logit regression
model_balanced = sm.MNLogit(y_bal, X_bal_num)
multinom_fit_balanced = model_balanced.fit()
print(multinom_fit_balanced.summary())
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
set.seed(352)
train_set_balanced <- upSample(x = train_set[, c("age","commute_distance","has_car_available", "weekend_commuter")],
                  y = train_set$transport_mode,
                  yname = "transport_mode")
multinom_fit_balanced <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = train_set_balanced)
print(summary(multinom_fit_balanced))
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
# The upsampled training set is created once in R (caret::upSample; see the
# R tab) and pulled into Python via reticulate's `r` object, so both languages
# fit the model on the identical balanced training set.
train_set_balanced = r.train_set_balanced.reset_index(drop=True)

X_bal = train_set_balanced[["age", "commute_distance", "has_car_available", "weekend_commuter"]]
y_bal = train_set_balanced["transport_mode"]
X_bal = sm.add_constant(X_bal)

X_bal_num = X_bal.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_bal_num[col] = X_bal_num[col].cat.codes

model_balanced = sm.MNLogit(y_bal, X_bal_num)
multinom_fit_balanced = model_balanced.fit()
print(multinom_fit_balanced.summary())
```

:::

## Goodness of Fit

Let us assess how well the fitted multinomial logistic regression explains and predicts the observed transport modes on the training set. We consider two complementary perspectives: **likelihood-based** measures that evaluate how much of the outcome's variation the model explains on the log-likelihood scale (similar to @sec-poisson-goodness-of-fit in [Classical Poisson Regression](https://alexrod61.github.io/regression-cookbook/book/10-classical-poisson.html)), and **classification-based** measures that evaluate how accurately the model assigns observations to their correct transport mode.

### Likelihood-Based Measures

The **null deviance** measures how well the intercept-only model (no regressors, only a baseline log-odds for each non-reference category) fits the data:

$$
D_{\text{null}} = -2\,\ell(\hat{\boldsymbol{\beta}}_{0,\text{null}}),
$$

where $\hat{\boldsymbol{\beta}}_{0,\text{null}}$ are the category-specific intercepts estimated without any regressor.

The **residual deviance** is the analogous quantity for the fitted model:

$$
D_{\text{resid}} = -2\,\ell(\hat{\boldsymbol{\beta}}).
$$

A well-fitting model should have a substantially smaller residual deviance than the null deviance. The reduction $D_{\text{null}} - D_{\text{resid}}$ is the likelihood-ratio (LR) statistic, which under $H_0$ (all slope coefficients across all non-baseline categories are zero) follows a $\chi^2$ distribution with degrees of freedom equal to the number of slope parameters $k$:

$$
G^2 = D_{\text{null}} - D_{\text{resid}} = -2\bigl(\ell_{\text{null}} - \ell_{\text{fitted}}\bigr) \;\overset{H_0}{\sim}\; \chi^2(k).
$$

**McFadden's pseudo-$R^2$** expresses the proportional reduction in log-likelihood relative to the null model:

$$
R^2_{\text{McFadden}} = 1 - \frac{\ell(\hat{\boldsymbol{\beta}})}{\ell(\hat{\boldsymbol{\beta}}_{0,\text{null}})}.
$$

Unlike the usual R-squared in OLS regression, McFadden's pseudo-R-squared should not be interpreted as the proportion of variance explained. Instead, it measures improvement in log-likelihood relative to the null model. Its values are typically much smaller than regular R-squared values, and values around $0.2$–$0.4$ are often considered indicative of a good/decent model fit in discrete-response models.

**AIC** and **BIC** scores penalize the log-likelihood for model complexity:

$$
\text{AIC} = -2\,\ell(\hat{\boldsymbol{\beta}}) + 2p, \qquad \text{BIC} = -2\,\ell(\hat{\boldsymbol{\beta}}) + p\log n,
$$

where $p$ is the total number of estimated parameters (the category-specific intercepts and slopes across all non-baseline categories) and $n$ is the sample size. Lower AIC and BIC scores indicate better fit, and they are most useful for comparing competing models (e.g. this imbalanced fit versus the upsampled fit).

We compute all of these measures below.

::: {.panel-tabset}

## R Code

```{.r}
library(nnet)

# Null model: intercept only, no regressors
multinom_null <- multinom(transport_mode ~ 1, data = train_set)

ll_fitted <- logLik(multinom_fit)
ll_null   <- logLik(multinom_null)

null_deviance  <- -2 * as.numeric(ll_null)
resid_deviance <- -2 * as.numeric(ll_fitted)
mcfadden_r2    <- 1 - as.numeric(ll_fitted) / as.numeric(ll_null)
lr_stat        <- as.numeric(-2 * (ll_null - ll_fitted))
lr_df          <- attr(ll_fitted, "df") - attr(ll_null, "df")
lr_pval        <- pchisq(lr_stat, df = lr_df, lower.tail = FALSE)

cat("Null deviance:      ", round(null_deviance,  2), "\n")
cat("Residual deviance:  ", round(resid_deviance, 2), "\n")
cat("McFadden R²:        ", round(mcfadden_r2, 4),   "\n")
cat("G²:                 ", round(lr_stat, 4),
    " (df =", lr_df, ", p =", formatC(lr_pval, format = "e", digits = 2), ")\n")
cat("AIC:                ", round(AIC(multinom_fit), 2), "\n")
cat("BIC:                ", round(BIC(multinom_fit), 2), "\n")
```

## Python Code

```{.python}
import numpy as np
from scipy.stats import chi2 as chi2dist

# The log-likelihood of the intercept-only (null) model is available
# directly on the fitted MNLogit result
ll_fitted = multinom_fit.llf
ll_null   = multinom_fit.llnull

null_deviance  = -2 * ll_null
resid_deviance = -2 * ll_fitted
mcfadden_r2    = 1 - ll_fitted / ll_null
lr_stat        = -2 * (ll_null - ll_fitted)
lr_df          = multinom_fit.df_model
lr_pval        = chi2dist.sf(lr_stat, lr_df)

print(f"Null deviance:      {null_deviance:.2f}")
print(f"Residual deviance:  {resid_deviance:.2f}")
print(f"McFadden R²:        {mcfadden_r2:.4f}")
print(f"G²:                 {lr_stat:.4f}  (df = {lr_df}, p = {lr_pval:.2e})")
print(f"AIC:                {multinom_fit.aic:.2f}")
print(f"BIC:                {multinom_fit.bic:.2f}")
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
library(nnet)

multinom_null <- multinom(transport_mode ~ 1, data = train_set)

ll_fitted <- logLik(multinom_fit)
ll_null   <- logLik(multinom_null)

null_deviance  <- -2 * as.numeric(ll_null)
resid_deviance <- -2 * as.numeric(ll_fitted)
mcfadden_r2    <- 1 - as.numeric(ll_fitted) / as.numeric(ll_null)
lr_stat        <- as.numeric(-2 * (ll_null - ll_fitted))
lr_df          <- attr(ll_fitted, "df") - attr(ll_null, "df")
lr_pval        <- pchisq(lr_stat, df = lr_df, lower.tail = FALSE)

cat("Null deviance:      ", round(null_deviance,  2), "\n")
cat("Residual deviance:  ", round(resid_deviance, 2), "\n")
cat("McFadden R²:        ", round(mcfadden_r2, 4),   "\n")
cat("G²:                 ", round(lr_stat, 4),
    " (df =", lr_df, ", p =", formatC(lr_pval, format = "e", digits = 2), ")\n")
cat("AIC:                ", round(AIC(multinom_fit), 2), "\n")
cat("BIC:                ", round(BIC(multinom_fit), 2), "\n")
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
import numpy as np
from scipy.stats import chi2 as chi2dist

ll_fitted = multinom_fit.llf
ll_null   = multinom_fit.llnull

null_deviance  = -2 * ll_null
resid_deviance = -2 * ll_fitted
mcfadden_r2    = 1 - ll_fitted / ll_null
lr_stat        = -2 * (ll_null - ll_fitted)
lr_df          = multinom_fit.df_model
lr_pval        = chi2dist.sf(lr_stat, lr_df)

print(f"Null deviance:      {null_deviance:.2f}")
print(f"Residual deviance:  {resid_deviance:.2f}")
print(f"McFadden R²:        {mcfadden_r2:.4f}")
print(f"G²:                 {lr_stat:.4f}  (df = {lr_df}, p = {lr_pval:.2e})")
print(f"AIC:                {multinom_fit.aic:.2f}")
print(f"BIC:                {multinom_fit.bic:.2f}")
```

:::

The residual deviance is smaller than the null deviance by $201.99$ units. This reduction is captured by the likelihood-ratio statistic $G^2 = 201.99$ on 12 degrees of freedom ($p < 0.001$), providing strong evidence that a model with age, commute distance, car availability, and weekend commuting status fits the training data better than the intercept-only model. McFadden's $R^2 = 0.156$ is still below the 0.2–0.4 range commonly associated with a good fit for discrete-response models, which indicates that these four regressors together explain only a modest share of the variation in `transport_mode` on the log-likelihood scale, even though their joint contribution is highly significant. The AIC ($1120.09$) and BIC ($1190.38$) reported here describe the imbalanced training model; they are most useful later for comparing this fit against the upsampled model or other model specifications.

### Classification-Based Measures

While likelihood-based measures evaluate the model on the probability scale, classification-based measures assess how well the model assigns each observation to its predicted transport mode by taking the category with the highest predicted probability as the point prediction.

Let's visualize and summarize goodness of fit for the train set. 

::: {.panel-tabset}

## R Code

```{.r}
y_pred <- predict(multinom_fit, newdata = train_set)
mode_levels <- c("Car", "Bicycle", "Public_Transit", "Walking")
conf_matrix <- confusionMatrix(factor(y_pred, levels = mode_levels),
                              factor(train_set$transport_mode, levels = mode_levels))

# Visualise
cm_df <- as.data.frame(as.table(conf_matrix$table))
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

# Order axes so the main diagonal (correct matches) runs top-left to
# bottom-right: Predicted increases left -> right (Car -> Walking), and
# Actual increases top -> bottom (Car -> Walking). ggplot draws the first
# factor level at the bottom of the y-axis, so we reverse Actual's levels to
# place "Car" at the top.
cm_df$Predicted <- factor(cm_df$Predicted, levels = levels(cm_df$Predicted))
cm_df$Actual    <- factor(cm_df$Actual,    levels = rev(levels(cm_df$Actual)))

cm_df |>
  ggplot(aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  labs(title = "Confusion Matrix (Training Set)",
       x = "Predicted label", y = "True label") +
  theme_bw(base_size = 13)
```

## Python Code

```{.python}
y_pred = multinom_fit.predict(X_numeric).idxmax(axis=1)

# Confusion matrix
cm = confusion_matrix(y.cat.codes, y_pred, labels=range(len(y.cat.categories)))
disp = ConfusionMatrixDisplay(confusion_matrix=cm,
                              display_labels=y.cat.categories)
disp.plot(cmap="Blues")
disp.ax_.set_title("Confusion Matrix (Training Set)")
fig = disp.figure_
fig.tight_layout()
fig
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| label: fig-confusion-matrix-train-R
#| fig-cap: "Confusion matrix of predicted versus actual transport modes on the training set."
#| echo: false
#| message: false
#| warning: false
y_pred <- predict(multinom_fit, newdata = train_set)
mode_levels <- c("Car", "Bicycle", "Public_Transit", "Walking")
conf_matrix <- confusionMatrix(factor(y_pred, levels = mode_levels),
                              factor(train_set$transport_mode, levels = mode_levels))

cm_df <- as.data.frame(as.table(conf_matrix$table))
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

# Order axes so the main diagonal (correct matches) runs top-left to
# bottom-right: Predicted increases left -> right (Car -> Walking), and
# Actual increases top -> bottom (Car -> Walking). ggplot draws the first
# factor level at the bottom of the y-axis, so we reverse Actual's levels to
# place "Car" at the top.
cm_df$Predicted <- factor(cm_df$Predicted, levels = levels(cm_df$Predicted))
cm_df$Actual    <- factor(cm_df$Actual,    levels = rev(levels(cm_df$Actual)))

cm_df |>
  ggplot(aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  labs(title = "Confusion Matrix (Training Set)",
       x = "Predicted label", y = "True label") +
  theme_bw(base_size = 13)
```

## Python Output

```{python}
#| label: fig-confusion-matrix-train-py
#| fig-cap: "Confusion matrix of predicted versus actual transport modes on the training set."
#| echo: false
#| message: false
#| warning: false
y_pred = multinom_fit.predict(X_numeric).idxmax(axis=1)

# Compute confusion matrix
cm = confusion_matrix(y.cat.codes, y_pred, labels=range(len(y.cat.categories)))
disp = ConfusionMatrixDisplay(confusion_matrix=cm,
                              display_labels=y.cat.categories)
disp.plot(cmap="Blues")
disp.ax_.set_title("Confusion Matrix (Training Set)")
fig = disp.figure_
fig.tight_layout()
fig
```

:::

You're encouraged to repeat these diagnostics on the balanced (upsampled) model and on the held‑out test set to compare whether balancing improves minority class recall without overly harming overall calibration.

Let's compute the metrics.

::: {.panel-tabset}

## R Code

```{.r}
# Classification predictions & confusion matrix
y_pred <- predict(multinom_fit, newdata = train_set)
conf_matrix <- confusionMatrix(factor(y_pred, levels = levels(train_set$transport_mode)),
                               factor(train_set$transport_mode, levels = levels(train_set$transport_mode)))
overall_accuracy <- conf_matrix$overall['Accuracy']
macro_recall <- mean(conf_matrix$byClass[,"Recall"], na.rm = TRUE)
macro_f1 <- mean(conf_matrix$byClass[,"F1"], na.rm = TRUE)
cat("Overall Accuracy:", round(overall_accuracy, 3), "\n")
cat("Macro Recall:", round(macro_recall, 3), "\n")
cat("Macro F1:", round(macro_f1, 3), "\n")
```

## Python Code

```{.python}
# Compute classification predictions & confusion matrix
probs = multinom_fit.predict(X_numeric)  # DataFrame of class probabilities
y_pred = probs.idxmax(axis=1)

# Compute the metrics
acc = accuracy_score(y.cat.codes, y_pred)
recall = recall_score(y.cat.codes, y_pred, average='macro')
# Average F1 only over the classes the model actually predicts, so a
# never-predicted class (undefined precision) is dropped rather than
# counted as 0 — matching caret's byClass F1 with na.rm = TRUE in the R tab.
predicted_labels = np.unique(y_pred)
f1 = f1_score(y.cat.codes, y_pred, labels=predicted_labels, average='macro')
print(f"Overall Accuracy: {acc:.3f}")
print(f"Macro Recall: {recall:.3f}")
print(f"Macro F1-score: {f1:.3f}")
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
y_pred <- predict(multinom_fit, newdata = train_set)
conf_matrix <- confusionMatrix(factor(y_pred, levels = levels(train_set$transport_mode)),
                               factor(train_set$transport_mode, levels = levels(train_set$transport_mode)))
overall_accuracy <- conf_matrix$overall['Accuracy']
macro_recall <- mean(conf_matrix$byClass[,"Recall"], na.rm = TRUE)
macro_f1 <- mean(conf_matrix$byClass[,"F1"], na.rm = TRUE)
cat("Overall Accuracy:", round(overall_accuracy, 3), "\n")
cat("Macro Recall:", round(macro_recall, 3), "\n")
cat("Macro F1:", round(macro_f1, 3), "\n")
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
probs = multinom_fit.predict(X_numeric)
y_pred = probs.idxmax(axis=1)
acc = accuracy_score(y.cat.codes, y_pred)
recall = recall_score(y.cat.codes, y_pred, average='macro')
# Average F1 only over the classes the model actually predicts, so a
# never-predicted class (undefined precision) is dropped rather than
# counted as 0 — matching caret's byClass F1 with na.rm = TRUE in the R tab.
predicted_labels = np.unique(y_pred)
f1 = f1_score(y.cat.codes, y_pred, labels=predicted_labels, average='macro')
print(f"Overall Accuracy: {acc:.3f}")
print(f"Macro Recall: {recall:.3f}")
print(f"Macro F1-score: {f1:.3f}")
```

:::

On the training set, the model achieves an overall accuracy of about **67.3\%**. On average, the model correctly identifies **33.1\%** of instances for each transport mode. Macro recall is relatively low, which suggests the model is biased to the majority classes, even though accuracy is decent. 
Macro F1 score is **65.6\%**, which means the model has a moderate balance of precision and recall on average per class. Macro F1 is much higher than macro recall, and this can happen if precision is high for the majority class, but recall for minority classes is low.
In summary, the model performs okay overall but is biased toward the dominant classes. It often fails to detect less frequent transport modes, which is why macro recall is low. Accuracy alone would overestimate the model’s effectiveness across all classes.
You are encouraged to explore the performance of the fitted model on the balanced sets. 

## Inference

After we finish estimating the coefficients, we look at how uncertain they are and whether they differ from zero. We get a variance estimate for each coefficient from the observed information matrix (the negative of the Hessian at the solution). Taking square roots of those variances gives standard errors. 
Similar as in [Poisson regression](https://alexrod61.github.io/regression-cookbook/book/10-classical-poisson.html), each coefficient divided by its standard error gives a (asymptotically normal) **Wald Z statistic** to test if the true effect is zero. Using the same standard errors we build confidence intervals: estimate $\pm$ (critical value $\times$ standard error). To make results easier to interpret, we often exponentiate a coefficient (or its confidence limits) to get an odds ratio—how the odds of choosing a category change for a one‑unit increase in the regressor, holding other variables fixed.

We can determine whether a regressor is statistically associated with the logarithm of the odds through hypothesis testing for the parameters $\beta^{(u,v)}_j$ by link function. Define the Wald statistic $z^{(u,v)}_j$ as:
$$
z_j^{(u, v)} = \frac{\hat{\beta}_j^{(u, v)}}{\mbox{se}\left(\hat{\beta}_j^{(u, v)}\right)}
$$ {#eq-wald-statistic}
and we will test the following hypotheses: 
$$
\begin{gather*}
H_0: \beta_j^{(u, v)} = 0\\
H_a: \beta_j^{(u, v)} \neq 0.
\end{gather*}
$$
Provided that the sample size $n$ is large enough, $z_j$ has an approximately Standard Normal distribution under $H_0$.

The corresponding $p$-values for each $\beta^{(u,v)}_j$ can be computed. The smaller the $p$-value, the stronger is the evidence against the null hypothesis $H_0$. As in the previous regression models, we would set a predetermined significance level  (usually taken to be $\alpha=0.05$) to infer if the $p$-value is small enough. If the $p$-value is smaller than the predetermined level $\alpha$, then you could claim that there is evidence to reject the null hypothesis. Hence, $p$-values that are small enough indicate that the data provides evidence in favour of association (or causation in the case of an experimental study!) between the response variable and the $j$th regressor.

Given a specified level of confidence where $\alpha$ is the significance level, we can construct approximate $(1-\alpha)\times 100\%$ confidence intervals for the corresponding true value of $\beta^{(u,v)}_j$:
$$
\begin{equation*}
\hat{\beta}_j^{(u, v)} \pm z_{\alpha/2}\mbox{se} \left( \hat{\beta}_j^{(u, v)} \right),
\end{equation*}
$$ {#eq-ci}
where $z_{\alpha/2}$ is the upper $\alpha/2$ quantile of the Standard Normal distribution.

### Fitting the Model

Now we will fit the model on the test set using the `nnet` package in `R` or the `statsmodel` package in `Python`. We will also compute the corresponding 95\% confidence intervals below and filter the statistically significant coefficients at the 5\% significance level.

::: {.panel-tabset}

## R Code

```{.r}
# Set "Car" as the baseline level
test_set$transport_mode <- relevel(test_set$transport_mode, ref = "Car")

# Fit multinomial logistic regression model on test set
multinom_test <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = test_set)

# Print the fitted model
print(summary(multinom_test))

# Significant coefficients only
cat("\nSignificant coefficients (5% threshold):\n")

broom::tidy(multinom_test, conf.int = TRUE, exponentiate = TRUE) |>
  dplyr::mutate_if(is.numeric, round, 3) |>
  dplyr::filter(p.value < 0.05) |>          # keep only significant coefficients
  rename(outcome_class = y.level) |>
  print(n = Inf, width = Inf)               # show full term names without truncation
```

## Python Code

```{.python}
# Factorize the response and set the baseline model
test_set["transport_mode"] = pd.Categorical(
    test_set["transport_mode"],
    categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)

# Drop outcome categories with no observations in this set. A class present in
# the full data may be absent from the smaller test set; statsmodels cannot
# estimate coefficients for an empty category and would otherwise return NaN.
test_set["transport_mode"] = test_set["transport_mode"].cat.remove_unused_categories()

# Design matrix and response
y = test_set["transport_mode"]
X = test_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]

# Add intercept
X = sm.add_constant(X)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_numeric = X.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes

# Fit multinomial logistic regression
model = sm.MNLogit(y, X_numeric)

# Fit with BFGS rather than the default Newton-Raphson method. Under
# quasi-complete separation the Newton-Raphson iterations overflow the
# log-likelihood and Hessian, returning NaN for every coefficient. If
# the BFGS solution itself overflows the observed information matrix,
# retry with fewer iterations; the likelihood is essentially flat along
# the separated directions, so stopping earlier sacrifices nothing.
for max_iterations in (1000, 200, 100, 50, 25, 10):
    multinom_test = model.fit(method="bfgs", maxiter=max_iterations, disp=False)
    information = -multinom_test.model.hessian(
        multinom_test.params.values.ravel(order="F"))
    if np.all(np.isfinite(information)):
        break

# statsmodels declines to invert an observed information matrix that is
# not strictly positive definite, which is exactly what quasi-complete
# separation produces: the log-likelihood is flat along the separated
# directions, so the corresponding eigenvalues are numerically zero. In
# that case, invert the information matrix with its eigenvalues floored
# at a small positive value. Directions in which the data carry
# essentially no information then receive very large variances, so the
# separated terms are reported with enormous standard errors and
# p-values near one instead of aborting the computation. The check must
# bypass the results wrapper via ._results, because the wrapper
# converts a missing covariance into a data frame of NaNs.
if multinom_test._results.normalized_cov_params is None:
    eigenvalues, eigenvectors = np.linalg.eigh(information)
    eigenvalues = np.maximum(eigenvalues, 1e-8)
    multinom_test._results.normalized_cov_params = (
        eigenvectors @ np.diag(1.0 / eigenvalues) @ eigenvectors.T)

# Print the fitted model
print(multinom_test.summary())

data_dict = {
    "estimate": np.exp(multinom_test.params),      # odds ratios
    "pvalue": multinom_test.pvalues,
    "std_err": multinom_test.bse,
    "statistic": multinom_test.tvalues
}

# Outcome class mapping. With "Car" as the baseline, the remaining (non-empty)
# categories are the comparison classes, in categorical order.
class_map = dict(enumerate(y.cat.categories[1:]))

# Container for processed tidy tables
tidy_dict = {}

for name, df in data_dict.items():
    tidy = (
        df.stack()               # long form
        .swaplevel()             # variable first, outcome second
        .reset_index()           # convert MultiIndex to columns
        .rename(columns={"level_0": "outcome_class", 
                         "level_1": "term", 
                         0: name})
    )
    tidy["outcome_class"] = tidy["outcome_class"].replace(class_map)
    tidy = tidy.sort_values(by="outcome_class").reset_index(drop=True)
    
    tidy_dict[name] = tidy

# CIs
conf = multinom_test.conf_int()
or_conf = np.exp(conf)
ci_lower_stacked = or_conf["lower"].reset_index()
ci_lower_stacked.columns = ['outcome_class', 'term', 'ci_lower']
ci_upper_stacked = or_conf["upper"].reset_index()
ci_upper_stacked.columns = ['outcome_class', 'term', 'ci_upper']

# Merge the result tables
ref_cols = ["outcome_class", "term"]
results = tidy_dict['estimate'] \
    .merge(tidy_dict['pvalue'], on=ref_cols) \
    .merge(tidy_dict['std_err'], on=ref_cols) \
    .merge(tidy_dict['statistic'], on=ref_cols) \
    .merge(ci_lower_stacked, on=ref_cols) \
    .merge(ci_upper_stacked, on=ref_cols)

# Keep only coefficients significant at the 5% level
results_sig = results[results["pvalue"] < 0.05].round(3)
print("\nSignificant coefficients (5% threshold):")
print(results_sig)
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
# Set "Car" as the baseline level
test_set$transport_mode <- relevel(test_set$transport_mode, ref = "Car")

# Fit multinomial logistic regression model on test set
multinom_test <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = test_set)

# Print the fitted model
print(summary(multinom_test))

# Significant coefficients only
cat("\nSignificant coefficients (5% threshold):\n")

broom::tidy(multinom_test, conf.int = TRUE, exponentiate = TRUE) |>
  dplyr::mutate_if(is.numeric, round, 3) |>
  dplyr::filter(p.value < 0.05) |>          # keep only significant coefficients
  rename(outcome_class = y.level) |>
  print(n = Inf, width = Inf)               # show full term names without truncation
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
# Factorize the response and set the baseline model
test_set["transport_mode"] = pd.Categorical(
    test_set["transport_mode"],
    categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)

# Drop outcome categories with no observations in this set. A class present in
# the full data may be absent from the smaller test set; statsmodels cannot
# estimate coefficients for an empty category and would otherwise return NaN.
test_set["transport_mode"] = test_set["transport_mode"].cat.remove_unused_categories()

# Design matrix and response
y = test_set["transport_mode"]
X = test_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]

# Add intercept
X = sm.add_constant(X)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_numeric = X.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes

# Fit multinomial logistic regression
model = sm.MNLogit(y, X_numeric)

# Fit with BFGS rather than the default Newton-Raphson method. Under
# quasi-complete separation the Newton-Raphson iterations overflow the
# log-likelihood and Hessian, returning NaN for every coefficient. If
# the BFGS solution itself overflows the observed information matrix,
# retry with fewer iterations; the likelihood is essentially flat along
# the separated directions, so stopping earlier sacrifices nothing.
for max_iterations in (1000, 200, 100, 50, 25, 10):
    multinom_test = model.fit(method="bfgs", maxiter=max_iterations, disp=False)
    information = -multinom_test.model.hessian(
        multinom_test.params.values.ravel(order="F"))
    if np.all(np.isfinite(information)):
        break

# statsmodels declines to invert an observed information matrix that is
# not strictly positive definite, which is exactly what quasi-complete
# separation produces: the log-likelihood is flat along the separated
# directions, so the corresponding eigenvalues are numerically zero. In
# that case, invert the information matrix with its eigenvalues floored
# at a small positive value. Directions in which the data carry
# essentially no information then receive very large variances, so the
# separated terms are reported with enormous standard errors and
# p-values near one instead of aborting the computation. The check must
# bypass the results wrapper via ._results, because the wrapper
# converts a missing covariance into a data frame of NaNs.
if multinom_test._results.normalized_cov_params is None:
    eigenvalues, eigenvectors = np.linalg.eigh(information)
    eigenvalues = np.maximum(eigenvalues, 1e-8)
    multinom_test._results.normalized_cov_params = (
        eigenvectors @ np.diag(1.0 / eigenvalues) @ eigenvectors.T)

# Print the fitted model
print(multinom_test.summary())

data_dict = {
    "estimate": np.exp(multinom_test.params),      # odds ratios
    "pvalue": multinom_test.pvalues,
    "std_err": multinom_test.bse,
    "statistic": multinom_test.tvalues
}

# Outcome class mapping. With "Car" as the baseline, the remaining (non-empty)
# categories are the comparison classes, in categorical order.
class_map = dict(enumerate(y.cat.categories[1:]))

# Container for processed tidy tables
tidy_dict = {}

for name, df in data_dict.items():
    tidy = (
        df.stack()               # long form
        .swaplevel()             # variable first, outcome second
        .reset_index()           # convert MultiIndex to columns
        .rename(columns={"level_0": "outcome_class", 
                         "level_1": "term", 
                         0: name})
    )
    tidy["outcome_class"] = tidy["outcome_class"].replace(class_map)
    tidy = tidy.sort_values(by="outcome_class").reset_index(drop=True)
    
    tidy_dict[name] = tidy

# CIs
conf = multinom_test.conf_int()
or_conf = np.exp(conf)
ci_lower_stacked = or_conf["lower"].reset_index()
ci_lower_stacked.columns = ['outcome_class', 'term', 'ci_lower']
ci_upper_stacked = or_conf["upper"].reset_index()
ci_upper_stacked.columns = ['outcome_class', 'term', 'ci_upper']

# Merge the result tables
ref_cols = ["outcome_class", "term"]
results = tidy_dict['estimate'] \
    .merge(tidy_dict['pvalue'], on=ref_cols) \
    .merge(tidy_dict['std_err'], on=ref_cols) \
    .merge(tidy_dict['statistic'], on=ref_cols) \
    .merge(ci_lower_stacked, on=ref_cols) \
    .merge(ci_upper_stacked, on=ref_cols)

# Keep only coefficients significant at the 5% level
results_sig = results[results["pvalue"] < 0.05].round(3)
print("\nSignificant coefficients (5% threshold):")
print(results_sig)
```

:::

On the test set, commute distance and the availability of cars are significant predictors of choosing public transit over driving at a significance threshold of $5\%$. The remaining rows in the table of significant coefficients in the `R` output should be read with caution. The test set contains only one `Walking` observation and four `Bicycle` observations, and none of these $5$ observations are weekend commuters; therefore, the `Walking` equation suffers from quasi-complete separation. The R output consequently flags these coefficients as significant, with implausible odds ratios (essentially zero, or an odds ratio near 20 per year of age for `Walking`), standard errors of essentially zero, and extreme test statistics. On the other hand, `statsmodels` handles the separation differently by assigning the separated terms very large standard errors, and yielding $p$-values close to $1$, so that they are correctly excluded from the significant coefficients in the `Python` output. Note that the `Python` model is fitted with the BFGS optimizer rather than the default Newton-Raphson method. Under quasi-complete separation the maximum likelihood estimates for the separated terms do not exist, and the Newton-Raphson iterations push these coefficients toward infinity until the log-likelihood and the Hessian overflow, at which point every entry of the summary table is reported as `nan`. BFGS does not rely on inverting the Hessian during the iterations and instead settles at finite, albeit extreme, estimates for the separated terms, which yields the very large standard errors and near-one $p$-values described above. In addition, `statsmodels` declines to invert an observed information matrix that is not strictly positive definite, and separation produces exactly that: the flat directions of the likelihood correspond to eigenvalues that are numerically zero, so the covariance matrix of the estimates is unavailable and any request for standard errors raises an error. We therefore stop the optimizer early enough that the information matrix remains finite, and invert it with its eigenvalues floored at a small positive value, so that the directions in which the data carry essentially no information receive very large variances rather than aborting the computation.

### Interpreting the Coefficients

The multinomial logistic regression model fitted on the test set yields the estimated coefficients above. Note that the estimates and confidence limits in the table of significant coefficients are exponentiated, so each entry is an odds ratio relative to the baseline category `Car`. A positive value of $\hat{\beta}_j^{(k,\texttt{Car})}$ corresponds to higher values of regressor $j$ being associated with a higher odds of category relative $k$ to `Car`, whereas a negative value of $\hat{\beta}_j^{(k,\texttt{Car})}$ corresponds to lower odds of the category $k$ relative to `Car` for higher values of regressor $j$. The significant regression coefficients belong to the `Public Transit` equation, and we will interpret them below. 

**Commute distance:** Commute distance is a significant predictor of choosing public transit over the car ($\text{OR} = 1.09$, $z = 3.51$, $p < 0.001$, 95% CI: $1.04, 1.14$). It has the following interpretation.

> *Holding age, car availability, and weekend commuting status constant, each additional kilometre of one-way commute distance is associated with 1.09 times higher odds of choosing public transit over car.*

Although the per-kilometre effect appears modest, it compounds over realistic differences in commute length: a commute that is $10$ kilometres longer multiplies the odds of choosing public transit over the car by approximately $1.09^{10} \approx 2.4$.

**Car availability:** Car availability is a significant predictor of choosing public transit over the car ($\text{OR} = 0.193$, $z = -4.79$, $p < 0.001$, 95% CI: $0.098, 0.378$). It has the following interpretation.

> *Holding commute distance, age, and weekend commuting status constant, individuals with a car available have 0.193 times the odds of choosing public transit over the car, relative to individuals without a car available; that is, car availability is associated with a reduction of approximately 81% in these odds.*

The remaining rows of the significant coefficients table in the `R` output are artifacts of the quasi-complete separation discussed above and are therefore not interpreted.

## Results

The fitted model can be written as 
$$
\log \left( \frac{\Pr(Y_i = k)}{\Pr(Y_i = \text{Car})} \right) = \hat{\beta}_{0}^{(k,\text{Car})} + \hat{\beta}_{1}^{(k,\text{Car})} x_{i,\texttt{distance}} + \hat{\beta}_{2}^{(k,\text{Car})} x_{i,\texttt{age}} + \hat{\beta}_{3}^{(k,\text{Car})}Z_{i,\texttt{car}} + \hat{\beta}_{4}^{(k,\text{Car})} Z_{i,\texttt{weekend}},
$$ {#eq-fitted-model}
where $k$ is one of the non-baseline levels: `Bicycle`, `Public Transit`, or `Walking`, and the regressors are denoted as $x_{i,\texttt{distance}}, x_{i,\texttt{age}}, Z_{i,\texttt{car}},$ and $Z_{i,\texttt{weekend}}$.

The estimated intercepts and slope coefficients of the multinomial logistic regression model fitted on the test set in the Inference section are summarized below, grouped by non-baseline category. Every estimate is interpreted relative to the baseline category `Car`, and the $z$ statistics, $p$-values, odds ratios, and $95\%$ confidence intervals are computed from the estimates and standard errors reported in the `R` output:

| Parameter | Estimate | Std. Error | $z$ | $p$-value | OR | 95% CI |
|-----------|----------|------------|-----|-----------|-----|--------|
| $\hat{\beta}_{0}^{(\text{Bicycle},\,\text{Car})}$ (intercept) | $-3.520$ | $2.095$ | $-1.68$ | $0.093$ | $0.030$ | $(0.000, 1.797)$ |
| $\hat{\beta}_{1}^{(\text{Bicycle},\,\text{Car})}$ (`commute_distance`) | $-0.110$ | $0.096$ | $-1.14$ | $0.252$ | $0.896$ | $(0.743, 1.081)$ |
| $\hat{\beta}_{2}^{(\text{Bicycle},\,\text{Car})}$ (`age`) | $0.029$ | $0.038$ | $0.76$ | $0.446$ | $1.029$ | $(0.956, 1.108)$ |
| $\hat{\beta}_{3}^{(\text{Bicycle},\,\text{Car})}$ (`has_car_availableYes`) | $-0.405$ | $1.192$ | $-0.34$ | $0.734$ | $0.667$ | $(0.065, 6.900)$ |
| $\hat{\beta}_{4}^{(\text{Bicycle},\,\text{Car})}$ (`weekend_commuterYes`) | $1.173$ | $1.251$ | $0.94$ | $0.349$ | $3.231$ | $(0.278, 37.510)$ |
| $\hat{\beta}_{0}^{(\text{Public\_Transit},\,\text{Car})}$ (intercept) | $-0.386$ | $0.617$ | $-0.63$ | $0.532$ | $0.680$ | $(0.203, 2.278)$ |
| $\hat{\beta}_{1}^{(\text{Public\_Transit},\,\text{Car})}$ (`commute_distance`) | $0.082$ | $0.024$ | $3.51$ | $< 0.001$ | $1.086$ | $(1.037, 1.137)$ |
| $\hat{\beta}_{2}^{(\text{Public\_Transit},\,\text{Car})}$ (`age`) | $-0.004$ | $0.012$ | $-0.37$ | $0.709$ | $0.996$ | $(0.973, 1.019)$ |
| $\hat{\beta}_{3}^{(\text{Public\_Transit},\,\text{Car})}$ (`has_car_availableYes`) | $-1.646$ | $0.343$ | $-4.79$ | $< 0.001$ | $0.193$ | $(0.098, 0.378)$ |
| $\hat{\beta}_{4}^{(\text{Public\_Transit},\,\text{Car})}$ (`weekend_commuterYes`) | $0.575$ | $0.486$ | $1.18$ | $0.236$ | $1.778$ | $(0.686, 4.608)$ |
| $\hat{\beta}_{0}^{(\text{Walking},\,\text{Car})}$ (intercept) | $-562.911$ | $0.004$ | $-1.61 \times 10^{5}$ | $< 0.001$ | $\approx 0$ | $(\approx 0, \approx 0)$ |
| $\hat{\beta}_{1}^{(\text{Walking},\,\text{Car})}$ (`commute_distance`) | $-0.406$ | $0.825$ | $-0.49$ | $0.623$ | $0.667$ | $(0.132, 3.357)$ |
| $\hat{\beta}_{2}^{(\text{Walking},\,\text{Car})}$ (`age`) | $8.880$ | $0.144$ | $61.51$ | $< 0.001$ | $7190$ | $(5418, 9541)$ |
| $\hat{\beta}_{3}^{(\text{Walking},\,\text{Car})}$ (`has_car_availableYes`) | $-24.968$ | $1.23 \times 10^{-5}$ | $-2.04 \times 10^{6}$ | $< 0.001$ | $\approx 0$ | $(\approx 0, \approx 0)$ |
| $\hat{\beta}_{4}^{(\text{Walking},\,\text{Car})}$ (`weekend_commuterYes`) | $-4.396$ | $6.74 \times 10^{-16}$ | $-6.53 \times 10^{15}$ | $< 0.001$ | $0.012$ | $(0.012, 0.012)$ |

The `Bicycle` and `Public Transit` blocks are estimated on adequate numbers of observations, and within these blocks only `commute_distance` and `has_car_availableYes` in the `Public Transit` equation are significant at the $5\%$ level, in agreement with the table of significant coefficients in the Inference section. The `Walking` block should not be interpreted: as discussed in the Inference section, the `Walking` equation suffers from quasi-complete separation, and its extreme estimates, essentially zero standard errors, and astronomical $z$ statistics are artifacts of a maximum likelihood estimate that does not exist rather than genuine effects.

### Prediction 

Let's fit the model on the training set, and predict on the held-out test set. The confusion matrix and performance metrics are reported below.

::: {.panel-tabset}

## R Code

```{.r}
# Fit the model on the training set
multinom_fit <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = train_set)

# Predict on the test set using the fitted model
y_pred <- predict(multinom_fit, newdata = test_set)

# Compute confusion matrix
mode_levels <- c("Car", "Bicycle", "Public_Transit", "Walking")
conf_matrix <- confusionMatrix(factor(y_pred, levels = mode_levels),
                              factor(test_set$transport_mode, levels = mode_levels))

# Visualise
cm_df <- as.data.frame(as.table(conf_matrix$table))
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

# Order axes so the main diagonal (correct matches) runs top-left to
# bottom-right: Predicted increases left -> right (Car -> Walking), and
# Actual increases top -> bottom (Car -> Walking). ggplot draws the first
# factor level at the bottom of the y-axis, so we reverse Actual's levels to
# place "Car" at the top.
cm_df$Predicted <- factor(cm_df$Predicted, levels = levels(cm_df$Predicted))
cm_df$Actual    <- factor(cm_df$Actual,    levels = rev(levels(cm_df$Actual)))

cm_df |>
  ggplot(aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  labs(title = "Confusion Matrix (Test Set)",
       x = "Predicted label", y = "True label") +
  theme_bw(base_size = 13)
```

## Python Code

```{.python}
# Factorize the response and set the baseline model
train_set["transport_mode"] = pd.Categorical(
    train_set["transport_mode"],
    categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)

# Design matrix and response
y = train_set["transport_mode"]
X = train_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]

# Add intercept
X = sm.add_constant(X)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_numeric = X.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes

# Fit multinomial logistic regression
model = sm.MNLogit(y, X_numeric)
multinom_fit = model.fit()

# Predict on the test set using the fitted model
test_set["transport_mode"] = pd.Categorical(
  test_set["transport_mode"],
  categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)
y_test = test_set["transport_mode"]
X_test = test_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]
X_test = sm.add_constant(X_test)
X_test_num = X_test.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_test_num[col] = X_test_num[col].cat.codes
y_pred_test = multinom_fit.predict(X_test_num).idxmax(axis=1)

# Confusion matrix
labels = y_test.cat.categories
cm = confusion_matrix(y_test.cat.codes, y_pred_test, labels=range(len(labels)))
disp = ConfusionMatrixDisplay(confusion_matrix=cm,
                              display_labels=y_test.cat.categories)
disp.plot(cmap="Blues")
disp.ax_.set_title("Confusion Matrix (Test Set)")
fig = disp.figure_
fig.tight_layout()
fig
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| label: fig-confusion-matrix-test-R
#| fig-cap: "Confusion matrix of predicted versus actual transport modes on the held-out test set."
#| echo: false
#| message: false
#| warning: false

# Fit the model on the training set
multinom_fit <- multinom(transport_mode ~ age + commute_distance + has_car_available + weekend_commuter, data = train_set)

y_pred <- predict(multinom_fit, newdata = test_set)
mode_levels <- c("Car", "Bicycle", "Public_Transit", "Walking")
conf_matrix <- confusionMatrix(factor(y_pred, levels = mode_levels),
                              factor(test_set$transport_mode, levels = mode_levels))

cm_df <- as.data.frame(as.table(conf_matrix$table))
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

# Order axes so the main diagonal (correct matches) runs top-left to
# bottom-right: Predicted increases left -> right (Car -> Walking), and
# Actual increases top -> bottom (Car -> Walking). ggplot draws the first
# factor level at the bottom of the y-axis, so we reverse Actual's levels to
# place "Car" at the top.
cm_df$Predicted <- factor(cm_df$Predicted, levels = levels(cm_df$Predicted))
cm_df$Actual    <- factor(cm_df$Actual,    levels = rev(levels(cm_df$Actual)))

cm_df |>
  ggplot(aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  labs(title = "Confusion Matrix (Test Set)",
       x = "Predicted label", y = "True label") +
  theme_bw(base_size = 13)
```

## Python Output

```{python}
#| label: fig-confusion-matrix-test-py
#| fig-cap: "Confusion matrix of predicted versus actual transport modes on the held-out test set."
#| echo: false
#| message: false
#| warning: false

# Factorize the response and set the baseline model
train_set["transport_mode"] = pd.Categorical(
    train_set["transport_mode"],
    categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)

# Design matrix and response
y = train_set["transport_mode"]
X = train_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]

# Add intercept
X = sm.add_constant(X)

# Convert categorical variables to numeric codes for statsmodels MNLogit
X_numeric = X.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_numeric[col] = X_numeric[col].cat.codes

# Fit multinomial logistic regression
model = sm.MNLogit(y, X_numeric)
multinom_fit = model.fit()

test_set["transport_mode"] = pd.Categorical(
  test_set["transport_mode"], 
  categories=["Car", "Bicycle", "Public_Transit", "Walking"],
    ordered=False
)
y_test = test_set["transport_mode"]
X_test = test_set[["age", "commute_distance", "has_car_available", "weekend_commuter"]]
X_test = sm.add_constant(X_test)
X_test_num = X_test.copy()
for col in ['has_car_available', 'weekend_commuter']:
    X_test_num[col] = X_test_num[col].cat.codes
y_pred_test = multinom_fit.predict(X_test_num).idxmax(axis=1)

# Confusion matrix
labels = y_test.cat.categories
cm = confusion_matrix(y_test.cat.codes, y_pred_test, labels=range(len(labels)))
disp = ConfusionMatrixDisplay(confusion_matrix=cm,
                              display_labels=y_test.cat.categories)
disp.plot(cmap="Blues")
disp.ax_.set_title("Confusion Matrix (Test Set)")
fig = disp.figure_
fig.tight_layout()
fig
```

:::

::: {.panel-tabset}

## R Code

```{.r}
# Compute the metrics
overall_accuracy <- conf_matrix$overall['Accuracy']
macro_recall <- mean(conf_matrix$byClass[,"Recall"], na.rm = TRUE)
macro_f1 <- mean(conf_matrix$byClass[,"F1"], na.rm = TRUE)
cat("Overall Accuracy:", round(overall_accuracy, 3), "\n")
cat("Macro Recall:", round(macro_recall, 3), "\n")
cat("Macro F1:", round(macro_f1, 3), "\n")
```

## Python Code

```{.python}
# Compute the metrics
acc = accuracy_score(y_test.cat.codes, y_pred_test)
recall = recall_score(y_test.cat.codes, y_pred_test, average='macro')
# Average F1 only over the classes the model actually predicts, so a
# never-predicted class (undefined precision) is dropped rather than
# counted as 0 — matching caret's byClass F1 with na.rm = TRUE in the R tab.
predicted_labels = np.unique(y_pred_test)
f1 = f1_score(y_test.cat.codes, y_pred_test, labels=predicted_labels, average='macro')
print(f"Overall Accuracy: {acc:.3f}")
print(f"Macro Recall: {recall:.3f}")
print(f"Macro F1-score: {f1:.3f}")
```

:::

::: {.panel-tabset}

## R Output

```{r}
#| echo: false
#| message: false
#| warning: false
overall_accuracy <- conf_matrix$overall['Accuracy']
macro_recall <- mean(conf_matrix$byClass[,"Recall"], na.rm = TRUE)
macro_f1 <- mean(conf_matrix$byClass[,"F1"], na.rm = TRUE)
cat("Overall Accuracy:", round(overall_accuracy, 3), "\n")
cat("Macro Recall:", round(macro_recall, 3), "\n")
cat("Macro F1:", round(macro_f1, 3), "\n")
```

## Python Output

```{python}
#| echo: false
#| message: false
#| warning: false
acc = accuracy_score(y_test.cat.codes, y_pred_test)
recall = recall_score(y_test.cat.codes, y_pred_test, average='macro')
# Average F1 only over the classes the model actually predicts, so a
# never-predicted class (undefined precision) is dropped rather than
# counted as 0 — matching caret's byClass F1 with na.rm = TRUE in the R tab.
predicted_labels = np.unique(y_pred_test)
f1 = f1_score(y_test.cat.codes, y_pred_test, labels=predicted_labels, average='macro')
print(f"Overall Accuracy: {acc:.3f}")
print(f"Macro Recall: {recall:.3f}")
print(f"Macro F1-score: {f1:.3f}")
```

:::

On the test set, the model achieves an overall accuracy of **67.6\%**, a macro recall of **33.5\%**, and a macro F1 of **66.5\%**. 

The `Car` and `Public Transit` categories are predicted most reliably with recalls of $ $ and $ $ respectively. The `Bicycle` and `Walking` observations in the test set are all misclassified as `Car` or `Public Transit`.

## Storytelling

To complete the data science workflow, let's step back from the coefficients, tests, and confusion matrices, and focus on how the results can be interpreted and applied in practice. This final stage, storytelling, is where statistical output is translated into insights that support real world decision making.

In this case study, the Multinomial Logistic regression model helps us understand how commute distance, age, car availability, and weekend commuting status relate to an individual's primary mode of transport, an outcome with four unordered categories: `Car`, `Public Transit`, `Bicycle`, and `Walking`. We will distill the key takeaways from the analysis and discuss what they imply for commuters, transportation planners, and policy makers.

### Key Insights From the Model

Car availability and commute distance are the two most consistently significant regressors across both the imbalanced and upsampled models. Having access to a car is associated with markedly lower odds of using public transit or walking instead of driving; the corresponding effect for cycling is not statistically significant, so car access alone does not appear to explain the choice to bike in this dataset. Commute distance moves the odds of public transit and cycling in opposite directions relative to driving, such that each additional kilometre increases the odds of choosing public transit over the car, while decreasing the odds of cycling over the car. This matches the intuition that the car and public transit become the practical options once a commute gets long, while cycling remains attractive mainly over short distances.

Age and weekend commuting status tell a more conditional story. On the original, imbalanced training data, neither regressor reaches statistical significance for most categories, largely because the sample contains too few cyclists and walkers to estimate their effects precisely. Once the minority classes are upsampled, both regressors become significant predictors of `Walking`, and age additionally becomes significant for `Bicycle`: older individuals and weekend commuters are more likely to walk than drive, once the class imbalance no longer masks the effect. This is a useful reminder that failing to reject $H_0$ is not the same as confirming $H_0$ because a lack of significance can potentially reflect insufficient data rather than a genuinely null effect for rare categories

Some estimates need a caveat rather than a substantive interpretation. The test set contains only one `Walking` observation and five `Bicycle` observations, none of the latter being weekend commuters, so the test set model suffers from complete separation. In R, the `weekend_commuter` effect for `Bicycle` and several `Walking` coefficients are flagged as significant with implausibly large or near-zero odds ratios, standard errors of essentially zero, and extreme test statistics. In Python, `statsmodels` assigns the same terms enormous standard errors and $p$-values near 1, and reports that the fit did not fully converge. None of this is evidence about the true effects; the terms are simply not identifiable from the available data. Analysts should flag this kind of estimate rather than interpret it, and should be cautious of near-zero or implausibly large odds ratios whenever a category is small.

As a predictive tool, the model favours the categories it has seen the most of. The confusion matrices for both the training and test sets show noticeably stronger performance for `Car` and `Public Transit`, the two dominant categories, than for `Bicycle` and `Walking`. Upsampling helps the model learn the minority classes better during estimation, but it cannot manufacture additional information at prediction time, so macro-averaged recall and F1 remain modest even after balancing. The model is best used as a tool for understanding how commute distance, car availability, age, and weekend habits relate to travel mode choice, rather than as a precise classifier for the least common commuting behaviours.

::: {.Tip}
::::{.Tip-header}
Using the Results Responsibly
::::
::::{.Tip-container}
It is important to remember that this analysis is associational, not causal. The model does not prove that changing someone's commute distance or car access would, on its own, change how they travel; it shows that, in this dataset, these regressors are consistently linked to the mode people choose once other measured factors are accounted for.

Nevertheless, the results provide a useful evidence-based starting point for discussion. They help identify which regressors are most closely linked to travel mode choice, and which associations, such as age and weekend commuting for the rarer categories, may only become visible with more balanced or more abundant data. For transportation planners, this kind of insight can help prioritize where infrastructure investment, such as bike lanes or improved transit access, is likely to shift commuting patterns the most.
::::
:::
